### 2022-07-04 general_sholokhov.tex
Import thesis proposal from the general exam as a starting point for my PhD thesis

\documentclass[11pt,letterpaper]{article}

% PRESETS OF PACKAGES AND MATH %
\input{latex-macros/useful_packages}
\input{latex-macros/math_macros}

%%%%% USER CONFIGURATION %%%%%
\newcommand{\userName}{Aleksei Sholokhov}
\newcommand{\userId}{Lesha}
\newcommand{\department}{Department of Applied Mathematics}
\newcommand{\institution}{University of Washington}
\newcommand{\projectNameShort}{Feature Selection for Mixed Models}
\newcommand{\doctitle}{Thesis Proposal}
\newcommand{\projectNameLong}{}

% Cast
\newcommand{\Peng}[1][]{\textbf{Peng\ifthenelse{\equal{#1}{}}{ }{ (#1):}}}
\newcommand{\Sasha}[1][]{\textbf{Sasha\ifthenelse{\equal{#1}{}}{ }{ (#1):}}}
\newcommand{\Lesha}[1][]{\textbf{\userId\ifthenelse{\equal{#1}{}}{ }{ (#1):}}}
\newcommand{\Brad}[1][]{\textbf{Brad\ifthenelse{\equal{#1}{}}{ }{ (#1):}}}
\newcommand{\Jim}[1][]{\textbf{Jim\ifthenelse{\equal{#1}{}}{ }{ (#1):}}}
\newcommand{\ouralgo}{R\&S-Mixed }

% Commands which are specific for this document 
\DeclareMathOperator{\bBeta}{\bar{\beta}}
\DeclareMathOperator{\bgamma}{\bar{\gamma}}
\DeclareMathOperator{\bSigma}{\bar{\Sigma}}

%%%%%%%%%%%%%%%
% PAGE FORMAT %
%%%%%%%%%%%%%%%

\pagestyle{fancy}
\setlength\parindent{0in}
\setlength\parskip{0.1in}
\setlength\headheight{15pt}
\linespread{1.18}

%%%%%%%%%%% HEADER / FOOTER %%%%%%%%%%%
\chead{\textit{\projectNameShort}}
\lhead{\textsc{\userName}}
\rhead{\textsc{\doctitle}}
\rfoot{}
\cfoot{\color{gray} \textsc{\thepage~/~\pageref*{LastPage}}}
\lfoot{}

% University Logo
%\newcommand{\univlogo}{%
%  \noindent % University Logo
%  \begin{wrapfigure}{r}{0.25\textwidth}
%    \vspace{-24pt}
%    \begin{center}
%      \includegraphics[width=0.25\textwidth]{Images/univ-logo.jpg}
%    \end{center}
%    \vspace{-10pt}
%  \end{wrapfigure}
%}

%\renewcommand{\thesection}{\Roman{section}}
% \renewcommand{\thesubsection}{\thesection.\Roman{subsection}}

%%%%%%%%%%%%
% CAPTIONS %
%%%%%%%%%%%%

%%%% Paragraph separation
%\setlength{\parskip}{.5em}

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

%%%%%%%%%%%%
% DOCUMENT %
%%%%%%%%%%%%

\begin{document}

\title{\doctitle}
~\\
{\center
{\large 
{\huge
Covariate Selection Techniques for Linear Mixed-Effects Models with Applications to Population Health Problems }
\\[10mm]
\userName
\\[40mm]
{\center
\textbf{Committee}\\
Aleksandr Aravkin (chair),\\
James Burke, \\
Nathan Kutz, \\
Jonathan Wakefield (GSR)
}
\\[40mm]
{A thesis proposal submitted in partial fulfillment\\ of the requirements for the degree of}
\\[5mm]
{Doctor of Philosophy}
\\[15mm]
\department
\\[5mm]
\institution
}
\\[10mm]
\textit{Last modified: \today}

}

\newpage
{\center \textbf{Abstract}\\[5mm]}

Linear Mixed-Effects models are widely used technique for modeling clustered data, such as cohort studies, longitudinal data analysis, and meta-research. In this work we address an important practical problem of simultaneous selection of fixed and random effects in this type of models. Our approach is based on the Relax-and-Split methodology, which uses a relaxation technique together with partial minimization to efficiently solve non-convex problems. We use it to relax the constraints on the maximal number of included covariates and then apply an interior point method to obtain a solution. The performance of the resulting estimator is studied on synthetic data, as well as on two real-data applications: covariates selection for modeling the contact rate for COVID-19, and the estimation of burden of anxiety and depression disorders as results of bullying. These experiments show that the proposed method is advantageous in terms of execution time and selection quality. In addition, several theoretical questions are discussed as a part of future work that would contribute to understanding of convergence conditions of Relax-and-Split based methods in general.

\newpage
\tableofcontents

\newpage
\section*{Notation}

\begin{center}
	\begin{equation}
		\begin{split}
			n & \text{ -- Total number of objects (observations) in a dataset} \\
			m & \text{ -- Total number of groups (clusters, studies) in a dataset} \\
			n_i & \text{ -- Number of objects (observations) in the group } i. \\
			p & \text{ -- Total number of fixed features (covariates, predictors)} \\
			q & \text{ -- Total number of random features (covariates, predictors)} \\
			X_i & \text{ -- Data (design) matrix of fixed features of the group $i$, } X_i \in \R^{n_i \times p} \\
			Z_i & \text{ -- Data (design) matrix of random features of the group $i$, } Z_i \in \R^{n_i \times q} \\
			Y_i & \text{ -- Target variable for the group } i \\
			\beta & \text{ -- Vector of fixed (mean) coefficients in a model, } \beta \in \R^{p} \\
			\Gamma & \text{ -- Covariance matrix of random effects in a model, } \Gamma \in \R^{q\times q} \\
			\gamma & \text{ -- Vector of variances of random effects when random effects are independent: } \Gamma = \diag(\gamma), \ \gamma \in \R^{q} \\
			u_i & \text{ -- Random effects for the group $i$, } u_i \in \R^q, \ u_i \sim \NN(0, \Gamma) \\
			\Lambda_i & \text{ -- Covariance matrix for measurement (observation) errors for the group $i$} \\
			\sigma_i & \text{ -- Vector of variances for measurement (observation) errors for the group $i$ when } \Lambda_i = \diag(\sigma_i) \\
			\varepsilon_i & \text{ -- Vector of measurement (observation) errors for the group $i$, } \varepsilon_i \sim \NN(0, \Lambda_i) \\
			\Omega_i & \text{ -- Covariance matrix of within-cluster observations, } \Omega_i = \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i, \ \Omega_i \in \R^{n_i \times n_i} \\
			k & \text{ -- Number of non-zero mean coefficients allowed in a model} \\
			s & \text{ -- Number of non-zero variances of random effects allowed in a model} \\
			\tbeta & \text{ -- Sparse counterpart of $\beta$, } \tbeta \in \R^p, \ \|\tbeta\|_0 = k \\
			\tgamma & \text{ -- Sparse counterpart of $\gamma$, } \tgamma \in \R^q, \ \|\tgamma\|_0 = s \\
			\LL(u_i\ |\ \beta, \gamma) & \text{ -- Negative log-likelihood of random effects of a linear mixed model}\\
			\LL_{ML}(\beta, \gamma) & \text{ -- Negative marginalized log-likelihood of parameters of a linear mixed model} \\
			\LL_{REML}(\beta, \gamma) & \text{ -- Negative marginalized restricted log-likelihood (REML) of a linear mixed model}\\
			\LL_{r}(\beta, \gamma) & \text{ -- Relaxed marginalized log-likelihood of a linear mixed model} \\
			a \circ b & \text{ -- Element-wise (Hadamard) product of $a$ and $b$} \\
		\end{split}
	\end{equation}
\end{center}

\newpage

\section{Introduction}
Linear Mixed-Effects Models are widely used technique for modeling clustered data, such as cohort studies, longitudinal data analysis, and meta-research. A key step in these types of works is feature selection, which goal is choosing the most useful predictors from a large set of possible covariates. In the next section we introduce the reader to the concept of linear mixed-effect models and provide a literature review on the most commonly used optimization methods for finding estimators for them, such as Newton method, EM-algorithm, and Interior Point Method. Section \ref{ch:feature_selection_for_mixed_models} gives reader an overview of existing covariates selection methods, such as subset selection, shrinkage penalties, and AIC/BIC-based methods. Section \ref{ch:relax_and_split_thesis_work} introduces Relax-and-Split approach to feature selection, outlines the proposed algorithm, and discusses optimization subroutines involved in it. In the Section \ref{ch:experiments_on_synthetic_data} we test our approach in two simulation setups, with the goals to compare the proposed algorithm to its competitors (Section \ref{ch:comparison_to_other_algo}) and to show its applicability conditions and current limitations (Section \ref{ch:scalability_experiments}). Section \ref{ch:experiments_on_real_data} shows the application of our approach to two real-world problems: the estimation of burden of anxiety and depression disorders as results of bullying (Section \ref{ch:bullying}), and the covariates selection for modeling the contact rate for COVID-19 (Section \ref{ch:covid}). Section \ref{ch:future_work} discusses shortcomings of the current version of the algorithm and outlines theoretical questions and practical improvements which we plan address as a part of the thesis work.

\label{ch:literature_review}
\subsection{Linear Mixed-Effects Models}
Mixed-effect regression models are a particular kind of regression models for clustered data. They describe the relationship between an outcome variable and its predictors in a setting where the observations are grouped according to one or many criteria. These models are used in a wide range of applications, such as longitudinal studies for clinical trials, population health problems, risk-outcome studies, and meta-research.

Mathematically, let there be $m$ groups of observations indexed by $i$, with sizes $n_i$ respectively, and the total number of observations to be $n = n_1 + n_2 + \dots + n_m$. Depending on the domain of application, these groups are sometimes referred to as studies or clusters. For each group, we have design matrices $X_i \in \R^{n_i \times p}$, which are called the matrices of fixed features, and $Z_i \in \R^{n_i \times q}$, called matrices of random features, and we observe vectors of outcomes $Y_i \in \R^{n_i}$. Typically, columns of $Z_i$ are a subset of columns $X_i$ but it does not have to be the case in general. Let $X = [X_1, X_2, \dots, \X_m]^T$ and $Z = [Z_1, Z_2, \dots, Z_m]^T$. According to \cite{Patterson1971, Pinheiro2000}, a linear mixed-effects model (LME) is defined as:

\eq{
	\label{eq:lmm_setup}
	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i, \quad i = 1 \dots m \\
	u_i & \sim \NN(0, \Gamma),\quad \Gamma \in \R_{++}^{q\times q} \\
	\varepsilon_i & \sim \NN(0, \Lambda_i), \quad \Lambda_i \in \R_{++}^{n_i \times n_i}
}

 where $\beta \in \R^p$ is a vector of fixed (mean) covariates and $u_i \in \R^{q}$ are the vectors of unobservable random effects, which are assumed to be distributed normally with zero mean and the unknown covariance matrix $\Gamma$. The observation error's variances $\Lambda_i$ are often considered to be known, or to have a parametric form, such as $\Lambda_i = \sigma I$ where $\sigma$ is unknown. In this work, we consider $\Lambda_i$ to be known, unless stated otherwise.

 If one considers $\omega_i = Z_i u_i + \varepsilon_i$ to be an unknown per-cluster noise vectors, then the model \ref{eq:lmm_setup} can also be viewed as a correlated noise model:

 \eq{
 \label{eq:lmm_correlated_noise_setup}
	Y_i = X_i\beta + \omega_i,\quad \omega_i \sim \NN(0, \Omega_i(\Gamma)), \quad \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i
}

For the sake of conciseness, since $\Omega_i$ is always a function of $\Gamma$, $\Omega_i(\Gamma)$ will further be spelled as just $\Omega_i$, without the parentheses. In fact, $\Omega_i$ will be the only terms depending on $\Gamma$ in the majority of expressions including the mixed model's likelihood and its derivatives described below.

 When fitting an LME, the goal is to get both the estimation of fixed covariance $\beta$ and the estimation of variance parameters of random effects $\Gamma$. One of the possible approaches is integrating random effects $u_i$ out to get a marginalized posterior distribution which depends only on $\beta$ and $\Gamma$, but not on $u_i$. According to Bayes Theorem:

\eq{
	\label{eq:integrating_re_out_short}
	f_{\beta, \Gamma | X = \hat{X}, Y=\hat{Y}}(\beta, \Gamma) = \frac{f_{\beta, \Gamma}(\beta, \Gamma)}{f_{X, Y}(\hat{X}, \hat{Y})}\int_{v}f_{X, Y | \beta, u = v}(\hat{X}, \hat{Y})f_{u|\Gamma}(v)dv 
}

where ($\hat{X}, \hat{Y}$) is a particular implementation of random variables $(X, Y)$; $f_{\beta, \Gamma | X=\hat{X}, Y=\hat{Y}}(\beta, \Gamma)$ is a posterior distribution of the model's parameters given data $(\hat{X}, \hat{Y})$, also known as the likelihood of the model's parameters; $f_{\beta, \Gamma}(\beta, \Gamma)$ is a joint prior distribution of the model's parameters; $f_{X, Y | \beta, u}(\hat{X}, \hat{Y})$ is likelihood of the data $(\hat{X}, \hat{Y})$ for the model (\ref{eq:lmm_setup}) given $\beta$ and $u_i$ are known; and $f_{u|\Gamma}(u)$ is a conditional distribution of random effects $u_i$ given that $\Gamma$ is known.

Omitting the parameters-independent multiplier $f_{X, Y}(\hat{X}, \hat{Y})$ and taking the negative log-likelihood of (\ref{eq:integrating_re_out_short}), we obtain $\LL_{ML}$:

\eq{
	\label{eq:lmm_objective}
	& \LL_{ML}(\beta, \Gamma)  \equiv -\log(f_{\beta, \Gamma| X=\hat{X}, Y=\hat{Y}}(\beta, \Gamma)) \sim \\ & \sim   \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma^{-1}Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \half\log{\det{\pa{Z_i \Gamma^{-1}Z_i^T + \Lambda_i}}} -\log(f(\beta, \Gamma)) = \\
	& = \sum_{i = 1}^m \half(y_i - X_i\beta)^T\Omega_i^{-1}(y_i - X_i\beta) + \half\log{\det{\Omega_i}} -\log(f_{\beta, \Gamma}(\beta, \Gamma))
}

This expression $\LL_{ML}(\beta, \Gamma)$ is known as the negative log likelihood function of a linear mixed-effects model. The term $-\log(f_{\beta, \Gamma}(\beta, \Gamma))$ represents prior knowledge about $\beta$ and $\Gamma$, and is often referred to as a regularization term. As an example, setting $-\log(f_{\beta, \Gamma}(\beta, \Gamma)) = \frac{1}{2\sigma^2_\beta}\|\beta\|_2^2 + \frac{1}{2\sigma^2_\Gamma}\|\Gamma\|_2^2$ would represent our prior belief that $\beta \sim \NN(0, \sigma^2_\beta)$ and $\Gamma \sim \NN(0, \sigma^2_\gamma)$.

The unknown coefficients $\beta$ and $\Gamma$ are to be obtained via solving the following optimization problem:

\eq{
	\label{eq:ml_lme_optimization_setup}
	\min_{\beta, \Gamma} & \ \LL_{ML}(\beta, \Gamma) \\
	s.t. & \ \Gamma \in \R_{++}^{s\times s}
}

In a simpler case where $\Gamma = \diag(\gamma)$ is diagonal (often referred to as \textit{the diagonal setup}), the positive definite constraint from (\ref{eq:ml_lme_optimization_setup}) transforms into a box constraint:

\eq{
	\label{eq:ml_lme_diagonal_setup}
	\min_{\beta, \gamma} & \ \LL_{ML}(\beta, \diag(\gamma)) \\
	s.t. & \ \gamma > 0
	}

\begin{tip}
\paragraph{Use of $\Gamma$ versus $\gamma$.} Starting from Chapter \ref{ch:relax_and_split_thesis_work} this work is focused on the case where the matrix $\Gamma$ is diagonal: $\Gamma = \diag(\gamma)$ (the diagonal setup (\ref{eq:ml_lme_diagonal_setup})), so the unknown covariance parameters of the problem are often referred to as just $\gamma$. This simpler case corresponds to the situation when random effects $u_{ij}$ corresponding to different mean effects $\beta_j$ are independent of each other. In this case the positive definiteness constraint is guaranteed to satisfy whenever $\gamma > 0$. Authors made the decision to start with the diagonal case because it covers their immediate needs in applications, and they see the extensions of the algorithm proposed in Chapter \ref{ch:proposed_algorithm} to the non-diagonal case to be a part of future work. This chapter, however, discusses both cases of $\gamma$ and $\Gamma$ to make connections between the general and the particular cases, and to preserve the notation consistency with the referenced works.
\end{tip}

%Let's consider a simpler case when $\Gamma$ is set to be a diagonal matrix $\diag(\gamma)$, which  The case of $\gamma_j = 0$ for some $j$ is related to $u_i$ being drawn from a degenerate distribution with the variances for $j$'th component to be zero. It implies that random, group dependent, part for $j$'th covariate is zero, and so the variability of the respective coefficient over groups does not help the model to explain the variance in the target variable. Since this particular case of some $\gamma_j$ being zero is directly related to selection of random features, we relax the constraint of $\Gamma$ to be non-negative definite instead of positive-definite. It also does not cause any singularities in the loss function since the matrices $\Omega_i$ are invertible and positive-definite for as long as $\Lambda_i$ are positive-definite and $\Gamma$ is non-negative definite.

The expected value of the posterior mean effects $\beta$ given $\Gamma$ has a closed form representation which immediately follows from (\ref{eq:lmm_correlated_noise_setup}):

\eq{
	\label{eq:beta_formula}
	\beta^{(k+1)} = \argmin_{\beta}\LL_{ML}(\beta, \Gamma^{(k)}) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i
}

There is, however, no closed-form solution for $\Gamma$ in general case, so it has to be obtained via an optimization subroutine. There are three aspects that make this problem challenging from the optimization point of view:

\begin{enumerate}
	\item \textbf{Non-convexity}: the loss function $\LL_{ML}$ is not convex with respect to $\Gamma$, which can prevent an optimization method from obtaining a global solution.
	\item \textbf{Overfitting}: the matrix $\Gamma$ contains as many as $\frac{q(q-1)}{2}$ free parameters, overfitting of which may result in near-singularities, especially when the sample size is small \cite{Lange1989}. The diagonal setup $\Gamma = \diag(\gamma)$ is still prone to overfitting, although has less freedom ($q$ free parameters) to do so.
	\item \textbf{Positive Constraint}: the requirement of keeping the matrix $\Gamma$ to be positive definite limits the scope of methods which can be used for optimization. The diagonal setup has a box constraint $\gamma > 0$ which also has to be accommodated by the optimization subroutine.
\end{enumerate}

When $\beta^*$ and $\Gamma^*$, the minimizers for (\ref{eq:ml_lme_optimization_setup}), are obtained, the individual random effects $u_i$ can be found through the minimization of conditional likelihood of random effects:

\eq{
\min_{u_i}\LL(u_i|\beta, \gamma) = \min_{u_i}\left(\half u_i^T\Gamma^{-1}u_i + \half (Z_iu_i - Y_i + F_i(\beta))^T\Lambda_i^{-1}(Z_iu_i - Y_i + F_i(\beta))\right),
}

This problem has a closed form solution which was first derived by \cite{Harville1976} as an extension to Gauss-Markov theorem:

 \eq{
	\label{eq:random_effects_formula}
	u_i = (\Gamma^{-1} + Z_i^T\Lambda_i^{-1}Z_i)^{-1}Z_i^T\Lambda_i^{-1}(Y_i - X_i\beta).	
}

As pointed by \cite{Harville1974}, $\Gamma^*$ -- the minimizer for the maximal likelihood formulation \ref{eq:ml_lme_optimization_setup} -- is known to be biased downwards because it does not take into account the loss of $k$ degrees of freedom from estimating $\beta^*$ within the same inference procedure. To overcome this effect they proposed a Restricted Maximal Likelihood (REML) formulation where the loss function is based on $n-k$ linearly independent error contrasts:

\eq{
	\label{eq:reml_lmm_objective}
	\LL_{REML}(\beta, \Gamma) = -\half\log\det\left(\sum_{i=1}^m X_i^T(Z_i\Gamma^{-1}Z_i^T + \Lambda_i)^{-1}X_i\right) + \LL_{ML}(\beta, \Gamma)
}

Although different from ML, the REML formulation does not introduce more complexity to optimization: all the solution methods for ML formulation discussed below also have an adaptation to REML.

Two main optimization approaches to solving mixed effects likelihood formulation are used: EM algorithm \cite{Dempster1977} and Newton-Raphson \cite{Lindstrom1988}.

\subsubsection{EM algorithm for Mixed Models}
One more way of thinking about mixed-effects models is considering \textbf{$\{Y_i\}_{i=1}^m$} to be incomplete data and \textbf{$\{Y_i, u_i\}_{i=1}^m$} to be complete data, where $u_i$ is the unobservable part. In this setting, the problem of fitting a mixed model can also be classified as a statistical inference problem in the presence of missing data, and so the methods for solving this kind of setup might be applicable. One example is EM-algorithm -- a general approach to estimating parameters in problems with missing data developed by \cite{Dempster1977}. As later discussed by \cite{Laird1982}, one can view EM-algorithm as optimizing the approximating function $Q(\beta, \Gamma | \hat{\beta}, \hat{\Gamma})$ -- the likelihood of $(\beta, \Gamma)$ given that the missing data $u_i$ comes from a normal distribution $\NN(\hat{\beta}, Z_i\hat{\Gamma} Z_i^T + \Lambda_i)$. EM-algorithm, in its general form, consists of two steps: Expectation step and Minimization step:

\eq{
	\text{E-step: }\quad & Q^{(k)}(\beta, \Gamma | \hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}) := p\left(\beta, \Gamma | \hat{u_i}^{(k)} =  \expectation{u_i|\hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}}\right) \\
	\text{M-step: }\quad & \hat\beta^{(k+1)}, \hat\Gamma^{(k)}  :=  \argmax_{\beta, \Gamma}\ Q(\beta, \Gamma^{k+1} | \hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}) 
}

On the E-step the function $Q(\beta, \Gamma | \hat{\beta}, \hat{\Gamma})$ is constructed using the expected values of missing data given the previous-iteration parameters $\hat{\beta}, \hat{\Gamma}$, and on M-step $Q$-function is minimized, and its minimizers $\hat{\beta^*}, \hat{\Gamma^*}$ are set to be next-iteration parameters. The process continues until convergence.

As discussed in \cite{Laird1987}, one can express the EM-algorithm's iterations in the linear mixed models setup as:

\eq{
	\beta^{(k+1)} & = \argmin_{\beta}\LL(\beta, \Gamma^{(k)}) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}(\Gamma^{(k)})X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}(\Gamma^{(k)})Y_i \\
	u_i^{(k+1)} & = ((\Gamma^{(k)})^{-1} + Z_i^T\Lambda_i^{-1}Z_i)^{-1}Z_i^T\Lambda_i^{-1}(Y_i - X_i\beta^{(k)}) \\
	\Gamma^{(k+1)} & = \frac{1}{m} \sum_{i=1}^m u_i^{(k)} (u_i^{(k)})^T + \Gamma^{(k)}(I - Z_i\Omega_i^{-1}(\Gamma^{(k)})Z_i^T\Gamma^{(k)})	
	}

EM-algorithm has two main advantages: simplicity of iterations and the absence of constraints in optimization subroutines. The later is especially important: EM-algorithm guarantees that if the initial guess $\Gamma^{(0)}$ was positive definite then all $\Gamma^{(k)}$, including, under certain conditions, the final result $\Gamma^*$, will be positive definite. This removes the necessity to handle the constraint $\Gamma \in \R^{s\times s}_{++}$ separately, e.g. via re-parametrization using Cholesky decomposition, which allows to keep the iterations fast and simple. As an apparent consequence, EM algorithm also correctly handles the box constraint in the diagonal setup~(\ref{eq:ml_lme_diagonal_setup}).

The main drawback of EM-algorithm is its slow convergence speed. In the form above, \cite{Dempster1977} showed that the algorithm monotonically converges, with geometric rate, as established by~\cite{Wu1983}, to a local maximal likelihood estimator. The geometric rate is due to the fact that EM-algorithm can be equivalently viewed as a proximal gradient descent on KL-divergence based likelihood. There have been wide effort on both discovering more statistical properties of the algorithm, as well as on improving its convergence rate. For instance, in his recent work \cite{Balakrishnan2017} provided a quantitative characterization of a region such that the EM-algorithm, initialized within that region, is guaranteed to converge to a "sufficiently good", fixed point: the one within statistical precision of the global optimum. There are also works which are empirically improving the convergence speed of EM in terms of computations and time via implementation-level improvements, like \cite{Liu1994, Liu1998}, to name a few.

\subsubsection{Newton Methods for Mixed Models}
\label{section:newton_methods_for_mixed_models}
Newton method as a second-order optimization method for estimating parameters of mixed models was proposed by \cite{Jennrich1986} and later improved by \cite{Lindstrom1988}. The core idea of this approach is to consider $\hat\beta(\Gamma)$ as a function of $\Gamma$ (see Eq. \ref{eq:beta_formula}), and apply Newton method to the resulting formulation:

\eq{
	\label{eq:lmm_loss_only_of_big_gamma}
	\min_{\Gamma}\ & \LL_{ML}(\beta(\Gamma), \Gamma) \\
	\text{s.t.}\ & \Gamma \in \R^{s\times s}_{+}
}

In a general setup when $\Gamma$ is not diagonal, one can treat the positive-definiteness constraint by reparametrizing the problem via modified Cholesky decomposition, as suggested by \cite{Chen2003}:
\eq{
	\label{eq:dlld_decomposition}
	\Gamma = DLL^TD
}
where $D$ is an $s\times s$ diagonal matrix, and $L$ is an $s\times s$ lower-triangular matrix with ones on the main diagonal. Applying this variables transformation to the equation (\ref{eq:lmm_loss_only_of_big_gamma}) we get an unconstrained optimization problem on diagonal elements of $D$ and on the off-diagonal elements of $L$:

\eq{
	\min_{D, L}\ & \LL_{ML}(\hat{\beta}(DLL^TD), DLL^TD) \\ 
	\text{s.t.}\ &  D_{ij} = 0, \ i \neq j \\ 
	& \ L_{ii} = 1,  \  i = 1 \dots s
}

In a simpler case of $\Gamma$ being diagonal ($\Gamma = \diag{\gamma}$), the setup (\ref{eq:lmm_loss_only_of_big_gamma}) simplifies to

\eq{
	\label{eq:lmm_loss_only_of_gamma}
	\min_{\gamma > 0} \LL(\hat{\beta}(\gamma), \gamma)
}
On each iteration, Newton-like family of methods is looking for $\gamma^{(k+1)}$ -- a minimizer for a quadratic approximation of the problem at the point $\gamma^{(k)}$:
\eq{
	\gamma^{(k+1)} = \argmin_{\Delta\gamma} & \LL_{ML}(\gamma^{(k)}) + \nabla_\gamma\LL_{ML}(\gamma^{(k)})^T \Delta\gamma + \Delta\gamma^T\nabla^2_{\gamma\gamma}\LL_{ML}(\gamma^{(k)})\Delta\gamma\\
	\text{ s.t. } & \gamma^{(k)} + \Delta\gamma \geq 0
}

The positivity constraint narrows the scope of optimization techniques which can be used. There is, however, still wide variety of methods to choose from like Truncated Newton Methods with Barriers \cite{Nash2000}, Projection Methods \cite{Bertsekas1982}, or Interior Point methods \cite{Potra2000}, one can refer to \cite{TrustRegionBook} for a detailed overview of these algorithms. The Interior Point method shall be discussed in more details since it's used in Chapter \ref{ch:proposed_algorithm} of this work.

Interior Point (IP) methods \cite{Kojima1991, Nesterov1994, Wright1997} target the Karush-Kuhn-Tucker optimality conditions for the problem (\ref{eq:lmm_loss_only_of_gamma}) directly by introducing ``margins'' -- non-negative axillary variables which transform inequality constraints into equality constraints:

\eq{
	\label{eq:lmm_loss_only_of_gamma}
	\min_{\gamma, d} & \LL(\gamma) \\
	\text{ s.t. } & \gamma - d = 0 \\
	& d \geq 0
}

The positivity constraint for $d$ is then handled via a logarithmic barrier. The Lagrangian and the KKT system for this problem are:

\eq{
	\label{eq:kkt_conditions}
	F_\mu(\gamma, d, v) & = \LL(\gamma) + v^T(\gamma - d) - \mu\sum_{i=1}^s\log(d_i) \\
	\nabla F & = \begin{bmatrix}
		\nabla_{\gamma} F_\mu \\
		\nabla_{v} F_\mu \\
		\nabla_{d} F_\mu 
	\end{bmatrix} = \begin{bmatrix}
		\nabla \LL(\gamma) + v \\
		\gamma - d \\
		v \circ d - \mu\circ 1
	\end{bmatrix} = 0
}

where $v$ is a vector of dual variables, and $\mu$ is a positive hyper parameter which define the steepness of the logarithmic barrier, and $v\circ d$ denotes the element-wise product.
Application of Newton root-finding method to this KKT system gives us the Interior Point method. In particular, the search direction $[\Delta\gamma, \Delta v, \Delta d]$ as the solution of the following system:

\eq{
	\label{eq:ip_system_as_equation}
	\nabla^2F_\mu\begin{bmatrix}
		\Delta \gamma \\
		\Delta v \\
		\Delta d
	\end{bmatrix} = -\nabla F_\mu\begin{bmatrix}
		\Delta \gamma \\
		\Delta v \\
		\Delta d
	\end{bmatrix}
}

The iteration then looks like
\eq{
	\gamma^{(k+1)} & = \gamma^{(k)} + \alpha \Delta\gamma \\
	v^{(k+1)} & = v^{(k)} + \alpha \Delta v \\
	d^{(k+1)} & = d^{(k)} + \alpha \Delta d
}

$\alpha$ is chosen via line search so all $\gamma^{(k+1)}$, $v^{(k+1)}$, and $d^{(k+1)}$ are positive, and a sufficient descent condition is satisfied:
\eq{
	\|\nabla F_\mu(\gamma^{(k+1)}, v^{(k+1)}, d^{(k+1)})\| \leq 0.99\|\nabla F_\mu(\gamma^{(k)}, v^{(k)}, d^{(k)})\| 
}

Since this is a root finding procedure for the equation (\ref{eq:kkt_conditions}), the natural stopping criterion is 
\eq{
\|\nabla F_\mu(\gamma^{(k)}, v^{(k)}, d^{(k)})\| \leq \varepsilon
}
for some sufficiently small $\varepsilon$.

Interior Point method is a second-order method, so, under certain assumptions, it has a quadratic rate convergence region around the optimal point. This leads to better performance in terms of the number of iterations comparing to EM-algorithm. The performance in terms of time, however, strongly depends on the problem's size. It's due to the fact that working with Hessian information requires solving a linear system, which is not needed for first order methods like EM-algorithm.

\subsection{Feature Selection for Mixed Models}
\label{ch:feature_selection_for_mixed_models}
In statistical learning setting, feature selection is defined as a process of selecting, or ranking, the most important predictors in a dataset. The goal is to get an interpretable model without sacrificing prediction quality. Selection process often happens simultaneously with fitting the model, where a predictor is considered to be selected if its respective coefficient in the model is non-zero. If the desired number of coefficients $k$ is given, then  the feature selection problem can be formulated as minimization of a loss function $f(\theta)$, which is  often proportionate to the negative log-likelihood $\log\, p(\theta| X, Y)$, with respect to its parameters $\theta$ subject to a zero-norm constraint:

\eq{
	\label{eq:general_feature_selection_zero_norm}
	&\min_{\theta}  f(\theta) \\
	& \text{ s.t. } \|\theta\|_0 \leq k 
	}

 In this form the problem is equivalent to the Knapsack problem, which is NP-complete. Many techniques based on exhaustive search, also known as subset selection process, were developed, and showed to be effective when the total number of predictors is small (see \cite{Muller2013}). The same setup appears to be intractable in a high-dimension setting due to the exponential growth of the number of subsets to check when $n \to \infty$, however, it's possible to get an approximate solution via relaxation techniques discussed below.

 One way to work with zero-norm constraint in the Eq.\ref{eq:general_feature_selection_zero_norm} is to relax it to a first norm constraint. An example of such approach is LASSO, or least absolute square shrinkage operator, studied by \cite{Tibshirani1996} for the least squares model:

 \eq{
 	\label{eq:lasso_regression}
 	\min_\beta ||y - X\theta||_2^2 + \lambda||\theta||_1
 }

  It combines the small bias of the least square estimator and the interpretability of the subset selection method, producing, however, biased estimates for the large coefficients, according to \cite{Zou2006}.  This approach was later extended to exhibit other desirable properties, such as lesser bias for larger coefficients (SCAD from \cite{Fan2001}) or simultaneous selection for highly correlated predictors (Elastic Net from \cite{Zou2005}), and to accommodate other model setups, including classification ($\ell_1$-SVM from \cite{Bradley1998, Shalev-Shwartz2011, Belyy2018}), clustering (\cite{Jajuga1991}), and reinforcement learning (\cite{Kim2019}), to name a few.

  Feature selection for linear mixed models is generally a more challenging problem than feature selection for linear regression problem. First, in linear regression setting the observation are independent, whereas in mixed-effects setup they're typically dependent. Second, one needs to do simultaneous selection of mean effect variables ($\beta$) and the ones related to their random effects' covariance structure ($\Gamma$), and these groups can have different relative importance. Finally, selecting random features implies that $\Gamma$ will have zero columns and rows, which prevents it being positive-definite. Consequently, selecting variance parameters leads to boundary issues which make both theoretical analysis and numerical computations more challenging.

   The shrinkage operator approach was first adapted to the mixed-effect models setting by \cite{Lan2006} to perform fixed effects selection. Selection of random effects is generally more difficult than selection of fixed effects alone because removing random effect from the model requires elimination of the entire respective row and column of $\Gamma$. This extension was provided by \cite{Krishna2008}, who  performed simultaneous fixed and random effects selection by combining three components: the modified Cholesky decomposition $\Gamma = DLL^TD$, as suggested by \cite{Chen2003}, the Adaptive LASSO penalty, as in \cite{Lan2006}, and EM algorithm (\cite{Dempster1977}). Their penalized Q-function for EM-algorithm is defined as:
  \eq{
  	\label{eq:krishna_lasso}
  	Q(\beta, D,L|\beta^{(k)}, D^{(k)}, L^{(k)}) = & \mathbb{E}_{u_i | \beta^{(k)}, D^{(k)}, L^{(k)}}\left[{\sum_{i=1}^m \|Y_i - X_i\beta - Z_iDLu_i\|^2}\right] + \\ + & \lambda\left(\sum_{j=1}^k\frac{|\beta_j|}{|\hat\beta_j|} + \sum_{j = 1}^s\frac{|D_{ii}|}{|\bar{D}_{ii}|}\right)
  }
   They suggest tuning the regularization parameter $\lambda$ using a BIC criterion. A similar formulation was used by \cite{Lin2013} who proposed a joint selection with two-stage ALASSO. On stage 1 they approximate the shrinkage penalty with local quadratic approximation and use Newton method to select random effects, and then, on stage 2 they select fixed effects via penalized log-likelihood where the covariance variables are fixed.

   Another type of regularization which is used for feature selection is SCAD \cite{Fan2001} which acts on the first derivative of the penalty function $p_\lambda(\theta)$:
   \eq{
   		p'_{\lambda}(|\theta|) = \lambda\left\{\ind{\theta \leq \lambda} +  \frac{(a\lambda - \theta)_+}{(a-1)\lambda}\ind{\theta - \lambda} \right\}
   }
   The adaptation of SCAD penalty to selecting both fixed and random features in linear mixed model setup was developed by \cite{Fan2012}.

  In high-dimensional setting marginal likelihood is often replaced with a hierarchical approximation of it known as h-likelihood (\cite{Lee1996}), which, for a given data $(\hat{X}, \hat{Y})$ is defined as

  \eq{
  \label{eq:h_likelihood}
  \LL_{H}(\beta, \Gamma, u) = -\log f_{X, Y | \beta, \Gamma, u}(\hat{X}, \hat{Y}) - \log f_{u|\Gamma}(u, \Gamma)
  }

  Where $f_{X, Y | \beta, \Gamma, u}(\hat{X}, \hat{Y})$ is a likelihood of data given $\beta, \Gamma,$ and $u$, and $f_{u|\Gamma}(u, \Gamma)$ is a conditional density of random effects $u$ given $\Gamma$. This formulation allows to avoid integration over random effects which is required when the classical marginal likelihood is used. Practically, the use of h-Likelihood eliminates the need to compute and to invert the covariance matrices $\Omega_i$ or their Cholesky factors, which is often the most expensive operation when computing the marginal likelihood numerically. When derived using h-Likelihood, the estimators of fixed effects $\beta$ and random effects $u_i$ are, as shown by \cite{Lee1996}, asymptotically best unbiased predictors. In order to estimate variance components simultaneously with fixed effects an Adjusted Profile likelihood, which is an hierarchical generalization of REML, is used.

  Similarly to marginal likelihood above, various feature selection approaches were adapted to be used with h-Likelihood. Similarly to what \cite{Krishna2008} proposed for marginal likelihood based selection, \cite{Xu2015} proposed to penalize the fixed effects and random effects variance components $\Omega_i$ in h-likelihood with LASSO-like family of penalties. They also showed, that, under certain assumptions, the proposed estimator possesses asymptotic normality property. The most recent development in this trend to date was provided by \cite{Xie2020} who utilized multi-armed bandits approach for selecting mean effects in high-dimensional, h-Likelihood, setting.

  Shrinkage penalties and norm-based penalties are not the only kind of relaxations which are used in feature selection. Recently \cite{Ghosh2018} studied SCAD regularization for selecting mean effects in a high-dimensional setting motivated by genomics problems. Other approaches to feature selection are based on information criteria such as AICs and BICs \cite{Jones2011}. According to \cite{Muller2013}, the most widely used AIC criterion is what \cite{Vaida2005} call the marginal AIC criterion:
  \eq{
  \label{eq:vaida_aic}
  AIC = 2\LL(\hat{\theta}) - 2\alpha_n(p+q)
  }
  where $\hat\theta$ includes all the estimated parameters in a mixed-effect model $(\beta, \Gamma)$, and $\alpha_n = n(n-p-q-1)$ for a finite sample case \cite{Sugiura1978}. As discussed by \cite{Fang2011}, AIC is asymptotically equivalent to leave-one-out cross-validation, so it can be used for choosing between a finite number of models.

  One can refer to \cite{Muller2013} for a detailed overview of different feature selection approaches.

  Variable projection is another approach to tackle the zero-norm constraint in (\ref{eq:general_feature_selection_zero_norm}). Relax-and-Split \cite{Zheng2018RelaxAndSplit} technique is one of implementations of this approach. It's been successfully applied in basis pursuit denoise problems \cite{Baraldi2019}, compressed sensing \cite{Zheng2019}, and sparse models identification \cite{Champion2020}. The key idea of this method is to use a set of auxiliary variables to relax the constraints:
\eq{
	\min_{\|\beta\|_0 \leq k} f(\beta) \quad \to \quad \min_{\beta;\ \|w\|_0 \leq k} f(\beta) + \frac{1}{2\nu}\|w - \beta\|_2^2
}
and then to apply an alternating minimization algorithm to solve the relaxation. This approach will be discussed in details in the next chapter.

\newpage
\section{Relax-and-Split for Mixed-Effects Selection}
\label{ch:relax_and_split_thesis_work}
\subsection{Introduction}
Recently \cite{Zheng2018RelaxAndSplit} proposed to use variable projection technique called Relax-and-Split for nonconvex-composite optimization problems of the form:

\eq{
	\min_\beta f(\beta) = h(A\beta) + g(\beta)
}
where $\beta \in \R^k$ are the decision variables, $A\in \R^{n \times k}$, $h(\beta)$ is non-smooth, non-convex, and separable, and $g(\beta)$ is convex. The key innovation of their work is to relax the original variable $\beta$ with a set of auxiliary variables $\tbeta$, and then use partial minimization techniques to develop effective algorithms:
\eq{
	\min_x f(\beta) = h(A\beta) + g(\beta)\quad \to \quad \min_{\beta, \tbeta} h(\tbeta) + \frac{1}{2\theta}\|A\beta - \tbeta\|_2^2 + g(\beta)
}
The same approach can be used for relaxing constraints in non-convex optimization problems. This approach was discussed by \cite{Zheng2019} as a part of SR3 framework: 
\eq{
	\min_{\beta \in \Delta} f(\beta) \quad \to \quad \min_{\tbeta\in \Delta;\ \beta} f(\beta) + \frac{1}{2\nu}\|\beta - \tbeta\|_2^2
}

As an example, the zero-norm constraint from (\ref{eq:general_feature_selection_zero_norm}) can be relaxed as follows:

\eq{
	\label{eq:relax_and_split_in_general_form_2}
	\min_\beta f(\beta)\ \text{ s.t. }\|\beta\|_0 \leq k \quad \to \quad \min_{\beta, \tbeta} \ &  f(\beta) + \lambda\|\beta - \tbeta\|_2^2 \ \text{ s.t. } \|\tbeta\|_0 \leq k
}
The second norm links $\beta$, an unconstrained variable, and its constrained counterpart $\tbeta$. The coefficient $\lambda$ represents the strength of this link and can be viewed as a Gaussian prior for the distance of $\beta$ from the "wall" -- a $k$-dimensional subspace. The larger $\lambda$ is -- the higher penalty is for $\beta$ violating the zero-norm constraint.

This relaxation admits effective application of alternating minimization methods:
\eq{
	\label{eq:relax_and_split_general_iteration}
	\beta^{(t+1)} = & \argmin_{\beta}\left[ f(\beta) + \lambda\|\beta - \tbeta^{(t)}\|^2_2\right] \\
	\tbeta^{(t+1)} = & \argmin_{\tbeta} \|\beta^{(t)} - \tbeta\|_2^2 \\
	& \text{ s.t. } \|\tbeta\|_0 \leq k
}
The first optimization subroutine is an unconstrained optimization problem. The second subroutine is a Euclidean projection of $\beta^{(t)}$ onto a $k$-dimensional subspace. This operation can be efficiently implemented since it only involves taking $k$ largest in absolute value elements of $\beta^{(t)}$. The coefficient $\lambda$ is iteratively increased until $\|\beta - \tbeta\| \leq \varepsilon$ for some small $\varepsilon$.

This relaxation has certain advantages over $\ell_1$-type relaxations:
\begin{itemize}
	\item \textbf{Interpretability of hyper-parameters.} Notice that this relaxation does not change the meaning of the only hyper-parameter $k$. In both formulations in Eq. (\ref{eq:relax_and_split_in_general_form_2}) the parameter $k$ denotes the number of non-zero coefficients in the resulting estimator $\beta^*$ whereas in the formulation (\ref{eq:lasso_regression}) $k$ is replaced with $\lambda$, which denotes the variance of Laplacian prior for $\beta^*$. There is no data-independent mapping between $\lambda$ and $k$ which would allow researcher to make an informed decision on the value of $\lambda$ before calling the LASSO fit, which leads to multiple trial-launches resulting in a overall production pipeline slowdown. 
	\item \textbf{No model bias given the subspace.} Notice that in (\ref{eq:relax_and_split_in_general_form_2}) the local minimizer for the relaxation when $\lambda \to \infty$ will also be a local minimizer for a non-relaxed setup (\ref{eq:general_feature_selection_zero_norm}). The solution for the formulation (\ref{eq:lasso_regression}) is, in a general case, not going to be a local minimizer for (\ref{eq:general_feature_selection_zero_norm}) since its non-zero coefficients are biased by the implicit Laplacian prior (\cite{Zou2006} provides a comprehensive discussion on the nature of this bias). Because of that, feature selection via LASSO is normally executed as a two-stage process, when, after the optimal subset of features is being selected on the stage one, the optimization is repeated on that features only without $\ell_1$ regularization to partially eliminate the model bias. There is, however, no need for the second stage when the formulation (\ref{eq:relax_and_split_in_general_form_2}) is used. 
\end{itemize}

The main disadvantage of the formulation (\ref{eq:relax_and_split_in_general_form_2}) comparing to LASSO (\ref{eq:lasso_regression}) is that the latter only needs to be solved once, whereas the former should be solved iteratively for increasing $\lambda$ until the relaxation is eliminated, which naturally takes longer. This, however, can be mitigated by applying warm-start technique, when the optimization after increasing $\lambda$ starts from the previous optimal $\beta^*$. This approach is especially efficient when the second order optimization subroutine is used for the step one of the algorithm (\ref{eq:relax_and_split_general_iteration}): the subsequent problem differs from the previous one by a quadratic term, so if the method previously reached the quadratic convergence region then increasing $\lambda$ won't change it.

% This relaxed problem also admits effective partial minimization schemes. Partial minimization over the original variables $x$ is now unconstrained optimization problem, and partial minimization over $w$ is a Euclidean projection problem onto $\Delta$. The later can be as complex as the set $\Delta$ is, but for many real-world applications the projections are accessible in analytic form. The main advantage of having a relaxation is that the complexity of projecting on $\Delta$ is now handled separately from the complexity of optimizing the function $f(x)$. This comes with a price of solving the relaxation iteratively with $\nu \to 0$ until $w = x$ and $x\in \Delta$, so the original constraint is satisfied.

\subsection{Proposed Algorithm}
\label{ch:proposed_algorithm}

Consider a mixed-effect model setting described in Eq. (\ref{eq:lmm_setup}) with $\Gamma$ being diagonal: $\Gamma = \diag(\gamma)$.  We wish to find a minimizer the following loss function:

\eq{
	\label{eq:fs_loss_intractable}
	\min_{\beta, \gamma} & \LL_{ML}(\beta, \diag{(\gamma)}) \\
	\text{s.t. } 	& \|\beta\|_0 \leq k \\
					& \|\gamma\|_0 \leq j \\
					& \gamma \geq 0
}

where $\LL_{ML}(\beta, \diag{(\gamma)})$ is a marginalized log-likelihood of a mixed model defined in Equation (\ref{eq:lmm_objective}).

In this form, the problem can have as many as $C^n_k$ local minimizers (at least one per $k$-subspace), and finding a global solution is equivalent in complexity to solving a weighted knapsack problem. However, finding a "sufficiently good" local minimizer is usually enough for practical purposes.

We relax the original problem using Relax-and-Split approach (\cite{Zheng2018RelaxAndSplit}) in order to both automate choosing a good subspace and to do optimization in the whole space instead of doing optimization in subspaces. Namely, we introduce two constrained auxiliary variables $\tbeta$ and $\tgamma$ and treat them as projections of now unconstrained variables $\beta$ and $\gamma$ on $k$- and $j$-subspaces respectively. The loss function $\LL_r(\beta, \gamma, \tbeta, \tgamma)$ of for the relaxed setup is defined as:
\eq{
	\LL_r(\beta, \gamma, \tbeta, \tgamma) \equiv \LL_{ML}(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta\|^2_2 + \frac{\lambda_{\gamma}}{2}\|\gamma - \tgamma\|_2^2
}

where $\LL_{ML}(\beta, \diag{(\gamma)})$ is defined in (\ref{eq:lmm_objective}):
\eq{
	\label{eq:lmm_objective_2}
	\LL_{ML}(\beta, \gamma) = \sum_{i = 1}^m \half(Y_i - X_i\beta)^T\Omega_i^{-1}(\gamma)(Y_i - X_i\beta) + \half\log{\det{\Omega_i(\gamma)}}
}
and $\Omega_i(\gamma)$ is a covariance matrix for the $i$'th group observation's noise:
\eq{
	\Omega_i(\gamma) = Z_i\diag(\gamma)Z_i^T + \Lambda_i
}

In order to solve a relaxed setup for given relaxation parameters $\lambda_\beta$ and $\lambda_\gamma$ we need to find a solution $(\beta^*, \gamma^*, \tbeta^*, \tgamma^*)$ for the following problem:

\eq{
	\label{eq:lmm_relax_and_split_full_setup}
	\min_{\beta, \gamma, \tbeta, \tgamma}& \LL_r(\beta, \gamma, \tbeta, \tgamma)\\
	\text{s.t. } 	& \|\tbeta\|_0 \leq k \\
					& \|\tgamma\|_0 \leq j \\
					& \gamma \geq 0 \\
					& \tgamma \geq 0
}

We apply variable projection technique to solve the relaxed problem (\ref{eq:lmm_relax_and_split_full_setup}). This approach outlines the Algorithm \ref{alg:feature_selection_relax_and_split_outline}, henceforth referred as \ouralgo,  as an iterative procedure with four projection steps within each iteration. In order to guarantee that $(\tbeta, \tgamma)$ from (\ref{eq:lmm_relax_and_split_full_setup}) is a solution for (\ref{eq:fs_loss_intractable}), we iteratively increase $\lambda_\beta$ and $\lambda_\gamma$ until $\beta=\tbeta$ and $\gamma=\tgamma$.

\begin{algorithm}
	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
	\Repeat{converges}{
		$\tbeta^{(k+1)} \leftarrow \proj_{\|\beta\|_0 \leq k}(\beta^{(k)})$ \tcc*[r]{\label{alg:feature_selection_relax_and_split_outline:tbeta_line}Take max k from $\beta$, rest to 0}
		$\tgamma^{(k+1)} \leftarrow \proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)})$ \tcc*[r]{\label{alg:feature_selection_relax_and_split_outline:tgamma_line}Take max j from $\gamma$, rest to 0}
		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$ \tcc*[r]{\label{alg:feature_selection_relax_and_split_outline:beta_gamma_line} IP Method}
	}
	}
	\BlankLine
\caption{\label{alg:feature_selection_relax_and_split_outline}\ouralgo: Relax-and-Split-based feature selection algorithm for linear mixed-effects models.}	
\end{algorithm}

The algorithm has three alternating minimization steps in its inner cycle. We shall discuss all three steps in details below.

\paragraph{Algorithm \ref{alg:feature_selection_relax_and_split_outline}, lines \ref{alg:feature_selection_relax_and_split_outline:tbeta_line} and \ref{alg:feature_selection_relax_and_split_outline:tgamma_line}: Projections on a $\ell_0$ box.} When $\beta$ and $\gamma$ are fixed the partial minimization solution for $\tbeta$ and $\tgamma$ is just projections of $\beta$ and $\gamma$ on respective $\ell_0$-boxes:

\eq{
	\tbeta^{(k+1)} = \argmin_{\tbeta} \|\tbeta - \beta^{(k)}\|_2^2 \text{ s.t. } \|\tbeta\|_0 \leq k \quad \implies & \quad  \tbeta^{(k+1)} = \text{ max $k$ components from } |\beta_1|, \dots, |\beta_p| \\
	\tgamma^{(k+1)} = \argmin_{\tgamma} \|\tgamma - \gamma^{(k)}\|_2^2 \text{ s.t. } \|\tgamma\|_0 \leq j \quad \implies & \quad  \tgamma^{(k+1)} = \text{ max $j$ components from } |\gamma_1|, \dots, |\gamma_q| \\
	& \text{... and the rest of components set to 0}
}

\paragraph{Algorithm \ref{alg:feature_selection_relax_and_split_outline}, line \ref{alg:feature_selection_relax_and_split_outline:beta_gamma_line}: Interior Point Method.}

As described in Section \ref{section:newton_methods_for_mixed_models}, Interior Point (IP) Methods are a class of second-order methods which come from barrier relaxation of constraints followed by direct application of Newton method to KKT optimality conditions. Following the same idea, we transform the inequality constraints to equality constraints and then relax these constraints via logarithmic barriers:

\eq{
	\label{eq:log_barrier_relaxation_for_ip}
	\min_{\beta, \gamma, d} & \LL(\beta, \gamma) +  \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2 + \mu\sum_{i=1}^{q}\log(d_i) \\
	\text{ s.t. } & \gamma - d = 0
}

where $d \in R^{q}$ are ``offset''  variables, and $\mu$ is a relaxation parameter: the smaller $\mu$ corresponds to ``sharper'' barrier approximation.  Lagrangian for this optimization problem can be written as:

\eq{
	F_\mu(v, d, \beta, \gamma) & = \LL(\beta, \gamma) +  \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2 + \mu\sum_{i=1}^{q}\log(d_i) + v^T(\gamma - d) \\
}

From the necessary optimality conditions on $d$ we have that 
\eq{
	\nabla_d F_\mu = \begin{bmatrix}
		v_1 - \mu/d_1 \\
		\dots \\
		v_q - \mu/d_q 
	\end{bmatrix} = \begin{bmatrix}
		0 \\
		\dots \\
		0 
	\end{bmatrix} \implies v_id_i = \mu \text{ for all } i
}
which can be written in a matrix form as 
\eq{
	\label{eq:complementary_slackness_kkt}
	\diag(v)d - \mu\circ 1 = 0
}
where ``$\circ$'' denotes an element-wise product.

Using $\gamma - d = 0$ to express $F_\mu$ as a function of $v$, $\beta$, and $\gamma$ alone, we can write down the KKT system:
\eq{
	\nabla F_\mu(v, \beta, \gamma) & = \begin{bmatrix}
		\nabla_v F_\mu \\
		\nabla_\beta F_\mu \\
		\nabla_\gamma F_\mu 
	\end{bmatrix} = \begin{bmatrix}
		v\circ \gamma - \mu\circ 1 \\
		\nabla_\beta \LL(\beta, \gamma) + \lambda_\beta(\beta - \tbeta^{(k)}) \\
		\nabla_\gamma \LL(\beta, \gamma) + \lambda_\gamma(\gamma - \tgamma^{(k)}) - v
	\end{bmatrix} = 0
}

Following the the approach of (\ref{eq:ip_system_as_equation}), on each iteration we choose the search direction $[\Delta v, \Delta \beta, \Delta \gamma]$ as the solution of the following system:

\eq{
	\label{eq:ip_iteration_rs_form}
	\nabla^2F_\mu\begin{bmatrix}
		\Delta v \\
		\Delta \beta \\
		\Delta \gamma
	\end{bmatrix} = -\nabla F_\mu\begin{bmatrix}
		\Delta v \\
		\Delta \beta \\
		\Delta \gamma
	\end{bmatrix}
}

Once the search direction $[\Delta v, \Delta \beta, \Delta \gamma]$, is established the method performs a step
\eq{
	v^{(k+1)} & = v^{(k)} + \alpha_k \Delta v \\
	\gamma^{(k+1)} & = \gamma^{(k)} + \alpha_k \Delta\gamma \\
	\beta^{(k+1)} & = \beta^{(k)} + \alpha_k \Delta \beta
}
where the step length $\alpha_k$ is chosen to satisfy positivity and sufficient descent conditions:
\eq{
	\text{\textit{Positivity:}} \qquad & \gamma \geq 0,\ v \geq 0 \\
	\text{\textit{Sufficient Descent:}}\qquad & \|\nabla F_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|\nabla F_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| 
}
On each iteration the relaxation parameter $\mu$ is adjusted as $\mu^{(k+1)} = \frac{{v^{(k)}}^T\gamma^{(k)}}{q}$. The method continues to iterate until the stopping criterion $\|\nabla F_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq \text{\texttt{tol}}$ is met.

The Hessian matrix $\nabla^2 F_\mu$ in Eq. (\ref{eq:ip_iteration_rs_form}) is defined as 
\eq{
	\nabla^2 F_\mu = \begin{bmatrix}
 		\diag(\gamma) & 0 & \diag(v) \\
 		0 & \nabla^2_{\beta\beta}\LL + \lambda_\beta I & \nabla^2_{\beta\gamma} \LL\\
 		-I & \nabla^2_{\gamma\beta}\LL & \nabla^2_{\gamma\gamma}\LL + \lambda_\gamma	 I
 	\end{bmatrix}
}

The precise form of the derivatives $\nabla\LL$ will depend on which likelihood is used: standard maximal likelihood $\LL_{ML}$ or restricted maximum likelihood $\LL_{REML}$. For instance, if the former is used then the derivatives are defined as: 
\eq{
	\label{eq:all_derivatives}
	\nabla_\beta\LL(\beta, \gamma) & = -\sum_{i=1}^m X_i^T\Omega_i^{-1}(Y_i - X_i\beta)\\
	\nabla_\gamma\LL(\beta, \gamma) & = \sum_{i = 1}^m  \diag{{Z_i}^T\Omega_i^{-1}Z_i} - ({Z_i}^T\Omega_i^{-T}(Y_i - X_i\beta))^{\circ 2}\\
	\nabla^2_{\beta\beta}\LL(\beta, \gamma) & =  \half\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i \\
	\nabla^2_{\beta\gamma}\LL(\beta, \gamma) & = -\half\sum_{i = 1}^m X_i^T\Omega_i^{-1}Z_iZ_i^T\Omega_i^{-1}(Y_i - X_i\beta) \\
	\nabla^2_{\gamma\gamma}\LL(\beta, \gamma) & 
	= \half\sum_{i = 1}^m -(Z_i^T\Omega_i^{-T}Z_i)^{\circ 2} + 2Z_i^T\Omega_i^{-T}(Y_i - X_i\beta)((Y_i - X_i\beta)^T\Omega^{-T}Z_i)^T\circ(Z_i^T\Omega^{-1}Z_i)
}

For the sake of completeness we provide the derivation of these formulas in Appendix \ref{appendix:derivatives_of_lmm}, although they can also be found in \cite{Lindstrom1988}.

\subsubsection{Implementation Details}
\paragraph{Hessians through Cholesky Decomposition}
Computing the derivatives in (\ref{eq:all_derivatives}) are the most computationally intensive part of Algorithm \ref{alg:feature_selection_relax_and_split_outline}. Here we provide the way to efficiently compute them via Cholesky decomposition.

Let's define $L_i$ to be a Cholesky factor of the matrix $\Omega_i$:
\eq{
	\Omega_i = L_iL_i^T
}
Having a Cholesky factor $L_i$ for $\Omega_i$ one can calculate a Cholesky factor of the inverse matrix $\Omega_i^{-1}$ in $\OO(max(m, n)^2)$ operations\footnote{See the implementation of LAPACK's \textit{\href{http://www.netlib.org/lapack/explore-html/da/dba/group__double_o_t_h_e_rcomputational_ga97c5ddb103f5f6bc2dc060886aaf2ffc.html}{dtrtri}} as an example.} Let's also define $\xi_i = Y_i - X_i\beta$. Then we can notice that all terms in formulae (\ref{eq:all_derivatives}) can be build out of three building blocks: $L^{-1}_iZ_i$, $L^{-1}_iX_i$, and $L^{-1}_i\xi_i$. Namely:

\eq{
	\label{eq:all_derivatives}
	\nabla_\beta\LL(\beta, \gamma) & = -\sum_{i=1}^m (L^{-1}_iX_i)^T(L_i^{-1}\xi_i)\\
	\nabla_\gamma\LL(\beta, \gamma) & = \sum_{i = 1}^m  \diag{((L_iZ_i)^T(L_iZ_i))} - ((L_i^{-1}Z_i)^T(L_i^{-1}\xi_i))^{\circ 2}\\
	\nabla^2_{\beta\beta}\LL(\beta, \gamma) & =  \half\sum_{i = 1}^m (L_i^{-1}X_i)^T(L_i^{-1}X_i) \\
	\nabla^2_{\beta\gamma}\LL(\beta, \gamma) & = -\half\sum_{i = 1}^m (L_i^{-1}X_i)^T(L_i^{-1}Z_i)(L_i^{-1}Z_i)^T(L_i\xi_i) \\
	\nabla^2_{\gamma\gamma}\LL(\beta, \gamma) & 
	= \half\sum_{i = 1}^m -((L_i^{-1}Z_i)^T(L_i^{-1}Z_i))^{\circ 2} + 2(L_i^{-1}Z_i)^T(L_i^{-1}\xi_i)((L_i^{-1}\xi_i)^T(L_i^{-1}Z_i))^T\circ((L_i^{-1}Z_i)^T(L_i^{-1}Z_i))
}
Apparently, one should calculate bigger building blocks like $(L_iZ_i)^TL_iZ_i$ only once and then reuse it for efficient performance.

\paragraph{Positive Approximation of Hessian}
Since the problem is non-convex with respect to $\gamma$ the Hessian is not in general positive definite. Negative definite Hessians are known to hamper the convergence of second-order methods \cite{nocedal2006numerical}. In order to prevent this we use a positive upper bound as an approximation of the true Hessian. Notice that true Hessian consists of two terms: positive definite and negative definite: 
\eq{
\nabla^2_{\gamma\gamma}\LL(\beta, \gamma) & 
	= \half\sum_{i = 1}^m \underbrace{-(Z_i^T\Omega_i^{-T}Z_i)^{\circ 2}}_{\text{negative definite}} + \underbrace{2Z_i^T\Omega_i^{-T}(Y_i - X_i\beta)((Y_i - X_i\beta)^T\Omega^{-T}Z_i)^T\circ(Z_i^T\Omega^{-1}Z_i)}_{\text{positive definite}}
}
We approximate Hessian with its positive definite part $H(\beta, \gamma)$:
\eq{
\nabla^2_{\gamma\gamma}\LL(\beta, \gamma) \prec H(\beta, \gamma) \equiv 2Z_i^T\Omega_i^{-T}(Y_i - X_i\beta)((Y_i - X_i\beta)^T\Omega^{-T}Z_i)^T\circ(Z_i^T\Omega^{-1}Z_i)
}

\subsection{Experiments on Synthetic Data}
\label{ch:experiments_on_synthetic_data}
\subsubsection{Comparison to Other Algorithms}
\label{ch:comparison_to_other_algo}

\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|l|c|c|c|c|c|}
\hline
Setup & Algoritm & \% C & \% CF & \% CR & MSE & TIME \\
\hline 
\hline
$n = 30$, $n_i = 5$ & \ouralgo & 58 & 72 & 78 & 0.66 & 0.015 \\
& rPQL & 88 & 98 & 88 & 0.88 & 26-59 \\ 
& M-ALASSO & 71 & 73 & 79 & - & -\\
& ALASSO & 79 & 81 & 96 & - & - \\
& SCAD-P & - & 90 & 86 & - & - \\
\hline 
$n = 60$, $n_i = 10$ & \ouralgo & 98 & 100  & 98 & 0.69& 0.018 \\
& rPQL & 98 & 99 & 98 & 0.97 &  26-59 \\ 
& M-ALASSO & 83 & 83 & 89 & - & - \\
& ALASSO & 95 & 96 & 99 & - & -\\
& SCAD-P & 100 & 100 & 100 & - & - \\
\hline

\end{tabular}
\end{center}
\caption{\label{table:krishna_setup_results} Comparison of feature selection algorithms. \% CF -- percent of models where true fixed effects were identified correctly, \% CR -- percent of models where true random effects were identified correctly, \% C -- both fixed and random effects were identified correctly.}	
\end{table}

In the first set of synthetic tests we replicate the experimental setup from \cite{Krishna2008}, which was used to estimate the performance of M-ALASSO. The same setup has been used for evaluating the performance of SCAD-P \cite{Fan2012} and two-stage adaptive LASSO of \cite{Lin2013}, which allows us to compare the performance of \ouralgo relatively to the performance of its competitors. We also include rPQL of \cite{Hui2017} who applied shrinkage operators to PQL quasi-likelihood to facilitate feature selection in Generalized Linear Mixed Models (GLMMs) for the sake of completeness, although rPQL is designed to address a more general setup then \ouralgo.

Two scenarios were considered. In each of them 200 datasets were generated from the multivariate normal distribution: 
\eq{
	y_i \sim \NN(X_i\beta,\ Z_i\Gamma Z_i^T + I)
} 
where:
\begin{itemize}
	\item \textbf{Scenario 1}: $n=30$, $n_i=5$, $p = 9$, $q = 4$, with true parameters $\beta = (1, 1, 0, \dots, 0)$ and the covariance matrix $\Gamma$ being: 
		\eq{
			\Gamma = \begin{bmatrix}
				9 & 4.8 & 0.6 & 0 \\
				4.8 & 4 & 1 & 0 \\
				0.6 & 1 & 1 & 0\\
				0 & 0 & 0 & 0
			\end{bmatrix}
		}
		Columns of $X_i$ and $Z_i$ were generated as i.i.d. vectors from a uniform distribution on $(-2, 2)$, and observation errors were generated from a standard normal distribution.
	\item \textbf{Scenario 2}: everything as in Scenario 1, but $n=60$ and $n_i=10$ to estimate performance on a larger dataset. 
\end{itemize}

The aspect which makes this setup particularly challenging for \ouralgo is that the matrix $\Gamma$ is not diagonal, so the true parameters $(\beta^*, \Gamma^*)$ are out of the model's parameters scope. This inevitably pulls the performance down as the model is unable to take the covariance structure of random effects into account properly, and so the respective model bias is introduced to both estimates of $\beta$ and $u_i$. It's still, however, possible to compare the performance of feature selection by comparing the subset of selected random covariates which is defined by non-zero elements of $\gamma$.

The evaluation results are listed in Table \ref{table:krishna_setup_results}. Data for all algorithms except \ouralgo was borrowed from \cite{Hui2017} who aggregated it from the original works. From this table it's clear that the performance of \ouralgo on Setup 1 is comparable to performance of other algorithms in terms of selecting correct fixed or random effects separately, but is worse in terms of making a correct selection simultaneously. This is expected given that for small datasets the finite sample error adds up to the error from restricting $\gamma$ to be diagonal creating a significant gap in performance. This difference, however, disappears when the algorithms are compared on a bigger Setup 2, where \ouralgo performs on par with its strongest competitors. This behavior suggests that the diagonal $\gamma$ restriction is not as restrictive as it might seem, and our algorithm can enjoy both optimization stability and high feature selection accuracy given a sufficiently large sample size even if true random effects are correlated. Besides, \ouralgo demonstrated lower mean-squared prediction error and four orders of magnitude shorter execution time comparing to rPQL\footnote{Authors of other above mentioned algorithms did not report their execution times.}.

\subsubsection{Scalability Experiments}
\label{ch:scalability_experiments}
\begin{figure}
	\centering
	\includegraphics[width=0.85\textwidth]{Images/scalability_mse_and_acc}
	\caption{\label{fig:scalability_experiment} This is a scalability experiment that examines the performance of \ouralgo on synthetic data. We looked at how algorithm's performance varies when the number of observations is fixed and the number of features changes. In each synthetic dataset a random half of covariates were assigned to be significant and were used to generate observations, the rest was not used. The top figure shows the accuracy of selecting true important covariates depending on the number of covariates available. The bottom figure shows mean squared error of the model's prediction on train and test data. Each synthetic experiment was repeated 300 times, the shaded areas indicate 95\% confidence intervals of the empirical distribution of outcomes of those experiments. Depending on the relation between the number of observations per group ($n_i$), total number of observations ($n = \sum_i n_i$), and the number of features $p$, we see three distinctive "performance zones". In Zone 1, which corresponds to ``tall'' datasets ($n_i \geq p$ for all $i$), algorithm identifies important covariates correctly in 100\% cases. In Zone 2 (``square'' datasets: $n_i < p$ but $n > p$) the performance falls around $p = n/2$ to the performance of random guessing, keeping this poor performance in Zone 3 (``wide'' datasets: $p > n$). Given that the train error is significantly lower than the test error in Zones 2 and 3, we conclude that the resulting models overfit the data when number of covariates is bigger than the number of observations.}
\end{figure}

In the second set of synthetic experiments we want to examine how the feature selection quality of \ouralgo changes when the number of features grows and the number of observations stays fixed.

The setting is similar to \textbf{Scenario 2} from Chapter \ref{ch:comparison_to_other_algo}: $n = 60$, $n_i = 10$, $q$ was set equal to $p$, and also all $X_i$ were equal to $Z_i$ for all $i$. The number of features $p$ was iteratively increasing by blocks of three, where the blocks of new columns $X_i$ and $Z_i$ were sampled i.i.d. from a multivariate normal distribution $\NN(0, \Psi)$ with the variance-covariance matrix $\Psi$:
	\eq{
		\Psi = \begin{bmatrix}
			9 & 4.8 & 0.6 \\
			4.8 & 4 & 1 \\
			0.6 & 1 & 1 \\
		\end{bmatrix}
	}

	For each experiment, random half of all features were active fixed features with 70\% of those active fixed features also having a random effect. To be precise, $\beta$ had a random 50\% subset of its coordinates equal to 1 with zeros elsewhere. Random 70\% out of those non-zero coordinates in $\beta$ had 1s in respective positions in $\gamma$ ($\Gamma := \diag(\gamma)$), with the rest coordinates of $\gamma$ to be 0.

	The results are presented on Figure \ref{fig:scalability_experiment}. One can immediately notice that the behavior of plots splits the picture into three different zones:
	\begin{enumerate}
		\item \textbf{Zone 1: ``tall'' datasets.} This zone corresponds to the situation when $n_i \geq p$ for all $i$, so every group has more observations than there are covariates in the dataset. The performance of the algorithm in this zone is nearly perfect, with all correct covariates to be selected. The train error is nearly the same to the test error which evidences that the model generalizes the data properly.
		\item \textbf{Zone 2: ``square'' datasets.} This zone corresponds to the situation when $n_i < p$ but $n = \sum_i n_i > p$, so a problem of fitting individual regression models would lead to an underdetermined system for each group, but the problem of fitting a linear mixed regression as a whole is still well-defined. Although for $p \leq n/2$ the performance is still nearly perfect, there is a clear drop in performance in terms of both selection accuracy and RMSE on test data once $p \geq n/2$. This is a strong sign that the model overfits the data which leads to generalization errors.
		\item \textbf{Zone 3: ``wide'' datasets.} This zone corresponds to solving underdetermined systems where $p > n$. We see that the performance of the method is roughly equal to performance of random guessing, which means that \ouralgo is not suitable for learning from ``wide'' datasets at this point.
	\end{enumerate}
	This analysis shows that the method can be successfully applied to the problems up until $p > n/2$, with ``wide'' dataset setups to be an apparent space for improvement.

\subsection{Experiments on Real Data}
\label{ch:experiments_on_real_data}
\subsubsection{Meta-Research on Internet Bullying} 
\label{ch:bullying}
\begin{figure}
	\includegraphics[width=0.95\textwidth]{Images/bullying_data.csv_intercept_only}
	\caption{\label{fig:bullying_data_intercept_only}Model fit for Internet Bullying Data from GBD 2020 using only intercept and time. We see that the groups are not separable, but the fit of a mixed model is reliable: there is a distinctive minimum of the loss function around $\gamma=0.17$.}
\end{figure}

\begin{figure}
	\caption{\label{fig:bullying_data_fixed_feature_selection}Fixed Feature Selection for Bullying Data from GBD 2020. The algorithm picks five historically significant covariates and two historically insignificant. See the Appendix \ref{appendix:bullying_covariates} for covariates description and assessment of significance. }
	\includegraphics[width=0.92\textwidth]{Images/bullying_data.csv_fixed_feature_selection}
\end{figure}

\begin{figure}
	\caption{\label{fig:bullying_data_random_feature_selection}Random Feature Selection for Bullying Data from GBD 2020. The algorithm picks five historically significant covariates and two historically insignificant. See the Appendix \ref{appendix:bullying_covariates} for covariates description and assessment of significance.}
	\includegraphics[width=0.92\textwidth]{Images/bullying_data.csv_random_feature_selection}
\end{figure}

In this section we apply \ouralgo to identify the most important covariates in research on relative risk of anxiety and depression disorders depending on the exposure to bullying in young age\footnote{The data is available at ...}. This research has been a part of Global Burden of Diseases study for the last several years. The end goal for this work was to estimate the burden (DALYs) of major depressive disorder (MDD) and anxiety disorders that caused by bullying. For this risk factor, the exposure is primarily concentrated in childhood and adolescents, but the risk for MDD and anxiety disorders is anticipated to continue well into adulthood. This elevated risk is, however, expected to decrease with time as other risk factors come into play in adulthood (unemployment, relationship issues, etc.). To accommodate this, the research team uses the models which estimate the relative risk (RR) of MDD and anxiety disorders among persons exposed to bullying depending on how many years it has been since the first exposure. Studies informing the model were sourced from a systematic review and consist of longitudinal cohort studies. They measure exposure to bullying at baseline, and then follow up years later and assess them for MDD or anxiety disorders. The detailed description of the covariates can be found in Appendix \ref{appendix:bullying_covariates} .

The feature selection process is illustrated on Figures \ref{fig:bullying_data_fixed_feature_selection} and \ref{fig:bullying_data_random_feature_selection}. Since there was no apparent prior on $k$ -- the number of features to keep in the model -- an AIC criterion from \cite{Vaida2005} was used to make a choice (see the plot on the first row, column two on Figure \ref{fig:bullying_data_fixed_feature_selection}). We see that there is a minimum around $k=9$, although the minimum is shallow so $k=8$ can also be considered. Five extra covariates (the ones outside \texttt{intercept} and \texttt{time}) were selected by \ouralgo: \texttt{cv\_unadjisted}, \texttt{cv\_threshold\_bullying},\texttt{cv\_b\_parent\_only}, \texttt{cv\_anx} and \texttt{percent\_female} (have been important in the previous years of GBD). The covariates \texttt{cv\_or} and \texttt{cv\_child\_baseline} has not been important previously but also got selected.

\subsubsection{COVID-19 Contact Rate Forecasting Data}
\label{ch:covid}
\begin{figure}
	\begin{subfigure}[b]{\textwidth}
		\includegraphics[width=\textwidth]{Images/fit_Alaska}
	\end{subfigure}
	\begin{subfigure}[b]{\textwidth}
		\includegraphics[width=\textwidth]{Images/fit_Slovenia}	
	\end{subfigure}
	\caption{\label{fig:covid_feature_selection_1} Comparison of fits of two different models (fully dense linear mixed model and \ouralgo) to the original IHME Projections for Alaska and Slovenia. The quality (RMSE) achieved by a sparse fit is within 10\% from a quality of both dense models which is evidences that the model picked up informative features.}
\end{figure}

\begin{figure}
	\begin{subfigure}[b]{\textwidth}
		\includegraphics[width=\textwidth]{Images/fit_Turkey}
	\end{subfigure}
	\begin{subfigure}[b]{\textwidth}
		\includegraphics[width=\textwidth]{Images/fit_Switzerland}	
	\end{subfigure}
	\caption{\label{fig:covid_feature_selection_2} Comparison of fits of two different models (fully dense linear mixed model and \ouralgo) to the original IHME Projections for Turkey and Switzerland. The quality (RMSE) achieved by a sparse fit is within 10\% from a quality of both dense models which is evidences that the model picked up informative features.}
\end{figure}

In this section we apply our method to a COVID-19 Contact Rate Forecasting problem. The global pandemic created an unprecedented need for robust and accurate disease transmission forecasting. Since the beginning of the pandemic Institute for Health Metrics and Evaluation has been providing guidance to local authorities across the world with their pioneering \href{https://covid19.healthdata.org/global?view=total-deaths&tab=trend}{COVID-19 Projections tool} \cite{IHME2020Projections}. The key methodology which was essential for success in their forecasting of the disease's dynamics was transforming the death data into the contact rate time series data, and then relating this contact rate to available covariates such as temperature, social mobility, population, and others \cite{IHME2020covidnature}. All these covariates were collected in real time using limited human resources, and identifying the most important covariates was crucial for distributing these resources effectively. In the example below we show how \ouralgo can be used for making covariate selection on IHME data.

The dataset consists of $m=60$ groups (countries or states), the detailed description of groups sizes ($n_i$) and time spans is provided in the Table \ref{table:covid_data_description} in the Appendix. The target variable $Y_i$ was the contact rate -- the coefficient $\beta(t)$ from an SEIR model (not to be confused with $\beta$ -- vector of fixed effects). The covariates -- columns of $X_i$ and $Z_i$ -- were: \texttt{intercept} --a column of ones,  \texttt{temperature} -- average air temperature in degrees Fahrenheit, \texttt{mobility\_lift} -- social mobility, \texttt{proportion\_over\_1k} -- population size threshold, and \texttt{testing\_reference} -- testing. The observation error's variance $\sigma_i$ was set to be 0.1.

Feature selection results for four particular locations (Alaska, Slovenia, Turkey, and Switherland) are presented on Figures \ref{fig:covid_feature_selection_1} and \ref{fig:covid_feature_selection_2}, with coefficients for the rest of the locations attached in supplementary materials. \ouralgo was charged with a task to produce a fit using only three covariates, one of which had to be mean-only (no random effects). On the left we see the original data (\texttt{Data}, in {\color{blue} blue}), and predictions of three models: IHME Projections (\texttt{IHME}, in {\color{orange} orange}), a linear mixed model fit with no selection (\texttt{Dense MM}, in {\color{red} red}), and a sparse fit of \ouralgo (\ouralgo in {\color{green} green}). The \ouralgo model has chosen to use \texttt{intercept}, \texttt{mobility\_lift}, and \texttt{temperature} covariates, with the later only as a fixed covariate; \texttt{testing\_reference} and \texttt{proportion\_over\_1k} were left out. On the plots to the left we see that the exclusion of two covariates did not significantly affect the quality of predictions. The residual errors (RMSE) to the right support this statement: the difference in RSME is within 10\% of RSME of a "dense" model which uses all the covariates, as well as from IHME's original model which fit all the locations via sequence of independent regressions. This choice also seems reasonable given the timespan: the proliferation of testing during the spring was not yet significant, so it did not inform predictions in a major way. The exclusion of \texttt{proportion\_over\_1k} could have been due to the scale of grouping: locations were grouped on the level of states and countries, not on the level of individual counties where the influence of population-based covariates could have been more significant.

\subsection{Software Implementation}
In order to ensure reproducibility of our research, the proposed algorithm (\ouralgo, Algorithm \ref{alg:feature_selection_relax_and_split_outline}) has been implemented as a part of \texttt{\href{https://github.com/aksholokhov/skmixed}{skmixed}}\footnote{Available at \href{https://github.com/aksholokhov/skmixed}{https://github.com/aksholokhov/skmixed}} library. This library implements functionality for fitting linear mixed models and selecting covariates. The user interface was designed to be fully compliant with the standards\footnote{\href{https://scikit-learn.org/stable/developers/develop.html}{https://scikit-learn.org/stable/developers/develop.html}} of \texttt{sklearn} library to minimize learning time.

\newpage
\section{Discussion and Future work}
\label{ch:future_work}

Although the proposed algorithm performs on par with its competitors in its current form (Algorithm \ref{alg:feature_selection_relax_and_split_outline}), there are several improvements which the author plans to work on in future. In this section we discuss these improvements and outline possible paths to them. We plan to test and deploy these methods on global health applications through our collaborations with researchers at the Institute for Health Metrics and Evaluation.

\subsection{Better Control over Relaxation Parameters}

Both Relax-and-Split and Interior Point methods introduce relaxations to the problem. In particular, Relax-and-Split approach relaxes $\ell_0$-constraints using auxiliary variables, which gives us relaxation parameters $\lambda_\beta$ and $\lambda_\gamma$, whereas IP method relaxes equality constraints in (\ref{eq:log_barrier_relaxation_for_ip}) introducing another parameter $\mu$. All three parameters $\lambda_\beta$, $\lambda_\gamma$, and $\mu$ need to be tightened iteratively ($\lambda_\beta$, $\lambda_\gamma \to \infty$ and $\mu \to 0$) to guarantee that the variables $\beta$ and $\gamma$ satisfy the original constraints from (\ref{eq:fs_loss_intractable}). One can see this triplet as one big relaxation for the original problem and tune them all at the same place.

 For the IP method $\mu$ is set to via a complementary slackness condition:
 \eq{
 	\mu^{(k)} := \sum_{i=1}^m \frac{v_i^{(k)}\gamma_i^{(k)}}{m}
 }
 that come from necessary optimality conditions on a KKT-system (\ref{eq:complementary_slackness_kkt}). At the same time, $\lambda_\beta$ and $\lambda_\gamma$ are increased in a static and rather arbitrary fashion: $\lambda := 2(1+\lambda)$. This approach is effective but not efficient: \cite{Zheng2018RelaxAndSplit} showed that the convergence is guaranteed, but one needs to iteratively solve the IP relaxation in full every time $\lambda_\beta$ and $\lambda_\gamma$ are increased. Adjusting them in the same place where $\mu$ is adjusted does not work because the steady exponential growth in $\lambda$ drives $\|F_\mu\|$ up faster than Newton method drives it down over iterations, which breaks the stopping criterion of IP method $\|F_\mu\| \leq \varepsilon$.\\

The proposed approach is to increase $\lambda_\beta$ and $\lambda_\gamma$ based on the current step length $\alpha_k$. The logic behind this choice is that if the relaxation parameters are too large then the method will struggle with making a step away from the projections $\tbeta$ and $\tgamma$, and so no further increase is required. The key theorem which lays a theoretical foundation behind it would be:
\begin{theorem}[Distance Between Minima]
\label{thm:distance_between_minima}
	For a fixed dataset $(X_i, Y_i)$ and relaxation parameters $\lambda_\beta$, $\lambda_\gamma$ the distance between $(\beta^*, \gamma^*)$, the unconstrained minimizers of relaxed problem, and their projections $(\tbeta^*, \tgamma^*)$ is bounded by a constant $M$ depending on $(X_i, Y_i)$ and the relaxation parameters.
\end{theorem}

There are two main reasons why the joint relaxation is more attractive than the current version. First, it will simplify the Algorithm \ref{alg:feature_selection_relax_and_split_outline} by removing the nested loop. This will accelerate the algorithm in terms of time and the number of iterations as both relaxations will be tuned within the same cycle with no further need to solve one relaxation per each step of another relaxation. Second, these theoretical findings may lead to developing a universal tuning technique for Relax-and-Split methods and might be used in other applications outside of mixed models setting.

\subsection{Asymptotic Properties}

Although \ouralgo can be classified as a penalized likelihood based method, regularity conditions from \cite{Fan2001}, which are being widely used by competitor algorithms, are not applicable here directly because the penalty is not of the form $P_{\lambda, n}(|\beta|, |\gamma|)$. Hence, a separate analysis of statistical properties of the proposed estimator is needed.

The first theorem establishes the foundation for proving three consequent theorems by outlining the conditions under which the proposed method is guaranteed to find the right $k$-subspace:

\begin{theorem}[Conditions for Convergence to True Estimator]
	Under certain conditions the method converges in a finite number of iterations to $(\hat\beta, \hat\gamma)$ which projections $(\tbeta, \tgamma)$ belong to a $k$- and $j$-subspaces respectively that contain the true minimum $(\beta^*,\gamma^*)$.
\end{theorem}
The key challenge is to find the set of conditions which satisfy the theorem and at the same time are sufficiently general to cover real-world cases. For example, the conditions of $X_i$ and $Z_i$ to be sets of orthonormal vectors are known to be sufficient due to \cite{Tibshirani1996}, but without costly preprocessing the real world datasets are rarely orthonormal matrices.

The next step is to prove a set of statements which is normally discussed in this type of works (see, for example \cite{Krishna2008, Hui2017}) using Theorem \ref{thm:distance_between_minima}.

\begin{theorem}[Consistency of Estimator]
	There exists a local minimizer $(\hat{\beta}, \hat{\gamma})$ for the proposed loss function, such that it is asymptotically consistent with true minimum $(\beta^*, \gamma^*)$. 
\end{theorem}
\textit{Proof Outline.} The proof is based on the fact that the set of stationary points for the original formulation (\ref{eq:fs_loss_intractable}) matches to the set of stationary points of the relaxation $\LL_r$ (\ref{eq:lmm_relax_and_split_full_setup}) when $\lambda_\beta, \lambda_\gamma \to \infty$. Hence, the first step would be providing a concentration bounds on how far a stationary point $(\hat{\beta}, \hat{\gamma})$  of the relaxation $\LL_r(\beta, \gamma)$ can be from a stationary point $(\beta^*, \gamma^*)$ of the original setup $\LL(\beta, \gamma)$ given a fixed $\lambda_\beta \gg 1$ and $\lambda_\gamma \gg 1$. This bound \textit{should be} $n$-independent meaning that the minima will match for any finite number of samples $n$.

\begin{theorem}[Consistency in Zeros]
	If some coordinates of the true minimizer ($\beta^*, \gamma^*$) are zero, then it is also zero in $(\hat{\beta}, \hat{\gamma})$, given that the later is sufficiently close to the former.
\end{theorem}

\textit{Proof Outline.} This is similar to the previous statement: the fact that the limiting minima of the relaxation $\LL_r(\beta, \gamma)$ match the minima of the original setup $\LL_{ML}(\beta, \gamma)$ implies that their zero coordinates also match.

\begin{theorem}[Asymptotic Normality]
	The proposed estimator $(\hat{\beta}, \hat{\gamma})$ asymptotically normally distributed around true minimizer ($\beta^*, \gamma^*$) in its true non-zero $k+j$-subspace.
\end{theorem}

These proofs would not only contribute to this particular work, but would also be a valuable development of Relax-and-Split approach in general. This is ongoing work which should be done by the time of Thesis Defense.

\subsection{Exponentially Smoothed Projection}
One of the aspects of Algorithm \ref{alg:feature_selection_relax_and_split_outline} which drags the selection accuracy downwards is premature, all-in, guessing of the sparse subspaces. Namely, $\tbeta$ and $\tgamma$ are assigned to be the direct projections of $\beta$ and $\gamma$, which may not identify the right subspace correctly during the first iterations when the problem is significantly relaxed. At the same time, the attraction to $\tbeta$ and $\tgamma$ chosen on the previous iteration biases the next $\beta$ and $\gamma$. One way of mitigating this effect is using exponential smoothing. In particular, lines \ref{alg:feature_selection_relax_and_split_outline:tbeta_line} and \ref{alg:feature_selection_relax_and_split_outline:tgamma_line} in the Algorithm \ref{alg:feature_selection_relax_and_split_outline} can be changed to the weighted exponential average. The proposed modification is outlined in Algorithm \ref{alg:exp_smoothing}.

\begin{algorithm}
	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
	\Repeat{converges}{
		$\tbeta^{(k+1)} \leftarrow \delta\proj_{\|\beta\|_0 \leq k}(\beta^{(k)}) + (1-\delta)\tbeta^{(k)}$\\
		$\tgamma^{(k+1)} \leftarrow \delta\proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)}) + (1-\delta)\tgamma^{(k)}$\\
		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$\\
	}
	}
	\BlankLine
\caption{\label{alg:exp_smoothing} A prospective version of \ouralgo with exponential smoothing for projected variables. This modification should increase accuracy of covariates selection.}	
\end{algorithm}

The hyper parameter $\delta$ can be assigned via cross-validation tuning. Alternatively, it's possible to analyze this scheme via Nesterov's Accelerated Methods framework \cite{Nesterov2004} or Polyak's Heavy Ball framework \cite{Polyak1964} in hope that a proper choice of $\delta$ will increase the convergence rate of the algorithm.

\subsection{Correlated Features}
In Algorithm \ref{alg:feature_selection_relax_and_split_outline}, when we project $\beta$ onto a $k$-subspace constraint to get $\tbeta$ we take $k$ maximum elements in their absolute value from $\beta$ and put them into $\tbeta$, setting the rest to $0$. This can also be viewed as a minimization of a local quadratic model with a (trivial) approximation of Hessian with an identity matrix. Namely, for the iteration $j$:

\eq{
	\tbeta^* := & \argmin_{\tbeta} \LL_r(\beta^{(j)}, \gamma^{(j)}) + \nabla_\beta\LL_r(\beta^{(j)}, \gamma^{(j)})^T(\tbeta - \beta^{(j)}) + \half(\tbeta - \beta^{(j)})^TI(\tbeta - \beta^{(j)}) \\
	& \text{ s.t. } \|\tbeta\|_0 \leq k
}
Since  $(\beta^{(j)}, \gamma^{(j)})$ is a partial minimizer of $\LL_r(\beta, \gamma, \tbeta, \tgamma)$ for the iteration $j$, we have that the second term equal to $0$. Notice $\tbeta^*$ would be a correct minimizer of the quadratic approximation of the loss function \textit{given that Hessian can be approximated by an identity}, which is not always the case. As we have from (\ref{eq:all_derivatives}):

\eq{
	\nabla^2_{\beta\beta}\LL(\beta, \gamma) & =  \half\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i
}
which is equal to $\half I$ in, for instance, the case of "orthonormal design": when all columns of $X_i$ are orthonormal for all $I$. This is not the case in general when fixed features are correlated, which suggests that \ouralgo might not work as well in a setup with highly correlated features. This formulation also suggests an extension:

\eq{
	\tbeta^* := & \argmin_{\tbeta} \LL_r(\beta^{(j)}, \gamma^{(j)}) + \nabla_\beta\LL_r(\beta^{(j)}, \gamma^{(j)})^T(\tbeta - \beta^{(j)}) + \half(\tbeta - \beta^{(j)})^T(I + \nabla^2_{\beta\beta}\LL(\beta^{(j)}, \gamma^{(j)}))(\tbeta - \beta^{(j)}) \\
	& \text{ s.t. } \|\tbeta\|_0 \leq k
}
This is a ``quadratic knapsack'' problem which is NP-complete. Nevertheless, there are effective approximation methods for solving this problem. The hope is that this version of the algorithm will perform ultimately better in the setup where covariates are highly correlated.

\bibliographystyle{apalike}
\bibliography{bibliography}
\clearpage

\appendix
\section*{Acknowledgement}
I am extremely grateful to my advisor, Sasha Aravkin, for providing support, guidance, and enthusiasm during my work on this project in the Department of Applied Mathematics and in Institute for Health Metrics and Evaluation.\\
I would also like to thank Jim Burke for his detailed review of my work which would hopefully lead to fruitful collaboration. \\ 
Last, but not least: I am grateful to Damian Santomauro, who introduced me to his work on the consequences of bullying, provided with a dataset, and with a valuable feedback on the performance of the proposed method. \\

\section{Derivatives of Marginalized Log-likelihood for Linear Mixed Models}
\label{appendix:derivatives_of_lmm}
For conciseness, let us define the mismatch $\xi_i = Y_i - X_i\beta$. We also omit the dependence on $\beta$, as it's fixed at this point. The loss function \ref{eq:lmm_objective} takes the form 
\eq{
	\label{eq:lmm_simplified_gamma}
	\LL\pa{\gamma} = \sum_{i = 1}^m\half\xi_i^T\pa{\Omega_i(\gamma)}^{-1}\xi_i + \half\log\det\pa{\Omega_i(\gamma)}.
}

The derivative of the objective w.r.t $\gamma_j$, the $j$'th diagonal element of the matrix $\Gamma$ is

\eq{
	\pderiv{\xi_i^T\Omega_i^{-1}\xi_i}{\Gamma_{jj}} & = \trace \left[\left(\pderiv{\xi_i^T\Omega_i^{-1}\xi_i}{\Omega_i}\right)\pderiv{\Omega}{\Gamma_{jj}}\right] = \trace\left[\left(-\Omega_i^{-T}\xi_i\xi_i^T\Omega_i^{-T}\right)^T Z_i\pderiv{\Gamma}{\Gamma_{jj}}Z_i^T\right] =  
}

where $\pderiv{\Gamma}{\Gamma_{jj}}$ is a structure matrix, which, in a general case, is equal to a single-entry matrix $J^{jj}$ with $jj$'th element is equal to 1 and zeroes elsewhere. Substituting this back we get
\eq{
	= \trace\left[\left(-\Omega_i^{-T}\xi_i\xi_i^T\Omega_i^{-T}\right)^T Z^j_i{Z^j_i}^T\right] = 
} 
where $Z^j_i$ is a $j$'th column of the matrix $Z_i$. Making a circular swap we end up with

\eq{
	= \trace\left[-{Z^j_i}^T\Omega_i^{-T}\xi_i\xi_i^T\Omega_i^{-T} Z^j_i\right] = - ({Z_i^j}^T\Omega_i^{-T}\xi_i)^2
}

Similarly,

\eq{
	\pderiv{\log\det\Omega_i}{\Gamma_{jj}} = \trace\left[\left(\pderiv{\log\det\Omega_i}{\Omega_i}\right)\pderiv{\Omega_i}{\Gamma_{jj}}\right] = \trace\left[\Omega_i^{-1}Z_i^j{Z_i^j}^T\right] = {Z_i^j}^T\Omega_i^{-1}Z_i^j
}

Taking into account that $\Omega_i$ is symmetric, we have

\eq{
	\left[\nabla_\gamma \LL\pa{\beta, \gamma}\right]_j = \sum_{i = 1}^m  - ({Z_i^j}^T\Omega_i^{-T}\xi_i)^2 + {Z_i^j}^T\Omega_i^{-1}Z_i^j = 
}

or, in vector form

\eq{
    = \sum_{i = 1}^m  \diag{{Z_i}^T\Omega_i^{-1}Z_i} - ({Z_i}^T\Omega_i^{-T}\xi_i)^{\circ 2} =
}

where $\circ$ denotes the Hadamard (element-wise) product. Using the Cholesky decomposition $\Omega_i = L_iL_i^T$ we can calculate it more effectively, using only one triangular matrix inversion:

\eq{
    = \sum_{i = 1}^m \left[ \sum_{\text{rows}}\left(L_i^{-1}Z_i\right)^{\circ 2} -  [(L_i^{-1}Z_i)^{T}(L_i^{-1}\xi_i)]^{\circ 2}\right]
}

Notice, that the loss function (\ref{eq:lmm_objective}) and the optimal $\beta$ solution (\ref{eq:beta_formula}) can also be effectively computed using Cholesky:

\eq{
	\LL\pa{\gamma} = \sum_{i = 1}^m \half\xi_i^T\pa{\Omega_i(\gamma)}^{-1}\xi_i + \half\log\det\pa{\Omega_i(\gamma)} = \sum_{i = 1}^m \half\ \|L_i^{-1}\xi_i\|^2 - \sum_{j = 1}^k \log{[L^{-1}_i]}_{jj}
}

\eq{
	\beta_{k+1} & = \argmin_{\beta}\LL(\beta, \gamma_{k}) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i = \\
	& = \left(\sum_{i = 1}^m (L_i^{-1}X_i)^TL_i^{-1}X_i\right)^{-1}\sum_{i = 1}^m (L_i^{-1}X_i)^TL_i^{-1}y_i 
}

The Hessian w.r.t. $\gamma$ also can be found:

\eq{
	\frac{\partial^2 \LL\pa{\beta, \gamma}}{\partial \gamma_j^2} & = \sum_{i=1}^m -2({Z_i^j}^T\Omega_i^{-T}\xi_i)\trace\left[\frac{\partial{Z_i^j}^T\Omega_i^{-T}\xi_i}{\partial \Omega_i}\frac{\partial \Omega_i}{\partial \Gamma_{jj}} \right] + \trace\left[\frac{\partial{Z_i^j}^T\Omega_i^{-1}Z_i^j}{\partial \Omega_i}\frac{\partial \Omega_i}{\partial \Gamma_{jj}} \right] = \\
	& = \sum_{i=1}^m 2({Z_i^j}^T\Omega_i^{-T}\xi_i)\trace\left[\Omega_i^{-1}Z_i^j{\xi_i}^T\Omega_i^{-1}Z_i^j{Z_i^j}^T \right] - ({Z_i^j}^T\Omega_i^{-T}Z_i^j)^2 = \\
	& = \sum_{i=1}^m 2({Z_i^j}^T\Omega_i^{-T}\xi_i)({Z_i^j}^T\Omega_i^{-1}Z_i^j)({\xi_i}^T\Omega_i^{-1}Z_i^j) - ({Z_i^j}^T\Omega_i^{-T}Z_i^j)^2
}

\eq{
	\frac{\partial^2 \LL\pa{\beta, \gamma}}{\partial \gamma_j\partial \gamma_k} & = \sum_{i=1}^m -2({Z_i^j}^T\Omega_i^{-T}\xi_i)\trace\left[\frac{\partial{Z_i^j}^T\Omega_i^{-T}\xi_i}{\partial \Omega_i}\frac{\partial \Omega_i}{\partial \Gamma_{kk}} \right] + \trace\left[\frac{\partial{Z_i^j}^T\Omega_i^{-1}Z_i^j}{\partial \Omega_i}\frac{\partial \Omega_i}{\partial \Gamma_{kk}} \right] = \\
	& = \sum_{i=1}^m 2({Z_i^j}^T\Omega_i^{-T}\xi_i)\trace\left[\Omega_i^{-1}Z_i^j{\xi_i}^T\Omega_i^{-1}Z_i^k{Z_i^k}^T \right] - ({Z_i^j}^T\Omega_i^{-T}Z_i^k)^2 = \\
	& = \sum_{i=1}^m 2(\xi_i^T\Omega_i^{-T}{Z_i^j})({Z_i^j}^T\Omega_i^{-1}Z_i^k)({Z_i^k}^T\Omega_i^{-1}{\xi_i}) - ({Z_i^j}^T\Omega_i^{-T}Z_i^k)^2
}

\eq{
	\label{eq:lmm_hessian_gamma}
	\nabla^2_\gamma\LL\pa{\beta, \gamma} = \half\sum_{i = 1}^m -(Z_i^T\Omega_i^{-T}Z_i)^{\circ 2}  + 2\diag(Z_i^T\Omega_i^{-T}\xi_i)(Z_i^T\Omega^{-1}Z_i)\diag(\xi_i^T\Omega^{-T}Z_i) = \\
	= \half\sum_{i = 1}^m -(Z_i^T\Omega_i^{-T}Z_i)^{\circ 2} + 2(Z_i^T\Omega_i^{-T}\xi_i)(\xi_i^T\Omega^{-T}Z_i)^T\circ(Z_i^T\Omega^{-1}Z_i)
}

\section{Description of Datasets}
\subsection{GBD Bullying Data}
\input{tables/bullying_features_description}

\subsection{COVID-19 Contact Rate Forecasting Data}
\input{tables/covid_groups_description}
\input{tables/covid_coefficients}

\end{document}

### 2022-07-04 presentation/general_presentation.tex
Import thesis proposal from the general exam as a starting point for my PhD thesis

\documentclass[8pt]{beamer}
\input{useful_packages}
\input{math_macros}

% \usetheme[progressbar=frametitle]{metropolis}
\usepackage{appendixnumberbeamer}

% PRESETS OF PACKAGES AND MATH %

\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
\newcommand{\ouralgo}{R\&S-Mixed }
\title{Feature Selection for Mixed-Effects Models}
\date{Thursday, 3\textsuperscript{rd} of December, 2020}
\author{Aleksei Sholokhov}
% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}

\begin{document}

\maketitle

\begin{frame}{Plan}
	\tableofcontents
\end{frame}

\section{Introduction}
\subsection{Linear Mixed-Effects Models}
\begin{frame}{Linear Mixed-Effect Models are a type of regression models for grouped data}

	\[
	\text{Dataset: m groups } (X_i, y_i),\quad i = 1, \dots m
    \]
	\begin{columns}[T,onlytextwidth]
    \column{0.5\textwidth}
	 \centering Standard Linear Regression:
   	\begin{figure}
   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_mean_prediction}
   	\end{figure}
   	\[
   		y = X\beta + \varepsilon, \quad \varepsilon \sim \NN(0, \Lambda) 
   	\]

    \column{0.5\textwidth}
    	\centering  Linear Mixed-Effect Model:
   	\begin{figure}
   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_random_prediction}
   	\end{figure}
   	   		\[
   		\begin{split}
   			y_i & = X_i(\beta + {\color{red} u_i}) + \varepsilon_i, \quad \varepsilon_i \sim \NN(0, \Lambda_i) \\
   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma)}
   		\end{split}
   		\]

  \end{columns}
\end{frame}

\begin{frame}{Notation}
	\eq{
   		y_i & = \alt<3->{X_i{\color<4,9>{red}\beta} + {\color<3>{red}Z_i}{\color<5,9>{red}u_i}}{{\color<2>{red}X_i}(\beta + u_i)} + \varepsilon_i \quad i = 1\dots m \\
   		{\color<7>{red}\varepsilon_i} & \sim \NN(0, {\color<8,9>{red}\Lambda_i}) \\
   		u_i & \sim \NN(0, {\color<6,9>{red}\Gamma})
   	}

   	\begin{itemize}
   		\item<2-> $X_i \in \R^{n_i \times p}$ -- \only<3>{{\color{red} fixed}}\only<4->{fixed} features, or covariates, \only<3>{{\color{red} $Z_i \in \R^{n_i \times q}$ -- random features }}\only<4->{ $Z_i \in R^{n_i \times q}$ -- random features}
   		\item<3-> $p$ -- number of fixed features, $q$ -- number of random effects.
   		\item<4-> $\beta \in \R^p$ -- fixed effects, or mean effects
   		\item<5-> $u_i \in \R^q$ -- random effects
   		\item<6-> $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects
   		\item<7-> $\varepsilon_i \in \R^{n_i}$ -- observation noise
   		\item<8-> $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for noise
   		\item<9-> Unknowns: $\beta$, $u_i$, $\Gamma$, sometimes $\Lambda_i$.

   	\end{itemize}
   	\uncover<10->{
   	\[
   		\Gamma = \diag{({\color{red}\gamma})},\ {\color{red}\gamma \in \R^{q}} \quad \text{ and } \Lambda_i \text{ are known.}
   	\]
   	}
\end{frame}

\begin{frame}[fragile]{Likelihood for Mixed Models}
Suppose we have a linear mixed effects model

\eq{
	\label{eq:lmm_setup}
	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i \\
	u_i & \sim \NN(0, \Gamma), \ \Gamma = \diag{\gamma} \\
	\varepsilon_i & \sim \NN(0, \Lambda_i)\\
}
where $\beta$, $\gamma$, and $u_i$ are unknown model parameters, and $\lambda_i$ are known.

	Noticing that $Z_iu_i + \varepsilon_i \sim \NN(0, Z_i\Gamma Z_i^T + \Lambda_i)$ we get the negative log-likelihood:
	 \eq{
	\label{eq:lmm_objective}
	\mathcal{L}(\beta, \gamma) & = \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \\ & + \half\log{\det{\pa{Z_i \Gamma Z_i^T + \Lambda_i}}}
	}

\end{frame}

\subsection{Feature Selection for Mixed Models}
\begin{frame}{Feature Selection for Mixed Models}

Consider the following feature selection setup:

\eq{
	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
	\text{s.t. } & \|\beta\|_0 \leq k \\
	& \|\gamma\|_0 \leq j \\
	& \gamma \geq 0
}

where $k$ and $j$ is the size of the sparse support for the fixed and random effects respectively.

This is a very hard problem to solve directly.

\end{frame}

\begin{frame}{Methods for Feature Selection}
There are several ways to address this setup:
  \begin{columns}[T,onlytextwidth]
    \column{0.33\textwidth}
    (0) Exhaustive Search:
\[
	\begin{split}
	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
	\text{s.t. } & \|\beta\|_0 \leq k \\
	& \|\gamma\|_0 \leq j \\
	& \gamma \geq 0
	\end{split}
\]

    \column{0.33\textwidth}
    \uncover<2->{
    \centering
    (1) $\ell_1$-relaxation:
\[
	\begin{split}
	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
	\text{s.t. } & \|\beta\|_{\color{red}1} \leq {\color{red} \lambda_\beta} \\
	& \|\gamma\|_{\color{red}1} \leq {\color{red} \lambda_\gamma} \\
	& \gamma \geq 0
	\end{split}
\]
	}
	\column{0.33\textwidth}
	\uncover<3->{
    \centering
    (3) Other types of sparsity-inducing penalties\\ ($\ell_p$, $SCAD$).
    }
  \end{columns}

 \onslide<1-3>{
	\begin{columns}[T, onlytextwidth]
	\column{0.33\textwidth}
	\onslide<1-3>{
	\centering
	AIC \cite{Vaida2005}\\
	BIC \cite{Jones2011}\\
	}
	\column{0.33\textwidth}
	\onslide<2-3>{
	\centering
	M-ALASSO \cite{Bondell2010} \\
	ALASSO \cite{Lin2013} \\
	}
	\column{0.33\textwidth}
	\onslide<3>{
	\centering
	SCAD \cite{Fan2001}\\
	SCAD-P \cite{Fan2012}
	}
	\end{columns}
}

\onslide<4->{
	\begin{center}
	(2) SR3-type Relaxation \cite{Zheng2019}
	\eq{
	\min_{\beta, \gamma, {\color{red}\hbeta}, {\color{red}\tgamma}} & \LL(\beta, \gamma) + {\color{red}\frac{\lambda_\beta}{2}\|\beta - \tbeta\|^2_2 + \frac{\lambda_{\gamma}}{2}\|\gamma - \tgamma\|_2^2}\\
	\text{s.t. } 	& \|{\color{red}\tbeta}\|_0 \leq k \\
					& \|{\color{red}\tgamma}\|_0 \leq j, \  \gamma \geq 0
	}
	\end{center}
	}

\end{frame}
\section{Proposed Algorithm}
\begin{frame}{Relax-and-Split from Optimization Perspective}
\eq{
	{\color<2, 3>{red}\min}_{{\color<2>{red}\beta, \gamma,} {\color<3>{red}\tbeta, \tgamma}} & {\color<2>{red}\LL(\beta, \gamma)} + {\color<2,3>{red}\frac{{\color<4>{red}\lambda_\beta}}{2}\|\beta - \tbeta\|^2_2} + {\color<2,3>{red}\frac{{\color<4>{red}\lambda_\gamma}}{2}\|\gamma - \tgamma\|_2^2}\\
	\text{s.t. } 	& {\color<3>{red}\|\tbeta\|_0 \leq k, \ \|\tgamma\|_0 \leq j} \\ 
		& {\color<2>{red}\gamma \geq 0}, \ {\color<3>{red}\tgamma \geq 0}
	}
We'll find $\beta^*$, $\gamma^*$, $\tbeta^*$, and $\tgamma^*$ via alternating partial minimization:

\begin{itemize}
	\item<2-> Minimization w.r.t. $\beta$ and $\gamma$ is done via \textbf{Interior Point Method}.
	\item<3-> Minimization w.r.t. $\tbeta$ and $\tgamma$ is \textbf{projection} of $\beta$ and $\gamma$ onto $\ell_0$-balls.
	\item<4-> $\lambda_\beta$ and $\lambda_\gamma$ are \textbf{iteratively increased} until $\tbeta \approx \beta$ and $\tgamma \approx \gamma$.
\end{itemize}
\only<-4>{
\visible<3>{
	\eq{
		\tbeta^{(k+1)} \leftarrow \proj_{\|\beta\|_0 \leq k}(\beta^{(k)}) \quad &\implies \quad \text{Take max k from } \beta \text{, rest to 0} \\
		\tgamma^{(k+1)} \leftarrow \proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)})\quad &\implies \quad \text{Take max j from }\gamma \text{, rest to 0}
		}
}
}
\only<5->{
\textbf{Advantages}:
\begin{enumerate}
	\item Interpretable hyperparameters.
	\item No model bias when the $k-$subspace is chosen.
\end{enumerate}
}
\end{frame}

%\begin{frame}{Interior Point Method}
%The Lagrangian of the barrier relaxation is:
%	\eq{
%	F_\mu(v, \beta, \gamma) & = \LL(\beta, \gamma) +  \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2 + \\ & + \mu\sum_{i=1}^{q}\log(\gamma_i) - v^T\gamma \\
%	}
%	The descent direction is chosen as:
%	\eq{
%	\nabla^2F_\mu\begin{bmatrix}
%		\Delta v \\
%		\Delta \beta \\
%		\Delta \gamma
%	\end{bmatrix} = -\nabla F_\mu\begin{bmatrix}
%		\Delta v \\
%		\Delta \beta \\
%		\Delta \gamma
%	\end{bmatrix}
%	\text{, step is: }
%	\begin{cases}
%	v^{(k+1)} = v^{(k)} + \alpha_k \Delta v \\
%	\gamma^{(k+1)} = \gamma^{(k)} + \alpha_k \Delta\gamma \\
%	\beta^{(k+1)} = \beta^{(k)} + \alpha_k \Delta \beta
%	\end{cases}
%	}
%	and $\alpha_k$ is chosen such that
%	\eq{
%	& \gamma^{(k+1)} \geq 0,\ v^{(k+1)} \geq 0 \\
%	& \|\nabla F_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|\nabla F_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| 
%	}
%\end{frame}

\begin{frame}{R\&S-Mixed}
The pseudocode of the proposed algorithm:
{\small
\begin{algorithm}[H]
	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
	\Repeat{converges}{
		$\tbeta^{(k+1)} \leftarrow \proj_{\|\beta\|_0 \leq k}(\beta^{(k)})$\\ 
		$\tgamma^{(k+1)} \leftarrow \proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)})$ \\
		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$
	}
	}
	\BlankLine
\end{algorithm}
	}
\end{frame}

\section{Experiments}
\subsection{Application to Synthetic Problems}

\begin{frame}{Performance in Comparison to Other Algorithms}
\begin{itemize}
	\item \textbf{Scenario 1}: $n=30$, $n_i=5$, $p = 9$, $q = 4$, with true parameters $\beta = (1, 1, 0, \dots, 0)$ and the covariance matrix $\Gamma$ being: 
		\eq{
			\Gamma = \begin{bmatrix}
				9 & 4.8 & 0.6 & 0 \\
				4.8 & 4 & 1 & 0 \\
				0.6 & 1 & 1 & 0\\
				0 & 0 & 0 & 0
			\end{bmatrix}
		}
	\item \textbf{Scenario 2}: everything as in Scenario 1, but $n=60$ and $n_i=10$.
\end{itemize}
\textbf{Competitors}:
\begin{itemize}
	\item \textbf{ALASSO:} 2 stage: A-LASSO+Newton and A-LASSO+PCO
	\item \textbf{M-ALASSO:} Adaptive LASSO + EM Algorithm
	\item \textbf{SCAD-P:} SCAD + Proxy Matrix for $\Gamma$
	\item \textbf{rPQL:} Quasi-Likelihood + Adaptive LASSO (for GLMMs)
\end{itemize}
\end{frame}

\begin{frame}{Performance in Comparison to Other Algorithms}
\begin{table}[H]
\begin{center}
\begin{tabular}{|l|l|c|c|c|c|c|}
\hline
Setup & Algoritm & \% C & \% CF & \% CR & MSE & TIME \\
\hline 
\hline
$n = 30$, $n_i = 5$ & \textbf{\ouralgo} & 58 & 72 & 78 & 0.66 & 0.015 \\
& rPQL & 88 & 98 & 88 & 0.88 & 26-59 \\ 
& M-ALASSO & 71 & 73 & 79 & - & -\\
& ALASSO & 79 & 81 & 96 & - & - \\
& SCAD-P & - & 90 & 86 & - & - \\
\hline 
$n = 60$, $n_i = 10$ & \textbf{\ouralgo} & 98 & 100  & 98 & 0.69& 0.018 \\
& rPQL & 98 & 99 & 98 & 0.97 &  26-59 \\ 
& M-ALASSO & 83 & 83 & 89 & - & - \\
& ALASSO & 95 & 96 & 99 & - & -\\
& SCAD-P & 100 & 100 & 100 & - & - \\
\hline

\end{tabular}
\end{center}
\caption{\label{table:krishna_setup_results} Comparison of feature selection algorithms. \% CF -- percent of models where true fixed effects were identified correctly, \% CR -- percent of models where true random effects were identified correctly, \% C -- both fixed and random effects were identified correctly.}	
\end{table}

\end{frame}

\begin{frame}{Scalability Experiment}
\begin{itemize}
	\item $n = 60$, $n_i = 10$
	\item $p = q \in [4, 7, 10, \dots, 90]$, 200 experiments for each.
	\item $X_i = Z_i$, columns are drawn form $\NN(0, \Psi)$ where
	\[
		\Psi = \begin{bmatrix}
			9 & 4.8 & 0.6 \\
			4.8 & 4 & 1 \\
			0.6 & 1 & 1 \\
		\end{bmatrix}
	\]
	\item 50\% random coordinate in $\beta$ are active
	\item 70\% of those are also active in $\gamma$
\end{itemize}
\end{frame}

\begin{frame}{Scalability Experiment}
		\begin{center}
			\includegraphics<+>[width=0.8\textwidth]{Figures/scalability_accuracy}
			%\includegraphics<+>[width=0.8\textwidth]{Figures/scalability_mse}
		\end{center}
\end{frame}

\subsection{Application to Real-World Problems}

\begin{frame}{Contact Rate Modeling for COVID-19 Forecasting}
	\begin{itemize}
		\item $n = 60$ groups (countries and US states), $n_i \approx 50$
		\item $Y_i$ -- contact rate for COVID SEIIR Model 
		\item $p=q=5$ covariates related to temperature, mobility, population, testing; plus intercept.
	\end{itemize}
\end{frame}

\begin{frame}{Contact Rate Modeling for COVID-19 Forecasting}
	\begin{center}
		\includegraphics<+>[width=\textwidth]{Figures/fit_Alaska}
		\includegraphics<+>[width=\textwidth]{Figures/fit_Slovenia}
		\includegraphics<+>[width=\textwidth]{Figures/fit_Switzerland}
		\includegraphics<+>[width=\textwidth]{Figures/fit_Turkey}
	\end{center}
\end{frame}

\begin{frame}{Burden of Anxiety and Depression as Result of Bullying}
	\begin{itemize}
		\item $m = 10$ cohort studies, $n = 77$, highly unbalanced 
		\item $p = 13$, $q = 12$ (\texttt{time} was preselected fixed-only)
		\item Covariates are related to studies' designs.
	\end{itemize}
\end{frame}

\begin{frame}{Burden of Anxiety and Depression as Result of Bullying}
	\begin{center}
		\includegraphics[width=\textwidth]{Figures/bullying_data.csv_inclusion}
		%\includegraphics<+>[width=\textwidth]{Figures/bullying_data.csv_fixed_feature_selection}
		%\includegraphics<+>[width=\textwidth]{Figures/bullying_data.csv_random_feature_selection}

	\end{center}
\end{frame}
\section{Future Work}
\begin{frame}{Future Work: Theory}
\begin{theorem}[Conditions for Convergence to True Estimator]
	Under certain conditions the method converges in a finite number of iterations to $(\hat\beta, \hat\gamma)$ which projections $(\tbeta, \tgamma)$ belong to a $k$- and $j$-subspaces respectively that contain the true minimum $(\beta^*,\gamma^*)$.
\end{theorem}
\begin{theorem}[Consistency of Estimator]
	There exists a local minimizer $(\hat{\beta}, \hat{\gamma})$ for the proposed loss function, such that it is asymptotically consistent with true minimum $(\beta^*, \gamma^*)$. 
\end{theorem}
\begin{theorem}[Consistency in Zeros]
	If some coordinates of the true minimizer ($\beta^*, \gamma^*$) are zero, then it is also zero in $(\hat{\beta}, \hat{\gamma})$, given that the later is sufficiently close to the former.
\end{theorem}
\begin{theorem}[Asymptotic Normality]
	The proposed estimator $(\hat{\beta}, \hat{\gamma})$ asymptotically normally distributed around true minimizer ($\beta^*, \gamma^*$) in its true non-zero $k+j$-subspace.
\end{theorem}	
\end{frame}

\begin{frame}{Future work: Algorithm}
	\textbf{Question:} Will exponential smoothing of projection improve the accuracy? 
	{\small
	\begin{algorithm}[H]
	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
	\Repeat{converges}{
		$\tbeta^{(k+1)} \leftarrow {\color{red}\delta}\proj_{\|\beta\|_0 \leq k}(\beta^{(k)}) + {\color{red}(1-\delta)\tbeta^{(k)}}$\\
		$\tgamma^{(k+1)} \leftarrow {\color{red}\delta}\proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)}) + {\color{red}(1-\delta)\tgamma^{(k)}}$\\
		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$
	}
	}
	\BlankLine
	\end{algorithm}
	}
	\end{frame}

\begin{frame}{Future Work: Implementation}
		Can we increase $\lambda_\beta$ and $\lambda_\gamma$ in a more careful way to avoid potential stacking? 
		The approach can be based on the theorem:
		\begin{theorem}[Distance Between Minima]
	For a fixed dataset $(X_i, Y_i)$ and relaxation parameters $\lambda_\beta$, $\lambda_\gamma$ the distance between $(\beta^*, \gamma^*)$, the unconstrained minimizers of relaxed problem, and their projections $(\tbeta^*, \tgamma^*)$ is bounded by a constant $M$ depending on $(X_i, Y_i)$ and the relaxation parameters.
		\end{theorem}
\end{frame}

\begin{frame}{The End}
	\begin{center}
		\large Thank you for your attention!
	\end{center}
\end{frame}

\begin{frame}[allowframebreaks]{References}

  \bibliography{bibliography}
  \bibliographystyle{alpha}

\end{frame}

\end{document}

### 2022-07-04 presentation/useful_packages.tex
Import thesis proposal from the general exam as a starting point for my PhD thesis

%%%%%%%%%%%%
% PACKAGES %
%%%%%%%%%%%%

% kek 
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[utf8]{inputenc} % For Spanish characters
\usepackage[english]{babel} % USEnglish localization
% =========== General Formatting =============
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{fancyhdr} % Allows for nice header and footer
\usepackage{appendix} % Enables appendices
\usepackage{enumerate} % Custom numerate, useful for i,ii,iii... I,II,III...
% \usepackage{hyperref} % For hyperlinks in the PDF
\usepackage{extramarks} % 
\usepackage{fontawesome} % For web icons

% =============== Math ===============
\usepackage{amsmath} % Standard math packages
\usepackage{amsthm} % Math Theorems
\usepackage{amssymb} % Math Symbols
\usepackage{amsfonts} % Math Fonts
\usepackage{mathtools} % Extra math tools such as PairedDelimiter
\usepackage{upgreek} % Nice sigma => \upsigma
\usepackage{array} % Enables array features
\usepackage{siunitx} % For SI Unit easy formatting
\usepackage{dsfont} % For math. indicators

% =============== Figures ===============
\usepackage{float} % Float features
\usepackage{wrapfig} % Al­lows fig­ures or ta­bles to have text wrapped around them
\usepackage{tikz} % Diagram and figure creation and rendering
% =============== Tables ===============
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi
\usepackage{multirow} % Combined rows in tables
\usepackage{multicol} % Combined columns in tables
\usepackage{colortbl} % Color cells
%\usepackage{longtable} % Tables than span multipages

% =============== Bibliography ==========
\usepackage{natbib} % Lots of useful versions of \cite

% % =============== Listings ============
\usepackage{listings} % Main package for inserting code
\usepackage[scaled]{beramono} % For using the beramono font
% % =============== Algorithms ===============
\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e} % Allows for algorithm description
\usepackage[noend]{algpseudocode}
% =============== Other ===============
\usepackage{datetime} % Date-Time formatting
\usepackage{ulem} % For strikethrough text \st{}
\usepackage[colorinlistoftodos]{todonotes} % useful for leaving todonotes
\usepackage{textcomp} %Text Com­pan­ion fonts

\usepackage{pdfpages} % Insert pdfs
\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
% \usepackage[space]{grffile} % insert files with spaces
\usepackage{pdflscape} % Individual horizontal pages
\usepackage[many]{tcolorbox} % Color boxes for comment
%\usepackage{xargs} % Expanded arguments features
%\usepackage{fix-cm} % Computer-Modern at arbitrarysizes
%\usepackage{eurosym} % Eurosymbol

%%%%%%%%%%%%%%%%%
% TODO COMMANDS %
%%%%%%%%%%%%%%%%%
\usepackage[colorinlistoftodos]{todonotes} % useful for leaving todonotes
\newcommand{\hightodo}[1]{\todo[inline,backgroundcolor=red!90!yellow!80!white]{\textbf{\userId}: #1}}
\newcommand{\lowtodo}[1]{\todo[inline,backgroundcolor=blue!30!white]{\textbf{\userId}: #1}}
\newcommand{\ask}[2]{\todo[inline,backgroundcolor=Orchid!40!white]{
	\textbf{Ask #1} #2}}
\newcommand{\critical}[1]{\todo[inline, backgroundcolor=BrickRed!50!white]{\textbf{Critical:} #1}}
\newcommand{\search}[1]{\todo[inline, backgroundcolor=blue!30!white]{\textbf{Search:} #1}}
\newcommand{\edit}[1]{\todo[inline, backgroundcolor=CornflowerBlue!30!white]{\textbf{Edit:} #1}}
\newcommand{\verify}[1]{\todo[inline, backgroundcolor=Apricot!50!white]{\textbf{Verify:} #1}}
\newcommand{\experiment}[1]{\todo[inline, backgroundcolor=Green!50!white]{\textbf{Verify:} #1}}
\newcommand{\delegate}[2]{\todo[inline, background=Lavender]{\textbf{For #1:} #2}}

% Tips
\newtcolorbox{tip}{colframe=white, colback=gray!10!white}

% Mettings
\newcounter{meetingscounter}
\setcounter{meetingscounter}{1}
\newtcolorbox{meeting}[4]{colframe=white, colback=white, colbacktitle=gray!10!white, 
	coltitle=black, fonttitle=\bfseries, breakable,
	subtitle style={boxrule=0.4pt, colback=white},
	title={
		\begin{tabular}{p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}l}
		\underline{Meeting \arabic{meetingscounter}}  &  \stepcounter{meetingscounter} \faCalendar\ #2 &  \faClockO\ #3 & \faMapMarker\ #4 & \faUsers\ #1
		\end{tabular}
		}
	}

% Thick strike out
\newcommand{\cancelled}[1]{%
    \renewcommand{\ULthickness}{3pt}%
       \sout{#1}%
    \renewcommand{\ULthickness}{.4pt}% Resetting to ulem default
}

\newcommand{\done}[1]{\sout{#1}}

### 2022-07-04 tables/bullying_features_description.tex
Import thesis proposal from the general exam as a starting point for my PhD thesis

\label{appendix:bullying_covariates}
The author acknowledges his colleague and collaborator Damian Santomauro\footnote{\href{mailto:d.santomauro@uq.edu.au}{d.santomauro@uq.edu.au}, Affiliate Assistant Professor of Health Metrics Sciences, Institute for Health Metrics and Evaluation, University of Washington} for providing the dataset, the description of its covariates, and the expert assessment of their historical importance in different rounds of GBD study below.
\begin{enumerate}
	\item \texttt{cv\_symptoms}
	\begin{itemize}
		\item 0 = study assesses participants for MDD or anxiety disorders via a diagnostic interview to determine whether they have a diagnosis. 
		\item 1 = study uses a symptom scale (e.g., Beck Depression Inventory) and uses an established cut-off on that scale to determine caseness. 
		\item Has not historically been significant. 
	\end{itemize}
	\item \texttt{cv\_unadjusted}
	\begin{itemize}
		\item 0 = RR is adjusted for potential confounders (e.g., SES, etc.)	
		\item 1 = RR is not adjusted for potential confounders
		\item Has been significant in the past.
	\end{itemize}
	\item \texttt{cv\_b\_parent\_only} 
	\begin{itemize}
		\item 0 = Child is involved in reporting their own exposure to bullying.
		\item 1 = Only parent is involved in reporting the child’s exposure to bullying
		\item This covariate has recently started becoming significant (but not consistently). 
	\end{itemize}
	\item \texttt{cv\_or}
	\begin{itemize}
		\item 0 = estimate is a RR
		\item 1 = estimate is an odds ratio (OR)
		\item ORs are always larger than RRs. However the magnitude may be very small / insignificant.
	\end{itemize}
	\item \texttt{cv\_multi\_reg}
	\begin{itemize}
		\item 0 = RR is the ratio of the rate of the outcome in persons exposed vs all persons unexposed (including persons exposed to low-threshold bullying victimization)
		\item 1 =  RRs are estimated via a logistic regression where exposure represented by 3 categories: 1) No exposure, 2) Occasional exposure, 3) Frequent exposure. The RR for occasional exposure will exclude participants with frequent exposure, and the RR for frequent exposure will exclude participants with occasional exposure. 
		\item Is expected to be significant.
	\end{itemize}
	\item \texttt{cv\_low\_threshold\_bullying}
	\begin{itemize}
		\item 0 = uses a ‘frequent’ exposure frequency threshold for classing someone as exposed to bullying.
		\item 1 = uses an ‘occasional’ exposure frequency threshold for classing someone as exposed to bullying.
		\item Has been consistently significant with a strong magnitude.
	\end{itemize}
	\item \texttt{cv\_anx}
	\begin{itemize}
		\item 0 = estimate represents risk for MDD
		\item 1 = estimate represents risk for anxiety disorders
	\end{itemize}
	\item \texttt{cv\_selection\_bias}
	\begin{itemize}
		\item 0 = < 15\% attrition at followup
		\item 1 = $\geqslant$ 15\% attrition at followup
		\item Has been significant in the past
	\end{itemize}
	\item \texttt{Percent\_female}
	\begin{itemize}
		\item Indicates \% of sample in estimate that are female. 
	\end{itemize}
	\item \texttt{cv\_child\_baseline}
	\begin{itemize}
		\item Has not been significant in the past.
	\end{itemize}
\end{enumerate}

### 2022-07-04 presentation/general_presentation.tex
Fill presentation up to algorithm's pseudocode

@@
 \newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
-\newcommand{\ouralgo}{R\&S-Mixed }
+\newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
 \title{Feature Selection for Mixed-Effects Models}
-\date{Thursday, 3\textsuperscript{rd} of December, 2020}
+\date{\today}
 \author{Aleksei Sholokhov}
@@
 
-\section{Introduction}
+\section{Feature Selection for Linear Mixed-Effect Models}
 \subsection{Linear Mixed-Effects Models}
-\begin{frame}{Linear Mixed-Effect Models are a type of regression models for grouped data}
+\begin{frame}{Linear Mixed-Effect Models}
 
-	\[
-	\text{Dataset: m groups } (X_i, y_i),\quad i = 1, \dots m
-    \]
+Linear Mixed-Effect (LME) models are often used for analyzing combined data across a range of groups. \newline
+
+They use use covariates to separate the population variability (fixed-effects) from the group variability (random effects). \newline
+
+LMEs borrow strength across groups to estimate key statistics in cases when data within units are sparse or highly variable. \newline
+
+\end{frame}
+
+\begin{frame}{Linear Mixed-Effect Models}
+
+	Dataset: $m$ groups $(X_i, Z_i, y_i),\quad i = 1, \dots m$, each has $n_i$ observations
+	\begin{itemize}
+		\item 	$X_i \in \R^{n_i \times p}$ -- group $i$ design matrix for fixed features
+		\item 	$Z_i \in \R^{n_i \times q}$ -- group $i$ design matrix for random effects
+		\item 	$y_i \in \R^{n_i}$ -- group $i$ observations  
+	\end{itemize}
+
 	\begin{columns}[T,onlytextwidth]
@@
    		\begin{split}
-   			y_i & = X_i(\beta + {\color{red} u_i}) + \varepsilon_i, \quad \varepsilon_i \sim \NN(0, \Lambda_i) \\
-   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma)}
+   			y_i & = X_i\beta + {\color{red}Z_i u_i} + \varepsilon_i, \quad \varepsilon_i \sim \NN(0, \Lambda_i) \\
+   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma4)}
    		\end{split}
@@
 	\eq{
-   		y_i & = \alt<3->{X_i{\color<4,9>{red}\beta} + {\color<3>{red}Z_i}{\color<5,9>{red}u_i}}{{\color<2>{red}X_i}(\beta + u_i)} + \varepsilon_i \quad i = 1\dots m \\
-   		{\color<7>{red}\varepsilon_i} & \sim \NN(0, {\color<8,9>{red}\Lambda_i}) \\
-   		u_i & \sim \NN(0, {\color<6,9>{red}\Gamma})
+   		y_i & = X_i\beta + Z_iu_i + \varepsilon_i \quad i = 1\dots m \\
+   		\varepsilon_i & \sim \NN(0, \Lambda_i) \\
+   		u_i & \sim \NN(0, \Gamma)
    	}
@@
    	\begin{itemize}
-   		\item<2-> $X_i \in \R^{n_i \times p}$ -- \only<3>{{\color{red} fixed}}\only<4->{fixed} features, or covariates, \only<3>{{\color{red} $Z_i \in \R^{n_i \times q}$ -- random features }}\only<4->{ $Z_i \in R^{n_i \times q}$ -- random features}
-   		\item<3-> $p$ -- number of fixed features, $q$ -- number of random effects.
-   		\item<4-> $\beta \in \R^p$ -- fixed effects, or mean effects
-   		\item<5-> $u_i \in \R^q$ -- random effects
-   		\item<6-> $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects
-   		\item<7-> $\varepsilon_i \in \R^{n_i}$ -- observation noise
-   		\item<8-> $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for noise
-   		\item<9-> Unknowns: $\beta$, $u_i$, $\Gamma$, sometimes $\Lambda_i$.
+   		\item $p$ -- number of fixed features, $q$ -- number of random effects.
+   		\item $\beta \in \R^p$ -- fixed effects, or mean effects
+   		\item $u_i \in \R^q$ -- random effects
+   		\item $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects, often $\Gamma = \diag{(\gamma)}$
+   		\item $\varepsilon_i \in \R^{n_i}$ -- observation noise
+   		\item $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for noise
+   	\end{itemize}
+   	Unknowns: $\beta$, $u_i$, $\gamma$, sometimes $\Lambda_i$.
+\end{frame}
 
+\begin{frame}{Likelihood for Mixed Models}
+Negative log-likelihood:
+\eq{
+	\label{eq:lmm_objective}
+	\mathcal{L}(\beta, \gamma) & = \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \\ & + \half\log{\det{\pa{Z_i \Gamma Z_i^T + \Lambda_i}}}, \quad \Gamma = \diag{(\gamma)}
+	}
+Maximum likelihood estimates for $\beta$ and $\gamma$ solve the problem:
+\eq{
+	\mathcal{LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma)
+}
+
+To select covariates we add a sparsity-promoting regularizer $R(\beta, \gamma)$
+\eq{
+	\mathcal{FS-LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
+}
+
+\begin{itemize}
+	\item $\LL(\beta, \gamma)$ is smooth on its domain, quadratic w.r.t. $\beta$ and $\bar\eta$-weakly-convex w.r.t. $\gamma$.
+	\item $R(\beta, \gamma)$ is closed, proper, convex, with easily computed \textit{prox operator}
 \end{itemize}
-   	\uncover<10->{
-   	\[
-   		\Gamma = \diag{({\color{red}\gamma})},\ {\color{red}\gamma \in \R^{q}} \quad \text{ and } \Lambda_i \text{ are known.}
-   	\]
+
+\end{frame}
+
+\begin{frame}{Regularization}
+\begin{itemize}
+	\item $R(\beta, \gamma)$ is closed, proper, convex, with easily computed \textit{prox operator}
+\end{itemize}
+
+\eq{
+	\prox_{\alpha R + \delta_{\CC}}(\tbeta, \tgamma) & := \argmin_{(\beta, \gamma) \in \CC} R(\beta, \gamma) + \frac{1}{2\alpha}\|(\beta, \gamma) - (\tbeta, \tgamma)\|_2^2, \\ & \text{ where } \CC := \R^p \times R^q_+ \\
+}
+
+Examples: 
+\begin{itemize}
+	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \cite{Bondell2010,Lin2013}
+	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \cite{Vaida2005,Jones2011}
+	\item $R(x)$ -- SCAD penalty (\cite{Fan2001,Fan2012})
+\end{itemize}
+
+TODO: add picture
+
+\end{frame}
+
+\begin{frame}{SR3-Relaxation for Mixed-Effect Models}
+Original problem $\mathcal{FS-LME}$:
+\eq{
+	\min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
+}
+Relaxed problem $\ouralgo$:
+\eq{
+	\label{eq:msr3_formulation_explicit}
+	\min_{\beta, \tbeta \in \R^p,\, \gamma, \tgamma \in \R^{q}_+} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma) + R(\beta, \gamma)
+}
+where the \textit{relaxation} $\kappa_\eta$ decouples the likelihood and the regularizer 
+\eq{
+	\kappa_\eta(\beta - \tbeta, \gamma - \tgamma) := \frac{\eta}{2}\|\beta - \tbeta\|^2_2 + \frac{\eta + \bar\eta}{2}\|\gamma - \tgamma\|_2^2
+}
+and the \textit{projection function} $\phi_\mu$ replaces $\gamma \geq 0$ with a log-barrier 
+\eq{
+	\phi_\mu(\gamma) := \begin{cases}
+		-\mu\sum_{i=1}^q \ln(\gamma_i/\mu), & \mu > 0 \\
+		\delta_{\R_+^q}(\gamma), & \mu = 0 \\
+		+\infty, & \mu < 0
+	\end{cases}
 } 
@@
 
-\begin{frame}[fragile]{Likelihood for Mixed Models}
-Suppose we have a linear mixed effects model
+\begin{frame}{Value Function Reformulation}
+$\ouralgo$-relaxation replaces the original likelihood $\LL$ with a \textit{value function} $u_{\eta,\mu}$:
+\eq{
+	\label{eq:value_function_definition}
+	u_{\eta,\mu}(\tbeta,\tgamma) & := \min_{(\beta, \gamma)} \LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\  & := \min_{(\beta, \gamma)} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma)
+}
+so $\ouralgo$-formulation~(\ref{eq:msr3_formulation_explicit}) becomes
+\eq{
+	\min_{(\tbeta, \tgamma) \in \CC} u_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
+}
+When $\bar\eta$ is larger than the weak-convexity constant
+\begin{itemize}
+	\item $u_{\eta,\mu}$ is well-defined and continuously differentiable.
+	\item Solutions $(\tbeta^*, \tgamma^*)$ for $\ouralgo$ converge to solutions $(\beta^*, \gamma^*)$ of $\mathcal{FS-LME}$ \\ when $\mu \rightarrow 0$ and $\eta \rightarrow \infty$.
+\end{itemize}
 
+\textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations is enough.
+\end{frame}
+
+\begin{frame}{Value Function Reformulation}
+TODO: insert image here
+\end{frame}
+
+\begin{frame}{Designing an Algorithm}
+	Gradient of a Lagrangian:
 	\eq{
-	\label{eq:lmm_setup}
-	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i \\
-	u_i & \sim \NN(0, \Gamma), \ \Gamma = \diag{\gamma} \\
-	\varepsilon_i & \sim \NN(0, \Lambda_i)\\
-}
-where $\beta$, $\gamma$, and $u_i$ are unknown model parameters, and $\lambda_i$ are known.
-
-	Noticing that $Z_iu_i + \varepsilon_i \sim \NN(0, Z_i\Gamma Z_i^T + \Lambda_i)$ we get the negative log-likelihood:
-	 \eq{
-	\label{eq:lmm_objective}
-	\mathcal{L}(\beta, \gamma) & = \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \\ & + \half\log{\det{\pa{Z_i \Gamma Z_i^T + \Lambda_i}}}
+		G_{\nu,\eta}((\beta,\gamma,v),(\tbeta,\tgamma)) := \begin{bmatrix}
+			\nabla_\beta \LL(\beta, \gamma) + \eta(\beta-\tbeta) \\
+			\nabla_\gamma \LL(\beta, \gamma) + (\bar\eta + \eta)(\gamma-\tgamma) - v\\
+			v \bigodot \gamma - \mu\textbf{1}
+		\end{bmatrix}
 	}
 
+\textbf{Lemma:} For every $(\mu,\eta) \in \R_+\times\R_{++}$,
+\eq{
+	(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
+	\text{(equivalent to)} \\
+	\exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
+}
+If $\mu > 0$, then $\hat{v} = -\nabla\phi_\mu(\hat{\gamma})$, and if $\mu = 0$, then $\hat{v}$ is the unique KKT multiplier associated with the constraint $0 \leq \gamma$.
 \end{frame}
 
-\subsection{Feature Selection for Mixed Models}
-\begin{frame}{Feature Selection for Mixed Models}
-
-Consider the following feature selection setup:
-
-\eq{
-	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
-	\text{s.t. } & \|\beta\|_0 \leq k \\
-	& \|\gamma\|_0 \leq j \\
-	& \gamma \geq 0
-}
-
-where $k$ and $j$ is the size of the sparse support for the fixed and random effects respectively.
-
-This is a very hard problem to solve directly.
-
-\end{frame}
-
-\begin{frame}{Methods for Feature Selection}
-There are several ways to address this setup:
-  \begin{columns}[T,onlytextwidth]
-    \column{0.33\textwidth}
-    (0) Exhaustive Search:
-\[
-	\begin{split}
-	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
-	\text{s.t. } & \|\beta\|_0 \leq k \\
-	& \|\gamma\|_0 \leq j \\
-	& \gamma \geq 0
-	\end{split}
-\]
-
-    \column{0.33\textwidth}
-    \uncover<2->{
-    \centering
-    (1) $\ell_1$-relaxation:
-\[
-	\begin{split}
-	\min_{\beta, \gamma} & \LL(\beta, \gamma) \\
-	\text{s.t. } & \|\beta\|_{\color{red}1} \leq {\color{red} \lambda_\beta} \\
-	& \|\gamma\|_{\color{red}1} \leq {\color{red} \lambda_\gamma} \\
-	& \gamma \geq 0
-	\end{split}
-\]
-	}
-	\column{0.33\textwidth}
-	\uncover<3->{
-    \centering
-    (3) Other types of sparsity-inducing penalties\\ ($\ell_p$, $SCAD$).
-    }
-  \end{columns}
-
- \onslide<1-3>{
-	\begin{columns}[T, onlytextwidth]
-	\column{0.33\textwidth}
-	\onslide<1-3>{
-	\centering
-	AIC \cite{Vaida2005}\\
-	BIC \cite{Jones2011}\\
-	}
-	\column{0.33\textwidth}
-	\onslide<2-3>{
-	\centering
-	M-ALASSO \cite{Bondell2010} \\
-	ALASSO \cite{Lin2013} \\
-	}
-	\column{0.33\textwidth}
-	\onslide<3>{
-	\centering
-	SCAD \cite{Fan2001}\\
-	SCAD-P \cite{Fan2012}
-	}
-	\end{columns}
-}
-
-\onslide<4->{
-	\begin{center}
-	(2) SR3-type Relaxation \cite{Zheng2019}
-	\eq{
-	\min_{\beta, \gamma, {\color{red}\hbeta}, {\color{red}\tgamma}} & \LL(\beta, \gamma) + {\color{red}\frac{\lambda_\beta}{2}\|\beta - \tbeta\|^2_2 + \frac{\lambda_{\gamma}}{2}\|\gamma - \tgamma\|_2^2}\\
-	\text{s.t. } 	& \|{\color{red}\tbeta}\|_0 \leq k \\
-					& \|{\color{red}\tgamma}\|_0 \leq j, \  \gamma \geq 0
-	}
-	\end{center}
-	}
-
-\end{frame}
-\section{Proposed Algorithm}
-\begin{frame}{Relax-and-Split from Optimization Perspective}
-\eq{
-	{\color<2, 3>{red}\min}_{{\color<2>{red}\beta, \gamma,} {\color<3>{red}\tbeta, \tgamma}} & {\color<2>{red}\LL(\beta, \gamma)} + {\color<2,3>{red}\frac{{\color<4>{red}\lambda_\beta}}{2}\|\beta - \tbeta\|^2_2} + {\color<2,3>{red}\frac{{\color<4>{red}\lambda_\gamma}}{2}\|\gamma - \tgamma\|_2^2}\\
-	\text{s.t. } 	& {\color<3>{red}\|\tbeta\|_0 \leq k, \ \|\tgamma\|_0 \leq j} \\ 
-		& {\color<2>{red}\gamma \geq 0}, \ {\color<3>{red}\tgamma \geq 0}
-	}
-We'll find $\beta^*$, $\gamma^*$, $\tbeta^*$, and $\tgamma^*$ via alternating partial minimization:
-
-\begin{itemize}
-	\item<2-> Minimization w.r.t. $\beta$ and $\gamma$ is done via \textbf{Interior Point Method}.
-	\item<3-> Minimization w.r.t. $\tbeta$ and $\tgamma$ is \textbf{projection} of $\beta$ and $\gamma$ onto $\ell_0$-balls.
-	\item<4-> $\lambda_\beta$ and $\lambda_\gamma$ are \textbf{iteratively increased} until $\tbeta \approx \beta$ and $\tgamma \approx \gamma$.
-\end{itemize}
-\only<-4>{
-\visible<3>{
-	\eq{
-		\tbeta^{(k+1)} \leftarrow \proj_{\|\beta\|_0 \leq k}(\beta^{(k)}) \quad &\implies \quad \text{Take max k from } \beta \text{, rest to 0} \\
-		\tgamma^{(k+1)} \leftarrow \proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)})\quad &\implies \quad \text{Take max j from }\gamma \text{, rest to 0}
-		}
-}
-}
-\only<5->{
-\textbf{Advantages}:
-\begin{enumerate}
-	\item Interpretable hyperparameters.
-	\item No model bias when the $k-$subspace is chosen.
-\end{enumerate}
-}
-\end{frame}
-
-%\begin{frame}{Interior Point Method}
-%The Lagrangian of the barrier relaxation is:
-%	\eq{
-%	F_\mu(v, \beta, \gamma) & = \LL(\beta, \gamma) +  \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2 + \\ & + \mu\sum_{i=1}^{q}\log(\gamma_i) - v^T\gamma \\
-%	}
-%	The descent direction is chosen as:
-%	\eq{
-%	\nabla^2F_\mu\begin{bmatrix}
-%		\Delta v \\
-%		\Delta \beta \\
-%		\Delta \gamma
-%	\end{bmatrix} = -\nabla F_\mu\begin{bmatrix}
-%		\Delta v \\
-%		\Delta \beta \\
-%		\Delta \gamma
-%	\end{bmatrix}
-%	\text{, step is: }
-%	\begin{cases}
-%	v^{(k+1)} = v^{(k)} + \alpha_k \Delta v \\
-%	\gamma^{(k+1)} = \gamma^{(k)} + \alpha_k \Delta\gamma \\
-%	\beta^{(k+1)} = \beta^{(k)} + \alpha_k \Delta \beta
-%	\end{cases}
-%	}
-%	and $\alpha_k$ is chosen such that
-%	\eq{
-%	& \gamma^{(k+1)} \geq 0,\ v^{(k+1)} \geq 0 \\
-%	& \|\nabla F_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|\nabla F_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| 
-%	}
-%\end{frame}
-
-\begin{frame}{R\&S-Mixed}
-The pseudocode of the proposed algorithm:
-{\small
-\begin{algorithm}[H]
-	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
-	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
-	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
-	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
-	\Repeat{converges}{
-		$\tbeta^{(k+1)} \leftarrow \proj_{\|\beta\|_0 \leq k}(\beta^{(k)})$\\ 
-		$\tgamma^{(k+1)} \leftarrow \proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)})$ \\
-		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$
-	}
-	}
-	\BlankLine
-\end{algorithm}
-	}
+\begin{frame}{$\ouralgo$ Algorithm}
+\input{../extras/08Pseudocode}
 \end{frame}

### 2022-07-04 presentation/general_presentation.tex
Finish first pass for presentation

@@
 \usepackage{xspace}
+\usepackage{caption}
+\usepackage{subcaption}
 \newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
 \newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
-\title{Feature Selection for Mixed-Effects Models}
+\title{A Relaxation Approach to Feature Selection for Mixed-Effects Models}
 \date{\today}
-\author{Aleksei Sholokhov}
+\author{Aleksei Sholokhov, James V. Burke, Peng Zheng, and Aleksandr Aravkin}
 % \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}
@@
 
-\begin{frame}{Plan}
-	\tableofcontents
-\end{frame}
-
-\section{Feature Selection for Linear Mixed-Effect Models}
+\section{Relaxed for Linear Mixed-Effect Models}
 \subsection{Linear Mixed-Effects Models}
@@
 
-TODO: add picture
+\begin{figure}
+     \centering
+     \begin{subfigure}[b]{0.25\textwidth}
+         \centering
+         \includegraphics[width=\textwidth]{Figures/l0_regularizer.pdf}
+         \caption{$\ell_0$}
+     \end{subfigure}%
+     \begin{subfigure}[b]{0.25\textwidth}
+         \centering
+         \includegraphics[width=\textwidth]{Figures/l1_regularizer.pdf}
+         \caption{$\ell_1$}
+     \end{subfigure}%
+     \begin{subfigure}[b]{0.25\textwidth}
+         \centering
+         \includegraphics[width=\textwidth]{Figures/lh_regularizer.pdf}
+         \caption{$\ell_p,\, p=1/2$}
+     \end{subfigure}%
+     \begin{subfigure}[b]{0.25\textwidth}
+         \centering
+         \includegraphics[width=\textwidth]{Figures/scad_regularizer.pdf}
+         \caption{SCAD}
+     \end{subfigure}
+     \caption{Three simple graphs}
+     \label{fig:three graphs}
+\end{figure}
 
@@
 	\label{eq:msr3_formulation_explicit}
-	\min_{\beta, \tbeta \in \R^p,\, \gamma, \tgamma \in \R^{q}_+} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma) + R(\beta, \gamma)
+	\min_{\beta, \tbeta \in \R^p,\, \gamma, \tgamma \in \R^{q}_+} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma) + R(\tbeta, \tgamma)
 }
@@
 
-\textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations is enough.
+\textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations keep the solution close to the central path.
 \end{frame}
@@
 \begin{frame}{Value Function Reformulation}
-TODO: insert image here
+\begin{figure}
+	\includegraphics[width=0.7\textwidth]{Figures/intuition_current.pdf}
+	\caption{\label{fig:intuition_sr3} Comparison of the level-sets for the original likelihood (left) and $\ouralgo$-likelihood (right), for fixed (top) and random (bottom) effects.}
+\end{figure}
 \end{frame}
@@
 \eq{
-	(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
-	\text{(equivalent to)} \\
-	\exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
+	&(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
+	& \iff \\
+	& \exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
 }
@@
 
-\begin{frame}{$\ouralgo$ Algorithm}
+\begin{frame}{$\ouralgo$-fast Algorithm}
 \input{../extras/08Pseudocode}
@@
 
-\begin{frame}{Performance in Comparison to Other Algorithms}
+\begin{frame}{Application to Synthetic Problems}
+\textbf{The Experiment} 
 \begin{itemize}
-	\item \textbf{Scenario 1}: $n=30$, $n_i=5$, $p = 9$, $q = 4$, with true parameters $\beta = (1, 1, 0, \dots, 0)$ and the covariance matrix $\Gamma$ being: 
-		\eq{
-			\Gamma = \begin{bmatrix}
-				9 & 4.8 & 0.6 & 0 \\
-				4.8 & 4 & 1 & 0 \\
-				0.6 & 1 & 1 & 0\\
-				0 & 0 & 0 & 0
-			\end{bmatrix}
-		}
-	\item \textbf{Scenario 2}: everything as in Scenario 1, but $n=60$ and $n_i=10$.
+	\item The number of fixed effects $p$ and random effects $q$ is 20.
+	\item $\beta = \gamma = \frac{1}{2}[1,2,3,\dots,10, 0\dots,0]$
+	\item 9 groups with sizes [10, 15, 4, 8, 3, 5, 18, 9, 6]
+	\item $X_i \sim \NN(0, I)^p$, $Z_i = X_i$, $\varepsilon_i \sim \NN(0, 0.3^2I)$
+	\item Each experiment is repeated 100 times.
+	\item Grid-search for $\eta \in [10^{-4}, 10^{2}]$, golden search for $\lambda \in [0, 10^5]$
+	\item Final model is chosen to maximize BIC
 \end{itemize}
-\textbf{Competitors}:
+\begin{table}
+	\input{../extras/performance_table_short_current}
+\end{table}
+\end{frame}
+
+\begin{frame}{Application to Synthetic Problems}
+\begin{figure}
+	\includegraphics[width=\textwidth]{Figures/performance_picture_current.pdf}
+\end{figure}
+Benefits:
 \begin{itemize}
-	\item \textbf{ALASSO:} 2 stage: A-LASSO+Newton and A-LASSO+PCO
-	\item \textbf{M-ALASSO:} Adaptive LASSO + EM Algorithm
-	\item \textbf{SCAD-P:} SCAD + Proxy Matrix for $\Gamma$
-	\item \textbf{rPQL:} Quasi-Likelihood + Adaptive LASSO (for GLMMs)
+	\item $\ouralgo$-relaxation has similar (and sometimes better!) feature selection performance than the original likelihood.
+	\item $\ouralgo$-fast optimization accelerates the compute time by $\sim 10^2$.
+\end{itemize}
+Setbacks:
+   \begin{itemize}
+   	\item $\eta$ is a new hyperparameter to tune.
    \end{itemize}
@@
 
-\begin{frame}{Performance in Comparison to Other Algorithms}
-\begin{table}[H]
-\begin{center}
-\begin{tabular}{|l|l|c|c|c|c|c|}
-\hline
-Setup & Algoritm & \% C & \% CF & \% CR & MSE & TIME \\
-\hline 
-\hline
-$n = 30$, $n_i = 5$ & \textbf{\ouralgo} & 58 & 72 & 78 & 0.66 & 0.015 \\
-& rPQL & 88 & 98 & 88 & 0.88 & 26-59 \\ 
-& M-ALASSO & 71 & 73 & 79 & - & -\\
-& ALASSO & 79 & 81 & 96 & - & - \\
-& SCAD-P & - & 90 & 86 & - & - \\
-\hline 
-$n = 60$, $n_i = 10$ & \textbf{\ouralgo} & 98 & 100  & 98 & 0.69& 0.018 \\
-& rPQL & 98 & 99 & 98 & 0.97 &  26-59 \\ 
-& M-ALASSO & 83 & 83 & 89 & - & - \\
-& ALASSO & 95 & 96 & 99 & - & -\\
-& SCAD-P & 100 & 100 & 100 & - & - \\
-\hline
-
-\end{tabular}
-\end{center}
-\caption{\label{table:krishna_setup_results} Comparison of feature selection algorithms.
-\% CF -- percent of models where true fixed effects were identified correctly, \% CR -- percent of models where true random effects were identified correctly, \% C -- both fixed and random effects were identified correctly.}
-\end{table}
-
+\begin{frame}{$\ell_0$-based Covariate Selection for Bullying Study from GBD}
+	\begin{figure}
+		\includegraphics[width=\textwidth]{Figures/bullying_data_assessment_selection.pdf}
+		\caption{Fixed and random covariate selection for Bullying dataset from~\cite{GBD}.
+The model selected 9 covariates, 7 of which were historically significant, and did not select 4 covariates, 1 of which was historically significant.}
+	\end{figure}
 \end{frame}
 
-\begin{frame}{Scalability Experiment}
+\begin{frame}{Thank You!}
+	The code is available on GitHub: \\
+	\href{github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}
 	\begin{itemize}
-	\item $n = 60$, $n_i = 10$
-	\item $p = q \in [4, 7, 10, \dots, 90]$, 200 experiments for each.
-	\item $X_i = Z_i$, columns are drawn form $\NN(0, \Psi)$ where
-	\[
-		\Psi = \begin{bmatrix}
-			9 & 4.8 & 0.6 \\
-			4.8 & 4 & 1 \\
-			0.6 & 1 & 1 \\
-		\end{bmatrix}
-	\]
-	\item 50\% random coordinate in $\beta$ are active
-	\item 70\% of those are also active in $\gamma$
+		\item All estimators are fully compatible to \texttt{sklearn} library.
+		\item Implements SR3 for linear, generalized-linear, and linear mixed-effect models.
+		\item Has tutorials, tests, and documentation.
 	\end{itemize}
-\end{frame}
 
-\begin{frame}{Scalability Experiment}
-		\begin{center}
-			\includegraphics<+>[width=0.8\textwidth]{Figures/scalability_accuracy}
-			%\includegraphics<+>[width=0.8\textwidth]{Figures/scalability_mse}
-		\end{center}
-\end{frame}
-
-\subsection{Application to Real-World Problems}
-
-\begin{frame}{Contact Rate Modeling for COVID-19 Forecasting}
-	\begin{itemize}
-		\item $n = 60$ groups (countries and US states), $n_i \approx 50$
-		\item $Y_i$ -- contact rate for COVID SEIIR Model 
-		\item $p=q=5$ covariates related to temperature, mobility, population, testing; plus intercept.
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Contact Rate Modeling for COVID-19 Forecasting}
-	\begin{center}
-		\includegraphics<+>[width=\textwidth]{Figures/fit_Alaska}
-		\includegraphics<+>[width=\textwidth]{Figures/fit_Slovenia}
-		\includegraphics<+>[width=\textwidth]{Figures/fit_Switzerland}
-		\includegraphics<+>[width=\textwidth]{Figures/fit_Turkey}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Burden of Anxiety and Depression as Result of Bullying}
-	\begin{itemize}
-		\item $m = 10$ cohort studies, $n = 77$, highly unbalanced 
-		\item $p = 13$, $q = 12$ (\texttt{time} was preselected fixed-only)
-		\item Covariates are related to studies' designs.
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Burden of Anxiety and Depression as Result of Bullying}
-	\begin{center}
-		\includegraphics[width=\textwidth]{Figures/bullying_data.csv_inclusion}
-		%\includegraphics<+>[width=\textwidth]{Figures/bullying_data.csv_fixed_feature_selection}
-		%\includegraphics<+>[width=\textwidth]{Figures/bullying_data.csv_random_feature_selection}
-
-	\end{center}
-\end{frame}
-\section{Future Work}
-\begin{frame}{Future Work: Theory}
-\begin{theorem}[Conditions for Convergence to True Estimator]
-	Under certain conditions the method converges in a finite number of iterations to $(\hat\beta, \hat\gamma)$ which projections $(\tbeta, \tgamma)$ belong to a $k$- and $j$-subspaces respectively that contain the true minimum $(\beta^*,\gamma^*)$.
-\end{theorem}
-\begin{theorem}[Consistency of Estimator]
-	There exists a local minimizer $(\hat{\beta}, \hat{\gamma})$ for the proposed loss function, such that it is asymptotically consistent with true minimum $(\beta^*, \gamma^*)$. 
-\end{theorem}
-\begin{theorem}[Consistency in Zeros]
-	If some coordinates of the true minimizer ($\beta^*, \gamma^*$) are zero, then it is also zero in $(\hat{\beta}, \hat{\gamma})$, given that the later is sufficiently close to the former.
-\end{theorem}
-\begin{theorem}[Asymptotic Normality]
-	The proposed estimator $(\hat{\beta}, \hat{\gamma})$ asymptotically normally distributed around true minimizer ($\beta^*, \gamma^*$) in its true non-zero $k+j$-subspace.
-\end{theorem}	
-\end{frame}
-
-\begin{frame}{Future work: Algorithm}
-	\textbf{Question:} Will exponential smoothing of projection improve the accuracy? 
-	{\small
-	\begin{algorithm}[H]
-	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
-	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
-	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
-	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
-	\Repeat{converges}{
-		$\tbeta^{(k+1)} \leftarrow {\color{red}\delta}\proj_{\|\beta\|_0 \leq k}(\beta^{(k)}) + {\color{red}(1-\delta)\tbeta^{(k)}}$\\
-		$\tgamma^{(k+1)} \leftarrow {\color{red}\delta}\proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)}) + {\color{red}(1-\delta)\tgamma^{(k)}}$\\
-		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$
-	}
-	}
-	\BlankLine
-	\end{algorithm}
-	}
-	\end{frame}
-
-\begin{frame}{Future Work: Implementation}
-		Can we increase $\lambda_\beta$ and $\lambda_\gamma$ in a more careful way to avoid potential stacking? 
-		The approach can be based on the theorem:
-		\begin{theorem}[Distance Between Minima]
-	For a fixed dataset $(X_i, Y_i)$ and relaxation parameters $\lambda_\beta$, $\lambda_\gamma$ the distance between $(\beta^*, \gamma^*)$, the unconstrained minimizers of relaxed problem, and their projections $(\tbeta^*, \tgamma^*)$ is bounded by a constant $M$ depending on $(X_i, Y_i)$ and the relaxation parameters.
-		\end{theorem}
-\end{frame}
-
-\begin{frame}{The End}
-	\begin{center}
-		\large Thank you for your attention!
-	\end{center}
-\end{frame}
-
-\begin{frame}[allowframebreaks]{References}
-
-  \bibliography{bibliography}
-  \bibliographystyle{alpha}
-
+	\textbf{References}: 
+	\bibliographystyle{plain}
+	\bibliography{bibliography.bib}
 \end{frame}

### 2022-07-23 presentation/general_presentation.tex
ICCOPT corrections from Jim and Sashsa

@@
 \newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
-\title{A Relaxation Approach to Feature Selection for Mixed-Effects Models}
+\title{A Relaxation Approach to Feature Selection \newline for Linear Mixed-Effects Models}
 \date{\today}
-\author{Aleksei Sholokhov, James V. Burke, Peng Zheng, and Aleksandr Aravkin}
-% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}
+\author{Aleksei Sholokhov, James V. Burke, Peng Zheng, Damian Santomauro, and Aleksandr Aravkin}
+\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
 
@@
 \section{Relaxed for Linear Mixed-Effect Models}
-\subsection{Linear Mixed-Effects Models}
-\begin{frame}{Linear Mixed-Effect Models}
+\subsection{Mixed-Effects Models}
+\begin{frame}{Mixed-Effect Models}
 
-Linear Mixed-Effect (LME) models are often used for analyzing combined data across a range of groups. \newline
+\begin{itemize}
+	\item Used for analyzing \textbf{combined data} across a range of \textbf{groups}.
+	\item Use covariates to separate the \textbf{population variability} from the \textbf{group variability}.
+	\item \textbf{Borrow strength} across groups to estimate key statistics. % when data within units are sparse or highly variable.
+	\item Often seek \textbf{sparse models} that only use \textbf{most informative} covariates. 
+\end{itemize}
 
-They use use covariates to separate the population variability (fixed-effects) from the group variability (random effects). \newline
-
-LMEs borrow strength across groups to estimate key statistics in cases when data within units are sparse or highly variable. \newline
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{Figures/ihme_predictions.png}\footnote{Picture is taken from \href{covid19.healthdata.org}{covid19.healthdata.org} }
+\end{figure}
 
@@
 
-\begin{frame}{Linear Mixed-Effect Models}
+\begin{frame}{Linear Mixed-Effect (LME) Models}
 
@@
    			y_i & = X_i\beta + {\color{red}Z_i u_i} + \varepsilon_i, \quad \varepsilon_i \sim \NN(0, \Lambda_i) \\
-   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma4)}
+   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma)}
    		\end{split}
@@
 \begin{itemize}
-	\item $R(\beta, \gamma)$ is closed, proper, convex, with easily computed \textit{prox operator}
+	\item $R(\beta, \gamma)$ is closed, proper, with easily computed \textit{prox operator}
 \end{itemize}
@@
      \end{subfigure}
-     \caption{Three simple graphs}
+     \caption{Four commonly-used regularizers which promote sparsity}
      \label{fig:three graphs}
@@
 
-\begin{frame}{SR3-Relaxation for Mixed-Effect Models}
+\begin{frame}{SR3-Relaxation for Mixed-Effect Models ($\ouralgo$)}
 Original problem $\mathcal{FS-LME}$:
@@
 \eq{
-	\kappa_\eta(\beta - \tbeta, \gamma - \tgamma) := \frac{\eta}{2}\|\beta - \tbeta\|^2_2 + \frac{\eta + \bar\eta}{2}\|\gamma - \tgamma\|_2^2
+	\kappa_\eta(\beta - \tbeta, \gamma - \tgamma) := \frac{\eta}{2}\|\beta - \tbeta\|^2_2 + \frac{\eta}{2}\|\gamma - \tgamma\|_2^2, \quad \eta > \bar\eta
 }
-and the \textit{projection function} $\phi_\mu$ replaces $\gamma \geq 0$ with a log-barrier 
+and the \textit{perspective mapping} $\phi_\mu$ replaces $\gamma \geq 0$ with a log-barrier 
 \eq{
@@
 }
-When $\bar\eta$ is larger than the weak-convexity constant
+When $\eta$ is larger than the weak-convexity constant
 \begin{itemize}
 	\item $u_{\eta,\mu}$ is well-defined and continuously differentiable.
-	\item Solutions $(\tbeta^*, \tgamma^*)$ for $\ouralgo$ converge to solutions $(\beta^*, \gamma^*)$ of $\mathcal{FS-LME}$ \\ when $\mu \rightarrow 0$ and $\eta \rightarrow \infty$.
+	\item As $\mu \rightarrow 0$ and $\eta \rightarrow \infty$, cluster points of solutions to $\ouralgo$ are first-order stationary points for $\mathcal{FS-LME}$ \\ 
 \end{itemize}
@@
 \textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations keep the solution close to the central path.
+\end{frame}
+
+\begin{frame}{Value Function Reformulation}
+\begin{figure}
+	\includegraphics[width=\textwidth]{Figures/intuition_prev_paper}
+	\caption{\label{fig:intuition_prev} Picture from \cite{Zheng2018RelaxAndSplit}: for a linear problem, value function relaxation ``squashes'' level-sets simplifying the optimization landscape.}
+\end{figure}
 \end{frame}
@@
 \begin{frame}{Designing an Algorithm}
-	Gradient of a Lagrangian:
+	$G_{\nu,\eta}$ encodes both gradient of a Lagrangian (lines 1-2) and the complementarity condition (line 3):
 	\eq{
@@
 			\nabla_\beta \LL(\beta, \gamma) + \eta(\beta-\tbeta) \\
-			\nabla_\gamma \LL(\beta, \gamma) + (\bar\eta + \eta)(\gamma-\tgamma) - v\\
+			\nabla_\gamma \LL(\beta, \gamma) + \eta(\gamma-\tgamma) - v\\
 			v \bigodot \gamma - \mu\textbf{1}
@@
 	}
-
+	We apply Newton method to $G$ while geometrically decreasing $\mu$. \\
 \textbf{Lemma:} For every $(\mu,\eta) \in \R_+\times\R_{++}$,
@@
 \begin{itemize}
-	\item $\ouralgo$-relaxation has similar (and sometimes better!) feature selection performance than the original likelihood.
+	\item $\ouralgo$-relaxation improves feature selection performance than the original likelihood.
 	\item $\ouralgo$-fast optimization accelerates the compute time by $\sim 10^2$.
 \end{itemize}
-Setbacks:
+Challenge:
    \begin{itemize}
-   	\item $\eta$ is a new hyperparameter to tune.
+   	\item Initialization of $\eta$ is problem-specific
    \end{itemize}
+\end{frame}
+
+\begin{frame}{Choice of $\eta$}
+	\begin{figure}
+		\includegraphics[width=\textwidth]{Figures/eta_L1.pdf}
+	\end{figure}
 \end{frame}
@@
 \begin{frame}{Thank You!}
-	The code is available on GitHub: \\
-	\href{github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}
+	The code is available on GitHub: \href{github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}
 	\begin{itemize}
@@
 	\end{itemize}
-
+\end{frame}
+\begin{frame}{References}
 	\textbf{References}: 

### 2023-04-06 appendix/acknowledgements.tex
Initiate PhD thesis structure

I am extremely grateful to my advisor, Sasha Aravkin, for providing support, guidance, and enthusiasm during my work on this project in the Department of Applied Mathematics and in Institute for Health Metrics and Evaluation.\\
I would also like to thank Jim Burke for his detailed review of my work which would hopefully lead to fruitful collaboration. \\ 
Last, but not least: I am grateful to Damian Santomauro, who introduced me to his work on the consequences of bullying, provided with a dataset, and with a valuable feedback on the performance of the proposed method. \\

### 2023-04-06 cs.tex
Initiate PhD thesis structure

\section{Application to Compressive Sensing}
\subsection{Introduction}
\paragraph{Prior Work}
\subsection{Method}
\paragraph{Compressive Sensing} In many applications we can not observe $x(t)$ in real time. Instead, we observe $p$ detectors, and each detector provides a linear combination of a small number of coordinates of $x(t)$ at a time:
\begin{equation}
    \label{eq:cs_definition}
    y(t) = A_t x(t), \quad A_t \in \mathbb{R}^{p\times n}
\end{equation}

In particular, we consider a single pixel camera setup where for every time instance $t$, $p$ acquisitions $y(t)$ are obtained by a high sampling rate photo-detector using the projection matrix $A_t$. The rows of the matrix $A_t$ correspond to a binary mask pattern that can be encoded using a digital micro-mirror device (DMD) where the incoming light from $x(t)$ is reflected from the DMD array and focused onto the photo-detector. Figure~\ref{fig:SPI} illustrates an example of the single pixel imaging setup where a gas plume is imaged using a DMD array and a medium infra-red (MIR) photo-detector. 
\begin{figure}
\centering
\includegraphics[width = 0.5\textwidth]{figures/SPI_setup.pdf}\label{fig:SPI}
\end{figure}

However, it is often possible to have access to a complete state $x(t)$ at the moment of model training. Thus, one can develop a ROM $(\psi_\theta, \phi_\theta, h_\theta)$ using the full state $x(t)$, and then utilize this ROM for real-time compressive sensing applications.

\paragraph{Training Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses. The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data. Formally, for a given set of trajectories $\bd{x}_i$, $i \in [1 \dots k]$, where each trajectory $\bd{x}_i \in \mathbb{R}^{n \times p}$ is a set of $p$ snapshots that correspond to the recorded states of the system for $p$ time-steps, $t_j$, $j \in [1, \dots, p]$, the loss function $\LL^{data}_\theta$ is defined as:

\begin{align}
    \label{eq:loss_training}
    \LL^{training}(\theta) & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
\end{align}
where $\sigma$ is the standard deviation of the observation noise.

To obtain a ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$, we minimize the loss above:

\begin{equation}
    \label{eq:training_minimization}
    \theta^* = \arg\min_{\theta} \LL^{training}(\theta)
\end{equation}

We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}). This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}. To simplify the notation without loss of generality, in the rest of the paper we assume that all trajectories were recorded over the same time-frame with an equal and uniform step-size.

\paragraph{Reconstruction Loss} 
We use the ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ above to forecast the dynamics based on partial observations in real time. Namely, instead of reconstructing $x(t)$ based on compressive-sensing observations $y(t)$ directly, we first reconstruct the latent dynamics $z(t)$ and then project it to the observable space using the decoder $\psi_{\theta^*}(z)$.

\begin{align}
    \label{eq:reconstruction_problem_differential}
    \min_{\{z_t\}_{t=1, \dots, T}} & \frac{1}{2}\sum_{t=1}^T \left\|y_t - A\psi_{\theta^*}(z_t)\right\|_2^2 \\
    \text{s.t. } & \dot{z} = h_{\theta^*}(z)
\end{align}

We integrate the constraint and write it in its Lagrangian form:
\begin{align}
\label{eq:reconstruction_problem_integral}
    \min_{\{z_t\}_{t=1, \dots, T}} \LL_{\theta^*}^{recon}(z)
\end{align}

where

\begin{equation}
    \label{eq:reconstruction_loss}
     \LL_{\theta^*}^{recon}(z) = \frac{1}{2}\sum_{t=1}^T \left\|y_t - A\psi_{\theta^*}(z_t)\right\|_2^2 + \frac{\lambda}{2} \sum_{t=1}^T \left\|z_{t-1} + \int_{t-1}^{t}h_{\theta^*}(z)dz - z_t\right\|_2^2
\end{equation}

where the parameter $\lambda$ controls the degree on which the compressing sensing algorithm relies on the latent dynamics $h_{\theta^*}$ during the signal reconstruction phase. We minimize the loss~\ref{eq:reconstruction_loss} using a gradient-based technique, with the gradients obtained using automatic differentiation frameworks.

\subsection{Experiments}
\subsection{Discussion and Conclusion}

### 2023-04-06 extras/table_glmm_current.tex
Initiate PhD thesis structure

\begin{tabular}{lrrrr}
\toprule
Algorithm &        MSR3-Fast ($\ell_1$)&         \texttt{glmmLasso}\footnote{\href{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}} \cite{groll2014variable}  &            \texttt{lmmLasso}\footnote{\href{https://rdrr.io/cran/lmmlasso/}{https://rdrr.io/cran/lmmlasso/}}\cite{schelldorfer2011estimation} & PGD ($\ell_1$) \\
\midrule
Accuracy, \% &     {\bf 88} &        48 &          66 & 73  \\
FE Accuracy, \% &      {\bf 86} &        52  &          47 & 56\\
RE Accuracy, \% &     {\bf 91} &        45  &         84 & {\bf 91} \\
Time, sec        & {\bf 0.19} &  1.37  &  11.51 & 38.39 \\
Iterations, num  &      34 &        50 &             - & 7693 \\
\bottomrule
\end{tabular}

### 2023-04-06 msr3.tex
Initiate PhD thesis structure

Linear Mixed-Effects (LME) models are a fundamental tool for modeling correlated data, 
including cohort studies, longitudinal data analysis, and meta-analysis.  Design and analysis of variable selection methods for LMEs is more difficult than for linear regression because LME models are nonlinear. In this work we 
propose a relaxation strategy and optimization methods that enable a wide range of variable selection methods for LMEs using both convex and nonconvex regularizers, including  $\ell_1$, Adaptive-$\ell_1$, SCAD, and $\ell_0$. 
The computational framework only requires the proximal operator for each regularizer to be available, and the implementation is available in an open source \texttt{python} package \texttt{pysr3}, consistent 
with the \texttt{sklearn} standard. 
The numerical results on simulated data sets indicate that the proposed strategy improves on the state of the art for both accuracy and compute time. 
The variable selection techniques are also validated on a real example using a data set on bullying victimization.

{\it Keywords:}  Mixed effects models, feature selection, nonconvex optimization 
%\vfill
%
%\newpage
%\spacingset{1.5} % DON'T change the spacing!

\section{Introduction}
Linear mixed-effects (LME) models use covariates 
%(also known as predictors or features) 
to explain the variability of target variables in a grouped data setting. For each group, the relationship between covariates and observations is modeled 
using group-specific coefficients that are linked by a common prior distribution
across all groups, allowing LMEs to borrow strength across groups
in order to estimate statistics for the common prior. 
LMEs are used in settings with insufficient data to resolve each group independently, making them 
fundamental tools for regression analysis in 
population health sciences (\cite{covid2020modeling,murray2020global}), meta-analysis (\cite{dersimonian1986meta, zheng2021trimmed}), life sciences, and 
as well as in many others domains (\cite{zuur2009mixed}).

Variable selection is a fundamental problem in all regression settings. In linear regression, the LASSO method~\citep{tibshirani1996regression} and related extensions have been widely used. % for this purpose. 
However, variable selection for LMEs is complicated by the nonlinear structure and relative sparsity of the within-group data. 
While standard methods and software are available for linear regression (see e.g. \texttt{glmnet}~\cite{glmnet}), there are few open source libraries for variable selection 
for LMEs. 
Many covariates selection algorithms for LMEs have been proposed over the last 20 years (see the survey \cite{Buscemi2019Survey}), but comparison of 
these strategies and practical application remains difficult. %practical application of these strategies is difficult. 
%Most solutions share the goal of improving selection quality but vary significantly in implementation details. 
Approaches vary by choice of likelihood (e.g. marginal, restricted, or h- likelihood),  
regularizer~(e.g. $\ell_1$~\citep{Krishna2008} or SCAD~\cite{ibrahim2011fixed}), and information criteria \citep{Vaida2005,Ibrahim2011}. 
%and 
%sometimes  estimate effects in stages \citep{Krishna2008}. 
Implementations vary as well, typically using regularizer-specific local quadratic approximations to apply
solution methods for smooth problems (Newton-Raphson, EM, sequential least squares) to fit the original nonsmooth model. 
All of these decisions make it  difficult to compare and evaluate performance of available 
variable selection strategies and to determine which method is best suited for a given task. 
This challenge  is exacerbated by the absence of standardized datasets %for different experimental designs
and open source libraries for each method. 
Our main practical goal to fill this gap by developing a unified methodological framework that
accommodates a wide variety of variable selection strategies based on a set
of easily implementable regularizers, and implemented in an open source library that makes 
it easy to use and to compare different methods.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{figures/summary_picture.pdf}
    \caption{Selection of fixed and random effects for LME likelihoods $\mathcal{L}$ using 
    `regularization-agnostic' framework and its SR3 extension using four regularizers. 
    %standard and SR3-relaxed linear mixed-effect likelihoods with popular sparsity-promoting regularizers for the task of simultaneous 
    %The original likelihood $\LL(\beta, \gamma)$ (Eq. \ref{eq:lmm_objective}) is relaxed with SR3 approach to obtain $u(\beta, \gamma)$ (Eq. \ref{eq:value_function_definition}). Both setups are then solved using the same Proximal Gradient Descent (PGD) algorithm. 
    SR3 relaxation accelerates algorithmic converge  (middle panel), and gives better robustness and improved performance on synthetic problems across regularizers (right panel)}.
    \label{fig:summary}
\end{figure}

In this work we develop a regularization-agnostic covariate selection strategy that (1) is fast and simple to implement, (2) provides robust models, and (3) is flexible enough to support most regularizers currently used in variable selection across different domains.  The baseline approach uses the proximal gradient descent (PGD) method, which has been studied by the optimization community for over 40 years, but has not been widely used in LME covariate selection. We provide proximal operators for commonly used regularizers and show how to apply the PGD method to the nonconvex LME setting. In particular we apply the PGD method to four regularizers, 
including the $\ell_0$ regularizer, which does not admit local quadratic approximations and has not been used before for LME variable selection strategies.

We also develop a new meta-approach that can improve the performance of LME selection methods for any regularizer. Specifically, we extend sparse relaxed regularized regression (SR3) framework~{(\cite{Zheng2019SR3})} to the LME setting. In linear regression, SR3 accelerates and improves the performance of  regularization strategies by introducing auxiliary variables that decouple the accuracy and sparsity requirements for the model coefficients.
We develop a conceptual and algorithmic approach necessary to extend the SR3 concept to LME.  This development is necessary because the LME problem is nonlinear, nonconvex, and includes constraints on variance parameters.  We show that the new approach yields superior results in terms of specificity and sensitivity of feature selection, and is also computationally efficient.

\edit{Add a paragraph summarizing results from the theory paper}

All new methods are implemented in an open-source library called \texttt{pysr3}, which fills a gap for python mixed-models selection tools in \texttt{Python} (\cite{Buscemi2019Survey}, Table 3). Our algorithms are  1-2 orders of magnitude faster than available LASSO-based libraries for  mixed effects selection in \texttt{R},  see Table \ref{table:glmmlasso}. \texttt{pysr3} enables a standardized comparison of different methods in the LME setting, and makes both the PGD framework and its SR3 extension available to practitioners working with LME models.

\edit{Extend on the package summary using JOSS paper}

%\subsection{Notation}
%\begin{enumerate}
%\item sets like $\N$, $\B$, $\bS$, $\SS$.
%\item nonsmooth stuff, subdifferential and normal cones
%\end{enumerate}

\section{Linear Mixed-Effects Models: Notation and Fundamentals}
Mixed-effect models describe the relationship between an outcome variable and its predictors when the observations are grouped, for example in studies or clusters.  To set the notation, consider $m$ groups of observations indexed by $i$, with sizes $n_i$, and the total number of observations equal to $n = n_1 + n_2 + \dots + n_m$. For each group, we have design matrices for fixed features $X_i \in \R^{n_i \times p}$,  and matrices of random features $Z_i \in \R^{n_i \times q}$, along with vectors of outcomes $Y_i \in \R^{n_i}$. 
%Typically, columns of $Z_i$ are a subset of columns $X_i$ but it does not have to be the case in general. 
Let 
{$X = [X_1^T, X_2^T, \dots, X_m^T]^T$ and $Z = [Z_1^T, Z_2^T, \dots, Z_m^T]^T$.} 
Following~\cite{Patterson1971, Pinheiro2000}, we define a Linear Mixed-Effects (LME) model as
\eq{
	\label{eq:lme_setup}
	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i, \quad i = 1 \dots m \\
	u_i & \sim \NN(0, \Gamma),\quad \Gamma \in \bS_{+}^{q} \\
	\varepsilon_i & \sim \NN(0, \Lambda_i), \quad \Lambda_i \in \bS_{++}^{n_i}
}
 where $\beta \in \R^p$ is a vector of fixed (mean) covariates, 
 $u_i \in \R^{q}$ are unobservable random effects assumed to be distributed normally with zero mean and the unknown covariance matrix $\Gamma$, and $\bS_{+}^{\nu}$ 
 and $\bS_{++}^{\nu}$ are the sets of
 real symmetric $\nu\times \nu$ positive semi-definite
 and positive definite matrices, respectively. 
Matrices $Z_i$ encode a wide variety of models, including
random intercepts ($Z_i$ are columns of 1's that add $u_i$ to all datapoints from the $i$th study)
and random slopes ($Z_i$ also scale $u_i$ according to the magnitude of a covarite), see e.g.~\cite{pinheiro2006mixed}. 
%In these typical examples, the columns of $Z_i$ are subsets of the columns
%of $X_i$.  
In our study, we assume that the observation error covariance matrices 
$\Lambda_i$ are given and that the random effects covariance matrix 
is an unknown diagonal matrix, i.e., $\Gamma = \Diag{\gamma}, \ \gamma \in \R^s_+$.
% either
% assumed to have a simple parametric form, such as $\Lambda_i=\sig_i I$
% with $\sig_i$ to be determiined, or given given  
 %; we focus on the latter case below.

Defining group-specific error terms $\omega_i = Z_i u_i + \varepsilon_i$, we get a compact formulation that 
recasts~\eqref{eq:lme_setup} as a correlated noise model:
 \eq{
 \label{eq:lmm_correlated_noise_setup}
	Y_i = X_i\beta + \omega_i,\quad \omega_i \sim \NN(0, \Omega_i(\Gamma)), \quad \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i.
}
For brevity, we refer to $\Omega_i(\Gamma)$ as just $\Omega_i$.
%without the parentheses. In fact, $\Omega_i$ will be the only terms depending on $\Gamma$ in the majority of expressions including the mixed model's likelihood and its derivatives described below.
The reformulation \eqref{eq:lmm_correlated_noise_setup}
yields the following marginalized  negative log-likelihood function of a linear mixed-effects model~\citep{Patterson1971}:
\eq{
	\label{eq:lmm_objective}
	\LL_{ML}(\beta, \Gamma)  :=
%	 \textabove{$\Delta$}{=}
	 \sum_{i = 1}^m \half(y_i - X_i\beta)^T\Omega_i^{-1}(y_i - X_i\beta) + \half\ln{\det{\Omega_i}}.
}
Maximum likelihood estimates for $\beta$ and $\Gam$ are obtained by
solving the optimization problem
\eq{
	\label{eq:ml_lme_optimization_setup}
	\min_{\beta, \Gamma} & \ \LL_{ML}(\beta, \Gamma)  \quad \mbox{s.t.} \quad \ \Gamma \in \bS_{+}^{q}.
}
At this point, we bring in three basic definitions from variational analysis~\cite{rockafellar2009variational}.

\begin{definition}[Epigraph and level sets]
The epigraph of a function $f:\mathbb{R}^n\rightarrow \mathbb{R} \cup \{\infty\}$ is defined as
\[
\epi f = \{(x,\alpha) : f(x) \leq \alpha\}. 
\]
For a given $\alpha$, the $\alpha$-level set of $f$ is defined as
\[
\text{lev}_\alpha f = \{x: f(x) \leq \alpha\}.
\] 
\end{definition}

\begin{definition}[Lower semicontinuity and level-boundedness]
A function $f:\mathbb{R}^n\rightarrow \mathbb{R} \cup \{\infty\}$ is lower semicontinous (lsc) when $\epi f$ is closed, 
and level-bounded when all level sets $\text{lev}_\alpha f$ are bounded.  
\end{definition}

\begin{definition}[Convexity]
A function $f:\mathbb{R}^n\rightarrow \mathbb{R} \cup \{\infty\}$ is convex when $\epi f$ is a convex set. Equivalently, 
\[
f(\lambda x + (1-\lambda)y) \leq \lambda f(x) + (1-\lambda) f(y) \quad \forall x,y\in\dom{f},\ \lambda \in (0,1),
\]
where $\dom{f}:=\lset{x\in\Rn}{f(x)<+\infty}$.
\end{definition}

\begin{definition}[Weak convexity]
A function $f:\mathbb{R}^n\rightarrow \mathbb{R} \cup \{\infty\}$ is $\eta$-weakly convex 
$f(\cdot)+\frac{\eta}{2}\|\cdot\|^2$ is convex. 
\end{definition}

The negative log likelihood~\eqref{eq:ml_lme_optimization_setup}  is nonlinear and nonconvex, and requires an iterative numerical solver.
However, it is convex with respect to $\beta$, and weakly convex with respect to $\gamma$, with a weak convexity constant $\overline \eta$
computed in~\cite[Section 5.1]{Theory1} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation 
%\eq{
%	\label{eq:beta_formula}
\[
	\beta(\Gamma) = \argmin_{\beta}\LL(\beta, \Gamma) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i.
\] %}
By using the simplification $\Gam=\Diag{\gam}$, we obtain the problem
%{In the meta-analysis setting under study, 
%In our study, we assume that $\Gam$ is a diagonal matrix of the form
%$\Gamma = \Diag{\gamma}, \ \gamma \in \R^s$ with the 
%$\Lambda_i \in \R_{++}^{n_i \times n_i}$ fixed and given 
%%the positive definite constraint from (\ref{eq:ml_lme_optimization_setup}) transforms into a box constraint:
%so that our problem takes the form}
\eq{
	\label{eq:lme_diagonal_setup}
	\min_{\beta\in\R^p, \gam\in\R^q_+} & \ \LL(\beta,\gam)
	:=\LL_{ML}(\beta, \Diag{\gamma}) 
%	\\
%	s.t. & \ \gamma \geq 0\ .
	}
%Under the assumption $\Lambda_i \in \R_{++}^{n_i \times n_i}$ in~\eqref{eq:lme_setup}, 
In this setting, when an
entry $\gamma_j$ takes the  value $0$ the corresponding coordinates of all random effects $u_{ij}$ are identically $0$ for all $i$.

Verification of the existence to solutions to \eqref{eq:lme_diagonal_setup}
and, more generally, \eqref{eq:ml_lme_optimization_setup} follows from the work of \cite{zheng2021trimmed}.
Standalone proofs for the existence of minimizers are developed in~\cite[Theorem 1]{Theory1},
and extended to the presence of regularizers in~\cite[Theorem 2]{Theory1}. %, restated below for convenience.

%%Here we streamline these techniques and extend to the regularized case, 
%%with basic results given below, with proofs in the appendix.  
%
%\begin{theorem}[Existence of a Minimizer]\label{thm:basic existence}
%Let the assumptions in the statement of problem \eqref{eq:ml_lme_optimization_setup} hold. Then optimal solutions to
%\eqref{eq:ml_lme_optimization_setup} exist.
%\end{theorem}
%
%Feature selection methods  include sparsity-inducing penalties or constraints. See~\cite[Theorem 1]{Theory1}
%for an existence result under weak assumptions on the regularizer. 
%
%
% 
%
%\begin{theorem}\label{thm:basic existence2}
%Let the assumptions in the statement of problem \eqref{eq:lme_diagonal_setup} hold,
%define $\LL(\beta,\gam):=\LL_{ML}(\beta, \Diag{\gamma})$
%and suppose $\map{\hR}{\R^p\times\R^q_+}{\R\cup\{+\infty\}}$ 
%is lsc and level bounded.
%Then $\LL+\hR$ is level bounded and solutions to
%the following optimization problem exist:
%\eq{
%\label{eq:extended loss}
%\min_{\beta\in\R^p, \gamma\in\R^q_+}  \LL(\beta, \gamma) + 
%\hR(\beta, \gamma) .
%}
%\end{theorem}

%\red{I don't get the point of the following paragraph. Indeed, it seems wrong.}

This paper focuses the case where $\Gamma$ is diagonal, 
  (often referred to as \textit{the diagonal setup}) 
 and all $\Lambda_i$ are known (see \eqref{eq:lme_diagonal_setup}), 
following the meta-analysis use-case~\citep{zheng2021trimmed} that is widely used in epidemiological studies~\cite{murray2020global}. 
While the proposed approach can be extended to the non-diagonal case, we leave it for future work, save for a brief discussion in Section~\ref{sec:applications}.

\subsection{Prior Work on Feature Selection for Mixed-Effects Models}
\label{sec:prior_work}

Variable (feature) selection models seeks to select or rank the most important predictors in a dataset in order to get a parsimonious model at a minimal cost to prediction quality. 
%The selection process often occurs simultaneously with fitting the model, and a predictor is `selected' if its respective coefficient in the model is non-zero. 
If the desired number of coefficients $k$ is given, then the feature selection problem can be formulated as the minimization of a loss function $f(\theta)$ (e.g. the negative log-likelihood) subject to a zero-norm constraint:
\eq{
	\label{eq:general_feature_selection_zero_norm}
	&\min_{\theta}  f(\theta) \quad \mbox{ s.t. }  \quad \|\theta\|_0 \leq k 
	}
where $\|\theta\|_0$ denotes the number of nonzero entries in $\theta$, see panel (c) of Figure~\ref{fig:regularizers}.

%\begin{figure}
%    \centering
%    \begin{tabular}{cc}
%    \includegraphics[width=0.3\textwidth]{figs/l0_regularizer.pdf}
%    &
%        \includegraphics[width=0.3\textwidth]{figs/l1_regularizer.pdf}\\
%        (a) $\ell_0$ & (b) $\ell_1$ \\
%           \includegraphics[width=0.3\textwidth]{figs/lh_regularizer.pdf}
%    &
%        \includegraphics[width=0.3\textwidth]{figs/scad_regularizer.pdf}\\
%              (a) $\ell_{1/2}$ & (b) Clipped Absolute Deviation (CAD) 
%    \end{tabular}
%    \caption{Common convex and non-convex regularizers used for feature selection.}
%    \label{fig:regularizers}
%\end{figure}

% In this form the problem is equivalent to the Knapsack problem, which is NP-complete. Many techniques based on exhaustive search, also known as subset selection process, were developed, and showed to be effective when the total number of predictors is small (see \cite{Muller2013}). The same setup appears to be intractable in a high-dimension setting due to the exponential growth of the number of subsets to check when $n \to \infty$, however, it's possible to get an approximate solution via relaxation techniques.

 \begin{figure}[H]
    \centering
    \begin{tabular}{ccc}
        \includegraphics[width=0.3\textwidth]{figures/l1_regularizer.pdf}
&
           \includegraphics[width=0.3\textwidth]{figures/lh_regularizer.pdf}
    &
    \includegraphics[width=0.3\textwidth]{figures/l0_regularizer.pdf}\\
  %      \includegraphics[width=0.3\textwidth]{figs/scad_regularizer.pdf}\\
              (a) $\ell_1$  & (b) $\ell_{1/2}$ & (c) $\ell_0$
    \end{tabular}
    \caption{Common convex and non-convex regularizers used for feature selection.}
    \label{fig:regularizers}
\end{figure}

The constraint in \eqref{eq:general_feature_selection_zero_norm} is 
combinatorial, and a common workaround is to relax it to a one-norm constraint, 
with $\|\theta\|_1$ equal to the sum of absolute values of the entries of $\theta$. 
The best-known example of this approach is the least absolute square shrinkage operator (LASSO) studied by \cite{Tibshirani1996} for linear regression, 
see panel (a) of Figure~\ref{fig:regularizers}.
% for the least squares model:
% \eq{
% 	\label{eq:lasso_regression}
% 	\min_\theta f(\theta) + \lambda||\theta||_1
 %

% LASSO combines the small bias of the least-square estimator and the interpretability of the subset selection method, producing
% %, however, 
% biased estimates for the large coefficients~\citep{Zou2006}.  This approach has been extended to leverage many other regularizers that exhibit useful properties including bias 
%reduction for larger coefficients (SCAD from \cite{Fan2001}), simultaneous selection for highly correlated predictors (Elastic Net from \cite{Zou2005}), or better selection accuracy using non-penalized solution's coordinates as weights (A-LASSO from \cite{Zou2006}).

 Feature selection for LMEs is more difficult than for linear regression models. 
 In linear regression the observations are independent, whereas in mixed-effects setup they are generally correlated. 
 In addition, LMEs have both mean effect variables  $\beta$ as well as random variance variables 
 $\Gamma$. % where the groups can have different relative importance. 
% Finally, selecting random features implies that $\Gamma$ will have zero columns and rows thereby forcing the optimization methods to the boundary of the feasible parameter set and making both the theoretical analysis and numerical computations more challenging.  
The shrinkage operator approach for linear regression 
\citep{Tibshirani1996} was first adapted 
to the problem of feature selection for the fixed effects in mixed-effect models by \cite{Lan2006}. 
The removal of a random effect from the model requires the elimination of an entire row and column from $\Gamma$. 
To make the problem more tractable, \cite{Chen2003} reparamtrized $\Gamma$ through a modified Cholesky decomposition $\Gamma(D,L): = DLL^TD$, where $D$ is a diagonal matrix and $L$ is a lower-triangular matrix with ones on the main diagonal, and focused on selecting elements of $D$. 
% In this case, one can select variance parameters by selecting the elements of the diagonal matrix $D$ while treating the non-diagonal entries of $L$ as free parameters. 
Based on this idea, \cite{Krishna2008} extended the Adaptive LASSO regularizer (\cite{Lan2006, Xu2015}) to mixed-effects setting using the objective
   \[
        \LL(\beta, \Gamma(D,L)) + \lambda\left(\sum_{i = 1}^p 
        \left|\frac{\beta_i}{\hat\beta_i}\right| + \sum_{j = 1}^q \frac{D_{ii}}{\hat{D_{ii}}} \right),
   \]
 where $\hat\beta$ and $\hat{D}$ are the solution of a non-penalized maximum likelihood problem and $\lambda$ is a tuning parameter for the weighted regularizer
 and is called the regularization parameter. 
 %This strategy has a well known issue: non-penalized estimates $\hat\beta$ and $\hat{D}$ may be inaccessible because the numerical algorithm may fail to converge, especially when the underlying solution is sparse. 
 \cite{Ibrahim2011} use a similar approach, penalizing  non-zero elements $\Gamma_{ij}$ directly. 
 Other methods that use Adaptive LASSO for simultaneous selection 
 of fixed and random effects are \cite{Lin2013, fan2014robust, pan2018simultaneous}. Adaptive LASSO is available to practitioners via \texttt{R} packages \texttt{glmmLasso}\footnote{\href{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}} (\cite{groll2014variable}) and \texttt{lmmLasso}\footnote{\href{https://rdrr.io/cran/lmmlasso/}{https://rdrr.io/cran/lmmlasso/}}(\cite{schelldorfer2011estimation}).

 A popular nonconvex regularizer used for feature selection is smoothed clipped absolute deviation (SCAD)~\cite{Fan2001}. 
 %The (non-smoothed) clipped absolute deviation penalty is depicted in panel (d) of Figure~\ref{fig:regularizers}.
% , and the algebraic forms for both CAD and SCAD are given in the appendix. 
% which acts on the first derivative of the penalty function $p_\lambda(\theta)$:
 %  \eq{
  % 		p'_{\lambda}(|\theta|) = \lambda\left\{\ind{\theta \leq \lambda} +  \frac{(a\lambda - \theta)_+}{(a-1)\lambda}\ind{\theta - \lambda} \right\}
  % }
The adaptation of the SCAD penalty to select both fixed and random features in 
linear mixed models was developed by \cite{Fan2012}. SCAD was also used by \cite{chen2015inference} for selecting fixed effects and establishing the existence of random effects in ANOVA-type models. Finally, \cite{Ghosh2018} studied SCAD regularization for selecting mean effects in high-dimensional genomics problems.

\edit{Add more methods for feature selection \cite{Buscemi2019Survey, Muller2013}}

To better compare methods, we need to consider the regularization parameter 
\plum{$\lambda$} and how it is tuned. 
The output of a shrinkage model critically depends on the tuning parameter 
$\lambda$. The entire range of possible $\lambda$ values is captured 
by the notion of a ``$\lambda$-path in the model space'', with  the best parameter and the final model chosen using information criteria. According to \cite{Muller2013}, the most widely used information criterion is the marginal AIC criterion (\cite{Vaida2005}):
  \eq{
  \label{eq:vaida_aic}
  AIC := 2\LL(\hat{\theta}) + 2\alpha_n(p+q)
  }
  where $\hat\theta$ includes all the estimated parameters $(\beta, \Gamma)$, and $\alpha_n := n(n-p-q-1)$ for the finite sample case (\cite{Sugiura1978}). 
  %As discussed by \cite{Fang2011}, AIC is asymptotically equivalent to leave-one-out cross-validation, so it can be used for choosing between a finite number of models. AIC, however, is also known to be positively-biased (\cite{Ibrahim2011}). 
  Alternatively, LASSO-type methods (\cite{Krishna2008, Ibrahim2011}) use a BIC-type information criterion:
  \eq{
    BIC := 2\LL(\hat{\theta}) + \log(n)(p+q).
  }
BIC performs well in practice, but does not have theoretical guarantees~(\cite{schelldorfer2011estimation}).
  %One can refer to \cite{Muller2013, Buscemi2019Survey} for a detailed overview of different feature selection approaches.

\edit{Add more ICs (e.g. Muller)}

\section{Algorithms for Feature Selection}
\label{sec:pgd_methods}
We approach feature selection by adding a regularizer to model \eqref{eq:lme_diagonal_setup}: 
%( $\Lam_i$ are known and given, and $\Gamma$ an unknown diagonal matrix,   
%to be estimated along with the fixed effects $\beta$. 
%and add a regularizer:
%to the objective to obtain the optimization problem described
%in Theorem \ref{thm:basic existence2}:

%Feature selection is imposed by adding
%Consider a mixed-effect model setting described in Eq. (\ref{eq:lme_setup}) with $\Gamma$ being diagonal: $\Gamma = \Diag{\gamma}$.  
%We wish to find a minimizer of the problem 
%\eq{
%	\label{eq:lme_loss_original}
%	\min_{\beta\in\R^p, \gamma\in\R^q_+} & \LL(\beta, \gamma) + R(\beta, \gamma), 
%	}
\eq{%\label{eq:main opt in x over C}
    \label{eq:lme_loss_original_in_x}
    \min_{x } & \LL(x) + R(x)+ \delta_{\CC}(x),
%    \\
%    \text{s.t. } & x \in \CC\ .
}
where $x = (\beta, \gamma)$, 
$\CC:=\R^p\times \R^q_+$,
$\map{R}{\R^P\times\R^q_+}{\eR_+:=\R_+\cup\{+\infty\}}$
is a 
lower semi-continuous (lsc) regularization term, and
$\delta_{\CC}$ is the convex indicator function
\[
    \delta_{\CC}(x) := \begin{cases} 0, &  x \in \CC \\ +\infty, & x \not\in \CC .\end{cases}
\]  
By \cite[Theorem 2]{Theory1}, solutions to \eqref{eq:lme_loss_original_in_x}
always exist when $R$ has compact lower level sets.
% or 
% $\LL(\beta, \gamma)=\LL_{REML}(\beta,\Diag(\gamma))$ in \eqref{eq:reml_objective} with 
% \[
% \B=[\zero,\gam_\mmax\one]:=\left\{ \gamma\, |\, 0\le \gamma_i\le \gam_\mmax,\ i=1,\dots,q\right\}
% \]
% for some $\gam_\mmax>0$ where $\one$ is the vector of the appropriate
% dimension each of whose components is one.
% Since the method applies to either likelihood, hereafter we omit the subscripts and use $\LL$ to denote the objective function. %, leaving this choice up to a practitioner.
%For conciseness, define $x = (\beta, \gamma)$ and 
%$\CC:=\R^p\times \R^q_+$.
%such that $\tbeta \in \R^p$ is unconstrained and $\tgamma \in \R^q_+$ is non-negative:
%\eq{
% x\in \CC \textabove{$\Delta$}{=} \{[\beta', \gamma']: \  \beta' \in \R^p, \ \gamma' \in \R^q_+\} \subset \R^{p+q} 
%}
%With this notation, problem \eqref{eq:lme_loss_original} becomes
%\eq{%\label{eq:main opt in x over C}
%    \label{eq:lme_loss_original_in_x}
%    \min_{x \in \CC} & \LL(x) + R(x),
%    \\
%    \text{s.t. } & x \in \CC\ .
%}
%Equivalently, the box constraint can be moved to the loss in a form of an indicator function:
%or equivalently
%\[
%\eq{
%	\label{eq:lme_loss_original}
%    \min_{x} \LL(x) + R(x) + \delta_{\CC}(x)\ ,
%}
%where 
%$x = (\beta, \gamma)$, 
%$\CC:=\R^p\times \R^q_+$,  
%\quad \text{where}\ \, 
%\delta_{\CC}$ is the convex indicator function
%\[
%    \delta_{\CC}(x) := \begin{cases} 0, &  x \in \CC \\ +\infty, & x \not\in \CC .\end{cases}
%\]
The most common regularizers are separable taking the form
\begin{equation}\label{eq:R sep}
R(x) = \sum_{i = 1}^p r_i(x_i), 
\end{equation}
with typical choices for the component functions $r_i$ given in Table \ref{table:proxes}.

\subsection{Variable Selection via Proximal Gradient Descent}

Since $\LL$ is differentiable on its domain and 
proximal operator
for $\alpha R + \delta_{\CC}$ is 
computationally tractable, 
the Proximal Gradient Descent (PGD) Algorithm (e.g. see \cite{AB17}) offers a simple numerical
strategy for estimating first-order stationary points for 
\eqref{eq:lme_loss_original_in_x}.
% when 
%the application of the . 
%
%whenever the application of the proximal operator \cite{}
%to $\alpha R + \delta_{\CC}$ is 
%computationally tractable, a simple numerical method for estimating
%first-order stationary points for \eqref{eq:lme_loss_original_in_x} is
%the Proximal Gradient Descent (PGD) Algorithm. 
The proximal operator for 
$\alpha R + \delta_{\CC}$ is defined as the mapping
\[ %\eq{
    \prox_{\alpha R + \delta_{\CC}}(z) := \argmin_{y\in \CC}\ R(y) + \frac{1}{2\alpha}\|y - z\|_2^2 ,
\] %}
and the PGD iteration is given by
\[
    x^+ = \prox_{\alpha R + \delta_{\CC}}(x - \alpha \nabla \LL(x)),
\]
where $\alpha$ is a stepsize.
When $R(x)$ has the form given in \eqref{eq:R sep}, we have
\[
 \prox_{R}(z) = (\prox_r(z_1), \dots, \prox_r(z_q)) .
\]
\begin{table}[H]
\small
    \centering
    \begin{tabular}{|p{25.4mm}|c|c|}
        \hline
         \!\!Regularizer & $r(x)$, $x \in \R$ & $\prox_{\alpha r}(z)  $ \\
         \hline \hline
         LASSO ($\ell_1$) \newline (\cite{tibshirani1996regression}) & $|x|$ & $\sign(z)(|z| - \alpha)_+$ \\
         \hline
         A-LASSO \newline (\cite{Fan2001}) & $\bar{w}|x|$, $\bar{w} \geq 0$ &  $\sign(z)(|z| - \alpha\bar{w})_+$\\
         \hline
         SCAD \newline (\cite{fan1997comments}) & $\begin{cases} \sigma |x|, & |x| \leq \sigma \\ \frac{-x^2 + 2\rho\sigma x - \sigma^2}{2(\rho - 1)}, & \sigma < |x| < \rho\sigma \\ \frac{\sigma^2(\rho + 1)}{2}, & |x| > \rho\sigma \end{cases}$ & $\begin{cases} \sign(z)(|z| - \sigma\alpha)_+, & |z| \leq \sigma(1+\alpha) \\ \frac{(\rho - 1)z - \sign(z)\rho \sigma\alpha}{\rho - 1 - \alpha}, & \sigma(1+\alpha) < |z| \\
         &\quad\leq \max(\rho, 1+\alpha)\sigma \\ z, & |z| > \max(\rho,1+\alpha)\sigma \end{cases}$ \\
%         \hline
%         $\ell_p$, $0 < p < 1$ & $|x|^p$ & Coordinate Newton (\cite{Zheng2019SR3}) \\
         \hline
         $\delta_{\|x\|_0 \leq k}$  \newline ($\ell_0$ ball) & $\begin{cases} 0, & \#\{|x_i| \neq 0\} \leq k\\ \infty, & \text{ otherwise}\end{cases}$ & keep $k$ largest $|x_i|$, set the rest to 0 \\
         \hline
    \end{tabular}
    \caption{Proximal operators for commonly used sparsity-promoting regularizers. }
    \label{table:proxes}
\end{table}

\edit{Make section "Evaluating proximal operators" using Appendix}

Table \ref{table:proxes} provides closed form expressions for the proximal operators 
of commonly used regularizers. 
%All regularizers in this list are separable except $R(x) = \delta_{\|x\|_0 \leq k}$, for which the solution also has an analytically closed form.
%Again when $R$ is separable with the $r_i$ coming from Table \ref{table:proxes}, 
%the evaluation of 
For all of these cases,  the following theorem gives closed form expressions for
$\prox_{\alpha R + \delta_{\CC}}(z)$.

\begin{theorem}[$\prox$ for bounded $\gamma$]     \label{thm:prox_of_positive_quadrant}
We consider modified regularizers $r(\gamma)$ from the Table \ref{table:proxes} 
% and let $v = (\beta, \gamma)$. % \in \CC \subset R^{p + q}$.
that include an additional constraint on $\gamma$ of the form 
\[
0 \leq \gamma \leq \bar\gamma,
\]
for $\bar\gamma\in[0,+\infty]$.
%which which the positive orthant as a special case (i.e. $\bar\gamma = \infty$). 
We have the following results. 
%    \begin{enumerate}
%    \item  For $R(x)$ given by LASSO, A-LASSO, CAD, and SCAD, we have  for all $i$ that 
    %\eq{
    %\prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+}}(v_{p+q}))
    %}
    %where 
 %   \eq{
 %   \prox_{\alpha r + \delta_{\R_+}}(v_{i}) = \begin{cases} \prox_{\alpha r}(v_i), & v_i \geq 0 \\ 0, & \text{ otherwise} \end{cases} 
%    \item For $R(x) = \delta_{\|x\|_0 \leq k}$, the $\prox_{\alpha R + \delta_{\CC}}(v)$ can be evaluated by taking $k$ largest non-negative coordinates of $v$, and setting the rest to $0$.
 %   \end{enumerate}
%    Next, suppose $\LL = \LL_{REML}$. 
    \begin{enumerate}
    \item  For SCAD, we have for all $i$ that  
   % \eq{
%    \prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+q}))
 %   }
  %  where 
%    \eq{
\[
    \prox_{(\alpha r + \del_{[0,\bgam]})} %\delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}
    (\gamma_{i}) = \begin{cases} 
        \prox_{\alpha r}(\gamma_i), & 0 \leq \gamma_i < \bar\gamma  \\
        \bar\gamma, & \gamma_i \geq \bar\gamma \\
        0, & \text{ otherwise} \end{cases}
\]
%    }
    \item  For LASSO, A-LASSO we have for all $i$ that 
 %   \eq{
  %  \prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+q}))
  %  }
  %  where 
%    \eq{
\[
\prox_{(\alpha r + \del_{[0,\bgam]})}(\gamma_i) 
%    \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{i}) 
    = \begin{cases} 
        \prox_{\alpha r}(\gamma_i), & 0 \leq \gamma_i < \bar\gamma + \alpha \\
        \bar\gamma, & \gamma_i \geq \bar\gamma + \alpha \\
        0, & \text{ otherwise} \end{cases}
\]
%    }
    \item For $R(\cdot) = \delta_{\lev{\norm{\cdot}_0}{k}}$  %\delta_{\|x\|_0 \leq k}$, 
    the $\prox_{\alpha R + \delta_{\CC}}(\gamma)$ can be evaluated by taking $k$ largest coordinates of $\gamma$ such that $0 \leq \gamma_i \leq \bar\gamma$, and setting the 
    remainder to $0$.
    \end{enumerate}
\end{theorem}
The proof of the Theorem \ref{thm:prox_of_positive_quadrant} is provided in Appendix \ref{ch:proxes_appendix}. 
%\textcolor{red}{Aleksei, please check proof of appendix is all in terms of $\gamma$, rather than $v = (\beta, \gamma)$.}
The  PGD algorithm is detailed in Algorithm~\ref{alg:pgd_for_lme}.
{The algorithm's step-size $\alpha$ depends on the Lipschitz constant; an upper-bound is given in Appendix \ref{sec:lipschitz_constant}. In practice, $\alpha$ is computed using a line-search, since the available estimate for $L$ is very conservative.}

\smallskip

\begin{algorithm}[H]
\SetAlgoLined
$x = x_0$, $\alpha < \frac{1}{L}$,\text{ where } $\LL$ \text{ is $L$-Lipschitz}\\
 \While{not converged}{
    $x^+ = \prox_{\alpha R + \delta_{\CC}}(x - \alpha \nabla \LL(x))$;\\
 }
 \caption{\label{alg:pgd_for_lme}Proximal Gradient Descent for Linear Mixed-Effect Models}
\end{algorithm}
\medskip

\noindent

The main advantages of Algorithm \ref{alg:pgd_for_lme} are its simplicity and flexibility.
The main loop needs only the gradient and prox operator, and the structure of the algorithm is independent of the choice of regularizer $R$.
Algorithm \ref{alg:pgd_for_lme} locates first-order stationary points
under weak assumptions, in particular neither the objective nor the regularizer need to be convex~\citep{AB17,attouch2013convergence}.

%we use it only as 
%a point of reference for the algorithm that is the focus of our study which is presented in
%the Section \ref{sec:sr3_adaptation_to_lme}.
%The interested reader should consult the \textcolor{blue}{references \cite{} for the 
%various possible convergence properties
%of Algorithm \ref{alg:pgd_for_lme}}.

\edit{Add other optimization methods for FS in LMEs (e.g. Newton)}

\search{Add table of cross-references between optimization methods and regularizers}

\search{Add table with rates of convergence wherever possible}

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{figures/intuition_current.pdf}
    \caption{Proximal Gradient Descent (PGD) for~\eqref{eq:value_function_optimization} converges far faster than for~\eqref{eq:lme_loss_original_in_x}, because the SR3-value function $u_\eta$ yields more spherical level-sets than original likelihood for both convex components $\beta$ (first row) and non-convex components $\gamma$ (second row). 
    %The effect is particularly strong for $\gamma$-minimizers near the boundary of the constraint region (bottom-left panel).
    }
    \label{fig:geometric_intuition_sr3}
\end{figure}

\subsection{Variable Selection via MSR3}

To develop an approach that is both more efficient and accurate, we extend the SR3 regularization of~\cite{Zheng2019SR3} to LMEs. 
We call the extension MSR3, since we are focusing on mixed effects models. 
Starting with the regularized likelihood~\eqref{eq:lme_loss_original_in_x} we introduce auxiliary parameters designed to discover the 
fixed and random features: 
\eq{%\label{eq:main opt in x over C}
    \label{eq:lme_loss_sr3}
    \min_{x, w} & \LL(x) + R(w)+ \delta_{\CC}(x) + \kappa_{\eta}(x-w),
%    \\
%    \text{s.t. } & x \in \CC\ .
}
where $\kappa_\eta$ penalizes deviations between $x$ and $w$, and also guarantees that the objective is convex with respect to the $\gamma$
components of $x$ for sufficiently large $\eta$:
\begin{equation}
\label{eq:kappa}
\kappa_{\eta} (\beta, \gamma) = \frac{\eta}{2}\|(\beta,\gamma)\|^2 ,
\end{equation}
with $\eta\ge\bar\eta$ where $\bar \eta$ is the weak convexity constant
computed in~\cite[Section 5.1]{Theory1}.
As $\eta \uparrow \infty$, the extended objective~\eqref{eq:lme_loss_sr3} converges in an epigraphical sense to the original objective~\eqref{eq:lme_loss_original_in_x}. 
However, feature selection accuracy does not require this continuation, and a fixed 
modest value can be used~\citep{Zheng2019SR3}, e.g. $\eta=1$.

To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we define a value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
\eq{
\label{eq:value_function_definition}
		u_\eta(w) & = \min_x \ \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w)\\
		S_\eta(w) & = \argmin_x \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w).
}
Substituting (\ref{eq:value_function_definition}) into (\ref{eq:lme_loss_sr3}) the problem transforms into the problem of optimizing a regularized value function:
\eq{
	\label{eq:value_function_optimization}
	\min_{w} u_\eta(w) + R(w)
}
%Thus at a conceptual level, 
Here we have transformed the original regularized likelihood~\eqref{eq:lme_loss_original_in_x}  %has been transformed 
through relaxation and partial 
minimization to obtain an equivalent problem~\eqref{eq:value_function_optimization} 
for $w$ with the same regularizer. The value function $u_\eta$ encapsulates 
global variational information on the function $\LL(x) + \delta_{\CC}(x)$
relative to $w$.
%Problem~\eqref{eq:value_function_optimization} 
%attempts to find sparse $w$ that are small with respect to $u_\eta$.

In the case of linear regression, 
the function $u_\eta$ has a closed form solution~\cite{Zheng2019SR3}. However, in both  the linear regression context of~\cite{Zheng2019SR3} 
and in the LME context studied here, we need only compute $S_\eta(w)$ 
in order to optimize~\eqref{eq:value_function_optimization}. 
Indeed, in \cite[Section 5]{Theory1} it is shown that $u_\gamma$ is well-defined, 
differentiable, and Lipschitz continuous, with
%and we have a closed form solution for the derivative: 
\begin{equation}\label{eq:grad_value_function_explicit}
\nabla u_\eta(w) = \nabla_w k_\eta(x-w)|_{x = S_\eta(w)}=\eta(w-S_\eta(w)) .  
\end{equation}
%where $\nabla_w k_\eta(x-w)$ is computed directly from~\eqref{eq:kappa}.

Our empirical studies indicate that 
(\ref{eq:value_function_optimization}) has advantages over (\ref{eq:lme_loss_original_in_x})
from an optimization perspective since $u_\eta(w)$ typically has nearly spherical level-sets while keeping the position of minima close to those of $\LL(x)$. This effect is extensively studied and validated for a quadratic loss function in the original work  of \cite{Zheng2019SR3}. 
%we provide empirical evidence that suggests that it also extends to more general non-linear functions, specifically to the marginal likelihood. 
In Figure \ref{fig:geometric_intuition_sr3}, we plot the level-sets of $\LL(x) + \|x\|_1$ (left column) and $u(w) + \|w\|_1$ (right column) for the same mixed-effect problem. 
%A more symmetrical 
The more spherical geometry of the latter allows the Algorithm \ref{alg:pgd_for_value} 
(described below) to converge in 21 iterations, whereas Algorithm \ref{alg:pgd_for_lme} takes 1284 iterations. The difference is most pronounced when the minimum sits on the boundary of the feasible set, which is always the case for the variable selection problems with sparse support.
%In effect, Algorithm \ref{alg:pgd_for_value} incorporates global variational information 
%about the function $\LL$ while Algorithm \ref{alg:pgd_for_lme}

%The value function $u_\eta(w)$ possesses the following properties: 
%
%\begin{theorem}[Properties of value function optimization]
%\label{thm:properties_of_value_function}
%Let $\eta > 0$, $R(x)$ is level-compact. Then
%\begin{enumerate}
%	\item The solutions for (\ref{eq:value_function_optimization}) always exists when $R(w)$ is level-compact \cite[Theorem 6]{jimtheory}.
%	\item The solution set $S_\eta(w)$ is well-defined, single-valued and continuous \cite[Theorem 9]{jimtheory}
%	\item $u_\eta(w)$ is well-defined and continuous \cite[Theorem 9]{jimtheory}.
%	\item $u_\eta(w)$ is differentiable with $\nabla_w u_\eta(w) = \nabla_w \kappa_{\eta}(\bar{x}-w)$ where $\bar{x} = S_\eta(w)$ \cite[Theorem 11]{jimtheory}
%\end{enumerate}
%\end{theorem}

We apply PGD to optimize the value function $u_\eta(w)$ which yields the iteration of the form
\eq{
	\label{eq:pgd_for_value_function}
	w^+ = \prox_{\alpha^{-1}R}(w - \alpha\nabla u_\eta(w))
}
Because of the results stated above, all components of the iteration~(\ref{eq:pgd_for_value_function}) are well-defined. 
The equivalence of Algorithm \ref{alg:pgd_for_value} and 
\eqref{eq:pgd_for_value_function} is established in the following lemma, 
which extends the relationship studied by~\cite{Zheng2019SR3} to the case of $x = (\beta, \gamma)$.

%To get a deeper intuition for Algorithm~\ref{alg:pgd_for_value} we make the following remark. 
\begin{lemma}[Equivalence of Algorithms]\label{lem:equivalence}
Algorithm~\ref{alg:pgd_for_value} is equivalent to~\eqref{eq:pgd_for_value_function}. 
\end{lemma}
\begin{proof}
%Specifically, we define $\alpha$ to be block-specific: 
%	\[
%	\alpha = [\underbrace{\eta^{-1}, \dots, \eta^{-1}}_{p}, \underbrace{(\eta + \overline\lambda)^{-1}, \dots (\eta + \overline\lambda)^{-1}}_q]
%	\]
%so that we have
%	\eq{
%		\label{eq:grad_value_function_explicit}
%		\nabla u_\eta(w) = \alpha \odot (\overline{x} - w)|_{\overline{x} = S_\eta(w)}
%	}
%where ``$\odot$'' denotes the Hadamard or element-wise product.  
Substituting~(\ref{eq:grad_value_function_explicit}) into~(\ref{eq:pgd_for_value_function}), 
we see that the iteration~(\ref{eq:pgd_for_value_function}) is equivalent to the alternating minimization scheme outlined in the Algorithm~\ref{alg:pgd_for_value}.
\end{proof}
\begin{algorithm}[H]
\SetAlgoLined
$w = w_0$ \\
 \While{not converged}{
    $x^+ = \arg\min_x \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w)$ \\
    $w^+ = \prox_{\alpha^{-1} R}(x^+)$
 }
 \caption{\label{alg:pgd_for_value}Proximal Gradient Descent for Value Function}
\end{algorithm}

While in linear regression setting of of~\cite{Zheng2019SR3}, Algorithm~\ref{alg:pgd_for_value} can be implemented exactly, in the nonlinear case 
evaluating $x^+$ requires an iterative algorithm. In particular, we use an interior point method to account for non-negativity of $\gamma$.  
This means that the value function $u_\eta$ is approximately evaluated, using a Newton iteration with a barrier term.  
The degree of the approximation is controlled by the convergence criteria of the interior point algorithm.

%The alternating strategy is best suited for the simpler context , where the $x^+$ update has a closed form solution. In the case of~\eqref{eq:lme_loss_sr3}, the regularized loss is convex in $x$ because of the structure of $\kappa_\eta$, but is still nonlinear, and requires an iterative algorithm.

\paragraph{An Interior Point Method for Approximating $u_\eta$.}

In order to solve for the $x^+$ update in line 2 of Algorithm~\ref{alg:pgd_for_value}, we must optimize a convex loss with linear inequality constraints, that is, 
for a fixed $w = (\hat \beta, \hat \gamma)$, we need to solve 
\begin{equation}
\label{eq:ineqLoss}
\min_{\beta, \gamma} \mathcal{L}(\beta, \gamma) + \kappa_{\eta}(\beta - \hat \beta, \gamma - \hat \gamma) \quad \mbox{s.t.} \quad 0 \leq \gamma. 
\end{equation}
This problem is well suited for an interior point approach~\citep{Kojima1991,Nesterov1994,Wright1997}. 
%We now describe how to apply IP methods \teal{to solve the inner problem~\eqref{eq:lme_loss_relaxed} with respect to $x = (\beta, \gamma)$, 
%evaluating the value function $u(w)$ in~\eqref{eq:value_function_definition}}. 
%\eqref{eq:lme_loss_original}. 
% when $\LL=\LL_{ML}$. The case where $\LL=\LL_{REML}$ follows a 
% similar pattern.
%To apply IP methods to our problem class, we first 
First, the inequality constraint $0\le \gamma$ is relaxed using a log-barrier penalty, obtaining a minimization problem for a relaxed objective $\mathcal{L}_{\mu,\eta}$: 
\eq{
	\label{eq:log_barrier_relaxation_for_ip}
	\min_{\beta, \gamma} &\left\{\mathcal{L}_{\mu,\eta}(\beta, \gamma) :=  \mathcal{L}(\beta, \gamma) + \kappa_{\eta}(\beta - \hat \beta, \gamma - \hat \gamma) - \mu\sum_{i=1}^{q}\ln(\gamma_i) \right\}.
}
Here the log-barrier penalty approximates the indicator function to the positive orthant as $\mu$ decreases; indeed, the function $\gamma\mapsto\mu\ln(\gamma)$ epi-converges to the indicator function $\delta_{\mathbb{R}^n_+}(\gamma)$ as $\mu \downarrow 0$ (\cite{rockafellar2009variational}). The penalty (homotopy) parameter $\mu$ is progressively decreased to $0$ as the algorithm proceeds as described below. 
%in the limit minimizing $\mathcal{L}_{\eta,\mu}$ gives the right $x^+$ update in Algorithm~\ref{alg:pgd_for_value}. 
The existence of solutions for the problem~(\ref{eq:log_barrier_relaxation_for_ip}) for any positive $\mu$ is shown in \cite[Theorem 5]{Theory1}, and the convergence of solutions to the MSR3 solution as $\mu\downarrow 0$ is shown in~\cite[Theorem 7]{Theory1}.  
Finally, \cite[Theorem 6]{Theory1} shows that the MSR3 relaxation is consistent with respect to the barrier, so that as the MSR3 parameter $\eta \uparrow \infty$,
 limit points of global solutions to the former are global solutions to the latter. 
 However, in the applications considered here,
 the empirical studies in Sections \ref{sec:synthetic} and \ref{sec:real}
 indicate that one does not need to make $\eta$ particularly large in order to accurately identify the correct sparsity pattern.

%The Lagrangian for~\eqref{eq:log_barrier_relaxation_for_ip} is obtained by dualizing the inequality $\gamma \geq 0$ constraint: 
% Define the objective for \eqref{eq:log_barrier_relaxation_for_ip} by
% \eq{
% 	F_\mu(\beta, \gamma) & = \LL(\beta, \gamma) +  \frac{\lambda_b}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_g}{2}\|\gamma - \tgamma^{(k)}\|_2^2 - \mu\sum_{i=1}^{q}\log(\gamma_i) %+ v^T\gamma \\
% }

For $\gamma>0$, the necessary optimality conditions 
for $\mathcal{L}_{\mu,\eta}$ in $\gamma$ give us the relation 
%between dual variables $v$ and the primal variables $\gamma$: 
\eq{
	\nabla_\gamma \mathcal{L}_{\mu,\eta}(\beta,\gamma) = 
	%\begin{bmatrix}
\nabla_\gam \LL(\beta,\gam)+\eta (\gamma - \hat\gamma)
-\mu \Diag{\gam}^{-1}\one
=0,
%		v_1 - \mu/\gamma_1 \\
%		\dots \\
%		v_q - \mu/\gamma_q 
%	\end{bmatrix} = \begin{bmatrix}
%		0 \\
%		\dots \\
%		0 
%	\end{bmatrix} 
%	\implies v_i\gamma_i = \mu \text{ for } i=1,\dots,q.
}
where $\one$ is the vector of all ones of the 
appropriate dimension.
By setting 
\[
v=\nabla_\gam \mathcal{L}_{\mu,\eta}(\beta,\gam)+\eta(\gamma - \hat \gamma),
\] 
we can rewrite this equation as
%This equation can be written in a matrix form as 
\eq{
	\label{eq:complementary_slackness_kkt}
	v\odot\gamma - \mu\one = 0,
}
where $\one$ is the vector of all ones of the appropriate dimension and
``$\odot$'' denotes the Hadamard (or simply element-wise) product.
%Together with the remaining optimality conditions, we obtain a set of nonlinear equations  that form the the KKT system for~\eqref{eq:log_barrier_relaxation_for_ip}:
The complete set of optimality conditions for 
\eqref{eq:log_barrier_relaxation_for_ip} can now be written as
\eq{\label{eq:IP equations}
	G_{\mu,\eta}(v, \beta, \gamma) & 
%	= \begin{bmatrix}
%		\nabla_v F_\mu \\
%		\nabla_\beta F_\mu \\
%		\nabla_\gamma F_\mu 
%	\end{bmatrix} 
:= \begin{bmatrix}
v\odot \gamma - \mu\one \\
\nabla_\beta \LL(\beta, \gamma) + \eta(\beta - \hat \beta) \\
\nabla_\gamma \LL(\beta, \gamma) + \eta(\gamma - \hat \gamma) - v
\end{bmatrix} = 0.
}
We then apply Newton's method to~\eqref{eq:IP equations}, 
that is, in each iteration the search direction $[\Delta v, \Delta \beta, \Delta \gamma]$ solves the linear system
\eq{
	\label{eq:ip_iteration_rs_form}
	\nabla G_{\mu,\eta}(v, \beta, \gamma)\begin{bmatrix}
		\Delta v \\
		\Delta \beta \\
		\Delta \gamma
	\end{bmatrix} = -G_{\mu,\eta}(v, \beta, \gamma).
}
where 
\eq{
	\nabla G_{\mu, \eta}(v, \beta, \gamma) = \begin{bmatrix}
 		\Diag{\gamma} & 0 & \Diag{v} \\
 		0 & \nabla^2_{\beta\beta}\LL + \eta I & \nabla^2_{\beta\gamma} \LL\\
 		-I & \nabla^2_{\gamma\beta}\LL & \nabla^2_{\gamma\gamma}\LL + (\eta + \overline{\lambda})	 I
 	\end{bmatrix}
}
and we have used the fact that  $v\odot \gamma=\Diag{v}\gam=\Diag{\gam}v$. 
The exact formulae for the derivatives of $\LL$ are provided in the Appendix~\ref{appendix:derivatives_of_lmm}.

The general structure of the algorithm is as follows.
Given a search direction
$[\Delta v^{(k)}, \Delta \beta^{(k)}, \Delta \gamma^{(k)}]$, 
choose a step of size $\alpha_k>0$
%\eq{
%    \label{eq:ip_step}
%	v^{(k+1)} & = v^{(k)} + \alpha_k \Delta v \\
%	\gamma^{(k+1)} & = \gamma^{(k)} + \alpha_k \Delta\gamma \\
%	\beta^{(k+1)} & = \beta^{(k)} + \alpha_k \Delta \beta
%}
%with $\alpha_k$ 
so that the update
\[
\begin{pmatrix}
v^{(k+1)}\\ \beta^{(k+1)}\\ \gamma^{(k+1)}
\end{pmatrix}
=
\begin{pmatrix}
v^{(k)}\\ \beta^{(k)}\\ \gamma^{(k)}
\end{pmatrix}
+\alf_k
\begin{pmatrix}
\Delta v^{(k)}\\ \Delta\beta^{(k)}\\ \Delta\gamma^{(k)}
\end{pmatrix}
\]
satisfies the conditions
% chosen to satisfy positivity and sufficient descent conditions:
\eq{
    \label{eq:ip_steplen_conditions}
	\text{\textit{Positivity:}} \qquad & \gamma^{(k+1)} > 0,\ v^{(k+1)} > 0 \\
	\text{\textit{Sufficient Descent:}}\qquad & \| G_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|G_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| .
}
At each iteration the relaxation parameter $\mu$ is updated by the formula 
\eq{
    \label{eq:ip_mu_update}
    \mu^{(k+1)} = {v^{(k)}}^T\gamma^{(k)}/q,
}
where ${v^{(k)}}^T\gamma^{(k)}$ is the duality gap at
iteration $k$. The algorithm terminates when the criteria 
\eq{
\label{eq:ip_convergence_criterion}
\begin{aligned}
\|G_{\mu,\eta}(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| &\leq \text{\texttt{tol}}\\
\mu &\leq  \text{\texttt{tol}} 
\end{aligned}
}
are both satisfied, so the interior point problem is nearly stationary, and closely approximates the original problem~\eqref{eq:ineqLoss}.
MSR3 is summarized in Algorithm~\ref{alg:MSR3}, which approximates Algorithm~\ref{alg:pgd_for_value} as the tolerance goes to $0$. 
In the numerical experiments, we use $\text{\texttt{tol}} = 10^{-5}$, and accuracy does not change as the tolerance parameter decreases.

\begin{algorithm}[H]
\SetAlgoLined
$w = w_0$ \\
 \While{not converged}{
    $x^+ \; \mbox{satisfies }\;  \|G_{\mu,\eta}(v^+, x^+)\| \leq \text{\texttt{tol}}, \; \mu \leq \text{\texttt{tol}} $\\
    $w^+ = \prox_{\alpha^{-1} R}(x^+)$
 }
 \caption{\label{alg:MSR3} MSR3}
\end{algorithm}

\paragraph{Positive Approximation of the Hessian}
For many datasets the weak convexity constant $\overline \eta$ can be extremely large 
and difficult to compute. However, if $\overline \eta$ is too small $\nabla^2_{\gamma\gamma}\LL(\beta, \gamma)$ is 
negative-(semi)definite. Negative definite Hessians can hamper the convergence of 
second-order methods (e.g., see \cite{nocedal2006numerical}). 
Therefore, one must take care in selecting $\eta$. For this, we recall from
\cite[Lemma 3]{Theory1} that
\begin{equation} %\label{eq:hess LL}
\nabla^2\LL{(\beta,\gam)}=\sum_{i=1}^m
S_i^T\begin{bmatrix}X_i^T\\ -Z_i^T\end{bmatrix}
\Omega_i(\gam)^{-1}
\begin{bmatrix}X_i& -Z_i\end{bmatrix}S_i
-\begin{bmatrix}
0&0\\ 0& \half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}
\end{bmatrix}.
\end{equation}
This implies that negative eigenvalues for the Hessian must arise from the
Hessian with respect to $\gamma$,
$\nabla^2_{\gamma\gamma}\LL(\beta, \gamma)$, and more specifically, the
term $(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}$. 
A positive semidefinite approximation to the Hessian is obtained by simply dropping this term.

\subsection{Relaxation and Efficient Algorithms: MSR3 and MSR3-Fast }
\label{sec:synthetic}

While algorithm~\eqref{alg:pgd_for_value} is modular, it requires solving a nonlinear optimization problem in $x = (\beta, \gamma)$ for each single update
of $w = (\hat \beta, \hat \gamma)$. To make the implementation as efficient as possible, we designed a more balanced updating scheme, that 
alternates Newton iterations as described in the interior point algorithm with $w$ updates. We update $w$ whenever we are sufficiently close 
to the `central path' in the interior point method, a condition that can be checked rigorously using optimality conditions. 
This scheme is detailed in Algorithm~\ref{alg:pgd_for_value_function_fast}.

\begin{algorithm}[H]
\SetAlgoLined
$\texttt{progress}\leftarrow \textbf{True}$; \quad \texttt{iter = 0}; \\
$\beta^+, \tbeta^+\leftarrow\beta_0$; 
\quad $\gamma^+, \tgamma^+\leftarrow\gamma_0$;  
\quad $v^+ \leftarrow 1 \in \R^q$; 
\quad  $\mu \leftarrow \frac{{v^+}^T\gamma^+}{10 q}$\\
 \While{\texttt{iter} $<$ \texttt{max\_iter}  \ and \ $\|G_\emu(\beta^+, \gamma^+, v^+)\|$ $>$ \texttt{tol}   \ and  \ \texttt{progress} \\}{
    $\beta \leftarrow \beta^+$; \quad $\gamma \leftarrow \gamma^+$; \quad $\tbeta \leftarrow \tbeta^+$; \quad $\tgamma \leftarrow \tgamma^+$ \\
%    $A \leftarrow \nabla G_\emu((\beta, \gamma, v), (\tbeta, \tgamma))$\\
  %  $b \leftarrow G_\emu((\beta, \gamma, v), (\tbeta, \tgamma))$\\
    $[dv, d\beta, d\gamma] \leftarrow  \nabla G_\emu((\beta, \gamma, v), (\tbeta, \tgamma))^{-1}  G_\emu((\beta, \gamma, v), (\tbeta, \tgamma))$ \tcp*[f]{Newton Iteration}\\ 
    $\alpha \leftarrow 0.99\times\min\left(1, -\frac{\gamma_i}{d\gamma_i}, \forall i :\ d\gamma_i < 0\right)$\\
    $\beta^+ \leftarrow \beta + \alpha d\beta$; \quad $\gamma^+ = \gamma + \alpha d\gamma$; \quad  $v^+ \leftarrow v + \alpha dv$\\
    \If{$\|\gamma^+\odot v^+ - q^{-1}{\gamma^+}^Tv^+ \mathbf{1}\| > 0.5q^{-1}{v^+}^T\gamma^+$}{
  %  	\tcp*[h]{Not in the neighborhood of the central path yet}\\
    	continue \tcp*[f]{Keep doing Newton iterations}\\
    }
    \Else{ 
        $\tbeta^+ = \prox_{\alpha R}(\beta^+)$;
        \    $\tgamma^+ = \prox_{\alpha R + \delta_{\R_+}}(\gamma^+)$; 
        \    $\mu = \frac{1}{10}\frac{{v^+}^T\gamma^+}{q}$ \tcp*[f]{Near central path} 
    }
%	\tcp*[h]{Keep iterating until convergence} \\
    \texttt{progress} = ($\|\beta^+ - \beta\| \geq \text{tol}$ or $\|\gamma^+ - \gamma\|  \geq \text{tol}$ or $\|\tbeta^+ - \tbeta\| \geq \text{tol}$ or $\|\tgamma^+ - \tgamma\| \geq \text{tol}$)\\
    \texttt{iter += 1}
 }
 \Return{$\tbeta^+$, $\tgamma^+$}
 \caption{\label{alg:pgd_for_value_function_fast}MSR3-fast (Optimized Proximal Gradient Descent for the Value function)}
\end{algorithm}

\section{Verifications}
\label{sec:applications}

\subsection{MSR3 for Covariate Selection}
\label{ch:sr3_l1}

\begin{table}
\centering
\input{tables/performance_table_short_current}
\caption{\label{table:comparison_of_algorithms} Comparison of performance of algorithms measured as accuracy of selecting the correct covariates and run-time. The L0 strategy stands out 
over other standard regularizers. MSR3 improves performance significantly for all regularizers, while MSR3-fast improves convergence speed while preserving the 
accuracy of MSR3.  
More detailed results are in the Table \ref{table:detailed_comparison_of_algorithms} of Appendix \ref{ch:appendix_detailed_comparison}.}
\end{table}

\begin{figure}
    \centering
	\includegraphics[width=1.0\textwidth]{figures/performance_picture_current}
	\caption{\label{fig:box_with_whiskers_for_synthetic_data}	Feature selection accuracy and execution time in seconds for PGD 
	(Algorithm \ref{alg:pgd_for_lme}), 
	MSR3 (Algorithm \ref{alg:pgd_for_value}), and MSR3-fast
	(Algorithm \ref{alg:pgd_for_value_function_fast}) with various regularizers. 
	MSR3-Fast has the same accuracy as 
	MSR3 and significantly decreases computation time.}
\end{figure}

In this section we 
{compare the numerical performance 
the feature selection accuracy and numerical efficiency of Algorithms 
\ref{alg:pgd_for_lme} and 
\ref{alg:pgd_for_value_function_fast} when using the 
LASSO, A-LASSO, SCAD, and L0 sparsity regularizers.
We begin by describing how the data is generated for our numerical simulations 
followed by a description of how the  
regularization parameter $\lambda$ and the coupling parameter $\eta$  were chosen.
Our experiments on real data are presented in Section \ref{sec:real}.}

%show the performance of the `regularization-agnostic' feature selection method, and also show that MSR3 improves performance across different regularization strategies. Specifically, we compare the feature selection accuracy and numerical efficiency of Algorithms  \ref{alg:pgd_for_lme} and 
%\ref{alg:pgd_for_value_function_fast} 
%with LASSO, A-LASSO, CAD, SCAD, and L0.

%\paragraph{LASSO path} The LASSO path is a set of solutions $x$ parametrized by $\lambda$, where $\lambda$ is sweeping over its range from $\lambda = 0$ to $\lambda \to \infty$, until no parameters are included in the model. It's known that a classic LASSO tends to include false positives early along the path (\cite{Su2017}).

\paragraph{Experimental Setup.} The number of fixed effects $p$ and random effects $q$ are set at $20$ with $\beta = \gamma = [\frac{1}{2}, \frac{2}{2}, \frac{3}{2}, \dots, \frac{10}{2}, 0, 0, 0, \dots, 0]$, i.e. the first 10 covariates are increasingly important and the last 10 covariates are not. The data is generated as 
\[
\begin{aligned}
y_i &= X_i\beta + Z_iu_i + \varepsilon_i, \quad  \varepsilon_i \sim \NN(0, 0.3^2 I) \\
X_i &\sim \NN(0, I)^p, \quad Z_i = X_i \\ 
u_i& \sim \NN(0, \Diag{\gamma}),\\ 
\end{aligned}
\]
with 9 groups of sizes $[10, 15, 4, 8, 3, 5, 18, 9, 6]$. 
The data generation is repeated 100 times in order to estimate
the uncertainty bounds. The smallest non-zero components in the generated signals are just above the level of observation noise. 
%\red{We need to say something about "the smallest non-zero components are such that they are just above the noise from observation error"}

\vskip -.8in
\begin{wrapfigure}{r}{7.5cm}
		\includegraphics[width=7.5cm]{figures/eta_L1.pdf}
		\caption{Dependence of model performance on the relaxation $\eta$ for a sample problem.}\label{fig:eta}
	\end{wrapfigure}

\paragraph{Parameter Selection.} The regularization parameter $\lambda$ 
multiplying $R$ and the 
coupling parameter $\eta$ restricting the difference between 
$(\beta,\gamma)$ and $(\tbeta,\tgam)$ are chosen to maximize a classic 
BIC criterion from \cite{Jones2011}. 
We begin by setting a log-uniform grid of 20 candidate values for the parameter $\eta \in [10^{-4}, 10^2]$, and then for each value of $\eta$
on this grid 
the BIC is optimized using a golden search in $\lambda \in [0, 10^5]$. The final values of $\eta$ and $\lambda$ are then chosen
to be those that maximize the BIC criterion.

	Figure \ref{fig:eta} shows the dependence of accuracy on the values of $\eta$ for the first data set generated in our test set. There are three distinct regions, corresponding to loose, moderate, and tight levels of coupling. When $\eta$ is small the coupling term does not have sufficient strength and the training does not progress far from the initial point (a fully dense vector \textbf{1} in this case). When the coupling is tight, the level-sets of the problem converge to those of the original problem, and thus the minimizer. For the values in between, the coupling significantly improves the model's accuracy. 
These results are consistent with experiments in the sparse linear regression setting~\cite{Zheng2019SR3}.

\paragraph{Results.}
The experimental results are presented in the Table \ref{table:comparison_of_algorithms} and Figure~\ref{fig:box_with_whiskers_for_synthetic_data}. MSR3 improves the selection accuracy of most regularization techniques 
described in Table \ref{table:proxes}, showing a near-perfect performance, while converging two orders of magnitude faster in wall-clock time.

\paragraph{Comparison to \texttt{glmmLasso} and \texttt{lmmLasso}.}
We used \cite[Table 3]{Buscemi2019Survey} as a reference 
for feature selection libraries. Of the 17 entries mentioned, the four libraries that successfully ran on our synthetic data described above were  packages \texttt{glmmLasso}\footnote{\href{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}} (\cite{groll2014variable}), \texttt{lmmLasso}\footnote{\href{https://rdrr.io/cran/lmmlasso/}{https://rdrr.io/cran/lmmlasso/}}(\cite{schelldorfer2011estimation}), \texttt{fence}\footnote{\href{https://rdrr.io/cran/fence/}{https://rdrr.io/cran/fence/}} (\cite{jiang2008fence}) and \texttt{PCO} (\cite{lin2013pco}) libraries. \texttt{fence} caused a memory overflow on the experimental system during the performance evaluation on the datasets described above. We could not evaluate {\texttt{PCO} because
it did not support  datasets where the total number of random effects $mq$ exceeded the total number of observations $n$. We compare performance of MSR3 
(available through the open source \texttt{pysr3} library) 
to the performance of the \texttt{R} packages \texttt{glmmLasso}\footnote{\href{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}} (\cite{groll2014variable}) and \texttt{lmmLasso}\footnote{\href{https://rdrr.io/cran/lmmlasso/}{https://rdrr.io/cran/lmmlasso/}} (\cite{schelldorfer2011estimation}) which are the functionally closest libraries available
} online. As of this writing, \texttt{glmmlasso} does not allow the user to specify $\Gamma$ as a diagonal matrix.  Since the diagonal specification simplifies the problem, this puts \texttt{glmmlasso} package at a disadvantage in our numerical comparison. We evaluate all algorithms' performance on the same set of problems as described above. We tuned the hyperparameters of \texttt{glmmLasso} and \texttt{lmmLasso} by minimizing the BIC scores provided by the libraries over $\lambda \in [0, 10^5]$. The results are presented in Table \ref{table:glmmlasso}. Overall, MSR3 executes, on average, 5 times faster in wall-clock time than \texttt{glmmLasso} and 60 times faster than \texttt{lmmLasso} and shows much higher accuracy of selecting correct fixed and random effects simultaneously. 
{The accuracy of \texttt{glmmLasso} is lower relative to the other libraries' scores likely due to its BIC selection criterion choosing dense models.} The package \texttt{lmmLasso} supports the diagonal specification of $\Gamma$, thus allowing a direct comparison with the scores from \texttt{pysr3}.  \texttt{lmmLasso} yields a competitive accuracy of selecting random effects but \texttt{lmmLasso} provides dense solutions for fixed effects $\beta$ for chosen values of $\lambda$.

\begin{table}[h]
\centering
\input{tables/table_glmm_current.tex}
\caption{\label{table:glmmlasso} Comparison of performance of MSR3-Fast for $\ell_1$ regularizer vs \texttt{glmmLasso}. MSR3-Fast executes 5 times faster in wall time and has higher accuracy of selecting correct covariates. 
%Importantly, the accuracy of \texttt{glmmLasso} is likely skewed downwards due to BIC selection criterion choosing dense ultimate models and due to the missing option to constrain the matrix $\Gamma$ to be diagonal. \texttt{lmmLasso} supports the diagonal specification of $\Gamma$ which translates into a competitive quality of selecting random effects. However, while finding sparse fixed effects, \texttt{lmmLasso} provides dense solutions for fixed effects $\beta$.
}
\end{table}

\subsection{Experiments on Real Data}
\label{sec:real}

\begin{figure}
    \centering
	\caption{\label{fig:bullying_data_random_feature_selection}Validation of Random Feature Selection for Bullying Data from GBD 2020. 
	Left panel shows coefficient paths across numbers of nonzero covariates allowed in the model using the $\ell_0$ regularizer. 
	Right panel evaluates each choice against expert knowledge. 
	The algorithm picks seven historically significant covariates and two historically insignificant, for the model selected using the BIC criteria. 
	See the Appendix \ref{appendix:bullying_covariates} for covariates description and assessment of significance.}
	\includegraphics[width=1\textwidth]{figures/bullying_data_assessment_selection}
\end{figure}

In this section we validate the MSR3-empowered $\ell_0$-regularized mixed-effect model ($R(x) = \delta_{\|x\|_0 \leq k}$ from Table \ref{table:proxes}) by using it to identify the most important covariates in real data on relative risk of anxiety and depressive disorders depending on the exposure to bullying in young age\footnote{Institute for Health Metrics and Evaluation (IHME). Bullying Victimization Relative Risk Bundle GBD 2020. Seattle, United States of America (USA), 2021.}. This research has been a part of Global Burden of Diseases study for the last several years. The end goal  is to estimate the burden 
{through disability adjusted life years} (DALYs)~\citep{murray1997understanding}
of major depressive disorder (MDD) and anxiety disorders that are caused by bullying. For this risk factor, the exposure is primarily concentrated in childhood and adolescents, but the risk for MDD and anxiety disorders is anticipated to continue well into adulthood. This elevated risk is, however, expected to decrease with time as other risk factors come into play in adulthood (unemployment, relationship issues, etc.). To accommodate this, the research team uses the models which estimate the relative risk (RR) of MDD and anxiety disorders among persons exposed to bullying depending on how many years it has been since the first exposure. Studies informing the model were sourced from a systematic review and consist of longitudinal cohort studies. They measure exposure to bullying at baseline, and then follow up years later and assess them for MDD or anxiety disorders. The detailed description of the covariates can be found in Appendix \ref{appendix:bullying_covariates}.

The feature selection process is illustrated on Figure~\ref{fig:bullying_data_random_feature_selection}. 
%Since there is no prior on $k$ -- the number of features to keep in the model -- 
Here, the BIC criterion from \cite{Jones2011} was used to select $k$, which suggests $k=4$ or $5$.  
 For the $k=4$ case, the selected covariates (\texttt{intercept}, \texttt{time}, \texttt{cv\_threshold\_bullying}, \texttt{cv\_b\_parent\_only}) 
are known as important and were used in the analysis in previous years of GBD. For the $k=5$ case, 
%five covariates (in addition to \texttt{intercept} and \texttt{time}) were selected by \ouralgo: \texttt{cv\_unadjusted}, \texttt{cv\_threshold\_bullying}, \texttt{cv\_b\_parent\_only}, \texttt{cv\_anx}, and \texttt{percent\_female}. 
%in addition to the covariates already known to be important, 
the algorithm also selects  \texttt{cv\_child\_baseline} and \texttt{cv\_or}, which were not used before. The  \texttt{cv\_child\_baseline} covariate describes whether the midpoint in the sample is above or below 13. 
The  \texttt{cv\_or} variable describes whether the estimate is a relative risk or odds ratio. The selection of these variables suggests a closer look at the data reporting mechanisms across studies. 
For example, there is an active literature on converting estimates between relative risks and odds ratios~\cite{grant2014converting,wang2013converting}.

\section{Theoretical Analysis}
\ask{Jim}{What would be the best way to introduce theoretical results to my thesis?}

\section{Software Implementation}
To ensure reproducibility of this research, all new algorithms have been implemented as a part of the \texttt{pysr3}\footnote{Available at \href{https://github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}} library. This library implements functionality for fitting linear mixed models and selecting covariates. The user interface was designed to be fully compliant with the standards\footnote{\href{https://scikit-learn.org/stable/developers/develop.html}{https://scikit-learn.org/stable/developers/develop.html}} of \texttt{sklearn} library to minimize learning time.

\section{Discussion}
\label{sec:discussion}

    In this paper, we developed and implemented a first-order variable selection framework for LMEs that handles convex and nonconvex regularizers. We also showed that 
    the MSR3 relaxation \eqref{eq:value_function_optimization} 
%a regularizer-agnostic transformation of the likelihood, 
improves the covariates selection accuracy of a wide group of popular sparsity-promoting regularizers. The fact that the relaxation improves accuracy, rather than just serving as a means to {numertical efficiency, %solve the original problem,
     is very interesting and deserves future study.} % result that can be studied in future work.

    Since the LME relaxation does not have a closed form, we used an interior method to evaluate the requisite value function. We also developed 
    a more efficient version of the algorithm  (MSR3-Fast) that interleaved interior point iterations with updates of the auxiliary variables, and this method was chosen for the 
    open source library \texttt{pysr3}. % that we provided. 
    %convergence of this method, and illustrate its performance on synthetic and real-world data. 
    %The relaxation shares minima with the original function while having more symmetrical level-sets which accelerate the convergence of gradient-based methods. 
{ Numerical experiments on synthetic data showed that the MSR3 approach for variable selection extends regions of hyper-parameter values where the highest accuracy is achieved, making it easier for information criteria to select the best model.} 
 The variable selection library for the accelerated method MSR3-Fast is much faster than currently available software, and allows the MSR3 approach to be easily applied to a range of regularizers that have 
{computationally efficient prox operators.} %prox operators available.

    The main analytic limitations of the proposed method stem from a lack of an analytical representation of the value function in the MSR3 relaxation for LMEs
    \eqref{eq:value_function_optimization}. 
    However, the MSR3 framework (Algorithm \ref{alg:pgd_for_value})
    incorporates global variational information about
    the likelihood $\LL$ into the PGD algorithm whereas the standard application of the PGD algorithm (Algorithm \ref{alg:pgd_for_lme}) only uses a local linear approximation to
    $\LL$ at each iteration. This difference reveals itself in both the increased speed and accurate of the MSR3 approach on this class of problems.
{In contrast to SR3 in linear regression settings, where the CG method can be efficiently used to evaluate the value function (see e.g.~\cite{Baraldi2019}),  
 the nonlinear optimization problem required for LMEs is more difficult.} 
%\red{(I'm not sure of the point of the previous sentence.)} 
{Although the use of
Hessian information makes each iteration computationally efficient, it limits the size of the problems to which the method can be applied. 
On the other hand, switching to first-order methods for the inner problem inside the relaxation may be prohibitively slow. 
A potential path to balance these limitations is to develop efficient upper-bounding models for the value function that can be evaluated more efficiently.
}

    The suggested methodology can be expanded to a wider class of models. In particular, one can extend MSR3 to the setting of non-linear mixed-effect models or generalized linear mixed models, which are known to be challenging setups for covariate selection tasks. Both of these problem classes face require optimizing highly nonlinear objective functions that arise when we consider marginal likelihoods. 
The SR3  approach may allow new avenues for more efficient strategies, analogous to what was done here for LMEs.

### 2023-04-06 phd_thesis.tex
Initiate PhD thesis structure

--- 2022-07-04 general_sholokhov.tex
@@
-\documentclass[11pt,letterpaper]{article}
+\documentclass[12pt,a4paper]{book}
 
@@
 \input{latex-macros/useful_packages}
-\input{latex-macros/math_macros}
+\input{math_macros-jim}
+
+\usepackage{natbib}
+% Color 
+\newcommand{\blue}[1]{\textcolor{blue}{#1}}
+\newcommand{\red}[1]{\textcolor{red}{#1}}
+\newcommand{\plum}[1]{\textcolor{Plum}{#1}}
+\newcommand{\teal}[1]{\textcolor{teal}{#1}}
 
@@
 \newcommand{\projectNameShort}{Feature Selection for Mixed Models}
-\newcommand{\doctitle}{Thesis Proposal}
+\newcommand{\doctitle}{Doctoral Thesis}
 \newcommand{\projectNameLong}{}
@@
 \textbf{Committee}\\
-Aleksandr Aravkin (chair),\\
+Aleksandr Aravkin (co-chair),\\
+Nathan Kutz (co-chair), \\
 James Burke, \\
-Nathan Kutz, \\
 Jonathan Wakefield (GSR)
@@
 \\[40mm]
-{A thesis proposal submitted in partial fulfillment\\ of the requirements for the degree of}
+{A thesis proposal submitted in fulfillment\\ of the requirements for the degree of}
 \\[5mm]
@@
 
-\newpage
-\section*{Notation}
+\listoftodos
 
-\begin{center}
-	\begin{equation}
-		\begin{split}
-			n & \text{ -- Total number of objects (observations) in a dataset} \\
-			m & \text{ -- Total number of groups (clusters, studies) in a dataset} \\
-			n_i & \text{ -- Number of objects (observations) in the group } i. \\
-			p & \text{ -- Total number of fixed features (covariates, predictors)} \\
-			q & \text{ -- Total number of random features (covariates, predictors)} \\
-			X_i & \text{ -- Data (design) matrix of fixed features of the group $i$, } X_i \in \R^{n_i \times p} \\
-			Z_i & \text{ -- Data (design) matrix of random features of the group $i$, } Z_i \in \R^{n_i \times q} \\
-			Y_i & \text{ -- Target variable for the group } i \\
-			\beta & \text{ -- Vector of fixed (mean) coefficients in a model, } \beta \in \R^{p} \\
-			\Gamma & \text{ -- Covariance matrix of random effects in a model, } \Gamma \in \R^{q\times q} \\
-			\gamma & \text{ -- Vector of variances of random effects when random effects are independent: } \Gamma = \diag(\gamma), \ \gamma \in \R^{q} \\
-			u_i & \text{ -- Random effects for the group $i$, } u_i \in \R^q, \ u_i \sim \NN(0, \Gamma) \\
-			\Lambda_i & \text{ -- Covariance matrix for measurement (observation) errors for the group $i$} \\
-			\sigma_i & \text{ -- Vector of variances for measurement (observation) errors for the group $i$ when } \Lambda_i = \diag(\sigma_i) \\
-			\varepsilon_i & \text{ -- Vector of measurement (observation) errors for the group $i$, } \varepsilon_i \sim \NN(0, \Lambda_i) \\
-			\Omega_i & \text{ -- Covariance matrix of within-cluster observations, } \Omega_i = \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i, \ \Omega_i \in \R^{n_i \times n_i} \\
-			k & \text{ -- Number of non-zero mean coefficients allowed in a model} \\
-			s & \text{ -- Number of non-zero variances of random effects allowed in a model} \\
-			\tbeta & \text{ -- Sparse counterpart of $\beta$, } \tbeta \in \R^p, \ \|\tbeta\|_0 = k \\
-			\tgamma & \text{ -- Sparse counterpart of $\gamma$, } \tgamma \in \R^q, \ \|\tgamma\|_0 = s \\
-			\LL(u_i\ |\ \beta, \gamma) & \text{ -- Negative log-likelihood of random effects of a linear mixed model}\\
-			\LL_{ML}(\beta, \gamma) & \text{ -- Negative marginalized log-likelihood of parameters of a linear mixed model} \\
-			\LL_{REML}(\beta, \gamma) & \text{ -- Negative marginalized restricted log-likelihood (REML) of a linear mixed model}\\
-			\LL_{r}(\beta, \gamma) & \text{ -- Relaxed marginalized log-likelihood of a linear mixed model} \\
-			a \circ b & \text{ -- Element-wise (Hadamard) product of $a$ and $b$} \\
-		\end{split}
-	\end{equation}
-\end{center}
+\chapter{MSR3}
+\include{msr3}
 
-\newpage
+\chapter{PINODE}
+\include{pinode}
 
-\section{Introduction}
-Linear Mixed-Effects Models are widely used technique for modeling clustered data, such as cohort studies, longitudinal data analysis, and meta-research.
-A key step in these types of works is feature selection, which goal is choosing the most useful predictors from a large set of possible covariates.
-In the next section we introduce the reader to the concept of linear mixed-effect models and provide a literature review on the most commonly used optimization methods for finding estimators for them, such as Newton method, EM-algorithm, and Interior Point Method.
-Section \ref{ch:feature_selection_for_mixed_models} gives reader an overview of existing covariates selection methods, such as subset selection, shrinkage penalties, and AIC/BIC-based methods.
-Section \ref{ch:relax_and_split_thesis_work} introduces Relax-and-Split approach to feature selection, outlines the proposed algorithm, and discusses optimization subroutines involved in it.
-In the Section \ref{ch:experiments_on_synthetic_data} we test our approach in two simulation setups, with the goals to compare the proposed algorithm to its competitors (Section \ref{ch:comparison_to_other_algo}) and to show its applicability conditions and current limitations (Section \ref{ch:scalability_experiments}).
-Section \ref{ch:experiments_on_real_data} shows the application of our approach to two real-world problems: the estimation of burden of anxiety and depression disorders as results of bullying (Section \ref{ch:bullying}), and the covariates selection for modeling the contact rate for COVID-19 (Section \ref{ch:covid}).
-Section \ref{ch:future_work} discusses shortcomings of the current version of the algorithm and outlines theoretical questions and practical improvements which we plan address as a part of the thesis work.
-
-\label{ch:literature_review}
-\subsection{Linear Mixed-Effects Models}
-Mixed-effect regression models are a particular kind of regression models for clustered data.
-They describe the relationship between an outcome variable and its predictors in a setting where the observations are grouped according to one or many criteria.
-These models are used in a wide range of applications, such as longitudinal studies for clinical trials, population health problems, risk-outcome studies, and meta-research.
-
-Mathematically, let there be $m$ groups of observations indexed by $i$, with sizes $n_i$ respectively, and the total number of observations to be $n = n_1 + n_2 + \dots + n_m$.
-Depending on the domain of application, these groups are sometimes referred to as studies or clusters.
-For each group, we have design matrices $X_i \in \R^{n_i \times p}$, which are called the matrices of fixed features, and $Z_i \in \R^{n_i \times q}$, called matrices of random features, and we observe vectors of outcomes $Y_i \in \R^{n_i}$.
-Typically, columns of $Z_i$ are a subset of columns $X_i$ but it does not have to be the case in general.
-Let $X = [X_1, X_2, \dots, \X_m]^T$ and $Z = [Z_1, Z_2, \dots, Z_m]^T$.
-According to \cite{Patterson1971, Pinheiro2000}, a linear mixed-effects model (LME) is defined as:
-
-\eq{
-	\label{eq:lmm_setup}
-	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i, \quad i = 1 \dots m \\
-	u_i & \sim \NN(0, \Gamma),\quad \Gamma \in \R_{++}^{q\times q} \\
-	\varepsilon_i & \sim \NN(0, \Lambda_i), \quad \Lambda_i \in \R_{++}^{n_i \times n_i}
-}
-
- where $\beta \in \R^p$ is a vector of fixed (mean) covariates and $u_i \in \R^{q}$ are the vectors of unobservable random effects, which are assumed to be distributed normally with zero mean and the unknown covariance matrix $\Gamma$.
-The observation error's variances $\Lambda_i$ are often considered to be known, or to have a parametric form, such as $\Lambda_i = \sigma I$ where $\sigma$ is unknown.
-In this work, we consider $\Lambda_i$ to be known, unless stated otherwise.
-
- If one considers $\omega_i = Z_i u_i + \varepsilon_i$ to be an unknown per-cluster noise vectors, then the model \ref{eq:lmm_setup} can also be viewed as a correlated noise model:
-
- \eq{
- \label{eq:lmm_correlated_noise_setup}
-	Y_i = X_i\beta + \omega_i,\quad \omega_i \sim \NN(0, \Omega_i(\Gamma)), \quad \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i
-}
-
-For the sake of conciseness, since $\Omega_i$ is always a function of $\Gamma$, $\Omega_i(\Gamma)$ will further be spelled as just $\Omega_i$, without the parentheses.
-In fact, $\Omega_i$ will be the only terms depending on $\Gamma$ in the majority of expressions including the mixed model's likelihood and its derivatives described below.
-
- When fitting an LME, the goal is to get both the estimation of fixed covariance $\beta$ and the estimation of variance parameters of random effects $\Gamma$.
-One of the possible approaches is integrating random effects $u_i$ out to get a marginalized posterior distribution which depends only on $\beta$ and $\Gamma$, but not on $u_i$.
-According to Bayes Theorem:
-
-\eq{
-	\label{eq:integrating_re_out_short}
-	f_{\beta, \Gamma | X = \hat{X}, Y=\hat{Y}}(\beta, \Gamma) = \frac{f_{\beta, \Gamma}(\beta, \Gamma)}{f_{X, Y}(\hat{X}, \hat{Y})}\int_{v}f_{X, Y | \beta, u = v}(\hat{X}, \hat{Y})f_{u|\Gamma}(v)dv 
-}
-
-where ($\hat{X}, \hat{Y}$) is a particular implementation of random variables $(X, Y)$; $f_{\beta, \Gamma | X=\hat{X}, Y=\hat{Y}}(\beta, \Gamma)$ is a posterior distribution of the model's parameters given data $(\hat{X}, \hat{Y})$, also known as the likelihood of the model's parameters; $f_{\beta, \Gamma}(\beta, \Gamma)$ is a joint prior distribution of the model's parameters; $f_{X, Y | \beta, u}(\hat{X}, \hat{Y})$ is likelihood of the data $(\hat{X}, \hat{Y})$ for the model (\ref{eq:lmm_setup}) given $\beta$ and $u_i$ are known; and $f_{u|\Gamma}(u)$ is a conditional distribution of random effects $u_i$ given that $\Gamma$ is known.
-
-Omitting the parameters-independent multiplier $f_{X, Y}(\hat{X}, \hat{Y})$ and taking the negative log-likelihood of (\ref{eq:integrating_re_out_short}), we obtain $\LL_{ML}$:
-
-\eq{
-	\label{eq:lmm_objective}
-	& \LL_{ML}(\beta, \Gamma)  \equiv -\log(f_{\beta, \Gamma| X=\hat{X}, Y=\hat{Y}}(\beta, \Gamma)) \sim \\ & \sim   \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma^{-1}Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \half\log{\det{\pa{Z_i \Gamma^{-1}Z_i^T + \Lambda_i}}} -\log(f(\beta, \Gamma)) = \\
-	& = \sum_{i = 1}^m \half(y_i - X_i\beta)^T\Omega_i^{-1}(y_i - X_i\beta) + \half\log{\det{\Omega_i}} -\log(f_{\beta, \Gamma}(\beta, \Gamma))
-}
-
-This expression $\LL_{ML}(\beta, \Gamma)$ is known as the negative log likelihood function of a linear mixed-effects model.
-The term $-\log(f_{\beta, \Gamma}(\beta, \Gamma))$ represents prior knowledge about $\beta$ and $\Gamma$, and is often referred to as a regularization term.
-As an example, setting $-\log(f_{\beta, \Gamma}(\beta, \Gamma)) = \frac{1}{2\sigma^2_\beta}\|\beta\|_2^2 + \frac{1}{2\sigma^2_\Gamma}\|\Gamma\|_2^2$ would represent our prior belief that $\beta \sim \NN(0, \sigma^2_\beta)$ and $\Gamma \sim \NN(0, \sigma^2_\gamma)$.
-
-The unknown coefficients $\beta$ and $\Gamma$ are to be obtained via solving the following optimization problem:
-
-\eq{
-	\label{eq:ml_lme_optimization_setup}
-	\min_{\beta, \Gamma} & \ \LL_{ML}(\beta, \Gamma) \\
-	s.t. & \ \Gamma \in \R_{++}^{s\times s}
-}
-
-In a simpler case where $\Gamma = \diag(\gamma)$ is diagonal (often referred to as \textit{the diagonal setup}), the positive definite constraint from (\ref{eq:ml_lme_optimization_setup}) transforms into a box constraint:
-
-\eq{
-	\label{eq:ml_lme_diagonal_setup}
-	\min_{\beta, \gamma} & \ \LL_{ML}(\beta, \diag(\gamma)) \\
-	s.t. & \ \gamma > 0
-	}
-
-\begin{tip}
-\paragraph{Use of $\Gamma$ versus $\gamma$.} Starting from Chapter \ref{ch:relax_and_split_thesis_work} this work is focused on the case where the matrix $\Gamma$ is diagonal: $\Gamma = \diag(\gamma)$ (the diagonal setup (\ref{eq:ml_lme_diagonal_setup})), so the unknown covariance parameters of the problem are often referred to as just $\gamma$.
-This simpler case corresponds to the situation when random effects $u_{ij}$ corresponding to different mean effects $\beta_j$ are independent of each other.
-In this case the positive definiteness constraint is guaranteed to satisfy whenever $\gamma > 0$.
-Authors made the decision to start with the diagonal case because it covers their immediate needs in applications, and they see the extensions of the algorithm proposed in Chapter \ref{ch:proposed_algorithm} to the non-diagonal case to be a part of future work.
-This chapter, however, discusses both cases of $\gamma$ and $\Gamma$ to make connections between the general and the particular cases, and to preserve the notation consistency with the referenced works.
-\end{tip}
-
-%Let's consider a simpler case when $\Gamma$ is set to be a diagonal matrix $\diag(\gamma)$, which  The case of $\gamma_j = 0$ for some $j$ is related to $u_i$ being drawn from a degenerate distribution with the variances for $j$'th component to be zero.
-It implies that random, group dependent, part for $j$'th covariate is zero, and so the variability of the respective coefficient over groups does not help the model to explain the variance in the target variable.
-Since this particular case of some $\gamma_j$ being zero is directly related to selection of random features, we relax the constraint of $\Gamma$ to be non-negative definite instead of positive-definite.
-It also does not cause any singularities in the loss function since the matrices $\Omega_i$ are invertible and positive-definite for as long as $\Lambda_i$ are positive-definite and $\Gamma$ is non-negative definite.
-
-The expected value of the posterior mean effects $\beta$ given $\Gamma$ has a closed form representation which immediately follows from (\ref{eq:lmm_correlated_noise_setup}):
-
-\eq{
-	\label{eq:beta_formula}
-	\beta^{(k+1)} = \argmin_{\beta}\LL_{ML}(\beta, \Gamma^{(k)}) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i
-}
-
-There is, however, no closed-form solution for $\Gamma$ in general case, so it has to be obtained via an optimization subroutine.
-There are three aspects that make this problem challenging from the optimization point of view:
-
-\begin{enumerate}
-	\item \textbf{Non-convexity}: the loss function $\LL_{ML}$ is not convex with respect to $\Gamma$, which can prevent an optimization method from obtaining a global solution.
-	\item \textbf{Overfitting}: the matrix $\Gamma$ contains as many as $\frac{q(q-1)}{2}$ free parameters, overfitting of which may result in near-singularities, especially when the sample size is small \cite{Lange1989}.
-The diagonal setup $\Gamma = \diag(\gamma)$ is still prone to overfitting, although has less freedom ($q$ free parameters) to do so.
-	\item \textbf{Positive Constraint}: the requirement of keeping the matrix $\Gamma$ to be positive definite limits the scope of methods which can be used for optimization.
-The diagonal setup has a box constraint $\gamma > 0$ which also has to be accommodated by the optimization subroutine.
-\end{enumerate}
-
-When $\beta^*$ and $\Gamma^*$, the minimizers for (\ref{eq:ml_lme_optimization_setup}), are obtained, the individual random effects $u_i$ can be found through the minimization of conditional likelihood of random effects:
-
-\eq{
-\min_{u_i}\LL(u_i|\beta, \gamma) = \min_{u_i}\left(\half u_i^T\Gamma^{-1}u_i + \half (Z_iu_i - Y_i + F_i(\beta))^T\Lambda_i^{-1}(Z_iu_i - Y_i + F_i(\beta))\right),
-}
-
-This problem has a closed form solution which was first derived by \cite{Harville1976} as an extension to Gauss-Markov theorem:
-
- \eq{
-	\label{eq:random_effects_formula}
-	u_i = (\Gamma^{-1} + Z_i^T\Lambda_i^{-1}Z_i)^{-1}Z_i^T\Lambda_i^{-1}(Y_i - X_i\beta).	
-}
-
-As pointed by \cite{Harville1974}, $\Gamma^*$ -- the minimizer for the maximal likelihood formulation \ref{eq:ml_lme_optimization_setup} -- is known to be biased downwards because it does not take into account the loss of $k$ degrees of freedom from estimating $\beta^*$ within the same inference procedure.
-To overcome this effect they proposed a Restricted Maximal Likelihood (REML) formulation where the loss function is based on $n-k$ linearly independent error contrasts:
-
-\eq{
-	\label{eq:reml_lmm_objective}
-	\LL_{REML}(\beta, \Gamma) = -\half\log\det\left(\sum_{i=1}^m X_i^T(Z_i\Gamma^{-1}Z_i^T + \Lambda_i)^{-1}X_i\right) + \LL_{ML}(\beta, \Gamma)
-}
-
-Although different from ML, the REML formulation does not introduce more complexity to optimization: all the solution methods for ML formulation discussed below also have an adaptation to REML.
-
-Two main optimization approaches to solving mixed effects likelihood formulation are used: EM algorithm \cite{Dempster1977} and Newton-Raphson \cite{Lindstrom1988}.
-
-\subsubsection{EM algorithm for Mixed Models}
-One more way of thinking about mixed-effects models is considering \textbf{$\{Y_i\}_{i=1}^m$} to be incomplete data and \textbf{$\{Y_i, u_i\}_{i=1}^m$} to be complete data, where $u_i$ is the unobservable part.
-In this setting, the problem of fitting a mixed model can also be classified as a statistical inference problem in the presence of missing data, and so the methods for solving this kind of setup might be applicable.
-One example is EM-algorithm -- a general approach to estimating parameters in problems with missing data developed by \cite{Dempster1977}.
-As later discussed by \cite{Laird1982}, one can view EM-algorithm as optimizing the approximating function $Q(\beta, \Gamma | \hat{\beta}, \hat{\Gamma})$ -- the likelihood of $(\beta, \Gamma)$ given that the missing data $u_i$ comes from a normal distribution $\NN(\hat{\beta}, Z_i\hat{\Gamma} Z_i^T + \Lambda_i)$.
-EM-algorithm, in its general form, consists of two steps: Expectation step and Minimization step:
-
-\eq{
-	\text{E-step: }\quad & Q^{(k)}(\beta, \Gamma | \hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}) := p\left(\beta, \Gamma | \hat{u_i}^{(k)} =  \expectation{u_i|\hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}}\right) \\
-	\text{M-step: }\quad & \hat\beta^{(k+1)}, \hat\Gamma^{(k)}  :=  \argmax_{\beta, \Gamma}\ Q(\beta, \Gamma^{k+1} | \hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}) 
-}
-
-On the E-step the function $Q(\beta, \Gamma | \hat{\beta}, \hat{\Gamma})$ is constructed using the expected values of missing data given the previous-iteration parameters $\hat{\beta}, \hat{\Gamma}$, and on M-step $Q$-function is minimized, and its minimizers $\hat{\beta^*}, \hat{\Gamma^*}$ are set to be next-iteration parameters.
-The process continues until convergence.
-
-As discussed in \cite{Laird1987}, one can express the EM-algorithm's iterations in the linear mixed models setup as:
-
-\eq{
-	\beta^{(k+1)} & = \argmin_{\beta}\LL(\beta, \Gamma^{(k)}) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}(\Gamma^{(k)})X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}(\Gamma^{(k)})Y_i \\
-	u_i^{(k+1)} & = ((\Gamma^{(k)})^{-1} + Z_i^T\Lambda_i^{-1}Z_i)^{-1}Z_i^T\Lambda_i^{-1}(Y_i - X_i\beta^{(k)}) \\
-	\Gamma^{(k+1)} & = \frac{1}{m} \sum_{i=1}^m u_i^{(k)} (u_i^{(k)})^T + \Gamma^{(k)}(I - Z_i\Omega_i^{-1}(\Gamma^{(k)})Z_i^T\Gamma^{(k)})	
-	}
-
-EM-algorithm has two main advantages: simplicity of iterations and the absence of constraints in optimization subroutines.
-The later is especially important: EM-algorithm guarantees that if the initial guess $\Gamma^{(0)}$ was positive definite then all $\Gamma^{(k)}$, including, under certain conditions, the final result $\Gamma^*$, will be positive definite.
-This removes the necessity to handle the constraint $\Gamma \in \R^{s\times s}_{++}$ separately, e.g.
-via re-parametrization using Cholesky decomposition, which allows to keep the iterations fast and simple.
-As an apparent consequence, EM algorithm also correctly handles the box constraint in the diagonal setup~(\ref{eq:ml_lme_diagonal_setup}).
-
-The main drawback of EM-algorithm is its slow convergence speed.
-In the form above, \cite{Dempster1977} showed that the algorithm monotonically converges, with geometric rate, as established by~\cite{Wu1983}, to a local maximal likelihood estimator.
-The geometric rate is due to the fact that EM-algorithm can be equivalently viewed as a proximal gradient descent on KL-divergence based likelihood.
-There have been wide effort on both discovering more statistical properties of the algorithm, as well as on improving its convergence rate.
-For instance, in his recent work \cite{Balakrishnan2017} provided a quantitative characterization of a region such that the EM-algorithm, initialized within that region, is guaranteed to converge to a "sufficiently good", fixed point: the one within statistical precision of the global optimum.
-There are also works which are empirically improving the convergence speed of EM in terms of computations and time via implementation-level improvements, like \cite{Liu1994, Liu1998}, to name a few.
-
-\subsubsection{Newton Methods for Mixed Models}
-\label{section:newton_methods_for_mixed_models}
-Newton method as a second-order optimization method for estimating parameters of mixed models was proposed by \cite{Jennrich1986} and later improved by \cite{Lindstrom1988}.
-The core idea of this approach is to consider $\hat\beta(\Gamma)$ as a function of $\Gamma$ (see Eq.
-\ref{eq:beta_formula}), and apply Newton method to the resulting formulation:
-
-\eq{
-	\label{eq:lmm_loss_only_of_big_gamma}
-	\min_{\Gamma}\ & \LL_{ML}(\beta(\Gamma), \Gamma) \\
-	\text{s.t.}\ & \Gamma \in \R^{s\times s}_{+}
-}
-
-In a general setup when $\Gamma$ is not diagonal, one can treat the positive-definiteness constraint by reparametrizing the problem via modified Cholesky decomposition, as suggested by \cite{Chen2003}:
-\eq{
-	\label{eq:dlld_decomposition}
-	\Gamma = DLL^TD
-}
-where $D$ is an $s\times s$ diagonal matrix, and $L$ is an $s\times s$ lower-triangular matrix with ones on the main diagonal.
-Applying this variables transformation to the equation (\ref{eq:lmm_loss_only_of_big_gamma}) we get an unconstrained optimization problem on diagonal elements of $D$ and on the off-diagonal elements of $L$:
-
-\eq{
-	\min_{D, L}\ & \LL_{ML}(\hat{\beta}(DLL^TD), DLL^TD) \\ 
-	\text{s.t.}\ &  D_{ij} = 0, \ i \neq j \\ 
-	& \ L_{ii} = 1,  \  i = 1 \dots s
-}
-
-In a simpler case of $\Gamma$ being diagonal ($\Gamma = \diag{\gamma}$), the setup (\ref{eq:lmm_loss_only_of_big_gamma}) simplifies to
-
-\eq{
-	\label{eq:lmm_loss_only_of_gamma}
-	\min_{\gamma > 0} \LL(\hat{\beta}(\gamma), \gamma)
-}
-On each iteration, Newton-like family of methods is looking for $\gamma^{(k+1)}$ -- a minimizer for a quadratic approximation of the problem at the point $\gamma^{(k)}$:
-\eq{
-	\gamma^{(k+1)} = \argmin_{\Delta\gamma} & \LL_{ML}(\gamma^{(k)}) + \nabla_\gamma\LL_{ML}(\gamma^{(k)})^T \Delta\gamma + \Delta\gamma^T\nabla^2_{\gamma\gamma}\LL_{ML}(\gamma^{(k)})\Delta\gamma\\
-	\text{ s.t. } & \gamma^{(k)} + \Delta\gamma \geq 0
-}
-
-The positivity constraint narrows the scope of optimization techniques which can be used.
-There is, however, still wide variety of methods to choose from like Truncated Newton Methods with Barriers \cite{Nash2000}, Projection Methods \cite{Bertsekas1982}, or Interior Point methods \cite{Potra2000}, one can refer to \cite{TrustRegionBook} for a detailed overview of these algorithms.
-The Interior Point method shall be discussed in more details since it's used in Chapter \ref{ch:proposed_algorithm} of this work.
-
-Interior Point (IP) methods \cite{Kojima1991, Nesterov1994, Wright1997} target the Karush-Kuhn-Tucker optimality conditions for the problem (\ref{eq:lmm_loss_only_of_gamma}) directly by introducing ``margins'' -- non-negative axillary variables which transform inequality constraints into equality constraints:
-
-\eq{
-	\label{eq:lmm_loss_only_of_gamma}
-	\min_{\gamma, d} & \LL(\gamma) \\
-	\text{ s.t. } & \gamma - d = 0 \\
-	& d \geq 0
-}
-
-The positivity constraint for $d$ is then handled via a logarithmic barrier. The Lagrangian and the KKT system for this problem are:
-
-\eq{
-	\label{eq:kkt_conditions}
-	F_\mu(\gamma, d, v) & = \LL(\gamma) + v^T(\gamma - d) - \mu\sum_{i=1}^s\log(d_i) \\
-	\nabla F & = \begin{bmatrix}
-		\nabla_{\gamma} F_\mu \\
-		\nabla_{v} F_\mu \\
-		\nabla_{d} F_\mu 
-	\end{bmatrix} = \begin{bmatrix}
-		\nabla \LL(\gamma) + v \\
-		\gamma - d \\
-		v \circ d - \mu\circ 1
-	\end{bmatrix} = 0
-}
-
-where $v$ is a vector of dual variables, and $\mu$ is a positive hyper parameter which define the steepness of the logarithmic barrier, and $v\circ d$ denotes the element-wise product.
-Application of Newton root-finding method to this KKT system gives us the Interior Point method.
-In particular, the search direction $[\Delta\gamma, \Delta v, \Delta d]$ as the solution of the following system:
-
-\eq{
-	\label{eq:ip_system_as_equation}
-	\nabla^2F_\mu\begin{bmatrix}
-		\Delta \gamma \\
-		\Delta v \\
-		\Delta d
-	\end{bmatrix} = -\nabla F_\mu\begin{bmatrix}
-		\Delta \gamma \\
-		\Delta v \\
-		\Delta d
-	\end{bmatrix}
-}
-
-The iteration then looks like
-\eq{
-	\gamma^{(k+1)} & = \gamma^{(k)} + \alpha \Delta\gamma \\
-	v^{(k+1)} & = v^{(k)} + \alpha \Delta v \\
-	d^{(k+1)} & = d^{(k)} + \alpha \Delta d
-}
-
-$\alpha$ is chosen via line search so all $\gamma^{(k+1)}$, $v^{(k+1)}$, and $d^{(k+1)}$ are positive, and a sufficient descent condition is satisfied:
-\eq{
-	\|\nabla F_\mu(\gamma^{(k+1)}, v^{(k+1)}, d^{(k+1)})\| \leq 0.99\|\nabla F_\mu(\gamma^{(k)}, v^{(k)}, d^{(k)})\| 
-}
-
-Since this is a root finding procedure for the equation (\ref{eq:kkt_conditions}), the natural stopping criterion is 
-\eq{
-\|\nabla F_\mu(\gamma^{(k)}, v^{(k)}, d^{(k)})\| \leq \varepsilon
-}
-for some sufficiently small $\varepsilon$.
-
-Interior Point method is a second-order method, so, under certain assumptions, it has a quadratic rate convergence region around the optimal point.
-This leads to better performance in terms of the number of iterations comparing to EM-algorithm.
-The performance in terms of time, however, strongly depends on the problem's size.
-It's due to the fact that working with Hessian information requires solving a linear system, which is not needed for first order methods like EM-algorithm.
-
-\subsection{Feature Selection for Mixed Models}
-\label{ch:feature_selection_for_mixed_models}
-In statistical learning setting, feature selection is defined as a process of selecting, or ranking, the most important predictors in a dataset.
-The goal is to get an interpretable model without sacrificing prediction quality.
-Selection process often happens simultaneously with fitting the model, where a predictor is considered to be selected if its respective coefficient in the model is non-zero.
-If the desired number of coefficients $k$ is given, then  the feature selection problem can be formulated as minimization of a loss function $f(\theta)$, which is  often proportionate to the negative log-likelihood $\log\, p(\theta| X, Y)$, with respect to its parameters $\theta$ subject to a zero-norm constraint:
-
-\eq{
-	\label{eq:general_feature_selection_zero_norm}
-	&\min_{\theta}  f(\theta) \\
-	& \text{ s.t. } \|\theta\|_0 \leq k 
-	}
-
- In this form the problem is equivalent to the Knapsack problem, which is NP-complete.
-Many techniques based on exhaustive search, also known as subset selection process, were developed, and showed to be effective when the total number of predictors is small (see \cite{Muller2013}).
-The same setup appears to be intractable in a high-dimension setting due to the exponential growth of the number of subsets to check when $n \to \infty$, however, it's possible to get an approximate solution via relaxation techniques discussed below.
-
- One way to work with zero-norm constraint in the Eq.\ref{eq:general_feature_selection_zero_norm} is to relax it to a first norm constraint.
-An example of such approach is LASSO, or least absolute square shrinkage operator, studied by \cite{Tibshirani1996} for the least squares model:
-
- \eq{
- 	\label{eq:lasso_regression}
- 	\min_\beta ||y - X\theta||_2^2 + \lambda||\theta||_1
- }
-
-  It combines the small bias of the least square estimator and the interpretability of the subset selection method, producing, however, biased estimates for the large coefficients, according to \cite{Zou2006}.
-This approach was later extended to exhibit other desirable properties, such as lesser bias for larger coefficients (SCAD from \cite{Fan2001}) or simultaneous selection for highly correlated predictors (Elastic Net from \cite{Zou2005}), and to accommodate other model setups, including classification ($\ell_1$-SVM from \cite{Bradley1998, Shalev-Shwartz2011, Belyy2018}), clustering (\cite{Jajuga1991}), and reinforcement learning (\cite{Kim2019}), to name a few.
-
-  Feature selection for linear mixed models is generally a more challenging problem than feature selection for linear regression problem.
-First, in linear regression setting the observation are independent, whereas in mixed-effects setup they're typically dependent.
-Second, one needs to do simultaneous selection of mean effect variables ($\beta$) and the ones related to their random effects' covariance structure ($\Gamma$), and these groups can have different relative importance.
-Finally, selecting random features implies that $\Gamma$ will have zero columns and rows, which prevents it being positive-definite.
-Consequently, selecting variance parameters leads to boundary issues which make both theoretical analysis and numerical computations more challenging.
-
-   The shrinkage operator approach was first adapted to the mixed-effect models setting by \cite{Lan2006} to perform fixed effects selection.
-Selection of random effects is generally more difficult than selection of fixed effects alone because removing random effect from the model requires elimination of the entire respective row and column of $\Gamma$.
-This extension was provided by \cite{Krishna2008}, who  performed simultaneous fixed and random effects selection by combining three components: the modified Cholesky decomposition $\Gamma = DLL^TD$, as suggested by \cite{Chen2003}, the Adaptive LASSO penalty, as in \cite{Lan2006}, and EM algorithm (\cite{Dempster1977}).
-Their penalized Q-function for EM-algorithm is defined as:
-  \eq{
-  	\label{eq:krishna_lasso}
-  	Q(\beta, D,L|\beta^{(k)}, D^{(k)}, L^{(k)}) = & \mathbb{E}_{u_i | \beta^{(k)}, D^{(k)}, L^{(k)}}\left[{\sum_{i=1}^m \|Y_i - X_i\beta - Z_iDLu_i\|^2}\right] + \\ + & \lambda\left(\sum_{j=1}^k\frac{|\beta_j|}{|\hat\beta_j|} + \sum_{j = 1}^s\frac{|D_{ii}|}{|\bar{D}_{ii}|}\right)
-  }
-   They suggest tuning the regularization parameter $\lambda$ using a BIC criterion.
-A similar formulation was used by \cite{Lin2013} who proposed a joint selection with two-stage ALASSO.
-On stage 1 they approximate the shrinkage penalty with local quadratic approximation and use Newton method to select random effects, and then, on stage 2 they select fixed effects via penalized log-likelihood where the covariance variables are fixed.
-
-   Another type of regularization which is used for feature selection is SCAD \cite{Fan2001} which acts on the first derivative of the penalty function $p_\lambda(\theta)$:
-   \eq{
-   		p'_{\lambda}(|\theta|) = \lambda\left\{\ind{\theta \leq \lambda} +  \frac{(a\lambda - \theta)_+}{(a-1)\lambda}\ind{\theta - \lambda} \right\}
-   }
-   The adaptation of SCAD penalty to selecting both fixed and random features in linear mixed model setup was developed by \cite{Fan2012}.
-
-  In high-dimensional setting marginal likelihood is often replaced with a hierarchical approximation of it known as h-likelihood (\cite{Lee1996}), which, for a given data $(\hat{X}, \hat{Y})$ is defined as
-
-  \eq{
-  \label{eq:h_likelihood}
-  \LL_{H}(\beta, \Gamma, u) = -\log f_{X, Y | \beta, \Gamma, u}(\hat{X}, \hat{Y}) - \log f_{u|\Gamma}(u, \Gamma)
-  }
-
-  Where $f_{X, Y | \beta, \Gamma, u}(\hat{X}, \hat{Y})$ is a likelihood of data given $\beta, \Gamma,$ and $u$, and $f_{u|\Gamma}(u, \Gamma)$ is a conditional density of random effects $u$ given $\Gamma$.
-This formulation allows to avoid integration over random effects which is required when the classical marginal likelihood is used.
-Practically, the use of h-Likelihood eliminates the need to compute and to invert the covariance matrices $\Omega_i$ or their Cholesky factors, which is often the most expensive operation when computing the marginal likelihood numerically.
-When derived using h-Likelihood, the estimators of fixed effects $\beta$ and random effects $u_i$ are, as shown by \cite{Lee1996}, asymptotically best unbiased predictors.
-In order to estimate variance components simultaneously with fixed effects an Adjusted Profile likelihood, which is an hierarchical generalization of REML, is used.
-
-  Similarly to marginal likelihood above, various feature selection approaches were adapted to be used with h-Likelihood.
-Similarly to what \cite{Krishna2008} proposed for marginal likelihood based selection, \cite{Xu2015} proposed to penalize the fixed effects and random effects variance components $\Omega_i$ in h-likelihood with LASSO-like family of penalties.
-They also showed, that, under certain assumptions, the proposed estimator possesses asymptotic normality property.
-The most recent development in this trend to date was provided by \cite{Xie2020} who utilized multi-armed bandits approach for selecting mean effects in high-dimensional, h-Likelihood, setting.
-
-  Shrinkage penalties and norm-based penalties are not the only kind of relaxations which are used in feature selection.
-Recently \cite{Ghosh2018} studied SCAD regularization for selecting mean effects in a high-dimensional setting motivated by genomics problems.
-Other approaches to feature selection are based on information criteria such as AICs and BICs \cite{Jones2011}.
-According to \cite{Muller2013}, the most widely used AIC criterion is what \cite{Vaida2005} call the marginal AIC criterion:
-  \eq{
-  \label{eq:vaida_aic}
-  AIC = 2\LL(\hat{\theta}) - 2\alpha_n(p+q)
-  }
-  where $\hat\theta$ includes all the estimated parameters in a mixed-effect model $(\beta, \Gamma)$, and $\alpha_n = n(n-p-q-1)$ for a finite sample case \cite{Sugiura1978}.
-As discussed by \cite{Fang2011}, AIC is asymptotically equivalent to leave-one-out cross-validation, so it can be used for choosing between a finite number of models.
-
-  One can refer to \cite{Muller2013} for a detailed overview of different feature selection approaches.
-
-  Variable projection is another approach to tackle the zero-norm constraint in (\ref{eq:general_feature_selection_zero_norm}).
-Relax-and-Split \cite{Zheng2018RelaxAndSplit} technique is one of implementations of this approach.
-It's been successfully applied in basis pursuit denoise problems \cite{Baraldi2019}, compressed sensing \cite{Zheng2019}, and sparse models identification \cite{Champion2020}.
-The key idea of this method is to use a set of auxiliary variables to relax the constraints:
-\eq{
-	\min_{\|\beta\|_0 \leq k} f(\beta) \quad \to \quad \min_{\beta;\ \|w\|_0 \leq k} f(\beta) + \frac{1}{2\nu}\|w - \beta\|_2^2
-}
-and then to apply an alternating minimization algorithm to solve the relaxation. This approach will be discussed in details in the next chapter.
-
-\newpage
-\section{Relax-and-Split for Mixed-Effects Selection}
-\label{ch:relax_and_split_thesis_work}
-\subsection{Introduction}
-Recently \cite{Zheng2018RelaxAndSplit} proposed to use variable projection technique called Relax-and-Split for nonconvex-composite optimization problems of the form:
-
-\eq{
-	\min_\beta f(\beta) = h(A\beta) + g(\beta)
-}
-where $\beta \in \R^k$ are the decision variables, $A\in \R^{n \times k}$, $h(\beta)$ is non-smooth, non-convex, and separable, and $g(\beta)$ is convex.
-The key innovation of their work is to relax the original variable $\beta$ with a set of auxiliary variables $\tbeta$, and then use partial minimization techniques to develop effective algorithms:
-\eq{
-	\min_x f(\beta) = h(A\beta) + g(\beta)\quad \to \quad \min_{\beta, \tbeta} h(\tbeta) + \frac{1}{2\theta}\|A\beta - \tbeta\|_2^2 + g(\beta)
-}
-The same approach can be used for relaxing constraints in non-convex optimization problems. This approach was discussed by \cite{Zheng2019} as a part of SR3 framework: 
-\eq{
-	\min_{\beta \in \Delta} f(\beta) \quad \to \quad \min_{\tbeta\in \Delta;\ \beta} f(\beta) + \frac{1}{2\nu}\|\beta - \tbeta\|_2^2
-}
-
-As an example, the zero-norm constraint from (\ref{eq:general_feature_selection_zero_norm}) can be relaxed as follows:
-
-\eq{
-	\label{eq:relax_and_split_in_general_form_2}
-	\min_\beta f(\beta)\ \text{ s.t. }\|\beta\|_0 \leq k \quad \to \quad \min_{\beta, \tbeta} \ &  f(\beta) + \lambda\|\beta - \tbeta\|_2^2 \ \text{ s.t. } \|\tbeta\|_0 \leq k
-}
-The second norm links $\beta$, an unconstrained variable, and its constrained counterpart $\tbeta$.
-The coefficient $\lambda$ represents the strength of this link and can be viewed as a Gaussian prior for the distance of $\beta$ from the "wall" -- a $k$-dimensional subspace.
-The larger $\lambda$ is -- the higher penalty is for $\beta$ violating the zero-norm constraint.
-
-This relaxation admits effective application of alternating minimization methods:
-\eq{
-	\label{eq:relax_and_split_general_iteration}
-	\beta^{(t+1)} = & \argmin_{\beta}\left[ f(\beta) + \lambda\|\beta - \tbeta^{(t)}\|^2_2\right] \\
-	\tbeta^{(t+1)} = & \argmin_{\tbeta} \|\beta^{(t)} - \tbeta\|_2^2 \\
-	& \text{ s.t. } \|\tbeta\|_0 \leq k
-}
-The first optimization subroutine is an unconstrained optimization problem.
-The second subroutine is a Euclidean projection of $\beta^{(t)}$ onto a $k$-dimensional subspace.
-This operation can be efficiently implemented since it only involves taking $k$ largest in absolute value elements of $\beta^{(t)}$.
-The coefficient $\lambda$ is iteratively increased until $\|\beta - \tbeta\| \leq \varepsilon$ for some small $\varepsilon$.
-
-This relaxation has certain advantages over $\ell_1$-type relaxations:
-\begin{itemize}
-	\item \textbf{Interpretability of hyper-parameters.} Notice that this relaxation does not change the meaning of the only hyper-parameter $k$.
-In both formulations in Eq.
-(\ref{eq:relax_and_split_in_general_form_2}) the parameter $k$ denotes the number of non-zero coefficients in the resulting estimator $\beta^*$ whereas in the formulation (\ref{eq:lasso_regression}) $k$ is replaced with $\lambda$, which denotes the variance of Laplacian prior for $\beta^*$.
-There is no data-independent mapping between $\lambda$ and $k$ which would allow researcher to make an informed decision on the value of $\lambda$ before calling the LASSO fit, which leads to multiple trial-launches resulting in a overall production pipeline slowdown.
-	\item \textbf{No model bias given the subspace.} Notice that in (\ref{eq:relax_and_split_in_general_form_2}) the local minimizer for the relaxation when $\lambda \to \infty$ will also be a local minimizer for a non-relaxed setup (\ref{eq:general_feature_selection_zero_norm}).
-The solution for the formulation (\ref{eq:lasso_regression}) is, in a general case, not going to be a local minimizer for (\ref{eq:general_feature_selection_zero_norm}) since its non-zero coefficients are biased by the implicit Laplacian prior (\cite{Zou2006} provides a comprehensive discussion on the nature of this bias).
-Because of that, feature selection via LASSO is normally executed as a two-stage process, when, after the optimal subset of features is being selected on the stage one, the optimization is repeated on that features only without $\ell_1$ regularization to partially eliminate the model bias.
-There is, however, no need for the second stage when the formulation (\ref{eq:relax_and_split_in_general_form_2}) is used.
-\end{itemize}
-
-The main disadvantage of the formulation (\ref{eq:relax_and_split_in_general_form_2}) comparing to LASSO (\ref{eq:lasso_regression}) is that the latter only needs to be solved once, whereas the former should be solved iteratively for increasing $\lambda$ until the relaxation is eliminated, which naturally takes longer.
-This, however, can be mitigated by applying warm-start technique, when the optimization after increasing $\lambda$ starts from the previous optimal $\beta^*$.
-This approach is especially efficient when the second order optimization subroutine is used for the step one of the algorithm (\ref{eq:relax_and_split_general_iteration}): the subsequent problem differs from the previous one by a quadratic term, so if the method previously reached the quadratic convergence region then increasing $\lambda$ won't change it.
-
-% This relaxed problem also admits effective partial minimization schemes.
-Partial minimization over the original variables $x$ is now unconstrained optimization problem, and partial minimization over $w$ is a Euclidean projection problem onto $\Delta$.
-The later can be as complex as the set $\Delta$ is, but for many real-world applications the projections are accessible in analytic form.
-The main advantage of having a relaxation is that the complexity of projecting on $\Delta$ is now handled separately from the complexity of optimizing the function $f(x)$.
-This comes with a price of solving the relaxation iteratively with $\nu \to 0$ until $w = x$ and $x\in \Delta$, so the original constraint is satisfied.
-
-\subsection{Proposed Algorithm}
-\label{ch:proposed_algorithm}
-
-Consider a mixed-effect model setting described in Eq. (\ref{eq:lmm_setup}) with $\Gamma$ being diagonal: $\Gamma = \diag(\gamma)$.  We wish to find a minimizer the following loss function:
-
-\eq{
-	\label{eq:fs_loss_intractable}
-	\min_{\beta, \gamma} & \LL_{ML}(\beta, \diag{(\gamma)}) \\
-	\text{s.t. } 	& \|\beta\|_0 \leq k \\
-					& \|\gamma\|_0 \leq j \\
-					& \gamma \geq 0
-}
-
-where $\LL_{ML}(\beta, \diag{(\gamma)})$ is a marginalized log-likelihood of a mixed model defined in Equation (\ref{eq:lmm_objective}).
-
-In this form, the problem can have as many as $C^n_k$ local minimizers (at least one per $k$-subspace), and finding a global solution is equivalent in complexity to solving a weighted knapsack problem.
-However, finding a "sufficiently good" local minimizer is usually enough for practical purposes.
-
-We relax the original problem using Relax-and-Split approach (\cite{Zheng2018RelaxAndSplit}) in order to both automate choosing a good subspace and to do optimization in the whole space instead of doing optimization in subspaces.
-Namely, we introduce two constrained auxiliary variables $\tbeta$ and $\tgamma$ and treat them as projections of now unconstrained variables $\beta$ and $\gamma$ on $k$- and $j$-subspaces respectively.
-The loss function $\LL_r(\beta, \gamma, \tbeta, \tgamma)$ of for the relaxed setup is defined as:
-\eq{
-	\LL_r(\beta, \gamma, \tbeta, \tgamma) \equiv \LL_{ML}(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta\|^2_2 + \frac{\lambda_{\gamma}}{2}\|\gamma - \tgamma\|_2^2
-}
-
-where $\LL_{ML}(\beta, \diag{(\gamma)})$ is defined in (\ref{eq:lmm_objective}):
-\eq{
-	\label{eq:lmm_objective_2}
-	\LL_{ML}(\beta, \gamma) = \sum_{i = 1}^m \half(Y_i - X_i\beta)^T\Omega_i^{-1}(\gamma)(Y_i - X_i\beta) + \half\log{\det{\Omega_i(\gamma)}}
-}
-and $\Omega_i(\gamma)$ is a covariance matrix for the $i$'th group observation's noise:
-\eq{
-	\Omega_i(\gamma) = Z_i\diag(\gamma)Z_i^T + \Lambda_i
-}
-
-In order to solve a relaxed setup for given relaxation parameters $\lambda_\beta$ and $\lambda_\gamma$ we need to find a solution $(\beta^*, \gamma^*, \tbeta^*, \tgamma^*)$ for the following problem:
-
-\eq{
-	\label{eq:lmm_relax_and_split_full_setup}
-	\min_{\beta, \gamma, \tbeta, \tgamma}& \LL_r(\beta, \gamma, \tbeta, \tgamma)\\
-	\text{s.t. } 	& \|\tbeta\|_0 \leq k \\
-					& \|\tgamma\|_0 \leq j \\
-					& \gamma \geq 0 \\
-					& \tgamma \geq 0
-}
-
-We apply variable projection technique to solve the relaxed problem (\ref{eq:lmm_relax_and_split_full_setup}).
-This approach outlines the Algorithm \ref{alg:feature_selection_relax_and_split_outline}, henceforth referred as \ouralgo,  as an iterative procedure with four projection steps within each iteration.
-In order to guarantee that $(\tbeta, \tgamma)$ from (\ref{eq:lmm_relax_and_split_full_setup}) is a solution for (\ref{eq:fs_loss_intractable}), we iteratively increase $\lambda_\beta$ and $\lambda_\gamma$ until $\beta=\tbeta$ and $\gamma=\tgamma$.
-
-\begin{algorithm}
-	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
-	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
-	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
-	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
-	\Repeat{converges}{
-		$\tbeta^{(k+1)} \leftarrow \proj_{\|\beta\|_0 \leq k}(\beta^{(k)})$ \tcc*[r]{\label{alg:feature_selection_relax_and_split_outline:tbeta_line}Take max k from $\beta$, rest to 0}
-		$\tgamma^{(k+1)} \leftarrow \proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)})$ \tcc*[r]{\label{alg:feature_selection_relax_and_split_outline:tgamma_line}Take max j from $\gamma$, rest to 0}
-		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$ \tcc*[r]{\label{alg:feature_selection_relax_and_split_outline:beta_gamma_line} IP Method}
-	}
-	}
-	\BlankLine
-\caption{\label{alg:feature_selection_relax_and_split_outline}\ouralgo: Relax-and-Split-based feature selection algorithm for linear mixed-effects models.}	
-\end{algorithm}
-
-The algorithm has three alternating minimization steps in its inner cycle. We shall discuss all three steps in details below.
-
-\paragraph{Algorithm \ref{alg:feature_selection_relax_and_split_outline}, lines \ref{alg:feature_selection_relax_and_split_outline:tbeta_line} and \ref{alg:feature_selection_relax_and_split_outline:tgamma_line}: Projections on a $\ell_0$ box.} When $\beta$ and $\gamma$ are fixed the partial minimization solution for $\tbeta$ and $\tgamma$ is just projections of $\beta$ and $\gamma$ on respective $\ell_0$-boxes:
-
-\eq{
-	\tbeta^{(k+1)} = \argmin_{\tbeta} \|\tbeta - \beta^{(k)}\|_2^2 \text{ s.t.
-} \|\tbeta\|_0 \leq k \quad \implies & \quad  \tbeta^{(k+1)} = \text{ max $k$ components from } |\beta_1|, \dots, |\beta_p| \\
-	\tgamma^{(k+1)} = \argmin_{\tgamma} \|\tgamma - \gamma^{(k)}\|_2^2 \text{ s.t.
-} \|\tgamma\|_0 \leq j \quad \implies & \quad  \tgamma^{(k+1)} = \text{ max $j$ components from } |\gamma_1|, \dots, |\gamma_q| \\
-	& \text{... and the rest of components set to 0}
-}
-
-\paragraph{Algorithm \ref{alg:feature_selection_relax_and_split_outline}, line \ref{alg:feature_selection_relax_and_split_outline:beta_gamma_line}: Interior Point Method.}
-
-As described in Section \ref{section:newton_methods_for_mixed_models}, Interior Point (IP) Methods are a class of second-order methods which come from barrier relaxation of constraints followed by direct application of Newton method to KKT optimality conditions.
-Following the same idea, we transform the inequality constraints to equality constraints and then relax these constraints via logarithmic barriers:
-
-\eq{
-	\label{eq:log_barrier_relaxation_for_ip}
-	\min_{\beta, \gamma, d} & \LL(\beta, \gamma) +  \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2 + \mu\sum_{i=1}^{q}\log(d_i) \\
-	\text{ s.t. } & \gamma - d = 0
-}
-
-where $d \in R^{q}$ are ``offset''  variables, and $\mu$ is a relaxation parameter: the smaller $\mu$ corresponds to ``sharper'' barrier approximation.
-Lagrangian for this optimization problem can be written as:
-
-\eq{
-	F_\mu(v, d, \beta, \gamma) & = \LL(\beta, \gamma) +  \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2 + \mu\sum_{i=1}^{q}\log(d_i) + v^T(\gamma - d) \\
-}
-
-From the necessary optimality conditions on $d$ we have that 
-\eq{
-	\nabla_d F_\mu = \begin{bmatrix}
-		v_1 - \mu/d_1 \\
-		\dots \\
-		v_q - \mu/d_q 
-	\end{bmatrix} = \begin{bmatrix}
-		0 \\
-		\dots \\
-		0 
-	\end{bmatrix} \implies v_id_i = \mu \text{ for all } i
-}
-which can be written in a matrix form as 
-\eq{
-	\label{eq:complementary_slackness_kkt}
-	\diag(v)d - \mu\circ 1 = 0
-}
-where ``$\circ$'' denotes an element-wise product.
-
-Using $\gamma - d = 0$ to express $F_\mu$ as a function of $v$, $\beta$, and $\gamma$ alone, we can write down the KKT system:
-\eq{
-	\nabla F_\mu(v, \beta, \gamma) & = \begin{bmatrix}
-		\nabla_v F_\mu \\
-		\nabla_\beta F_\mu \\
-		\nabla_\gamma F_\mu 
-	\end{bmatrix} = \begin{bmatrix}
-		v\circ \gamma - \mu\circ 1 \\
-		\nabla_\beta \LL(\beta, \gamma) + \lambda_\beta(\beta - \tbeta^{(k)}) \\
-		\nabla_\gamma \LL(\beta, \gamma) + \lambda_\gamma(\gamma - \tgamma^{(k)}) - v
-	\end{bmatrix} = 0
-}
-
-Following the the approach of (\ref{eq:ip_system_as_equation}), on each iteration we choose the search direction $[\Delta v, \Delta \beta, \Delta \gamma]$ as the solution of the following system:
-
-\eq{
-	\label{eq:ip_iteration_rs_form}
-	\nabla^2F_\mu\begin{bmatrix}
-		\Delta v \\
-		\Delta \beta \\
-		\Delta \gamma
-	\end{bmatrix} = -\nabla F_\mu\begin{bmatrix}
-		\Delta v \\
-		\Delta \beta \\
-		\Delta \gamma
-	\end{bmatrix}
-}
-
-Once the search direction $[\Delta v, \Delta \beta, \Delta \gamma]$, is established the method performs a step
-\eq{
-	v^{(k+1)} & = v^{(k)} + \alpha_k \Delta v \\
-	\gamma^{(k+1)} & = \gamma^{(k)} + \alpha_k \Delta\gamma \\
-	\beta^{(k+1)} & = \beta^{(k)} + \alpha_k \Delta \beta
-}
-where the step length $\alpha_k$ is chosen to satisfy positivity and sufficient descent conditions:
-\eq{
-	\text{\textit{Positivity:}} \qquad & \gamma \geq 0,\ v \geq 0 \\
-	\text{\textit{Sufficient Descent:}}\qquad & \|\nabla F_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|\nabla F_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| 
-}
-On each iteration the relaxation parameter $\mu$ is adjusted as $\mu^{(k+1)} = \frac{{v^{(k)}}^T\gamma^{(k)}}{q}$.
-The method continues to iterate until the stopping criterion $\|\nabla F_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq \text{\texttt{tol}}$ is met.
-
-The Hessian matrix $\nabla^2 F_\mu$ in Eq. (\ref{eq:ip_iteration_rs_form}) is defined as 
-\eq{
-	\nabla^2 F_\mu = \begin{bmatrix}
- 		\diag(\gamma) & 0 & \diag(v) \\
- 		0 & \nabla^2_{\beta\beta}\LL + \lambda_\beta I & \nabla^2_{\beta\gamma} \LL\\
- 		-I & \nabla^2_{\gamma\beta}\LL & \nabla^2_{\gamma\gamma}\LL + \lambda_\gamma	 I
- 	\end{bmatrix}
-}
-
-The precise form of the derivatives $\nabla\LL$ will depend on which likelihood is used: standard maximal likelihood $\LL_{ML}$ or restricted maximum likelihood $\LL_{REML}$.
-For instance, if the former is used then the derivatives are defined as:
-\eq{
-	\label{eq:all_derivatives}
-	\nabla_\beta\LL(\beta, \gamma) & = -\sum_{i=1}^m X_i^T\Omega_i^{-1}(Y_i - X_i\beta)\\
-	\nabla_\gamma\LL(\beta, \gamma) & = \sum_{i = 1}^m  \diag{{Z_i}^T\Omega_i^{-1}Z_i} - ({Z_i}^T\Omega_i^{-T}(Y_i - X_i\beta))^{\circ 2}\\
-	\nabla^2_{\beta\beta}\LL(\beta, \gamma) & =  \half\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i \\
-	\nabla^2_{\beta\gamma}\LL(\beta, \gamma) & = -\half\sum_{i = 1}^m X_i^T\Omega_i^{-1}Z_iZ_i^T\Omega_i^{-1}(Y_i - X_i\beta) \\
-	\nabla^2_{\gamma\gamma}\LL(\beta, \gamma) & 
-	= \half\sum_{i = 1}^m -(Z_i^T\Omega_i^{-T}Z_i)^{\circ 2} + 2Z_i^T\Omega_i^{-T}(Y_i - X_i\beta)((Y_i - X_i\beta)^T\Omega^{-T}Z_i)^T\circ(Z_i^T\Omega^{-1}Z_i)
-}
-
-For the sake of completeness we provide the derivation of these formulas in Appendix \ref{appendix:derivatives_of_lmm}, although they can also be found in \cite{Lindstrom1988}.
-
-\subsubsection{Implementation Details}
-\paragraph{Hessians through Cholesky Decomposition}
-Computing the derivatives in (\ref{eq:all_derivatives}) are the most computationally intensive part of Algorithm \ref{alg:feature_selection_relax_and_split_outline}.
-Here we provide the way to efficiently compute them via Cholesky decomposition.
-
-Let's define $L_i$ to be a Cholesky factor of the matrix $\Omega_i$:
-\eq{
-	\Omega_i = L_iL_i^T
-}
-Having a Cholesky factor $L_i$ for $\Omega_i$ one can calculate a Cholesky factor of the inverse matrix $\Omega_i^{-1}$ in $\OO(max(m, n)^2)$ operations\footnote{See the implementation of LAPACK's \textit{\href{http://www.netlib.org/lapack/explore-html/da/dba/group__double_o_t_h_e_rcomputational_ga97c5ddb103f5f6bc2dc060886aaf2ffc.html}{dtrtri}} as an example.} Let's also define $\xi_i = Y_i - X_i\beta$.
-Then we can notice that all terms in formulae (\ref{eq:all_derivatives}) can be build out of three building blocks: $L^{-1}_iZ_i$, $L^{-1}_iX_i$, and $L^{-1}_i\xi_i$.
-Namely:
-
-\eq{
-	\label{eq:all_derivatives}
-	\nabla_\beta\LL(\beta, \gamma) & = -\sum_{i=1}^m (L^{-1}_iX_i)^T(L_i^{-1}\xi_i)\\
-	\nabla_\gamma\LL(\beta, \gamma) & = \sum_{i = 1}^m  \diag{((L_iZ_i)^T(L_iZ_i))} - ((L_i^{-1}Z_i)^T(L_i^{-1}\xi_i))^{\circ 2}\\
-	\nabla^2_{\beta\beta}\LL(\beta, \gamma) & =  \half\sum_{i = 1}^m (L_i^{-1}X_i)^T(L_i^{-1}X_i) \\
-	\nabla^2_{\beta\gamma}\LL(\beta, \gamma) & = -\half\sum_{i = 1}^m (L_i^{-1}X_i)^T(L_i^{-1}Z_i)(L_i^{-1}Z_i)^T(L_i\xi_i) \\
-	\nabla^2_{\gamma\gamma}\LL(\beta, \gamma) & 
-	= \half\sum_{i = 1}^m -((L_i^{-1}Z_i)^T(L_i^{-1}Z_i))^{\circ 2} + 2(L_i^{-1}Z_i)^T(L_i^{-1}\xi_i)((L_i^{-1}\xi_i)^T(L_i^{-1}Z_i))^T\circ((L_i^{-1}Z_i)^T(L_i^{-1}Z_i))
-}
-Apparently, one should calculate bigger building blocks like $(L_iZ_i)^TL_iZ_i$ only once and then reuse it for efficient performance.
-
-\paragraph{Positive Approximation of Hessian}
-Since the problem is non-convex with respect to $\gamma$ the Hessian is not in general positive definite.
-Negative definite Hessians are known to hamper the convergence of second-order methods \cite{nocedal2006numerical}.
-In order to prevent this we use a positive upper bound as an approximation of the true Hessian.
-Notice that true Hessian consists of two terms: positive definite and negative definite:
-\eq{
-\nabla^2_{\gamma\gamma}\LL(\beta, \gamma) & 
-	= \half\sum_{i = 1}^m \underbrace{-(Z_i^T\Omega_i^{-T}Z_i)^{\circ 2}}_{\text{negative definite}} + \underbrace{2Z_i^T\Omega_i^{-T}(Y_i - X_i\beta)((Y_i - X_i\beta)^T\Omega^{-T}Z_i)^T\circ(Z_i^T\Omega^{-1}Z_i)}_{\text{positive definite}}
-}
-We approximate Hessian with its positive definite part $H(\beta, \gamma)$:
-\eq{
-\nabla^2_{\gamma\gamma}\LL(\beta, \gamma) \prec H(\beta, \gamma) \equiv 2Z_i^T\Omega_i^{-T}(Y_i - X_i\beta)((Y_i - X_i\beta)^T\Omega^{-T}Z_i)^T\circ(Z_i^T\Omega^{-1}Z_i)
-}
-
-\subsection{Experiments on Synthetic Data}
-\label{ch:experiments_on_synthetic_data}
-\subsubsection{Comparison to Other Algorithms}
-\label{ch:comparison_to_other_algo}
-
-\begin{table}[h!]
-\begin{center}
-\begin{tabular}{|l|l|c|c|c|c|c|}
-\hline
-Setup & Algoritm & \% C & \% CF & \% CR & MSE & TIME \\
-\hline 
-\hline
-$n = 30$, $n_i = 5$ & \ouralgo & 58 & 72 & 78 & 0.66 & 0.015 \\
-& rPQL & 88 & 98 & 88 & 0.88 & 26-59 \\ 
-& M-ALASSO & 71 & 73 & 79 & - & -\\
-& ALASSO & 79 & 81 & 96 & - & - \\
-& SCAD-P & - & 90 & 86 & - & - \\
-\hline 
-$n = 60$, $n_i = 10$ & \ouralgo & 98 & 100  & 98 & 0.69& 0.018 \\
-& rPQL & 98 & 99 & 98 & 0.97 &  26-59 \\ 
-& M-ALASSO & 83 & 83 & 89 & - & - \\
-& ALASSO & 95 & 96 & 99 & - & -\\
-& SCAD-P & 100 & 100 & 100 & - & - \\
-\hline
-
-\end{tabular}
-\end{center}
-\caption{\label{table:krishna_setup_results} Comparison of feature selection algorithms.
-\% CF -- percent of models where true fixed effects were identified correctly, \% CR -- percent of models where true random effects were identified correctly, \% C -- both fixed and random effects were identified correctly.}
-\end{table}
-
-In the first set of synthetic tests we replicate the experimental setup from \cite{Krishna2008}, which was used to estimate the performance of M-ALASSO.
-The same setup has been used for evaluating the performance of SCAD-P \cite{Fan2012} and two-stage adaptive LASSO of \cite{Lin2013}, which allows us to compare the performance of \ouralgo relatively to the performance of its competitors.
-We also include rPQL of \cite{Hui2017} who applied shrinkage operators to PQL quasi-likelihood to facilitate feature selection in Generalized Linear Mixed Models (GLMMs) for the sake of completeness, although rPQL is designed to address a more general setup then \ouralgo.
-
-Two scenarios were considered. In each of them 200 datasets were generated from the multivariate normal distribution: 
-\eq{
-	y_i \sim \NN(X_i\beta,\ Z_i\Gamma Z_i^T + I)
-} 
-where:
-\begin{itemize}
-	\item \textbf{Scenario 1}: $n=30$, $n_i=5$, $p = 9$, $q = 4$, with true parameters $\beta = (1, 1, 0, \dots, 0)$ and the covariance matrix $\Gamma$ being: 
-		\eq{
-			\Gamma = \begin{bmatrix}
-				9 & 4.8 & 0.6 & 0 \\
-				4.8 & 4 & 1 & 0 \\
-				0.6 & 1 & 1 & 0\\
-				0 & 0 & 0 & 0
-			\end{bmatrix}
-		}
-		Columns of $X_i$ and $Z_i$ were generated as i.i.d. vectors from a uniform distribution on $(-2, 2)$, and observation errors were generated from a standard normal distribution.
-	\item \textbf{Scenario 2}: everything as in Scenario 1, but $n=60$ and $n_i=10$ to estimate performance on a larger dataset. 
-\end{itemize}
-
-The aspect which makes this setup particularly challenging for \ouralgo is that the matrix $\Gamma$ is not diagonal, so the true parameters $(\beta^*, \Gamma^*)$ are out of the model's parameters scope.
-This inevitably pulls the performance down as the model is unable to take the covariance structure of random effects into account properly, and so the respective model bias is introduced to both estimates of $\beta$ and $u_i$.
-It's still, however, possible to compare the performance of feature selection by comparing the subset of selected random covariates which is defined by non-zero elements of $\gamma$.
-
-The evaluation results are listed in Table \ref{table:krishna_setup_results}.
-Data for all algorithms except \ouralgo was borrowed from \cite{Hui2017} who aggregated it from the original works.
-From this table it's clear that the performance of \ouralgo on Setup 1 is comparable to performance of other algorithms in terms of selecting correct fixed or random effects separately, but is worse in terms of making a correct selection simultaneously.
-This is expected given that for small datasets the finite sample error adds up to the error from restricting $\gamma$ to be diagonal creating a significant gap in performance.
-This difference, however, disappears when the algorithms are compared on a bigger Setup 2, where \ouralgo performs on par with its strongest competitors.
-This behavior suggests that the diagonal $\gamma$ restriction is not as restrictive as it might seem, and our algorithm can enjoy both optimization stability and high feature selection accuracy given a sufficiently large sample size even if true random effects are correlated.
-Besides, \ouralgo demonstrated lower mean-squared prediction error and four orders of magnitude shorter execution time comparing to rPQL\footnote{Authors of other above mentioned algorithms did not report their execution times.}.
-
-\subsubsection{Scalability Experiments}
-\label{ch:scalability_experiments}
-\begin{figure}
-	\centering
-	\includegraphics[width=0.85\textwidth]{Images/scalability_mse_and_acc}
-	\caption{\label{fig:scalability_experiment} This is a scalability experiment that examines the performance of \ouralgo on synthetic data.
-We looked at how algorithm's performance varies when the number of observations is fixed and the number of features changes.
-In each synthetic dataset a random half of covariates were assigned to be significant and were used to generate observations, the rest was not used.
-The top figure shows the accuracy of selecting true important covariates depending on the number of covariates available.
-The bottom figure shows mean squared error of the model's prediction on train and test data.
-Each synthetic experiment was repeated 300 times, the shaded areas indicate 95\% confidence intervals of the empirical distribution of outcomes of those experiments.
-Depending on the relation between the number of observations per group ($n_i$), total number of observations ($n = \sum_i n_i$), and the number of features $p$, we see three distinctive "performance zones".
-In Zone 1, which corresponds to ``tall'' datasets ($n_i \geq p$ for all $i$), algorithm identifies important covariates correctly in 100\% cases.
-In Zone 2 (``square'' datasets: $n_i < p$ but $n > p$) the performance falls around $p = n/2$ to the performance of random guessing, keeping this poor performance in Zone 3 (``wide'' datasets: $p > n$).
-Given that the train error is significantly lower than the test error in Zones 2 and 3, we conclude that the resulting models overfit the data when number of covariates is bigger than the number of observations.}
-\end{figure}
-
-In the second set of synthetic experiments we want to examine how the feature selection quality of \ouralgo changes when the number of features grows and the number of observations stays fixed.
-
-The setting is similar to \textbf{Scenario 2} from Chapter \ref{ch:comparison_to_other_algo}: $n = 60$, $n_i = 10$, $q$ was set equal to $p$, and also all $X_i$ were equal to $Z_i$ for all $i$.
-The number of features $p$ was iteratively increasing by blocks of three, where the blocks of new columns $X_i$ and $Z_i$ were sampled i.i.d.
-from a multivariate normal distribution $\NN(0, \Psi)$ with the variance-covariance matrix $\Psi$:
-	\eq{
-		\Psi = \begin{bmatrix}
-			9 & 4.8 & 0.6 \\
-			4.8 & 4 & 1 \\
-			0.6 & 1 & 1 \\
-		\end{bmatrix}
-	}
-
-	For each experiment, random half of all features were active fixed features with 70\% of those active fixed features also having a random effect.
-To be precise, $\beta$ had a random 50\% subset of its coordinates equal to 1 with zeros elsewhere.
-Random 70\% out of those non-zero coordinates in $\beta$ had 1s in respective positions in $\gamma$ ($\Gamma := \diag(\gamma)$), with the rest coordinates of $\gamma$ to be 0.
-
-	The results are presented on Figure \ref{fig:scalability_experiment}. One can immediately notice that the behavior of plots splits the picture into three different zones:
-	\begin{enumerate}
-		\item \textbf{Zone 1: ``tall'' datasets.} This zone corresponds to the situation when $n_i \geq p$ for all $i$, so every group has more observations than there are covariates in the dataset.
-The performance of the algorithm in this zone is nearly perfect, with all correct covariates to be selected.
-The train error is nearly the same to the test error which evidences that the model generalizes the data properly.
-		\item \textbf{Zone 2: ``square'' datasets.} This zone corresponds to the situation when $n_i < p$ but $n = \sum_i n_i > p$, so a problem of fitting individual regression models would lead to an underdetermined system for each group, but the problem of fitting a linear mixed regression as a whole is still well-defined.
-Although for $p \leq n/2$ the performance is still nearly perfect, there is a clear drop in performance in terms of both selection accuracy and RMSE on test data once $p \geq n/2$.
-This is a strong sign that the model overfits the data which leads to generalization errors.
-		\item \textbf{Zone 3: ``wide'' datasets.} This zone corresponds to solving underdetermined systems where $p > n$.
-We see that the performance of the method is roughly equal to performance of random guessing, which means that \ouralgo is not suitable for learning from ``wide'' datasets at this point.
-	\end{enumerate}
-	This analysis shows that the method can be successfully applied to the problems up until $p > n/2$, with ``wide'' dataset setups to be an apparent space for improvement.
-
-\subsection{Experiments on Real Data}
-\label{ch:experiments_on_real_data}
-\subsubsection{Meta-Research on Internet Bullying} 
-\label{ch:bullying}
-\begin{figure}
-	\includegraphics[width=0.95\textwidth]{Images/bullying_data.csv_intercept_only}
-	\caption{\label{fig:bullying_data_intercept_only}Model fit for Internet Bullying Data from GBD 2020 using only intercept and time.
-We see that the groups are not separable, but the fit of a mixed model is reliable: there is a distinctive minimum of the loss function around $\gamma=0.17$.}
-\end{figure}
-
-\begin{figure}
-	\caption{\label{fig:bullying_data_fixed_feature_selection}Fixed Feature Selection for Bullying Data from GBD 2020.
-The algorithm picks five historically significant covariates and two historically insignificant.
-See the Appendix \ref{appendix:bullying_covariates} for covariates description and assessment of significance.
-}
-	\includegraphics[width=0.92\textwidth]{Images/bullying_data.csv_fixed_feature_selection}
-\end{figure}
-
-\begin{figure}
-	\caption{\label{fig:bullying_data_random_feature_selection}Random Feature Selection for Bullying Data from GBD 2020.
-The algorithm picks five historically significant covariates and two historically insignificant.
-See the Appendix \ref{appendix:bullying_covariates} for covariates description and assessment of significance.}
-	\includegraphics[width=0.92\textwidth]{Images/bullying_data.csv_random_feature_selection}
-\end{figure}
-
-In this section we apply \ouralgo to identify the most important covariates in research on relative risk of anxiety and depression disorders depending on the exposure to bullying in young age\footnote{The data is available at ...}.
-This research has been a part of Global Burden of Diseases study for the last several years.
-The end goal for this work was to estimate the burden (DALYs) of major depressive disorder (MDD) and anxiety disorders that caused by bullying.
-For this risk factor, the exposure is primarily concentrated in childhood and adolescents, but the risk for MDD and anxiety disorders is anticipated to continue well into adulthood.
-This elevated risk is, however, expected to decrease with time as other risk factors come into play in adulthood (unemployment, relationship issues, etc.).
-To accommodate this, the research team uses the models which estimate the relative risk (RR) of MDD and anxiety disorders among persons exposed to bullying depending on how many years it has been since the first exposure.
-Studies informing the model were sourced from a systematic review and consist of longitudinal cohort studies.
-They measure exposure to bullying at baseline, and then follow up years later and assess them for MDD or anxiety disorders.
-The detailed description of the covariates can be found in Appendix \ref{appendix:bullying_covariates} .
-
-The feature selection process is illustrated on Figures \ref{fig:bullying_data_fixed_feature_selection} and \ref{fig:bullying_data_random_feature_selection}.
-Since there was no apparent prior on $k$ -- the number of features to keep in the model -- an AIC criterion from \cite{Vaida2005} was used to make a choice (see the plot on the first row, column two on Figure \ref{fig:bullying_data_fixed_feature_selection}).
-We see that there is a minimum around $k=9$, although the minimum is shallow so $k=8$ can also be considered.
-Five extra covariates (the ones outside \texttt{intercept} and \texttt{time}) were selected by \ouralgo: \texttt{cv\_unadjisted}, \texttt{cv\_threshold\_bullying},\texttt{cv\_b\_parent\_only}, \texttt{cv\_anx} and \texttt{percent\_female} (have been important in the previous years of GBD).
-The covariates \texttt{cv\_or} and \texttt{cv\_child\_baseline} has not been important previously but also got selected.
-
-\subsubsection{COVID-19 Contact Rate Forecasting Data}
-\label{ch:covid}
-\begin{figure}
-	\begin{subfigure}[b]{\textwidth}
-		\includegraphics[width=\textwidth]{Images/fit_Alaska}
-	\end{subfigure}
-	\begin{subfigure}[b]{\textwidth}
-		\includegraphics[width=\textwidth]{Images/fit_Slovenia}	
-	\end{subfigure}
-	\caption{\label{fig:covid_feature_selection_1} Comparison of fits of two different models (fully dense linear mixed model and \ouralgo) to the original IHME Projections for Alaska and Slovenia.
-The quality (RMSE) achieved by a sparse fit is within 10\% from a quality of both dense models which is evidences that the model picked up informative features.}
-\end{figure}
-
-\begin{figure}
-	\begin{subfigure}[b]{\textwidth}
-		\includegraphics[width=\textwidth]{Images/fit_Turkey}
-	\end{subfigure}
-	\begin{subfigure}[b]{\textwidth}
-		\includegraphics[width=\textwidth]{Images/fit_Switzerland}	
-	\end{subfigure}
-	\caption{\label{fig:covid_feature_selection_2} Comparison of fits of two different models (fully dense linear mixed model and \ouralgo) to the original IHME Projections for Turkey and Switzerland.
-The quality (RMSE) achieved by a sparse fit is within 10\% from a quality of both dense models which is evidences that the model picked up informative features.}
-\end{figure}
-
-In this section we apply our method to a COVID-19 Contact Rate Forecasting problem.
-The global pandemic created an unprecedented need for robust and accurate disease transmission forecasting.
-Since the beginning of the pandemic Institute for Health Metrics and Evaluation has been providing guidance to local authorities across the world with their pioneering \href{https://covid19.healthdata.org/global?view=total-deaths&tab=trend}{COVID-19 Projections tool} \cite{IHME2020Projections}.
-The key methodology which was essential for success in their forecasting of the disease's dynamics was transforming the death data into the contact rate time series data, and then relating this contact rate to available covariates such as temperature, social mobility, population, and others \cite{IHME2020covidnature}.
-All these covariates were collected in real time using limited human resources, and identifying the most important covariates was crucial for distributing these resources effectively.
-In the example below we show how \ouralgo can be used for making covariate selection on IHME data.
-
-The dataset consists of $m=60$ groups (countries or states), the detailed description of groups sizes ($n_i$) and time spans is provided in the Table \ref{table:covid_data_description} in the Appendix.
-The target variable $Y_i$ was the contact rate -- the coefficient $\beta(t)$ from an SEIR model (not to be confused with $\beta$ -- vector of fixed effects).
-The covariates -- columns of $X_i$ and $Z_i$ -- were: \texttt{intercept} --a column of ones,  \texttt{temperature} -- average air temperature in degrees Fahrenheit, \texttt{mobility\_lift} -- social mobility, \texttt{proportion\_over\_1k} -- population size threshold, and \texttt{testing\_reference} -- testing.
-The observation error's variance $\sigma_i$ was set to be 0.1.
-
-Feature selection results for four particular locations (Alaska, Slovenia, Turkey, and Switherland) are presented on Figures \ref{fig:covid_feature_selection_1} and \ref{fig:covid_feature_selection_2}, with coefficients for the rest of the locations attached in supplementary materials.
-\ouralgo was charged with a task to produce a fit using only three covariates, one of which had to be mean-only (no random effects).
-On the left we see the original data (\texttt{Data}, in {\color{blue} blue}), and predictions of three models: IHME Projections (\texttt{IHME}, in {\color{orange} orange}), a linear mixed model fit with no selection (\texttt{Dense MM}, in {\color{red} red}), and a sparse fit of \ouralgo (\ouralgo in {\color{green} green}).
-The \ouralgo model has chosen to use \texttt{intercept}, \texttt{mobility\_lift}, and \texttt{temperature} covariates, with the later only as a fixed covariate; \texttt{testing\_reference} and \texttt{proportion\_over\_1k} were left out.
-On the plots to the left we see that the exclusion of two covariates did not significantly affect the quality of predictions.
-The residual errors (RMSE) to the right support this statement: the difference in RSME is within 10\% of RSME of a "dense" model which uses all the covariates, as well as from IHME's original model which fit all the locations via sequence of independent regressions.
-This choice also seems reasonable given the timespan: the proliferation of testing during the spring was not yet significant, so it did not inform predictions in a major way.
-The exclusion of \texttt{proportion\_over\_1k} could have been due to the scale of grouping: locations were grouped on the level of states and countries, not on the level of individual counties where the influence of population-based covariates could have been more significant.
-
-\subsection{Software Implementation}
-In order to ensure reproducibility of our research, the proposed algorithm (\ouralgo, Algorithm \ref{alg:feature_selection_relax_and_split_outline}) has been implemented as a part of \texttt{\href{https://github.com/aksholokhov/skmixed}{skmixed}}\footnote{Available at \href{https://github.com/aksholokhov/skmixed}{https://github.com/aksholokhov/skmixed}} library.
-This library implements functionality for fitting linear mixed models and selecting covariates.
-The user interface was designed to be fully compliant with the standards\footnote{\href{https://scikit-learn.org/stable/developers/develop.html}{https://scikit-learn.org/stable/developers/develop.html}} of \texttt{sklearn} library to minimize learning time.
-
-\newpage
-\section{Discussion and Future work}
-\label{ch:future_work}
-
-Although the proposed algorithm performs on par with its competitors in its current form (Algorithm \ref{alg:feature_selection_relax_and_split_outline}), there are several improvements which the author plans to work on in future.
-In this section we discuss these improvements and outline possible paths to them.
-We plan to test and deploy these methods on global health applications through our collaborations with researchers at the Institute for Health Metrics and Evaluation.
-
-\subsection{Better Control over Relaxation Parameters}
-
-Both Relax-and-Split and Interior Point methods introduce relaxations to the problem.
-In particular, Relax-and-Split approach relaxes $\ell_0$-constraints using auxiliary variables, which gives us relaxation parameters $\lambda_\beta$ and $\lambda_\gamma$, whereas IP method relaxes equality constraints in (\ref{eq:log_barrier_relaxation_for_ip}) introducing another parameter $\mu$.
-All three parameters $\lambda_\beta$, $\lambda_\gamma$, and $\mu$ need to be tightened iteratively ($\lambda_\beta$, $\lambda_\gamma \to \infty$ and $\mu \to 0$) to guarantee that the variables $\beta$ and $\gamma$ satisfy the original constraints from (\ref{eq:fs_loss_intractable}).
-One can see this triplet as one big relaxation for the original problem and tune them all at the same place.
-
- For the IP method $\mu$ is set to via a complementary slackness condition:
- \eq{
- 	\mu^{(k)} := \sum_{i=1}^m \frac{v_i^{(k)}\gamma_i^{(k)}}{m}
- }
- that come from necessary optimality conditions on a KKT-system (\ref{eq:complementary_slackness_kkt}).
-At the same time, $\lambda_\beta$ and $\lambda_\gamma$ are increased in a static and rather arbitrary fashion: $\lambda := 2(1+\lambda)$.
-This approach is effective but not efficient: \cite{Zheng2018RelaxAndSplit} showed that the convergence is guaranteed, but one needs to iteratively solve the IP relaxation in full every time $\lambda_\beta$ and $\lambda_\gamma$ are increased.
-Adjusting them in the same place where $\mu$ is adjusted does not work because the steady exponential growth in $\lambda$ drives $\|F_\mu\|$ up faster than Newton method drives it down over iterations, which breaks the stopping criterion of IP method $\|F_\mu\| \leq \varepsilon$.\\
-
-The proposed approach is to increase $\lambda_\beta$ and $\lambda_\gamma$ based on the current step length $\alpha_k$.
-The logic behind this choice is that if the relaxation parameters are too large then the method will struggle with making a step away from the projections $\tbeta$ and $\tgamma$, and so no further increase is required.
-The key theorem which lays a theoretical foundation behind it would be:
-\begin{theorem}[Distance Between Minima]
-\label{thm:distance_between_minima}
-	For a fixed dataset $(X_i, Y_i)$ and relaxation parameters $\lambda_\beta$, $\lambda_\gamma$ the distance between $(\beta^*, \gamma^*)$, the unconstrained minimizers of relaxed problem, and their projections $(\tbeta^*, \tgamma^*)$ is bounded by a constant $M$ depending on $(X_i, Y_i)$ and the relaxation parameters.
-\end{theorem}
-
-There are two main reasons why the joint relaxation is more attractive than the current version.
-First, it will simplify the Algorithm \ref{alg:feature_selection_relax_and_split_outline} by removing the nested loop.
-This will accelerate the algorithm in terms of time and the number of iterations as both relaxations will be tuned within the same cycle with no further need to solve one relaxation per each step of another relaxation.
-Second, these theoretical findings may lead to developing a universal tuning technique for Relax-and-Split methods and might be used in other applications outside of mixed models setting.
-
-\subsection{Asymptotic Properties}
-
-Although \ouralgo can be classified as a penalized likelihood based method, regularity conditions from \cite{Fan2001}, which are being widely used by competitor algorithms, are not applicable here directly because the penalty is not of the form $P_{\lambda, n}(|\beta|, |\gamma|)$.
-Hence, a separate analysis of statistical properties of the proposed estimator is needed.
-
-The first theorem establishes the foundation for proving three consequent theorems by outlining the conditions under which the proposed method is guaranteed to find the right $k$-subspace:
-
-\begin{theorem}[Conditions for Convergence to True Estimator]
-	Under certain conditions the method converges in a finite number of iterations to $(\hat\beta, \hat\gamma)$ which projections $(\tbeta, \tgamma)$ belong to a $k$- and $j$-subspaces respectively that contain the true minimum $(\beta^*,\gamma^*)$.
-\end{theorem}
-The key challenge is to find the set of conditions which satisfy the theorem and at the same time are sufficiently general to cover real-world cases.
-For example, the conditions of $X_i$ and $Z_i$ to be sets of orthonormal vectors are known to be sufficient due to \cite{Tibshirani1996}, but without costly preprocessing the real world datasets are rarely orthonormal matrices.
-
-The next step is to prove a set of statements which is normally discussed in this type of works (see, for example \cite{Krishna2008, Hui2017}) using Theorem \ref{thm:distance_between_minima}.
-
-\begin{theorem}[Consistency of Estimator]
-	There exists a local minimizer $(\hat{\beta}, \hat{\gamma})$ for the proposed loss function, such that it is asymptotically consistent with true minimum $(\beta^*, \gamma^*)$. 
-\end{theorem}
-\textit{Proof Outline.} The proof is based on the fact that the set of stationary points for the original formulation (\ref{eq:fs_loss_intractable}) matches to the set of stationary points of the relaxation $\LL_r$ (\ref{eq:lmm_relax_and_split_full_setup}) when $\lambda_\beta, \lambda_\gamma \to \infty$.
-Hence, the first step would be providing a concentration bounds on how far a stationary point $(\hat{\beta}, \hat{\gamma})$  of the relaxation $\LL_r(\beta, \gamma)$ can be from a stationary point $(\beta^*, \gamma^*)$ of the original setup $\LL(\beta, \gamma)$ given a fixed $\lambda_\beta \gg 1$ and $\lambda_\gamma \gg 1$.
-This bound \textit{should be} $n$-independent meaning that the minima will match for any finite number of samples $n$.
-
-\begin{theorem}[Consistency in Zeros]
-	If some coordinates of the true minimizer ($\beta^*, \gamma^*$) are zero, then it is also zero in $(\hat{\beta}, \hat{\gamma})$, given that the later is sufficiently close to the former.
-\end{theorem}
-
-\textit{Proof Outline.} This is similar to the previous statement: the fact that the limiting minima of the relaxation $\LL_r(\beta, \gamma)$ match the minima of the original setup $\LL_{ML}(\beta, \gamma)$ implies that their zero coordinates also match.
-
-\begin{theorem}[Asymptotic Normality]
-	The proposed estimator $(\hat{\beta}, \hat{\gamma})$ asymptotically normally distributed around true minimizer ($\beta^*, \gamma^*$) in its true non-zero $k+j$-subspace.
-\end{theorem}
-
-These proofs would not only contribute to this particular work, but would also be a valuable development of Relax-and-Split approach in general.
-This is ongoing work which should be done by the time of Thesis Defense.
-
-\subsection{Exponentially Smoothed Projection}
-One of the aspects of Algorithm \ref{alg:feature_selection_relax_and_split_outline} which drags the selection accuracy downwards is premature, all-in, guessing of the sparse subspaces.
-Namely, $\tbeta$ and $\tgamma$ are assigned to be the direct projections of $\beta$ and $\gamma$, which may not identify the right subspace correctly during the first iterations when the problem is significantly relaxed.
-At the same time, the attraction to $\tbeta$ and $\tgamma$ chosen on the previous iteration biases the next $\beta$ and $\gamma$.
-One way of mitigating this effect is using exponential smoothing.
-In particular, lines \ref{alg:feature_selection_relax_and_split_outline:tbeta_line} and \ref{alg:feature_selection_relax_and_split_outline:tgamma_line} in the Algorithm \ref{alg:feature_selection_relax_and_split_outline} can be changed to the weighted exponential average.
-The proposed modification is outlined in Algorithm \ref{alg:exp_smoothing}.
-
-\begin{algorithm}
-	$\lambda_\beta = 0$; $\lambda_\gamma = 0$ \\
-	\Repeat{$\tbeta \approx \beta, \tgamma \approx \gamma$}{
-	$\lambda_\beta \leftarrow 2(1 + \lambda_\beta)$ \\
-	$\lambda_\gamma \leftarrow 2(1 + \lambda_\gamma)$\\
-	\Repeat{converges}{
-		$\tbeta^{(k+1)} \leftarrow \delta\proj_{\|\beta\|_0 \leq k}(\beta^{(k)}) + (1-\delta)\tbeta^{(k)}$\\
-		$\tgamma^{(k+1)} \leftarrow \delta\proj_{\|\gamma\|_0 \leq s}(\gamma^{(k)}) + (1-\delta)\tgamma^{(k)}$\\
-		$\beta^{(k+1)}, \gamma^{(k+1)} \leftarrow \argmin_{\gamma \geq 0, \beta}\LL(\beta, \gamma) + \frac{\lambda_\beta}{2}\|\beta - \tbeta^{(k)}\|^2_2 + \frac{\lambda_\gamma}{2}\|\gamma - \tgamma^{(k)}\|_2^2$\\
-	}
-	}
-	\BlankLine
-\caption{\label{alg:exp_smoothing} A prospective version of \ouralgo with exponential smoothing for projected variables. This modification should increase accuracy of covariates selection.}	
-\end{algorithm}
-
-The hyper parameter $\delta$ can be assigned via cross-validation tuning.
-Alternatively, it's possible to analyze this scheme via Nesterov's Accelerated Methods framework \cite{Nesterov2004} or Polyak's Heavy Ball framework \cite{Polyak1964} in hope that a proper choice of $\delta$ will increase the convergence rate of the algorithm.
-
-\subsection{Correlated Features}
-In Algorithm \ref{alg:feature_selection_relax_and_split_outline}, when we project $\beta$ onto a $k$-subspace constraint to get $\tbeta$ we take $k$ maximum elements in their absolute value from $\beta$ and put them into $\tbeta$, setting the rest to $0$.
-This can also be viewed as a minimization of a local quadratic model with a (trivial) approximation of Hessian with an identity matrix.
-Namely, for the iteration $j$:
-
-\eq{
-	\tbeta^* := & \argmin_{\tbeta} \LL_r(\beta^{(j)}, \gamma^{(j)}) + \nabla_\beta\LL_r(\beta^{(j)}, \gamma^{(j)})^T(\tbeta - \beta^{(j)}) + \half(\tbeta - \beta^{(j)})^TI(\tbeta - \beta^{(j)}) \\
-	& \text{ s.t. } \|\tbeta\|_0 \leq k
-}
-Since  $(\beta^{(j)}, \gamma^{(j)})$ is a partial minimizer of $\LL_r(\beta, \gamma, \tbeta, \tgamma)$ for the iteration $j$, we have that the second term equal to $0$.
-Notice $\tbeta^*$ would be a correct minimizer of the quadratic approximation of the loss function \textit{given that Hessian can be approximated by an identity}, which is not always the case.
-As we have from (\ref{eq:all_derivatives}):
-
-\eq{
-	\nabla^2_{\beta\beta}\LL(\beta, \gamma) & =  \half\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i
-}
-which is equal to $\half I$ in, for instance, the case of "orthonormal design": when all columns of $X_i$ are orthonormal for all $I$.
-This is not the case in general when fixed features are correlated, which suggests that \ouralgo might not work as well in a setup with highly correlated features.
-This formulation also suggests an extension:
-
-\eq{
-	\tbeta^* := & \argmin_{\tbeta} \LL_r(\beta^{(j)}, \gamma^{(j)}) + \nabla_\beta\LL_r(\beta^{(j)}, \gamma^{(j)})^T(\tbeta - \beta^{(j)}) + \half(\tbeta - \beta^{(j)})^T(I + \nabla^2_{\beta\beta}\LL(\beta^{(j)}, \gamma^{(j)}))(\tbeta - \beta^{(j)}) \\
-	& \text{ s.t. } \|\tbeta\|_0 \leq k
-}
-This is a ``quadratic knapsack'' problem which is NP-complete.
-Nevertheless, there are effective approximation methods for solving this problem.
-The hope is that this version of the algorithm will perform ultimately better in the setup where covariates are highly correlated.
+\input{cs}
 
@@
 \bibliography{bibliography}
+
 \clearpage
@@
 \appendix
-\section*{Acknowledgement}
-I am extremely grateful to my advisor, Sasha Aravkin, for providing support, guidance, and enthusiasm during my work on this project in the Department of Applied Mathematics and in Institute for Health Metrics and Evaluation.\\
-I would also like to thank Jim Burke for his detailed review of my work which would hopefully lead to fruitful collaboration. \\ 
-Last, but not least: I am grateful to Damian Santomauro, who introduced me to his work on the consequences of bullying, provided with a dataset, and with a valuable feedback on the performance of the proposed method.
-\\
+\chapter{Appendix}
+\section{Acknowledgement}
+\input{appendix/acknowledgements}
 
@@
 \label{appendix:derivatives_of_lmm}
-For conciseness, let us define the mismatch $\xi_i = Y_i - X_i\beta$. We also omit the dependence on $\beta$, as it's fixed at this point. The loss function \ref{eq:lmm_objective} takes the form 
-\eq{
-	\label{eq:lmm_simplified_gamma}
-	\LL\pa{\gamma} = \sum_{i = 1}^m\half\xi_i^T\pa{\Omega_i(\gamma)}^{-1}\xi_i + \half\log\det\pa{\Omega_i(\gamma)}.
-}
-
-The derivative of the objective w.r.t $\gamma_j$, the $j$'th diagonal element of the matrix $\Gamma$ is
-
-\eq{
-	\pderiv{\xi_i^T\Omega_i^{-1}\xi_i}{\Gamma_{jj}} & = \trace \left[\left(\pderiv{\xi_i^T\Omega_i^{-1}\xi_i}{\Omega_i}\right)\pderiv{\Omega}{\Gamma_{jj}}\right] = \trace\left[\left(-\Omega_i^{-T}\xi_i\xi_i^T\Omega_i^{-T}\right)^T Z_i\pderiv{\Gamma}{\Gamma_{jj}}Z_i^T\right] =
-}
-
-where $\pderiv{\Gamma}{\Gamma_{jj}}$ is a structure matrix, which, in a general case, is equal to a single-entry matrix $J^{jj}$ with $jj$'th element is equal to 1 and zeroes elsewhere.
-Substituting this back we get
-\eq{
-	= \trace\left[\left(-\Omega_i^{-T}\xi_i\xi_i^T\Omega_i^{-T}\right)^T Z^j_i{Z^j_i}^T\right] = 
-} 
-where $Z^j_i$ is a $j$'th column of the matrix $Z_i$. Making a circular swap we end up with
-
-\eq{
-	= \trace\left[-{Z^j_i}^T\Omega_i^{-T}\xi_i\xi_i^T\Omega_i^{-T} Z^j_i\right] = - ({Z_i^j}^T\Omega_i^{-T}\xi_i)^2
-}
-
-Similarly,
-
-\eq{
-	\pderiv{\log\det\Omega_i}{\Gamma_{jj}} = \trace\left[\left(\pderiv{\log\det\Omega_i}{\Omega_i}\right)\pderiv{\Omega_i}{\Gamma_{jj}}\right] = \trace\left[\Omega_i^{-1}Z_i^j{Z_i^j}^T\right] = {Z_i^j}^T\Omega_i^{-1}Z_i^j
-}
-
-Taking into account that $\Omega_i$ is symmetric, we have
-
-\eq{
-	\left[\nabla_\gamma \LL\pa{\beta, \gamma}\right]_j = \sum_{i = 1}^m  - ({Z_i^j}^T\Omega_i^{-T}\xi_i)^2 + {Z_i^j}^T\Omega_i^{-1}Z_i^j = 
-}
-
-or, in vector form
-
-\eq{
-    = \sum_{i = 1}^m  \diag{{Z_i}^T\Omega_i^{-1}Z_i} - ({Z_i}^T\Omega_i^{-T}\xi_i)^{\circ 2} =
-}
-
-where $\circ$ denotes the Hadamard (element-wise) product. Using the Cholesky decomposition $\Omega_i = L_iL_i^T$ we can calculate it more effectively, using only one triangular matrix inversion:
-
-\eq{
-    = \sum_{i = 1}^m \left[ \sum_{\text{rows}}\left(L_i^{-1}Z_i\right)^{\circ 2} -  [(L_i^{-1}Z_i)^{T}(L_i^{-1}\xi_i)]^{\circ 2}\right]
-}
-
-Notice, that the loss function (\ref{eq:lmm_objective}) and the optimal $\beta$ solution (\ref{eq:beta_formula}) can also be effectively computed using Cholesky:
-
-\eq{
-	\LL\pa{\gamma} = \sum_{i = 1}^m \half\xi_i^T\pa{\Omega_i(\gamma)}^{-1}\xi_i + \half\log\det\pa{\Omega_i(\gamma)} = \sum_{i = 1}^m \half\ \|L_i^{-1}\xi_i\|^2 - \sum_{j = 1}^k \log{[L^{-1}_i]}_{jj}
-}
-
-\eq{
-	\beta_{k+1} & = \argmin_{\beta}\LL(\beta, \gamma_{k}) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i = \\
-	& = \left(\sum_{i = 1}^m (L_i^{-1}X_i)^TL_i^{-1}X_i\right)^{-1}\sum_{i = 1}^m (L_i^{-1}X_i)^TL_i^{-1}y_i 
-}
-
-The Hessian w.r.t. $\gamma$ also can be found:
-
-\eq{
-	\frac{\partial^2 \LL\pa{\beta, \gamma}}{\partial \gamma_j^2} & = \sum_{i=1}^m -2({Z_i^j}^T\Omega_i^{-T}\xi_i)\trace\left[\frac{\partial{Z_i^j}^T\Omega_i^{-T}\xi_i}{\partial \Omega_i}\frac{\partial \Omega_i}{\partial \Gamma_{jj}} \right] + \trace\left[\frac{\partial{Z_i^j}^T\Omega_i^{-1}Z_i^j}{\partial \Omega_i}\frac{\partial \Omega_i}{\partial \Gamma_{jj}} \right] = \\
-	& = \sum_{i=1}^m 2({Z_i^j}^T\Omega_i^{-T}\xi_i)\trace\left[\Omega_i^{-1}Z_i^j{\xi_i}^T\Omega_i^{-1}Z_i^j{Z_i^j}^T \right] - ({Z_i^j}^T\Omega_i^{-T}Z_i^j)^2 = \\
-	& = \sum_{i=1}^m 2({Z_i^j}^T\Omega_i^{-T}\xi_i)({Z_i^j}^T\Omega_i^{-1}Z_i^j)({\xi_i}^T\Omega_i^{-1}Z_i^j) - ({Z_i^j}^T\Omega_i^{-T}Z_i^j)^2
-}
-
-\eq{
-	\frac{\partial^2 \LL\pa{\beta, \gamma}}{\partial \gamma_j\partial \gamma_k} & = \sum_{i=1}^m -2({Z_i^j}^T\Omega_i^{-T}\xi_i)\trace\left[\frac{\partial{Z_i^j}^T\Omega_i^{-T}\xi_i}{\partial \Omega_i}\frac{\partial \Omega_i}{\partial \Gamma_{kk}} \right] + \trace\left[\frac{\partial{Z_i^j}^T\Omega_i^{-1}Z_i^j}{\partial \Omega_i}\frac{\partial \Omega_i}{\partial \Gamma_{kk}} \right] = \\
-	& = \sum_{i=1}^m 2({Z_i^j}^T\Omega_i^{-T}\xi_i)\trace\left[\Omega_i^{-1}Z_i^j{\xi_i}^T\Omega_i^{-1}Z_i^k{Z_i^k}^T \right] - ({Z_i^j}^T\Omega_i^{-T}Z_i^k)^2 = \\
-	& = \sum_{i=1}^m 2(\xi_i^T\Omega_i^{-T}{Z_i^j})({Z_i^j}^T\Omega_i^{-1}Z_i^k)({Z_i^k}^T\Omega_i^{-1}{\xi_i}) - ({Z_i^j}^T\Omega_i^{-T}Z_i^k)^2
-}
-
-\eq{
-	\label{eq:lmm_hessian_gamma}
-	\nabla^2_\gamma\LL\pa{\beta, \gamma} = \half\sum_{i = 1}^m -(Z_i^T\Omega_i^{-T}Z_i)^{\circ 2}  + 2\diag(Z_i^T\Omega_i^{-T}\xi_i)(Z_i^T\Omega^{-1}Z_i)\diag(\xi_i^T\Omega^{-T}Z_i) = \\
-	= \half\sum_{i = 1}^m -(Z_i^T\Omega_i^{-T}Z_i)^{\circ 2} + 2(Z_i^T\Omega_i^{-T}\xi_i)(\xi_i^T\Omega^{-T}Z_i)^T\circ(Z_i^T\Omega^{-1}Z_i)
-}
+\input{appendix/derivatives_of_lmms}
 
@@
 \subsection{GBD Bullying Data}
-\input{tables/bullying_features_description}
+\input{appendix/bullying_features_description}
 

### 2023-04-06 pinode.tex
Initiate PhD thesis structure

\section{Introduction}

        Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs). 
        High-fidelity simulations of such dynamical systems can become intractable especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}. A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}. At the moment of writing this paper such systems are incapable of real-time complex simulations, but they already can run low-dimensional pre-trained models, thus inviting the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation} . ROMs, therefore, are essential for enabling design optimization, uncertainty propagation, predictive models, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}

        To be suitable for control, a ROM training method needs to find a low-dimensional manifold and a dynamics on it that together yield both high-accuracy predictions and long-term stability \cite{ahmed2021closures,noack2011reduced}. Most traditional ROMs are projection-based, e.g. dynamic mode decomposition (DMD) \cite{kutz2016dynamic,tu2013dynamic} and proper orthogonal decomposition (POD) \cite{holmes2012turbulence}, which transform the trajectories of a high-dimensional dynamical system into a suitable, and in some sense optimal, low-dimensional subspace. This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of basis functions (spatial modes) \cite{benner2015survey}. 
        One challenge for POD methods is their intrusive nature, i.e. requiring access to the solver codes.
       To overcome this, operator inference approaches \cite{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and then exploit lifting to fit the latent space dynamics data into
        polynomial, typically quadratic, models.
        These models, however, are (i) limited in representation power (up to quadratic, e.g. for lift and learn approach) and (ii) require a custom-tailored SVD-based optimization technique.

        In a thrust to overcome these challenges, significant effort has been invested into developing autoencoder-based reduced-order models, as a popular nonlinear ROM technique, which can yield both accurate and stable ROMs \cite{lee2020model,gin2021deep,champion2019data,kim2019deep}. In practice, however, they require datasets that densely cover a hypothetical infinite dimensional phase portrait of the dynamical system. Large demand for data significantly limits the use of such models in physics applications where the data can be expensive to obtain.

        Another severe challenge of utilizing ROMs comes from their poor out-of-distribution performance \cite{fries2022lasdi,cranmer2020discovering,gin2021deep}, especially when it is fundamentally impossible for a practitioner to obtain data that covers the entire distribution of possible data inputs. For example, in HVAC applications, one may collect data from a room with two windows but not from one room for every possible number of windows. In atmospheric LiDAR applications, we may conduct experiments on a certain terrain but we can never conduct experiments on all sorts of terrains \cite{nabi2020improving}. In such situations embedding the knowledge of physics into a model becomes necessary to improve extrapolation performance, and for which several approaches have recently been proposed.
        For instance, the seminal works \cite{bongard2007automated,schmidt2009distilling} have tried to determine the underlying structure of a nonlinear dynamical system from data using symbolic regression. Recently, Cranmer et al. \cite{cranmer2020discovering} employed symbolic regression in conjunction with graph neural network (GNN), while encouraging sparse latent representation, to extract explicit physical relations. They showed that the symbolic expressions extracted from the GNN generalized to out-of-distribution- data better than the GNN itself. However, symbolic regression also suffers from excessive computational costs,  and may be prone to overfitting.

        Another example of incorporating physics in ROMs is the use of parametric models at the latent space, e.g. by using the sparse identification of nonlinear dynamics (SINDy) \cite{brunton2016discovering,champion2019data}. For instance, \cite{fries2022lasdi,he2022glasdi} used a chain-rule based loss that ties latent-space derivatives to the observable-space derivatives for simultaneous training of the auto-encoder and the latent dynamics. However, such loss is highly sensitive to noise in data, especially when evaluating time-derivatives with finite differences is required\cite{delahunt2022sindynoise}. 
        Collocation-based enforcement of the physics, i.e. projection of the candid functions in the governing equations to enforce the chain rule instead of finite difference, could address such numerical difficulties. Recently, Liu et al. \cite{liu2022physics} used auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM. However, linearity may not be expressive enough for complex dynamics with multiple basins of attraction \cite{page2019koopman}.
        Finally, recent works on NeuralODE (NODE) \cite{chen2018neuralode,rackauckas2020udes} show a way to fit an arbitrary non-linear model (e.g. a network) as a latent space dynamics model, significantly extending the set of models for the latent dynamics that one can train efficiently.

        In this paper, we employ autoencoders to perform nonlinear model reduction along with NODE at the latent space to model complex and nonlinear dynamics. Our goal is to reduce their demand for data and improve overall forecasting stability under challenging training conditions. To achieve that we build on ideas from classical collocation methods of numerical analysis to embed knowledge from a known governing equation into the latent-space dynamics of a ROM, as described in Section 2. Then, in Section 3, we show that addition of our physics-informed loss allows for exceptional data supply strategies that improves the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible. We demonstrate that such an approach not only reduces the need for large training data-sets and produces highly-accurate and long-term stable models, but also leads to the discovery of more compact latent spaces, which is especially important for applications in compressed sensing and control.

            \begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/graphical-abstract.png}
    \caption{We propose utilizing a collocation points technique from numerical analysis to transfer knowledge of physics into continuous-time reduced-order models (ROMs). Such physics-informed models yield orders of magnitude more accurate predictions in tasks of far-out forecasting, predictions for out-of-distribution initial conditions, and learning from high-noise data. Good performance on those tasks is crucial for using ROMs in problems of compressive sensing and control.}
    \label{fig:architecture}
\end{figure}

\section{Method}
\label{sec:method}
\paragraph{Reduced-Order Model with Non-Linear Latent Dynamics}
We consider an autonomous dynamical system on a finite space $\XX \subseteq \mathbb{R}^n$

\begin{equation}
    \label{eq:generic_stationary_ode}
        \ddt\bd{x}(t) = \bd{f}(\bd{x}(t))
\end{equation}

In real-world applications it is often expensive to solve the relationship~(\ref{eq:generic_stationary_ode}) directly because $x(t)$ can be very high-dimensional. However, a variety of works provided both theoretical \cite{holmes2012turbulence} and practical \cite{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$. In that space, the dynamics evolve according to a (generally unknown) function  $\bd{h}(\bd{z})$:

\begin{equation}
    \label{eq:latent_stationary_ode}
        \ddt\bd{z}(t) = \bd{h}(\bd{z}(t))
\end{equation}

We call the space $\XX$ an observable space, and $\ZZ$ a latent space. When an invertible mapping $\psi:\ \ZZ \to \XX$ between the observable and the latent spaces is known, one can predict the dynamics of the system $\bd{x}$ at a future time $T$ by projecting the initial condition $\bd{x}(0)$ into the latent space, performing an integration there, and mapping the resulting trajectory back to the observable space:
\begin{equation}
\label{eq:integration_in_latent_space}
\begin{split}
    \bd{z}(0) & = \psi^{-1}(\bd{x}(0)) \\
    \bd{z}(T) & = \bd{z}(0) + \int_{0}^T\bd{h}(\bd{z}(t))dt \\
    \bd{x}(T) & = \psi(\bd{z}(T))
\end{split}
\end{equation}

When $m << n$ we refer to the triplet $(\psi, \psi^{-1}, \bd{h})$ as a Reduced-Order Model (ROM) of $\bd{f}$. It is often the case that for a given system $\bd{f}$ there exists no ROM $(\psi, \psi^{-1}, \bd{h})$ such that the relation~(\ref{eq:integration_in_latent_space}) holds exactly. In this case, we seek an \textit{approximation} ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ that minimizes the difference between the data $x(t)$ and the prediction $\hat{x}(t)$ over a chosen class of models $(\psi_\theta, \phi_\theta, h_\theta)$ parameterized by $\theta$.

Multiple real-world applications necessitate using ROMs instead of integrating the relation~(\ref{eq:generic_stationary_ode}) directly. For example, integrating~(\ref{eq:generic_stationary_ode}) may be computationally intractable especially on platforms with limited computing capability such as embedded and autonomous devices. For instance, in an HVAC system, solving~(\ref{eq:generic_stationary_ode}) means solving a Navier-Stocks equation on a fine grid in real time, which exceeds the computing capabilities of current-generation appliances. On the other hand, integrating~(\ref{eq:integration_in_latent_space}) may be cheap when $m << n$. Finally, even when solving~(\ref{eq:generic_stationary_ode}) is possible in real time (e.g. by utilizing a remote cluster), executing control over the resulting model, which is an end-goal for an HVAC system, may still be intractable. Indeed, executing control requires \textit{multiple} iterative evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\cite{duriez2017machine}.

\paragraph{Architecture} In this work we model $\psi$, $\psi^{-1}$, and $\bd{h}$ with fully-connected neural networks $\psi_\theta$, $\phi_\theta$, and $h_\theta$, respectively. Specifically, the pair ($\psi$, $\psi^{-1}$) is modelled with an auto-encoder $(\psi_\theta, \phi_\theta)$, and $\bd{h}$ is modelled with a fully-connected network $h_\theta$. Figure~\ref{fig:architecture} visualizes the architecture of the model.

\paragraph{Data-Driven Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses. The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data. Formally, for a given set of trajectories $\bd{x}_i$, $i \in [1 \dots k]$, where each trajectory $\bd{x}_i \in \mathbb{R}^{n \times p}$ is a set of $p$ snapshots that correspond to the recorded states of the system for $p$ time-steps, $t_j$, $j \in [1, \dots, p]$, the loss function $\LL^{data}_\theta$ is defined as:

\begin{align}
    \label{eq:loss_data_driven}
    \LL^{data}_\theta & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
\end{align}
where $\sigma$ is the standard deviation of the observation noise. We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}). This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}. To simplify the notation without loss of generality, in the rest of the paper we assume that all trajectories were recorded over the same time-frame with an equal and uniform step-size.

\paragraph{Physics-Informed Loss} In their recent work, \cite{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics. To extend this approach to the non-linear regime, we note that for a true mapping $\phi$ the following holds:
\begin{equation}
    \label{eq:chain_rule_1}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \frac{d\bd{z}}{d\bd{x}}\frac{d\bd{x}}{dt} = \nabla\phi(\bd{x}(t))^T\bf{f}(\bd{x(t)})
\end{equation}
On the other hand, by the definition of $\psi$ and $\bd{h}$ we have that
\begin{equation}
    \label{eq:chain_rule_2}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \bd{h}(\phi(\bd{x}(t))
\end{equation}
Combining Equations~(\ref{eq:chain_rule_1}) and~(\ref{eq:chain_rule_2}) we get that
\begin{equation}
    \label{eq:physsics_informed_equation}
    \bd{h}(\phi(\bd{x}(t)) = \nabla\phi(\bd{x})^T\bd{f}(\bd{x})
\end{equation}
Equation~(\ref{eq:physsics_informed_equation}) links the dynamics $\bd{h}(\bd{z})$ and the encoder $\phi(\bd{x})$ with the known equation $\bd{f}(\bd{x})$ and is true for all $z \in \ZZ$ and $x \in \XX$. Hence, knowledge of $\bd{f}$ can be assimilated into the model by evaluating Equation~(\ref{eq:physsics_informed_equation}) on a set of $N$ carefully sampled points $\bar{\bd{x}}_i \in \XX$, $i \in [1, \dots, N]$:

\begin{equation}
    \label{eq:loss_physics_informed}
    \LL^{physics}_\theta = \sum_{i = 1}^N \left[\frac{\omega_3}{N}\left\|h_\theta(\phi_\theta(\bar{\bd{x}}_i)) - \nabla \phi_\theta(\bar{\bd{x}}_i) \bd{f}(\bar{\bd{x}_i})\right\|^2 + \frac{\omega_4}{N}\left\|\bar{\bd{x}}_i - \psi_\theta(\phi_\theta(\bar{\bd{x}}_i))\right\|\right]
\end{equation}

We refer to the points $\bar{\bd{x}}_i$ as \textit{collocations}.

\paragraph{Collocations}
\label{sec:collocations_conditions} We define a collocation as pair $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$. Collocations are samples from the space $\XX \times Im_{f}(\XX)$, and they should satisfy three conditions, ordered by importance:
\begin{enumerate}
    \item \textbf{Simplicity}: $\bd{f}(\bar{\bd{x}}_j)$ should be computationally cheap to evaluate. It is especially important for PDE systems, where $\bd{f}$ may involve high-order derivatives.
    \item \textbf{Representativeness}: $\bar{\bd{x}}_j$ should cover the space of states where one aims to improve the model's performance or stability. Collocation points that a model might encounter and that are not represented by data snapshots are the best candidates.
    \item \textbf{Feasibility}: $\bar{\bd{x}}_j \in \XX$. In other words, $x_j$ should be an attainable state of the system. Collocation points outside of $\XX$ may downgrade the performance of the autoencoder by forcing it to be an invertible function on a domain outside of $\XX$ where a true mapping $\psi$ operates on.
\end{enumerate}
Thus, an optimal sampling procedure for collocations $\bd{\bar{x}}_j$ is domain-specific and should be designed given a particular system $\bd{f}$ and available data $\bd{x}_i$. We show examples of how these conditions can be implemented for real systems in the following sections.

Our definition of collocation points follows recent works\cite{liu2022physics}, which is a further development of the original definition \cite{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.

\paragraph{Combined Loss Function} We train the model by optimizing a sum of the physics-informed loss~(\ref{eq:loss_physics_informed}) and the data-driven loss~(\ref{eq:loss_data_driven}):
\begin{equation}
    \label{eq:loss_combined}
    \min_\theta \left[\LL^{physics}_\theta + \LL^{data}_\theta\right]  
\end{equation}
When $\omega_1 = \omega_2 = 0$ we have that $\LL^{data}_\theta = 0$, so we say that the model is (purely) \textit{\textbf{Physics-Informed}}. Similarly, when $\omega_3 = \omega_4 = 0$ we have that $\LL^{physics}_\theta = 0$ and we say that the model is (purely) \textit{\textbf{Data-Driven}}. When $\omega_i \neq 0, \, \forall i$, we say that the model is \textit{\textbf{Hybrid}}.

We use a \texttt{pytorch}~\cite{NEURIPS2019_9015} implementation of Adam algorithm~\cite{kingma2014adam} for optimization. To evaluate $\nabla_\theta \LL^{physics}_\theta$ and $\nabla_\theta \LL^{data}_\theta$ we use \texttt{torchdiffeq}~\cite{chen2018neural} -- a \texttt{pytorch}-compatible implementation of Neural ODE framework.

To the best of our knowledge, this is the first framework that combines non-linear latent-dynamics (Neural ODE), autoencoders, and a physics-informed loss term~(\ref{eq:loss_physics_informed}). Thus, we call our framework \textit{Physics-Informed Neural ODE}, or PINODE.

\section{Experiments}
\label{sec:exp}
The experiments section is organized as follows. First, to illustrate the ideas behind the framework we study its performance on a high-dimensional ODE -- a lifted Duffing oscillator. We show how a non-linear latent dynamics $\bd{h}(\bd{z})$ overcomes the limitations of DMD and Koopman networks from \cite{liu2022physics} by handling multiple basins of attraction within one model. We also show that using physics-informed loss is sufficient for reconstructing the behaviour for basins of attraction that are not represented by the data. Finally, we demonstrate that a purely data-driven model may be highly-accurate short-term and highly unstable long-term, even when the data is abundant, and show that the physics-informed approach improves long-term stability of such models by multiple orders of magnitude.

Next, we study the framework's performance on Burgers' equation. We show that (i) the non-linear latent dynamics yields more compact latent spaces than its linear counterpart for the same accuracy; (ii) these compact latent spaces allow for more stable long-term predictions; (iii) in the presence of significant noise in data, the use of collocations improves stability by providing an extra source of information that is noise-free, and (iv) in certain scenarios, training \textit{only} on collocations yields \textit{better} models than training on data, even when a vast amount of data is available. The latter shows that the contribution of physics-informed loss~(\ref{eq:loss_physics_informed}) may surpass the one of the data-based loss~(\ref{eq:loss_data_driven}), especially when the data is severely limited or noisy.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/duff_first_exp_abstract.png}
    \caption{We use a toy example -- a Lifted Duffing Oscillator -- to show that it is possible to ``fill the gaps'' in data with collocations. Namely, Hybrid model from Figure~\ref{fig:architecture} is able to learn the dynamics of two additional basins of attraction that were not represented in the dataset. As shown in the top-rightmost frame, without the collocations the model does not infer the dynamics in the unseen regions correctly.}
    \label{fig:duffing_pinode}
\end{figure}

\newpage
\subsection{Lifted Duffing Oscillator}
\label{sec:duffing}

A Duffing oscillator is a dynamical system $d\bd{z}/dt = \bd{h}(\bd{z})$ such that

\begin{equation}
    \label{eq:duffing_definition}
    \begin{split}
    \frac{dz_1}{dt} & = z_2 \\ 
    \frac{dz_2}{dt} & = z_1 - z_1^3
    \end{split}
\end{equation}

A phase portrait for 300 randomly sampled trajectories from this system is visualized on Figure~\ref{fig:duffing_pinode}, left frame. Depending on the total energy, each trajectory always stays in one of three regions: the left lobe, the right lobe, or the outer area, visualized in red, green, and blue respectively. To create a synthetic high-dimensional system that retains this property, we lift the Duffing trajectories into a higher-dimensional space by applying an invertible transformation $\mathcal{A}(\bd{z})$:

\begin{equation}
    \label{eq:duffing_true_decoder}
    \bd{x} := \mathcal{A}(\bd{z}) = A\bd{z}^3, \quad A \in \mathbb{R}^{128 \times 2}, \quad A_{ij} \sim_{i.i.d.} \mathcal{N}(0, 1)
\end{equation}

\begin{wrapfigure}{r}{0.25\textwidth}
  \begin{center}
    \includegraphics[width=0.23\textwidth]{figures/duffing_comparison.pdf}
  \end{center}
  \caption{Non-linearity in latent dynamics and autoencoder is important for long-term extrapolation. PIKN Hybrid model was unable to extrapolate the dynamics from collocations, while PINODE Hybrid model was.}%\vspace{-0.5in}
  \label{fig:duffing_comparison}
\end{wrapfigure}
Hence, for this system $z \in \ZZ = \mathbb{R}^2$ and $\bd{x} \in \XX = \text{span}\{A_{:,1}, A_{:,2}\} \subseteq \mathbb{R}^{128}$. We treat $\XX$ as an observable space, in which the dynamical system~(\ref{eq:duffing_definition}) obeys the following:
\begin{equation}
    \frac{d\bd{x}}{dt} = \bd{f}(\bd{x}) = \nabla((A^TA)^{-1}A^T\bd{x}^{1/3})^T\bd{h}((A^TA)^{-1}A^T\bd{x}^{1/3})
\end{equation}
Thus, we created a high-dimensional dynamical system with multiple basins of attraction for which the dynamics $\bd{f}$ is known.

For the experiment, we generate 6144 trajectories $\bd{x}_i$, $t=[0, 1]$, $\Delta t = 0.1$, all taken from the left lobe region (in red). We also sample 50000 collocations $\bar{\bd{x}}_j$ from the right (green) and the outer (blue) regions each by sampling $\bar{\bd{z}}_j \in U\left([-3/2,\, 3/2] \times [-1, 1]\right)$ and then applying the transformation~(\ref{eq:duffing_true_decoder}). For this example the conditions for collocations discussed in Section~\ref{sec:method} are trivially satisfied.

We train two PINODE models: Data-Driven model that only uses the trajectories and Hybrid model that uses both trajectories and collocations. The models share the same architecture (Figure~\ref{fig:architecture}) and training parameters that are detailed in Appendix~\ref{appendix:duffing_pinode}. After training, we invert the mapping~(\ref{eq:duffing_true_decoder}) to project the models' high-dimensional predictions for unseen initial conditions onto the true low-dimensional manifold; those are visualized in  Figure~\ref{fig:duffing_pinode}.

\begin{figure}[t]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/duffing_periods.pdf}
    \caption{ The time is measured in numbers of training time periods, i.e. $x=3T$ refers the time-range between two and three training time-periods away. The errors of a purely data-driven model grow exponentially when forecasting multiple times father in time than they were trained to do. In contrast, purely Physics-Informed model starts with worst interpolation accuracy but maintains nearly the same error when forecasting far out. Hybrid model combines the short-term accuracy of Data-Driven model and long-term stability of Physics-Informed model.}%\vspace{-0.4in}
    \label{fig:duffing_periods}
\end{figure}

We make two observations from the results displayed in Figure~\ref{fig:duffing_pinode}. First, a purely data-driven model is unable to extrapolate outside its training region using only the data from that region. This observation is consistent with the conclusions from related works~\cite{gin2021deep} that neural networks interpolate well but struggle with extrapolation tasks. Second, we see that collocations provided enough extra information for the model to predict nearly perfectly in regions from which no trajectories were provided. This observation suggests that one can use collocations to ``cover the gaps'' in data and improve the extrapolation accuracy of the model.

The non-linearity of the Neural ODE plays a crucial role in modeling the latent dynamics. To illustrate, we train a Hybrid PIKN model~\cite{liu2022physics} using the same dataset. PIKN differs from PINODE in using a linear latent dynamics $\frac{dz}{dt} = Lz$, where $L$ is a finite-dimensional approximation of Koopman operator, instead of a general non-linear dynamics $\frac{dz}{dt} = h_\theta(z)$. For PIKN we set $z \in \mathbb{R}^{16}$, an 8 times expansion of the dimension of the true manifold. In Figure~(\ref{fig:duffing_comparison}) we see that PIKN is unable to extrapolate the dynamics to unseen areas correctly using collocations: eventually, all trajectories "collapse" onto the same attractor. DMD shows even worse performance which could be attributed to the linear model reduction.

In the next experiment we show that collocations stabilize long-term predictions of the model even when data from all parts of space is available. To illustrate it we generate a dataset of 6144 trajectories (2048 per red, green, and blue area each) and 50000 collocations uniformly distributed among all three lobes. Next, we train three models: Data-Driven, Physics-Informed, and Hybrid versions of PINODE. Finally, we evaluate their relative performance for ever-increasing forecast time. In Figure~(\ref{fig:duffing_periods}), the x-axis represents test time-period. For example, $x = 2T$ represents prediction errors on a time-period $[2T, 3T)$ where the model was trained on trajectories of length $T$, and on $y$-axis we plot the distribution of prediction errors for 300 unseen trajectories within that period $[2T, 3T)$.

The results on Figure~(\ref{fig:duffing_periods}) show that the performance of the Data-Driven model degrades quickly when the forecasting time-period increases despite an excellent performance when forecasting within its train time-period. Physics-Informed model starts with modest performance but this performance does not degrade when forecasting far ahead. The Hybrid model, in its turn, combines both near-term accuracy with long-term stability, yielding best results over each time period.

\subsection{Burger's equation}
\label{sec:burger_eqn}
We now study the performance of our framework on Burger's equation with $[-\pi, \pi]$-periodic boundary conditions:
\begin{equation}
\begin{split}
\label{eq:burgers_equation}
    & u_t  + uu_x = \nu u_{xx} \\
    & u(-\pi, t) = u(\pi, t),\quad \forall t \in [0, T]
\end{split}
\end{equation}
where $u_t$, $u_x$, and $u_{xx}$ represent partial derivatives in time, the first, and second spatial derivatives, respectively. Burgers' equation is a PDE occurring in applications to acoustics, gas and fluid dynamics, and traffic flows \cite{burgers1948mathematical}. When $\nu$ is significantly smaller than one, the system exhibits strong non-linear behaviour and is called ``advection-dominated'', otherwise when $\nu$ is large the system is called ``diffusion-dominated''. 
In the case of the former, linear projection methods such as POD become inaccurate as the true solution space has a slow decaying Kolmogorov n-width, manifesting itself in slow decaying singular values \cite{peherstorfer2022breaking}. Therefore, in this section we focus on the advection-dominated Burger for which we set $\nu = 0.01$.

To generate trajectories, we discretize the spacial domain $[-\pi,\,\pi]$ with 128 grid-points, and we solve Equation~\ref{eq:burgers_equation} for $t \in [0, 2]$ with $\Delta t = 0.1$ using a spectral solver~\cite{trefethen2000spectral}. To generate a diverse set of initial conditions we sum the first 10 harmonic terms with random coefficients:
\begin{equation}
    \label{eq:burger_initial_condition}
    u(x, 0) = \frac{1}{10}\sum_{k = 1}^{10} a_k\cos(kx) + b_k\sin((k+1)x), \quad a_k, b_k \sim \mathcal{N}(0, 1)
\end{equation}

To generate collocations we use the same family of functions as we used for the initial conditions in Equation~(\ref{eq:burger_initial_condition}), and we additionally randomize the presence of individual frequencies in the sum:
\begin{equation}
    \label{eq:burger_collocations}
    \bar{u}(x) = \frac{1}{10}\sum_{k = 1}^{10} p_ka_k\cos(kx) + q_kb_k\sin((k+1)x), \quad a_k, b_k \sim \mathcal{N}(0, 1), \quad p_k, q_k \sim Be(1/2).
\end{equation}
We choose this family of collocations to meet the conditions~(\ref{sec:collocations_conditions}). First, this family is representative of the state space $\XX\times Im_f(\XX)$ in the region of interest (moving wave-fronts). Second, (\ref{eq:burger_collocations})~is a smooth set of functions that does not contain unattainable states. Finally, and more importantly, the values $u_x$ and $u_{xx}$ and, consequently $u_t$ can be computed analytically, which makes it especially cheap to sample large amounts of collocations.

\subsubsection{Compressibility of the Latent Space}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/burgers_examples_of_ics.pdf}
    \caption{Examples of "harmonic", "bell-curve", and "bump" initial conditions, as well as the resulting solutions, in columns 1, 2, and 3 respectively.}
    \label{fig:burgers_examples_of_ics}
\end{figure}

\label{sec:compressibility}
\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/compressibility.pdf}
    \caption{PINODE Hybrid model utilized the latent space dimension 5 times more efficiently by MSE than PIKN Hybrid model when modelling low-viscosity (highly-nonlinear) Burger's equation (left frame). The difference in performance grows to x100, when forecasting two times farther than the train period (central frame). The reason behind PIKN's long-term instability is the precsence of eigenvalues with poitive real part in the latent dynamics matrix (right frame).}
    \label{fig:burgers_compressibility}
\end{figure}

In Section~\ref{sec:duffing} we showed that a non-linear finite-dimensional latent dynamics model can be a necessity for building a compact ROM for a high-dimensional system. It is \textit{not} the case for Burgers' equations since there is Cole-Hopf transformation that linearize the dynamics. However, a latent-space non-linearity can, in principle, be utilized for finding a more compact latent space, or for increasing the forecast accuracy for the same size of the latent space. In this section we give an example of how PINODE achieves both goals.

For this experiment we generate 16384 trajectories as described in~(\ref{eq:burger_initial_condition}). We also generate 100000 collocations as described in~(\ref{eq:burger_collocations}). We use the exceedingly large amount of data to allow the models to achieve the best performance for the allowed dimension of the latent space. We evaluate the performance of the models on test data with two different time-frames: (1) same as the training data (\textit{interpolation}), and (2) two times longer than training data (\textit{extrapolation}). More details of the experimental setup are provided in Appendix~(\ref{appendix:burgers_compressibility}).

We compare the performance of three models: DMD, PIKN Hybrid, and PINODE Hybrid. First, we notice that DMD, despite achieving small loss values on training data ($\sim 10^{-3}$), does not perform well on test data. This observation is consistent with earlier works~(\cite{kalur2021robust,kutz2016dynamic}); it illustrates well that a combination of a linear encoder and a linear latent dynamics may not be sufficient for modelling highly-nonlinear phenonmena. Second, we notice that PINODE achieves better performance within a given latent space budget. For instance, for $m = 16$ (Figure~(\ref{fig:burgers_compressibility}), left pane), PINODE achieves $\sim 5$ times lower mean squared error than PIKN, which achieves the same performance only when $m = 512$. More importantly, PINODE maintains low error when predicting for a longer-term horizon (extrapolation in time), which is not the case for PIKN (Figure~(\ref{fig:burgers_compressibility}, right pane). This is a consequence of the latent-dynamics matrix ($h(z) = Lz$) of PIKN having eigenvalues with positive real parts, which implies long-term instability (Figure~(\ref{fig:burgers_compressibility}), right pane). Although there has been progress in the literature~\cite{kojimalearning}, further research is needed to understand (i) how to enforce stability constraints for PIKN, and (ii) why one does not need the same enforcement for PINODE to exhibit stable behaviour.

\subsection{Training in Low-Data Regime with Collocations}
\label{sec:data_vs_collocations}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/data_vs_collocations.pdf}
    \caption{The figure shows a comparison of the achievable MSE relative to the full data regime (1024 trajectories). When the data is scarce, collocations-based physics-informed loss improves the forecasting accuracy of ROMs by an average of 5 times lower MSE compared to the data-only regime, as shown in this experiment with Burger's equation. Moreover, when other types of initial conditions (``harmonic'', ``bell-curve'') are used, the physics-only model (top-right corner of the right frame) outperformed the most data-rich model in our experiment (bottom-left corner). }
    \label{fig:burger_data_vs_collocations}
\end{figure}

In the next experiment we study relative efficiency of using collocations against using data in a low-data regime. It is frequently the case that only a small number of simulations (or measurements) can be obtained for a physical system of interest due to the computational, time, or budget constraints. We would like to compensate the lack of data with providing collocations which are considerably cheaper to generate. The collocations, however, can be a ``weaker'' signal for the model than real data, especially when time passes and the shape of the solution moves out of the collocation family's span. In this section we show that collocations can be effectively used for this goal, when chosen right, and their contribution to a model's accuracy may even surpass the contribution of the data.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/example_burgers.pdf}
    \caption{The first subplot shows the relative error of solving Burger's equations on 100 test (unseen) initial conditions for two models: PINODE Hybrid and PINODE Data-Driven. Both models interpolate well but a purely data-driven model fails to extrapolate past the time-period that it was trained on (to the left of the red line). PINODE-Hybrid provides stable long-term predictions that indicates a high-quality of a discovered low-dimensional manifold dynamics. An example of such effect is provided on subplots~2-4.}
    \label{fig:burgers_example}
\end{figure}

To illustrate the trade-off between data and collocations, we train one model per various combinations of number of trajectories vs collocations in their training sets. To gauge the extrapolation power of our models, we use trajectories with three types of initial conditions: ``harmonic'', ``bell-curve'', and ``bumps'' (see Figure~(\ref{fig:burgers_examples_of_ics}) for equations and illustrations). For training, we use 2048 trajectories with ``bumps'' initial conditions. For collocations, we use the harmonic family as described in~(\ref{eq:burger_collocations}). We use two test datasets: (1) 100 trajectories with ``bump'' ICs to assess within-distributuion performance, left frame), and (2) a mix of trajectories with ``bump'', ``bell-curve'', and ``harmonic'' initial conditions, 100 of each, to assess out-of-distribution performance. All test data trajectories were two times longer than the train trajectories. More details on the experimental setup are provided in Appendix~\ref{appendix:burgers_data_vs_collocations}. We fit one PINODE model with the latent dimension $m = 16$ per each combination of the amount of data and collocations. On Figure~(\ref{fig:burger_data_vs_collocations}) we report its performance on two test datasets relative to the richest data-driven model with no collocations (bottom-left corner).

From the results displayed on Figure~(\ref{fig:burger_data_vs_collocations}) we see that adding collocations always improved the model in our experiments.  Moreover, when a sufficient number of collocations is added in training, the model with fewer data snapshots was always able to outperform the model with the maximum amount of data but no collocations. On average, a collocation-aided model was \textit{5 times better} at both within-distribution and out-of-distribution performance relative to a purely data-driven version of the model. In addition, we noticed that a model that used only collocations can perform better than a data-rich model, especially when predicting the dynamics of the unseen initial conditions (Figure~(\ref{fig:burger_data_vs_collocations}), right pane, top-right vs bottom-left corner).

We also notice that the Hybrid models yield more stable and accurate predictions, relative to their purely data-driven counterparts, when forecasting far beyond the training time-period. In Figure~(\ref{fig:burgers_example}) we visualize the predictions for a test IC for two models: Data-Driven model from the bottom-left corner of Figure~(\ref{fig:burger_data_vs_collocations}), and a Hybrid model from the bottom-right corner of Figure~(\ref{fig:burger_data_vs_collocations}). The red line separates the time-period of training from the time-period of forecasting. The hybrid model's errors stay below $10^{-2}$ even when forecasting 10 times farther than what it was trained on. In contrast, Data-Driven model shows similarly low errors within its training time-region but the forecast errors grow quickly when forecasting beyond that.

\subsection{Robustness to Noise in Low-Data Regime}
\label{sec:burger_noise}
\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/burgers_noise.pdf}
    \caption{Physics-informed loss works as a safeguard that prevents unbounded performance drop when quality of the data degrades due to noise. Namely, the solution of the hybrid loss~(\ref{eq:loss_combined}) converges to the solution of the physics-informed loss~(\ref{eq:loss_physics_informed}), when the data-driven loss~(\ref{eq:loss_physics_informed}) becomes uninformative. The performance of purely data-driven methods (Data-Driven, DMD) grows unbounded since these models don't have an alternative noise-independent source of information.}
    \label{fig:burger_noise}
\end{figure}

In this section we show that the use of collocations improves the ROM's robustness when the data is noisy by providing an alternative, noise-free, source of information.

For this experiment, we use the datasets from the bottom-right cell of Figure~(\ref{fig:burger_data_vs_collocations}), Pane~2. Namely, we use 1024 trajectories with "bump" initial conditions (an example is on Figure~(\ref{fig:burgers_examples_of_ics})), and we use 65536 "harmonic" collocations as defined in Equation~\ref{eq:burger_collocations}. We then add i.i.d. Gaussian noise to the trajectories, with the variance ranging from $\sigma = 10^{-4}$ to $\sigma = 10$. For reference, most of the data values lie between $0$ and $1$, so a noise level with $\sigma > 1$ dominates the data. We train four models: PINODE Hybrid, PINODE Data-Driven, PINODE Physics-Informed, and DMD. To measure the models' out-of-distribution prediction errors we use the Test Dataset 2 from the previous section; this dataset consists of three types of noise-free trajectories (see Figure~\ref{fig:burgers_examples_of_ics}), 100 each. The errors are displayed on Figure~(\ref{fig:burger_noise}), left pane. The error of a purely Physics-Informed model (in red) is flat because the collocations are noise-free and thus the same for all noise-levels, so we only train this model once.

On Figure~(\ref{fig:burger_noise}) we see that, when noise is high, the error of purely data-driven models grows unbounded, whereas the performance of the hybrid method converges to the performance of the Physics-Informed model. We hypothesise that it happens because the second part~($\LL_{\theta}^{data}$) of the combined loss~(Eq.~\ref{eq:loss_combined}) turns into noise, and so its derivative also turns into noise.
\begin{equation}
    \nabla \LL_{\theta} = \underbrace{\nabla\LL_{\theta}^{physics}}_{\text{informative}} + \underbrace{\nabla\LL_{\theta}^{data}}_{\text{noise}}
\end{equation}
Thus, one can think about optimizing a hybrid model~(\ref{eq:loss_combined}) as about training Physics-Informed model~(\ref{eq:loss_physics_informed}) using a noisy gradient descent with a fixed-variance noise. From the optimization literature \cite{friedlander2012hybrid,patel2021global,shapiro2021lectures} we know that, under certain conditions, such SGD converges to a neighbourhood of a local minima of its loss (in this case $\LL_{\theta}^{physics}$) with high probability. So instead of diverging, a hybrid model turns into Physics-Informed model; where the latter works as a performance safeguard in the high-noise regime.  On the right hand-side of Figure~(\ref{fig:burger_noise}) we show an example of the prediction performance of each of the models described above. The data-driven and hybrid models yield visually similar solutions when $\sigma = 10^-3$. However, the former provides inadequate performance when the data is dominated by noise (which is expected), whereas a hybrid model in this regime produces a solution that is visually similar to the one that Physics-Informed model produces. A more rigorous analysis of this phenomenon seems possible but lies outside of the scope of this paper.

\section{Discussion and Conclusions}

In this work we demonstrated how a collocation point technique can improve the performance of an emerging class of continuous-time neural-network based reduced-order models. First, it can ``cover the gaps'' in datasets and inform the model about underrepresented basins of attraction. It alleviates the demand for large volumes of data that is common in network-based models, which is crucial in applications where the data is scarce and expensive. Second, the physics-informed loss may work as a safeguard, providing a noise-independent source of truth.
Third, collocations can stabilize the model's long-term predictions, allowing to accurately forecast far beyond the training horizon. Finally, together with using a NODE-based non-linear latent dynamics, adding physics-informed loss leads to discovery of more compact latent spaces that yield more accurate models. Simultaneous stability and compactness is especially important if one aims to use models together with compressive sensing and control algorithms.

One clear limitation of the current work is that the choice of an efficient collocation family is a design decision that a practitioner makes. The authors believe that such decisions can be automated by adopting existing approaches from classic works on numerical approximations of PDEs, which we leave for future research. Another automation that prompts future research is more efficient way of sampling collocation points, possibly via applying modern adaptive learning techniques \cite{subramanian2022adaptive}. Finally, although Section~\ref{sec:burger_noise} provides some rationale why one may expect robustness of Hybrid models under noise, the authors believe that a more rigorous analysis is possible; particularly one that provides conditions under which such robustness is guaranteed.

### 2023-04-06 tables/tables/bullying_features_description.tex
Initiate PhD thesis structure

--- 2022-07-04 tables/bullying_features_description.tex
@@
 		\item 1 = study uses a symptom scale (e.g., Beck Depression Inventory) and uses an established cut-off on that scale to determine caseness. 
-		\item Has not historically been significant. 
+		\item Has not  been significant in the past. 
 	\end{itemize}
@@
 		\item 1 = Only parent is involved in reporting the child’s exposure to bullying
-		\item This covariate has recently started becoming significant (but not consistently). 
+		\item Has recently been significant. 
 	\end{itemize}
@@
 		\item 1 = estimate is an odds ratio (OR)
-		\item ORs are always larger than RRs. However the magnitude may be very small / insignificant.
+		\item ORs are always larger than RRs; covariate may not be significant. 
 	\end{itemize}
@@
 	\begin{itemize}
-		\item 0 = uses a ‘frequent’ exposure frequency threshold for classing someone as exposed to bullying.
-		\item 1 = uses an ‘occasional’ exposure frequency threshold for classing someone as exposed to bullying.
-		\item Has been consistently significant with a strong magnitude.
+		\item 0 = uses a ‘frequent’ exposure  threshold for classing someone as exposed to bullying.
+		\item 1 = uses an ‘occasional’ exposure threshold for classing someone as exposed to bullying.
+		\item Has been significant in the past. 
 	\end{itemize}
@@
 		\item 1 = $\geqslant$ 15\% attrition at followup
-		\item Has been significant in the past
+		\item Has been significant in the past.
 	\end{itemize}
@@
 	\begin{itemize}
+	    \item Indicates whether mid-age of sample is above or below 13.
 		\item Has not been significant in the past.

### 2023-04-06 presentation/pinode.tex
Restructure presentation

\section{Reduced-Order Models}
\subsection{Intro to Reduced-Order Models (ROMs)}

\begin{frame}{Reduced-Order Models}

Schematics, terminology, notation

\end{frame}

\begin{frame}{Variations of ROMs}
	Linear, nonlinear, citations
\end{frame}

\begin{frame}{Physics-Informed Loss}
Definition of Physics-informed ML.

Derivation of PI loss through chain rule.

Problem with evaluating RHS of differential equations.

Interpretation with cross-borrowing strength. 
\end{frame}

\begin{frame}{Physics-Informed Koopman Networks}
	Linear latent dynamics

	Paper

	Not a rom so not suitable for downstream tasks. 
\end{frame}

\begin{frame}{Physics-Informed Neural ODEs (PINODE)}
	Nonlinearity in latent space
	- require a lot of data
	- picky to initialization
	- good interpolation bad extrapolation	
\end{frame}

\begin{frame}{PINODE Results on Burgers}
	Add PI loss => Fixes most of the issues
	(example with Burgers)
	Points learned
	- Inform the model about unseen conditions
	Paper reference
\end{frame}

\begin{frame}{PINODE Active}
	> PIKN: "we use it and it works"
	> PINODE: "this is how you need to use it right"
	> PINODE-Active: can we use it right automatically?

	Loop scheme
\end{frame}

\begin{frame}{PINODE Active Results}
	Best result (Burgers)
\end{frame}

\begin{frame}{PINODE-CS}
	Application of the above 
	Loss and setup
\end{frame}

\begin{frame}{PINODE-CS Results}
	Results on Burgers
	Maybe results on a harder problem
\end{frame}

### 2023-04-06 presentation/pysr3.tex
Restructure presentation

--- 2022-07-23 presentation/general_presentation.tex
@@
-\documentclass[8pt]{beamer}
-\input{useful_packages}
-\input{math_macros}
+\section{Linear Mixed-Effect Models}
+\subsection{ Mixed-Effects Models}
+\begin{frame}{Feature Selection for Mixed-Effect Models}
 
-% \usetheme[progressbar=frametitle]{metropolis}
-\usepackage{appendixnumberbeamer}
-
-% PRESETS OF PACKAGES AND MATH %
-
-\usepackage{xspace}
-\usepackage{caption}
-\usepackage{subcaption}
-\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
-\newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
-\title{A Relaxation Approach to Feature Selection \newline for Linear Mixed-Effects Models}
-\date{\today}
-\author{Aleksei Sholokhov, James V. Burke, Peng Zheng, Damian Santomauro, and Aleksandr Aravkin}
-\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
-
-\begin{document}
-
-\maketitle
-
-\section{Relaxed for Linear Mixed-Effect Models}
-\subsection{Mixed-Effects Models}
-\begin{frame}{Mixed-Effect Models}
-
+Mixed-effect models
 \begin{itemize}
@@
 	\item \textbf{Borrow strength} across groups to estimate key statistics. % when data within units are sparse or highly variable.
-	\item Often seek \textbf{sparse models} that only use \textbf{most informative} covariates. 
 \end{itemize}
@@
 
+\begin{frame}{Feature Selection for Mixed-Effect Models}
+Practitioners:
+\begin{itemize}
+	\item Often seek \textbf{sparse models} that select \textbf{most informative} covariates. 
+	\item Want to be \textbf{flexible but efficient} in using various sparsity-promoting terms.
+	\item Want a library to be \textbf{universal and compatible} with e.g. \texttt{sklearn}.
+\end{itemize}
+\vspace{1em}
+
+Sparse Relaxed Regularized Regression ($\mathcal{SR}3$) \cite{Zheng2018RelaxAndSplit} showed great results for t linear models:
+
+\begin{figure}
+	\includegraphics[width=0.7\textwidth]{Figures/intuition_prev_paper.png}
+\end{figure}
+
+\vspace{1em}
+\textit{\textbf{Goal}: create a feature selection library that uses a relaxation approach for feature-selection in mixed-effect models.}
+
+\end{frame}
+
 \begin{frame}{Linear Mixed-Effect (LME) Models}
@@
 		\item 	$X_i \in \R^{n_i \times p}$ -- group $i$ design matrix for fixed features
-		\item 	$Z_i \in \R^{n_i \times q}$ -- group $i$ design matrix for random effects
+		\item 	$Z_i \in \R^{n_i \times q}$ -- group $i$ design matrix for random features
 		\item 	$y_i \in \R^{n_i}$ -- group $i$ observations  
@@
 	\begin{columns}[T,onlytextwidth]
+
     \column{0.5\textwidth}
-	 \centering Standard Linear Regression:
-   	\begin{figure}
-   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_mean_prediction}
-   	\end{figure}
+    \vspace{3em}
+    Model:
 	 	\[
-   		y = X\beta + \varepsilon, \quad \varepsilon \sim \NN(0, \Lambda) 
+   		\begin{split}
+   			y_i & = X_i\beta + {\color{red}Z_i u_i} + \varepsilon_i \\
+   			 \varepsilon_i & \sim \NN(0, \Lambda_i) \\
+   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma)}
+   		\end{split}
+   		\]
+
+   	Equivalently:
+   		\[
+   		\begin{split}
+   			y_i & = X_i\beta + \textcolor{red}{\omega_i} \\
+   			{\color{red} \omega_i} & {\color{red}\sim \NN(0, Z_i\Gamma Z_i^T + \Lambda_i)}
+   		\end{split}
+   		\]
+
+   	Simplifying assumption: 
+   	\[
+   		\Gamma = \diag(\gamma)
    	\]
@@
     \column{0.5\textwidth}
-    	\centering  Linear Mixed-Effect Model:
+    	\centering  
    	\begin{figure}
@@
    	\end{figure}
-   	   		\[
-   		\begin{split}
-   			y_i & = X_i\beta + {\color{red}Z_i u_i} + \varepsilon_i, \quad \varepsilon_i \sim \NN(0, \Lambda_i) \\
-   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma)}
-   		\end{split}
-   		\]
 
@@
 \begin{frame}{Likelihood for Mixed Models}
-Negative log-likelihood:
+%Maximum likelihood estimates for $\beta$ and $\gamma$ solve the problem:
+%\eq{
+%	\mathcal{LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma)
+%}
+
+Optimization problem:
+\eq{
+	\mathcal{FS-LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
+}
+
+Where $\LL$:
 \eq{
@@
 	}
-Maximum likelihood estimates for $\beta$ and $\gamma$ solve the problem:
-\eq{
-	\mathcal{LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma)
-}
-
-To select covariates we add a sparsity-promoting regularizer $R(\beta, \gamma)$
-\eq{
-	\mathcal{FS-LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
-}
 
@@
 	\item $\LL(\beta, \gamma)$ is smooth on its domain, quadratic w.r.t. $\beta$ and $\bar\eta$-weakly-convex w.r.t. $\gamma$.
-	\item $R(\beta, \gamma)$ is closed, proper, convex, with easily computed \textit{prox operator}
+	\item $R(\beta, \gamma)$ is closed, proper, with easily computed \textit{prox operator}
 \end{itemize}
@@
 	\label{eq:value_function_definition}
-	u_{\eta,\mu}(\tbeta,\tgamma) & := \min_{(\beta, \gamma)} \LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\  & := \min_{(\beta, \gamma)} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma)
+	v_{\eta,\mu}(\tbeta,\tgamma) & := \min_{(\beta, \gamma)} \LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\  & := \min_{(\beta, \gamma)} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma)
 }
@@
 \eq{
-	\min_{(\tbeta, \tgamma) \in \CC} u_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
+	\min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} v_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
 }
@@
 \begin{itemize}
-	\item $u_{\eta,\mu}$ is well-defined and continuously differentiable.
+	\item $v_{\eta,\mu}$ is well-defined and continuously differentiable.
 	\item As $\mu \rightarrow 0$ and $\eta \rightarrow \infty$, cluster points of solutions to $\ouralgo$ are first-order stationary points for $\mathcal{FS-LME}$ \\ 
+	\item $v_{\eta, \mu}$ don't need to be evaluated precisely.
 \end{itemize}
 
-\textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations keep the solution close to the central path.
+% \textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations keep the solution close to the central path.
 \end{frame}
 
-\begin{frame}{Value Function Reformulation}
-\begin{figure}
-	\includegraphics[width=\textwidth]{Figures/intuition_prev_paper}
-	\caption{\label{fig:intuition_prev} Picture from \cite{Zheng2018RelaxAndSplit}: for a linear problem, value function relaxation ``squashes'' level-sets simplifying the optimization landscape.}
-\end{figure}
-\end{frame}
+%\begin{frame}{Value Function Reformulation}
+%\begin{figure}
+%	\includegraphics[width=\textwidth]{Figures/intuition_prev_paper}
+%	\caption{\label{fig:intuition_prev} Picture from \cite{Zheng2018RelaxAndSplit}: for a linear problem, value function relaxation ``squashes'' level-sets simplifying the optimization landscape.}
+%\end{figure}
+%\end{frame}
 
@@
 \begin{frame}{Application to Synthetic Problems}
-\textbf{The Experiment} 
 \begin{itemize}
@@
 \end{itemize}
-\begin{table}
-	\input{../extras/performance_table_short_current}
-\end{table}
+\begin{figure}
+	\includegraphics[width=\textwidth]{Figures/performance_without_msr3.pdf}
+\end{figure}
+\begin{itemize}
+	\item[\textcolor{green}{+}] $\ouralgo$-relaxation improves feature selection performance of the original likelihood.
+	\item[\textcolor{green}{+}] $\ouralgo$-fast optimization accelerates the compute time by $\sim 10^2$.
+   	\item[\textcolor{red}{--}] Initialization of $\eta$ is problem-specific
+\end{itemize}
 \end{frame}
 
-\begin{frame}{Application to Synthetic Problems}
-\begin{figure}
-	\includegraphics[width=\textwidth]{Figures/performance_picture_current.pdf}
-\end{figure}
-Benefits:
-\begin{itemize}
-	\item $\ouralgo$-relaxation improves feature selection performance than the original likelihood.
-	\item $\ouralgo$-fast optimization accelerates the compute time by $\sim 10^2$.
-\end{itemize}
-Challenge:
-   \begin{itemize}
-   	\item Initialization of $\eta$ is problem-specific
-   \end{itemize}
+\begin{frame}{Comparison to Other Libraries}
+\begin{table}
+	\input{../extras/table_glmm_current.tex}
+\end{table}
 \end{frame}
@@
 
-\begin{frame}{Thank You!}
+\begin{frame}{Software}
 	The code is available on GitHub: \href{github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}
@@
 \end{frame}
-\begin{frame}{References}
-	\textbf{References}: 
-	\bibliographystyle{plain}
-	\bibliography{bibliography.bib}
-\end{frame}
-
-\end{document}

### 2023-04-06 presentation/pysr3_only_pres.tex
Restructure presentation

--- 2022-07-04 presentation/general_presentation.tex
@@
 
-\section{Relaxed for Linear Mixed-Effect Models}
-\subsection{Linear Mixed-Effects Models}
-\begin{frame}{Linear Mixed-Effect Models}
+% pysr3 part
+\input{pysr3.tex}
 
-Linear Mixed-Effect (LME) models are often used for analyzing combined data across a range of groups. \newline
-
-They use use covariates to separate the population variability (fixed-effects) from the group variability (random effects). \newline
-
-LMEs borrow strength across groups to estimate key statistics in cases when data within units are sparse or highly variable. \newline
-
-\end{frame}
-
-\begin{frame}{Linear Mixed-Effect Models}
-
-	Dataset: $m$ groups $(X_i, Z_i, y_i),\quad i = 1, \dots m$, each has $n_i$ observations
-	\begin{itemize}
-		\item 	$X_i \in \R^{n_i \times p}$ -- group $i$ design matrix for fixed features
-		\item 	$Z_i \in \R^{n_i \times q}$ -- group $i$ design matrix for random effects
-		\item 	$y_i \in \R^{n_i}$ -- group $i$ observations  
-	\end{itemize}
-
-	\begin{columns}[T,onlytextwidth]
-    \column{0.5\textwidth}
-	 \centering Standard Linear Regression:
-   	\begin{figure}
-   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_mean_prediction}
-   	\end{figure}
-   	\[
-   		y = X\beta + \varepsilon, \quad \varepsilon \sim \NN(0, \Lambda) 
-   	\]
-
-    \column{0.5\textwidth}
-    	\centering  Linear Mixed-Effect Model:
-   	\begin{figure}
-   		\includegraphics[width=0.9\textwidth]{Figures/lme_example_random_prediction}
-   	\end{figure}
-   	   		\[
-   		\begin{split}
-   			y_i & = X_i\beta + {\color{red}Z_i u_i} + \varepsilon_i, \quad \varepsilon_i \sim \NN(0, \Lambda_i) \\
-   			{\color{red} u_i} & {\color{red}\sim \NN(0, \Gamma4)}
-   		\end{split}
-   		\]
-
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Notation}
-	\eq{
-   		y_i & = X_i\beta + Z_iu_i + \varepsilon_i \quad i = 1\dots m \\
-   		\varepsilon_i & \sim \NN(0, \Lambda_i) \\
-   		u_i & \sim \NN(0, \Gamma)
-   	}
-
-   	\begin{itemize}
-   		\item $p$ -- number of fixed features, $q$ -- number of random effects.
-   		\item $\beta \in \R^p$ -- fixed effects, or mean effects
-   		\item $u_i \in \R^q$ -- random effects
-   		\item $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects, often $\Gamma = \diag{(\gamma)}$
-   		\item $\varepsilon_i \in \R^{n_i}$ -- observation noise
-   		\item $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for noise
-   	\end{itemize}
-   	Unknowns: $\beta$, $u_i$, $\gamma$, sometimes $\Lambda_i$.
-\end{frame}
-
-\begin{frame}{Likelihood for Mixed Models}
-Negative log-likelihood:
-\eq{
-	\label{eq:lmm_objective}
-	\mathcal{L}(\beta, \gamma) & = \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \\ & + \half\log{\det{\pa{Z_i \Gamma Z_i^T + \Lambda_i}}}, \quad \Gamma = \diag{(\gamma)}
-	}
-Maximum likelihood estimates for $\beta$ and $\gamma$ solve the problem:
-\eq{
-	\mathcal{LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma)
-}
-
-To select covariates we add a sparsity-promoting regularizer $R(\beta, \gamma)$
-\eq{
-	\mathcal{FS-LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
-}
-
-\begin{itemize}
-	\item $\LL(\beta, \gamma)$ is smooth on its domain, quadratic w.r.t. $\beta$ and $\bar\eta$-weakly-convex w.r.t. $\gamma$.
-	\item $R(\beta, \gamma)$ is closed, proper, convex, with easily computed \textit{prox operator}
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}{Regularization}
-\begin{itemize}
-	\item $R(\beta, \gamma)$ is closed, proper, convex, with easily computed \textit{prox operator}
-\end{itemize}
-
-\eq{
-	\prox_{\alpha R + \delta_{\CC}}(\tbeta, \tgamma) & := \argmin_{(\beta, \gamma) \in \CC} R(\beta, \gamma) + \frac{1}{2\alpha}\|(\beta, \gamma) - (\tbeta, \tgamma)\|_2^2, \\ & \text{ where } \CC := \R^p \times R^q_+ \\
-}
-
-Examples: 
-\begin{itemize}
-	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \cite{Bondell2010,Lin2013}
-	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \cite{Vaida2005,Jones2011}
-	\item $R(x)$ -- SCAD penalty (\cite{Fan2001,Fan2012})
-\end{itemize}
-
-\begin{figure}
-     \centering
-     \begin{subfigure}[b]{0.25\textwidth}
-         \centering
-         \includegraphics[width=\textwidth]{Figures/l0_regularizer.pdf}
-         \caption{$\ell_0$}
-     \end{subfigure}%
-     \begin{subfigure}[b]{0.25\textwidth}
-         \centering
-         \includegraphics[width=\textwidth]{Figures/l1_regularizer.pdf}
-         \caption{$\ell_1$}
-     \end{subfigure}%
-     \begin{subfigure}[b]{0.25\textwidth}
-         \centering
-         \includegraphics[width=\textwidth]{Figures/lh_regularizer.pdf}
-         \caption{$\ell_p,\, p=1/2$}
-     \end{subfigure}%
-     \begin{subfigure}[b]{0.25\textwidth}
-         \centering
-         \includegraphics[width=\textwidth]{Figures/scad_regularizer.pdf}
-         \caption{SCAD}
-     \end{subfigure}
-     \caption{Three simple graphs}
-     \label{fig:three graphs}
-\end{figure}
-
-\end{frame}
-
-\begin{frame}{SR3-Relaxation for Mixed-Effect Models}
-Original problem $\mathcal{FS-LME}$:
-\eq{
-	\min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
-}
-Relaxed problem $\ouralgo$:
-\eq{
-	\label{eq:msr3_formulation_explicit}
-	\min_{\beta, \tbeta \in \R^p,\, \gamma, \tgamma \in \R^{q}_+} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma) + R(\tbeta, \tgamma)
-}
-where the \textit{relaxation} $\kappa_\eta$ decouples the likelihood and the regularizer 
-\eq{
-	\kappa_\eta(\beta - \tbeta, \gamma - \tgamma) := \frac{\eta}{2}\|\beta - \tbeta\|^2_2 + \frac{\eta + \bar\eta}{2}\|\gamma - \tgamma\|_2^2
-}
-and the \textit{projection function} $\phi_\mu$ replaces $\gamma \geq 0$ with a log-barrier 
-\eq{
-	\phi_\mu(\gamma) := \begin{cases}
-		-\mu\sum_{i=1}^q \ln(\gamma_i/\mu), & \mu > 0 \\
-		\delta_{\R_+^q}(\gamma), & \mu = 0 \\
-		+\infty, & \mu < 0
-	\end{cases}
-} 
-\end{frame}
-
-\begin{frame}{Value Function Reformulation}
-$\ouralgo$-relaxation replaces the original likelihood $\LL$ with a \textit{value function} $u_{\eta,\mu}$:
-\eq{
-	\label{eq:value_function_definition}
-	u_{\eta,\mu}(\tbeta,\tgamma) & := \min_{(\beta, \gamma)} \LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\  & := \min_{(\beta, \gamma)} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma)
-}
-so $\ouralgo$-formulation~(\ref{eq:msr3_formulation_explicit}) becomes
-\eq{
-	\min_{(\tbeta, \tgamma) \in \CC} u_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
-}
-When $\bar\eta$ is larger than the weak-convexity constant
-\begin{itemize}
-	\item $u_{\eta,\mu}$ is well-defined and continuously differentiable.
-	\item Solutions $(\tbeta^*, \tgamma^*)$ for $\ouralgo$ converge to solutions $(\beta^*, \gamma^*)$ of $\mathcal{FS-LME}$ \\ when $\mu \rightarrow 0$ and $\eta \rightarrow \infty$.
-\end{itemize}
-
-\textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations keep the solution close to the central path.
-\end{frame}
-
-\begin{frame}{Value Function Reformulation}
-\begin{figure}
-	\includegraphics[width=0.7\textwidth]{Figures/intuition_current.pdf}
-	\caption{\label{fig:intuition_sr3} Comparison of the level-sets for the original likelihood (left) and $\ouralgo$-likelihood (right), for fixed (top) and random (bottom) effects.}
-\end{figure}
-\end{frame}
-
-\begin{frame}{Designing an Algorithm}
-	Gradient of a Lagrangian:
-	\eq{
-		G_{\nu,\eta}((\beta,\gamma,v),(\tbeta,\tgamma)) := \begin{bmatrix}
-			\nabla_\beta \LL(\beta, \gamma) + \eta(\beta-\tbeta) \\
-			\nabla_\gamma \LL(\beta, \gamma) + (\bar\eta + \eta)(\gamma-\tgamma) - v\\
-			v \bigodot \gamma - \mu\textbf{1}
-		\end{bmatrix}
-	}
-
-\textbf{Lemma:} For every $(\mu,\eta) \in \R_+\times\R_{++}$,
-\eq{
-	&(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
-	& \iff \\
-	& \exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
-}
-If $\mu > 0$, then $\hat{v} = -\nabla\phi_\mu(\hat{\gamma})$, and if $\mu = 0$, then $\hat{v}$ is the unique KKT multiplier associated with the constraint $0 \leq \gamma$.
-\end{frame}
-
-\begin{frame}{$\ouralgo$-fast Algorithm}
-\input{../extras/08Pseudocode}
-\end{frame}
-
-\section{Experiments}
-\subsection{Application to Synthetic Problems}
-
-\begin{frame}{Application to Synthetic Problems}
-\textbf{The Experiment} 
-\begin{itemize}
-	\item The number of fixed effects $p$ and random effects $q$ is 20.
-	\item $\beta = \gamma = \frac{1}{2}[1,2,3,\dots,10, 0\dots,0]$
-	\item 9 groups with sizes [10, 15, 4, 8, 3, 5, 18, 9, 6]
-	\item $X_i \sim \NN(0, I)^p$, $Z_i = X_i$, $\varepsilon_i \sim \NN(0, 0.3^2I)$
-	\item Each experiment is repeated 100 times.
-	\item Grid-search for $\eta \in [10^{-4}, 10^{2}]$, golden search for $\lambda \in [0, 10^5]$
-	\item Final model is chosen to maximize BIC
-\end{itemize}
-\begin{table}
-	\input{../extras/performance_table_short_current}
-\end{table}
-\end{frame}
-
-\begin{frame}{Application to Synthetic Problems}
-\begin{figure}
-	\includegraphics[width=\textwidth]{Figures/performance_picture_current.pdf}
-\end{figure}
-Benefits:
-\begin{itemize}
-	\item $\ouralgo$-relaxation has similar (and sometimes better!) feature selection performance than the original likelihood.
-	\item $\ouralgo$-fast optimization accelerates the compute time by $\sim 10^2$.
-\end{itemize}
-Setbacks:
-   \begin{itemize}
-   	\item $\eta$ is a new hyperparameter to tune.
-   \end{itemize}
-\end{frame}
-
-\begin{frame}{$\ell_0$-based Covariate Selection for Bullying Study from GBD}
-	\begin{figure}
-		\includegraphics[width=\textwidth]{Figures/bullying_data_assessment_selection.pdf}
-		\caption{Fixed and random covariate selection for Bullying dataset from~\cite{GBD}.
-The model selected 9 covariates, 7 of which were historically significant, and did not select 4 covariates, 1 of which was historically significant.}
-	\end{figure}
-\end{frame}
-
-\begin{frame}{Thank You!}
-	The code is available on GitHub: \\
-	\href{github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}
-	\begin{itemize}
-		\item All estimators are fully compatible to \texttt{sklearn} library.
-		\item Implements SR3 for linear, generalized-linear, and linear mixed-effect models.
-		\item Has tutorials, tests, and documentation.
-	\end{itemize}
-
+\begin{frame}{References}
 	\textbf{References}: 

### 2023-04-10 appendix/lipschitz_constant.tex
Add theory paper to the thesis

Recall that a function $\LL(x)$ is called L-Lipschitz smooth when

\eq{
    \|\nabla \LL(x) - \nabla \LL(y)\|_2 \leq L\|x-y\|_2
}

To find the Lipschitz-constant of the function $\LL_{ML}$ (\ref{eq:lmm_objective}) we will use the fact that $\LL(x)$ is L-Lipschitz if and only if $\|\nabla^2 \LL(x)\| \leq L$ for any $x$. Hence, to upper-bound L we need to upper-bound the norms of Hessians. % from (\ref{eq:all_derivatives}).
Assume that $\|y_i - X_i\beta\| \leq \rho$ where $\rho > 0$. We get % Using the same technique as in (\ref{eq:trick_with_norm}), we get that

\eq{
    \left\|\nabla^2\LL(x)\right\|_2 & = \left\|\begin{bmatrix} \nabla^2_{\beta\beta}\LL(\beta, \gamma) & \nabla^2_{\beta\gamma}\LL(\beta, \gamma) \\ \nabla^2_{\gamma\beta}\LL(\beta, \gamma) & \nabla^2_{\gamma\gamma}\LL(\beta, \gamma) \end{bmatrix} \right\| \leq \sum_{i=1}^m \left\| \begin{bmatrix}\frac{\|X_i\|_2^2}{\|\Lambda_i\|_2} & \frac{\rho\|X_i\|_2\|Z_i\|^2_2}{\|\Lambda_i\|^2} \\ \frac{\rho\|X_i\|_2\|Z_i\|^2_2}{\|\Lambda_i\|^2} & \frac{\rho\|Z_i\|_2^4}{\|\Lambda_i\|_2^3} \end{bmatrix} \right\| \\  & \leq \sum_{i=1}^m\max\left(\frac{\|X_i\|_2^2}{\|\Lambda_i\|_2},  \frac{\rho\|X_i\|_2\|Z_i\|^2_2}{\|\Lambda_i\|^2},  \frac{\rho\|X_i\|_2\|Z_i\|^2_2}{\|\Lambda_i\|^2}, \frac{\rho\|Z_i\|_2^4}{\|\Lambda_i\|_2^3}\right) = L
}

The assumption $\|y_i - X_i\beta\| \leq \rho$ is typically enforced artificially by introducing a box-constraint for $\beta$. Unfortunately, if the assumption is violated then $\nabla\LL$ is not globally Lipschitz on its domain, as noted in \cite{aravkin2022jimtheory}. Nonetheless, it is possible to obtain convergence results with the inclusion of a line search or trust region strategy, see e.g. \cite{Burke-Engle18}.

### 2023-04-10 appendix/proxes.tex
Add theory paper to the thesis

\paragraph{SCAD}

For a scalar variable $x \in \R$, SCAD-regularizer is defined as 
\eq{
r(x) = \begin{cases} 
    \sigma |x|, & |x| \leq \sigma \\
    \frac{-x^2 + 2\rho\sigma |x| - \sigma^2}{2(\rho - 1)}, & \sigma < |x| < \rho\sigma \\ \frac{\sigma^2(\rho + 1)}{2}, & |x| > \rho\sigma 
    \end{cases}
}
To evaluate the $\prox_{\alpha r}$ operator we need to solve the following minimization problem:

\eq{
    \label{eq:prox_separated_minimization}
    \min_{x} r(x) + \frac{1}{2\alpha}(x - z)^2
}

For $\alpha = 1$, the solution was derived by \cite{fan1997comments}. 
Here we extend it  for an arbitrary $\alpha$. %to use it with PGD algorithm, this derivation closely follows its original source; we provide it for the sake of completeness. 
To identify the set of stationary points $\{x^*\}$ of a non-smooth function $f(x)$, we the optimality condition
\eq{
    0 \in \partial_x f(x^*)
}
where $\partial_x f(x)$ denotes a sub-differential set of $f$ at the point $x$. For the prox problem, we get % \ref{eq:prox_separated_minimization}  yields:
\eq{
    0 \in \frac{1}{\alpha}(x^* - z) + \partial r(x)_{x = x^*}
}
Since $r(x)$ is piece-wise defined the precise value of $\partial r(x)_{x = x^*}$ will depend on $x^*$:
\begin{enumerate}
    \item Let $0 < x^* \leq \sigma$, then we have $\partial r(x)_{x = x^*} = \{x^*\}$ and so
    \eq{
        x = z - \sigma\alpha, \quad z \in [\sigma\alpha, \sigma + \sigma\alpha]
    }
    \item Let $-\sigma\alpha \leq x^* < 0$, then we have $\partial r(x)_{x = x^*} = \{-x^*\}$ and so
    \eq{
        x = z + \sigma\alpha, \quad z \in [-\sigma - \sigma\alpha, -\sigma\alpha]
    }
    \item Let $x^* = 0$, then $\partial r(x)_{x = x^*} = [-1, 1]$, which yields
    \eq{
        \frac{1}{\alpha}(x^* - z) \in -\sigma[-1, 1] \quad \Rightarrow \quad      z \in [-\sigma\alpha, \sigma\alpha]
    }
%    or
%%    \eq{
%%        x^* \in z + [-\sigma\alpha, \sigma\alpha]
%%    }
%%    Given that $x^* = 0$, it provides us with the condition on $z$:
%    \eq{
%        z \in [-\sigma\alpha, \sigma\alpha]
%    }
    \item Let $\sigma < x^* < \rho\sigma$, then $r(x)_{x = x^*} = \frac{-{x^*}^2 + 2\rho\sigma x^* - \sigma^2}{2(\rho - 1)}$, which gives us 
    \eq{
        \frac{1}{\alpha}(x^* - z) = \frac{x^* - \rho\sigma}{\rho - 1}
    }
    To ensure that the stationary point is indeed a minimizer, we need to ensure that 
    \eq{ 
        \frac{1}{\alpha} - \frac{1}{\rho - 1} > 0 \quad \Rightarrow \quad \alpha < \rho - 1.
    }
Rearranging the terms to express $x^*$ as a function of $z$ we get % the value of prox minimizer:
    \eq{
        x^* = \frac{(\rho - 1)z - \lambda\rho\sigma}{\rho - 1 - \alpha} \quad \Rightarrow \quad         z \in [\sigma + \alpha\sigma, \rho\sigma]
    }
%    Flipping this expression back to $z$ being a function of $x^*$ we get
%    \eq{
%        z = \frac{x^*(\rho - 1 + \lambda) + \alpha\rho\sigma}{\rho - 1} \in\left[\frac{\sigma(\rho - 1 - \alpha) + \alpha\rho\sigma}{\rho - 1}, \frac{\sigma\rho(\rho - 1 - \alpha) + \alpha\rho\sigma}{\rho - 1} \right]
%    }
%    which, after algebraic simplifications, becomes
%    \eq{
%        z \in [\sigma + \alpha\sigma, \rho\sigma]
%    }
    \item Let $-\rho\sigma < x^* < -\sigma$, then, similarly to the previous case, we get 
    \eq{
        \frac{1}{\alpha}(x^* - z) = \frac{x^*+\rho\sigma}{\rho - 1}
    }
    %where the sign of the second term in the denominator changes due presence of absolute value. 
    Rearranging the terms to express $x$ in terms of $z$ we get:
    \eq{
        x^* = \frac{(\rho - 1)z + \lambda\rho\sigma}{\rho - 1 - \alpha} \quad \Rightarrow \quad       z \in [-\sigma -\alpha\sigma, -\sigma]
    }
%    Rearranging the terms to express $z$ in terms of $x^*$, and using that $-\sigma\rho < x^* < \sigma$ we get \eq{
%        z \in [-\sigma -\alpha\sigma, -\sigma]
%    }
    \item Finally, when $|x^*| \geq \sigma\rho$ we have $\partial r(x)_{x = x^*} = \{0\}$ and so 
    \eq{
        x^* = z, \quad |z| \geq \sigma\rho
    }
    Bundling all six cases together,  we have
    %when $\alpha + 1  < \rho$ and we get 
    \eq{
    \label{eq:scadProx}
        \prox_{\alpha r}(z) = \begin{cases} \sign(z)(|z| - \sigma\alpha)_+, & |z| \leq \sigma(1+\alpha) \\ \frac{(\rho - 1)z - \sign(z)\rho \sigma\alpha}{\rho - 1 - \alpha}, & \sigma(1+\alpha) < |z| \leq \max(\rho,1+\alpha)\sigma \\ z, & |z| > \max(\rho, 1+\alpha)\sigma \end{cases}
    }
The middle branch is active only when $\rho > 1+\alpha$. One special case of this is when $\alpha = 1$, and then~\eqref{eq:scadProx} recovers    the classic result by \cite{Fan2001}.

To get $\prox_{\alpha r + \delta_{\R_+}}(z)$ from $\prox_{\alpha r}(z)$ we only need to notice that (1) the minimizer $x^*$ of 
\eq{
    \min_x r(x) + \delta_{\R_+} + \frac{1}{\alpha}(x - z)^2
}
can never be negative, and that (2) when the minimizer $x^*$ is exactly zero we get:
\eq{
    \frac{1}{\alpha}(x^* - z) \in -\partial(r(x)|_{x = x^*} + \delta_{\R_+}(x)|_{x = x^*})% = -[-\infty, 0] + [-\sigma, \sigma]) = [-\sigma, +\infty]
    \quad \Rightarrow \quad     z \in [-\infty, \sigma\alpha]
}
%which gives us the condition on $z$:
%\eq{
%    z \in [-\infty, \sigma\alpha]
%}
%Similarly, when $\C$ also includes the upper-bound constraint for REML likelihood, the ``largest'' branch $|x^*| \geq \rho$ gets affected. Namely, when $\rho\sigma \leq |x^*| < \bar\gamma$ we still get $\partial r(x)_{x = x^*} = \{0\}$ and so
%\eq{
%    x^* = z,  \sigma\rho \leq \quad |z| < \bar\gamma
%}
%However, in case of $x^* = \bar\gamma$ we get
%\eq{
%    \frac{1}{\alpha}(x^* - z) \in -\partial(r(x)|_{x = x^*} + \delta_{\R_{\leq \bar\gamma}}(x)|_{x = x^*}) = -(\{0\} + [0, +\infty])  = [-\infty, 0]
%}
%which happens when $z \geq \bar\gamma$.

\end{enumerate}

\paragraph{A-LASSO}

A-LASSO regularizer is defined as 
\eq{
    r(x) = w|x|
}
where $w=1/|\hat{x}|$ with $\hat{x}$ the solution of a non-regularized problem (\cite{Zou2006}). 
%For the purposes of deriving the proximal operator of A-LASSO, $w$ is a constant and positive scalar value. 
The derivation of the proximal operator of A-LASSO nearly matches the steps 1, 2, and 3 that of SCAD above. We wish to evaluate
\eq{
    \min_{x} w|x| + \frac{1}{2\alpha}(x - z)^2
}
as a function of $z$.
The sub-differential optimality criterion yields
\eq{
    0 \in \frac{1}{\alpha}(x^* - z) + w\partial |x|
}
\begin{enumerate}
\item Let $0 < x^*$, then we have $\partial r(x)_{x = x^*} = \{x^*\}$ and so
    \eq{
        x^* = z - \alpha w, \quad z > \alpha w
    }
    \item Let $x^* < 0$, then we have $\partial r(x)_{x = x^*} = \{-x^*\}$ and so
    \eq{
        x^* = z + \alpha w, \quad z < -\alpha w
    }
    \item Let $x^* = 0$, then $\partial r(x)_{x = x^*} = [-1, 1]$, which yields
    \eq{
        \frac{1}{\alpha}(x^* - z) \in [-w, w] \quad \Rightarrow \quad      z \in [-\alpha w, \alpha w]
    }
%    or
%    \eq{
%        x^* \in z + [-\alpha w, \alpha w]
%    }
%    Given that $x^* = 0$, it yields the condition on $z$:
%    \eq{
%        z \in [-\alpha w, \alpha w]
%    }
\end{enumerate}
Combining all cases together we get
\eq{
    \prox_{\alpha r}(z) = \sign(z)(|z| - \alpha w)_+
}
Finally, $\prox_{\alpha r + \delta_{\R}}(z)$ can be derived by noticing that, in this case, (1) $x^* \geq 0$, and (2) when $x^* = 0$ the sub-differential changes due to the presence of the delta-function:
\eq{
    x^* = 0 \implies \frac{1}{\alpha}(x^* - z) \in -([-\alpha w, \alpha w] + [-\infty, 0]) = [-\alpha w, +\infty]
}
which gives us the condition 
\eq{
    x^* = z, \quad z \in [-\infty, \alpha w].
}

%Similar situation arises when $\C$ includes the upper-bound box constraint for REML. For $|x^*| = \bar\gamma$ we get that 
%\eq{
%    x^* = 0 \implies \frac{1}{\alpha}(x^* - z) \in -(w + [0, +\infty]) = [-\alpha w, +\infty]
%}
%which gives us the condition 
%\eq{
%    x^* = \bar\gamma, \text{ when } \quad z \geq \bar\gamma + \alpha w
%}

\paragraph{LASSO} LASSO is a particular case of A-LASSO above when $w = 1$.

\paragraph{$\ell_0$-regularizer} Comparing to its counterparts above, the regularizer $R(x) = \delta_{\|x\|\leq k}(x)$ is non-separable. However, the proximal operator of it can still be evaluated analytically:
\eq{
    \left[\prox_{\alpha R}(z)\right]_i = \left[\argmin_{\|x\| \leq k} \frac{1}{2\alpha}\|x - z\|^2\right]_i = \begin{cases}z_i, & i \in \II_{k}\\ 0, & \text{ otherwise}\end{cases}
}
where $\II_k$ is a set of $k$ largest in their absolute value coordinates of $z$. To get $\prox_{\alpha R + \delta_{\R_+}}$ we replace $\II_k$ with a set of $k$ largest positive coordinates of $z$, and set the rest of the coordinates to $0$.

### 2023-04-10 msr3-theory.tex
Add theory paper to the thesis

\subsection{Relaxation of the mixed-effects variable selection model}
\label{sec:relax}

Our strategy for obtaining approximate solutions to the
mixed-effects variable selection problem \eqref{eq:lme_loss_original_in_x} is
motivated by the sparse relaxed regularization regression (SR3) strategy
developed in \cite{Zheng2019SR3}. That is, $0\le\gam$ is relaxed
using the perspective of the negative log, i.e %log-barrier function
$\map{\vphi}{\R^q\times\R}{\R\cup\{\infty\}}$ given by
\[
\vphi(\gam,\mu):=
\begin{cases}
-\mu\sum_{i=1}^q\ln (\gam_i/\mu)&,\ \mu>0,\\
\del_{\R^q_+}(\gam)&,\ \mu=0,\\
+\infty&,\ \mu<0.
\end{cases}
\]
The mapping $\vphi$ is known to be a closed proper convex function and, 
for $\mu>0$, it is
essentially equivalent to the well-known log-barrier function in \eqref{eq:log_barrier_relaxation_for_ip}. 
%since in this case
%\[
%\phi_\mu(\gam)=q\mu\ln(\mu)-\mu\sum_{i=1}^q\ln (\gam_i).
%\]
For more information on the perspective mapping, its calculus,
and perspective duality, we refer the reader to \cite{ABDFM18,ABF13}.
%\red{Should we mention the perspective of $\psi(\gam)=-\ln (\gam)$, or perhaps the closure of the epi-limit? The connection to the perspective map might be useful since
%we get the joint convexity in $\mu$ and $\gam$.}
We call $\eta$ the coupling parameter and $\mu$ the log-barrier parameter
and write $\phimu(\cdot):=\vphi(\cdot,\mu)$.
%The parameter $0\le\blam$ is fixed and is related to the weak convexity of $\LL$
%discussed in Lemma \ref{lem:LL weak cvx}.
%}
The relaxed problem employs 
auxiliary variables $\tbg$ 
and relaxation parameters $0\le\eta$ and $0\le \mu$
to obtain the problem 
\begin{equation}\label{eq:relax} 
\begin{aligned}%\eq{\label{eq:lme_loss_relaxed}
%\nu_\emu:=	
\min_{\bg, \tbg} & \LL(\beta, \gamma) 
	+\phi_\mu(\gam)+\kappa_\eta(\beta-\tbeta,\gam-\tgam)
%\frac{\eta}{2}\norm{\beta - \tbeta}^2 + \frac{\blam+\eta}{2}\norm{\gamma - \tgamma}^2  
+ R(\tbeta, \tgamma) 
\\
\text{s.t. } & \tgamma \geq 0\ .
\end{aligned}
\end{equation}
We rewrite \eqref{eq:relax} %\eqref{eq:lme_loss_relaxed} 
so as to separate the smooth and nonsmooth components to obtain
\eq{ \label{eq:relax2}
    \min_{\bg, \tbg} & \LL_{\emu}(\bg,\tbg) + R\tbg +\del_{\R^q_+}(\tgam)\ ,
%    \\
%    \text{s.t. } & 0\le\tgam,
}
where 
%$x=(\beta, \gamma)$, 
%$w=(\tbeta, \tgamma)$, $\CC=\R^p\times\R^q_+$, and
%$\LL_\eta$ is defined in \eqref{eq:defn L sub eta}.
\eq{\label{eq:LLlam}
    \LL_{\emu}(\bg, \tbg) := 
    \LL\bg + \phimu(\gam)+\keta(\beta - \tbeta, \gamma - \tgamma).
}
Observe that, for all $\mu,\eta\in\R_+$, $\LL_{\emu},\ \nabla \LL_{\emu}$ and $\nabla^2 \LL_{\emu}$ are continuous
on $(\Rp\times\dom(\phimu))\times(\Rp\times\Rq)$ 
(see Appendix \ref{appendix:derivatives_of_lmm}) 
so that $\LL_{\emu}$ is smooth on its domain.
As in \cite{Zheng2019SR3}, we use the 
decoupling to write \eqref{eq:relax2}
as an iterated optimization problem over the smooth components of the objective.
This yields a representation of the form
\begin{equation}
    \label{eq:relax3}
    \min_{\tbg} \uemu\tbg + R\tbg+\del_{\Rqp}(\tgam),
\end{equation}
where %$w=(\tbeta,\tgam)$ and
\begin{equation}
    \label{eq:value_function_definition}
    \uemu\tbg := \min_{\bg } 
    \LL_{\emu}(\bg, \tbg)\ .
%    \hLL(\bg,\tgam)+ %\del_{\Rqp}(\gam)+
%    \frac{\eta}{2}\norm{\bg-\tbg}^2
    %\LL(x) + \lambda\|x - w\|_2^2 
\end{equation}
This is the formulation of the mixed-effects variable selection problem we study.
Our focus is on the optimal value function $\uemu$ which captures global 
variational information about the function $\LL$ over its domain.
We show that $\uemu$ has a locally Lipschitz continuous gradient
and that the evaluation of $\uemu$ and $\nabla \uemu$ is accomplished
by optimizing a well conditioned strongly convex function. This
allows us to apply the PGD algorithm to the function $\uemu$ rather than
the function $\LL$. Our numerical studies show that the global information 
captured by $\uemu$ significantly improves both the accuracy of the solution
obtained and the overall numerical efficiency of the algorithm.
%gives the problem structure used in the design of our
%numerical solution method. 
%Specifically, we design a 
%(PGD) %prox gradient 
%method 
%for solving this problem (see Section \ref{sec:pgd}). For this we must show that the function $u_\emu$
%is continuously differentiable with Lipschitz continuous gradient.

%\section{Prox Gradient Descent}
\subsection{Proximal Gradient Descent for $\uemu\tbg + R\tbg+\del_{\Rqp}(\tgam)$} 
\label{sec:pgd}

We follow the analysis of the PGD algorithm given in \cite[Chapter 10]{AB17}
as it applies to the objective
%The application of the (PGD) %proximal gradient descent 
%algorithm to 
\begin{equation}\label{eq:varphi}
\Phi_\emu\tbg:=\uemu\tbg +\tR\tbg,\ \ \text{where}\ 
\tR\tbg:=R\tbg+\del_{\Rqp}(\tgam).
\end{equation}
%
%We now state the (PGD) %prox gradient descent 
%algorithm for finding a minimum for \eqref{eq:relax3}. 
Since $u_\emu$ is nonconvex, one typically 
applies a line search method to select stepsize. 
However, this is often not required in practice. 
For this reason we state the algorithm with and without a line search.
\medskip

%, as outlined in Algorithms \ref{alg:pgd_for_lme} and \ref{alg:MSR3}. 
%Set $\hR\tbg=R\tbg+\del_{\R^q_+}(\tgam)$ and 
%$\phi\tbg:=u_\emu\tbg+\hR\tbg$.
%See also Appendix 
%\smallskip

%\noindent
%{\bf Algorithm 1:} Proximal Gradient Descent with fixed stepsize (PGD)

\begin{algorithm}[H]
\SetAlgoLined
$\tw = \tw^0$, $\gammax>\tgam^0$, $\alpha < \frac{2}{L}$ 
for $L>0$.\\
 \While{not converged and $\tgam\le\gammax$}{
    $\tw^+ = \prox_{\alpha \tR }(\tw - \alpha \nabla \uemu (\tw))$\\
 }
\caption{\label{alg:MSR3}Proximal Gradient Descent for 
$\Phi_\emu$}
\end{algorithm}
\medskip

%\noindent
%{\bf Algorithm 2:} Proximal Gradient Descent with Backtracking (PGD-B)

\begin{algorithm}[H]
\SetAlgoLined
{\bf Initialize:} 
$\theta\in(0,1),\ \tau\in(0,1)$,
$\eta>0,\ \mu>0$, $\eps_{\mbox{\tiny Tol}}\ge 0$
$k=0$, $t_0>0$, $\bw^{0}=(\tbeta^0,\tgam^0)\in \Rp\times\R^q_+$
with $\inf\Phi_\emu <\Phi_\emu(\bw^0)$, 
$\gammax>\tgam^0$, 
$w^0=\prox_{t_0\tR}(\bw^{0}-t_0\nabla u_\emu(\bw^0))$.
%$, \ \eta_{Tol}\ge\eta_0>0,\ \mu_0\ge \mu_{Tol}>0$.
\\
\While{{\small $\norm{w^k-\bw^k}>\eps_{\mbox{\tiny Tol}}$
and $\gam^k\le\gammax$}}{\small\smallskip
\begin{enumerate}
\item[(i)] $t_{k+1}=\max\lset{t}{
\begin{aligned}&s\in\bW,\, t=t_0\theta^s,\ 
w=\prox_{t\tR}(w^k-t\nabla u_\emu(w^k))\\ &
\phi(w)
\le \phi(w^k)-\tau t\norm{w^k-w}^2
\end{aligned}}$.
\item[(ii)] $\bw^{k+1}=w^{k}$
\item[]$w^{k+1}=\prox_{t_{k+1}\tR}(w^k-t_{k+1}\nabla u_\emu(w^k))$
\item[]$k=k+1$
\end{enumerate}
}
\smallskip
\caption{\label{alg:pgd with bt}Proximal Gradient Descent 
fo $\Phi_\emu$ with Backtracking}
\end{algorithm}
\medskip

In Algorithm \ref{alg:MSR3}, 
the parameter $L$ is assumed to be a global Lipschitz
constant for $\nabla \uemu$. In Section \ref{sec:convergence pgd}, we show that the existence of $L$ is not needed.
In both algorithms we introduce the requirement that $\gam^k\le\gammax$.
While it is possible to include an explicit constraint of this form in the 
optimal variable selection problem \eqref{eq:lme_loss_original_in_x}, 
we do not do so since we assume that $\gammax$ is chosen 
so large that, from a practical perspective, 
the violation of this constraint indicates that the model is 
poorly posed and the algorithm needs to be terminated.
We base our analysis of the convergence properties of
Algorithms 1 and 2 on \cite[Theorem 10.15]{AB17} which makes use
of the following three basic assumptions:
\smallskip

\noindent
{\bf Basic Assumptions for the PGD Algorithm}
\begin{enumerate}
\item[(A)] $\map{\tR}{\Rp\times\Rq}{\eR}$ is a closed proper convex function.
\item[(B)] $\map{\uemu}{\Rp\times\Rq}{\eR}$ is closed and proper, 
$\dom{\uemu}$ is
convex, $\dom{\tR}\subset\intr{\dom{\uemu}}$, and
$\uemu$ is $L_\emu$-smooth over $\intr{\dom{\uemu}}$.
\item[(C)] Problem \eqref{eq:relax3} has an optimal solution with optimal 
value $\Phi_{\mbox{\tiny OPT}}$.
\end{enumerate}

\noindent
We assume that (A) holds. This is not an overly
restrictive assumption since it is satisfied by most of the standard
variable selection regularizers.
We show that (C) holds when $R$ satisfies
an additional coercivity hypothesis (Theorem \ref{thm:relaxed existence}). 
On the other hand, establishing that (B) holds
in a concrete setting such as ours
can be quite difficult. 
In particular, just as with $\LL$, $\uemu$ may fail to be globally Lipschitz.
Validating Assumption (B) as well as developing a technique for circumventing
the need for a global Lipschitz constant for $\nabla\uemu$ 
consumes the majority of the theoretical development. 
%Assumption (B) is particularly difficult since $\uemu$ is an optimal value function.
%We establish $B$ established in several steps.
%However, we show that the global Lipschitz continuity of $\nabla \uemu$
%is not required in practice for the theory developed in 
%\cite{AB17} to apply. 
%We begin by establishing the smoothness of $\uemu$.

%\begin{figure}[H]
%    \centering
%    \includegraphics[width=\textwidth]{Images/intuition_current.pdf}
%    \caption{Proximal Gradient Descent (PGD) takes an order of magnitude fewer iterations to converge when optimizing a value function instead of the original likelihood, because SR3-relaxation improves the conditioning of the problem, which results in a more spherical level-sets. This effect persists for both quadratic convex (first row) and non-convex components (second row) of the likelihood. The effect is stronger for  minima near the boundary of the constraint region 
%    (bottom-left panel).}
%    \label{fig:geometric_intuition_sr3}
%\end{figure}

\subsection{The Smoothness of $u_\emu$}\label{sec:smoothness}
We investigate the relationship between the problems \eqref{eq:lme_loss_original_in_x}
and \eqref{eq:relax2}, the existence of solutions to 
 \eqref{eq:relax2}, and the properties of the function $\uemu$ and its derivative.
%We begin with an examination of the function $\LL$
%and 
%The first step is to understand the role of the parameter $\blam$.

\subsubsection{Underlying convexity}

\begin{lemma}[$\LL+\phimu$ is Weakly Convex]\label{lem:LL weak cvx}
Let $\LL$ be as given in \eqref{eq:lme_gam}. Then
\begin{equation}\label{eq:hess LL}
\nabla^2\LL{(\beta,\gam)}=\sum_{i=1}^m
S_i^T\begin{bmatrix}X_i^T\\ -Z_i^T\end{bmatrix}
\Omega_i(\gam)^{-1}
\begin{bmatrix}X_i& -Z_i\end{bmatrix}S_i
-\begin{bmatrix}
0&0\\ 0& \half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}
\end{bmatrix},
\end{equation}
for all
$(\beta,\gam)\in\R^p\times\R^q_+$, where 
\[
S_i:=\begin{bmatrix}
I_q&0\\ 0&\Diag{Z_i^T\Omega_i^{-1}(X_i\beta-Y_i)}
\end{bmatrix}
\]
and, for any $A\in\R^{t\times t}$, $A^{\circ 2}:=A\circ A$. In particular, this implies
that the matrix
\begin{equation}\label{eq:psd for LL}
\begin{bmatrix}
\nabla_{\beta\beta}\LL(\beta,\gam)&\nabla_{\gam\beta}\LL(\beta,\gam)\\
\nabla_{\beta\gam}\LL(\beta,\gam)&\nabla_{\gam\gam}\LL(\beta,\gam)
+\blam I\end{bmatrix}
\end{equation}
is positive semidefinite
for $\bar\eta= \nu m$, where
\[
\nu:=max\lset{(1/2) \mu_\mmin(\Lam_i)^{-2}\sig^4_\mmax(Z_i)}{i=1,\dots,m},
\]
$\mu_\mmin(\Lam_i)$ is the smallest eigenvalue of $\Lam_i$, and
 $\sig_\mmax(Z_i)$ is the largest singular value of $Z_i,\, i=1,\dots,m$.
Consequently, for any $\tbg\in\dom{\tR}$ and $\mu\ge 0$, the mapping 
\(
\bg\mapsto \LL_{\emu}(\bg, \tbg)
%\hLL(\bg,\tgam) %:=\LL(\beta,\gam)+\frac{\blam}{2}\norm{\gam-\hgam}^2
%\sum_{i=1}^m\frac{\nu_i}{2}\norm{\gam-\hat\gam}^2,
\)
is convex for all $\eta\ge\bar\eta:=\nu m$.
In particular, this implies that $\LL+\phimu$ is weakly convex for any $\mu\ge 0$, 
and the mapping
\(
%\begin{aligned}
\bg\mapsto\LL_{\emu}(\bg,\tbg)
%&=
%\LL\bg+\phimu(\gam)+
%\keta(\beta-\tbeta,\beta-\tbeta)
%\\ &=
%\hLL(\bg,\tgam)+    \frac{\eta}{2}\norm{\bg-\tbg}^2 
%\end{aligned}
\)
is strongly convex for $\eta> \bar\eta$ with modulus of strong convexity $(\eta-\bar\eta)$
regardless of the choice of $\tbg\in\Rp\times\Rq$.
%Moreover, the mapping 
%\eq{\label{eq:defn L sub eta}
%(\beta, \gamma)\mapsto    
%\LL_{\eta}(x, w) := \tLL(x, w) + \frac{\eta}{2}\norm{x-w}^2 
%}
%is strongly convex for every $\eta>0$ and any choice of 
%$(\tbeta,\tgam)\in\R^p\times\R^q$.
\end{lemma}
\begin{proof}
The formula for $\nabla^2\LL$ is given in Appendix \ref{appendix:derivatives_of_lmm}
(see \eqref{eq:hess LL}).
By \cite[Theorem 3.1]{ABBP2021}, 
$\mu_\mmax(\left(Z_i^T\Omega_i(\gam)^{-1}Z\right)\le
\lam_\mmin^{-1}\sig_\mmax^2(Z_i)$, and since 
$\mu_\mmax(H^{\circ 2})\le \mu_\mmax^2(H)$
for all $H\in\bS^q_+$ \cite{HJ85}, we have
\[
\mu_\mmax\left(\half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}\right)
\le (1/2) \lam_\mmin^{-2}
\sig^4_\mmax(Z_i)=:\nu_i
\qquad  i=1,\dots,m.
\]
This establishes that the matrix in \eqref{eq:psd for LL} is positive semidefinite.
Since $\phimu$ is convex, the mapping
\[
(\beta,\gam)\mapsto \LL_{\emu}(\bg,\tbg) %\hLL((\beta,\gam),\tgam)
\]
is strongly convex for any choice
of $\blam> \nu m$, where 
\eq{\label{eq:nu}
\nu:=\max_{i=1,\dots,m}\nu_i.
}
%The strong convexity statement follows from 
%the fact that
%$\nabla^2 \LL_{\eta}((\beta, \gamma), (\tbeta, \tgamma))
%=\nabla^2 \LL(\beta,\gam)+\eta I$ where $\nabla^2 \tLL(\beta,\gam)$
%is positive semidfinite.
\end{proof}
For the remainder of the paper, we assume that 
\eq{\label{eq:blam1}\eta> \nu m=:\bar\eta} 
so that the mapping
$(\beta,\gam)\mapsto \LL_{\emu}(\bg,\tbg)$ %\hLL((\beta,\gam),\tgam)$
is strongly convex with 
%$\nabla_{\bg\bg}^2\hLL(\bg,\tgam)$ 
positive definite Hessian
regardless of the 
choice of $\tgam\in\R^q$. With this in mind, the function $\uemu$ defined by
\eqref{eq:value_function_definition} resembles a Moreau envelope.
%of the convex function $\hLL+\phi_\mu$.
However, this is misleading since, in particular,
we are not even assured of the existence of solutions to
the optimization problem defining $\uemu$.

\subsubsection{Existence and consistency}\label{ssec:existence and consistency}

To establish the existence of solutions to the relaxed
optimization problems \eqref{eq:relax2} and the
problems defining the parametrized family $\uemu$ in
\eqref{eq:value_function_definition}, we assume that $R$ is $1$-coercive.
\begin{lemma}\label{lem:lb plus 1c}
Given $\mu>0$ let $\phimu$ be as defined above, and assume that 
$\map{R}{\Rp\times\Rq}{\R\cup\{+\infty\}}$
is 1-coercive, i.e., 
\(
\liminf_{\Vert{\tbg}\Vert\rightarrow\infty}\Vert{\tbg}\Vert^{-1}R\tbg>0.
\)
Then 
%the objective function in \eqref{eq:relax},
$\phimu+R$ 
%\(
%\wLL_\emu(\bg,\tbg):=\LL_{\emu}(\bg, \tbg)+R\tbg+\del_{R^q_+}(\tgam)
%\),
is level compact.
\end{lemma}
%\begin{proof}
%By Lemma \ref{lem:LL weak cvx}, for every $\tbg \in\Rp\times\Rq$,
%the mapping 
%$\bg\mapsto \LL_{\emu}(\bg, \tbg)$
%is strongly convex with modulus of strong convexity $\eta$.
%Consequently, for every $\tbg \in\Rp\times\Rq$,
%\[
%\begin{aligned}
%\LL_{\emu}((0,\one),\tbg)&+
%\ip{\nabla_{\bg}\LL_{\emu}((0,\one),\tbg)}{\bg-(0,\one)}+
%\frac{\eta}{2}\norm{\bg-(0,\one)}^2
%\\
%&\le \LL_{\emu}(\bg, \tbg)
%\qquad\forall\, \bg\in\Rp\times\R^q_{++},
%\end{aligned}
%\]
%which implies that
%\[\begin{aligned}
%&\LL(0,\one)+q\mu\ln(\mu)+\kappa_\eta(\tbeta,\tgam-\one)
%-\norm{\nabla_\beta \LL(0,\one)}\norm{\beta}-
%\norm{\nabla_\gam \LL(0,\one)}\norm{\gam-\one}-\mu^2\sqrt{q}\norm{\gam-\one}
%\\ &
%-\eta\ip{\tbeta}{\beta}+(\blam+\eta)\ip{\one-\tgam}{\gam-\one}
%+\frac{\eta}{2}\norm{(\beta,\gam)-(0,\one)}^2
%\\ &
%\le \LL_{\emu}(\bg, \tbg)
%\qquad\qquad\forall\, \bg\in\Rp\times\R^q_{++}.
%\end{aligned}
%\]
%That is, there are constants $\xi_0\in\R,\ \xi_1,\xi_2,\xi_3,\xi_4>0$ such that
%\[
%\begin{aligned}
%\xi_0+(\xi_2\norm{\beta}^2-\xi_1\norm{\beta})
%+(\xi_4\norm{\gam-1}^2-\xi_3\norm{\gam-\one})
%-\eta\ip{\tbeta}{\beta-\tbeta}-(\blam+\eta)\ip{\one-\tgam}{\gam-\one}
%\\ 
%\leq\LL_{\emu}(\bg, \tbg)
%\quad\forall\, \bg\in\Rp\times\R^q_{++}.
%\end{aligned}
%\]
%\end{proof}
%
\begin{proof}
If $\mu=0$, then the result is trivially true, so we assume that $\mu>0$.
Let $\{\bgk\}\subset \Rp\times\Rqp$ be such that $\norm{\bgk}\uparrow\infty$.
We need to show that $\phimu(\gam^k)+R\bgk\rightarrow \infty$.
If $\{\gam^k\}$ is bounded, then $\phimu(\gam^k)+R\bgk\rightarrow \infty$
since in this case $\phimu(\gam^k)$ is bounded below.
So assume that $\{\gam^k\}$ is unbounded which implies that
$\phimu(\gam^k)\rightarrow -\infty$. Since $R$ is 1-coercive,
we know that there is an $\hat\alf>0$ such that, for $k$ sufficiently large,
$R\bgk\ge\hat\alf\sum_{i=1}^q\gam^k_i$. But then
$\phimu(\gam^k)+R\bgk\ge \sum_{i=1}^q(\hat\alf\gam^k_i-\mu\ln(\gam^k_i))$
where the right-hand side diverges to $+\infty$ as $k\uparrow\infty$.
Hence, $\phimu(\gam^k)+R\bgk\rightarrow \infty$.
\end{proof}

%We have the following result concerning the existence of solutions to
%\eqref{eq:relax2}.

\begin{theorem}\label{thm:relaxed existence}
Let $\LL$  be as in Theorem \ref{thm:basic existence2}
and let $\eta> 0$ satisfy \eqref{eq:blam1}.
Let $\mu\ge 0$. If $\mu=0$, assume that $\map{R}{\Rp\times\Rq}{\R_+\cup\{+\infty\}}$
is level compact; otherwise, assume $R$
is 1-coercive.
%Then, for every $\tgam\in\Rq$, there exists 
%$(\beta^*_\tgam,\gam^*_\tgam)\in\Rp\times\Rqp$
%such that 
%\[
%\LL(\beta^*_\tgam,\gam^*_\tgam)+\phimu(\gam^*_\tgam)
%\le \LL_{\emu}(\bg,\tbg)\quad\forall\ (\bg,\tbg)\in(\Rp\times\Rqp)\times(\Rp\times\Rq).
%\]
%In addition, suppose $R$ satisfies
%\eq{\label{eq:R condition}
%\left[\eta>0\, \text{and}\ \Xi(\eta,\rho)\!:=\!\lset{(x,w)\in\CC\times\CC}{\LL_\eta(x,w)\le \rho}\ne\emptyset
%\right]\implies 
%\emptyset\ne\argmin_{(x,w)\in \Xi(\eta,\rho)}R(w).
%}
%Furthermore, 
%if $R$ is level compact, t
Then solutions to \eqref{eq:relax2} always exist.
\end{theorem}
\begin{proof}
Let 
\(
v^* %\!:=\!\inf\!\lset{\!\LL_\emu(\bg,\tbg)\!+\!R\tbg\!}{\!0\!\le\!\tgam\!}
\)
be the optimal value in \eqref{eq:relax2}
and let
\(
\!\{(\bgk,\tbgk)\}\!\subset (\Rp\times\Rqp)^2
\) 
be such that
\(
\LL_\emu(\bgk,\tbgk)\!+\!R\tbgk\downarrow v^*\!.
\)
By \eqref{eq:eigbd} and \eqref{eq:eig1}, 
it must be the case that 
\begin{equation}\label{eq:lower bd}
%\max\{\ln(\norm{r(\beta^k}^2/n),\ln\norm{\hOmega(\gam^k)}\} + 
\begin{aligned}
\LL_\emu&(\bgk, \tbgk)+R\tbgk
\\ &\ge
\frac{n\!+\!1}{2}\ln(\talf)+\phimu(\gam^k)+
\keta(\beta^k-\tbeta^k,\gam^k-\tgam^k)+R\tbgk
\\ &\ge
\frac{n\!+\!1}{2}\ln(\talf)+\phimu(\gam^k)+\frac{\bar\eta}{2}\norm{\gam^k-\tgam^k}^2
+R\tbgk.
\end{aligned}
\end{equation}
%where $\bar\eta:=\nu m$.

If $v^*=-\infty$, then \eqref{eq:lower bd} tells us that
\begin{equation}\label{eq:minus infty1}
%\max\{\ln(\norm{r(\beta^k}^2/n),\ln\norm{\hOmega(\gam^k)}\} + 
\phimu(\gam^k)
+\frac{\blam}{2}\norm{\gam^k-\tgam^k}^2+R\tbgk\rightarrow-\infty.
\end{equation}
This in turn implies that $\mu>0$, $\phimu(\gam^k)\rightarrow-\infty$
and $\norm{\gam^k}\rightarrow \infty$.
Since $R$ is 1-coercive and $\norm{\gam^k}\rightarrow \infty$, 
we can assume with no loss in generality that there is an $\balf>0$
such that $R\tbgk\ge \balf \sum_{i=1}^q\tgam^k_i$ for all $k\in\N$. 
Consequently,
\eq{\label{eq:unbdd1}\begin{aligned}
\phimu(\gam^k)
+&\frac{\blam}{2}\norm{\gam^k-\tgam^k}^2+R\tbgk
\\ &\ge
\sum_{i=1}^q\left(-\mu\ln (\gam^k_i/\mu)+\frac{\blam}{2}(\gam^k_i-\tgam^k_i)^2
+\balf\tgam^k_i\right)
\\ &= 
\sum_{i=1}^q\left(
(-\mu\ln (\gam^k_i/\mu)+\balf\gam^k_i)+\frac{\blam}{2}(\gam^k_i-\tgam^k_i)^2
-\balf (\gam^k_i-\tgam^k_i)
\right)
\\ &=
\sum_{i=1}^q\left(
(-\mu\ln (\gam^k_i/\mu)+\balf\gam^k_i)+\frac{\blam}{2}
\left[(\gam^k_i-\tgam^k_i-\frac{\balf}{\blam})^2-(\frac{\balf}{\blam})^2\right]
\right)
\\ &\ge -q\frac{\balf^2}{2\blam}+\sum_{i=1}^q
(-\mu\ln(\gam^k_i/\mu)+\balf\gam^k_i)
\\ &= -q\frac{\balf^2}{2\blam}+\phimu(\gam^k)+\balf\norm{\gam^k}_1\ 
\rightarrow +\infty,
\end{aligned}}
which is a contradiction. Hence $v^*>-\infty$.

Let $\rho> v^*>-\infty$. If $\{\gam^k\}\subset\Rqp$ is unbounded, we may assume with no
loss in generality that $\norm{\gam^k}\rightarrow+\infty$. 
If $\mu=0$, then, by  \eqref{eq:lower bd},
\(
\rho>\frac{n\!+\!1}{2}\ln(\talf)+R\tbgk
\uparrow +\infty,
\)
a contradiction, and so we can assume that $\mu>0$ and $R$ is 1-coercive.
%Lemma \ref{lem:lb plus 1c} tells us that
%$\phimu+R$ is coercive.
Using \eqref{eq:lower bd} we may proceed as in \eqref{eq:unbdd1} to find that
\begin{equation}\label{eq:lbd2}
%\begin{aligned}
\rho > \frac{n\!+\!1}{2}\ln(\talf)
-q\frac{\balf^2}{2\blam}+\sum_{i=1}^q
(-\mu\ln \gam^k_i+\balf\gam^k_i)\ 
\rightarrow +\infty,
\end{equation}
again a contradiction, so the sequence $\{\gam^k\}$ is bounded.
Therefore, the first inequality in \eqref{eq:lower bd} tells us that the entire sequence 
$\{(\bgk,\tbgk)\}$ is necessarily bounded.
Consequently, a limit point of the sequence $\{(\bgk,\tbgk)\}$ exists and,
since $R$ is lsc, any such
limit point is a solution to 
\eqref{eq:relax2}.
\end{proof}

%
%\begin{proof}
%The proof follows the pattern given for Theorem \ref{thm:basic existence}.
%Since 
%Let $\map{A}{(\R^p\times\R^q)\times(\R^p\times\R^q)}
%{\R^n\times\bS^n\times(\R^p\times\R^q)\times(\R^p\times\R^q)}$ be the affine transformation 
%\[
%A((\beta,\gam),(\tbeta,\tgam))=
%(r(\beta),\hOmega(\gam),(\beta-\tbeta,\gam-\tgam),(\tbeta,\tgam)),
%\]
%and let $\map{\hf}{\R^n\times\bS^n_{++}\times(\R^p\times\R^q)\times(\R^p\times\R^q)}{\R}$ be defined by
%\[
%\hf(r,M,(y,z),(\tbeta,\tgam)):=f(r,M)+ \phimu(\gam)+
%\kappa_\eta(y,z) %\frac{\eta}{2}\norm{y}^2 + \frac{\blam+\eta}{2}\norm{z}^2 
%+R(\tbeta,\tgam).   
%\]
%%where $\map{\kappa_\eta}{\R^p\times\R^q}{\R}$ is the coercive mapping
%%\begin{equation}\label{eq:kappa}
%%\kappa_\eta(y,z):=\frac{\eta}{2}\norm{y}^2 + \frac{\blam+\eta}{2}\norm{z}^2.
%%\end{equation}
%Then
%\[
%\LL_{\emu}((\beta,\gam),(\tbeta,\tgam)) + R(\tbeta,\tgam)
%=\hf(A((\beta,\gam),(\tbeta,\tgam))).
%\]
%Let 
%\(
%\rho>v^*:=\inf\lset{\LL_\emu(\bg,\tbg)+R\tbg}{0\le\tgam}
%\) 
%and set
%\(
%\MM_\rho:=\lev{\LL_\eta+R}{\rho}\cap(\Rp\times\Rq\times\Rp\times\R^q_+).
%%\lset{(x,w)\in\CC\times\CC}{\LL_\eta(x,w)+R(w)\le \rho}.
%\)
%Since $\LL(x)\le \LL_\eta(x,w)\le \rho$, $\kappa_\eta(y,z)\le\rho$ and
%$R(w)\le\rho$ for every $(x,w)\in\MM_\rho$,
%\[
%\FF:=
%\lset{A(x,w)}{(x,w)\in\CC^2\cap\MM_\rho}\subset\hDD_{\rho\talf\eta}:=
%\DD_{\rho\talf}\times\lev{f}{\rho}\times\lev{R}{\rho},
%%\lset{(r(\beta),\hOmega(\gam)}}{\exists\, w\in\CC\ \text{s.t.}\ 
%%((\beta,\gam),w)\in\MM_\rho}\subset\DD_{\rho\talf},
%\]
%where $\talf:=\mu_\mmin(\Lam)$ and 
%$\hDD_{\rho\talf\eta}$ is compact by
%Theorem \ref{lem:levelcompact1} and our coercivity assumption on $R$. 
%By taking $R\equiv0$ in
%Corollary \ref{cor:basic existence}, there exists $x^*\in\CC$ %$x^*=(\beta^*,\gam^*)$
%such that $\LL(x^*)=\min_{x\in\CC}\LL(x)$, and so $\LL(x^*)\le v^*$. 
%Let $\{(x^k,w^k)\}\subset\CC^2$ be such that 
%$\LL_{\eta}(x^k,w^k) + R(w^k)\downarrow v^*$.  
%Since $\{A(x^k,w^k)\}\subset\hDD_{\rho\talf\eta}$, we may assume with no loss in 
%generality that $A(x^k,w^k)\rightarrow(\br,\bOmega,(\by,\bz),\hw)$
%for some 
%$(\br,\bOmega,(\by,\bz),\hw)\in(\R^n\times\bS_+^n\times(\R^p\times\R^q)^2)
%\cap\im{A}$. Since $(\tbeta^k,\tgam^k)=w^k\rightarrow\hw=(\hat\beta,\hat\gam)$,
%we have $(\beta^k,\gam^k)\rightarrow(\by+\hat\beta,\bz+\hat\gam)
%=:(\bar\beta,\bar\gam)$ and
%$(r(\beta^k),\hOmega(\gam^k))\rightarrow
%(r(\bar\beta),\hOmega(\bar\gam))$. Therefore, 
%$\LL_{\eta}((\bar\beta,\bar\gam),(\hat\beta,\hat\gam)) + R(\hat\beta,\hat\gam)
%=v^*$ since $\LL_\eta+R$ is continuous on $\MM_\rho$.
%\end{proof}

Next we fix $\mu\ge 0$ and show that as 
$\eta\uparrow\infty$ the solutions to 
\eqref{eq:relax2} converge to solutions of
\begin{equation}\label{eq:log-barrier problem}
\min_{\bg\in\CC}\LL\bg+\phimu(\gam)+R\bg.
\end{equation}
In particular, for $\mu=0$, they converge to solutions of 
\eqref{eq:vs_lme}.

\begin{theorem}[Consistency as $\eta\rightarrow\infty$]\label{thm:eta consistency}
Let $\LL$ and $R$ be as in Theorem \ref{thm:relaxed existence}
and fix $\mu\ge 0$.
Let $\{\eta_k\}\subset\R_{++}$ be such that $\eta_k<\eta_{k+1}$
with $\eta_k\uparrow\infty$, and let $(\bgk,\tbgk)$ be an optimal solution to 
\eqref{eq:relax2} for $(\emu)=(\eta_k,\mu)$, $k\in\N$.
Then any limit point (equivalently, cluster point) 
$(\bbg,\hbg)$ of $\{(\bgk,\tbgk)\}$ satisfies
$\bbg=\hbg$ with $\bbg$ being an optimal solution to 
\eqref{eq:log-barrier problem}.
\end{theorem}
\begin{proof}
With no loss in generality $\eta_k>\bar\eta$ for all $k$.
Set 
\[\left.\begin{aligned}
a_k(x,w)&:=\LL_{\eta_k,\mu}(x,w)+R(w)\\ 
b_k(x,w)&:=\LL(x)+\phimu(\gam)+R(w)\\
c_k(x,w)&:=\kappa_{\eta_k}(\beta-\tbeta,\gam-\tgam)
\end{aligned}\right\}\quad \forall k\in\N,
\] 
where 
$x=(\beta,\gam)$ and $w=(\tbeta,\tgam)$ with $\kappa_\eta$ defined in
\eqref{eq:kappa}. Set $x^k=\bgk,\ \bx=\bbg,\ w^k=\tbgk$ and $\hw=\hbg$.
By Lemma \ref{lem:lb plus 1c} and Theorem \ref{thm:basic existence2} 
with $\hR=\phimu+R$, 
%\ref{thm:relaxed existence}, 
there is an optimal solution $\xmu$ 
to \eqref{eq:log-barrier problem} %\eqref{eq:vs_lme} 
yielding an optimal value of $\vmu$ for which 
$a_k(x^k,w^k)\le a_k(\xmu,\xmu)=\vmu$ for all $k\in\N$. Hence, the sequence
$\{a_k(x^k,w^k)\}$ is upper bounded by $\vmu$.
Since
\[
a_k(x^k,w^k)\le a_k(x^{k+1},w^{k+1})\le a_{k+1}(x^{k+1},w^{k+1}),
\]
there exists $\tv$ such that
$a_k(x^k,w^k)\uparrow\tv\le \vmu$.
Next, observe that
\[
a_k(x^k,w^k) \le a_k(x^{k+1},w^{k+1})\ \ \text{ and }\ \ 
a_{k+1}(x^{k+1},w^{k+1})\le a_{k+1}(x^{k},w^{k}).
\]
By adding these inequalities together we find that 
$\norm{x^{k+1}-w^{k+1}}\le \norm{x^{k}-w^{k}}$ so that 
$\norm{x^{k}-w^{k}}\downarrow\tkappa$ for some $\tkappa\ge 0$.
We also have
\[\begin{aligned}
b_k(x^k,w^k)+(\eta_k/2)\norm{x^{k}-w^{k}}
&=a_k(x^k,w^k)
\\ &\le a_k(x^{k+1},w^{k+1})
\\ &=b_{k+1}(x^{k+1},w^{k+1})+(\eta_k/2)\norm{x^{k+1}-w^{k+1}}
\\ &\le b_{k+1}(x^{k+1},w^{k+1})+(\eta_k/2)\norm{x^{k}-w^{k}},
\end{aligned}\]
which gives $b_k(x^k,w^k)\le b_{k+1}(x^{k+1},w^{k+1})\le \tv$.
Therefore, $b_k(x^k,w^k)\uparrow \hv$ for some $\hv\le\tv$.
Consequently,
\[
\tkappa=\lim_k\norm{x^{k}-w^{k}}=
\lim_k\eta_k^{-1}[a_k(x^k,w^k)-b_k(x^k,w^k)]=0.
\]
Therefore, if $(\bx,\bw)$ is any limit point of the sequence $\{(x^k,w^k)\}$,
then $\bx=\bw$ and $\LL(\bx)+\phimu(\bgam)+R(\bx)=\vmu$ 
since $\LL(x^k)+\phimu(\gam^k)+R(w^k)\le a_k(x^k,w^k)\le \vmu$
for all $k\in\N$.
\end{proof}
We now pair Theorem \ref{thm:eta consistency} with a consistency result for the barrier parameter $\mu$.

\begin{theorem}[Consistency as $\mu\rightarrow 0$]
\label{thm:mu consistency}
Let $\LL$ and $R$ be as in Theorem \ref{thm:relaxed existence}. 
For every $\mu\ge 0$, problem 
\eqref{eq:log-barrier problem} has a solution $(\beta_\mu,\gam_\mu)$.
Moreover, if $\{\mu_k\}\subset\R_{++}$ is such that $\mu_k\downarrow 0$,
then the sequence $\{(\beta_{\mu_k},\gam_{\mu_k})\}$ is bounded
and every limit point of the sequence %$\{(\beta_\mu,\gam_\mu):\ \mu>0\}$
%$(\bbeta,\bgam)$ 
%as $\mu\rightarrow 0$ 
is a 
solution to \eqref{eq:lme_loss_original_in_x}.
\end{theorem}
\begin{proof}
The existence of $(\beta_\mu,\gam_\mu)$ for all $\mu\ge 0$ 
follows immediately from 
Lemma \ref{lem:lb plus 1c} and Theorem \ref{thm:basic existence2} with
$\hR=R+\phimu$. 
%Let $(\bbeta,\bgam)$ be a 
%limit point of $\{(\beta_\mu,\gam_\mu):\ \mu>0\}$
%as $\mu\rightarrow 0$. 
%That is, there is a sequence 
Let $\mu_k\downarrow 0$
and set $(\beta^k,\gam^k):=(\beta_{\mu_k},\gam_{\mu_k})$.
%\rightarrow (\bbeta,\bgam)$.
Set $\tLL:=\LL+R+\del_{\Rp\times\Rqp}$ so that 
the objective in \eqref{eq:log-barrier problem} is $\tLL+\phimu$
and the objective in \eqref{eq:vs_lme} is $\tLL$
with $(\beta_0,\gam_0)$ a solution to \eqref{eq:lme_loss_original_in_x}
by definition. 
Observe that
\[\begin{aligned}
\tLL\bgk+\phi_{\mu_k}(\gam^k)&\le 
\tLL(\beta^{k+1},\gam^{k+1})+\phi_{\mu_k}(\gam^{k+1})\quad \text{and}
\\
\tLL(\beta^{k+1},\gam^{k+1})+\phi_{\mu_{k+1}}(\gam^{k+1})&\le 
\tLL(\beta^{k},\gam^{k})+\phi_{\mu_{k+1}}(\gam^k)
\end{aligned}\]
Summing these inequalities yields the inquality
\[
(\mu_k-\mu_{k+1})\sum_{i=1}^q\ln(\gam^{k}_i)
\ge 
(\mu_k-\mu_{k+1})\sum_{i=1}^q\ln(\gam^{k+1}_i),
\]
so $\{\sum_{i=1}^q\ln(\gam^{k}_i)\}$ is a non-increasing sequence.
Therefore,
\[\begin{aligned}
\tLL(\beta^{k+1},\gam^{k+1})+\phi_{\mu_{k+1}}(\gam^{k+1})&\le 
\tLL(\beta^{k},\gam^{k})+\phi_{\mu_{k+1}}(\gam^k)
\\ &\le
\tLL(\beta^{k},\gam^{k})+\phi_{\mu_{k+1}}(\gam^{k+1})
\end{aligned}\]
which implies that $\{\tLL(\beta^{k},\gam^{k})\}$
is also a non-increasing sequence and bounded below by 
$\tLL(\beta_0,\gam_0)$. 
Since
Theorem \ref{thm:basic existence2} tells us that $\tLL$ is level compact,
the sequence $\{\bgk\}$ is bounded. 
Let $(\bbeta,\bgam)\in\Rp\times\Rqp$ be any
limit point of $\{\bgk\}$ and let $J\subset\N$ be such that
$\bgk\overset{J}{\rightarrow}(\bbeta,\bgam)$. Then
\[
\tLL\bgk+\phi_{\mu_k}(\gam^k)\le
\tLL\bg+\phi_{\mu_k}(\gam) \quad \forall\,\bg\in\Rp\times\R^q_{++}.
\]
Since $\tLL$ is continuous on $\Rp\times\Rqp$ and the perspective function
$\phi_{\mu}(\gam)=\varphi(\mu,\gam)$ is lsc on $\Rp\times\Rqp$, we have
\[
\tLL(\bbeta,\bgam)\le\liminf_{k\in J}(\tLL\bgk+\phi_{\mu_k}(\gam^k))\le
\tLL\bg \quad \forall\,\bg\in\Rp\times\R^q_{++}.
\]
Consequently, the continuity of $\tLL$ on $\Rp\times\Rqp$ implies
that $(\bbeta,\bgam)$ solves \eqref{eq:vs_lme}.
\end{proof}

%The auxiliary variables allows us to write \eqref{eq:relax2}
%as an iterated optimization problem of the form
%\begin{equation}
%    \label{eq:relax3}
%    \min_{w} u(w) + R(w)+\del_{\CC}(w),
%\end{equation}
%where $w=(\tbeta,\tgam)$ and
%\begin{equation}
%    \label{eq:value_function_definition}
%    u(\tbeta,\tgam) := \min_{(\beta,\gam) } 
%    \hLL((\beta,\gam),\tgam)+\del_\CC(\beta,\gam)+
%    \frac{\eta}{2}\norm{(\beta,\gam)-(\tbeta,\tgam)}^2
%%    \LL_\lambda(x, w).
%    %\LL(x) + \lambda\|x - w\|_2^2 
%\end{equation}
%with %where $x=(\beta,\gam)$, $w=(\tbeta,\tgam)$, and 
%\[
%\hLL((\beta,\gam),\tgam):=\LL(\beta,\gam)+\frac{\blam}{2}\norm{\gam-\tgam}^2.
%\]
%\blue{
%We also consider a family of log-barrier relaxations to the problem 
%\eqref{eq:relax3}. For this we define
%$\map{\phi_\mu}{\R^q}{\R\cup\{\infty\}}$ by
%\[
%\phi_\mu(\gam):=
%\begin{cases}
%\del_{\R^q_+}(\gam)&,\ \mu=0,\\
%-\mu\sum_{i=1}^q\ln \gam_i&,\ \mu>0.
%\end{cases}
%\]
%Observe that $\del_\CC(\beta,\gam)=\phi_0(\gam)$.
%\red{mention epi-convergence}
%The log-barrier relaxations to \eqref{eq:relax3}
%are obtained by replacing $u$ with
%\begin{equation}\label{eq:barrier u}
%    u_\mu(\tbeta,\tgam) := \min_{(\beta,\gam) } 
%    \hLL((\beta,\gam),\tgam)+\phi_mu(\gam)+
%    \frac{\eta}{2}\norm{(\beta,\gam)-(\tbeta,\tgam)}^2,
%\end{equation}
%where $\mu>0$ is the barrier parameter.
%In order to facilitate our discussion of the family of relaxed problems
%related to $u_\mu$, we  
%refer to the problem \eqref{eq:relax3}
%with $u$ replaced by $u_\mu$ as problem 
%\eqref{eq:relax3}$_\mu$, and write
%$u_0:=u$ and
%\eqref{eq:relax3}$_0$:=\eqref{eq:relax3}. 
%}

\subsubsection{The continuity and differentiability of $\uemu$}

%The problems in \eqref{eq:relax3} all 
%have the same form as the unrelaxed problem  
%\eqref{eq:vs_lme} 
%with the optimal value function $\uemu$ replacing the likelihood $\LL$.
%We propose applying the (PGD) %Proximal Gradient Descent 
%algorithm 
%to \eqref{eq:relax3}.
%%to these relaxations of problem 
%%\eqref{eq:log-barrier problem}. %\eqref{eq:relax2}.
%The first step is to
%%in this program we 
%examine 
The continuity of $\uemu$ is closely tied to the continuity of 
the associated solution mapping
$\map{\SS_\emu}{\Rp\times\Rq}{\Rp\times\dom(\phimu)}$ given by 
\begin{equation}\label{eq:argmin for u}
\SS_\emu(\tbeta,\tgam):= 
\argmin_{(\beta,\gam) } \LL_{\emu}(\bg,\tbg)\ .
%    \hLL((\beta,\gam),\tgam)+    
%    \frac{\eta}{2}\norm{(\beta,\gam)-(\tbeta,\tgam)}^2.
\end{equation}

%
%We begin by examining the function $\LL$.
%
%From the optimization perspective, $\LL$ has a very useful structural
%property. Specifically, $\LL$
%is weakly convex, that is, there exists $\lam>0$ such that
%$\LL(\beta,\lam)+\frac{\lam}{2}\norm{(\beta,\lam)}^2$ is convex.
%%The proof is given in Appendix \ref{}.
%
%\begin{lemma}[$\LL$ is Weakly Convex]\label{lem:LL weak cvx}
%Let $\LL$ be as given in \eqref{eq:lme_gam}. Then
%\begin{equation}\label{eq:hess LL}
%\nabla^2\LL{(\beta,\gam)}=\sum_{i=1}^m
%S_i^T\begin{bmatrix}X_i^T\\ -Z_i^T\end{bmatrix}
%\Omega_i(\gam)^{-1}
%\begin{bmatrix}X_i^T& -Z_i^T\end{bmatrix}S_i
%-\begin{bmatrix}
%0&0\\ 0& \half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}
%\end{bmatrix},
%\end{equation}
%for all
%$(\beta,\gam)\in\R^p\times\R^q_+$, where 
%\[
%S_i:=\begin{bmatrix}
%I_q&0\\ 0&\Diag{Z_i^T\Omega_i^{-1}(X_i\beta-Y_i)}
%\end{bmatrix}
%\]
%and, for any $A\in\R^{t\times t}$, $A^{\circ 2}:=A\circ A$. In particular, this implies
%that, for any $\tgam\in\R^q_+$, the mapping 
%\(
%x\mapsto\hLL(x,\tgam) %:=\LL(\beta,\gam)+\frac{\blam}{2}\norm{\gam-\hgam}^2
%%\sum_{i=1}^m\frac{\nu_i}{2}\norm{\gam-\hat\gam}^2,
%\)
%is convex for $\blam\ge \nu m$, where
%\[
%\nu:=max\lset{(1/2) \mu_\mmin(\Lam_i)^{-2}\sig^4_\mmax(Z_i)}{i=1,\dots,m}
%\]
%and $\sig_\mmax(Z_i)$ is the largest singular value of $Z_i,\, i=1,\dots,m$.
%%Moreover, the mapping 
%%\eq{\label{eq:defn L sub eta}
%%(\beta, \gamma)\mapsto    
%%\LL_{\eta}(x, w) := \tLL(x, w) + \frac{\eta}{2}\norm{x-w}^2 
%%}
%%is strongly convex for every $\eta>0$ and any choice of 
%%$(\tbeta,\tgam)\in\R^p\times\R^q$.
%\end{lemma}
%\begin{proof}
%The formula for $\nabla^2\LL$ is given in Appendix \ref{appendix:derivatives_of_lmm}
%(see \eqref{eq:hess LL}).
%By \cite[Theorem 3.1]{ABBP2021}, 
%$\mu_\mmax(\left(Z_i^T\Omega_i(\gam)^{-1}Z\right)\le
%\lam_\mmin^{-1}\sig_\mmax^2(Z_i)$, and since 
%$\mu_\mmax(H^{\circ 2})\le \mu_\mmax^2(H)$
%for all $H\in\bS^q_+$ \cite{HJ85}, we have
%\[
%\mu_\mmax\left(\half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}\right)
%\le (1/2) \lam_\mmin^{-2}
%\sig^4_\mmax(Z_i)=:\nu_i
%\qquad  i=1,\dots,m.
%\]
%Consequently, the mapping
%\[
%(\beta,\gam)\mapsto\hLL((\beta,\gam),\tgam)
%\]
%is a convex function for any choice
%of $\tgam\in\R^q_+$, where 
%\eq{\label{eq:nu}
%\nu:=\max_{i=1,\dots,m}\nu_i.
%}
%%The strong convexity statement follows from 
%%the fact that
%%$\nabla^2 \LL_{\eta}((\beta, \gamma), (\tbeta, \tgamma))
%%=\nabla^2 \LL(\beta,\gam)+\eta I$ where $\nabla^2 \tLL(\beta,\gam)$
%%is positive semidfinite.
%\end{proof}
%
%Henceforth we assume that $\blam\ge \nu m$ so that the mapping
%$(\beta,\gam)\mapsto\hLL((\beta,\gam),\tgam)$ is convex 
%with $\nabla_{xx}^2\hLL(x,\tgam)$ positive semi-definite
%regardless of the 
%choice of $\tgam\in\R^q$. With this in mind, the function $u_\mu$ defined by
%\eqref{eq:value_function_definition} resembles the Moreau envelope
%of the convex function $\hLL+\phi_\mu$.
%However, this is not quite correct since $\hLL$ is also a function of $\tgam$.
%We begin with the following result.

%\begin{theorem}[$u$ is Convex]\label{thm:u cvx}
%The function $u$ as defined in \eqref{eq:value_function_definition}
%is well-defined and convex on $\Rp\times\Rq$.
%\end{theorem}
%\begin{proof}
%Since the optimization problem in \eqref{eq:value_function_definition}
%has a strongly convex objective and non-empty feasible region, 
%this problem has a unique solution. Therefore, $u$ is well defined.
%To see that $u$ is convex on $\Rp\times\Rq$, let $\tau\in[0,1]$
%and $(\tbeta_1,\tgam_1),\, (\tbeta_2,\tgam_2)\in\CC$, and set
%$(\tbeta_\tau,\tgam_\tau):=(1-\tau)(\tbeta_1,\tgam_1)+\tau(\tbeta_2,\tgam_2)$.
%Since $\CC$ is convex, we have
%\[
%\begin{aligned}
%u(\tbeta_\tau,\tgam_\tau)&=
%\min_{(\beta,\gam) } 
%\hLL((\beta,\gam),\tgam_\tau)+\del_\CC(\beta,\gam)+
%\frac{\eta}{2}\norm{(\beta,\gam)-(\tbeta_\tau,\tgam_\tau)}^2
%\\
%&=
%\min_{(\beta_1,\gam_1), (\beta_2,\gam_2)} 
%\hLL((1-\tau)(\beta_1,\gam_1)+\tau(\beta_2,\gam_2),\tgam_\tau)+
%\del_\CC(\beta_1,\gam_1)+\del_\CC(\beta_2,\gam_2)
%\frac{\eta}{2}\norm{(1-\tau)(\beta_1,\gam_1)+\tau(\beta_2,\gam_2)-(\tbeta_\tau,\tgam_\tau)}^2
%\
%&\le
%\min_{(\beta_1,\gam_1), (\beta_2,\gam_2)}
%
%\end{aligned}
%\]
%\end{proof}

%%START HERE

\begin{theorem}[Continuity of $\uemu$ and $\SS_\emu$]\label{thm:u cont}
Let the assumptions of Theorem \ref{thm:relaxed existence} hold.
For every $(\mu,\eta)\in\R_+\times\R_{++}$,
the function $\uemu$ defined in \eqref{eq:value_function_definition}
is well-defined and continuous on $\Rp\times\Rq$.
In addition, the solution mapping $\SS_\emu$
%for $\CC:=\Rp\times\Rqp$, 
%the mapping $\map{\SS_\emu}{\Rp\times\Rq}{\Rp\times\dom(\phimu)}$ 
%given by
%\[
%\SS_\emu(\tbeta,\tgam):= 
%\argmin_{(\beta,\gam) } 
%    \hLL((\beta,\gam),\tgam)+    
%    \frac{\eta}{2}\norm{(\beta,\gam)-(\tbeta,\tgam)}^2
%\]
is well-defined, single-valued and continuous on $\Rp\times\Rq$.
\end{theorem}

\begin{proof}
Since $\eta>\blam= \nu m$, Lemma \ref{lem:LL weak cvx} 
tells us that the objective in \eqref{eq:value_function_definition}
is strongly convex, 
and so \eqref{eq:value_function_definition} has a unique solution. 
Consequently, $\SS_\emu$ is well-defined and single-valued
on $\Rp\times\Rq$.
This implies that $\uemu$ is also well defined on $\Rp\times\Rq$ since
\[
\uemu(\tbeta,\tgam)= \LL_{\emu}(\SS_\emu(\tbeta,\tgam),\tbg)
%\hLL(\SS_\emu(\tbeta,\tgam),\tgam)+ %\del_\CC(\SS(\tbeta,\tgam))+
%    \frac{\eta}{2}\norm{\SS_\emu(\tbeta,\tgam)-(\tbeta,\tgam)}^2
    \quad\forall\, (\tbeta,\tgam)\in \Rp\times\Rq.
\]
%Therefore, $u$ and $\SS$ are 
%well-defined with $\SS$ being single-valued on $\Rp\times\Rq$.
The result follows once it is shown that $\SS_\emu$ is continuous.

Let $\{(\tbeta^k,\tgam^k)\}\subset\Rp\times\Rq$ and 
$(\tbeta^*,\tgam^*)\in\Rp\times\Rq$ be such that 
$(\tbeta^k,\tgam^k)\rightarrow(\tbeta^*,\tgam^*)$.  
Set $(\hbeta^k,\hgam^k):=\SS_\emu(\tbeta^k,\tgam^k),\ k\in\N$
and $(\bbeta,\bgam)=\SS_\emu(\tbeta^*,\tgam^*)$. 
We must show that $(\hbeta^k,\hgam^k)\rightarrow (\bbeta,\bgam)$.
We begin by showing that the sequence $\{(\hbeta^k,\hgam^k)\}$
is bounded.
By Lemma \ref{lem:LL weak cvx}, the mapping $\bg\mapsto \LL_{\emu}(\bg,\tbg)$ is strongly convex with modulus of strong convexity $\eta$ for all $\tbg\in\Rp\times\Rq$.
In particular, this implies that
\eq{\label{eq:bd limit}\begin{aligned}
\LL_{\emu}((\bbeta,\bgam),\tbgk)&+
\ip{\nabla_{\bg} \LL_{\emu}((\bbeta,\bgam),\tbgk)}
{(\hbeta^k,\hgam^k)-(\bbeta,\bgam)}
\\ &+
\frac{\eta}{2}\norm{(\hbeta^k,\hgam^k)-(\bbeta,\bgam)}^2
\\ &\le
\LL_{\emu}((\hbeta^k,\hgam^k),\tbgk)
\\ &\le
\LL_{\emu}((\bbeta,\bgam),\tbgk).
%\rightarrow \LL_{\emu}((\bbeta,\bgam),(\tbeta^*,\tgam^*)).
\end{aligned}}
Since $(\tbeta^k,\tgam^k)\rightarrow(\tbeta^*,\tgam^*)$ and both
$\nabla_{\bg} \LL_{\emu}((\bbeta,\bgam),\cdot)$ and 
$\LL_{\emu}((\bbeta,\bgam),\cdot)$ are continuous at 
$(\tbeta^*,\tgam^*)$, we can assume with no loss in generality that there 
is a constant 
$c>0$ such that
\[
\norm{\nabla_{\bg} \LL_{\emu}((\bbeta,\bgam),\tbgk)}\le c\text{ and }
|\LL_{\emu}((\bbeta,\bgam),\tbgk)|\le c\quad\forall\, k\in \N.
\]
Plugging this into \eqref{eq:bd limit} and simplifying gives
\[
\frac{\eta}{2}\norm{(\hbeta^k,\hgam^k)-(\bbeta,\bgam)}^2
\le c(1+\norm{(\hbeta^k,\hgam^k)-(\bbeta,\bgam)}).
\]
Therefore the sequence $\{(\hbeta^k,\hgam^k)\}$ must be bounded.

%\vskip .5in
%
%By taking $\hR\bg=\phimu(\gam)+\frac{\blam}{2}\norm{\gam-\tgam^*}^2$
%in Theorem \ref{thm:basic existence2}, we find that there exists 
%$(\beta^*,\gam^*)\in\Rp\times\Rqp$ %$x^*=(\beta^*,\gam^*)$
%such that 
%$\hLL((\beta^*,\gam^*),\tgam^*)=\min_{\bg}\hLL(\bg,\tgam^*)$.
%Set
%\[
%h((\beta,\gam),(\tbeta,\tgam)):=
%\hLL((\beta,\gam),\tgam)+ %\del_\CC(\beta,\gam)+
%    \frac{\eta}{2}\norm{(\beta,\gam)-(\tbeta,\tgam)}^2.
%\]
%%where $\hLL$ is defined in \eqref{eq:L plus phimu plus blam}.
%Then, for all $k\in\N$,
%\begin{equation}\label{eq:T cont1}
%h((\hbeta^k,\hgam^k),(\tbeta^k,\tgam^k))\le 
%h((\bbeta,\bgam),(\tbeta^k,\tgam^k))\quad \forall\, k\in\N
%\end{equation}
%and
%\begin{equation}\label{eq:T cont2}
%\begin{aligned}
%\hLL((\beta^*,\gam^*),\tgam^*)
%+ \frac{\eta}{2}\norm{\hbeta^k - \tbeta^k}^2 + 
%\frac{\eta}{2}\norm{\hgam^k - \tgam^k}^2
%\le h((\hbeta^k,\hgam^k),(\tbeta^k,\tgam^k))
%\\ \le h((\beta^*,\gam^*),(\tbeta^k,\tgam^k))
%=
%\hLL((\beta^*,\gam^*),\tgam^k)
%+ \frac{\eta}{2}\norm{\beta^* - \tbeta^k}^2 + 
%\frac{\eta}{2}\norm{\gam^* - \tgam^k}^2
%\end{aligned}
%\end{equation}
%where
%\[
%h((\beta,\gam),(\tbeta,\tgam)):=
%\hLL((\beta,\gam),\tgam)+\del_\CC(\beta,\gam)+
%    \frac{\eta}{2}\norm{(\beta,\gam)-(\tbeta,\tgam)}^2.
%\]
%The inequality \eqref{eq:T cont2} tells us that
%\[
%\frac{\eta}{2}\norm{\hbeta^k - \tbeta^k}^2 + 
%\frac{\blam+\eta}{2}\norm{\hgam^k - \tgam^k}^2
%\le
%\frac{\eta}{2}\norm{\beta^* - \tbeta^k}^2 + 
%\frac{\blam+\eta}{2}\norm{\gam^* - \tgam^k}^2
%\quad\forall\, k\in\N,
%\]
%which implies that the sequence $\{(\hbeta^k,\hgam^k)\}$
%is bounded since the sequence $\{(\tbeta^k,\tgam^k)\}$
%is convergent. 
Let $(\beta_0,\gam_0)$ be any limit point of $\{(\hbeta^k,\hgam^k)\}$
and let $J\subset\N$ 
be such that
$(\hbeta^k,\hgam^k)\overset{J}{\rightarrow}(\beta_0,\gam_0)$. Then, by
the final inequality in 
\eqref{eq:bd limit}, we can take the limit in $k\in J$ to find
that 
\(
\LL_{\emu}((\beta_0,\gam_0),(\tbeta^*,\tgam^*))
\le
\LL_{\emu}((\bbeta,\bgam),(\tbeta^*,\tgam^*)).
\)
The uniqueness of $(\bbeta,\bgam)$ tells us that
$(\beta_0,\gam_0)=(\bbeta,\bgam)$. 
Since 
$(\beta_0,\gam_0)$ was any limit point of the bounded sequence 
$\{(\hbeta^k,\hgam^k)\}$,
we have $(\hbeta^k,\hgam^k)\rightarrow (\bbeta,\bgam)$ which
implies that $\SS_\emu$ is continuous on $\Rp\times\Rq$.
\end{proof}

We now consider the differentiability of $\uemu$. For this we make use of the
following lemma.

\begin{lemma}[Local uniform level boundedness of $\LL_\emu$]
\label{lem:level bdd}
Let $\mu\ge 0$, $\eta>\blam$ and suppose that the assumptions of Theorem \ref{thm:relaxed existence} hold.
Set $x=\bg$ and $w=\tbg$.
% and define $\CC_\mu=\Rp\times\dom(\phimu)$.
Then
the function
$\LL_\emu(\bg,\tbg)$ is level bounded in $\bg$ locally uniformly in $\tbg$
for all $\tbg\in \Rp\times\Rq$. That is, for every 
$\tbg\in\Rp\times\Rq$ and $\rho\in\R$,  
there are $\nu\in\R$ and $\eps>0$ such that
\(
\lset{\bg}{\LL_\emu(\bg,(\bbeta,\bgam)\le\rho}\subset \nu\B
\)
for all $\tbg\in(\bbeta,\bgam)+\eps\B$.
\end{lemma}
\begin{proof}
Set $x=\bg,\ w=\tbg$, and $\bw=(\bbeta,\bgam)$.
If the result is false, there exists $\bw\in\Rp\times\Rq$, $\rho>0$, and a sequence
$\{(x^k,w^k)\}\subset(\Rp\times\dom(\phimu))\times(\Rp\times\Rq)$ 
such that
$w^k\rightarrow\bw$ and $\norm{x^k}\uparrow\infty$ with 
$x^k\in\lset{x}{\LL_\emu(x,w^k)\le\rho}$
for all $k\in\N$. 
By Lemma \ref{lem:LL weak cvx}, the mappings 
$x\mapsto\LL_\emu(x,w)$ are strongly convex with modulus 
$\hat\eta:=\eta-\blam >0$
for all $w\in\Rp\times\Rq$. Let $\hx\in\Rp\times\dom{\phi_\mu}$. 
Then $(\hx,\bw)$
is a point of continuity for $\nabla_x\LL_\emu$, so with no loss in 
generality there is a $c_0>0$ such that 
$\norm{\nabla_x\LL_\emu(\hx,w^k)}\le c_0$ for all $k\in\N$.
Then strong convexity implies that
\[\begin{aligned}
\LL_\emu(\hx,w^k)&-c_0\norm{x^k-\hx}+\frac{\hat\eta}{2}\norm{x^k-\hx}^2
\\ &\le
\LL_\emu(\hx,w^k)+\ip{\nabla_x\LL_\emu(\hx,w^k)}{x^k-\hx}
+\frac{\hat\eta}{2}\norm{x^k-\hx}^2
\\ &\le
\LL_\emu(x^k,w^k)\le\rho.
\end{aligned}\]
But $\LL_\emu(\hx,w^k)-c_0\norm{x^k-\hx}+\frac{\hat\eta}{2}\norm{x^k-\hx}^2\uparrow\infty$ since $\norm{x^k}\uparrow\infty$. This contradiction 
establishes the result.
\end{proof}

\begin{theorem}[Differentiability of $\uemu$]\label{thm:u diff}
Let $\mu\ge 0$, $\eta>\blam$ and suppose that the assumptions of Theorem \ref{thm:relaxed existence} hold.
Then
the function $\uemu$ defined in \eqref{eq:value_function_definition}
is continuously differentiable on $\Rp\times\Rq$ with 
\begin{equation}\label{eq:grad u}
\nabla \uemu\tbg\!=\!\nabla_{\tbg}\kappa_\eta(\tbeta\!-\!\hbeta,\tgam\!-\!\hgam)
\!=\!\eta
\begin{pmatrix}\tbeta\!-\!\hbeta\\ \tgam\!-\!\hgam\end{pmatrix}
\!,\,
\text{where}\ (\hbeta,\hgam)\!=\!\SS_\emu\tbg.
\end{equation}
\end{theorem}

\begin{proof} 
We show that the result follows from \cite[Theorem 10.58]{rockafellar2009variational}.
Set $x=\bg$ and $w=\tbg$. The objective function in the definition of
$\uemu$ is $\LL_\emu(x,w)$, where $\LL_\emu$  is proper and lsc.
Moreover, Lemma \ref{lem:level bdd} tells us that
$\LL_\emu(x,w)$ is level bounded in $x$ locally uniformly in $w$
for all $w\in \Rp\times\Rq$. We have already observed that, 
for all $\mu\in\R_+$ and $\eta>\blam$, $\LL_{\emu}$ and  $\nabla \LL_{\emu}$ 
are continuous
on $(\Rp\times\dom(\phimu))\times(\Rp\times\Rq)$.
Therefore, by \cite[Theorem 10.58]{rockafellar2009variational}
and Theorem \ref{thm:u cont}, $\uemu$ is locally upper-$\CC^1$
and strictly differentiable at every point $w\in\Rp\times\Rq$ with
$\nabla \uemu(w)=\nabla _w\LL_{\emu}(\SS_\emu(w))$.
In addition, $\SS_\emu$ is continuous on 
$\Rp\times\Rq$.
The result follows since $\nabla _w\LL_{\emu}(\bg,\tbg)
=\eta\begin{pmatrix}\tbeta-\beta\\ \tgam-\gam\end{pmatrix}
$.
\end{proof}

\subsubsection{The Lipschitz Continuity of $\nabla \uemu\tbg$}

Since our goal is to employ the PGD algorithm to solve the relaxed problems
\eqref{eq:relax}, we 
require that $\nabla \uemu\tbg$ be Lipschitz continuous. 
Formula \eqref{eq:grad u} tells us that the Lipschitz continuity of
$\nabla \uemu\tbg$ is equivalent to that of the solution mapping 
$\SS_\emu$. 
To study the Lipschitz continuity of $\SS_\emu$ we make use of 
%observe that 
%the optimization problem defining $\uemu$ in \eqref{eq:value_function_definition} 
%is convex
%with polyhedral convex constraints, consequently
%Karush-Kuhn-Tucker
%(KKT) multipliers exist for all $(\tbeta,\tgam)\in \Rp\times\Rq$. 
%We study the properties of these KKT points with the aid of the
the mapping
$\map{G}{\Rp\times\Rq\times\Rp}{\Rp\times\Rq}$ 
be given by
\begin{equation}\label{eq:G}
G_\emu((\beta, \gam,v),(\tbeta,\tgam)):=
\begin{bmatrix}
\nabla_\beta \LL(\beta, \gam) + \eta(\beta - \tbeta) \\
\nabla_\gam \LL(\beta, \gam) + 
\eta(\gam - \tgam) - v
\\
v\odot \gamma  -\mu\one
\end{bmatrix}.
\end{equation}
Observe that, for $\mu>0$,  
$(\hbeta, \hgam)=\SS_\emu(\tbeta,\tgam)$ 
%is an optimal solution to \eqref{eq:value_function_definition}
%with KKT multipier $\hv\in\Rp$ 
if and only if 
\begin{equation}\label{eq:kkt}
\hgam,\hv\in\R_+^q\ \text{ and }\ G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))=0,
\end{equation} 
since the equation $v\odot \gamma  =\mu\one$ implies that
$v=-\nabla \phimu(\gam)$. In addition, when $\mu=0$, condition \eqref{eq:kkt}
is equivalent to $(\hbeta, \hgam,v)$ being a KKT point for 
the optimization problem in
\eqref{eq:value_function_definition}
which, in turn, is equivalent to $(\hbeta, \hgam)=\SS_\emu(\tbeta,\tgam)$
by Theorem \ref{thm:u cont}. We record these observations in the following lemma.

\begin{lemma}\label{lem:oc for u}
Let the assumptions of Theorem \ref{thm:relaxed existence} hold.
Then, for every $(\mu,\eta)\in\R_+\times\R_{++}$, 
$(\hbeta, \hgam)=\SS_\emu(\tbeta,\tgam)$ if and only if there is a vector
$\hv\in\R^q_+$ such that $G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))=0$.
If $\mu>0$, then $\hv=-\nabla \phimu(\hgam)$, and if $\mu=0$, then
$\hv$ is the unique KKT multiplier associated with the constraint 
$0\le\gam$.
%$\gam\in\R^q_+$.
\end{lemma}

Our approach to establishing the Lipschitz continuity of $\SS_\emu$ is to 
first show that $\SS_\emu$ is differentiable and then obtain a bound on its Jacobian.
As usual, diffentiability follows by applying the implicit function theorem to 
$G_\emu$.

\begin{lemma}[The invertibility of $\nabla_{(\beta, \gam,v)}G_\emu$]\label{lem:invert}
Let the assumptions of Theorem \ref{thm:relaxed existence} hold and let
$G_\emu$ be as given in \eqref{eq:G}.
Let 
$(\tbeta,\tgam)\in\Rp\times\Rq$ and  
$(\hbeta, \hgam,\hv)\in\Rp\times\R_+^q\times\R_+^q$ be such that
$G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))=0$. 
Then $\nabla_{(\beta, \gam,v)}G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))$
is invertible if and only if
\begin{equation}\label{eq:str cs}
0<\hv_i+\hgam_i,\ i=1,\dots,q,\qquad\mbox{(strict complementary slackness)}
\end{equation}
which automatically holds if $\mu>0$.
In this case, the inverse is given by
\begin{equation}\label{eq:grad G inv}
\begin{bmatrix}
H^{-1}\!\!-\!\begin{bmatrix}\hat R\\ \hat H_2\end{bmatrix}\!
(D(\hgam)\!+\!D(\hv)\hat H_2)^{-1}D(\hv)[\hat R^T\ \hat H_2]
& 
\!\begin{bmatrix}\hat R\\ \hat H_2\end{bmatrix}\!(D(\hgam)\!+\!D(\hv)\hat H_2)^{-1}
\\
\!-\,(D(\hgam)\!+\!D(\hv)\hat H_2)^{-1}D(\hv)[\hat R^T\ \hat H_2]
&\!
(D(\hgam)\!+\!D(\hv)\hat H_2)^{-1}
\end{bmatrix},
\end{equation}
where $D(\hgam):=\Diag{\hgam}$, $D(\hv):=\Diag{\hv}$, 
\[
\begin{aligned}
H&=
\begin{bmatrix}
H_1&R\\ R^T& H_2
\end{bmatrix}
:=\begin{bmatrix}
\nabla_{\beta\beta}\LL(\hbeta,\hgam)\!+\!\eta I&\nabla_{\gam\beta}\LL(\hbeta,\hgam)
\\
\nabla_{\beta\gam}\LL(\hbeta,\hgam)&\nabla_{\gam\gam}\LL(\hbeta,\hgam)
\!+\!\eta I
\end{bmatrix}\qquad\text{and}
\\
H^{-1}&=
\begin{bmatrix}
H_1^{-1}+H_1^{-1}R(H_2-R^TH_1^{-1}R)^{-1}R^TH_1^{-1}& -H_1^{-1}R(H_2-R^TH_1^{-1}R)^{-1}
\\
-(H_2-R^TH_1^{-1}R)^{-1}R^TH_1^{-1}&(H_2-R^TH_1^{-1}R)^{-1}
\end{bmatrix}
\\
&=:
\begin{bmatrix}
\hat H_1&\hat R\\ \hat R^T& \hat H_2
\end{bmatrix}.
\end{aligned}
\]
\end{lemma}
\begin{proof}
Observe that
\begin{equation}\label{eq:nabla G}
\nabla_{(\beta, \gam,v)}G_\emu((\beta, \gam,v),(\tbeta,\tgam))
\!=\!\!
\begin{bmatrix}
\nabla_{\beta\beta}\LL(\beta,\gam)\!+\!\eta I&
\nabla_{\gam\beta}\LL(\beta,\gam)&0\\
\nabla_{\beta\gam}\LL(\beta,\gam)&\nabla_{\gam\gam}\LL(\beta,\gam)
\!+\!\eta I&-I
\\
0&\Diag{v}&\Diag{\gam}
\end{bmatrix}.
\end{equation}
Let us first assume that 
$\nabla_{(\beta, \gam,v)}G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))$
is invertible and, for simplicity write  
\[
\nabla_{(\beta, \gam,v)}G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))
=\begin{bmatrix}H&A\\ B^T&D\end{bmatrix},
\]
where $A:=[0,\ -I]^T$, $B=[0,\ \Diag{\hv}]^T$ and $D:=\Diag{\hgam}$. 
Since $H\in\bS^{p+q}_{++}$,
the matrix
\begin{equation}\label{eq:grad G inv1}
\begin{aligned}
\begin{bmatrix}I&0\\ -B^TH^{-1}&I\end{bmatrix}
\begin{bmatrix}H&A\\ B^T&D\end{bmatrix}
\begin{bmatrix}I&-H^{-1}A\\ 0&I\end{bmatrix}&=
\begin{bmatrix}H&0\\ 0&D-B^TH^{-1}A\end{bmatrix}
\\ &=\begin{bmatrix}H&0\\ 0&\Diag{\hgam}+\Diag{\hv}\hat H_{2}\end{bmatrix}
\end{aligned}
\end{equation}
is nonsingular. In particular, the matrix $\Diag{\hgam}+\Diag{\hv}\hat H_{2}$ is
necessarily invertible.
But if there is an $i$ such that $0=\hgam_i+\hv_i$, then $\hgam_i=\hv_i=0$
so that the matrix $\Diag{\hgam}+\Diag{\hv}H_{22}$ has a zero row and so
is singular. Since this cannot be that case,
\eqref{eq:str cs} must hold.

Conversely, suppose $(r^T,s^T,t^T)^T$ is in the nullspace of 
$\nabla_{(\beta, \gam,v)}G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))$. Then
$0=\Diag{\hv}s+\Diag{\hgam}t$. This combined with \eqref{eq:str cs} implies
that $s^Tt=0$. Consequently,
\[
0=\begin{pmatrix}r\\ s\end{pmatrix}^T\begin{pmatrix}0\\ t\end{pmatrix}
=\begin{pmatrix}r\\ s\end{pmatrix}^TH\begin{pmatrix}r\\ s\end{pmatrix},
\]
which implies that $(r,s)=(0,0)$ since $H$ is positive definite.
Therefore, $t=0$ which shows that 
$\nabla_{(\beta, \gam,v)}G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))$ is
nonsingular.

The formula for the inverse follows from \eqref{eq:grad G inv1} which tells us that
\begin{equation}\label{eq:grad G inv2}
\begin{bmatrix}H&A\\ B^T&D\end{bmatrix}^{-1}=
\begin{bmatrix}I&0\\ -B^TH^{-1}&I\end{bmatrix}
\begin{bmatrix}H^{-1}&0\\ 0&(\Diag{\hgam}+\Diag{\hv}\hat H_{2})^{-1}\end{bmatrix}
\begin{bmatrix}I&-H^{-1}A\\ 0&I\end{bmatrix}.
\end{equation}
Alternatively, one can apply the formulas in \cite{LS02}.
\end{proof}

Using Lemma \ref{lem:invert}, we
apply the implicit function theorem to the equation 
\[
G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))=0
\] 
and obtain the following result.
%to establish
%the differentiability of
%the solution mapping 
%$\SS_\emu(\tbeta,\tgam)$ and obtain a formula for its derivative.

\begin{theorem}[Differentiability of $\SS_\emu$]\label{thm:diff sol map}
Let the hypotheses and notation of Lemma \ref{lem:invert} hold and let $\SS_\emu$
be as defined in \eqref{eq:argmin for u}. Given $\emu\in\R_+\times\R_+$,
define
$\map{\hSS_\emu}{\R^p\times\R^q}{\R^p \times\R^q_+\times\R^q_+}$ by
\begin{equation}\label{eq:hSS}
\hSS_\emu(\tbeta,\tgam)=
\lset{(\hbeta, \hgam,\hv)}{\hgam,\hv\in\R^q_+\ \text{and}\ 
0=G_\emu((\hbeta, \hgam,\hv),(\tbeta,\tgam))}
\end{equation}
Suppose
$(\tbeta,\tgam)\in\Rp\times\Rq$ and  
$(\bbeta, \bgam,\bv)=\hSS_\emu(\tbeta,\tgam)$
with $\bgam,\bv\in\R_+^q$ and such that \eqref{eq:str cs} holds. Then there
exist open neighborhoods $\widetilde\NN$ of $(\tbeta,\tgam)$ and 
such that 
$\SS_\emu$ and $\hSS_\emu$ are differentiable on $\widetilde\NN$ with
%and we have the 
%following formulas for these derivatives:
\[
\begin{aligned}
\nabla \hSS_\emu(\beta,\gam)&=\eta
\begin{bmatrix}
H^{-1}\!\!-\!\begin{bmatrix}\hat R\\ \hat H_2\end{bmatrix}\!
(D(\hgam)\!+\!D(\hv)\hat H_2)^{-1}D(\hv)[\hat R^T\ \hat H_2]
\\
\!-\,(D(\hgam)\!+\!D(\hv)\hat H_2)^{-1}D(\hv)[\hat R^T\ \hat H_2]
\end{bmatrix},
\\
\nabla \SS_\emu(\beta,\gam)&=\eta
\begin{bmatrix}
H^{-1}\!\!-\!\begin{bmatrix}\hat R\\ \hat H_2\end{bmatrix}\!
(D(\hgam)\!+\!D(\hv)\hat H_2)^{-1}D(\hv)[\hat R^T\ \hat H_2]
\end{bmatrix}
\end{aligned}
\]
for all $(\beta,\gam)\in \widetilde\NN$ and 
$(\hbeta,\hgam,\hv)=\hSS_\emu(\beta,\gam)$.
In particular, this implies that both $\hSS_\emu$ and $\SS_\emu$
are continuously differentiable on $\Rp\times\Rq$.
\end{theorem}

Using the notation of Lemma \ref{lem:invert} the expression for 
$\nabla \SS_\emu(\beta,\gam)$ in 
Theorem \ref{thm:diff sol map} can be simplified when $\mu>0$ to
{\small\[
\nabla \SS_\emu(\beta,\gam)\!=\eta
\!\left[ H^{-1}\!\!-\!
\begin{bmatrix}\!-\!H_1^{-1}R\\ I\end{bmatrix}
\hat H_2(\mu^{-1}\Diag{\hgam}^2\!+\!\hat H_2)^{-1}\!\hat H_2
[-R^TH_1^{-1}\ I]
\right]\! .
\]}
By combining this with the Shur complement formula
(e.g., see \eqref{eq:grad G inv1} and \eqref{eq:grad G inv2})
\[
H^{-1}=\begin{bmatrix}I&-H_1^{-1}R\\ 0&I\end{bmatrix}
\begin{bmatrix}H_1^{-1}&0\\ 0&(H_2-R^TH_1^{-1}R)^{-1}\end{bmatrix}
\begin{bmatrix}I&0\\ -R^TH_1^{-1}&I\end{bmatrix},
\]
where $\hat H_2=(H_2-R^TH_1^{-1}R)^{-1}$ is positive definite, we obtain  
{\small\[
\nabla \SS_\emu(\beta,\gam)\!=\!\eta
\begin{bmatrix}I&\!\!\!\!-H_1^{-1}R\\ 0&\!\!\!\!I\end{bmatrix}\!\!
\begin{bmatrix}H_1^{-1}&\!\!\!\!\!0\\ 
0&\!\!\!\!\!\hat H_2\!-\!\hat H_2(\mu^{-1}\Diag{\hgam}^2\!+\!\hat H_2)^{-1}\hat H_2
\end{bmatrix}\!\!
\begin{bmatrix}\!I&\!\!\!\!0\\ \!-R^TH_1^{-1}&\!\!\!\!I\end{bmatrix}
\]}
%where we recall that $\hat H_2^{-1}=H_2-R^TH_1^{-1}R$ is positive definite.
Since the matrix
\[
\hat H_2-\hat H_2(\mu^{-1}\Diag{\hgam}^2+\hat H_2)^{-1}\hat H_2=
\hat H_2^{1/2}[I-(I+\mu^{-1}\hat H_2^{-1/2}\Diag{\hgam}^2\hat H_2^{-1/2})^{-1}]
\hat H_2^{1/2}
\]
is positive definite, we have that
\begin{equation}\label{eq:grad S bound}
\norm{\nabla \SS_\emu(\beta,\gam)}\le
\eta(1+\norm{H_1^{-1}R}^2)
\max\{\norm{H_1^{-1}},\norm{(H_2-R^TH_1^{-1}R)^{-1}}\}.
\end{equation}
Since $H_1=\nabla_{\beta\beta}\LL(\hbeta,\hgam)+\eta I$, we have 
\(
\norm{H_1^{-1}}\le \eta^{-1}<(\eta-\bar\eta)^{-1}.
\)
We now show that  $(\eta-\bar\eta)^{-1}$ bounds
$\norm{(H_2-R^TH_1^{-1}R)^{-1}}$.
For this it is sufficient to show that $(\eta-\bar\eta)\le \mu_\mmin(H_2-R^TH_1^{-1}R)$.
%By Lemma \ref{lem:LL weak cvx}, the function $\LL(\beta,\gam)+\frac{\blam}{2}
%\norm{\gam}^2$ is convex. 
By Lemma \ref{lem:LL weak cvx}, the matrix in \eqref{eq:psd for LL} is positive
semidefinite.
%Therefore, $\nabla^2 \LL\bg+\bar\eta I$
%\[
%\begin{bmatrix}
%\nabla_{\beta\beta}\LL(\beta,\gam)&\nabla_{\gam\beta}\LL(\beta,\gam)\\
%\nabla_{\beta\gam}\LL(\beta,\gam)&\nabla_{\gam\gam}\LL(\beta,\gam)
%+\blam I\end{bmatrix}
%\]
%is positive semidefinite. 
Since 
$\nabla_{\beta\beta}\LL(\beta,\gam)$ 
is positive
definite, the Shur complement 
\(
%\begin{gather*}
\nabla_{\gam\gam}\LL(\beta,\gam)+\blam I-
\nabla_{\beta\gam}\LL(\beta,\gam)\nabla_{\beta\beta}
\LL(\beta,\gam)^{-1}
\nabla_{\gam\beta}\LL(\beta,\gam)
%\\
%=\nabla_{\gam\gam}\LL(\beta,\gam)+\blam I-
%R^T\nabla_{\beta\beta}(\LL(\beta,\gam)+\bar\eta I)^{-1}R
%\end{gather*}
\)
is positive semidefinite. Consequently,
\[
\begin{aligned}
H_2-R^TH_1^{-1}R=
(\eta-\bar\eta) I&+ (\nabla_{\gam\gam}\LL(\beta,\gam)+\blam I-
R^T\nabla_{\beta\beta}\LL(\beta,\gam)^{-1}R)
\\ &+R^T(\nabla_{\beta\beta}\LL(\beta,\gam)^{-1}-(\nabla_{\beta\beta}\LL(\beta,\gam)+\eta I)^{-1})R
\ \succeq (\eta-\bar\eta) I,
\end{aligned}
\]
since $\nabla_{\beta\beta}\LL(\beta,\gam)^{-1}-(\nabla_{\beta\beta}\LL(\beta,\gam)+\eta I)^{-1}$ is positive definite.
Therefore, $(\eta-\bar \eta)\le \mu_\mmin(H_2-R^TH_1^{-1}R)$.
By combining this with \eqref{eq:grad S bound} we obtain the bound
\begin{equation}\label{eq:S lip bd 1}
\norm{\nabla \SS_\emu(\beta,\gam)}\le
%\frac{\blam+\eta}{\eta}(1+\eta^{-2}\norm{\nabla_{\gam\beta}\LL(\hbeta,\hgam)}^2)
\frac{\eta}{\eta-\blam}\left(1+\norm{H_1^{-1}R}^2\right),
\end{equation}
where 
\[
\begin{aligned}
H_1^{-1}R&=-(X^T\Omega(\hgam)^{-1}X+\eta I)^{-1}
\sum_{i=1}^mX_i^T\Omega_i(\hgam)^{-1}Z_i
\Diag{Z_i^T\Omega_i(\hgam)^{-1}(X_i\hbeta-Y_i)}
\\ &=
-(X^T\Omega(\hgam)^{-1}X+\eta I)^{-1}X^T\Omega(\hgam)^{-1}
\hZ\,\Diag{\hZ^T\Omega(\hgam)^{-1}r(\hbeta)},
\end{aligned}
\]
with $r(\hbeta):=X\beta-y$ and 
\(
\hZ=\Diag{Z_1,Z_2,\dots,Z_m}.
\)
Therefore, as in Lemma \ref{lem:LL weak cvx}, we obtain the bound
\begin{equation}\label{eq:S lip bd 2}
\norm{H_1^{-1}R}\le \eta^{-1}\mu^{-2}_\mmin(\Lam)          
\sig_\mmax(X)\sig_\mmax^2(Z)\norm{X\hbeta -y}.
\end{equation}
This inequality can be used to show that $\nabla \uemu$ is bounded on the lower
level sets of $\uemu\tbg + R\tbg+\del_{\Rqp}(\tgam)$ if
$\LL_{\emu}(\bg,\tbg) + R\tbg+\del_{\R^q_+}(\tgam)$ is level
compact. However, we only know that this is true if we can bound the values 
of $\hgam$ over these sets. In practice, the values of $\hgam$ are bounded if the
model is well posed since these values are tied to the variances of
the random effects. One can accommodate this by adding a constraint of the 
form $\gam\le\gammax$ for $\gammax\in\R^q_{++}$ chosen sufficiently large.

\begin{lemma}[Lipschitz Continuity of $\nabla\uemu$]\label{lem:grad u lip}
Let the assumptions of Theorem \ref{thm:relaxed existence} hold and suppose
$\mu>0$ and $\gammax\in\R^q_{++}$. Let $\zeta\in\R$ and set 
\[
\begin{aligned}
\widehat\VV(\eta,\mu,\gammax,\zeta)&:=
\lset{(\bg,\tbg)}{\begin{array}{c}
\LL_{\emu}(\bg,\tbg) + R\tbg+\del_{\R^q_+}(\tgam)\le\zeta,\\ \gam,\ \tgam\le\gammax\end{array}},
\text{and}
\\
\VV(\eta,\mu,\gammax,\zeta)&:=\lset{\tbg}{\uemu\tbg + R\tbg+\del_{\Rqp}(\tgam)\le\zeta,\ \tgam\le\gammax}.
\end{aligned}
\]
Then 
\begin{enumerate}
\item
both $\widehat\VV(\eta,\mu,\gammax,\zeta)$ and 
$\VV(\eta,\mu,\gammax,\zeta)$ 
are compact with
\begin{equation}\label{eq:VV inclusion}
\VV(\eta,\mu,\gammax,\zeta)\subset\lset{\tbg}{\begin{aligned}\tgam\le\gammax
 \ \text{and}\ 
\exists\,\bg\in\Rp\times\R^q_{++}
\ s.t.
\\
(\bg,\tbg)\in\widehat\VV(\eta,\mu,\gammax,\zeta)
\end{aligned}},
\end{equation}
\item
the set $\VV(\eta,\mu,\gammax,\zeta)$ has nonempty interior if
$\zeta>\uemu\tbg$ for some $\tbg\in\Rp\times\Rq$, and
\item
the set
$\widetilde\VV(\eta,\mu,\gammax,\zeta,\omega):=\clco(\VV(\eta,\mu,\gammax,\zeta)+\omega\uball)$ is a compact, convex set with nonempty interior
whenever $\VV(\eta,\mu,\gammax,\zeta)\ne\emptyset$.
\end{enumerate}
Moreover, $\nabla u_\emu$ is Lipschitz on $\clco(\VV(\eta,\mu,\gammax,\zeta)+\omega\uball)$
for every $\omega\ge 0$, where 
\[
\uball:=\lset{\bg\in\Rp\times\R^q}{\norm{\bg}\le 1}.
\]
%Then $\nabla u_\emu$ is Lipschitz on $\VV(\eta,\mu,\zeta)$ if the set
%\[
%\hRR:=\lset{X\hbeta-y}{(\hbeta,\hgam)=\SS_\emu\tbg,\ \tbg\in \VV(\eta,\mu,\zeta)}
%\]
%is bounded.
%In particular, this holds if the set
%\[
%\widehat\GG:=\lset{\hgam}{(\hbeta,\hgam)=\SS_\emu\tbg,\ \tbg\in \VV(\eta,\mu,\zeta)}
%\]
%is bounded since in this case the function 
%$\LL_{\emu}(\bg,\tbg) + R\tbg+\del_{\R^q_+}(\tgam)$
%is level compact.
\end{lemma}
\begin{proof}
%The first statement of the corollary follows from formula for $\nabla u_\emu$ in Theorem \ref{thm:u diff} and the bounds \eqref{eq:S lip bd 1} and 
%\eqref{eq:S lip bd 2}. 
Since Theorem \ref{thm:basic existence} tells us that
$\LL$ is bounded below, 
$\LL_{\emu}(\bg,\tbg) + R\tbg+\del_{\R^q_+}(\tgam)$ is not level compact
if and only if there is an unbounded sequence
in a lower level set of 
$\LL_{\emu}(\bg,\tbg) + R\tbg+\del_{\R^q_+}(\tgam)$ 
for which $\gam^k\uparrow\infty$.
Therefore, the compactness of $\widehat\VV(\eta,\mu,\gammax,\zeta)$ 
follows from the lower semicontinuity of  
$\LL_{\emu}(\bg,\tbg) + R\tbg+\del_{\R^q_+}(\tgam)$.
%Hence, the second statement of the corollary follows.
Since
\[
\uemu\tbg + R\tbg+\del_{\Rqp}(\tgam)\le 
\LL_{\emu}(\bg,\tbg) + R\tbg+\del_{\R^q_+}(\tgam)\quad
\forall\, \bg\in\Rp\times\Rq,
\]
the inclusion \eqref{eq:VV inclusion} holds.
In addition, the set on the right hand side of \eqref{eq:VV inclusion}
is the projection of $\widehat\VV(\eta,\mu,\gammax,\zeta)$ onto
its
first components $\bg$ and so is compact. This in turns tells us that 
$\VV(\eta,\mu,\gammax,\zeta)$ is compact. Hence, 
$\clco(\VV(\eta,\mu,\gammax,\zeta)+\omega\uball)$
is also compact. The continuity of $\uemu$ implies that 
$\VV(\eta,\mu,\gammax,\zeta)$ has nonempty interior if
$\zeta>\uemu\tbg$ for some $\tbg\in\Rp\times\Rq$.
Theorem \ref{thm:u cont} shows that $\SS_\emu$ is continuous on 
$\Rp\times\Rq$ so the bound \eqref{eq:S lip bd 2} combined with 
Theorem \ref{thm:diff sol map} implies that $\SS_\emu$ is locally
Lipschitz on $\Rp\times\Rq$. Hence, by \eqref{eq:grad u},
$\nabla \uemu$ is locally Lipschitz on $\Rp\times\Rq$.
The compactness of 
$\clco(\VV(\eta,\mu,\gammax,\zeta)+\omega\uball)$
tells us that $\nabla \uemu$ is Lipschitz on
this set for all $\omega\ge 0$.
\end{proof}

%\begin{remark}
%One can ensure that the set $\hGG$ is bounded by 
%choosing an upper bound $\gam^u\in\R^q_{++}$ and adding a term 
%%that bounds $\hgam$ from above 
%of the form 
%$\del_{\{\gam\le\gam^u\}}(\gam)$
%to 
%$\uemu\tbg + R\tbg+\del_{\Rqp}(\tgam)$.
%%the objective in \eqref{eq:relax2}. 
%%The bound  is user spe
%\end{remark}

\subsection{Convergence of the PGD Algorithm for $\Phi_\emu$}
\label{sec:convergence pgd}

The convergence of the PGD algorithm for fixed valued of the relaxation parameters
$\eta$ and $\mu$ appeals to the standard convergence theory as presented in
\cite[Chapter 10]{AB17} which requires the use of Assumptions (A)--(C) in Section 
\ref{sec:pgd}. We assume that the
variable selection regularizer $R$ is chosen so that
Assumption (A) holds. 
In addition, under the assumptions of Theorem 5, 
%we have show that solutions to 
%\eqref{eq:relax2} exist, 
Theorem \ref{thm:u diff} tells us that the function $\uemu$ is well defined and continuously differentiable on all of $\Rp\times\Rq$
with the solution mapping 
$\SS_\emu\tbg$ well defined, single valued, and differentiable on $\Rp\times\Rq$ 
(Theorem \ref{thm:diff sol map}).
 Therefore, Assumption (C)
is satisfied as is much of assumption (B).
However, as is commonly the case in a specific application, the 
$L_\emu$-smoothness of $\uemu$ over 
$\intr{\dom{\uemu}}=\Rp\times\Rq$ fails. This drawback is remedied by
observing that the 
PGD algorithm is a descent algorithm. This allows us to focus on the behavior of
the functions over the lower level sets
described in Lemma \ref{lem:grad u lip}.

Let $\bw^{0}=(\tbeta^0,\tgam^0)\in \Rp\times\R^q_+$ be the point at which Algorithm \ref{alg:MSR3}
is initiated and let
$\zeta>\LL_{\emu}(\bg,(\tbeta^0,\tgam^0)) 
+ R(\tbeta^0,\tgam^0)+\del_{\R^q_+}(\tgam^0)$ for any 
$\bg \in \Rp\times\R^q_{++}$. 
For $\omega\ge 0$ and $\eps\ge 0$, define
\(
\Dfr(\omega,\eps):=\widetilde
\VV(\eta,\mu,\gammax+\eps\one,\zeta+\eps,\omega+\eps)
\)
and set
\begin{equation}\label{eq:hats}
\hat u_\emu:=\uemu+\delta_{\Dfr(\bomega,\beps)}\quad\text{ and }\quad
\hR:=R+\del_{\Dfr(\bomega,0)},
\end{equation}
where $\beps>0$, $\widetilde\VV$ is defined in Lemma \ref{lem:grad u lip} and
\[
\bomega:=1+t_0\,\max\lset{\nabla \uemu\tbg}{\tbg\in\VV(\eta,\mu,\gammax,\zeta)}.
\]
Observe that all iterates of Algorithm \ref{alg:MSR3} 
lie in the set $\VV(\eta,\mu,\gammax,\zeta)$
since it is a descent algorithm.
Moreover, since the prox operator is nonexpansive 
(e.g., see  
\cite[Theorem 6.42(a)]{AB17} or \cite[Theorem 12.19]{rockafellar2009variational}),
all of the points tested in the backtracking line search in Algorithm 
\ref{alg:MSR3} 
must also
lie in the set $\widetilde\VV(\eta,\mu,\gammax,\zeta,\bomega)$ by construction.
Therefore, the iterates of Algorithm \ref{alg:MSR3} 
are identical to those obtained
when the algorithm is applied to $\hat u_\emu$ with 
$\tR:=R+\del_{\Dfr(\bomega,0)}$.
That is, we can assume
that the Algorithm \ref{alg:MSR3} 
is being applied to $\hat u_\emu$. 
Observe that $\hat u_\emu$ is closed and proper,
$\dom{\hat u_\emu}=\Dfr(\bomega,\beps)$ is convex, and
 $\dom{\hat u_\emu}=\Dfr(\bomega,\beps)$
has nonempty interior (by Lemma \ref{lem:grad u lip}(3)) with $\dom{\tR}\subset \intr{\dom{\hat u_\emu}}$ since $\beps>0$. 
In addition, the final statement of Lemma \ref{lem:grad u lip}
tells us that there is an $L_{(\emu,\gammax,\zeta)}>0$ 
%depending on $(\tbeta^0,\tgam^0)$ 
such that $\hat u_\emu$ is 
$L_{(\emu,\gammax)}$-smooth over $\intr{\dom{\hat u_\emu}}$. Hence, 
Assumptions (A)-(C) are satisfied by $\hat u_\emu$ and $\tR$ and so the 
convergence properties in \cite[Theorem 10.15]{AB17} hold for Algorithms
1 and 2 applied to $\uemu$ and $R$ under the assumptions of Theorem  
\ref{thm:relaxed existence}. By applying these observation
to \cite[Theorem 10.15]{AB17}, we obtain the following convergence result.

\begin{theorem}[Convergence of Algorithms \ref{alg:MSR3} and \ref{alg:pgd with bt}]\label{thm:convergence}
Let the assumptions of Theorem \ref{thm:relaxed existence} hold, and
let $\Phi_\emu$ be as defined in \eqref{eq:varphi}. 
%and suppose
%$\gammax\in\R^q_{++}$. 
Let $\{\tbgk\}$ be a sequence generated
by either Algorithm \ref{alg:MSR3} or \ref{alg:pgd with bt} 
with parameters 
$\theta\in(0,1),\ \tau\in(0,1)$,
$\eta>0,\ \mu>0$, $\eps_{\mbox{\tiny Tol}}=0$,
$t_0>0$, and $\gammax>\tgam^0$. 
Then, given $\zeta>\uemu(\tbeta^0,\tgam^0)$ there is an 
$L_{(\emu,\gammax,\zeta)}>0$ such that $\nabla\uemu$ is 
$L_{(\emu,\gammax,\zeta)}$-smooth over 
$\widetilde\VV((\emu,\gammax,\zeta,1)$. 
In Algorithm \ref{alg:pgd with bt}, replace 
$L_\emu$ with $L_{(\emu,\gammax,\zeta)}$ and set
 \[
 \begin{aligned}
 M&:=\begin{cases}
 \alf(1-\alf\frac{L_{(\emu,\gammax,\zeta)}}{2}),&
 \mbox{in Algorithm \ref{alg:pgd with bt}},
 \\
 \frac{2t_0\theta^2(1-\tau)}{\max\{2\theta(1-\tau),t_0L_{(\emu,\gammax,\zeta)}\}},&
 \mbox{in Algorithm \ref{alg:MSR3}},
 \end{cases}
 \qquad\text{and}
 \\
 r&:=
 \begin{cases}
 \alf,&
 \mbox{in Algorithm \ref{alg:pgd with bt}},
 \\
t_0,&
 \mbox{in Algorithm \ref{alg:MSR3}}.
 \end{cases}
 \end{aligned}\]
Then either $\tgam^k>\gammax$ after a finite number of iterations and the algorithms terminate, or the 
following hold:
\begin{enumerate}
\item
The sequence $\Phi_\emu\tbgk$ is nondecreasing.
In addition, \[\Phi_\emu(\tbgk)\tbeta^{k+1},\tgam^{k+1})<\Phi_\emu\tbgk\]
if and only if $\tbgk$ is not a stationary point of \eqref{eq:relax3}.
\item
$\norm{\tbgk-\prox_{r\tR}(\tbgk-r\nabla u_\emu\tbgk)}\rightarrow 0$ with
\[
\min_{i=0,1,\dots,k}\norm{\tbgk-\prox_{r\tR}(\tbgk-r\nabla u_\emu\tbgk)}
\le \frac{\sqrt{\Phi_\emu(\tbeta^0,\tgam^0)-\Phi_\emu^\mathsf{OPT}}}
{\sqrt{M(k+1)}},
\]
where $\Phi_\emu^\mathsf{OPT}:=\inf \Phi_\emu$.
\item
All limit points of the sequence $\{\tbgk\}$ are stationary points of
problem \eqref{eq:relax3}.
\end{enumerate}
\end{theorem}
\begin{proof}
As observed prior to the statement of the theorem, 
both Algorithm \ref{alg:MSR3} and \ref{alg:pgd with bt}
behave as if they were applied to the the functions 
$\hat u_\emu$ and $\hR$ defined in \eqref{eq:hats}. 
It was also shown that the functions $\hat u_\emu$ and $\hR$
satisfy the Assumptions (A)-(C) required by \cite[Theorem 10.15]{AB17}.
Hence, the consequences of \cite[Theorem 10.15]{AB17} hold.
By translating the notion of \cite[Theorem 10.15]{AB17} to that of this paper,
we obtain the result.
\end{proof}

\subsection{A Hybrid Algorithms for Feature Selection in Mixed Effects Models}
\label{sec:hybrid alg}

 In the previous section we established the convergence properties of the 
 PGD algorithm applied to the function $\Phi_\emu$ for fixed values of
 $\eta$ and $\mu$. In subsection \ref{ssec:existence and consistency},
 two consistency results are established for the relaxed problem \eqref{eq:relax3}.
 Theorem \ref{thm:eta consistency} shows that, for fixed $\mu\ge 0$,
 every limit point of solutions to \eqref{eq:relax3} as $\eta\uparrow\infty$
 is a solution to \eqref{eq:log-barrier problem}, while Theorem
 \ref{thm:mu consistency} tells us that every
 limit point of solutions \eqref{eq:log-barrier problem} as $\mu\downarrow 0$
 is a solution to the variable selection problem \eqref{eq:vs_lme}. 
These results suggest a range of numerical approachs to obtaining approximate 
solutions to the target problem \eqref{eq:vs_lme}.
 %%
 The issue of foremost concern is the method for approximating 
 solutions to \eqref{eq:value_function_definition} since the accuracy in this approximation
 determines the accuracy in both $\uemu$ and $\nabla \uemu$.
 %%
 To address this concern, we view the algorithm from an interior point
 perspective where every point on the {\it central path} is a solution
 to the optimization problem \eqref{eq:value_function_definition}
 defining $\uemu$ for the associated value of the homotopy parameter $\mu$. 
 An approximate solution is then
 considered acceptable if it is sufficiently close to the central path where
 proximity to the central path is measured in terms of the notion of the 
 {\it neighborhood} of the central path , e.g. see \cite{Wright-IP-book}. 
 %An interior point approach is very efficient
 Due to the convexity of the optimization problems \eqref{eq:value_function_definition}, 
 this is an efficient algorithm for approximating 
 $\uemu$ to high accuracy.
 %%

 The next issue we addressed is the method for initializing and adjusting the parameter 
 $\eta$.  This is particularly significant since the initial value of $\eta$ must be chosen
 to assure the convexity of the problems in \eqref{eq:value_function_definition}.
 Lemma \ref{lem:LL weak cvx} gives us guidance in this regard, but the necessary 
 computations to obtain a lower bound on $\eta$ can be arduous, and, in general,
 produce a wildly pessimistic lower bound. For this reason, we take a somewhat different 
 approach by proposing a variable metric strategy for solving the optimization problems
 in \eqref{eq:value_function_definition}. In this approach, we replace 
 the Hessian matrix $\nabla^2\LL_{\emu}\bg$ in
 the Newton equation

 \centerline{$
 G_\emu((\beta, \gamma, v), (\tbeta, \tgamma))+
 \nabla G_\emu((\beta, \gamma, v), (\tbeta, \tgamma))
 [dv, d\beta, d\gamma] =0
$}

\noindent
by the positive semi-definite approximation

 \centerline{$
\nabla^2\LL{(\beta,\gam)} \approx \sum_{i=1}^m
S_i^T\begin{bmatrix}X_i^T\\ -Z_i^T\end{bmatrix}
\Omega_i(\gam)^{-1}
\begin{bmatrix}X_i& -Z_i\end{bmatrix}S_i
%-\begin{bmatrix}
%0&0\\ 0& \half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}
%\end{bmatrix},
$}

\noindent
which is motivated by the expression for $\nabla^2\LL_{\emu}\bg$ given in
\eqref{eq:hess LL}. That is, we simply drop the negative semi-definite term
$-\sum_{i=1}^m\begin{bmatrix}
0&0\\ 0& \half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}
\end{bmatrix}$.
With this modification, the subproblems we solve are strongly convex for all $\eta>0$.
Consequently, the problem of initializing $\eta$ is less problematic. Our numerical experiments indicate that the performance of the algorithm is robust with respect to 
$\eta$. For this reason, we choose an initial value for $\eta$ and 
then leave it fixed over all iterations. Our method for choosing $\eta$ is described
in \cite[Section 4, Figure 5]{Practice}. Briefly, we maximize the 
Baysian Information Criterion (BIC) over a grid of values for $\eta$. 
 The resulting BIC response curve shows 
that the method is robust with respect to the 
 choice of $\eta$ and choosing $\eta\in [1,10]$ yields 
 accurate solutions for our selected test problems. 
 Once $\eta$ is fixed the
 PGD algorithm can be applied to solve the problem \eqref{eq:relax3}
 for decreasing values of $\mu$.

### 2023-04-10 msr3.tex
Add theory paper to the thesis

@@
 }
+
+\begin{theorem}[Existence of a Minimizer, Theorem 1 from \cite{aravkin2022jimtheory}]\label{thm:basic existence}
+Let the assumptions in the statement of problem \eqref{eq:ml_lme_optimization_setup} hold. Then optimal solutions to
+\eqref{eq:ml_lme_optimization_setup} exist.
+\end{theorem}
+
 At this point, we bring in three basic definitions from variational analysis~\cite{rockafellar2009variational}.
@@
 
-Table \ref{table:proxes} provides closed form expressions for the proximal operators 
+Under the assumption that $R$ is level-compact \cite{aravkin2022jimtheory} extended Theorem \ref{thm:basic existence} to the regularized case as follows:
+
+\begin{theorem}[Existence of a Minimizer for \ref{eq:lme_loss_original_in_x}, \cite{aravkin2022jimtheory}, Theorem 2]
+
+\label{thm:basic existence2}
+Let the assumptions in the statement of problem \eqref{eq:lme_diagonal_setup} hold.
+%define $\LL(\beta,\gam):=\LL_{ML}(\beta, \Diag{\gamma})$
+Suppose $\map{\hR}{\R^p\times\R^q_+}{\R\cup\{+\infty\}}$ 
+is lsc and level compact (i.e., $\epi{R}:=\lset{(\bg,\nu)}{R\bg\le\nu}$ is closed
+and $\lset{\bg}{R\bg\le\nu}$ is bounded for all $\nu\in\R$).
+Then $\LL+\hR$ is level compact and solutions to
+the following optimization problem exist:
+\eq{
+\label{eq:extended loss}
+\min_{\beta\in\R^p, \gamma\in\R^q_+}  \LL(\beta, \gamma) + 
+\hR(\beta, \gamma) .
+}
+\end{theorem}
+
+In practice, it is often advisable to include a constraint of the form 
+$\gamma\le\gammax$ for $\gam\in\R^q_{++}$ chosen
+sufficiently large since an excessively large variance usually indicates 
+that the model is poorly posed and needs review. 
+Such a constraint is also numerically expedient since it prevents $\gam$ from diverging. Table \ref{table:proxes} provides closed form expressions for the proximal operators 
 of commonly used regularizers. 
@@
 
-\begin{theorem}[$\prox$ for bounded $\gamma$]     \label{thm:prox_of_positive_quadrant}
+\begin{theorem}[Proximal operator for bounded $\gamma$]     \label{thm:prox_of_positive_quadrant}
 We consider modified regularizers $r(\gamma)$ from the Table \ref{table:proxes} 
@@
 \end{theorem}
-The proof of the Theorem \ref{thm:prox_of_positive_quadrant} is provided in Appendix \ref{ch:proxes_appendix}. 
+The proof of the Theorem \ref{thm:prox_of_positive_quadrant} is provided in Appendix \ref{appendix:proxes}. 
 %\textcolor{red}{Aleksei, please check proof of appendix is all in terms of $\gamma$, rather than $v = (\beta, \gamma)$.}
 The  PGD algorithm is detailed in Algorithm~\ref{alg:pgd_for_lme}.
-{The algorithm's step-size $\alpha$ depends on the Lipschitz constant; an upper-bound is given in Appendix \ref{sec:lipschitz_constant}.
+{The algorithm's step-size $\alpha$ depends on the Lipschitz constant; an upper-bound is given in Appendix \ref{appendix:lipschitz_constant}.
 In practice, $\alpha$ is computed using a line-search, since the available estimate for $L$ is very conservative.}
@@
 \eq{
-\label{eq:value_function_definition}
 		u_\eta(w) & = \min_x \ \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w)\\
@@
 }
-Substituting (\ref{eq:value_function_definition}) into (\ref{eq:lme_loss_sr3}) the problem transforms into the problem of optimizing a regularized value function:
+Substituting the above into (\ref{eq:lme_loss_sr3}) the problem transforms into the problem of optimizing a regularized value function:
 \eq{
@@
 \section{Theoretical Analysis}
-\ask{Jim}{What would be the best way to introduce theoretical results to my thesis?}
+\input{msr3-theory}
 

### 2023-04-11 appendix/pinode_appendix.tex
Add appendices; connect MSR3 theory and practice definitions; compiles with NO WARNINGS

\paragraph{Hardware} All experiments were computed using an \texttt{Intel(R) Xeon(R) CPU E5-2630 v4 @ 2.20 GHz} equipped with a \texttt{Tesla K80} GPU. The computer had \texttt{Linux 4.15} installed as an OS.

\subsection{Lifted Duffing Oscillator: Learning Unseen Basins with Collocations (Figure~\ref{fig:duffing_pinode})}
\label{appendix:duffing_pinode}
In this experiment we trained two models; both share the same architecture but differ in the input data, as described in Chapter~\ref{sec:duffing}.

\paragraph{Architecture}
The network consists of two blocks: the autoencoder pair $\phi_\theta(\bd{x}),\ \psi_\theta(\bd{z})$, and the latent dynamics $h_\theta(\bd{z})$, where $\theta$ represents the combined weights of all networks.

Both $\phi$ and $\psi$ were fully-connected networks with three layers, the input-output dimension of 128, the bottleneck-space dimension of 2, and the hidden-layer dimensions of 256. The hidden layers had ReLU activations except for the output layers of $\phi$ and $\psi$ which had linear activation.

The network $h$ was a fully-connected network with three layers, with the input and output dimension of 2 and the hidden-layer dimensions of 128. All hidden layers had ReLU activation, the output layer had a linear activation.

We used standard network classes of \texttt{pytorch}\footnote{\href{https://pytorch.org}{https://pytorch.org}} by~\cite{pytorch} to implement the networks, and we used a differentiable integrator \texttt{torchdiffeq}\footnote{\href{https://github.com/rtqichen/torchdiffeq}{https://github.com/rtqichen/torchdiffeq}} by~\cite{chen2018neuralode} for evaluating derivatives of the loss function.

\paragraph{Data} For data \textbf{snapshots} we used 6144 trajectories 10 steps long each, with a step-size $dt = 0.1$. For \textbf{collocations} we generated a set of $10^5$ 2-dimensional points $\bar{\bd{z}}_j \in U\left([-3/2,\, 3/2] \times [-1, 1]\right)$, excluded those belonging to the left (red) lobe, and projected them to the observable space $\mathcal{X}$ using the true decoder~(\ref{eq:duffing_true_decoder}). We then used those high-dimensional points $\bar{\bd{x}}_j$ as collocations.

\paragraph{Training} We trained the combined model for 400 epochs, with the learning rate of $10^{-4}$. All weights $\omega_i$ were set to 1 for a hybrid model, and $\omega_2$, $\omega_4$ were set to 0 for a data-driven model.

In each batch we had 64 trajectories and, in case of hybrid models, 640 collocation points. The rationale behind this ratio is that one trajectory contains number-of-steps snapshots of the system. Hence, to balance the amount of information that comes from both sources within a batch, we were taking the number-of-steps more collocations than trajectories for every batch. The same rationale holds for all other instances of training of a hybrid model in this paper.

\subsection{Lifted Duffing Oscillator: Far-Out Forecasting (Figure~\ref{fig:duffing_periods}).}
\label{appendix:duffing_forecast}
In this experiment we trained three models, all sharing the same architecture but differ in their loss functions, namely in the coefficients $w_i$.

\paragraph{Architecture} The networks architectures match to the one described in Appendix~\ref{appendix:duffing_pinode}.
\paragraph{Data} For data \textbf{snapshots} we used 6144 trajectories (2048 for each of the attractors) 10 steps long each, with a step-size $dt = 0.1$. For \textbf{collocations} we generated a set of $10^5$ 2-dimensional points $\bar{\bd{z}}_j \in U\left([-3/2,\, 3/2] \times [-1, 1]\right)$ and projected them to the observable space $\mathcal{X}$ using the true decoder~(\ref{eq:duffing_true_decoder}). We then used those high-dimensional points $\bar{\bd{x}}_j$ as collocations.

\paragraph{Training} We trained the combined model for 400 epochs, with the learning rate of $10^{-4}$. All weights $\omega_i$ were set to 1 for a hybrid model, $\omega_3$, $\omega_4$ were set to 0 for a data-driven model, and $\omega_1 = \omega_2 = 0$ for Physics-Informed model.

\subsection{Lifted Duffing Oscillator: Role of Non-Linear Latent Dynamics (Figure~\ref{fig:duffing_comparison}).}
\label{appendix:duffing_non_linearity}
In this experiment we trained three models: DMD, PIKN, and PINODE.

\paragraph{Architecture} The PINODE model's architecture matches to the one described in Appendix~\ref{appendix:duffing_pinode}. The PIKN model's architecture is the same except that the latent dynamics $h(z) = Lz$ is linear: it consists of one fully-connected linear layer of width 16 with no bias and no activation function. DMD model had the latent space of 16; the implementation is faithful to the original works \cite{kutz2016dynamic}.
\paragraph{Data} The datasets for both trajectories and collocations match to the ones described in Apendix~\ref{appendix:duffing_forecast}

\paragraph{Training} We trained the combined model for 400 epochs, with the learning rate of $10^{-4}$. All weights $\omega_i$ were set to 1 for both PIKN and PINODE. DMD model used no collocations, whereas PIKN and PINODE used both trajectories and collocations.

\subsection{Burgers' Equation: Compressibility (Figure~\ref{fig:burgers_compressibility})}
\label{appendix:burgers_compressibility}
In this section we study how efficiently different ROMs use the same size of the latent space.

\paragraph{Architecture} In \textbf{PINODE} and \textbf{PIKN} models, both $\phi$ and $\psi$ were fully-connected networks with three layers, the input-output dimension of 128 and the hidden-layer dimensions of 512. The hidden layers had ReLU activation except for the output layers of $\phi$ and $\psi$ which had linear activation.
The size of the latent space was varying from 2 to 512, see the x-axis of Figure~(\ref{fig:burgers_compressibility}). 
In \textbf{PINODE}, the network $h$ was a fully-connected network with three layers with the hidden-layer dimensions of 512. All hidden layers had ReLU activation, the output layer had a linear activation. For \textbf{PIKN} the network had one layer of the latent space size with no bias and linear activation. In other words, $h(z) = Az$.
For \textbf{DMD} we used a classic algorithm from~\cite{kutz2016dynamic}, and set the number of DMD modes to be equal to the size of the latent dimension. 
\paragraph{Data}For \textbf{snapshots}, we generated 16384 trajectories of the system for our train dataset and 300 trajectories for our test dataset. Each trajectory had 40 time-steps with $dt = 0.1$. We used randomly-generated functions from Equation~\ref{eq:burger_initial_condition} as initial conditions for the trajectories. We used only first 20 time-steps of train trajectories for training. We used the first 20 steps of the test trajectories to evaluate \textit{interpolation} performance of the model, and the next 20 time-steps for evaluating \textit{extrapolation} performance. All performance measures on Figure~(\ref{appendix:burgers_compressibility}) are based on the test dataset.
For \textbf{collocations}, we used Equation~(\ref{eq:burger_collocations}) to generate $10^5$ collocation points. 
\paragraph{Training} We trained models for 500 epochs, with the learning rate of $10^-4$. All weights $\omega_i$ were set to 1 for a hybrid model, and $\omega_3$, $\omega_4$ were set to 0 for a data-driven model. Every batch contained 64 trajectories and 1280 (that is, 64*20) collocation points, with the same rationale as in Appendix~\ref{appendix:duffing_pinode}.

\subsection{Burgers' Equation: Data-vs-Collocations (Figure \ref{fig:burger_data_vs_collocations})}
In this section we study the relative impact of data and collocations as training datasets to the performance of the resulting models.
\label{appendix:burgers_data_vs_collocations}
\paragraph{Architecture} In \textbf{PINODE} models both $\phi$ and $\psi$ were fully-connected networks with three layers, the input-output dimension of 128, the latent-space dimension of 16 and the hidden-layer dimensions of 512. The hidden layers had ReLU activation except for the output layers of $\phi$ and $\psi$ which had linear activation. The network $h$ was a fully-connected network with three layers with the input-output dimension of 16, and the hidden-layer dimensions of 512. All hidden layers had ReLU activation, the output layer had a linear activation.

\paragraph{Data} For \textbf{snapshots}, we generated 2048 trajectories of the system for our train dataset. Each trajectory had 20 time-steps with $dt = 0.1$. We used randomly-generated functions from Equation~\ref{eq:burger_initial_condition} as initial conditions for the train trajectories. 
For \textbf{collocations}, we used Equation~(\ref{eq:burger_collocations}) to generate $65536$ collocation points. 
We generated 300 trajectories for our test datasets, 100 per kind of initial conditions from Figure~(\ref{fig:burgers_examples_of_ics}). We used all 40 steps of the test trajectories to evaluate the performance of the models. All performance measures on Figure~(\ref{appendix:burgers_data_vs_collocations}) are based on this test dataset.

\paragraph{Training} We trained models for 500 epochs, with the learning rate of $10^-4$. All weights $\omega_i$ were set to 1 for a hybrid model, and $\omega_3$, $\omega_4$ were set to 0 for a data-driven model. Every batch contained 64 trajectories and 1280 (that is, 64*20) collocation points, with the same rationale as in Appendix~\ref{appendix:duffing_pinode}.

\subsection{Burgers' Equation: Robustness to Noise (Figure~\ref{fig:burger_noise})}
In this section we examine robustness to noise for four models: PINODE (Data-Driven, Physics-Informed, Hybrid). We also add DMD to the comparison as a reference point.
\label{appendix:burgers_noise}
\paragraph{Architecture} All PINODE models share the same architecture as described in Appendix~\ref{appendix:burgers_data_vs_collocations}.

\paragraph{Data} We start with the same data as described in Appendix~\ref{appendix:burgers_data_vs_collocations}. Then we apply 11 different levels of Gaussian noise with the mean 0 and the variances distributed log-uniformly between $[10^{-4}, 10^1]$. We only apply noise to the train snapshots; the test snapshots are noise-free.

\paragraph{Training} We trained models for 500 epochs, with the learning rate of $10^-4$. All weights $\omega_i$ were set to 1 for a hybrid model, $\omega_1$, $\omega_2$ , and $\omega_3$, $\omega_4$ were set to 0 for a data-driven model. Every batch contained 64 trajectories and 1280 (that is, 64*20) collocation points, with the same rationale as in Appendix~\ref{appendix:duffing_pinode}.

### 2023-04-11 msr3-theory.tex
Add appendices; connect MSR3 theory and practice definitions; compiles with NO WARNINGS

@@
-\subsection{Relaxation of the mixed-effects variable selection model}
-\label{sec:relax}
-
-Our strategy for obtaining approximate solutions to the
-mixed-effects variable selection problem \eqref{eq:lme_loss_original_in_x} is
-motivated by the sparse relaxed regularization regression (SR3) strategy
-developed in \cite{Zheng2019SR3}. That is, $0\le\gam$ is relaxed
-using the perspective of the negative log, i.e %log-barrier function
-$\map{\vphi}{\R^q\times\R}{\R\cup\{\infty\}}$ given by
-\[
-\vphi(\gam,\mu):=
-\begin{cases}
--\mu\sum_{i=1}^q\ln (\gam_i/\mu)&,\ \mu>0,\\
-\del_{\R^q_+}(\gam)&,\ \mu=0,\\
-+\infty&,\ \mu<0.
-\end{cases}
-\]
-The mapping $\vphi$ is known to be a closed proper convex function and, 
-for $\mu>0$, it is
-essentially equivalent to the well-known log-barrier function in \eqref{eq:log_barrier_relaxation_for_ip}. 
-%since in this case
-%\[
-%\phi_\mu(\gam)=q\mu\ln(\mu)-\mu\sum_{i=1}^q\ln (\gam_i).
-%\]
-For more information on the perspective mapping, its calculus,
-and perspective duality, we refer the reader to \cite{ABDFM18,ABF13}.
-%\red{Should we mention the perspective of $\psi(\gam)=-\ln (\gam)$, or perhaps the closure of the epi-limit? The connection to the perspective map might be useful since
-%we get the joint convexity in $\mu$ and $\gam$.}
-We call $\eta$ the coupling parameter and $\mu$ the log-barrier parameter
-and write $\phimu(\cdot):=\vphi(\cdot,\mu)$.
-%The parameter $0\le\blam$ is fixed and is related to the weak convexity of $\LL$
-%discussed in Lemma \ref{lem:LL weak cvx}.
-%}
-The relaxed problem employs 
-auxiliary variables $\tbg$ 
-and relaxation parameters $0\le\eta$ and $0\le \mu$
-to obtain the problem 
-\begin{equation}\label{eq:relax} 
-\begin{aligned}%\eq{\label{eq:lme_loss_relaxed}
-%\nu_\emu:=	
-\min_{\bg, \tbg} & \LL(\beta, \gamma) 
-	+\phi_\mu(\gam)+\kappa_\eta(\beta-\tbeta,\gam-\tgam)
-%\frac{\eta}{2}\norm{\beta - \tbeta}^2 + \frac{\blam+\eta}{2}\norm{\gamma - \tgamma}^2  
-+ R(\tbeta, \tgamma) 
-\\
-\text{s.t. } & \tgamma \geq 0\ .
-\end{aligned}
-\end{equation}
-We rewrite \eqref{eq:relax} %\eqref{eq:lme_loss_relaxed} 
-so as to separate the smooth and nonsmooth components to obtain
-\eq{ \label{eq:relax2}
-    \min_{\bg, \tbg} & \LL_{\emu}(\bg,\tbg) + R\tbg +\del_{\R^q_+}(\tgam)\ ,
-%    \\
-%    \text{s.t. } & 0\le\tgam,
-}
-where 
-%$x=(\beta, \gamma)$, 
-%$w=(\tbeta, \tgamma)$, $\CC=\R^p\times\R^q_+$, and
-%$\LL_\eta$ is defined in \eqref{eq:defn L sub eta}.
-\eq{\label{eq:LLlam}
-    \LL_{\emu}(\bg, \tbg) := 
-    \LL\bg + \phimu(\gam)+\keta(\beta - \tbeta, \gamma - \tgamma).
-}
-Observe that, for all $\mu,\eta\in\R_+$, $\LL_{\emu},\ \nabla \LL_{\emu}$ and $\nabla^2 \LL_{\emu}$ are continuous
-on $(\Rp\times\dom(\phimu))\times(\Rp\times\Rq)$ 
-(see Appendix \ref{appendix:derivatives_of_lmm}) 
-so that $\LL_{\emu}$ is smooth on its domain.
-As in \cite{Zheng2019SR3}, we use the 
-decoupling to write \eqref{eq:relax2}
-as an iterated optimization problem over the smooth components of the objective.
-This yields a representation of the form
-\begin{equation}
-    \label{eq:relax3}
-    \min_{\tbg} \uemu\tbg + R\tbg+\del_{\Rqp}(\tgam),
-\end{equation}
-where %$w=(\tbeta,\tgam)$ and
-\begin{equation}
-    \label{eq:value_function_definition}
-    \uemu\tbg := \min_{\bg } 
-    \LL_{\emu}(\bg, \tbg)\ .
-%    \hLL(\bg,\tgam)+ %\del_{\Rqp}(\gam)+
-%    \frac{\eta}{2}\norm{\bg-\tbg}^2
-    %\LL(x) + \lambda\|x - w\|_2^2 
-\end{equation}
-This is the formulation of the mixed-effects variable selection problem we study.
-Our focus is on the optimal value function $\uemu$ which captures global 
-variational information about the function $\LL$ over its domain.
-We show that $\uemu$ has a locally Lipschitz continuous gradient
-and that the evaluation of $\uemu$ and $\nabla \uemu$ is accomplished
-by optimizing a well conditioned strongly convex function. This
-allows us to apply the PGD algorithm to the function $\uemu$ rather than
-the function $\LL$. Our numerical studies show that the global information 
-captured by $\uemu$ significantly improves both the accuracy of the solution
-obtained and the overall numerical efficiency of the algorithm.
 %gives the problem structure used in the design of our
@@
 %\section{Prox Gradient Descent}
-\subsection{Proximal Gradient Descent for $\uemu\tbg + R\tbg+\del_{\Rqp}(\tgam)$} 
+\subsection{Proximal Gradient Descent for \texorpdfstring{$\uemu\tbg + R\tbg+\del_{\Rqp}(\tgam)$}{}} 
 \label{sec:pgd}
@@
 %{\bf Algorithm 1:} Proximal Gradient Descent with fixed stepsize (PGD)
-
-\begin{algorithm}[H]
-\SetAlgoLined
-$\tw = \tw^0$, $\gammax>\tgam^0$, $\alpha < \frac{2}{L}$ 
-for $L>0$.\\
- \While{not converged and $\tgam\le\gammax$}{
-    $\tw^+ = \prox_{\alpha \tR }(\tw - \alpha \nabla \uemu (\tw))$\\
- }
-\caption{\label{alg:MSR3}Proximal Gradient Descent for 
-$\Phi_\emu$}
-\end{algorithm}
-\medskip
 
@@
 
-%\begin{figure}[H]
+%\begin{figure}[t]
 %    \centering
@@
 
-\subsection{The Smoothness of $u_\emu$}\label{sec:smoothness}
+\subsection{The Smoothness of \texorpdfstring{$u_\emu$}{}}\label{sec:smoothness}
 We investigate the relationship between the problems \eqref{eq:lme_loss_original_in_x}
@@
 \begin{lemma}[$\LL+\phimu$ is Weakly Convex]\label{lem:LL weak cvx}
-Let $\LL$ be as given in \eqref{eq:lme_gam}. Then
+Let $\LL$ be as given in \eqref{eq:lme_diagonal_setup}. Then
 \begin{equation}\label{eq:hess LL}
@@
 In particular, for $\mu=0$, they converge to solutions of 
-\eqref{eq:vs_lme}.
+\eqref{eq:lme_loss_original_in_x}.
 
@@
 there is an optimal solution $\xmu$ 
-to \eqref{eq:log-barrier problem} %\eqref{eq:vs_lme} 
+to \eqref{eq:log-barrier problem} %\eqref{eq:lme_loss_original_in_x} 
 yielding an optimal value of $\vmu$ for which 
@@
 the objective in \eqref{eq:log-barrier problem} is $\tLL+\phimu$
-and the objective in \eqref{eq:vs_lme} is $\tLL$
+and the objective in \eqref{eq:lme_loss_original_in_x} is $\tLL$
 with $(\beta_0,\gam_0)$ a solution to \eqref{eq:lme_loss_original_in_x}
@@
 Consequently, the continuity of $\tLL$ on $\Rp\times\Rqp$ implies
-that $(\bbeta,\bgam)$ solves \eqref{eq:vs_lme}.
+that $(\bbeta,\bgam)$ solves \eqref{eq:lme_loss_original_in_x}.
 \end{proof}
@@
 
-\subsubsection{The continuity and differentiability of $\uemu$}
+\subsubsection{The continuity and differentiability of \texorpdfstring{$\uemu$}{}}
 
@@
 %have the same form as the unrelaxed problem  
-%\eqref{eq:vs_lme} 
+%\eqref{eq:lme_loss_original_in_x} 
 %with the optimal value function $\uemu$ replacing the likelihood $\LL$.
@@
 
-\subsubsection{The Lipschitz Continuity of $\nabla \uemu\tbg$}
+\subsubsection{The Lipschitz Continuity of \texorpdfstring{$\nabla \uemu\tbg$}{}}
 
@@
 
-\subsection{Convergence of the PGD Algorithm for $\Phi_\emu$}
+\subsection{Convergence of the PGD Algorithm for \texorpdfstring{$\Phi_\emu$}{}}
 \label{sec:convergence pgd}
@@
  limit point of solutions \eqref{eq:log-barrier problem} as $\mu\downarrow 0$
- is a solution to the variable selection problem \eqref{eq:vs_lme}. 
+ is a solution to the variable selection problem \eqref{eq:lme_loss_original_in_x}. 
 These results suggest a range of numerical approachs to obtaining approximate 
-solutions to the target problem \eqref{eq:vs_lme}.
+solutions to the target problem \eqref{eq:lme_loss_original_in_x}.
  %%

### 2023-04-11 msr3.tex
Add appendices; connect MSR3 theory and practice definitions; compiles with NO WARNINGS

@@
 
-\begin{figure}[H]
+\begin{figure}[t]
     \centering
@@
 %\eq{
-%	\label{eq:beta_formula}
-\[
+%	
+\[
+\label{eq:beta_formula}
 	\beta(\Gamma) = \argmin_{\beta}\LL(\beta, \Gamma) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i.
@@
 
- \begin{figure}[H]
+ \begin{figure}[t]
     \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[t]
     \centering
@@
 
-To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we define a value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
-\eq{
+To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we, for this section alone, define a simplified version of the value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
+\eq{
+	\label{eq:value_function_definition_only_eta}
 		u_\eta(w) & = \min_x \ \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w)\\
@@
 %To apply IP methods to our problem class, we first 
-First, the inequality constraint $0\le \gamma$ is relaxed using a log-barrier penalty, obtaining a minimization problem for a relaxed objective $\mathcal{L}_{\mu,\eta}$: 
-\eq{
-	\label{eq:log_barrier_relaxation_for_ip}
-	\min_{\beta, \gamma} &\left\{\mathcal{L}_{\mu,\eta}(\beta, \gamma) :=  \mathcal{L}(\beta, \gamma) + \kappa_{\eta}(\beta - \hat \beta, \gamma - \hat \gamma) - \mu\sum_{i=1}^{q}\ln(\gamma_i) \right\}.
-}
-Here the log-barrier penalty approximates the indicator function to the positive orthant as $\mu$ decreases; indeed, the function $\gamma\mapsto\mu\ln(\gamma)$ epi-converges to the indicator function $\delta_{\mathbb{R}^n_+}(\gamma)$ as $\mu \downarrow 0$ (\cite{rockafellar2009variational}).
+First, the inequality constraint $0\le\gam$ is relaxed
+using the perspective of the negative log, i.e %log-barrier function
+$\map{\vphi}{\R^q\times\R}{\R\cup\{\infty\}}$ given by
+\[
+\vphi(\gam,\mu):=
+\begin{cases}
+-\mu\sum_{i=1}^q\ln (\gam_i/\mu)&,\ \mu>0,\\
+\del_{\R^q_+}(\gam)&,\ \mu=0,\\
++\infty&,\ \mu<0.
+\end{cases}
+\]
+The mapping $\vphi$ is known to be a closed proper convex function and, 
+for $\mu>0$, it is
+essentially equivalent to the well-known log-barrier function.
+%since in this case
+%\[
+%\phi_\mu(\gam)=q\mu\ln(\mu)-\mu\sum_{i=1}^q\ln (\gam_i).
+%\]
+For more information on the perspective mapping, its calculus,
+and perspective duality, we refer the reader to \cite{ABDFM18,ABF13}.
+%\red{Should we mention the perspective of $\psi(\gam)=-\ln (\gam)$, or perhaps the closure of the epi-limit? The connection to the perspective map might be useful since
+%we get the joint convexity in $\mu$ and $\gam$.}
+We call $\eta$ the coupling parameter and $\mu$ the log-barrier parameter
+and write $\phimu(\cdot):=\vphi(\cdot,\mu)$.
+%The parameter $0\le\blam$ is fixed and is related to the weak convexity of $\LL$
+%discussed in Lemma \ref{lem:LL weak cvx}.
+%}
+The relaxed problem employs 
+auxiliary variables $\tbg$ 
+and relaxation parameters $0\le\eta$ and $0\le \mu$
+to obtain the problem 
+\begin{equation}\label{eq:relax} 
+\begin{aligned}%\eq{\label{eq:lme_loss_relaxed}
+%\nu_\emu:=	
+\min_{\bg, \tbg} & \LL(\beta, \gamma) 
+	+\phi_\mu(\gam)+\kappa_\eta(\beta-\tbeta,\gam-\tgam)
+%\frac{\eta}{2}\norm{\beta - \tbeta}^2 + \frac{\blam+\eta}{2}\norm{\gamma - \tgamma}^2  
++ R(\tbeta, \tgamma) 
+\\
+\text{s.t. } & \tgamma \geq 0\ .
+\end{aligned}
+\end{equation}
+
+We rewrite \eqref{eq:relax} %\eqref{eq:lme_loss_relaxed} 
+so as to separate the smooth and nonsmooth components to obtain
+\eq{ \label{eq:relax2}
+    \min_{\bg, \tbg} & \LL_{\emu}(\bg,\tbg) + R\tbg +\del_{\R^q_+}(\tgam)\ ,
+%    \\
+%    \text{s.t. } & 0\le\tgam,
+}
+where 
+%$x=(\beta, \gamma)$, 
+%$w=(\tbeta, \tgamma)$, $\CC=\R^p\times\R^q_+$, and
+%$\LL_\eta$ is defined in \eqref{eq:defn L sub eta}.
+\eq{\label{eq:LLlam}
+    \LL_{\emu}(\bg, \tbg) := 
+    \LL\bg + \phimu(\gam)+\keta(\beta - \tbeta, \gamma - \tgamma).
+}
+Observe that, for all $\mu,\eta\in\R_+$, $\LL_{\emu},\ \nabla \LL_{\emu}$ and $\nabla^2 \LL_{\emu}$ are continuous
+on $(\Rp\times\dom(\phimu))\times(\Rp\times\Rq)$ 
+(see Appendix \ref{appendix:derivatives_of_lmm}) 
+so that $\LL_{\emu}$ is smooth on its domain.
+As in \cite{Zheng2019SR3}, we use the 
+decoupling to write \eqref{eq:relax2}
+as an iterated optimization problem over the smooth components of the objective.
+This yields a representation of the form
+\begin{equation}
+    \label{eq:relax3}
+    \min_{\tbg} \uemu\tbg + R\tbg+\del_{\Rqp}(\tgam),
+\end{equation}
+where %$w=(\tbeta,\tgam)$ and
+\begin{equation}
+    \label{eq:value_function_definition}
+    \uemu\tbg := \min_{\bg } 
+    \LL_{\emu}(\bg, \tbg)\ .
+%    \hLL(\bg,\tgam)+ %\del_{\Rqp}(\gam)+
+%    \frac{\eta}{2}\norm{\bg-\tbg}^2
+    %\LL(x) + \lambda\|x - w\|_2^2 
+\end{equation}
+This is the formulation of the mixed-effects variable selection problem we study. 
+Notice that this value function differs from a simplified 
+definition in \eqref{eq:value_function_definition_only_eta}: we replaced 
+the linearity constraint with a perspective function and got another parameter $\mu$ due to it.
+The log-barrier penalty approximates the indicator function to the positive orthant as $\mu$ decreases; indeed, the function $\gamma\mapsto\mu\ln(\gamma)$ epi-converges to the indicator function $\delta_{\mathbb{R}^n_+}(\gamma)$ as $\mu \downarrow 0$ (\cite{rockafellar2009variational}).
 The penalty (homotopy) parameter $\mu$ is progressively decreased to $0$ as the algorithm proceeds as described below.
 %in the limit minimizing $\mathcal{L}_{\eta,\mu}$ gives the right $x^+$ update in Algorithm~\ref{alg:pgd_for_value}. 
-The existence of solutions for the problem~(\ref{eq:log_barrier_relaxation_for_ip}) for any positive $\mu$ is shown in \cite[Theorem 5]{Theory1}, and the convergence of solutions to the MSR3 solution as $\mu\downarrow 0$ is shown in~\cite[Theorem 7]{Theory1}.
+ We refer to \eqref{eq:value_function_definition} when we say ``value function'' from now on.
+
+Our focus is on the optimal value function $\uemu$ which captures global 
+variational information about the function $\LL$ over its domain.
+We show that $\uemu$ has a locally Lipschitz continuous gradient
+and that the evaluation of $\uemu$ and $\nabla \uemu$ is accomplished
+by optimizing a well conditioned strongly convex function. This
+allows us to apply the PGD algorithm to the function $\uemu$ rather than
+the function $\LL$. Our numerical studies show that the global information 
+captured by $\uemu$ significantly improves both the accuracy of the solution
+obtained and the overall numerical efficiency of the algorithm.
+
+To motivate the reader, the theoretical results that support the design of Algorithm \ref{alg:MSR3} go as follows.
+The existence of solutions for the problem~(\ref{eq:value_function_definition}) for any positive $\mu$ is shown in \cite[Theorem 5]{Theory1}, and the convergence of solutions to the MSR3 solution as $\mu\downarrow 0$ is shown in~\cite[Theorem 7]{Theory1}.
 Finally, \cite[Theorem 6]{Theory1} shows that the MSR3 relaxation is consistent with respect to the barrier, so that as the MSR3 parameter $\eta \uparrow \infty$,
  limit points of global solutions to the former are global solutions to the latter. 
- However, in the applications considered here,
- the empirical studies in Sections \ref{sec:synthetic} and \ref{sec:real}
+ However, the empirical studies in Sections \ref{sec:synthetic} and \ref{sec:real}
  indicate that one does not need to make $\eta$ particularly large in order to accurately identify the correct sparsity pattern.
-
-%The Lagrangian for~\eqref{eq:log_barrier_relaxation_for_ip} is obtained by dualizing the inequality $\gamma \geq 0$ constraint: 
-% Define the objective for \eqref{eq:log_barrier_relaxation_for_ip} by
+These results were first published in the joint works \cite{sholokhov2022relaxation, aravkin2022jimtheory}, constitute the original work of the author of the thesis, and were mostly taken verbatim per the co-authors'  permission.
+
+%The Lagrangian for~\eqref{eq:value_function_definition} is obtained by dualizing the inequality $\gamma \geq 0$ constraint: 
+% Define the objective for \eqref{eq:value_function_definition} by
 % \eq{
@@
 ``$\odot$'' denotes the Hadamard (or simply element-wise) product.
-%Together with the remaining optimality conditions, we obtain a set of nonlinear equations  that form the the KKT system for~\eqref{eq:log_barrier_relaxation_for_ip}:
+%Together with the remaining optimality conditions, we obtain a set of nonlinear equations  that form the the KKT system for~\eqref{eq:value_function_definition}:
 The complete set of optimality conditions for 
-\eqref{eq:log_barrier_relaxation_for_ip} can now be written as
+\eqref{eq:value_function_definition} can now be written as
 \eq{\label{eq:IP equations}
@@
 so that the update
-\[
+\eq{
+\label{eq:ip_one_step}
 \begin{pmatrix}
@@
 \end{pmatrix}
-\]
+}
 satisfies the conditions
@@
 \SetAlgoLined
-$w = w_0$ \\
- \While{not converged}{
-    $x^+ \; \mbox{satisfies }\;  \|G_{\mu,\eta}(v^+, x^+)\| \leq \text{\texttt{tol}}, \; \mu \leq \text{\texttt{tol}} $\\
+$w^0 = (\tbeta^0, \tgamma^0)$, $x^0 = (\beta^0, \gamma^0)$, $\gammax>\tgam^0$, $\alpha < \frac{2}{L}$ 
+for $L>0$.\\
+ \While{not converged and $\tgam\le\gammax$}{
+    \text{Find } $x^+$ \text{ using \eqref{eq:ip_one_step}}$\mbox{ which satisfies }\;  \|G_{\mu,\eta}(v^+, x^+)\| \leq \text{\texttt{tol}}, \; \mu \leq \text{\texttt{tol}} $\\
     $w^+ = \prox_{\alpha^{-1} R}(x^+)$
@@
 \end{algorithm}
+\medskip
 
@@
 accuracy of MSR3.  
-More detailed results are in the Table \ref{table:detailed_comparison_of_algorithms} of Appendix \ref{ch:appendix_detailed_comparison}.}
+More detailed results are in the Table \ref{table:detailed_comparison_of_algorithms} of Appendix \ref{appendix:detailed_comparison}.}
 \end{table}

### 2023-04-11 phd_thesis.tex
Add theory paper to the thesis
Add appendices; connect MSR3 theory and practice definitions; compiles with NO WARNINGS

@@
 
-\section{Description of Datasets}
-\subsection{GBD Bullying Data}
+\section{Derivation of Selected Proximal Operators from Table \ref{table:proxes}}
+\label{appendix:proxes}
+\input{appendix/proxes}
+
+\section{Existence of Minimizers (Theorems~\ref{thm:basic existence} and~\ref{thm:basic existence2})}
+\label{adx:basic existence}
+\input{appendix/existence_of_solutions_for_lme}
+
+\section{Lipschitz-constant for Likelihood of a Linear Mixed-Effects Model}
+\label{appendix:lipschitz_constant}
+\input{appendix/lipschitz_constant}
+
+\section{Detailed Results from Simulation from Table \ref{table:comparison_of_algorithms}}
+\label{appendix:detailed_comparison}
+\begin{table}[t]
+    \centering
+    \resizebox{\columnwidth}{!}{\input{tables/performance_table_long_current}}
+    \caption{Comparison of performance of algorithms}
+    \label{table:detailed_comparison_of_algorithms}
+\end{table}
+
+\section{Description of Datasets: GBD Bullying Data}
 \input{appendix/bullying_features_description}
 
-\subsection{COVID-19 Contact Rate Forecasting Data}
-\input{tables/covid_groups_description}
-\input{tables/covid_coefficients}
+%\subsection{COVID-19 Contact Rate Forecasting Data}
+%\input{tables/covid_groups_description}
+%\input{tables/covid_coefficients}
+
+\section{Detailed Design of Experiments for PINODE}
+\input{appendix/pinode_appendix}
 

### 2023-04-12 appendix/pinode_appendix.tex
Add PIKN chapter; Add todos across the doc

@@
 
+\subsection{Burger's Equation: Compressibility of Linear Latent Space for PIKN}
+\label{appendix:pikn_burgers}
+\paragraph{Architecture} 
+In the PIKN architecture for Burger's equation, the encoder is implemented as a a feed-forward network with the size of input layer set to be equal 128 (the number of spatial grid-points).
+It has two hidden layers with 512 neurons.
+The size of the output layer of the encoder -- the size of the latent dimension -- varies from 16 to 256 to study the latent space compressibility (see Figure~\ref{fig:pikn_burgers_compression}).
+The decoder's architecture mirrors the architecture of the encoder with the sizes of the inputs and outputs switched.
+An Adam optimizer has been applied for $500$ epochs for training.
+We use an adaptive learning rate: initially set to $10^{-4}$, it keeps decreasing by a factor of $0.5$ if no improvement has been made over the recent $20$ epochs, until it hits the minimal value of $10^{-7}$.
+\\
+Each network was trained in two different ways.
+Namely, we set different values for the weights $\omega_1, \omega_2, \omega_3, \omega_4$ in the loss function, leading to two different training regimes:
+\begin{itemize}
+    \item $\omega_1=0, \omega_2=0, \omega_3=1, \omega_4=1$. These settings lead to a purely data-driven learning; it corresponds to green lines on Figure~\ref{fig:pikn_burgers_compression}.
+    \item $\omega_1=1, \omega_2=1, \omega_3=1, \omega_4=1$.
+These settings represent a model that uses both data trajectories and collocations (a hybrid model); it corresponds to blue lines on Figure~\ref{fig:pikn_burgers_compression}.
+\end{itemize}
+
+\paragraph{Training} 
+We sample $80000$ functions $\{u^{(1)}, u^{(2)}, \dots, u^{(80000)}\}$ to use them as collocations.
+Each function $u^{(j)}$ is a superposition of the Fourier modes that satisfy the periodic boundary conditions.
+We call them "harmonic" initial conditions, five examples of which are displayed on the middle-right pane of Figure~\ref{fig:pikn_burgers_compression}.
+We used $\sin{(kx)}$ and $\cos{(kx)}$ for $k=0, 1, 2, \dots$ evaluated on $x \in [-\pi; \pi]$-interval as basis functions.
+The value of $k$ is restricted to be no greater than $10$.
+To evaluate $u_t$ at each collocation point we evaluated the spatial derivatives $\{(u_{x}^{(i)}, u_{xx}^{(i)})\}_{i=1}^{80000}$ using a numerical spectral method.
+% \paragraph{Train Data}
+
+For training trajectories, we use simulation data obtained from a solver base on spectral method.
+We run the simulation with $1024$ different initial states.
+Each initial state of $u$ is a Gaussian (bell-) curve with mean 0 and a randomly generated variance $\sigma \sim U(0.1,1)$, evaluated on $[-\pi; \pi]$-interval.
+Then we sample snapshots of the state $u$ with a temporal gap $\Delta t=0.1$ for $20$ steps (i.e.
+$p=20$).
+An example trajectory is displayed on the top-right pane of Figure~\ref{fig:pikn_burgers_compression}.
+
+% \paragraph{Test Data} 
+To evaluate and compare the performance of both models we use 200 trajectories: 100 trajectories with harmonic ICs and 100 trajectories with Gaussian ICs generated exactly in the same way as described above.
+Then we sample snapshots of the state $u$ with a temporal gap $\Delta t=0.1$ for $20$ steps (i.e.
+$p=20$).
+
+\paragraph{Results} 
+The results are aggregated in Table~\ref{table:pikn_burgers_performance} and visualized on Figure~\ref{table:pikn_burgers_performance}.
+
+\begin{table}[t!]
+    \centering
+    \resizebox{\columnwidth}{!}{\input{tables/pikn_performance_table}}
+    \caption{Performance measures for Burger's experiment depending on the dimensions of the latent space (Figure~\ref{fig:pikn_burgers_compression})}
+    \label{table:pikn_burgers_performance}
+\end{table}
+
 \subsection{Burgers' Equation: Data-vs-Collocations (Figure \ref{fig:burger_data_vs_collocations})}

### 2023-04-12 msr3.tex
Add PIKN chapter; Add todos across the doc

@@
+\edit{Rewrite the MSR3 abstract into a summary of whole work with citations}
 Linear Mixed-Effects (LME) models are a fundamental tool for modeling correlated data, 
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \centering
-\input{tables/table_glmm_current.tex}
+\resizebox{\columnwidth}{!}{\input{tables/table_glmm_current.tex}}
+
 \caption{\label{table:glmmlasso} Comparison of performance of MSR3-Fast for $\ell_1$ regularizer vs \texttt{glmmLasso}.
@@
 
-\subsection{Experiments on Real Data}
+\subsection{Sensitivity Analysis}
+\label{ch:msr3_sensitivity_analysis}
+\edit{Add sensitivity analysis from MSR3 practice paper}
+
+\subsection{Application to Real Data: Anxiety and Depression as a Result of Bullying Victimization}
 \label{sec:real}
@@
 
+\subsection{Application to Real Data: COVID-19 Transmission Factor}
+\edit{Add Covid example from my general}
+
 \section{Theoretical Analysis}
@@
 \section{Software Implementation}
+\edit{Move pysr3 software paper here}
 To ensure reproducibility of this research, all new algorithms have been implemented as a part of the \texttt{pysr3}\footnote{Available at \href{https://github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}} library.

### 2023-04-12 pinode.tex
Add PIKN chapter; Add todos across the doc

@@
 \section{Introduction}
-
+\edit{Write a big-picture intro to machine learning for physics}
+\edit{Replace mentions of paper with work}
+\subsection{Related Work}
+\edit{Split original intro into prior works paragraphs: DL for Ph, ROMs, PIML}
+\edit{Add physics-informed ML overview (PINN, DeepONet, NFO)}
         Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs).
@@
 We demonstrate that such an approach not only reduces the need for large training data-sets and produces highly-accurate and long-term stable models, but also leads to the discovery of more compact latent spaces, which is especially important for applications in compressed sensing and control.
+        \edit{Add Koopman example to the intro}
+
+        \edit{Add compressive sensing application to the intro}
 
@@
 \paragraph{Architecture} In this work we model $\psi$, $\psi^{-1}$, and $\bd{h}$ with fully-connected neural networks $\psi_\theta$, $\phi_\theta$, and $h_\theta$, respectively.
-Specifically, the pair ($\psi$, $\psi^{-1}$) is modelled with an auto-encoder $(\psi_\theta, \phi_\theta)$, and $\bd{h}$ is modelled with a fully-connected network $h_\theta$.
+Specifically, the pair ($\psi$, $\psi^{-1}$) is modeled with an auto-encoder $(\psi_\theta, \phi_\theta)$, and $\bd{h}$ is modeled with a fully-connected network $h_\theta$.
 Figure~\ref{fig:architecture} visualizes the architecture of the model.
@@
 In other words, $x_j$ should be an attainable state of the system.
-Collocation points outside of $\XX$ may downgrade the performance of the autoencoder by forcing it to be an invertible function on a domain outside of $\XX$ where a true mapping $\psi$ operates on.
+Collocation points outside of $\XX$ may downgrade the performance of the auto-encoder by forcing it to be an invertible function on a domain outside of $\XX$ where a true mapping $\psi$ operates on.
 \end{enumerate}
@@
 
-To the best of our knowledge, this is the first framework that combines non-linear latent-dynamics (Neural ODE), autoencoders, and a physics-informed loss term~(\ref{eq:loss_physics_informed}).
+To the best of our knowledge, this is the first framework that combines non-linear latent-dynamics (Neural ODE), auto-encoders, and a physics-informed loss term~(\ref{eq:loss_physics_informed}).
 Thus, we call our framework \textit{Physics-Informed Neural ODE}, or PINODE.
@@
 We show how a non-linear latent dynamics $\bd{h}(\bd{z})$ overcomes the limitations of DMD and Koopman networks from \cite{liu2022physics} by handling multiple basins of attraction within one model.
-We also show that using physics-informed loss is sufficient for reconstructing the behaviour for basins of attraction that are not represented by the data.
+We also show that using physics-informed loss is sufficient for reconstructing the behavior for basins of attraction that are not represented by the data.
 Finally, we demonstrate that a purely data-driven model may be highly-accurate short-term and highly unstable long-term, even when the data is abundant, and show that the physics-informed approach improves long-term stability of such models by multiple orders of magnitude.
@@
 Burgers' equation is a PDE occurring in applications to acoustics, gas and fluid dynamics, and traffic flows \cite{burgers1948mathematical}.
-When $\nu$ is significantly smaller than one, the system exhibits strong non-linear behaviour and is called ``advection-dominated'', otherwise when $\nu$ is large the system is called ``diffusion-dominated''.
+When $\nu$ is significantly smaller than one, the system exhibits strong non-linear behavior and is called ``advection-dominated'', otherwise when $\nu$ is large the system is called ``diffusion-dominated''.
 In the case of the former, linear projection methods such as POD become inaccurate as the true solution space has a slow decaying Kolmogorov n-width, manifesting itself in slow decaying singular values \cite{peherstorfer2022breaking}.
@@
     \includegraphics[width=0.9\textwidth]{figures/compressibility.pdf}
-    \caption{PINODE Hybrid model utilized the latent space dimension 5 times more efficiently by MSE than PIKN Hybrid model when modelling low-viscosity (highly-nonlinear) Burger's equation (left frame).
+    \caption{PINODE Hybrid model utilized the latent space dimension 5 times more efficiently by MSE than PIKN Hybrid model when modeling low-viscosity (highly-nonlinear) Burger's equation (left frame).
 The difference in performance grows to x100, when forecasting two times farther than the train period (central frame).
-The reason behind PIKN's long-term instability is the precsence of eigenvalues with poitive real part in the latent dynamics matrix (right frame).}
+The reason behind PIKN's long-term instability is the presence of eigenvalues with positive real part in the latent dynamics matrix (right frame).}
     \label{fig:burgers_compressibility}
@@
 First, we notice that DMD, despite achieving small loss values on training data ($\sim 10^{-3}$), does not perform well on test data.
-This observation is consistent with earlier works~(\cite{kalur2021robust,kutz2016dynamic}); it illustrates well that a combination of a linear encoder and a linear latent dynamics may not be sufficient for modelling highly-nonlinear phenonmena.
+This observation is consistent with earlier works~(\cite{kalur2021robust,kutz2016dynamic}); it illustrates well that a combination of a linear encoder and a linear latent dynamics may not be sufficient for modeling highly-nonlinear phenomena.
 Second, we notice that PINODE achieves better performance within a given latent space budget.
@@
 This is a consequence of the latent-dynamics matrix ($h(z) = Lz$) of PIKN having eigenvalues with positive real parts, which implies long-term instability (Figure~(\ref{fig:burgers_compressibility}), right pane).
-Although there has been progress in the literature~\cite{kojimalearning}, further research is needed to understand (i) how to enforce stability constraints for PIKN, and (ii) why one does not need the same enforcement for PINODE to exhibit stable behaviour.
+Although there has been progress in the literature~\cite{kojimalearning}, further research is needed to understand (i) how to enforce stability constraints for PIKN, and (ii) why one does not need the same enforcement for PINODE to exhibit stable behavior.
+
+    \begin{figure}[t!]
+        \centering
+        \includegraphics[width=\textwidth]{figures/burgers_compression.png}
+        \caption{The right plots give examples of data snapshots used: trajectories with bell-curve ICs (top) and harmonic ICs (bottom).
+The middle-right pane shows harmonic collocations used by hybrid models in addition to the snapshots.
+The left plot compares the prediction errors (MSE) of two models, data-driven and hybrid with 128-dimensional latent spaces, on harmonic initial conditions that were not present in the trained data.
+The shaded regions represent 95\% confidence intervals based on 100 test trajectories.
+The middle plot shows the prediction error (MSE) on each type of test data for a variety of models with different latent-space sizes.
+We see that the use of harmonic collocations significantly improves the model performance on unseen harmonic ICs without increasing the errors on bell-curve ICs.}
+        \label{fig:pikn_burgers_compression}
+    \end{figure}
+
+    However, we observe that PIKN benefits from using physics-informed loss as well.
+In particular, we show that one can use collocations to improve model's performance on types of initial conditions that are missing in the available training data.
+To illustrate that we train two models.
+The data-driven model only uses 1024 trajectories with bell-curve initial conditions (ICs) (we provide an example at the top-right frame of the Figure~\ref{fig:pikn_burgers_compression}).
+The hybrid model additionally observes 80000 harmonic collocations formed by summing first 10 sinusoidal modes with random coefficients (Figure~\ref{fig:pikn_burgers_compression}, middle-right frame), for which we evaluate $u_t$ analytically using Equation~\ref{eq:burgers_equation}.
+Next, we evaluate the performance of both models using unseen trajectories with both harmonic and bell-curve ICs, 100 trajectories each.
+We observe that the Hybrid model predicts the sinusoidal trajectories ~10 times better than the data-driven one (Figure~\ref{fig:pikn_burgers_compression}, left frame, shown for the 128-dimensional latent-space model).
+Since neither models had any trajectories of that type in its training set we conclude that the difference in performance comes from using harmonic collocations.
+We also note that better performance of the hybrid model on harmonic ICs does not come at an expense of worse performance on bell-curve ICs, as shown in the central frame of Figure~\ref{fig:pikn_burgers_compression}.
+This evidence suggests that one can improve a model's extrapolation power  by supplementing its training with sufficiently diverse set of collocations, especially when additional simulations are expensive to obtain but the collocations are cheap to generate.
+The details of the network's architecture and training procedure are provided in the Appendix \ref{appendix:pikn_burgers}.
 

### 2023-04-13 msr3.tex
Add sensitivity+scalability analyses and covid example

@@
 
-\begin{figure}[t]
+\begin{figure}[ht]
     \centering
@@
 
- \begin{figure}[t]
+ \begin{figure}[ht]
     \centering
@@
 
-\subsection{Sensitivity Analysis}
+\subsection{Scalability and Sensitivity Analysis}
 \label{ch:msr3_sensitivity_analysis}
-\edit{Add sensitivity analysis from MSR3 practice paper}
+\subsubsection{Scalability}
+We tested the scalability of the new approach (MSR3-fast) compared to proximal gradient descent and proximal gradient descent with line search.
+To do this, chose an initially small problem and we scaled the number of features in the data from 100 to 1000, while scaling the number of observations proportionally, and tested the time to completion
+of these three methods, averaged over 100 replicates.
+Namely, each problem had 8 groups of $10A$ observations each, $\beta$ and $\gamma$ had $20A$ features equally split between 0 and 1.
+To get the problems of different sizes we assigned $A$ to be 1, 2, 5, 10, 20, 50, and 100, and for each choice of A we generated 100 random problems.
+Since MSR3-fast has a relaxation parameter $\eta$, we evaluated MSR3-fast across different $\eta$ values to also test the effect of $\eta$ on timing.
+For each experiment, we also computed
+the accuracy of the feature selection, to make sure that there was no degradation in performance.
+The results are presented in Tables~\ref{table:scalability_time} and~\ref{table:scalability_accuracy}.
+In terms of timing, we indeed see a superlinear increase in computational complexity with respect to the number of features.
+Nonetheless, MSR3-fast is competitive with the alternatives across the experiments, and the results are far more accurate.
+Larger problems could likely significantly benefit from iterated solvers within the interior point framework.
+
+\begin{table}[ht]
+	\centering
+    \resizebox{\columnwidth}{!}{\input{tables/scaling_time.tex}}
+    \caption{Execution time for feature selection problems of varying sizes.
+Each cell shows total time, including grid-search with respect to the sparsity parameter $\lambda$.
+Each cell shows averaged value over 100 randomly-generated problems.}
+    \label{table:scalability_time}
+
+\end{table}
+
+\begin{table}[ht]
+	\centering
+    %\resizebox{\columnwidth}{!}{}
+    \input{tables/scaling_acc.tex}
+    \caption{Accuracy of feature selection problems of varying sizes. Each cell shows averaged value over 100 randomly-generated problems.}
+    \label{table:scalability_accuracy}
+\end{table}
+
+\subsubsection{Closeness to the Central Path for IP}
+The $\tau$ parameter of MSR3-fast controls how close the interior point method gets to the central path before taking a prox-gradient step.
+This is a heuristic parameter in the algorithm, and to understand its impact we tested
+the sensitivity of the execution time and accuracy for a problem with 200 features for four selections of relaxation parameter $\eta$.
+The problems were identical to those from the second row of Table~\ref{table:scalability_time}.
+The results are reported in Tables~\ref{table:central_path_time} and~\ref{table:central_path_accuracy}.
+Neither time nor accuracy were affected by $\tau$ across the levels of $\eta$.
+
+\begin{table}[ht]
+	\centering
+    \input{tables/central_path_time.tex}
+    \caption{Execution time of MSR3-fast for different values of $\tau$ - a parameter that controls how close the IP needs to be to the central path before doing a projection step.
+Each cell shows total time, including grid-search with respect to the sparsity parameter $\lambda$.
+Each cell shows averaged value over 100 randomly-generated problems.}
+    \label{table:central_path_time}
+\end{table}
+
+\begin{table}[ht]
+	\centering
+    %\resizebox{\columnwidth}{!}{}
+    \input{tables/central_path_acc.tex}
+    \caption{Accuracy of MSR3-fast for different values of $\tau$ - a parameter that controls how close the IP needs to be to the central path before doing a projection step.
+Each cell shows averaged value over 100 randomly-generated problems.}
+    \label{table:central_path_accuracy}
+\end{table}
 
@@
 \subsection{Application to Real Data: COVID-19 Transmission Factor}
-\edit{Add Covid example from my general}
+\label{ch:covid}
+\begin{figure}
+	\begin{subfigure}[b]{\textwidth}
+		\includegraphics[width=\textwidth]{figures/fit_Alaska}
+	\end{subfigure}
+	\begin{subfigure}[b]{\textwidth}
+		\includegraphics[width=\textwidth]{figures/fit_Slovenia}	
+	\end{subfigure}
+	\caption{\label{fig:covid_feature_selection_1} Comparison of fits of two different models (fully dense linear mixed model(MM) and MSR3-$\ell_0$ (R\&S) to the original IHME Projections for Alaska and Slovenia.
+The quality (RMSE) achieved by a sparse fit is within 10\% from a quality of both dense models which is evidences that the model picked up informative features.}
+\end{figure}
+
+\begin{figure}
+	\begin{subfigure}[b]{\textwidth}
+		\includegraphics[width=\textwidth]{figures/fit_Turkey}
+	\end{subfigure}
+	\begin{subfigure}[b]{\textwidth}
+		\includegraphics[width=\textwidth]{figures/fit_Switzerland}	
+	\end{subfigure}
+	\caption{\label{fig:covid_feature_selection_2} Comparison of fits of two different models (fully dense linear mixed model (MM) and MSR3-$\ell_0$) (R\&S) to the original IHME Projections for Turkey and Switzerland.
+The quality (RMSE) achieved by a sparse fit is within 10\% from a quality of both dense models which is evidences that the model picked up informative features.}
+\end{figure}
+
+In this section we apply our method to a COVID-19 Contact Rate Forecasting problem.
+The global pandemic created an unprecedented need for robust and accurate disease transmission forecasting.
+Since the beginning of the pandemic Institute for Health Metrics and Evaluation has been providing guidance to local authorities across the world with their pioneering \href{https://covid19.healthdata.org/global?view=total-deaths&tab=trend}{COVID-19 Projections tool} \cite{IHME2020Projections}.
+The key methodology which was essential for success in their forecasting of the disease's dynamics was transforming the death data into the contact rate time series data, and then relating this contact rate to available covariates such as temperature, social mobility, population, and others \cite{IHME2020covidnature}.
+All these covariates were collected in real time using limited human resources, and identifying the most important covariates was crucial for distributing these resources effectively.
+In the example below we show how \ouralgo can be used for making covariate selection on IHME data.
+
+The dataset consists of $m=60$ groups (countries or states), the detailed description of groups sizes ($n_i$) and time spans is provided in the Appendix~\ref{ch:covid}.
+The target variable $Y_i$ was the contact rate -- the coefficient $\beta(t)$ from an SEIR model (not to be confused with $\beta$ -- vector of fixed effects).
+The covariates -- columns of $X_i$ and $Z_i$ -- were: \texttt{intercept} --a column of ones,  \texttt{temperature} -- average air temperature in degrees Fahrenheit, \texttt{mobility\_lift} -- social mobility, \texttt{proportion\_over\_1k} -- population size threshold, and \texttt{testing\_reference} -- testing.
+The observation error's variance $\sigma_i$ was set to be 0.1.
+
+Feature selection results for four particular locations (Alaska, Slovenia, Turkey, and Switherland) are presented on Figures \ref{fig:covid_feature_selection_1} and \ref{fig:covid_feature_selection_2}, with coefficients for the rest of the locations attached in supplementary materials.
+\ouralgo was charged with a task to produce a fit using only three covariates, one of which had to be mean-only (no random effects).
+On the left we see the original data (\texttt{Data}, in {\color{blue} blue}), and predictions of three models: IHME Projections (\texttt{IHME}, in {\color{orange} orange}), a linear mixed model fit with no selection (\texttt{Dense MM}, in {\color{red} red}), and a sparse fit of \ouralgo (\ouralgo in {\color{green} green}).
+The \ouralgo model has chosen to use \texttt{intercept}, \texttt{mobility\_lift}, and \texttt{temperature} covariates, with the later only as a fixed covariate; \texttt{testing\_reference} and \texttt{proportion\_over\_1k} were left out.
+On the plots to the left we see that the exclusion of two covariates did not significantly affect the quality of predictions.
+The residual errors (RMSE) to the right support this statement: the difference in RSME is within 10\% of RSME of a "dense" model which uses all the covariates, as well as from IHME's original model which fit all the locations via sequence of independent regressions.
+This choice also seems reasonable given the timespan: the proliferation of testing during the spring was not yet significant, so it did not inform predictions in a major way.
+The exclusion of \texttt{proportion\_over\_1k} could have been due to the scale of grouping: locations were grouped on the level of states and countries, not on the level of individual counties where the influence of population-based covariates could have been more significant.
 

### 2023-04-19 introduction.tex
Add draft of the intro

\textbf{\textit{The current intro is a draft that captures content and, for now, disregards the form and grammar. It will be rewritten later in a more professional cadence.}}
\edit{Rewrite the introduction: improve style and grammar, add citations}

Machine Learning became an important tool in applied mathematics problems as data-driven approaches have been increasingly outperforming the first-principle ones over the last decade. The most successful approaches, however, combine first-principle and data-driven components. These first-principle components often come from our domain expertise and prior beliefs. Incorporating prior knowledge into a data-driven method is often challenging and requires a good share of ingenuity. This thesis showcases two quite different examples of such work. Together these examples cover a variety of challenges of incorporating priors knowledge into a model and describes approaches to overcome them.

One example is the feature selection problems for mixed-effect models. Feature selection problems typically arise when one believes that a small subset of features in dataset captures the majority of the variance of the quantity in question; this subset of features is often referred to as parse support. The canonical methods typically provide control over the degree of sparsity for the solution and outputs a candidate support for it. For a practitioner who has no additional information about the support besides its likely sparsity these methods provide sufficient toolkit.

The real-world practitioner, however, almost universally knows more than that. As a consequence, they want to incorporate their expert knowledge that mostly comes as a set of constraints. For example, a practitioner wants a model that uses at most 6 features out of 40, with features A and B being included at all times. If a feature C is included then it has to be negative, and a feature D should not depend on a grouping parameter. An attempt to tackle these constraints directly frequently leads to mathematically intractable problems, e.g. the first constraint makes the optimization problem N-P complete. Moreover, a practitioner wants to swap those priors in and out as more hypotheses and knowledge arise. It implies that the implementation of the optimization routine should not rely on the priors, otherwise it forces the practitioner to re-implement the algorithm after every such swap, rendering the instrument impractical. Chapter~\ref{ch:msr3} describes my work on developing such a universal feature selection approach for linear mixed-effects models while simultaneously tackling numerical challenges that arise from it.

The second example illustrates how one can embed a partial knowledge of physics into a deep neural network. The field of data-driven modeling of physical phenomena soared in popularity over the last decade with the advancements of GPU computations and automatic differentiation tools. It has proven to be especially useful in the field of reduced-order modeling where a neural network can find a representation of a system on a reduced-order manifold via finding a non-linear transformation to that space based on data. The classic methods such as Galerkin POD or DMD are fundamentally unable to find such spaces because these methods are confined to linear, albeit optimal, transformations. However, as the no-free-lunch theorem goes, with great approximation power comes great instability. In particular, the networks tend to find manifolds that overfit the data and fail at extrapolation tasks. Such failures jeopardize the performance of the whole system where the network is used, e.g. in control, model identification, or compressive sensing applications.

A natural idea for improving neural-network based models would be incorporating a partial knowledge of physics of the phenomenon that the model predicts. For example,  we can find a better reduced-order manifold if we know that the model predicts the behavior of a shock wave that obeys Burger's equation. Chapter~\ref{ch:pinode} is devoted to the development of PINODE ROMs -- a framework for building Reduced-Order Models with Physics-Informed Neural Ordinary Differential Equations. It illustrates how one can transfer knowledge from a known equation to a ROM that predicts solutions of said equation using collocation points technique. The resulting ROMs extrapolate significantly better forward in time as well as for unseen initial conditions, and are much less sensitive to noise. Finally, in Chapter~\ref{ch:cs} I show how PINODE models can be put to work in compressive sensing applications where it "stitches" separate timeframes together and allows to significantly surpass the Nyqvist sampling limit.

\paragraph{Attribution} Throughout my PhD I was fortunate to work with incredibly talented advisors and collaborators who significantly contributed to the content of this work. Chapter~\ref{ch:msr3} captures a sequence of joint works that were completed between 2019 and 2023 at the University of Washington's Institute for Health Metrics and Evaluations (IHME) together with Dr. Aleksandr Aravkin, Dr. James Burke, and Dr. Peng Zheng. The MSR3 framework algorithm was introduced in \cite{sholokhov2022relaxation}, with the theoretical foundations developed in \cite{aravkin2022jimtheory}, and a software was later published in \cite{sholokhov2023msr3joss}. Chapter~\ref{ch:pinode} represents my work at Mitsubishi Electric Research Labs (MERL) that I accomplished under the guidance of Dr. Hassan Mansour and Dr. Saleh Nabi. The PINODE framework was developed in \cite{sholokhov2023pinode} and is based on our earlier joint work with Dr. Yuying Liu on Physics-Informed Koopman Networks~(\cite{liu2022physics}). In particular, Figure~\ref{fig:pikn_burgers_compression} and the related example come from that work. Chapter~\ref{ch:cs} was developed in~\cite{sholokhov2023cs} together with Dr. J. Nathan Kutz and Dr. Steven Brunton at the University of Washington. Many parts of the works listed above were taken verbatim to this document per an explicit permission of my co-authors.

### 2023-04-19 phd_thesis.tex
Add sensitivity+scalability analyses and covid example
Add draft of the intro

@@
 \newcommand{\userName}{Aleksei Sholokhov}
-\newcommand{\userId}{Lesha}
+\newcommand{\userId}{aksh}
 \newcommand{\department}{Department of Applied Mathematics}
 \newcommand{\institution}{University of Washington}
-\newcommand{\projectNameShort}{Feature Selection for Mixed Models}
+\newcommand{\projectNameShort}{}
 \newcommand{\doctitle}{Doctoral Thesis}
-\newcommand{\projectNameLong}{}
+\newcommand{\projectNameLong}{Optimization Methods for Parameter Identifications \\ in Settings with Only Partial Knowledge}
 
@@
 \newcommand{\Jim}[1][]{\textbf{Jim\ifthenelse{\equal{#1}{}}{ }{ (#1):}}}
-\newcommand{\ouralgo}{R\&S-Mixed }
+\newcommand{\ouralgo}{MSR3 }
 
@@
 {\large 
-{\huge
-Covariate Selection Techniques for Linear Mixed-Effects Models with Applications to Population Health Problems }
+{\large
+\projectNameLong
+}
 \\[10mm]
 \userName
-\\[40mm]
+\\[20mm]
+{A dissertation \\
+ submitted in partial fulfillment of the\\ requirements for the degree of}
+\\[5mm]
+{Doctor of Philosophy}
+\\[20mm]
 {\center
-\textbf{Committee}\\
+\textbf{Reading Committee}\\
 Aleksandr Aravkin (co-chair),\\
-Nathan Kutz (co-chair), \\
+J. Nathan Kutz (co-chair), \\
 James Burke, \\
@@
 }
-\\[40mm]
-{A thesis proposal submitted in fulfillment\\ of the requirements for the degree of}
+\\[20mm]
+Program Authorized to Offer Degree:
 \\[5mm]
-{Doctor of Philosophy}
-\\[15mm]
 \department
@@
 \\[10mm]
-\textit{Last modified: \today}
+©Copyright 2023\\
+\userName
 
@@
 \newpage
-{\center \textbf{Abstract}\\[5mm]}
+{
+\center 
+\institution
+\\[10mm]
+\textbf{Abstract}
+\\[10mm]
+\projectNameLong
+\\[10mm]
+\userName
+\\[10mm]
+Chairs of the Supervisory Committee:\\
+Aleksandr Aravkin and J. Nathan Kutz\\
+\department
+\\[10mm]
+}
 
-Linear Mixed-Effects models are widely used technique for modeling clustered data, such as cohort studies, longitudinal data analysis, and meta-research.
-In this work we address an important practical problem of simultaneous selection of fixed and random effects in this type of models.
-Our approach is based on the Relax-and-Split methodology, which uses a relaxation technique together with partial minimization to efficiently solve non-convex problems.
-We use it to relax the constraints on the maximal number of included covariates and then apply an interior point method to obtain a solution.
-The performance of the resulting estimator is studied on synthetic data, as well as on two real-data applications: covariates selection for modeling the contact rate for COVID-19, and the estimation of burden of anxiety and depression disorders as results of bullying.
-These experiments show that the proposed method is advantageous in terms of execution time and selection quality.
-In addition, several theoretical questions are discussed as a part of future work that would contribute to understanding of convergence conditions of Relax-and-Split based methods in general.
+My abstract goes here. I will probably write it at the very end.
 
-\newpage
 \tableofcontents
@@
 
-\chapter{MSR3}
+\chapter{Introduction}
+\include{introduction}
+
+\chapter{MSR3: Sparse Relaxed Regularized Regression for Mixed-Effect Models}
+\label{ch:msr3}
 \include{msr3}
 
-\chapter{PINODE}
+\chapter{PINODE: Physics-Informed Neural Ordinary Differential Equations}
+\label{ch:pinode}
 \include{pinode}
@@
 \input{cs}
+
+\chapter{Conclusion}
 
@@
 
-\section{Description of Datasets: GBD Bullying Data}
+\section{Description of Real-World Datasets}
+\subsection{GBD Bullying Data}
 \input{appendix/bullying_features_description}
 
-%\subsection{COVID-19 Contact Rate Forecasting Data}
-%\input{tables/covid_groups_description}
-%\input{tables/covid_coefficients}
+\subsection{COVID-19 Contact Rate Forecasting Data}
+\input{tables/covid_groups_description}
+\input{tables/covid_coefficients}
 

### 2023-04-24 msr3.tex
Incorporate all Sasha and Jim's edits from the rebuttal

@@
 variable selection strategies and to determine which method is best suited for a given task. 
-This challenge  is exacerbated by the absence of standardized datasets %for different experimental designs
+This challenge  is exacerbated by the absence of standardized datasets 
+%for different experimental designs
 and open source libraries for each method. 
@@
 accommodates a wide variety of variable selection strategies based on a set
-of easily implementable regularizers, and implemented in an open source library that makes 
-it easy to use and to compare different methods.
-
-\begin{figure}[ht]
+of easily implementable regularizers, and 
+made available in an open source library, \texttt{pysr3}\footnote{https://github.com/aksholokhov/pysr3} that is 
+easy to use and to compare different methods. All experiments in the paper can be reproduced 
+using \texttt{pysr3} and code in the reproducibility guide\footnote{https://github.com/aksholokhov/msr3-paper}.
+
+\begin{figure}[h!]
     \centering
@@
 
-In this work we develop a regularization-agnostic covariate selection strategy that (1) is fast and simple to implement, (2) provides robust models, and (3) is flexible enough to support most regularizers currently used in variable selection across different domains.
+In this work we introduce a regularization-agnostic covariate selection strategy that (1) is fast and simple to implement, (2) provides robust models, and (3) is flexible enough to support most regularizers currently used in variable selection across different domains.
 The baseline approach uses the proximal gradient descent (PGD) method, which has been studied by the optimization community for over 40 years, but has not been widely used in LME covariate selection.
-We provide proximal operators for commonly used regularizers and show how to apply the PGD method to the nonconvex LME setting.
-In particular we apply the PGD method to four regularizers,
-including the $\ell_0$ regularizer, which does not admit local quadratic approximations and has not been used before for LME variable selection strategies.
-
-We also develop a new meta-approach that can improve the performance of LME selection methods for any regularizer.
-Specifically, we extend sparse relaxed regularized regression (SR3) framework~{(\cite{Zheng2019SR3})} to the LME setting.
+%We provide proximal operators for commonly used regularizers and show how to apply the PGD method to the nonconvex LME setting. In particular we apply the PGD method to four regularizers, 
+%including the $\ell_0$ regularizer, which does not admit local quadratic approximations and has not been used before for LME variable selection strategies. 
+In our initial numerical experiments, using a naive PGD approach
+indicated that, at best, the method yields only a marginal improvement over the
+equally unsatisfactory alternative methods   
+in accurately determining the correct variables in our variable selection test problems.
+We conjecture that the weakness of the naive PGD is a result of the 
+first-order likelihood approximation used in PGD.
+To overcome this problem, we propose an alternative likelihood approximation 
+with the goal of 
+incorporating global variational properties of the likelihood. For this purpose,  
+we extend the sparse relaxed regularized regression (SR3) framework~{(\cite{Zheng2019SR3})} to the LME setting. 
+This idea is supported
+by the success of SR3 in the context of linear regression where it
+accelerates and improves the performance of regularization strategies.
+This extension and its mathematical foundations in \cite{Theory1} constitute
+the major innovations of this work.
+Here we introduce the modeling framework and its relaxation, discuss the 
+resulting algorithms and their implementation details, and  validate the method on both
+simulated and real data sets, while the 
+%highly technical 
+mathematical 
+foundations are presented in \cite{Theory1}.
+The SR3 framework introduces auxiliary variables $x$ and $w$ to 
+decouple the likelihood $\LL$ and sparsity regularizer $R$ which are tied
+together by adding a multiple of the norm squared difference 
+$\frac{\eta}{2}\norm{x-w}^2$ to the
+objective. 
+%We then smooth the positive semi-definite constraint in the 
+%likelihood with a log barrier term yielding the smoothed likelihood $\mathcal{L}_\mu$
+%where $\mu$ is the smoothing parameter.
+Then, fixing the variables $w$ dedicated to the nonsmooth regularizer $R$, 
+the smooth function $\LL(x)+\frac{\eta}{2}\norm{x-w}^2$ is
+globally optimized over the variables $x$ to obtain an optimal value function
+$u_\eta(w)$. We then show that $u_\eta(w)$ is smooth. This opens
+the door to the application of the PGD algorithm to minimizing $u_\eta+R$ 
+where now $u_\eta$ contains global variational information
+on the likelihood function $\LL$. The main obstacle in the
+application of this approach is the evaluation of $u_\eta$ and its gradient.
+%%%
+In Section \ref{sec:MSR3} we present a method for overcoming this difficulty 
+using variable metric techniques and interior point technology.
+%%%%%%
+%To overcome this difficulty we turn to interior point technology. This choice requires 
+%smoothing the positive semi-definite constraint on the covariance matrices by
+%adding log-barrier term to the likelihood to obtain the smooth function
+%$\LL_{\mu,\eta}$ where $\mu$ is the log-barrier relaxation parameter, and
+%an associated optimal value function $u_\emu$ whose value and gradients 
+%can be quickly and accurately computed.
+%%by applying variable metric techniques along with
+%%well established properties of the interior point (IP) algorithm to obtain 
+%%sufficiently accurate values for
+%%both $u_\eta$ and its gradient. 
+%Using this underlying framework, we develop two related algorithms,
+%MSR3 and MR3-fast where MSR3-fast employs a kind of continuation approach 
+%that interlaces approximate IP optimality conditions for the evaluation of 
+%$u_\emu$ and its gradient with the prox computation in the PGD algorithm.
+%%using a {\it central path} proximity condition from the IP literature.  
+%The resulting algorithms are then applied to four regularizers, 
+%including the $\ell_0$ regularizer 
+%which does not admit local quadratic approximations and 
+%has not been used before for LME variable selection strategies.
+%The numerical results show that both MSR3 and MR3-fast
+%yield superior results over the naive PGD method in terms of specificity and sensitivity of feature selection
+%with a slight edge to MR3-fast. However, MR3-fast delivers a spectacular improvement
+%in speed over both the naive PGD and the MSR3 algorithms
+%(see Table \ref{table:comparison_of_algorithms}).
+
+%We also develop a new meta-approach that can improve the performance of LME selection methods for any regularizer.
+Specifically, we extend the sparse relaxed regularized regression (SR3) framework~{(\cite{Zheng2019SR3})} to the LME setting.
 In linear regression, SR3 accelerates and improves the performance of  regularization strategies by introducing auxiliary variables that decouple the accuracy and sparsity requirements for the model coefficients.
-We develop a conceptual and algorithmic approach necessary to extend the SR3 concept to LME.
+%We develop a conceptual and algorithmic approach necessary to extend the SR3 concept to LME.
 This development is necessary because the LME problem is nonlinear, nonconvex, and includes constraints on variance parameters.
 We show that the new approach yields superior results in terms of specificity and sensitivity of feature selection, and is also computationally efficient.
-
-\edit{Add a paragraph summarizing results from the theory paper}
 
@@
 \texttt{pysr3} enables a standardized comparison of different methods in the LME setting, and makes both the PGD framework and its SR3 extension available to practitioners working with LME models.
-
-\edit{Extend on the package summary using JOSS paper}
-
 %\subsection{Notation}
@@
 
-\section{Linear Mixed-Effects Models: Notation and Fundamentals}
+%Roadmap
+We begin in Section \ref{sec:LMEM} by giving a precise description of the LME
+model and set the notation for the remainder of the paper. 
+This is followed by a brief discussion of prior work on LMEs.
+In Section \ref{sec:pgd_methods} we present our algorithms for the LME model 
+starting with the naive PGD algorithm. This is followed by a description of the 
+variable splitting technique used in \cite{Zheng2019SR3} to incorporate global 
+variation information on the likelihood function into the direction finding subproblem
+for the PGD algorithm. 
+%This lays the foundation for the MSR3 algorithm. 
+Next we tackle the problem of how to approximate the 
+resulting optimal value function $u_\eta$ and its gradient where $\eta>0$ is the 
+decoupling parameter. As noted, this is done using variable metric techniques and 
+and interior point technology. We conclude Section \ref{sec:pgd_methods} 
+with a discussion of the
+MSR3 and MSR3-fast algorithms. In Section \ref{sec:applications} we discuss how
+the underlying algorithmic parameters are set and test the algorithm on both 
+simulated problems and a problem with real data. The paper is concluded in
+Section \ref{sec:discussion} with a brief discussion of the contributions.
+
+\edit{Add a paragraph summarizing the theory paper}
+\edit{Add a paragraph summarizing software paper}
+
+\section{Linear Mixed-Effects Models: Notation and Fundamentals}\label{sec:LMEM}
 Mixed-effect models describe the relationship between an outcome variable and its predictors when the observations are grouped, for example in studies or clusters.
@@
 Following~\cite{Patterson1971, Pinheiro2000}, we define a Linear Mixed-Effects (LME) model as
-\eq{
-	\label{eq:lme_setup}
+\begin{equation}
+\begin{aligned}
 	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i, \quad i = 1 \dots m \\
@@
 	\varepsilon_i & \sim \NN(0, \Lambda_i), \quad \Lambda_i \in \bS_{++}^{n_i}
-}
+	\end{aligned}
+	\label{eq:lme_setup}
+\end{equation}
  where $\beta \in \R^p$ is a vector of fixed (mean) covariates, 
@@
 random intercepts ($Z_i$ are columns of 1's that add $u_i$ to all datapoints from the $i$th study)
-and random slopes ($Z_i$ also scale $u_i$ according to the magnitude of a covarite), see e.g.~\cite{pinheiro2006mixed}. 
+and random slopes ($Z_i$ also scale $u_i$ according to the magnitude of a covariate), see e.g.~\cite{pinheiro2006mixed}. 
 %In these typical examples, the columns of $Z_i$ are subsets of the columns
@@
 is an unknown diagonal matrix, i.e., $\Gamma = \Diag{\gamma}, \ \gamma \in \R^s_+$.
+This assumption corresponds to the meta-analysis and meta-regression branch of mixed effects problems, 
+which is the primary focus of our applied collaborations~ (see e.g. \cite{zheng2022burden,lescinsky2022health,razo2022effects,stanaway2022health,dai2022health}.)
+The theoretical developments in this work allow extensions to other types of repeated measure models, 
+but practical implementation requires significant additional effort, and we leave these extensions to future work.
+
 % either
@@
 Defining group-specific error terms $\omega_i = Z_i u_i + \varepsilon_i$, we get a compact formulation that 
-recasts~\eqref{eq:lme_setup} as a correlated noise model:
+recasts~(\ref{eq:lme_setup}) as a correlated noise model:
  \eq{
@@
 computed in~\cite[Section 5.1]{Theory1} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation
-%\eq{
-%	
-\[
+
+\begin{equation}
 	\label{eq:beta_formula}
 	\beta(\Gamma) = \argmin_{\beta}\LL(\beta, \Gamma) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i.
-\] %}
+\end{equation}
+
 By using the simplification $\Gam=\Diag{\gam}$, we obtain the problem
@@
 	\min_{\beta\in\R^p, \gam\in\R^q_+} & \ \LL(\beta,\gam)
-	:=\LL_{ML}(\beta, \Diag{\gamma}) 
+	:=\LL_{ML}(\beta, \Diag{\gamma}) \, .
 %	\\
@@
  and all $\Lambda_i$ are known (see \eqref{eq:lme_diagonal_setup}), 
-following the meta-analysis use-case~\citep{zheng2021trimmed} that is widely used in epidemiological studies~\cite{murray2020global}. 
+following the meta-analysis use-case~\citep{zheng2021trimmed} that is widely 
+employed in epidemiological studies~\cite{murray2020global}. 
 While the proposed approach can be extended to the non-diagonal case, we leave it for future work, save for a brief discussion in Section~\ref{sec:applications}.
@@
 Variable (feature) selection models seeks to select or rank the most important predictors in a dataset in order to get a parsimonious model at a minimal cost to prediction quality. 
+Feature selection may be performed both on $\beta$, to find the sparse set of covariates that best explains the mean, and on $\gamma$, to find the sparse set of covariates that best accounts for 
+variation between groups. Both types of selection have been studied in the literature, and both are accessible using the methods developed here.
 %The selection process often occurs simultaneously with fitting the model, and a predictor is `selected' if its respective coefficient in the model is non-zero. 
@@
 
- \begin{figure}[ht]
+ \begin{figure}[h!]
     \centering
@@
 Based on this idea, \cite{Krishna2008} extended the Adaptive LASSO regularizer (\cite{Lan2006, Xu2015}) to mixed-effects setting using the objective
-   \[
+   \(
         \LL(\beta, \Gamma(D,L)) + \lambda\left(\sum_{i = 1}^p 
         \left|\frac{\beta_i}{\hat\beta_i}\right| + \sum_{j = 1}^q \frac{D_{ii}}{\hat{D_{ii}}} \right),
-   \]
+   \)
  where $\hat\beta$ and $\hat{D}$ are the solution of a non-penalized maximum likelihood problem and $\lambda$ is a tuning parameter for the weighted regularizer
@@
 
-\edit{Add more methods for feature selection \cite{Buscemi2019Survey, Muller2013}}
-
-To better compare methods, we need to consider the regularization parameter 
-\plum{$\lambda$} and how it is tuned. 
+To better compare methods, we need to consider the tuning of the regularization parameter 
+$\lambda$. 
 The output of a shrinkage model critically depends on the tuning parameter 
-$\lambda$. The entire range of possible $\lambda$ values is captured 
+$\lambda$. The entire range of $\lambda$ values is captured 
 by the notion of a ``$\lambda$-path in the model space'', with  the best parameter and the final model chosen using information criteria.
-According to \cite{Muller2013}, the most widely used information criterion is the marginal AIC criterion (\cite{Vaida2005}):
-  \eq{
-  \label{eq:vaida_aic}
-  AIC := 2\LL(\hat{\theta}) + 2\alpha_n(p+q)
-  }
+According to \cite{Muller2013}, the most widely used information criterion is the marginal AIC criterion (\cite{Vaida2005}),
+ \(  AIC := 2\LL(\hat{\theta}) + 2\alpha_n(p+q)\), 
   where $\hat\theta$ includes all the estimated parameters $(\beta, \Gamma)$, and $\alpha_n := n(n-p-q-1)$ for the finite sample case (\cite{Sugiura1978}). 
@@
 AIC, however, is also known to be positively-biased (\cite{Ibrahim2011}).
-  Alternatively, LASSO-type methods (\cite{Krishna2008, Ibrahim2011}) use a BIC-type information criterion:
-  \eq{
-    BIC := 2\LL(\hat{\theta}) + \log(n)(p+q).
-  }
+  Alternatively, LASSO-type methods (\cite{Krishna2008, Ibrahim2011}) use a BIC-type information criterion,  \(BIC := 2\LL(\hat{\theta}) + \log(n)(p+q)\).
 BIC performs well in practice, but does not have theoretical guarantees~(\cite{schelldorfer2011estimation}).
@@
 
-\edit{Add more ICs (e.g. Muller)}
+  \edit{Add more ICs, e.g. Muller}
 
@@
 %	}
+%\vskip -.5in
 \eq{%\label{eq:main opt in x over C}
+     \min_{x } & \LL(x) + R(x)+ \delta_{\CC}(x),
       \label{eq:lme_loss_original_in_x}
-    \min_{x } & \LL(x) + R(x)+ \delta_{\CC}(x),
 %    \\
@@
 lower semi-continuous (lsc) regularization term, and
-$\delta_{\CC}$ is the convex indicator function
-\[
-    \delta_{\CC}(x) := \begin{cases} 0, &  x \in \CC \\ +\infty, & x \not\in \CC .\end{cases}
-\]  
+$\delta_{\CC}$ is the convex indicator function, where
+$\delta_{\CC}(x) :=0$ for $x \in \CC$ and $+\infty$ otherwise.
+%\[
+%    \delta_{\CC}(x) := \begin{cases} 0, &  x \in \CC \\ +\infty, & x \not\in \CC .\end{cases}
+%\]  
 By \cite[Theorem 2]{Theory1}, solutions to \eqref{eq:lme_loss_original_in_x}
@@
 $\alpha R + \delta_{\CC}$ is defined as the mapping
-\[ %\eq{
+\( %\eq{
     \prox_{\alpha R + \delta_{\CC}}(z) := \argmin_{y\in \CC}\ R(y) + \frac{1}{2\alpha}\|y - z\|_2^2 ,
-\] %}
+\) %}
 and the PGD iteration is given by
-\[
+\(
     x^+ = \prox_{\alpha R + \delta_{\CC}}(x - \alpha \nabla \LL(x)),
-\]
+\)
 where $\alpha$ is a stepsize.
 When $R(x)$ has the form given in \eqref{eq:R sep}, we have
-\[
+\(
  \prox_{R}(z) = (\prox_r(z_1), \dots, \prox_r(z_q)) .
-\]
+\)
+Table \ref{table:proxes} provides closed form expressions for the proximal operators 
+of commonly used regularizers. 
+For all of these cases,  the following theorem gives closed form expressions for
+$\prox_{\alpha R + \delta_{\CC}}(z)$.
+
 \begin{table}[H]
@@
          \hline \hline
-         LASSO ($\ell_1$) \newline (\cite{tibshirani1996regression}) & $|x|$ & $\sign(z)(|z| - \alpha)_+$ \\
+         LASSO ($\ell_1$) \newline & $|x|$ & $\sign(z)(|z| - \alpha)_+$ \\
          \hline
-         A-LASSO \newline (\cite{Fan2001}) & $\bar{w}|x|$, $\bar{w} \geq 0$ &  $\sign(z)(|z| - \alpha\bar{w})_+$\\
+         A-LASSO \newline  & $\bar{w}|x|$, $\bar{w} \geq 0$ &  $\sign(z)(|z| - \alpha\bar{w})_+$\\
          \hline
-         SCAD \newline (\cite{fan1997comments}) & $\begin{cases} \sigma |x|, & |x| \leq \sigma \\ \frac{-x^2 + 2\rho\sigma x - \sigma^2}{2(\rho - 1)}, & \sigma < |x| < \rho\sigma \\ \frac{\sigma^2(\rho + 1)}{2}, & |x| > \rho\sigma \end{cases}$ & $\begin{cases} \sign(z)(|z| - \sigma\alpha)_+, & |z| \leq \sigma(1+\alpha) \\ \frac{(\rho - 1)z - \sign(z)\rho \sigma\alpha}{\rho - 1 - \alpha}, & \sigma(1+\alpha) < |z| \\
-         &\quad\leq \max(\rho, 1+\alpha)\sigma \\ z, & |z| > \max(\rho,1+\alpha)\sigma \end{cases}$ \\
+         SCAD \newline & $\begin{cases} \sigma |x|, & |x| \leq \sigma \\ \frac{-x^2 + 2\rho\sigma x - \sigma^2}{2(\rho - 1)}, & \sigma < |x| < \rho\sigma \\ \frac{\sigma^2(\rho + 1)}{2}, & |x| > \rho\sigma \end{cases}$ & $\begin{cases} \sign(z)(|z| - \sigma\alpha)_+, & |z| \leq \sigma(1+\alpha) \\ \frac{(\rho - 1)z - \sign(z)\rho \sigma\alpha}{\rho - 1 - \alpha}, & \sigma(1+\alpha) < |z| \leq \rho\sigma
+         %\max(\rho, 1+\alpha)\sigma 
+         \\ z, & |z| > \max(\rho,1+\alpha)\sigma \end{cases}$ \\
 %         \hline
@@
 \end{table}
-
-\edit{Make section "Evaluating proximal operators" using Appendix}
 
@@
 of commonly used regularizers.
+
 %All regularizers in this list are separable except $R(x) = \delta_{\|x\|_0 \leq k}$, for which the solution also has an analytically closed form.
@@
 %the evaluation of
-For all of these cases,  the following theorem gives closed form expressions for
-$\prox_{\alpha R + \delta_{\CC}}(z)$.
 
@@
 that include an additional constraint on $\gamma$ of the form 
-\[
+\(
 0 \leq \gamma \leq \bar\gamma,
-\]
+\)
 for $\bar\gamma\in[0,+\infty]$.
@@
     \begin{enumerate}
-    \item  For SCAD, we have for all $i$ that  
+    \item  For SCAD, we have for all $i$ that \;
    % \eq{
@@
 %    \eq{
-\[
+\(
     \prox_{(\alpha r + \del_{[0,\bgam]})} %\delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}
@@
         \bar\gamma, & \gamma_i \geq \bar\gamma \\
-        0, & \text{ otherwise} \end{cases}
-\]
+        0, & \text{ otherwise} \end{cases}.
+\)
 %    }
-    \item  For LASSO, A-LASSO we have for all $i$ that 
+    \item  For LASSO, A-LASSO we have for all $i$ that \;
  %   \eq{
@@
 %    \eq{
-\[
+\(
 \prox_{(\alpha r + \del_{[0,\bgam]})}(\gamma_i) 
@@
         \bar\gamma, & \gamma_i \geq \bar\gamma + \alpha \\
-        0, & \text{ otherwise} \end{cases}
-\]
+        0, & \text{ otherwise} \end{cases}.
+\)
 %    }
@@
 The main advantages of Algorithm \ref{alg:pgd_for_lme} are its simplicity and flexibility.
-The main loop needs only the gradient and prox operator, and the structure of the algorithm is independent of the choice of regularizer $R$.
+The main loop needs only the gradient and prox operator, and the structure of the algorithm is independent of the choice of $R$.
 Algorithm \ref{alg:pgd_for_lme} locates first-order stationary points
-under weak assumptions, in particular neither the objective nor the regularizer need to be convex~\citep{AB17,attouch2013convergence}.
+under weak assumptions, in particular neither the objective nor the regularizer need be convex~\citep{AB17,attouch2013convergence}.
 
@@
 
-\edit{Add other optimization methods for FS in LMEs (e.g. Newton)}
-
-\search{Add table of cross-references between optimization methods and regularizers}
-
-\search{Add table with rates of convergence wherever possible}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figures/intuition_current.pdf}
-    \caption{Proximal Gradient Descent (PGD) for~\eqref{eq:value_function_optimization} converges far faster than for~\eqref{eq:lme_loss_original_in_x}, because the SR3-value function $u_\eta$ yields more spherical level-sets than original likelihood for both convex components $\beta$ (first row) and non-convex components $\gamma$ (second row).
+%\begin{figure}[h!]
+  %  \centering
+ %   \includegraphics[width=1\textwidth]{Images/intuition_current.pdf}
+ %   \caption{Proximal Gradient Descent (PGD) for~\eqref{eq:value_function_optimization} converges far faster than for~\eqref{eq:lme_loss_original_in_x}, because the SR3-value function $u_\eta$ yields more spherical level-sets than original likelihood for both convex components $\beta$ (first row) and non-convex components $\gamma$ (second row).
     %The effect is particularly strong for $\gamma$-minimizers near the boundary of the constraint region (bottom-left panel).
-    }
-    \label{fig:geometric_intuition_sr3}
-\end{figure}
-
-\subsection{Variable Selection via MSR3}
+%    }
+ %   \label{fig:geometric_intuition_sr3}
+%\end{figure}
+
+\subsection{Variable Selection via MSR3}\label{sec:MSR3}
 
@@
 We call the extension MSR3, since we are focusing on mixed effects models. 
-Starting with the regularized likelihood~\eqref{eq:lme_loss_original_in_x} we introduce auxiliary parameters designed to discover the 
-fixed and random features: 
+Starting with the regularized likelihood~\eqref{eq:lme_loss_original_in_x} we introduce auxiliary parameters designed to discover the fixed and random features: 
 \eq{%\label{eq:main opt in x over C}
@@
 }
-where $\kappa_\eta$ penalizes deviations between $x$ and $w$, and also guarantees that the objective is convex with respect to the $\gamma$
-components of $x$ for sufficiently large $\eta$:
+where $\kappa_\eta$ penalizes deviations between $x = (\beta, \gamma)$ and $w = (\hat \beta, \hat \gamma)$, and also guarantees that the objective is convex with respect to the $\gamma$
+components of $x$ for sufficiently large $\eta$: \\
 \begin{equation}
 \label{eq:kappa}
-\kappa_{\eta} (\beta, \gamma) = \frac{\eta}{2}\|(\beta,\gamma)\|^2 ,
+\kappa_\eta (x-w) = \frac{\eta}{2}\|x-w\|^2 = 
+%\frac{\eta}{2}\left\| \begin{pmatrix} \beta - \hat \beta \\ \gamma - \hat \gamma\end{pmatrix}\right\|^2
+\frac{\eta}{2}\| \beta - \hat \beta \|^2 + 
+\frac{\eta}{2}\| \gamma - \hat \gamma\|^2
+%\kappa_{\eta} (\beta, \gamma) = \frac{\eta}{2}\|(\beta,\gamma)\|^2 ,
 \end{equation}
+
 with $\eta\ge\bar\eta$ where $\bar \eta$ is the weak convexity constant
@@
 As $\eta \uparrow \infty$, the extended objective~\eqref{eq:lme_loss_sr3} converges in an epigraphical sense to the original objective~\eqref{eq:lme_loss_original_in_x}. 
-However, feature selection accuracy does not require this continuation, and a fixed 
-modest value can be used~\citep{Zheng2019SR3}, e.g. $\eta=1$.
-
-To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we, for this section alone, define a simplified version of the value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
+However, feature selection accuracy does not require this continuation, 
+indeed, we show that a fixed 
+modest value such as $\eta=1$ can be used~\citep{Zheng2019SR3}.
+
+To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we define an optimal value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
 \eq{
@@
 }
-Substituting the above into (\ref{eq:lme_loss_sr3}) the problem transforms into the problem of optimizing a regularized value function:
+Substituting (\ref{eq:value_function_definition_only_eta}) into (\ref{eq:lme_loss_sr3}) 
+%the problem 
+transforms (\ref{eq:lme_loss_sr3}) into 
+%the problem of optimizing a regularized value function:
 \eq{
@@
 in order to optimize~\eqref{eq:value_function_optimization}. 
-Indeed, in \cite[Section 5]{Theory1} it is shown that $u_\gamma$ is well-defined, 
+Indeed, in \cite[Section 5]{Theory1} it is shown that 
+there exists a computable $\bar\eta>0$, which we have called the weak convexity constant, such that
+$\LL+\delta_\CC+\kappa_\eta(\cdot -w)$ is strongly convex for all 
+$\eta>\bar\eta$ regardless of the choice of $w$. This allows us to show
+that 
+$u_\eta$ is well-defined, 
 differentiable, and Lipschitz continuous, with
@@
 \end{equation}
+%We call $\bar\eta$ the weak convexity constant for 
+%$\LL+\delta_\CC+\kappa_\eta(\cdot -w)$.
 %where $\nabla_w k_\eta(x-w)$ is computed directly from~\eqref{eq:kappa}.
@@
 (\ref{eq:value_function_optimization}) has advantages over (\ref{eq:lme_loss_original_in_x})
-from an optimization perspective since $u_\eta(w)$ typically has nearly spherical level-sets while keeping the position of minima close to those of $\LL(x)$.
+from an optimization perspective since $u_\eta$ typically has nearly spherical level-sets while keeping the position of minima close to those of $\LL(x)$.
 This effect is extensively studied and validated for a quadratic loss function in the original work  of \cite{Zheng2019SR3}.
 %we provide empirical evidence that suggests that it also extends to more general non-linear functions, specifically to the marginal likelihood. 
-In Figure \ref{fig:geometric_intuition_sr3}, we plot the level-sets of $\LL(x) + \|x\|_1$ (left column) and $u(w) + \|w\|_1$ (right column) for the same mixed-effect problem. 
+In the center panel of Figure~\ref{fig:summary}, we plot the level-sets of $\LL(x) + \|x\|_1$ (left column) and $u_\eta + \|\cdot\|_1$ (right column) for the same mixed-effect problem. 
 %A more symmetrical 
@@
 
-We apply PGD to optimize the value function $u_\eta(w)$ which yields the iteration of the form
+We apply PGD to optimize the regularized value function $u_\eta$ which yields the iteration 
 \eq{
@@
 }
-Because of the results stated above, all components of the iteration~(\ref{eq:pgd_for_value_function}) are well-defined. 
+The results in \cite{Theory1} show that all components of the iteration~(\ref{eq:pgd_for_value_function}) are well-defined. 
 The equivalence of Algorithm \ref{alg:pgd_for_value} and 
@@
 \end{algorithm}
-
-While in linear regression setting of of~\cite{Zheng2019SR3}, Algorithm~\ref{alg:pgd_for_value} can be implemented exactly, in the nonlinear case 
-evaluating $x^+$ requires an iterative algorithm. In particular, we use an interior point method to account for non-negativity of $\gamma$.  
-This means that the value function $u_\eta$ is approximately evaluated, using a Newton iteration with a barrier term.  
-The degree of the approximation is controlled by the convergence criteria of the interior point algorithm.
+\smallskip
+
+In \cite[Theorem 6]{Theory1}, it is shown that
+for any sequence $\eta_k\uparrow\infty$ the associated optimal solutions $(x^k,w^k)$ 
+to (11) satisfy
+\(
+\mathcal{L}(x^k)+R(w^k)\uparrow 
+\inf_{x\in\Rp\times\Rq_+}\mathcal{L}(x)+R(x) \;
+\text{with}\; \Vert{x^k-w^k}\Vert\rightarrow 0.
+\)
+In particular, every cluster point of the sequences $\{x^k\}$ and $\{w^k\}$ are solutions
+to \eqref{eq:lme_diagonal_setup}, where such cluster points exist whenever
+the function $R$ is coercive, i.e. $\lim_{\norm{x}\uparrow\infty}R(x)=+\infty$.
+Just how close $w^k$ is to a solution to \eqref{eq:lme_diagonal_setup} remains
+an open question, however, our numerical studies in Section \ref{sec:applications}
+show that $\eta$ can be chosen surprisingly small. Indeed, we typically take $\eta=1$.
+
+In the linear regression setting of~\cite{Zheng2019SR3}, 
+Algorithm~\ref{alg:pgd_for_value} can be implemented exactly. 
+In the nonlinear case, evaluating $x^+$ requires an iterative algorithm. 
+For this we use an interior point method which replaces the indicator function
+$\delta_\CC$ by a smooth log-barrier term. This allows us to approximate both
+$u_\eta$ and its gradient where  
+the degree of the approximation is controlled by the convergence 
+criteria of the interior point algorithm.
 
@@
 \end{equation}
-This problem is well suited for an interior point approach~\citep{Kojima1991,Nesterov1994,Wright1997}. 
+This problem is well suited for an interior point approach~\citep{Kojima1991,Nesterov1994,Wright1997,vanderbei99}. 
 %We now describe how to apply IP methods \teal{to solve the inner problem~\eqref{eq:lme_loss_relaxed} with respect to $x = (\beta, \gamma)$, 
@@
 
+\edit{Connect this paragraph to the theory block}
+
 To motivate the reader, the theoretical results that support the design of Algorithm \ref{alg:MSR3} go as follows.
@@
  limit points of global solutions to the former are global solutions to the latter. 
- However, the empirical studies in Sections \ref{sec:synthetic} and \ref{sec:real}
+ However, in the applications considered here,
+ the empirical studies in Sections \ref{sec:synthetic} and \ref{sec:real}
  indicate that one does not need to make $\eta$ particularly large in order to accurately identify the correct sparsity pattern.
-These results were first published in the joint works \cite{sholokhov2022relaxation, aravkin2022jimtheory}, constitute the original work of the author of the thesis, and were mostly taken verbatim per the co-authors'  permission.
-
-%The Lagrangian for~\eqref{eq:value_function_definition} is obtained by dualizing the inequality $\gamma \geq 0$ constraint: 
-% Define the objective for \eqref{eq:value_function_definition} by
+
+%The Lagrangian for~\eqref{eq:log_barrier_relaxation_for_ip} is obtained by dualizing the inequality $\gamma \geq 0$ constraint: 
+% Define the objective for \eqref{eq:log_barrier_relaxation_for_ip} by
 % \eq{
@@
 By setting 
-\[
+\(
 v=\nabla_\gam \mathcal{L}_{\mu,\eta}(\beta,\gam)+\eta(\gamma - \hat \gamma),
-\] 
+\) 
 we can rewrite this equation as
@@
 ``$\odot$'' denotes the Hadamard (or simply element-wise) product.
-%Together with the remaining optimality conditions, we obtain a set of nonlinear equations  that form the the KKT system for~\eqref{eq:value_function_definition}:
+%Together with the remaining optimality conditions, we obtain a set of nonlinear equations  that form the the KKT system for~\eqref{eq:log_barrier_relaxation_for_ip}:
 The complete set of optimality conditions for 
@@
 that is, in each iteration the search direction $[\Delta v, \Delta \beta, \Delta \gamma]$ solves the linear system
-\eq{
-	\label{eq:ip_iteration_rs_form}
+\[
 	\nabla G_{\mu,\eta}(v, \beta, \gamma)\begin{bmatrix}
@@
 		\Delta \gamma
-	\end{bmatrix} = -G_{\mu,\eta}(v, \beta, \gamma).
-}
-where 
-\eq{
+	\end{bmatrix} = -G_{\mu,\eta}(v, \beta, \gamma), \quad
 	\nabla G_{\mu, \eta}(v, \beta, \gamma) = \begin{bmatrix}
@@
  	\end{bmatrix}
-}
+\]
 and we have used the fact that  $v\odot \gamma=\Diag{v}\gam=\Diag{\gam}v$. 
@@
 so that the update
-\eq{
-\label{eq:ip_one_step}
+\[
 \begin{pmatrix}
-v^{(k+1)}\\ \beta^{(k+1)}\\ \gamma^{(k+1)}
+v^{(k+1)}& \beta^{(k+1)}& \gamma^{(k+1)}
 \end{pmatrix}
@@
 \begin{pmatrix}
-v^{(k)}\\ \beta^{(k)}\\ \gamma^{(k)}
+v^{(k)}& \beta^{(k)} &\gamma^{(k)}
 \end{pmatrix}
@@
 \begin{pmatrix}
-\Delta v^{(k)}\\ \Delta\beta^{(k)}\\ \Delta\gamma^{(k)}
+\Delta v^{(k)} & \Delta\beta^{(k)}& \Delta\gamma^{(k)}
 \end{pmatrix}
-}
+\]
 satisfies the conditions
 % chosen to satisfy positivity and sufficient descent conditions:
-\eq{
-    \label{eq:ip_steplen_conditions}
+\[
+\begin{aligned}
 	\text{\textit{Positivity:}} \qquad & \gamma^{(k+1)} > 0,\ v^{(k+1)} > 0 \\
-	\text{\textit{Sufficient Descent:}}\qquad & \| G_\mu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|G_\mu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\| .
-}
+	\text{\textit{Sufficient Descent:}}\qquad & \| G_\emu(v^{(k+1)}, \beta^{(k+1)}, \gamma^{(k+1)})\| \leq 0.99\|G_\emu(v^{(k)}, \beta^{(k)}, \gamma^{(k)})\|,
+\end{aligned}
+\]
+where the parameter $0.99$ is used to bias toward the acceptance of a full Newton step.
 At each iteration the relaxation parameter $\mu$ is updated by the formula 
-\eq{
-    \label{eq:ip_mu_update}
+\( %\eq{
+%    \label{eq:ip_mu_update}
     \mu^{(k+1)} = {v^{(k)}}^T\gamma^{(k)}/q,
-}
+\) %}
 where ${v^{(k)}}^T\gamma^{(k)}$ is the duality gap at
 iteration $k$. The algorithm terminates when the criteria 
-\eq{
-\label{eq:ip_convergence_criterion}
+\[
 \begin{aligned}
@@
 \end{aligned}
-}
+\]
 are both satisfied, so the interior point problem is nearly stationary, and closely approximates the original problem~\eqref{eq:ineqLoss}.
@@
 \SetAlgoLined
-$w^0 = (\tbeta^0, \tgamma^0)$, $x^0 = (\beta^0, \gamma^0)$, $\gammax>\tgam^0$, $\alpha < \frac{2}{L}$ 
-for $L>0$.\\
- \While{not converged and $\tgam\le\gammax$}{
-    \text{Find } $x^+$ \text{ using \eqref{eq:ip_one_step}}$\mbox{ which satisfies }\;  \|G_{\mu,\eta}(v^+, x^+)\| \leq \text{\texttt{tol}}, \; \mu \leq \text{\texttt{tol}} $\\
+$w = w_0$ \\
+ \While{not converged}{
+    $x^+ \; \mbox{satisfies }\;  \|G_{\mu,\eta}(v^+, x^+)\| \leq \text{\texttt{tol}}, \; \mu \leq \text{\texttt{tol}} $\\
     $w^+ = \prox_{\alpha^{-1} R}(x^+)$
@@
 \end{algorithm}
-\medskip
 
@@
 For many datasets the weak convexity constant $\overline \eta$ can be extremely large 
-and difficult to compute. However, if $\overline \eta$ is too small $\nabla^2_{\gamma\gamma}\LL(\beta, \gamma)$ is 
+and difficult to compute. However, if $ \eta$ is too small 
+$\nabla^2_{\gamma\gamma}\LL_{\mu,\eta}(\beta, \gamma)$ is 
 negative-(semi)definite. Negative definite Hessians can hamper the convergence of 
@@
 \cite[Lemma 3]{Theory1} that
-\begin{equation} %\label{eq:hess LL}
+\[
 \nabla^2\LL{(\beta,\gam)}=\sum_{i=1}^m
@@
 \end{bmatrix}.
-\end{equation}
+\]
 This implies that negative eigenvalues for the Hessian must arise from the
@@
 
-\begin{algorithm}[H]
+In designing Algorithm~\ref{alg:pgd_for_value_function_fast}, we chose a particular central path parameter, $\tau = 0.5$ in line 8, 
+that controls how far the interior point method needs to proceed before we take a proximal gradient step. We explored the effect of this parameter
+on performance and timing in Section~\ref{ch:msr3_sensitivity_analysis} and found that it did not have any effect on either for values between $0.1$ and $0.9$. MSR3-fast 
+was competitive with respect to time compared to PGD and PGD with line search (also as reported in Chapter~\ref{ch:msr3_sensitivity_analysis}) for problems up to 1000 features.
+
+\begin{algorithm}[h!]
 \SetAlgoLined
@@
 
+\verify{Ensure that short and long performance tables are consistent with each other}
 \begin{table}
@@
     \centering
-	\includegraphics[width=1.0\textwidth]{figures/performance_picture_current}
+	\includegraphics[width=0.8\textwidth]{figures/benchmark}
 	\caption{\label{fig:box_with_whiskers_for_synthetic_data}	Feature selection accuracy and execution time in seconds for PGD 
@@
 
-In this section we 
-{compare the numerical performance 
-the feature selection accuracy and numerical efficiency of Algorithms 
+In this section we compare
+%compare the numerical performance 
+the feature selection accuracy and the numerical efficiency of Algorithms 
 \ref{alg:pgd_for_lme} and 
@@
 regularization parameter $\lambda$ and the coupling parameter $\eta$  were chosen.
-Our experiments on real data are presented in Section \ref{sec:real}.}
+Our experiments on real data are presented in Section \ref{sec:real}.
 
@@
 \vskip -.8in
-\begin{wrapfigure}{r}{7.5cm}
-		\includegraphics[width=7.5cm]{figures/eta_L1.pdf}
+\begin{figure}
+\center
+		\includegraphics[width=9.5cm]{figures/eta_L1.pdf}
 		\caption{Dependence of model performance on the relaxation $\eta$ for a sample problem.}\label{fig:eta}
-	\end{wrapfigure}
+\end{figure}
 
@@
 BIC criterion from \cite{Jones2011}. 
-We begin by setting a log-uniform grid of 20 candidate values for the parameter $\eta \in [10^{-4}, 10^2]$, and then for each value of $\eta$
-on this grid 
-the BIC is optimized using a golden search in $\lambda \in [0, 10^5]$. The final values of $\eta$ and $\lambda$ are then chosen
-to be those that maximize the BIC criterion.
+We set a log-uniform grid of 20 candidate values for the parameter $\eta \in [10^{-3}, 10^2]$. For each value of $\eta$,
+the BIC is optimized using a golden search in $\lambda \in [0, 10^5]$. The final values of $\eta$ and $\lambda$ are  chosen
+to maximize the BIC criterion.
 
@@
 When $\eta$ is small the coupling term does not have sufficient strength and the training does not progress far from the initial point (a fully dense vector \textbf{1} in this case).
-When the coupling is tight, the level-sets of the problem converge to those of the original problem, and thus the minimizer.
+When the coupling is tight, the level-sets and minimizers are closer to those of the the original problem.
 For the values in between, the coupling significantly improves the model's accuracy.
@@
 The results are presented in Table \ref{table:glmmlasso}.
-Overall, MSR3 executes, on average, 5 times faster in wall-clock time than \texttt{glmmLasso} and 60 times faster than \texttt{lmmLasso} and shows much higher accuracy of selecting correct fixed and random effects simultaneously.
+Overall, MSR3 executes, on average, 5 times faster in wall-clock time than \texttt{glmmLasso} and 60 times faster than \texttt{lmmLasso} and shows much higher accuracy in selecting correct fixed and random effects simultaneously.
 {The accuracy of \texttt{glmmLasso} is lower relative to the other libraries' scores likely due to its BIC selection criterion choosing dense models.} The package \texttt{lmmLasso} supports the diagonal specification of $\Gamma$, thus allowing a direct comparison with the scores from \texttt{pysr3}.
@@
 
-\begin{table}[ht]
+\begin{table}[h]
 \centering
 \resizebox{\columnwidth}{!}{\input{tables/table_glmm_current.tex}}
-
 \caption{\label{table:glmmlasso} Comparison of performance of MSR3-Fast for $\ell_1$ regularizer vs \texttt{glmmLasso}.

### 2023-04-26 introduction.tex
Rewrite the intro to a more official cadence

@@
-\textbf{\textit{The current intro is a draft that captures content and, for now, disregards the form and grammar. It will be rewritten later in a more professional cadence.}}
-\edit{Rewrite the introduction: improve style and grammar, add citations}
+Machine Learning has emerged as an important tool in applied mathematics problems as data-driven approaches have been increasingly outperforming the first-principle modeling in the tasks of classification, recognition, autonomous control, protein folding, and even playing video games.
+However, due to the absence of prior knowledge  purely data-driven approaches typically require a large amount of data to approximate behavior of first-principle models even in simplest scenarios.
+Thus the most promising modeling approaches of today are those that integrate both data-driven and first-principle components.
+These latter components frequently stem from domain expertise and pre-existing convictions.
+Incorporating prior knowledge into a data-driven method is a formidable task that demands a considerable amount of ingenuity.
+This thesis serves to illustrate two distinct examples of such endeavors.
+Together, these examples represent a diverse range of challenges in incorporating prior knowledge into a model and describe the strategies employed to surmount them.
 
-Machine Learning became an important tool in applied mathematics problems as data-driven approaches have been increasingly outperforming the first-principle ones over the last decade.
-The most successful approaches, however, combine first-principle and data-driven components.
-These first-principle components often come from our domain expertise and prior beliefs.
-Incorporating prior knowledge into a data-driven method is often challenging and requires a good share of ingenuity.
-This thesis showcases two quite different examples of such work.
-Together these examples cover a variety of challenges of incorporating priors knowledge into a model and describes approaches to overcome them.
+An instance of such a problem is feature selection for mixed-effect models.
+Feature selection problems typically arise when one believes that a small subset of features within a dataset captures the majority of the variance of the quantity of interest; this subset of features is commonly known as sparse support.
+The canonical feature selection methods typically provide control over the degree of sparsity for the solution and outputs a candidate support for it.
+In the absence of additional information about the support apart from its probable sparsity, these methods furnish a satisfactory toolkit for practitioners.
 
-One example is the feature selection problems for mixed-effect models.
-Feature selection problems typically arise when one believes that a small subset of features in dataset captures the majority of the variance of the quantity in question; this subset of features is often referred to as parse support.
-The canonical methods typically provide control over the degree of sparsity for the solution and outputs a candidate support for it.
-For a practitioner who has no additional information about the support besides its likely sparsity these methods provide sufficient toolkit.
+ Real-world practitioners, however, poses expert knowledge about the covariates in their data and wish to incorporate that knowledge as a set of priors and constraints.
+For instance, a practitioner may desire a model that employs at most 6 out of the 40 available features, with features A and B being mandatory components.
+Furthermore, if feature C is integrated, it must have a negative coefficient, and feature D must be independent of any grouping parameter.
+Attempts to directly address such constraints often result in the formulation of mathematically intractable problems.
+For example, the first constraint above makes the underlying optimization problem N-P complete.
+Additionally, practitioners require the ability to swap these priors as new hypotheses and knowledge come to light.
+This necessitates the implementation of an optimization routine that does not rely upon these priors, otherwise it would force the practitioner to re-implement the algorithm after every such swap, rendering the instrument impractical.
+Chapter~\ref{ch:msr3} elucidates my efforts in developing a universal feature selection approach for linear mixed-effects models that simultaneously confronts the numerical challenges that accompany it.
 
-The real-world practitioner, however, almost universally knows more than that.
-As a consequence, they want to incorporate their expert knowledge that mostly comes as a set of constraints.
-For example, a practitioner wants a model that uses at most 6 features out of 40, with features A and B being included at all times.
-If a feature C is included then it has to be negative, and a feature D should not depend on a grouping parameter.
-An attempt to tackle these constraints directly frequently leads to mathematically intractable problems, e.g.
-the first constraint makes the optimization problem N-P complete.
-Moreover, a practitioner wants to swap those priors in and out as more hypotheses and knowledge arise.
-It implies that the implementation of the optimization routine should not rely on the priors, otherwise it forces the practitioner to re-implement the algorithm after every such swap, rendering the instrument impractical.
-Chapter~\ref{ch:msr3} describes my work on developing such a universal feature selection approach for linear mixed-effects models while simultaneously tackling numerical challenges that arise from it.
+The second example serves to demonstrate how one can incorporate a degree of partial knowledge of physics into a deep neural network.
+The methods for data-driven modeling of physical phenomena surged in popularity, in large part due to the advancements in GPU computations and automatic differentiation tools.
+These methods have proven particularly useful in the tasks of reduced-order modeling.
+Using data a neural network can find a representation of a system on a reduced-order manifold via a non-linear transformation from an observable space to that manifold.
+Classic techniques, such as Galerkin POD or DMD, are fundamentally incapable of finding such spaces because these methods are confined to linear, albeit optimal, transformations.
+However, as the no-free-lunch theorem implies, the exceptional approximation capabilities of deep networks come at the cost of increased instability.
+In particular, the networks tend to identify manifolds that overfit the data and, consequently, fail at extrapolation tasks.
+Such deficiencies compromise the efficacy of the entire system in which the network is utilized, including applications in control, model identification, and compressive sensing.
 
-The second example illustrates how one can embed a partial knowledge of physics into a deep neural network.
-The field of data-driven modeling of physical phenomena soared in popularity over the last decade with the advancements of GPU computations and automatic differentiation tools.
-It has proven to be especially useful in the field of reduced-order modeling where a neural network can find a representation of a system on a reduced-order manifold via finding a non-linear transformation to that space based on data.
-The classic methods such as Galerkin POD or DMD are fundamentally unable to find such spaces because these methods are confined to linear, albeit optimal, transformations.
-However, as the no-free-lunch theorem goes, with great approximation power comes great instability.
-In particular, the networks tend to find manifolds that overfit the data and fail at extrapolation tasks.
-Such failures jeopardize the performance of the whole system where the network is used, e.g.
-in control, model identification, or compressive sensing applications.
-
-A natural idea for improving neural-network based models would be incorporating a partial knowledge of physics of the phenomenon that the model predicts.
-For example,  we can find a better reduced-order manifold if we know that the model predicts the behavior of a shock wave that obeys Burger's equation.
+Recent works have shown that one can improve a neural-network based model by incorporating partial knowledge of physics.
+For example, a model can find a better reduced-order manifold if it knows that the it predicts the behavior of a shock wave that obeys Burger's equation.
 Chapter~\ref{ch:pinode} is devoted to the development of PINODE ROMs -- a framework for building Reduced-Order Models with Physics-Informed Neural Ordinary Differential Equations.
-It illustrates how one can transfer knowledge from a known equation to a ROM that predicts solutions of said equation using collocation points technique.
-The resulting ROMs extrapolate significantly better forward in time as well as for unseen initial conditions, and are much less sensitive to noise.
-Finally, in Chapter~\ref{ch:cs} I show how PINODE models can be put to work in compressive sensing applications where it "stitches" separate timeframes together and allows to significantly surpass the Nyqvist sampling limit.
+It illustrates how one can transfer knowledge from a known equation to a model that predicts solutions of said equation using collocation points technique.
+The resulting ROMs are able to extrapolate forward in time considerably more effectively, as well as perform better for unseen initial conditions and exhibit less sensitivity to noise.
+Finally, in Chapter~\ref{ch:cs} I demonstrate the utility of PINODE models in compressive sensing applications, where they effectively "stitch" together separate timeframes and significantly exceed the Nyquist sampling limit.
 
@@
 Peng Zheng.
-The MSR3 framework algorithm was introduced in \cite{sholokhov2022relaxation}, with the theoretical foundations developed in \cite{aravkin2022jimtheory}, and a software was later published in \cite{sholokhov2023msr3joss}.
-Chapter~\ref{ch:pinode} represents my work at Mitsubishi Electric Research Labs (MERL) that I accomplished under the guidance of Dr.
+The MSR3 framework algorithm was introduced in \cite{sholokhov2022relaxation}, with the theoretical foundations developed in \cite{aravkin2022jimtheory}, and the software implementation published in \cite{sholokhov2023pysr3}.
+Chapter~\ref{ch:pinode} represents my work at Mitsubishi Electric Research Labs (MERL) under the guidance of Dr.
 Hassan Mansour and Dr.
@@
 Chapter~\ref{ch:cs} was developed in~\cite{sholokhov2023cs} together with Dr.
+Hassan Mansour, Dr.
+Saleh Nabi, Dr.
 J.
-Nathan Kutz and Dr.
+Nathan Kutz, and Dr.
 Steven Brunton at the University of Washington.

### 2023-04-26 msr3.tex
Add a snipet from pysr3 paper to software implementation chapter

@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \centering
@@
 
- \begin{figure}[h!]
+ \begin{figure}[ht!]
     \centering
@@
 
-\begin{algorithm}[h!]
+\begin{algorithm}[!ht]
 \SetAlgoLined
@@
 \edit{Move pysr3 software paper here}
-To ensure reproducibility of this research, all new algorithms have been implemented as a part of the \texttt{pysr3}\footnote{Available at \href{https://github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}} library.
+To ensure reproducibility of this research, all new algorithms have been implemented as a part of the \texttt{pysr3}\footnote{Available at \href{https://github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}} library~\cite{sholokhov2023pysr3}.
 This library implements functionality for fitting linear mixed models and selecting covariates.
 The user interface was designed to be fully compliant with the standards\footnote{\href{https://scikit-learn.org/stable/developers/develop.html}{https://scikit-learn.org/stable/developers/develop.html}} of \texttt{sklearn} library to minimize learning time.
+
+The baseline optimization method of PySR3 is proximal gradient descent (PGD).
+Each regularizer included in PySR3 can also be used in its relaxed SR3 form.
+More information about the structure of the library can be found in documentation\footnote{https://aksholokhov.github.io/pysr3/},
 

### 2023-05-02 pinode.tex
Add sensitivity+scalability analyses and covid example
Incorporate edits after submitting the rebuttal

@@
         Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs).
-        High-fidelity simulations of such dynamical systems can become intractable especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}.
+        High-fidelity simulations of such dynamical systems can become intractable, especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}.
 A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}.
-At the moment of writing this paper such systems are incapable of real-time complex simulations, but they already can run low-dimensional pre-trained models, thus inviting the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation} .
-ROMs, therefore, are essential for enabling design optimization, uncertainty propagation, predictive models, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}
+At the time of writing this paper, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation}.
+Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}
 
-        To be suitable for control, a ROM training method needs to find a low-dimensional manifold and a dynamics on it that together yield both high-accuracy predictions and long-term stability \cite{ahmed2021closures,noack2011reduced}.
+        In order to enable control of high dimensional dynamical systems, a ROM training method needs to  identify a low-dimensional manifold along with dynamics on the manifold that together yield high-accuracy predictions and long-term stability \cite{ahmed2021closures,noack2011reduced}.
 Most traditional ROMs are projection-based, e.g.
 dynamic mode decomposition (DMD) \cite{kutz2016dynamic,tu2013dynamic} and proper orthogonal decomposition (POD) \cite{holmes2012turbulence}, which transform the trajectories of a high-dimensional dynamical system into a suitable, and in some sense optimal, low-dimensional subspace.
-This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of basis functions (spatial modes) \cite{benner2015survey}.
-        One challenge for POD methods is their intrusive nature, i.e. requiring access to the solver codes.
-       To overcome this, operator inference approaches \cite{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and then exploit lifting to fit the latent space dynamics data into
-        polynomial, typically quadratic, models.
-        These models, however, are (i) limited in representation power (up to quadratic, e.g. for lift and learn approach) and (ii) require a custom-tailored SVD-based optimization technique.
+This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of the basis functions (spatial modes) \cite{benner2015survey}.
+One challenge for POD methods is their intrusive nature, i.e.
+requiring access to the solver codes.
+To overcome this, operator inference approaches \cite{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and  exploit lifting to fit the latent space dynamics data into polynomial, typically quadratic, models.
+These models, however, are (i) limited in representation power (up to quadratic, e.g.
+for lift and learn approach) and (ii) require a custom-tailored SVD-based optimization technique.
 
         In a thrust to overcome these challenges, significant effort has been invested into developing autoencoder-based reduced-order models, as a popular nonlinear ROM technique, which can yield both accurate and stable ROMs \cite{lee2020model,gin2021deep,champion2019data,kim2019deep}.
-In practice, however, they require datasets that densely cover a hypothetical infinite dimensional phase portrait of the dynamical system.
-Large demand for data significantly limits the use of such models in physics applications where the data can be expensive to obtain.
+In practice, however, autoencoder-based ROMs require datasets that densely cover a hypothetical infinite dimensional phase portrait of the dynamical system.
+Moreover, the large demand for training data significantly limits the use of such models in physics applications where the data can be expensive to obtain.
 
         Another severe challenge of utilizing ROMs comes from their poor out-of-distribution performance \cite{fries2022lasdi,cranmer2020discovering,gin2021deep}, especially when it is fundamentally impossible for a practitioner to obtain data that covers the entire distribution of possible data inputs.
-For example, in HVAC applications, one may collect data from a room with two windows but not from one room for every possible number of windows.
+For example, in HVAC applications, one may collect data from a room with two windows but not from a room for every possible number of windows.
 In atmospheric LiDAR applications, we may conduct experiments on a certain terrain but we can never conduct experiments on all sorts of terrains \cite{nabi2020improving}.
-In such situations embedding the knowledge of physics into a model becomes necessary to improve extrapolation performance, and for which several approaches have recently been proposed.
+In such situations embedding the knowledge of physics into a model becomes necessary to improve the extrapolation performance, and for which several approaches have recently been proposed.
         For instance, the seminal works \cite{bongard2007automated,schmidt2009distilling} have tried to determine the underlying structure of a nonlinear dynamical system from data using symbolic regression.
@@
 \cite{cranmer2020discovering} employed symbolic regression in conjunction with graph neural network (GNN), while encouraging sparse latent representation, to extract explicit physical relations.
-They showed that the symbolic expressions extracted from the GNN generalized to out-of-distribution- data better than the GNN itself.
+They showed that the symbolic expressions extracted from the GNN generalized to out-of-distribution data better than the GNN itself.
 However, symbolic regression also suffers from excessive computational costs,  and may be prone to overfitting.
@@
 For instance, \cite{fries2022lasdi,he2022glasdi} used a chain-rule based loss that ties latent-space derivatives to the observable-space derivatives for simultaneous training of the auto-encoder and the latent dynamics.
-However, such loss is highly sensitive to noise in data, especially when evaluating time-derivatives with finite differences is required\cite{delahunt2022sindynoise}.
+However, such loss is highly sensitive to noise in the data, especially when evaluating time-derivatives with finite differences is required\cite{delahunt2022sindynoise}.
         Collocation-based enforcement of the physics, i.e.
-projection of the candid functions in the governing equations to enforce the chain rule instead of finite difference, could address such numerical difficulties.
+projection of the candidate functions in the governing equations to enforce the chain rule instead of finite difference, could address such numerical difficulties.
 Recently, Liu et al.
-\cite{liu2022physics} used auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM.
+\cite{liu2022physics} used an auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM.
 However, linearity may not be expressive enough for complex dynamics with multiple basins of attraction \cite{page2019koopman}.
@@
 
-        In this paper, we employ autoencoders to perform nonlinear model reduction along with NODE at the latent space to model complex and nonlinear dynamics.
-Our goal is to reduce their demand for data and improve overall forecasting stability under challenging training conditions.
-To achieve that we build on ideas from classical collocation methods of numerical analysis to embed knowledge from a known governing equation into the latent-space dynamics of a ROM, as described in Section 2.
-Then, in Section 3, we show that addition of our physics-informed loss allows for exceptional data supply strategies that improves the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible.
+        In this paper, we employ autoencoders to perform nonlinear model reduction along with NODE in the latent space to model complex and nonlinear dynamics.
+We choose Neural ODEs in the latent space dynamics representation because of their ability to model highly non-linear dynamics, which is especially important when applications limit the size of the latent space dimension.
+Our goal is to reduce the demand for training data and improve the overall forecasting stability under challenging training conditions.
+To that end, we build on ideas from classical collocation methods of numerical analysis to embed knowledge from a known governing equation into the latent-space dynamics of a ROM, as described in Section 2.
+In Section 3, we show that the addition of our physics-informed loss allows for exceptional data supply strategies that improves the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible.
 We demonstrate that such an approach not only reduces the need for large training data-sets and produces highly-accurate and long-term stable models, but also leads to the discovery of more compact latent spaces, which is especially important for applications in compressed sensing and control.
+
         \edit{Add Koopman example to the intro}
@@
 
-            \begin{figure}[t]
-    \centering
-    \includegraphics[width=\textwidth]{figures/graphical-abstract.png}
-    \caption{We propose utilizing a collocation points technique from numerical analysis to transfer knowledge of physics into continuous-time reduced-order models (ROMs).
+%\begin{figure}[ht]
+%    \centering
+%    \includegraphics[width=\textwidth]{figures/graphical-abstract.png}
+%    \caption{We propose utilizing a collocation points technique from numerical analysis to transfer knowledge of physics into continuous-time reduced-order models (ROMs).
 Such physics-informed models yield orders of magnitude more accurate predictions in tasks of far-out forecasting, predictions for out-of-distribution initial conditions, and learning from high-noise data.
 Good performance on those tasks is crucial for using ROMs in problems of compressive sensing and control.}
-    \label{fig:architecture}
-\end{figure}
+%    \label{fig:architecture}
+%\end{figure}
 
-\section{Method}
+\section{Methods}
 \label{sec:method}
 \paragraph{Reduced-Order Model with Non-Linear Latent Dynamics}
-We consider an autonomous dynamical system on a finite space $\XX \subseteq \mathbb{R}^n$
+We consider an autonomous dynamical system on a finite space $\XX \subseteq \mathbb{R}^n$ given by
 
@@
     \label{eq:generic_stationary_ode}
-        \ddt\bd{x}(t) = \bd{f}(\bd{x}(t))
+        \ddt\bd{x}(t) = \bd{f}(\bd{x}(t)).
 \end{equation}
 
-In real-world applications it is often expensive to solve the relationship~(\ref{eq:generic_stationary_ode}) directly because $x(t)$ can be very high-dimensional.
-However, a variety of works provided both theoretical \cite{holmes2012turbulence} and practical \cite{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$.
+In real-world applications, it is often expensive to solve equation~(\ref{eq:generic_stationary_ode}) directly because $x(t)$ can be very high-dimensional.
+However, a variety of works provided both theoretical \cite{holmes2012turbulence} and empirical \cite{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$.
 In that space, the dynamics evolve according to a (generally unknown) function  $\bd{h}(\bd{z})$:
@@
 We call the space $\XX$ an observable space, and $\ZZ$ a latent space.
-When an invertible mapping $\psi:\ \ZZ \to \XX$ between the observable and the latent spaces is known, one can predict the dynamics of the system $\bd{x}$ at a future time $T$ by projecting the initial condition $\bd{x}(0)$ into the latent space, performing an integration there, and mapping the resulting trajectory back to the observable space:
+When an invertible mapping $\psi:\ \ZZ \to \XX$ between the observable and the latent spaces is known, one can predict the dynamics of the system $\bd{x}$ at a future time $T$ by projecting the initial condition $\bd{x}(0)$ into the latent space, integrating the dynamics in the latent space, and mapping the resulting trajectory back to the observable space:
 \begin{equation}
@@
 When $m << n$ we refer to the triplet $(\psi, \psi^{-1}, \bd{h})$ as a Reduced-Order Model (ROM) of $\bd{f}$.
-It is often the case that for a given system $\bd{f}$ there exists no ROM $(\psi, \psi^{-1}, \bd{h})$ such that the relation~(\ref{eq:integration_in_latent_space}) holds exactly.
+It is often the case that for a given system $\bd{f}$, there exists no ROM $(\psi, \psi^{-1}, \bd{h})$ such that the relation~(\ref{eq:integration_in_latent_space}) holds exactly.
 In this case, we seek an \textit{approximation} ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ that minimizes the difference between the data $x(t)$ and the prediction $\hat{x}(t)$ over a chosen class of models $(\psi_\theta, \phi_\theta, h_\theta)$ parameterized by $\theta$.
@@
 For example, integrating~(\ref{eq:generic_stationary_ode}) may be computationally intractable especially on platforms with limited computing capability such as embedded and autonomous devices.
-For instance, in an HVAC system, solving~(\ref{eq:generic_stationary_ode}) means solving a Navier-Stocks equation on a fine grid in real time, which exceeds the computing capabilities of current-generation appliances.
+For instance, in an HVAC system, solving~(\ref{eq:generic_stationary_ode}) means solving a Navier-Stokes equation on a fine grid in real time, which exceeds the computing capabilities of current-generation appliances.
 On the other hand, integrating~(\ref{eq:integration_in_latent_space}) may be cheap when $m << n$.
@@
 by utilizing a remote cluster), executing control over the resulting model, which is an end-goal for an HVAC system, may still be intractable.
-Indeed, executing control requires \textit{multiple} iterative evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\cite{duriez2017machine}.
+Indeed, executing control requires \textit{multiple} evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\cite{duriez2017machine}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/abstract_data_driven.png}
+    \caption{Illustration of the autoencoer structure with neural ODE in the latent space.
+The data-driven part of the loss function aims to minimize a sum of two objectives: the prediction loss and the reconstruction loss.
+The prediction loss minimizes the difference between the data trajectories and their model predictions to ensure temporal consistency of the latent space dynamics.
+The reconstruction loss ensures accurate reconstruction of individual snapshots, ensuring that the autoencoder behaves as an invertible mapping on all snapshots.}
+    \label{fig:data_driven_loss}
+\end{figure}
 
 \paragraph{Architecture} In this work we model $\psi$, $\psi^{-1}$, and $\bd{h}$ with fully-connected neural networks $\psi_\theta$, $\phi_\theta$, and $h_\theta$, respectively.
-Specifically, the pair ($\psi$, $\psi^{-1}$) is modeled with an auto-encoder $(\psi_\theta, \phi_\theta)$, and $\bd{h}$ is modeled with a fully-connected network $h_\theta$.
-Figure~\ref{fig:architecture} visualizes the architecture of the model.
+Specifically, the pair ($\psi$, $\psi^{-1}$) is modelled with an auto-encoder $(\psi_\theta, \phi_\theta)$, and $\bd{h}$ is modelled with a fully-connected network $h_\theta$.
+Figure~\ref{fig:data_driven_loss} visualizes the architecture of the model.
 
 \paragraph{Data-Driven Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses.
-The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data.
+The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data, as illustrated on Figure~\ref{fig:data_driven_loss}.
+
  Formally, for a given set of trajectories $\bd{x}_i$, $i \in [1 \dots k]$, where each trajectory $\bd{x}_i \in \mathbb{R}^{n \times p}$ is a set of $p$ snapshots that correspond to the recorded states of the system for $p$ time-steps, $t_j$, $j \in [1, \dots, p]$, the loss function $\LL^{data}_\theta$ is defined as:
@@
 where $\sigma$ is the standard deviation of the observation noise.
-We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}).
+We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and may use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}).
 This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}.
-To simplify the notation without loss of generality, in the rest of the paper we assume that all trajectories were recorded over the same time-frame with an equal and uniform step-size.
+To simplify the notation, without loss of generality, in the rest of the paper we assume that all trajectories are recorded over the same time-frame with the same uniform step-size.
+To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs)\cite{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}).
+Neural ODEs have demonstrated a better ability to model highly non-linear dynamics compared to linear models when the dimensionality of the dynamics variable is limited.
+This is especially useful in applications where the size of the latent space dimension needs to be small~\cite{lee2020model,gin2021deep,champion2019data,kim2019deep}.
 
-\paragraph{Physics-Informed Loss} In their recent work, \cite{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics.
+\paragraph{Physics-Informed Loss} In their recent work, Liu et al.~\cite{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics.
 To extend this approach to the non-linear regime, we note that for a true mapping $\phi$ the following holds:
@@
     \label{eq:chain_rule_1}
-    \frac{d\bd{z}(\bd{x}(t))}{dt} = \frac{d\bd{z}}{d\bd{x}}\frac{d\bd{x}}{dt} = \nabla\phi(\bd{x}(t))^T\bf{f}(\bd{x(t)})
+    \frac{d\bd{z}(\bd{x}(t))}{dt} = \frac{d\bd{z}}{d\bd{x}}\frac{d\bd{x}}{dt} = \nabla\phi(\bd{x}(t))^T\bd{f}(\bd{x(t)})
 \end{equation}
@@
 \end{equation}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/abstract_physics_informed.png}
+    \caption{The physics-informed loss function compares gradient fields in the current latent space with what a correctly-learned field should be in this latent space on set of collocation points.}
+    \label{fig:physics_informed_loss}
+\end{figure}
+
 Equation~(\ref{eq:physsics_informed_equation}) links the dynamics $\bd{h}(\bd{z})$ and the encoder $\phi(\bd{x})$ with the known equation $\bd{f}(\bd{x})$ and is true for all $z \in \ZZ$ and $x \in \XX$.
-Hence, knowledge of $\bd{f}$ can be assimilated into the model by evaluating Equation~(\ref{eq:physsics_informed_equation}) on a set of $N$ carefully sampled points $\bar{\bd{x}}_i \in \XX$, $i \in [1, \dots, N]$:
+Hence, as shown on Figure~\ref{fig:physics_informed_loss}, knowledge of $\bd{f}$ can be assimilated into the model by evaluating Equation~(\ref{eq:physsics_informed_equation}) on a set of $N$ carefully sampled points $\bar{\bd{x}}_i \in \XX$, $i \in [1, \dots, N]$:
 
@@
 
-We refer to the points $\bar{\bd{x}}_i$ as \textit{collocations}.
+We refer to the points $\bar{\bd{x}}_i$ as \textit{collocation points}.
 
-\paragraph{Collocations}
+\paragraph{Collocation Points}
 \label{sec:collocations_conditions} We define a collocation as pair $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$.
-Collocations are samples from the space $\XX \times Im_{f}(\XX)$, and they should satisfy three conditions, ordered by importance:
+collocation points are samples from the space $\XX \times Im_{f}(\XX)$, and they should satisfy three conditions, ordered by importance:
 \begin{enumerate}
@@
 In other words, $x_j$ should be an attainable state of the system.
-Collocation points outside of $\XX$ may downgrade the performance of the auto-encoder by forcing it to be an invertible function on a domain outside of $\XX$ where a true mapping $\psi$ operates on.
+Collocation points outside of $\XX$ may downgrade the performance of the autoencoder by forcing it to be an invertible function on a domain outside of $\XX$.
 \end{enumerate}
-Thus, an optimal sampling procedure for collocations $\bd{\bar{x}}_j$ is domain-specific and should be designed given a particular system $\bd{f}$ and available data $\bd{x}_i$.
+Thus, an optimal sampling procedure for collocation points $\bd{\bar{x}}_j$ is domain-specific and should be designed given a particular system $\bd{f}$ and available data $\bd{x}_i$.
 We show examples of how these conditions can be implemented for real systems in the following sections.
 
-Our definition of collocation points follows recent works\cite{liu2022physics}, which is a further development of the original definition \cite{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.
+The above definition of collocation points is not to be confused with a classic notion of collocation points for finding numerical solutions for differential equations~\cite{fornberg1998practical, trefethen2022numerical}.
+The classic notion refers to a set of points in time $[t_0, t_0 + c_1h, t_0 + c_2h, \dots, t_0 + h]$, $0 < c1 < c2 < \dots < 1$ which are chosen to obtain an optimal local interpolant of a solution of a differential equation for a time-period between $t_0$ and $t_0 + h$.
+For example, $s$ collocation points for Runge-Kutta methods are defined to provide an optimal Gauss-Legendre interpolant of order $s$; the coefficients $c_1, \dots, c_s$ come from a respective Butcher table.
+In contrast, we define collocation points as pairs $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$ which are  examples of mapping $x \to f(x)$.
+Our definition is built around solving an \textit{inverse} problem of approximating $\dot{x} = f(x)$ with $f_\theta(x)$ and follows a recent work~\cite{liu2022physics} which develops upon a definition from \cite{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.
 
@@
 \end{equation}
-When $\omega_1 = \omega_2 = 0$ we have that $\LL^{data}_\theta = 0$, so we say that the model is (purely) \textit{\textbf{Physics-Informed}}.
-Similarly, when $\omega_3 = \omega_4 = 0$ we have that $\LL^{physics}_\theta = 0$ and we say that the model is (purely) \textit{\textbf{Data-Driven}}.
+When $\omega_1 = \omega_2 = 0$ we have $\LL^{data}_\theta = 0$, so we say that the model is (purely) \textit{\textbf{Physics-Informed}}.
+Similarly, when $\omega_3 = \omega_4 = 0$ we have $\LL^{physics}_\theta = 0$ and we say that the model is (purely) \textit{\textbf{Data-Driven}}.
 When $\omega_i \neq 0, \, \forall i$, we say that the model is \textit{\textbf{Hybrid}}.
 
-We use a \texttt{pytorch}~\cite{NEURIPS2019_9015} implementation of Adam algorithm~\cite{kingma2014adam} for optimization.
-To evaluate $\nabla_\theta \LL^{physics}_\theta$ and $\nabla_\theta \LL^{data}_\theta$ we use \texttt{torchdiffeq}~\cite{chen2018neural} -- a \texttt{pytorch}-compatible implementation of Neural ODE framework.
+The coefficients $\omega_i$ are hyper-parameters which need to be tuned using a validation dataset.
+However, in all experiments of this paper we set $\omega_i$ to be either $0$ or $1$, and we balance $\LL^{physics}_\theta$ and $\LL^{data}_\theta$ the choice of samples in a batch of training data.
+Specifically, we set the number of collocation points per batch $N_{batch}$ to be equal to the number of trajectories per batch $k_{batch}$ times the number of time-steps$T$: $N_{batch} = Tk_{batch}$.
+In this way both $\LL^{physics}_\theta$ and $\LL^{data}_\theta$ represent the loss for $Tk_{batch}$ snapshots of the system, providing on average a similar contribution of information to the overall loss function.
+More laborious approaches of hyper-parameter tuning did not yield sufficient systematic advantage to justify the labour compared to this simple strategy.
 
-To the best of our knowledge, this is the first framework that combines non-linear latent-dynamics (Neural ODE), auto-encoders, and a physics-informed loss term~(\ref{eq:loss_physics_informed}).
+We use a \texttt{pytorch}~\cite{NEURIPS2019_9015} implementation of the Adam algorithm~\cite{kingma2014adam} for optimization.
+To evaluate $\nabla_\theta \LL^{physics}_\theta$ and $\nabla_\theta \LL^{data}_\theta$ we use \texttt{torchdiffeq}~\cite{chen2018neural} -- a \texttt{pytorch}-compatible implementation of the Neural ODE framework.
+
+To the best of our knowledge, this is the first framework that combines non-linear latent-dynamics (Neural ODE), autoencoders, and a physics-informed loss term~(\ref{eq:loss_physics_informed}).
 Thus, we call our framework \textit{Physics-Informed Neural ODE}, or PINODE.
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/duff_first_exp_abstract.png}
+    \caption{We use a toy example -- a Lifted Duffing Oscillator -- to show that it is possible to ``fill the gaps'' in data with collocation points.
+Specifically, the Hybrid model is able to learn the dynamics of two additional basins of attraction that were not represented in the dataset.
+As shown in the top-rightmost frame, without the collocation points the model does not infer the dynamics in the unseen regions correctly.}
+    \label{fig:duffing_pinode}
+\end{figure}
 
@@
 We show how a non-linear latent dynamics $\bd{h}(\bd{z})$ overcomes the limitations of DMD and Koopman networks from \cite{liu2022physics} by handling multiple basins of attraction within one model.
-We also show that using physics-informed loss is sufficient for reconstructing the behavior for basins of attraction that are not represented by the data.
-Finally, we demonstrate that a purely data-driven model may be highly-accurate short-term and highly unstable long-term, even when the data is abundant, and show that the physics-informed approach improves long-term stability of such models by multiple orders of magnitude.
+We also show that using physics-informed loss is sufficient for reconstructing the behaviour for basins of attraction that are not represented by the data.
+Finally, we demonstrate that a purely data-driven model may be highly-accurate in the short-term and highly unstable in the long-term, even when the data is abundant, and show that the physics-informed approach improves long-term stability of such models by multiple orders of magnitude.
 
 Next, we study the framework's performance on Burgers' equation.
-We show that (i) the non-linear latent dynamics yields more compact latent spaces than its linear counterpart for the same accuracy; (ii) these compact latent spaces allow for more stable long-term predictions; (iii) in the presence of significant noise in data, the use of collocations improves stability by providing an extra source of information that is noise-free, and (iv) in certain scenarios, training \textit{only} on collocations yields \textit{better} models than training on data, even when a vast amount of data is available.
-The latter shows that the contribution of physics-informed loss~(\ref{eq:loss_physics_informed}) may surpass the one of the data-based loss~(\ref{eq:loss_data_driven}), especially when the data is severely limited or noisy.
+We show that (i) the non-linear latent dynamics model yields more compact latent space representations than its linear counterpart for the same accuracy; (ii) the compact latent space representations allow for more stable long-term predictions; (iii) in the presence of significant noise in the data, the use of collocation points improves stability by providing an extra source of information that is noise-free, and (iv) in certain scenarios, training \textit{only} on collocation points yields \textit{better} models than training on data, even when a vast amount of data is available.
+The last observation shows that the contribution of the physics-informed loss~(\ref{eq:loss_physics_informed}) may surpass that of the data-based loss~(\ref{eq:loss_data_driven}), especially when the data is severely limited or noisy.
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=\textwidth]{figures/duff_first_exp_abstract.png}
-    \caption{We use a toy example -- a Lifted Duffing Oscillator -- to show that it is possible to ``fill the gaps'' in data with collocations.
-Namely, Hybrid model from Figure~\ref{fig:architecture} is able to learn the dynamics of two additional basins of attraction that were not represented in the dataset.
-As shown in the top-rightmost frame, without the collocations the model does not infer the dynamics in the unseen regions correctly.}
-    \label{fig:duffing_pinode}
-\end{figure}
-
-\newpage
 \subsection{Lifted Duffing Oscillator}
@@
 
+\begin{wrapfigure}{r}{0.25\textwidth}
+  \begin{center}
+    \includegraphics[width=0.23\textwidth]{figures/duffing_comparison.pdf}
+  \end{center}
+  \caption{Non-linearity in the latent dynamics and the autoencoder employed in teh PINODE Hybrid model are important for accurate long-term extrapolation.
+The DMD model and PIKN Hybrid model were unable to extrapolate the dynamics from collocation points.}%\vspace{-0.5in}
+  \label{fig:duffing_comparison}
+\end{wrapfigure}
 A phase portrait for 300 randomly sampled trajectories from this system is visualized on Figure~\ref{fig:duffing_pinode}, left frame.
-Depending on the total energy, each trajectory always stays in one of three regions: the left lobe, the right lobe, or the outer area, visualized in red, green, and blue respectively.
+Depending on the total energy, each trajectory always stays in one of three regions: the left lobe, the right lobe, or the outer area, visualized in red, green, and blue, respectively.
 To create a synthetic high-dimensional system that retains this property, we lift the Duffing trajectories into a higher-dimensional space by applying an invertible transformation $\mathcal{A}(\bd{z})$:
@@
 
-\begin{wrapfigure}{r}{0.25\textwidth}
-  \begin{center}
-    \includegraphics[width=0.23\textwidth]{figures/duffing_comparison.pdf}
-  \end{center}
-  \caption{Non-linearity in latent dynamics and autoencoder is important for long-term extrapolation.
-PIKN Hybrid model was unable to extrapolate the dynamics from collocations, while PINODE Hybrid model was.}%\vspace{-0.5in}
-  \label{fig:duffing_comparison}
-\end{wrapfigure}
 Hence, for this system $z \in \ZZ = \mathbb{R}^2$ and $\bd{x} \in \XX = \text{span}\{A_{:,1}, A_{:,2}\} \subseteq \mathbb{R}^{128}$.
@@
 \end{equation}
-Thus, we created a high-dimensional dynamical system with multiple basins of attraction for which the dynamics $\bd{f}$ is known.
+Thus, we created a high-dimensional dynamical system with multiple basins of attraction for which the dynamics $\bd{f}$ are known.
 
 For the experiment, we generate 6144 trajectories $\bd{x}_i$, $t=[0, 1]$, $\Delta t = 0.1$, all taken from the left lobe region (in red).
-We also sample 50000 collocations $\bar{\bd{x}}_j$ from the right (green) and the outer (blue) regions each by sampling $\bar{\bd{z}}_j \in U\left([-3/2,\, 3/2] \times [-1, 1]\right)$ and then applying the transformation~(\ref{eq:duffing_true_decoder}).
-For this example the conditions for collocations discussed in Section~\ref{sec:method} are trivially satisfied.
+We also sample 50000 collocation points $\bar{\bd{x}}_j$ from the right (green) and the outer (blue) regions each by sampling $\bar{\bd{z}}_j \in U\left([-3/2,\, 3/2] \times [-1, 1]\right)$ and then applying the transformation~(\ref{eq:duffing_true_decoder}).
+For this example the conditions for collocation points discussed in Section~\ref{sec:method} are trivially satisfied.
 
-We train two PINODE models: Data-Driven model that only uses the trajectories and Hybrid model that uses both trajectories and collocations.
-The models share the same architecture (Figure~\ref{fig:architecture}) and training parameters that are detailed in Appendix~\ref{appendix:duffing_pinode}.
+We train two PINODE models: a Data-Driven model that only uses the trajectories, and a Hybrid model that uses both trajectories and collocation points.
+The models share the same architecture and training parameters that are detailed in Appendix~\ref{appendix:duffing_pinode}.
 After training, we invert the mapping~(\ref{eq:duffing_true_decoder}) to project the models' high-dimensional predictions for unseen initial conditions onto the true low-dimensional manifold; those are visualized in  Figure~\ref{fig:duffing_pinode}.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.75\textwidth]{figures/duffing_periods.pdf}
-    \caption{ The time is measured in numbers of training time periods, i.e.
-$x=3T$ refers the time-range between two and three training time-periods away.
-The errors of a purely data-driven model grow exponentially when forecasting multiple times father in time than they were trained to do.
-In contrast, purely Physics-Informed model starts with worst interpolation accuracy but maintains nearly the same error when forecasting far out.
-Hybrid model combines the short-term accuracy of Data-Driven model and long-term stability of Physics-Informed model.}%\vspace{-0.4in}
-    \label{fig:duffing_periods}
-\end{figure}
 
@@
 This observation is consistent with the conclusions from related works~\cite{gin2021deep} that neural networks interpolate well but struggle with extrapolation tasks.
-Second, we see that collocations provided enough extra information for the model to predict nearly perfectly in regions from which no trajectories were provided.
-This observation suggests that one can use collocations to ``cover the gaps'' in data and improve the extrapolation accuracy of the model.
+Second, we see that collocation points provided enough extra information for the model to predict nearly perfectly in regions from which no trajectories were provided.
+This observation suggests that one can use collocation points to ``cover the gaps'' in data and improve the extrapolation accuracy of the model.
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.7\textwidth]{figures/duffing_periods.pdf}
+    \caption{ Box plots of the prediction error for three PINODE models: Data-Driven, Physics-Informed, and Hybrid.
+The time is measured in multiples of the training time period, i.e.
+$x=3T$ refers the time-range between two and three training time-periods away.
+}%\vspace{-0.4in}
+    \label{fig:duffing_periods}
+\end{figure}
 
-The non-linearity of the Neural ODE plays a crucial role in modeling the latent dynamics.
-To illustrate, we train a Hybrid PIKN model~\cite{liu2022physics} using the same dataset.
-PIKN differs from PINODE in using a linear latent dynamics $\frac{dz}{dt} = Lz$, where $L$ is a finite-dimensional approximation of Koopman operator, instead of a general non-linear dynamics $\frac{dz}{dt} = h_\theta(z)$.
-For PIKN we set $z \in \mathbb{R}^{16}$, an 8 times expansion of the dimension of the true manifold.
-In Figure~(\ref{fig:duffing_comparison}) we see that PIKN is unable to extrapolate the dynamics to unseen areas correctly using collocations: eventually, all trajectories "collapse" onto the same attractor.
-DMD shows even worse performance which could be attributed to the linear model reduction.
+The ability of Neural ODE to model nonlinear dynamics in the latent space is demonstrated in Figure~\ref{fig:duffing_comparison}.
+The figure shows a comparison between the Hybrid PINODE model, the Hybrid PIKN model~\cite{liu2022physics}, and DMD, all of which have been trained using the same dataset.
+PIKN differs from PINODE in that it uses linear latent dynamics $\frac{dz}{dt} = Lz$, where $L$ is a finite-dimensional approximation of the Koopman operator, instead of a general non-linear dynamics operator $\frac{dz}{dt} = h_\theta(z)$.
+For PIKN, we set $z \in \mathbb{R}^{16}$, an 8 times expansion of the dimension of the true manifold.
+We observe in Figure~\ref{fig:duffing_comparison} that PIKN is unable to extrapolate the dynamics to unseen areas correctly using the collocation points: eventually, all trajectories "collapse" onto the same attractor.
+It can also be seen that DMD shows even worse performance which could be attributed to its linear model reduction.
 
-In the next experiment we show that collocations stabilize long-term predictions of the model even when data from all parts of space is available.
-To illustrate it we generate a dataset of 6144 trajectories (2048 per red, green, and blue area each) and 50000 collocations uniformly distributed among all three lobes.
-Next, we train three models: Data-Driven, Physics-Informed, and Hybrid versions of PINODE.
-Finally, we evaluate their relative performance for ever-increasing forecast time.
-In Figure~(\ref{fig:duffing_periods}), the x-axis represents test time-period.
-For example, $x = 2T$ represents prediction errors on a time-period $[2T, 3T)$ where the model was trained on trajectories of length $T$, and on $y$-axis we plot the distribution of prediction errors for 300 unseen trajectories within that period $[2T, 3T)$.
+In the next experiment, we show that collocation points stabilize long-term predictions of the model even when data from all parts of the space are available.
+To illustrate, we generate a dataset of 6144 trajectories (2048 trajectories per red, green, and blue area) and 50000 collocation points uniformly distributed among all three lobes.
+We train three models: Data-Driven, Physics-Informed, and Hybrid versions of PINODE.
+The relative performance of the three models is evaluated in Figure~\ref{fig:duffing_periods}, where the x-axis represents the test time-horizon as multiples of the training trajectory length $T$.
+The $y$-axis shows box plots of the prediction mean squared error (MSE) corresponding to 300 unseen trajectories within the specific period.
+For example, $x = 2T$ represents the time-period $[2T, 3T)$, and the $y$-axis shows the distribution of the prediction errors within the period $[2T, 3T)$.
+Figure~\ref{fig:duffing_periods} shows that the performance of the Data-Driven model degrades quickly when the forecasting time-period increases despite its excellent performance when forecasting within its training time-period.
+The Physics-Informed model starts with modest performance over the training time horizon but maintains a stable performance when forecasting far ahead.
+The Hybrid model, in its turn, combines both near-term accuracy with long-term stability, yielding the best results over each time period.
 
-The results on Figure~(\ref{fig:duffing_periods}) show that the performance of the Data-Driven model degrades quickly when the forecasting time-period increases despite an excellent performance when forecasting within its train time-period.
-Physics-Informed model starts with modest performance but this performance does not degrade when forecasting far ahead.
-The Hybrid model, in its turn, combines both near-term accuracy with long-term stability, yielding best results over each time period.
-
-\subsection{Burger's equation}
+\subsection{Burgers' equation}
 \label{sec:burger_eqn}
-We now study the performance of our framework on Burger's equation with $[-\pi, \pi]$-periodic boundary conditions:
+We now study the performance of our framework on Burgers' equation with $[-\pi, \pi]$-periodic boundary conditions:
 \begin{equation}
@@
 where $u_t$, $u_x$, and $u_{xx}$ represent partial derivatives in time, the first, and second spatial derivatives, respectively.
-Burgers' equation is a PDE occurring in applications to acoustics, gas and fluid dynamics, and traffic flows \cite{burgers1948mathematical}.
-When $\nu$ is significantly smaller than one, the system exhibits strong non-linear behavior and is called ``advection-dominated'', otherwise when $\nu$ is large the system is called ``diffusion-dominated''.
+Burgers' equation is a PDE occurring in applications in acoustics, gas and fluid dynamics, and traffic flows \cite{burgers1948mathematical}.
+When $\nu$ is significantly smaller than one, the system exhibits strong non-linear behaviour and is called ``advection-dominated'', otherwise when $\nu$ is large the system is called ``diffusion-dominated''.
 In the case of the former, linear projection methods such as POD become inaccurate as the true solution space has a slow decaying Kolmogorov n-width, manifesting itself in slow decaying singular values \cite{peherstorfer2022breaking}.
-Therefore, in this section we focus on the advection-dominated Burger for which we set $\nu = 0.01$.
+Therefore, in this section we focus on the advection-dominated Burgers' equation for which we set $\nu = 0.01$.
 
-To generate trajectories, we discretize the spacial domain $[-\pi,\,\pi]$ with 128 grid-points, and we solve Equation~\ref{eq:burgers_equation} for $t \in [0, 2]$ with $\Delta t = 0.1$ using a spectral solver~\cite{trefethen2000spectral}.
+To generate trajectories, we discretize the spatial domain $[-\pi,\,\pi]$ into 128 grid-points, and solve Equation~\ref{eq:burgers_equation} for $t \in [0, 2]$ with $\Delta t = 0.1$ using a spectral solver~\cite{trefethen2000spectral}.
 To generate a diverse set of initial conditions we sum the first 10 harmonic terms with random coefficients:
@@
 
-To generate collocations we use the same family of functions as we used for the initial conditions in Equation~(\ref{eq:burger_initial_condition}), and we additionally randomize the presence of individual frequencies in the sum:
+To generate collocation points we use the same family of functions as we used for the initial conditions in Equation~(\ref{eq:burger_initial_condition}), and additionally randomize the presence of individual frequencies in the sum:
 \begin{equation}
@@
 \end{equation}
-We choose this family of collocations to meet the conditions~(\ref{sec:collocations_conditions}).
+We choose this family of collocation points to meet the conditions~(\ref{sec:collocations_conditions}).
 First, this family is representative of the state space $\XX\times Im_f(\XX)$ in the region of interest (moving wave-fronts).
 Second, (\ref{eq:burger_collocations})~is a smooth set of functions that does not contain unattainable states.
-Finally, and more importantly, the values $u_x$ and $u_{xx}$ and, consequently $u_t$ can be computed analytically, which makes it especially cheap to sample large amounts of collocations.
+Finally, and more importantly, the values $u_x$ and $u_{xx}$ and, consequently $u_t$ can be computed analytically, which makes it especially cheap to sample large numbers of collocation points.
 
-\subsubsection{Compressibility of the Latent Space}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{figures/burgers_examples_of_ics.pdf}
-    \caption{Examples of "harmonic", "bell-curve", and "bump" initial conditions, as well as the resulting solutions, in columns 1, 2, and 3 respectively.}
-    \label{fig:burgers_examples_of_ics}
-\end{figure}
+\subsection{Compressibility of the Latent Space}
 
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{figures/compressibility.pdf}
-    \caption{PINODE Hybrid model utilized the latent space dimension 5 times more efficiently by MSE than PIKN Hybrid model when modeling low-viscosity (highly-nonlinear) Burger's equation (left frame).
-The difference in performance grows to x100, when forecasting two times farther than the train period (central frame).
-The reason behind PIKN's long-term instability is the presence of eigenvalues with positive real part in the latent dynamics matrix (right frame).}
+    \includegraphics[width=0.8\textwidth]{figures/compressibility.pdf}
+    \caption{PINODE Hybrid model utilized the latent space dimension 5 times more efficiently in terms of MSE than PIKN Hybrid model when modelling low-viscosity (highly-nonlinear) Burgers' equation (left frame).
+The difference in performance grows to x100, when forecasting two times farther than the training period (central frame).
+PIKN suffers from long-term instability due to the presence of eigenvalues with positive real part in the latent dynamics matrix (right frame).
+In this frame we plot all the eigenvalues of the latent-space matrix for each PIKN model from frames 1-2.
+The legend in the right frame refers to the dimension of the latent space used by the corresponding PIKN model.}
     \label{fig:burgers_compressibility}
@@
 
-In Section~\ref{sec:duffing} we showed that a non-linear finite-dimensional latent dynamics model can be a necessity for building a compact ROM for a high-dimensional system.
-It is \textit{not} the case for Burgers' equations since there is Cole-Hopf transformation that linearize the dynamics.
-However, a latent-space non-linearity can, in principle, be utilized for finding a more compact latent space, or for increasing the forecast accuracy for the same size of the latent space.
-In this section we give an example of how PINODE achieves both goals.
+In Section~\ref{sec:duffing}, we showed that a non-linear finite-dimensional latent dynamics model can be necessary for building a compact ROM for the high-dimensional lifted Duffing system.
+That is \textit{not} necessarily the case for Burgers' equation since there exists the Cole-Hopf transformation that linearizes the dynamics for Burgers' equation.
+However, a latent-space non-linearity can, in principle, be utilized for finding a more compact latent space representation, or for increasing the forecast accuracy for a fixed latent space dimension.
+In this section, we demonstrate how PINODE can achieve both goals.
 
 For this experiment we generate 16384 trajectories as described in~(\ref{eq:burger_initial_condition}).
-We also generate 100000 collocations as described in~(\ref{eq:burger_collocations}).
-We use the exceedingly large amount of data to allow the models to achieve the best performance for the allowed dimension of the latent space.
-We evaluate the performance of the models on test data with two different time-frames: (1) same as the training data (\textit{interpolation}), and (2) two times longer than training data (\textit{extrapolation}).
-More details of the experimental setup are provided in Appendix~(\ref{appendix:burgers_compressibility}).
+We also generate 100000 collocation points as described in~(\ref{eq:burger_collocations}).
+The purpose of using such a large amount of data is to allow the trained models to achieve the best performance for the specified latent space dimension.
+We evaluate the performance of the models on test data with two different time-frames: (1) same as that of training data (\textit{interpolation}), and (2) two times longer than that of the training data (\textit{extrapolation}).
+More details on the experimental setup are provided in Appendix~(\ref{appendix:burgers_compressibility}).
 
-We compare the performance of three models: DMD, PIKN Hybrid, and PINODE Hybrid.
-First, we notice that DMD, despite achieving small loss values on training data ($\sim 10^{-3}$), does not perform well on test data.
-This observation is consistent with earlier works~(\cite{kalur2021robust,kutz2016dynamic}); it illustrates well that a combination of a linear encoder and a linear latent dynamics may not be sufficient for modeling highly-nonlinear phenomena.
-Second, we notice that PINODE achieves better performance within a given latent space budget.
-For instance, for $m = 16$ (Figure~(\ref{fig:burgers_compressibility}), left pane), PINODE achieves $\sim 5$ times lower mean squared error than PIKN, which achieves the same performance only when $m = 512$.
-More importantly, PINODE maintains low error when predicting for a longer-term horizon (extrapolation in time), which is not the case for PIKN (Figure~(\ref{fig:burgers_compressibility}, right pane).
-This is a consequence of the latent-dynamics matrix ($h(z) = Lz$) of PIKN having eigenvalues with positive real parts, which implies long-term instability (Figure~(\ref{fig:burgers_compressibility}), right pane).
-Although there has been progress in the literature~\cite{kojimalearning}, further research is needed to understand (i) how to enforce stability constraints for PIKN, and (ii) why one does not need the same enforcement for PINODE to exhibit stable behavior.
+In Figure~\ref{fig:burgers_compressibility}, we compare the performance of the three models: DMD, PIKN Hybrid, and PINODE Hybrid.
+First, we notice that DMD does not perform well on the test data, despite achieving a training loss ($\sim 10^{-3}$).
+This observation is consistent with earlier works~(\cite{kalur2021robust,kutz2016dynamic}); and illustrates well that a combination of a linear encoder and a linear latent dynamics operator may not be sufficient for modelling highly-nonlinear phenomena.
+Second, we notice that PINODE achieves better performance for a given latent space dimension compared to PIKN.
+For instance, for $m = 16$ (Figure~\ref{fig:burgers_compressibility}, left pane), PINODE achieves $\sim 5$ times lower mean squared error than PIKN, which achieves the same performance only when $m = 512$.
+More importantly, PINODE maintains a low prediction error over a longer-term horizon (extrapolation in time), which is not the case for PIKN (Figure~\ref{fig:burgers_compressibility}, center pane).
+This is a consequence of the latent-dynamics matrix ($h(z) = Lz$) of PIKN having eigenvalues with positive real parts, which implies long-term instability (Figure~\ref{fig:burgers_compressibility}, right pane).
+Although there has been progress in the literature~\cite{kojimalearning}, further research is needed to understand (i) how to enforce stability constraints for PIKN, and (ii) why one does not need the same enforcement for PINODE to exhibit stable behaviour.
 
-    \begin{figure}[t!]
+	\begin{figure}[ht!]
         \centering
@@
 
-\subsection{Training in Low-Data Regime with Collocations}
+\subsection{Training in Low-Data Regime with Collocation Points}
 \label{sec:data_vs_collocations}
 
-\begin{figure}
+In the next experiment, we study the relative efficiency of using collocation points against using data in a low-data regime.
+It is frequently the case that only a small number of simulations (or measurements) can be obtained for a physical system of interest due to the computational, time, or budget constraints.
+We would like to compensate the lack of sufficient data with providing collocation points which are considerably cheaper to generate.
+In this section, we show that, when chosen appropriately, collocation points can be effectively used for training a model in the low-data regime, and their contribution to a model's accuracy may even surpass the contribution of the data.
+\begin{figure}[ht]
     \centering
-    \includegraphics[width=\textwidth]{figures/data_vs_collocations.pdf}
-    \caption{The figure shows a comparison of the achievable MSE relative to the full data regime (1024 trajectories).
-When the data is scarce, collocations-based physics-informed loss improves the forecasting accuracy of ROMs by an average of 5 times lower MSE compared to the data-only regime, as shown in this experiment with Burger's equation.
-Moreover, when other types of initial conditions (``harmonic'', ``bell-curve'') are used, the physics-only model (top-right corner of the right frame) outperformed the most data-rich model in our experiment (bottom-left corner).
+    \includegraphics[width=0.8\textwidth]{figures/burgers_examples_of_ics.pdf}
+    \caption{Examples of "harmonic", "bell-curve", and "bump" initial conditions, as well as the resulting solutions, in columns 1, 2, and 3, respectively.}
+    \label{fig:burgers_examples_of_ics}
+\end{figure}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/data_vs_collocations.pdf}
+    \caption{Comparison of the achievable MSE relative to the full data regime (1024 trajectories).
+When the data is scarce, collocations-based physics-informed loss improves the forecasting accuracy of ROMs by an average of 5 times lower MSE compared to the data-only regime, as shown in this experiment with Burgers' equation.
+When other types of initial conditions (``harmonic'', ``bell-curve'') are used, the physics-only model (top-right corner of the right frame) outperformed the most data-rich model in our experiment (bottom-left corner).
 }
@@
 
-In the next experiment we study relative efficiency of using collocations against using data in a low-data regime.
-It is frequently the case that only a small number of simulations (or measurements) can be obtained for a physical system of interest due to the computational, time, or budget constraints.
-We would like to compensate the lack of data with providing collocations which are considerably cheaper to generate.
-The collocations, however, can be a ``weaker'' signal for the model than real data, especially when time passes and the shape of the solution moves out of the collocation family's span.
-In this section we show that collocations can be effectively used for this goal, when chosen right, and their contribution to a model's accuracy may even surpass the contribution of the data.
+To illustrate the trade-off between data and collocations, we train one model using varying combinations of the number of trajectories vs collocation points in their training datasets.
+To gauge the extrapolation power of our models, we use trajectories with three types of initial conditions: ``harmonic'', ``bell-curve'', and ``bumps'' (see Figure~\ref{fig:burgers_examples_of_ics} for illustrations).
+We generate 1024 trajectories with ``bumps'' initial conditions for the training data, and use the harmonic family of initial conditions as described in~(\ref{eq:burger_collocations}) for generating the training collocations.
+We use two test datasets: (1) 100 trajectories with ``bump'' ICs to assess within-distributuion performance, left frame), and (2) a mix of trajectories with ``bump'', ``bell-curve'', and ``harmonic'' initial conditions, 100 trajectories each, to assess out-of-distribution performance.
+All test data trajectories are two times longer than the training trajectories.
+More details on the experimental setup are provided in Appendix~\ref{appendix:burgers_data_vs_collocations}.
+Figure~\ref{fig:burger_data_vs_collocations} presents the reconstruction MSE of the test datasets obtained from a PINODE models that were trained on varying combinations of trajectories and collocation points as a percentage of the MSE achievable by a PINODE model that was trained on the full 1024 trajectories alone (no collocations).
+The PINODE models all use a latent space dimension $m=16$.
 
-\begin{figure}
+Figure~\ref{fig:burger_data_vs_collocations} demonstrates that adding collocation points consistently improves the model performance in our experiments.
+Moreover, when a sufficient number of collocation points is added in training, the model with fewer training trajectories was always able to outperform the model that was trained on all the available trajectories and no collocations.
+On average, a collocation-aided model was \textit{5 times better} at both within-distribution and out-of-distribution reconstruction relative to a purely data-driven version of the model.
+In addition, we noticed that a model that used only collocation points can perform better than a data-rich model, especially when predicting the dynamics of the unseen initial conditions (Figure~\ref{fig:burger_data_vs_collocations}, right pane, top-right vs bottom-left corner).
+
+\begin{figure}[ht]
     \centering
-    \includegraphics[width=\textwidth]{figures/example_burgers.pdf}
-    \caption{The first subplot shows the relative error of solving Burger's equations on 100 test (unseen) initial conditions for two models: PINODE Hybrid and PINODE Data-Driven.
-Both models interpolate well but a purely data-driven model fails to extrapolate past the time-period that it was trained on (to the left of the red line).
-PINODE-Hybrid provides stable long-term predictions that indicates a high-quality of a discovered low-dimensional manifold dynamics.
-An example of such effect is provided on subplots~2-4.}
+    \includegraphics[width=0.9\textwidth]{figures/example_burgers.pdf}
+    \caption{The first subplot shows the relative error of solving Burgers' equations on 100 test (unseen) initial conditions for two models: PINODE Hybrid and PINODE Data-Driven.
+Both models interpolate well but a purely data-driven model fails to extrapolate past the training time-horizon (left of the red vertical line).
+PINODE-Hybrid provides stable long-term predictions that points to its ability to correctly discover the low-dimensional manifold dynamics.}
     \label{fig:burgers_example}
@@
 
-To illustrate the trade-off between data and collocations, we train one model per various combinations of number of trajectories vs collocations in their training sets.
-To gauge the extrapolation power of our models, we use trajectories with three types of initial conditions: ``harmonic'', ``bell-curve'', and ``bumps'' (see Figure~(\ref{fig:burgers_examples_of_ics}) for equations and illustrations).
-For training, we use 2048 trajectories with ``bumps'' initial conditions.
-For collocations, we use the harmonic family as described in~(\ref{eq:burger_collocations}).
-We use two test datasets: (1) 100 trajectories with ``bump'' ICs to assess within-distributuion performance, left frame), and (2) a mix of trajectories with ``bump'', ``bell-curve'', and ``harmonic'' initial conditions, 100 of each, to assess out-of-distribution performance.
-All test data trajectories were two times longer than the train trajectories.
-More details on the experimental setup are provided in Appendix~\ref{appendix:burgers_data_vs_collocations}.
-We fit one PINODE model with the latent dimension $m = 16$ per each combination of the amount of data and collocations.
-On Figure~(\ref{fig:burger_data_vs_collocations}) we report its performance on two test datasets relative to the richest data-driven model with no collocations (bottom-left corner).
-
-From the results displayed on Figure~(\ref{fig:burger_data_vs_collocations}) we see that adding collocations always improved the model in our experiments.
-Moreover, when a sufficient number of collocations is added in training, the model with fewer data snapshots was always able to outperform the model with the maximum amount of data but no collocations.
-On average, a collocation-aided model was \textit{5 times better} at both within-distribution and out-of-distribution performance relative to a purely data-driven version of the model.
-In addition, we noticed that a model that used only collocations can perform better than a data-rich model, especially when predicting the dynamics of the unseen initial conditions (Figure~(\ref{fig:burger_data_vs_collocations}), right pane, top-right vs bottom-left corner).
-
 We also notice that the Hybrid models yield more stable and accurate predictions, relative to their purely data-driven counterparts, when forecasting far beyond the training time-period.
-In Figure~(\ref{fig:burgers_example}) we visualize the predictions for a test IC for two models: Data-Driven model from the bottom-left corner of Figure~(\ref{fig:burger_data_vs_collocations}), and a Hybrid model from the bottom-right corner of Figure~(\ref{fig:burger_data_vs_collocations}).
+In Figure~\ref{fig:burgers_example} we visualize the predictions for a test IC for two models: Data-Driven model from the bottom-left corner of Figure~(\ref{fig:burger_data_vs_collocations}), and a Hybrid model from the bottom-right corner of Figure~(\ref{fig:burger_data_vs_collocations}).
 The red line separates the time-period of training from the time-period of forecasting.
 The hybrid model's errors stay below $10^{-2}$ even when forecasting 10 times farther than what it was trained on.
-In contrast, Data-Driven model shows similarly low errors within its training time-region but the forecast errors grow quickly when forecasting beyond that.
+In contrast, the Data-Driven model shows low errors within its training time-region but the forecast errors grow quickly when forecasting beyond that.
 
-\subsection{Robustness to Noise in Low-Data Regime}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/burgers_collocations_rmse.pdf}
+    \caption{Collocation points improve results of all three models but they don't fix models' inherent shortcomings like instabilities in linear latent dynamics.\label{fig:burgers_collocations_rmse}}
+\end{figure}
+
+Finally, we observe that using collocation points can benefit other models, like DMD and PIKN.
+To illustrate, we replicate the experiments from Figure~\ref{fig:burger_data_vs_collocations} where the number of trajectories is 256 and with Bump ICs for PINODE, PIKN, and DMD.Figure~\ref{fig:burgers_collocations_rmse} shows the root mean squared error (RMSE) for the test data predictions as a function of the number of collocation points that were used in training.
+The figure illustrates the prediction error for increasing prediction horizons going from left to right, and demonstrates that in all cases, PINODE benefits from the available collocation points.
+The leftmost panel shows that every model improves its one-step-ahead predictions, with DMD quickly achieving near-optimal performance.
+However, once the forecast horizon is increased to 20 timesteps ahead (length of the training trajectories) and above, DMD failed to correctly forecast the long-term trajectories and was removed from those figure to improve legibility.
+The PIKN models improved the one-step-ahead (1st pane) and interpolation performance (2nd pane) by a factor of 4.
+It also improved the extrapolation performance for 40-steps prediction (3rd pane) but failed to extrapolate for 80 steps (4th pane, removed for legibility).
+We attribute this behavior of PIKN to the possibility that the latent dynamics operator of PIKN contains positive eigenvalues despite the use of collocation points.
+
+\subsection{Robustness to Noise in the Low-Data Regime}
 \label{sec:burger_noise}
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/burgers_noise.pdf}
+    \includegraphics[width=0.9\textwidth]{figures/burgers_noise.pdf}
     \caption{Physics-informed loss works as a safeguard that prevents unbounded performance drop when quality of the data degrades due to noise.
@@
 
-In this section we show that the use of collocations improves the ROM's robustness when the data is noisy by providing an alternative, noise-free, source of information.
+In this section we show that the use of collocation points improves the ROMs' robustness to noise in the data by providing an alternative, noise-free, source of information.
 
-For this experiment, we use the datasets from the bottom-right cell of Figure~(\ref{fig:burger_data_vs_collocations}), Pane~2.
-Namely, we use 1024 trajectories with "bump" initial conditions (an example is on Figure~(\ref{fig:burgers_examples_of_ics})), and we use 65536 "harmonic" collocations as defined in Equation~\ref{eq:burger_collocations}.
+For this experiment, we use the Burgers' equation dataset containing 1024 trajectories with "bump" initial conditions, and 65536 "harmonic" collocation points as defined in Equation~\ref{eq:burger_collocations}.
 We then add i.i.d.
-Gaussian noise to the trajectories, with the variance ranging from $\sigma = 10^{-4}$ to $\sigma = 10$.
+Gaussian noise to the trajectories, with variance ranging from $\sigma = 10^{-4}$ to $\sigma = 10$.
 For reference, most of the data values lie between $0$ and $1$, so a noise level with $\sigma > 1$ dominates the data.
 We train four models: PINODE Hybrid, PINODE Data-Driven, PINODE Physics-Informed, and DMD.
-To measure the models' out-of-distribution prediction errors we use the Test Dataset 2 from the previous section; this dataset consists of three types of noise-free trajectories (see Figure~\ref{fig:burgers_examples_of_ics}), 100 each.
-The errors are displayed on Figure~(\ref{fig:burger_noise}), left pane.
-The error of a purely Physics-Informed model (in red) is flat because the collocations are noise-free and thus the same for all noise-levels, so we only train this model once.
+To measure the models' out-of-distribution prediction errors, we use the test dataset with Bump, Gaussian, and Harmonic initial conditions, as described in the previous subsection.
+The prediction errors are displayed in Figure~\ref{fig:burger_noise}, left pane.
+The prediction error of a purely Physics-Informed model (in red) is flat because the collocation points are noise-free.
 
-On Figure~(\ref{fig:burger_noise}) we see that, when noise is high, the error of purely data-driven models grows unbounded, whereas the performance of the hybrid method converges to the performance of the Physics-Informed model.
-We hypothesise that it happens because the second part~($\LL_{\theta}^{data}$) of the combined loss~(Eq.~\ref{eq:loss_combined}) turns into noise, and so its derivative also turns into noise.
+Figure~(\ref{fig:burger_noise}) shows that in the high noise setting, the error of purely data-driven models (DMD and PINODE Data-Driven) grows unbounded, whereas the performance of the hybrid model converges to the performance of the Physics-Informed model as the noise level increases.
+We hypothesise that such behavior is due to the second part~($\LL_{\theta}^{data}$) of the combined loss~(Eq.~\ref{eq:loss_combined}) turns into noise, and so its derivative also turns into noise.
 \begin{equation}
@@
 \end{equation}
-Thus, one can think about optimizing a hybrid model~(\ref{eq:loss_combined}) as about training Physics-Informed model~(\ref{eq:loss_physics_informed}) using a noisy gradient descent with a fixed-variance noise.
-From the optimization literature \cite{friedlander2012hybrid,patel2021global,shapiro2021lectures} we know that, under certain conditions, such SGD converges to a neighbourhood of a local minima of its loss (in this case $\LL_{\theta}^{physics}$) with high probability.
-So instead of diverging, a hybrid model turns into Physics-Informed model; where the latter works as a performance safeguard in the high-noise regime.
-On the right hand-side of Figure~(\ref{fig:burger_noise}) we show an example of the prediction performance of each of the models described above.
+Thus, one can think about optimizing a hybrid model~(\ref{eq:loss_combined}) as about training a Physics-Informed model~(\ref{eq:loss_physics_informed}) using a noisy gradient descent with a fixed-variance noise.
+From the optimization literature \cite{friedlander2012hybrid,patel2021global,shapiro2021lectures} we know that, under certain conditions, such SGD converges to a neighbourhood of a local minimum of its loss (in this case $\LL_{\theta}^{physics}$) with high probability.
+So instead of diverging, a hybrid model turns into a Physics-Informed model; where the latter works as a performance safeguard in the high-noise regime.
+On the right hand-side of Figure~(\ref{fig:burger_noise}), we show an example of the prediction performance of each of the models described above.
 The data-driven and hybrid models yield visually similar solutions when $\sigma = 10^-3$.
-However, the former provides inadequate performance when the data is dominated by noise (which is expected), whereas a hybrid model in this regime produces a solution that is visually similar to the one that Physics-Informed model produces.
+However, the former provides inadequate performance when the data is dominated by noise, whereas a hybrid model in this regime produces a solution that is visually similar to the one that the Physics-Informed model produces.
 A more rigorous analysis of this phenomenon seems possible but lies outside of the scope of this paper.
@@
 
-In this work we demonstrated how a collocation point technique can improve the performance of an emerging class of continuous-time neural-network based reduced-order models.
-First, it can ``cover the gaps'' in datasets and inform the model about underrepresented basins of attraction.
-It alleviates the demand for large volumes of data that is common in network-based models, which is crucial in applications where the data is scarce and expensive.
-Second, the physics-informed loss may work as a safeguard, providing a noise-independent source of truth.
-Third, collocations can stabilize the model's long-term predictions, allowing to accurately forecast far beyond the training horizon.
-Finally, together with using a NODE-based non-linear latent dynamics, adding physics-informed loss leads to discovery of more compact latent spaces that yield more accurate models.
+In this work, we demonstrated how a collocation point-based technique can improve the performance of an emerging class of continuous-time physics-informed neural-network based reduced-order models.
+First, we demonstrated that the incorporation of collocation points in training data can ``cover the gaps'' in training trajectories and inform the model about underrepresented basins of attraction.
+Such an approach alleviates the demand for large volumes of data that is common in network-based models, which is crucial in applications where data is scarce and expensive.
+Second, the physics-informed loss may work as a safeguard, providing a noise-free source of underlying dynamics.
+Third, collocation points can stabilize the model's long-term predictions, allowing for accurate forecasting far beyond the training time horizon.
+Finally, together with using a NODE-based non-linear latent dynamics, adding physics-informed loss leads to the discovery of more compact latent space representations that also yield more accurate models.
 Simultaneous stability and compactness is especially important if one aims to use models together with compressive sensing and control algorithms.
+With respect to the computational complexity, we note that adding $Tk$ collocation points to the training imposes less of a computational burden than adding $k$ data trajectories because collocation points do not require computing integrals forward in time as in the case of data trajectories.
 
@@
 The authors believe that such decisions can be automated by adopting existing approaches from classic works on numerical approximations of PDEs, which we leave for future research.
-Another automation that prompts future research is more efficient way of sampling collocation points, possibly via applying modern adaptive learning techniques \cite{subramanian2022adaptive}.
-Finally, although Section~\ref{sec:burger_noise} provides some rationale why one may expect robustness of Hybrid models under noise, the authors believe that a more rigorous analysis is possible; particularly one that provides conditions under which such robustness is guaranteed.
+Another automation that prompts future research is deriving efficient ways of sampling collocation points, possibly via applying modern adaptive learning techniques \cite{subramanian2022adaptive}.
+Finally, although Section~\ref{sec:burger_noise} provides some rationale for why one may expect robustness of Hybrid models under noise, the authors believe that a more rigorous analysis is possible; particularly one that provides conditions under which such robustness is guaranteed.

### 2023-05-02 conclusion.tex
Add conclusions

In this document I gave an example of how partial knowledge can be utilized for model identification. First, in Section~\ref{sec:MSR3} we developed an optimization algorithm for feature selection in mixed-effect models. We showed how complex prior knowledge can be embedded into the model via proximal operator and the resulting model can be minimized using a proximal gradient descent. To tackle ill-conditioning arising from such combination of priors we propose a MSR3 relaxation and prove that its solution approaches the solution of the original problem as the relaxation tightens. Next, we showed how a partial knowledge of physics can be embedded into a reduced-order models. We achieve it by projecting true dynamics into a chosen latent space and requiring the gradient field in that space matching the expected dynamics. We show that this methodology helps to identify more stable latent manifolds. It leads to more accurate predictions and lesser sensitivity to noise. We also showed how the ROMs above can be used for compressive sensing and showed that they accurately reconstruct the dynamics from partial observations, significantly surpassing current state-of-the-are methods.

\paragraph{Perspective Research Directions} Due to the limited resources many directions were left for future work. First, given the success of SR3 framework in linear and linear mixed-effect models it should be extended to generalized linear and generalized mixed-effects models. The fundamental difficulty that arises in such extension is the absence of closed-form marginal log-likelihood. It implies that when one applies an SR3 relaxation on top of the likelihood they would have three nested levels of optimization. A solution for such obstacle could be in re-using the recipe that we developed in Section~\ref{sec:MSR3}. Namely, Algorithm~\ref{alg:pgd_for_value} had two nested optimization methods: for evaluating the value function and for optimizing over it. However we showed that one can ``blend'' both levels of optimization and achieve superior performance results. The same methodology of ``blending'' could possibly be applied to more than two nested optimization loops.

	Our developments in PINODE are also incomplete and invite future work in several promising directions. First, a more systematic way of choosing collocation points would greatly benefit the framework. In Section~\ref{sec:collocations_conditions} we define collocation points and provide a criteria which a good choice of colocations meets. However, we ultimately leave the reader without a constructive algorithm of identifying a collocation family suitable for their problem. At the same time, it has been known that for certain families of differential equations there are basis functions that can span their solutions. Such basis functions frequently poses easily-computable derivatives, which makes them perfect candidates for collocation points for PINODE. Bridging the gap between this new methodology and those classic results would undoubtedly yield a fruitful line of works. Finally, one could utilize PINODE models not only in compressive sensing applications but in any application where one needs to quickly search over a large space of possible simulations. These examples include applications such as medical imaging, model discovery, and online control.

### 2023-05-03 conclusion.tex
Rewrite the conclusion

@@
-In this document I gave an example of how partial knowledge can be utilized for model identification.
-First, in Section~\ref{sec:MSR3} we developed an optimization algorithm for feature selection in mixed-effect models.
-We showed how complex prior knowledge can be embedded into the model via proximal operator and the resulting model can be minimized using a proximal gradient descent.
-To tackle ill-conditioning arising from such combination of priors we propose a MSR3 relaxation and prove that its solution approaches the solution of the original problem as the relaxation tightens.
-Next, we showed how a partial knowledge of physics can be embedded into a reduced-order models.
-We achieve it by projecting true dynamics into a chosen latent space and requiring the gradient field in that space matching the expected dynamics.
-We show that this methodology helps to identify more stable latent manifolds.
-It leads to more accurate predictions and lesser sensitivity to noise.
-We also showed how the ROMs above can be used for compressive sensing and showed that they accurately reconstruct the dynamics from partial observations, significantly surpassing current state-of-the-are methods.
+In this work I gave two examples of how partial knowledge can be utilized to improve performance of machine learning models.
+First, in Section~\ref{sec:MSR3} we showed how complex priors can be embedded into a linear mixed-effects model via a suitable proximal operator, and how the resulting algorithm can be implemented using a proximal gradient descent.
+We note that a complex combination of priors often yields ill-conditioned problems, and we propose a MSR3 relaxation to address it.
+We also prove that a solution of such relaxation approaches the solution of the original problem as the relaxation tightens.
+Next, we showed how partial knowledge of physics can be embedded into deep-learning-based reduced-order models.
+Namely, we show how to project known physical equations that govern behavior of the system in an observable space into a chosen latent space.
+In that space one can require that the gradient field should match the projected true dynamics.
+We show that this methodology leads to identification of more stable latent manifolds, and using such manifolds for forecasting leads to more accurate predictions and lesser sensitivity to noise.
+We also showed how the reduced-order models described above can be used for compressive sensing.
+We demonstrated that they can accurately reconstruct the dynamics from partial observations, significantly surpassing current state-of-the-art compressive-sensing methods.
+We hope that these developments pave the way to higher-quality machine learning models which effectively marry data and the humanity's hard-won expert knowledge.
 
-\paragraph{Perspective Research Directions} Due to the limited resources many directions were left for future work.
-First, given the success of SR3 framework in linear and linear mixed-effect models it should be extended to generalized linear and generalized mixed-effects models.
-The fundamental difficulty that arises in such extension is the absence of closed-form marginal log-likelihood.
-It implies that when one applies an SR3 relaxation on top of the likelihood they would have three nested levels of optimization.
-A solution for such obstacle could be in re-using the recipe that we developed in Section~\ref{sec:MSR3}.
-Namely, Algorithm~\ref{alg:pgd_for_value} had two nested optimization methods: for evaluating the value function and for optimizing over it.
+\paragraph{Perspective Research Directions} Due to the limited time and resources we left many promising directions for future exploration.
+One such direction is extending sparse relaxed regularized regression (SR3) to generalized linear and generalized mixed-effects models.
+The fundamental difficulty that arises in such extension is the absence of closed-form marginal log-likelihoods.
+It implies that one would have three nested levels of optimization when they apply an SR3 relaxation on top of such likelihood: evaluating the likelihood, evaluating a value function, and optimizing over the value function.
+Re-using the recipe that we developed in Section~\ref{sec:MSR3} could help to surmount such obstacle.
+For example, Algorithm~\ref{alg:pgd_for_value} had two nested optimization methods: for evaluating the value function and for optimizing over it.
 However we showed that one can ``blend'' both levels of optimization and achieve superior performance results.
-The same methodology of ``blending'' could possibly be applied to more than two nested optimization loops.
+The same methodology of ``blending'' could possibly be applied to more than two nested optimization loops to yield algorithms that, in certain scenarios, work faster than non-nested optimization methods.
 
-	Our developments in PINODE are also incomplete and invite future work in several promising directions.
-First, a more systematic way of choosing collocation points would greatly benefit the framework.
-In Section~\ref{sec:collocations_conditions} we define collocation points and provide a criteria which a good choice of colocations meets.
-However, we ultimately leave the reader without a constructive algorithm of identifying a collocation family suitable for their problem.
-At the same time, it has been known that for certain families of differential equations there are basis functions that can span their solutions.
+	Our development of physics-informed reduced-order models also invites future work in several promising directions.
+First, a more systematic way of choosing collocation points would greatly benefit the method.
+In Section~\ref{sec:collocations_conditions} we define collocation points and provide criteria for a good choice of a family of colocation points.
+However, we ultimately leave the reader without a constructive algorithm for identifying candidate families of collocation points suitable for their problem.
+At the same time, certain differential equations admit families of basis functions that can span their solutions.
 Such basis functions frequently poses easily-computable derivatives, which makes them perfect candidates for collocation points for PINODE.
@@
 Finally, one could utilize PINODE models not only in compressive sensing applications but in any application where one needs to quickly search over a large space of possible simulations.
-These examples include applications such as medical imaging, model discovery, and online control.
+These examples include medical imaging, model discovery, online control, fast simulations, and many others.
+
+\paragraph{Acknowledgements} I would like to thank ... (TBD)

### 2023-05-03 introduction.tex
Incorporate Alanna's edits for my intro

@@
-Machine Learning has emerged as an important tool in applied mathematics problems as data-driven approaches have been increasingly outperforming the first-principle modeling in the tasks of classification, recognition, autonomous control, protein folding, and even playing video games.
-However, due to the absence of prior knowledge  purely data-driven approaches typically require a large amount of data to approximate behavior of first-principle models even in simplest scenarios.
-Thus the most promising modeling approaches of today are those that integrate both data-driven and first-principle components.
-These latter components frequently stem from domain expertise and pre-existing convictions.
+\edit{Add citations all over the intro}
+
+Machine learning has emerged as an important tool in applied mathematics problems as data-driven approaches have begun to outperform  first-principle modeling in the tasks of classification, recognition, autonomous control, protein folding, and even video games.
+However, due to the absence of prior knowledge,  purely data-driven approaches typically require a large amount of data to approximate the behavior of first-principle models even in the simplest scenarios.
+Thus, the most promising modeling approaches of today are those that integrate both data-driven and first-principle components.
+These latter components frequently stem from domain expertise and pre-existing beliefs.
 Incorporating prior knowledge into a data-driven method is a formidable task that demands a considerable amount of ingenuity.
@@
 Feature selection problems typically arise when one believes that a small subset of features within a dataset captures the majority of the variance of the quantity of interest; this subset of features is commonly known as sparse support.
-The canonical feature selection methods typically provide control over the degree of sparsity for the solution and outputs a candidate support for it.
+The canonical feature selection methods typically provide control over the  sparsity of the solution and output a candidate support for it.
 In the absence of additional information about the support apart from its probable sparsity, these methods furnish a satisfactory toolkit for practitioners.
 
- Real-world practitioners, however, poses expert knowledge about the covariates in their data and wish to incorporate that knowledge as a set of priors and constraints.
+ Real-world practitioners, however, pose expert knowledge about the covariates in their data and wish to incorporate that knowledge as a set of priors and constraints.
 For instance, a practitioner may desire a model that employs at most 6 out of the 40 available features, with features A and B being mandatory components.
@@
 Attempts to directly address such constraints often result in the formulation of mathematically intractable problems.
-For example, the first constraint above makes the underlying optimization problem N-P complete.
 Additionally, practitioners require the ability to swap these priors as new hypotheses and knowledge come to light.
-This necessitates the implementation of an optimization routine that does not rely upon these priors, otherwise it would force the practitioner to re-implement the algorithm after every such swap, rendering the instrument impractical.
-Chapter~\ref{ch:msr3} elucidates my efforts in developing a universal feature selection approach for linear mixed-effects models that simultaneously confronts the numerical challenges that accompany it.
+This necessitates the implementation of an optimization routine that does not rely upon these priors; otherwise it would force the practitioner to re-implement the algorithm after every such swap, rendering the method impractical.
+Chapter~\ref{ch:msr3} elucidates my efforts in developing a universal feature selection approach for linear mixed-effects models that simultaneously offers flexibility for incorporating a variety of priors and confronts the numerical challenges that accompany such flexibility.
 
 The second example serves to demonstrate how one can incorporate a degree of partial knowledge of physics into a deep neural network.
-The methods for data-driven modeling of physical phenomena surged in popularity, in large part due to the advancements in GPU computations and automatic differentiation tools.
-These methods have proven particularly useful in the tasks of reduced-order modeling.
-Using data a neural network can find a representation of a system on a reduced-order manifold via a non-linear transformation from an observable space to that manifold.
-Classic techniques, such as Galerkin POD or DMD, are fundamentally incapable of finding such spaces because these methods are confined to linear, albeit optimal, transformations.
-However, as the no-free-lunch theorem implies, the exceptional approximation capabilities of deep networks come at the cost of increased instability.
+Methods for data-driven modeling of physical phenomena have been increasingly popular in large part due to the advancements in GPU computations and automatic differentiation tools.
+These methods have proven particularly useful in reduced-order modeling tasks.
+Using data, a neural network can find a representation of a system on a reduced-order manifold via a non-linear transformation from an observable space to that manifold.
+Classic techniques, such as Galerkin Proper Orthogonal Decomposition (POD) or Dynamic Mode Decomposition (DMD), are fundamentally incapable of finding such spaces because these methods are confined to linear, albeit optimal, transformations.
+However, the exceptional approximation capabilities of deep networks come at the cost of increased instability.
 In particular, the networks tend to identify manifolds that overfit the data and, consequently, fail at extrapolation tasks.
@@
 Recent works have shown that one can improve a neural-network based model by incorporating partial knowledge of physics.
-For example, a model can find a better reduced-order manifold if it knows that the it predicts the behavior of a shock wave that obeys Burger's equation.
-Chapter~\ref{ch:pinode} is devoted to the development of PINODE ROMs -- a framework for building Reduced-Order Models with Physics-Informed Neural Ordinary Differential Equations.
-It illustrates how one can transfer knowledge from a known equation to a model that predicts solutions of said equation using collocation points technique.
-The resulting ROMs are able to extrapolate forward in time considerably more effectively, as well as perform better for unseen initial conditions and exhibit less sensitivity to noise.
-Finally, in Chapter~\ref{ch:cs} I demonstrate the utility of PINODE models in compressive sensing applications, where they effectively "stitch" together separate timeframes and significantly exceed the Nyquist sampling limit.
+For example, a model can find a better reduced-order manifold if it knows that it predicts the behavior of a shock wave that obeys Burger's equation.
+Chapter~\ref{ch:pinode} is devoted to the development of a framework for training Reduced-Order Models (ROMs) with Physics-Informed Neural Ordinary Differential Equations (PINODE).
+It illustrates how one can transfer knowledge from a known equation to a model that predicts solutions of said equation using collocation points.
+The resulting ROMs are able to extrapolate forward in time considerably more effectively,  perform better for unseen initial conditions, and exhibit less sensitivity to noise.
+Finally, in Chapter~\ref{ch:cs} I demonstrate the utility of PINODE models in compressive sensing applications, where they effectively "stitch" together separate timeframes and go below the Nyquist sampling limit.
 
@@
 Steven Brunton at the University of Washington.
-Many parts of the works listed above were taken verbatim to this document per an explicit permission of my co-authors.
+Many parts of the works listed above were taken verbatim to this document per the explicit permission of my co-authors.

### 2023-05-03 phd_thesis.tex
Incorporate all Sasha and Jim's edits from the rebuttal
Add a snipet from pysr3 paper to software implementation chapter
Add conclusions
Write a first version of the abstract

@@
 \\[5mm]
-\department
+Applied Mathematics
 \\[5mm]
@@
 
-My abstract goes here. I will probably write it at the very end.
+\paragraph{Abstract} This work summarizes two projects focused on incorporating prior knowledge into machine learning models.
+In the first project we developed universal feature selection methods for linear mixed-effect models.
+Namely, we extended Sparse Relaxed Regularized Regression (SR3) to Linear Mixed-Effect (LME) likelihood and showed how one can minimize such likelihood with a proximal gradient descent.
+We also develop theoretical underpinnings of the proposed extension, including consistency results, variational properties, implementability of optimization methods, and convergence results.
+In particular we provide convergence analyses for a basic implementation of SR3 for LME (called MSR3) and an accelerated hybrid algorithm (called MSR3-fast).
+Numerical results show the utility and speed of these algorithms on realistic simulated datasets.
+Finally, we provide an open-source implementation of both algorithms in an open source python package \texttt{pysr3}.
+This package offers complete compatibility with \texttt{scikit-learn}, so all \texttt{pysr3} models can be used in pipeline with classic modelling blocks such as data pre-processors, randomized grid search, cross-validation, and quality metrics.
+
+The second line of work develops a framework for training Reduced-Order Models (ROMs) with Physics-Informed Neural Ordinary Differential Equations (PINODE).
+Our innovation builds on ideas from classical collocation methods of numerical analysis and illustrates how one can use collocation points to transfer knowledge from a known equation to a model that approximates solutions of that equation.
+We show that the addition of our physics-informed loss allows for exceptional data supply strategies that improves the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible.
+The resulting ROMs are able to extrapolate forward in time considerably more effectively,  perform better for unseen initial conditions, and exhibit less sensitivity to noise.
+Finally, we show how such ROMs can be used as strong regularizers in single-pixel imaging (SPI) allowing to reduce data-intake by an order of magnitude relatively to current state-of-the-art algorithms.
 
@@
 \chapter{Conclusion}
+\include{conclusion}
 

### 2023-05-04 msr3-theory.tex
Add PIKN chapter; Add todos across the doc
Add sensitivity+scalability analyses and covid example
Address multiple TODO items in MSR3 chapter

@@
-%gives the problem structure used in the design of our
-%numerical solution method. 
-%Specifically, we design a 
-%(PGD) %prox gradient 
-%method 
-%for solving this problem (see Section \ref{sec:pgd}). For this we must show that the function $u_\emu$
-%is continuously differentiable with Lipschitz continuous gradient.
-
-%\section{Prox Gradient Descent}
-\subsection{Proximal Gradient Descent for \texorpdfstring{$\uemu\tbg + R\tbg+\del_{\Rqp}(\tgam)$}{}} 
+In this section we provide a 
+theoretical justification for the algorithmic approach to the solution
+of the marginalized maximum likelihood estimation problems presented
+in~\cite{aravkin2022jimtheory} as well as in Section~\ref{sec:MSR3}. The approach is motivated by the sparse relaxed regularization regression (SR3) strategy
+developed in \cite{Zheng2019SR3}. While the analysis of~\cite{Zheng2019SR3} relies on the least squares data-fitting term,
+here we develop the algorithmic design and analysis required for the nonlinear and nonconvex LME extension.
+
+This section proceeds as follows. 
+% Section 3
+We study the relaxation strategy defined in~\eqref{eq:relax} where the the relaxation depends on a decoupling
+parameter $\eta$ and log-barrier smoothing parameter $\mu$. In particular, we introduce
+the optimal value function $\uemu$~\eqref{eq:value_function_definition} 
+used to exploit the decoupling of the likelihood
+function from the sparsity regularizer. Here $\uemu$ is obtained by partially minimizing
+the decoupled variables in the likelihood while keeping those in the regularizer fixed. 
+% Section 4
+Section~\ref{sec:pgd} introduces the MSR3 algorithm
+as the PGD algorithm 
+applied to the sum of $\uemu$ and the sparsity regularizer. 
+A brief discussion the basic assumptions typically required
+for establishing the viability of the PGD algorithm for this formulation is given.
+% Section 5
+Section~\ref{sec:smoothness} is the theoretical core of the work. In this section
+we show that the optimal value function function $\uemu$ satisfies the properties
+necessary for the application of the PGD algorithm. 
+In particular, we establish the Lipschitz continuity of $\nabla \uemu$ 
+(Lemma \ref{lem:grad u lip}). 
+% Section 6
+The convergence results for MSR3 are presented in Section~\ref{sec:convergence pgd}
+for fixed values of $\eta$ and $\mu$.
+% Section 7
+In Section~\ref{sec:hybrid alg} we address the key issues surrounding the initialization
+of the coupling and smoothing parameters $\eta$ and $\mu$ when only approximate
+values for $\uemu$ and $\nabla \uemu$ are known. Here we appeal to both
+variable metric ideas as well as properties of the interior point algorithm.
+
+\subsection{Proximal Gradient Descent for the Regularized Value Function} 
 \label{sec:pgd}
@@
 
-\begin{algorithm}[H]
+\begin{algorithm}[ht]
 \SetAlgoLined
@@
 
-%\begin{figure}[t]
+%\begin{figure}[ht]
 %    \centering
@@
 
-\subsection{The Smoothness of \texorpdfstring{$u_\emu$}{}}\label{sec:smoothness}
+%\subsection{The Smoothness of \texorpdfstring{$u_\emu$}{}}
+\subsection{The Smoothness of the Value Function}
+\label{sec:smoothness}
 We investigate the relationship between the problems \eqref{eq:lme_loss_original_in_x}
@@
 
-\subsubsection{The Lipschitz Continuity of \texorpdfstring{$\nabla \uemu\tbg$}{}}
+%\subsubsection{The Lipschitz Continuity of \texorpdfstring{$\nabla \uemu\tbg$}{}}
+\subsubsection{The Lipschitz Continuity of the Gradient of the Value Function}
 
@@
 
-\subsection{Convergence of the PGD Algorithm for \texorpdfstring{$\Phi_\emu$}{}}
+\subsection{Convergence of the PGD Algorithm for Regularized MSR3 Likelihood}
 \label{sec:convergence pgd}

### 2023-05-04 msr3.tex
Address multiple TODO items in MSR3 chapter

@@
-\edit{Rewrite the MSR3 abstract into a summary of whole work with citations}
-Linear Mixed-Effects (LME) models are a fundamental tool for modeling correlated data, 
+% \edit{Rewrite the MSR3 abstract into a summary of whole work with citations}
+\paragraph{Abstract for Chapter~\ref{ch:msr3}} Linear Mixed-Effects (LME) models are a fundamental tool for modeling correlated data, 
 including cohort studies, longitudinal data analysis, and meta-analysis.
@@
 the underlying algorithmic parameters are set and test the algorithm on both 
-simulated problems and a problem with real data. The paper is concluded in
-Section \ref{sec:discussion} with a brief discussion of the contributions.
-
-\edit{Add a paragraph summarizing the theory paper}
-\edit{Add a paragraph summarizing software paper}
+simulated problems and a problem with real data.
+In Section~\ref{sec:theory-paper} we develop theoretical underpinnings of the proposed extension, including consistency results, variational properties, implementability of optimization methods, and convergence results for both MSR3 and MSR3-fast.
+Section~\ref{sec:software-paper} outlines the software implementation of MSR3 and MSR3-fast in \texttt{pysr3} package.
+The chapter is concluded in Section \ref{sec:discussion} with a brief discussion of the contributions.
 
@@
 
+The individual random effects $u_i$ can be found through the minimization of conditional likelihood of random effects given $\beta$ and $\Gamma$:
+
+\eq{
+\min_{u_i}\LL(u_i|\beta, \Gamma) = \min_{u_i}\left(\half u_i^T\Gamma^{-1}u_i + \half (Z_iu_i - Y_i + F_i(\beta))^T\Lambda_i^{-1}(Z_iu_i - Y_i + F_i(\beta))\right),
+}
+
+This problem has a closed form solution commonly known as Best Linear Unbiased Predictor, or BLUP. It was first derived by \cite{Harville1976} as an extension to Gauss-Markov theorem.
+
+ \eq{
+	\label{eq:random_effects_formula}
+	u_i = (\Gamma^{-1} + Z_i^T\Lambda_i^{-1}Z_i)^{-1}Z_i^T\Lambda_i^{-1}(Y_i - X_i\beta).	
+}
+
 By using the simplification $\Gam=\Diag{\gam}$, we obtain the problem
@@
 Finally, \cite{Ghosh2018} studied SCAD regularization for selecting mean effects in high-dimensional genomics problems.
+One can refer to \cite{Muller2013, Buscemi2019Survey} for a detailed overview of different feature selection approaches.
 
@@
 BIC performs well in practice, but does not have theoretical guarantees~(\cite{schelldorfer2011estimation}).
-  %One can refer to \cite{Muller2013, Buscemi2019Survey} for a detailed overview of different feature selection approaches.
-
-  \edit{Add more ICs, e.g. Muller}
+Finally, \cite{Hui2017} suggests penalizing the number of non-zero random effects in the resulting model instead of penalizing the number of random features with non-zero variance, especially when random effects are optimized directly instead of using BLUP~\eqref{eq:random_effects_formula}.
 
@@
 variational information about the function $\LL$ over its domain.
-We show that $\uemu$ has a locally Lipschitz continuous gradient
+In Section~\ref{sec:theory-paper} we show that $\uemu$ has a locally Lipschitz continuous gradient
 and that the evaluation of $\uemu$ and $\nabla \uemu$ is accomplished
@@
 allows us to apply the PGD algorithm to the function $\uemu$ rather than
-the function $\LL$. Our numerical studies show that the global information 
+the function $\LL$. Our numerical studies in Section~\ref{sec:applications} show that the global information 
 captured by $\uemu$ significantly improves both the accuracy of the solution
 obtained and the overall numerical efficiency of the algorithm.
-
-\edit{Connect this paragraph to the theory block}
-
-To motivate the reader, the theoretical results that support the design of Algorithm \ref{alg:MSR3} go as follows.
-The existence of solutions for the problem~(\ref{eq:value_function_definition}) for any positive $\mu$ is shown in \cite[Theorem 5]{Theory1}, and the convergence of solutions to the MSR3 solution as $\mu\downarrow 0$ is shown in~\cite[Theorem 7]{Theory1}.
-Finally, \cite[Theorem 6]{Theory1} shows that the MSR3 relaxation is consistent with respect to the barrier, so that as the MSR3 parameter $\eta \uparrow \infty$,
- limit points of global solutions to the former are global solutions to the latter. 
- However, in the applications considered here,
- the empirical studies in Sections \ref{sec:synthetic} and \ref{sec:real}
- indicate that one does not need to make $\eta$ particularly large in order to accurately identify the correct sparsity pattern.
 
@@
 
-\verify{Ensure that short and long performance tables are consistent with each other}
 \begin{table}
@@
 
-\vskip -.8in
 \begin{figure}
@@
 \section{Theoretical Analysis}
+\label{sec:theory-paper}
 \input{msr3-theory}
@@
 \section{Software Implementation}
-\edit{Move pysr3 software paper here}
-To ensure reproducibility of this research, all new algorithms have been implemented as a part of the \texttt{pysr3}\footnote{Available at \href{https://github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}} library~\cite{sholokhov2023pysr3}.
+\label{sec:software-paper}
+To ensure reproducibility of this research, all new algorithms have been implemented as a part of the \texttt{pysr3}\footnote{Available at \href{https://github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}} library~(\cite{sholokhov2023pysr3}).
 This library implements functionality for fitting linear mixed models and selecting covariates.
@@
 Each regularizer included in PySR3 can also be used in its relaxed SR3 form.
-More information about the structure of the library can be found in documentation\footnote{https://aksholokhov.github.io/pysr3/},
+More information about the structure of the library can be found in the documentation\footnote{https://aksholokhov.github.io/pysr3/} as well as in the package's companion paper~(\cite{sholokhov2023pysr3}).
 

### 2023-05-06 cs.tex
Add appendices; connect MSR3 theory and practice definitions; compiles with NO WARNINGS
Add draft of the intro
Write CS paper to date

@@
 \section{Application to Compressive Sensing}
+\label{ch:cs}
 \subsection{Introduction}
-\paragraph{Prior Work}
-\subsection{Method}
-\paragraph{Compressive Sensing} In many applications we can not observe $x(t)$ in real time.
-Instead, we observe $p$ detectors, and each detector provides a linear combination of a small number of coordinates of $x(t)$ at a time:
+% SPI definition and main stages
+Single-pixel imaging (SPI) is a relatively recent imaging technique that replaces the classic two dimensional sensor array with a single photodetector~\cite{bian2018experimental}.
+An SPI camera modulates the incoming light from a scene using a light modulator, such as a diffuser~\cite{guo2016multilayer} or a digital micro-mirror device (DMD)~\cite{pittman1995optical, sampsell1993overview}, and focuses the resulting correlated light patterns onto a photodiode.
+The configuration of the modulator and the recorded intensity of light are then processed to reconstruct an image of the target scene.
+
+% advantages of SPI
+SPI owes its popularity to multiple advantages it achieves over traditional imaging approaches that employ 2D sensor arrays.
+First, since all reflected light is focused onto a single detector, SPI has better sensitivity under dim lighting conditions and enjoys a broader spectral range\cite{Edgar2019}.
+Second, SPI moves the cost of acquisition and processing from the encoding/compression stage to the decoding/reconstruction stage.
+For example, a DMD array orients its mirrors to reflect light either towards the detector (+1) or away from it (0).
+Importantly, these orientations do not depend on the scene that is being acquired to ensure proper reconstruction: \cite{baraniuk2008simple}~showed that DMD patterns can be drawn as independent identically distributed (i.i.d.) samples from a uniform Bernoulli distribution.
+In contrast, sample-then-compress approaches\cite{skodras2001jpeg} require an expensive projection of multiple single-pixel snapshots onto a chosen basis, such as Fourier or top-K singular vectors, during their encoding step.
+Doing the heavy-lifting job during decoding is often preferred in applications where sensors have to be small and low-power, whereas decoding happens offline on high-performance computers.
+The simplicity of encoding keeps the computational requirements of SPI sensors (and thus their price) low, inviting their application in multispectral imaging \cite{bian2016multispectral,li2017efficient}, optical encryption~\cite{chen2013ghost}, remote sensing~\cite{zhao2012ghost}, object tracking~\cite{li2014ghost}, and many other fields~\cite{sun20133d,zhang2010correlated,cheng2009ghost}.
+
+% introducing CS
+Compressive Sensing (CS)~\cite{donoho2006compressed, candes2006compressive, baraniuk2014introduction, baraniuk2008simple} emerged in the mid-2000s as an alternative signal acquisition regime to classical Shannon-Nyquist sampling.
+In the classical regime, a band-limited signal is sampled at $N$ specific points in time (and/or space) so that the signal can be reconstructed from the acquired samples using a simple Sinc interpolation.
+Compressive sensing, on the other hand, proposes to acquire the signal using a small number of inner products between the signal and general non-adaptive sampling vectors.
+If the signal to be acquired has a $K$-sparse, or more generally, parsimonious, representation some transform domain, then the signal can be guaranteed to be recovered using as few as $\mathcal{O}(K\log(N/K))$ non-adaptive measurements, which is much smaller than the $N$ samples that are required by sample-then-compress approaches such as JPEG and JPEG-2000~\cite{skodras2001jpeg}.
+The development of compressive sensing theory and efficient reconstruction algorithms has further promoted the deployment of this technique in SPI~\cite{duarte2008single}.
+
+% Compressive Sensing (CS) is an approach to SPI which assumes that a signal is spare in a certain basis.
+Under this assumption a small collection of non-adaptive linear measurements can contain enough information for reconstructing the signal~\cite{duarte2008single, candes2006compressive}.
+Namely, when an original signal made of $N$ pixels has a $K$-sparse representation in a certain basis, the basis coefficients can be recovered using only $\mathcal{O}(K\log(N/K))$ which is much smaller than $N$ required by sample-then-compress approaches such as JPEG and JPEG-2000~\cite{skodras2001jpeg}.
+
+% Recent progress on deep SPI
+Because each CS measurement contains highly-compressed information about the observed frame, one needs to sample a large number of such measurements.
+Thus, many recent works propose approaches to reduce the number of required samples per frame (SPF).
+These approaches can be roughly grouped into two buckets: signal-dependent control of the mirrors and more sophisticated reconstruction techniques.
+The former ensures that each single-pixel measurement contains as much information as possible by focusing the mirrors on the selected regions of the frame~\cite{zhang2017fast,sun2017russian,xu20181000}.
+Although highly effective, such an approach cannot be applied to detectors that do not offer fine-grained control of their mirrors.
+The latter group focuses on creating more efficient optimization algorithms for the decoding step~\cite{katz2009compressive}.
+It includes non-iterative algorithms such as differential ghost imaging (DGI) \cite{gong2010method}, as well as iterative methods such as total variation (TV)~\cite{suo2016signal} regularization.
+Notably, recent approaches adopted deep learning (DL) for SPI reconstruction.
+These approaches showed that using deep neural networks can dramatically reduce sampling ratio and offer near-real-time performance \cite{lyu2017deep,higham2018deep,wang2019learning,wang2022single}.
+However, most of these methods use multiple samples for single-frame reconstruction and do not yield satisfactory reconstruction of multiple sequential frames when the number of samples per frame is small, even if the total number of samples is large, as we show below.
+
+% Our place among CS literature
+Our work aims to bridge this gap.
+It falls into the second group of decoding-centered approaches which aid CS reconstruction with deep learning.
+In particular, we show how a Reduced-Order Model (ROM) can be used to regularize CS reconstruction of a video of a physical phenomenon, even if all the measurements are taken at different moments in time.
+We achieve that by pre-training a ROM to compress simulations into a latent space and calculate the system's evolution there using neural ordinary differential equations (neural ODEs).
+At the time of reconstruction, we recover a trajectory in the latent space which yields compressing sensing measurements that match what the detector recorded when projected to the observable space.
+Effectively, we solve an ODE-regularized CS reconstruction in an efficient way using the adjoint-sensitivity method to obtain necessary derivatives~\cite{chen2018neuralode}.
+Such ODE-based regularization serves as a strong induced bias and reduces the number of samples per frame required for a successful reconstruction.
+In fact, our method requires fewer samples per frame than the current state-of-the-art methods.
+We introduce the methodology in Section~\ref{sec:cs_methods} and study its performance on two examples: Burger's equation and Kolmogorov flow.
+
+\subsection{Methods}
+\label{sec:cs_methods}
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/cs_abstract.png}
+    \caption{\label{fig:cs_abstract}We propose a compressive-sensing algorithm that reconstructs observations by identifying their trajectories in a latent space of a Reduced-Order Model (ROM).
+The model provides strong inductive bias and reduces data intake requirements for compressive-sensing hardware.}
+\end{figure}
+
+\paragraph{Reduced-Order Model with Non-Linear Latent Dynamics} We assume the same notation as in Section~\ref{sec:method} and define a reduced-order model (ROM) as a tuple $(\psi_\theta, \phi_\theta, h_\theta)$, which consists of a decoder $\psi_\theta(z)$, an encoder $\phi_\theta(z)$, and the latent dynamics $h(z)$, respectively, all parametrized by trainable weights $\theta$.
+We also assume that it is technically possible\footnote{For example, if $h(z)$ is a neural network, then one can evaluate such an integral using adjoint-sensitivity-based techniques, e.g.
+\texttt{torchdiffeq} package by \cite{chen2018neuralode}.} to evaluate a differentiable integral over the latent dynamics $h(z)$ for a given period of time.
+Figure~\ref{fig:cs_abstract} illustrates the architecture of the ROM.
+Since this section assumes that such a ROM comes pre-trained, we consider the weights $\theta$ to be frozen and omit them from the notation.
+
+\paragraph{Compressive Sensing} In many applications, we can not observe $x(t)$ in real time.
+Instead, as illustrated on Figure~\ref{fig:cs_abstract} (Step 2), we observe $p$ scalar values per timeframe, with each scalar a linear combination of a small number of coordinates of $x(t)$ at a time:
+
 \begin{equation}
     \label{eq:cs_definition}
-    y(t) = A_t x(t), \quad A_t \in \mathbb{R}^{p\times n}
+    \tilde{y}(t) = A_t x(t), \quad A_t \in \mathbb{R}^{p\times n}
 \end{equation}
 
-In particular, we consider a single pixel camera setup where for every time instance $t$, $p$ acquisitions $y(t)$ are obtained by a high sampling rate photo-detector using the projection matrix $A_t$.
-The rows of the matrix $A_t$ correspond to a binary mask pattern that can be encoded using a digital micro-mirror device (DMD) where the incoming light from $x(t)$ is reflected from the DMD array and focused onto the photo-detector.
-Figure~\ref{fig:SPI} illustrates an example of the single pixel imaging setup where a gas plume is imaged using a DMD array and a medium infra-red (MIR) photo-detector.
+When recorded over a discretized period of time $[t_1, \dots, t_T]$, such a set of observations is stored as a vector, $y \in \mathcal{R}^{T \times p} = [y(t_1), \dots, y(t_T)]^T$.
+
 \begin{figure}
 \centering
-\includegraphics[width = 0.5\textwidth]{figures/SPI_setup.pdf}\label{fig:SPI}
+\includegraphics[width = 0.5\textwidth]{figures/SPI_setup.pdf}
+\caption{SPI setup with DMD \label{fig:SPI}}
 \end{figure}
 
-However, it is often possible to have access to a complete state $x(t)$ at the moment of model training.
+In particular, we consider a single pixel camera setup in which, for every time instance $t$, $p$ acquisitions $y(t)$ are obtained by a high sampling rate photo-detector using the projection matrix $A_t$.
+The rows of the matrix $A_t$ correspond to a binary mask pattern that can be encoded using a digital micro-mirror device (DMD, \cite{pittman1995optical, sampsell1993overview}).
+Similarly to the setup from~\cite{duarte2008single}, the mirrors of our (idealized) DMD array represent pixels of the future reconstructed image, and the slopes of the mirrors represent the weights.
+The light from a scene lands on a DMD, which slopes each of its mirrors to reflect light either towards the detector (+1) or away from it (-1) according to a uniform Bernoulli variable.
+Next, the (+1) pixels go through a double-convex lens that focuses them on a signal photon detector.
+Finally, the detector integrates the signal and provides the final observation as its output voltage, which is later digitized by an A/D converter.
+The slopes of each mirror are also recorded in the detector's memory.
+Figure~\ref{fig:SPI} illustrates an example of the single pixel imaging setup where a gas plume is imaged using a DMD array and a medium infrared (MIR) photo-detector.
+
+A modern DMD supports snapshot frequencies of around 10~kHz (\cite{wang2022single}).
+When the dynamics of the recorded scene are much slower, one can assume that such a device can take multiple measurements simultaneously\footnote{When the dynamics change rapidly, one would have to use an array of DMD devices to achieve more than one sample per frame}.
+The number of samples recorded per  frame is referred to as \textit{samples per frame}, or SPF.
+It is often divided by the total number of mirrors of the DMD device and multiplied by 100\%, in which case it is called a \textit{sample per frame rate} $\beta$.
+For a compressive-sensing-based approach to be practically useful it should require a relatively small $\beta$ (e.g.
+2-10\%) for a successful reconstruction  of the scene.
+
+\paragraph{Reconstruction Loss} It is often possible to have access to the complete state $x(t)$ at the moment of model training.
 Thus, one can develop a ROM $(\psi_\theta, \phi_\theta, h_\theta)$ using the full state $x(t)$, and then utilize this ROM for real-time compressive sensing applications.
-
-\paragraph{Training Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses.
-The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data.
-Formally, for a given set of trajectories $\bd{x}_i$, $i \in [1 \dots k]$, where each trajectory $\bd{x}_i \in \mathbb{R}^{n \times p}$ is a set of $p$ snapshots that correspond to the recorded states of the system for $p$ time-steps, $t_j$, $j \in [1, \dots, p]$, the loss function $\LL^{data}_\theta$ is defined as:
-
-\begin{align}
-    \label{eq:loss_training}
-    \LL^{training}(\theta) & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
-     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
-\end{align}
-where $\sigma$ is the standard deviation of the observation noise.
-
-To obtain a ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$, we minimize the loss above:
-
-\begin{equation}
-    \label{eq:training_minimization}
-    \theta^* = \arg\min_{\theta} \LL^{training}(\theta)
-\end{equation}
-
-We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}).
-This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}.
-To simplify the notation without loss of generality, in the rest of the paper we assume that all trajectories were recorded over the same time-frame with an equal and uniform step-size.
-
-\paragraph{Reconstruction Loss} 
-We use the ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ above to forecast the dynamics based on partial observations in real time.
-Namely, instead of reconstructing $x(t)$ based on compressive-sensing observations $y(t)$ directly, we first reconstruct the latent dynamics $z(t)$ and then project it to the observable space using the decoder $\psi_{\theta^*}(z)$.
+In particular, we use the ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ from Figure~\ref{fig:cs_abstract}~(Frame 1) to forecast the dynamics based on partial observations in real time.
+Namely, instead of reconstructing $x(t)$ based on compressive-sensing observations $y(t)$ directly, we first reconstruct the latent dynamics $z(t)$ and then project it to the observable space using the decoder $\psi_{\theta^*}(z)$, as illustrated in Figure~\ref{fig:cs_abstract}, bottom-right pane.
 
@@
 where the parameter $\lambda$ controls the degree on which the compressing sensing algorithm relies on the latent dynamics $h_{\theta^*}$ during the signal reconstruction phase.
-We minimize the loss~\ref{eq:reconstruction_loss} using a gradient-based technique, with the gradients obtained using automatic differentiation frameworks.
+We minimize the loss~\eqref{eq:reconstruction_loss} using a gradient-based technique, with the gradients obtained using automatic differentiation frameworks.
 
 \subsection{Experiments}
-\subsection{Discussion and Conclusion}
+
+\subsubsection{Burger's Equation}
+We study the performance of our framework on Burger's equation with $[-\pi, \pi]$-periodic boundary conditions:
+
+\begin{equation}
+\begin{split}
+    & u_t  + uu_x = \nu u_{xx} \\
+    & u(-\pi, t) = u(\pi, t),\quad \forall t \in [0, T]
+\end{split}
+\end{equation}
+where $u_t$, $u_x$, and $u_{xx}$ denote the first partial derivative in time, and the first and second spatial derivatives, respectively.
+
+\paragraph{ROM Training} To obtain training data for the ROM, we replicate the experimental setup from Section~\ref{sec:compressibility}.
+Namely, we generate 1024 trajectories on a discretized spacial domain $[-\pi,\,\pi]$ with 128 grid points.
+To generate a diverse set of initial conditions, we sum the first 10 harmonic terms with random coefficients:
+\begin{equation}
+    \label{eq:burger_initial_condition_sine}
+    u(x, 0) = \frac{1}{10}\sum_{k = 1}^{10} a_k\cos(kx) + b_k\sin((k+1)x), \quad a_k, b_k \sim \mathcal{N}(0, 1)
+\end{equation}
+ We solve Equation~\ref{eq:burgers_equation} for $t \in [0, 2]$ with $\Delta t = 0.1$ using a spectral solver by~\cite{trefethen2000spectral}.
+
+\paragraph{Compressive Sensing}
+For the compressive sensing phase, we generate 128 trajectories with ``bump'' initial conditions -- a smooth approximation of a bump with two opposing steeply-curved sigmoids:
+
+\begin{equation}
+    \label{eq:burger_initial_condition_bumps}
+    u(x, 0) = \frac{1}{1 + exp(-k(x-a))} - \frac{1}{1 + exp(-k(x-b))}
+\end{equation}
+
+where $a < b$ are sampled uniformly in $[-\pi, \pi]$ and $k = 20$.
+We choose this shape to ensure that the training and sensing trajectories are sufficiently different.
+We choose compressive sensing matrices $A_t$ to be binary (0-1) matrices with each row having 64 non-zero components out of 128 sampled uniformly.
+
+We compare our approach (PINODE) with three alternatives: Digital Ghost Imaging by~\cite{ferri2010differential} and~\cite{gong2010method}, Total Variation Regularization  (TVR) by~\cite{bian2018experimental}, and an approach by~\cite{bora2017compressed} which uses an autoencoder (AE).
+We measure reconstruction accuracy with three commonly-used metrics: the residual mean-squared error (RMSE), the peak signal-to-noise ratio (PSNR), and the structural similarity index measure (SSIM).
+We evaluate both metrics for Burger's trajectories as 2D-images in spatial and in temporal domains.
+
+For two 2D images $a$ and $b$, PSNR is defined as the log-ratio of the maximal value from the true image:
+
+\[
+	\text{PSNR}(a, b) = 20\log\left(\frac{\max_{i,j}(|a[i,j]|)}{\|a - b\|_2}\right)
+\]
+
+SSIM is defined as a product of relative luminance $l(a,b)$, contrast $c(a,b)$, and structure $s(x,y)$:
+
+\[
+	\text{SSIM}(a, b) = l(a,b) \cdot c(a,b) \cdot s(a,b) = \left(\frac{2\mu_a\mu_b + c_1}{\mu_a^2 + \mu_b^2 + c_1}\right) \cdot \left(\frac{2\sigma_a\sigma_b + c_2}{\sigma_a^2 + \sigma_b^2 + c_2} \right) \cdot \left(\frac{\sigma_{ab} + c_2/2}{\sigma_a \sigma_b + c_2/2}\right)
+\]
+Finally, RMSE is defined as the mean residual squares for every state, averaged over the temporal axis:
+
+\[
+	\text{RMSE}(a,b) = \|a - b\|_F
+\]
+
+where, for a picture $x$, $\mu_x$ is the pixel sample mean, $\sigma_x$ is the standard deviation,  and $c_i$ are the constants which stabilize the division when the denominator approaches 0.
+Typically, $c_i$ are set to be proportional to the square of the dynamic range of the pixel values, e.g.
+$c_i = (k_i*2^{\text{bits per pixel}} - 1)^2$, with $k_1 = 0.01$ and $k_2 = 0.03$ being default choices.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/cs_burgers_psnr.png}
+    \caption{\label{fig:cs_burgers_spf} Median PSNR, RMSE, and SSIM with 25-27th percentile intervals over different number of samples per frame taken.
+We see that PINODE achieves 70\% higher peak signal-to-noise-ratio (PSNR), and 4 times higher structural similarity index measure (SSIM) when SPF is low, than its nearest competitor (AE).}
+
+\end{figure}
+
+\begin{figure}
+    \centering
+    	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_64.pdf}
+    \caption{\label{fig:cs_burgers_example_large_spf} Example reconstruction with 64 samples per frame (SPF). Unsurprisingly, all handle the job well since SPF is equal to the number of pixels.}
+\end{figure}
+
+\begin{figure}
+    \centering
+    	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_4.pdf}
+    \caption{\label{fig:cs_burgers_example_small_spf}Example reconstruction with 4 samples per frame (SPF).
+All algorithms except our proposed method fail to recover the image due to the insufficient SPF rate (~6\%).
+In contrast, our proposed method achieves nearly the same reconstruction quality as in Figure~\ref{fig:cs_burgers_example_large_spf} with SPF rate of 100\%.}
+\end{figure}
+
+\paragraph{Results} The results are presented in Figure~\ref{fig:cs_burgers_spf}.
+We plot median PSNR, RMSE, and SSIM over different number of samples per frame taken.
+We see that PINODE achieves 70\% higher peak signal-to-noise-ratio (PSNR) and 4 times higher structural similarity index measure (SSIM) when SPF is low relative to its nearest competitor (AE).
+We also see that the DGI and TVR approaches do not reconstruct images faithfully until SPF approaches the number of pixels in a snapshot; Figure~\ref{fig:cs_burgers_example_large_spf} provides an example of high-SPF reconstruction.
+In contrast, PINODE is able to reconstruct most trajectories with high accuracy using a very small number of samples per frame e.g.
+4, as visualized on Figure~\ref{fig:cs_burgers_example_small_spf}.
+This happens because the reduced-order model of PINODE serves as a strong prior: the model reconstructs the trajectory as a whole instead of reconstructing every snapshot separately, as all other algorithms do in this comparison.
+It allows PINODE to borrow strength across snapshots, even when SPF per each snapshot is small.
+
+\subsubsection{Kolmogorov Flow}
+We study the performance of our framework on Kolmogorov-like flow first-described by~\cite{dovzhenko1981generation} and later analyzed by \cite{tithof2017bifurcations,suri2017forecasting}.
+The following equation describes the behavior of a 2D velocity field $\bd{u}(x, y, t)$:
+
+\begin{align}
+\label{eq:kolmogorov_flow_definition}
+	& \partial_t \bd{u} + \bd{u} \cdot \nabla \bd{u} = -\nabla p + \nu\nabla^2\bd{u} + f \\
+	& \nabla \cdot \bd{u} = 0
+\end{align}
+
+where $p$ is a 2D pressure field, $f = \alpha\sin(ky)x$ represents the driving force with amplitude $A$ and wavenumber $k$, and $\nu = 1/\text{Re}$ is the non-dimensional viscosity equal to inverse of the Reynolds number.
+In all our experiments, $k = 4$ and $\text{Re} = 40$; we chose this setup to match the one from~\cite{wan2018data}, who identified that this combination of hyper-parameters leads to the occurrence of extreme instability events, making the prediction of such flow truly challenging.
+
+To train a ROM which predicts $u(x, y, t)$ we generate 8192 data trajectories as solutions of~\eqref{eq:kolmogorov_flow_definition} using a spectral solver\footnote{\href{https://github.com/zhong1wan/data-assisted/blob/master/Kolmogorov/kol2d_odd.py}{https://github.com/zhong1wan/data-assisted/blob/master/Kolmogorov/kol2d\_odd.py}} from~\cite{wan2018data}.
+Each solution was simulated from t=0 to t=100 to get past the transient regime and then recorded from t=100 to t=110, with the time-step $dt=0.5$.
+As in Section~\ref{sec:method}, the ROM consists of an encoder $\phi(u)$, a decoder $\psi(z)$, and the latent dynamics $h(z)$.
+It was trained by minimizing the data-driven loss~\eqref{eq:loss_data_driven} until the prediction RMSE on a holdout set stopped improving.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{figures/kolmogorov_example.pdf}
+	\caption{\label{fig:kolmogorov_example} An example of a compressive-sensing reconstruction of Kolmogorov flow using a ROM-regularized method.
+The rows represent three different moments in time.
+The first column represents the true flow.
+The second column represents a prediction of the ROM if the initial state of the system were to be known.
+The third column displays the results of CS reconstruction of our algorithm with 512 samples per frame.}
+\end{figure}
+
+A sample result is displayed on Figure~\ref{fig:kolmogorov_example}.
+The rows represent three different moments in time.
+The first column represents the true flow.
+The second column represents the prediction of the ROM if the initial state of the system were to be known.
+It is unknown, so the second row serves as a reference for the maximal theoretical performance of a ROM.
+The third column displays the results of CS reconstruction of our algorithm with 512 samples per frame.
+Each frame is $66\times 66=4356$ pixels, which yields a sampling-per-frame rate of 12\%.
+We observe that the reconstruction faithfully recovered the signal.
+Moreover, we see that the CS algorithm is not limited in its performance by the performance of the underlying ROM: the third row indicates that the model is unable to completely recover shifting dynamics, whereas the CS algorithm aids the model's predictions with CS observations and fully reconstructs the signal.
+
+\subsection{Discussion} 
+In this work, we demonstrated how a collocation-point-based technique can improve the performance of an emerging class of continuous-time physics-informed neural-network-based reduced-order models.
+First, we demonstrated that the incorporation of collocation points in training data can ``cover the gaps'' in training trajectories and inform the model about underrepresented basins of attraction.
+Such an approach alleviates the demand for large volumes of data that is common in network-based models.
+Reducing the amount of data required to faithfully reconstruct a signal is extremely valuable in applications where data is scarce and expensive.
+Second, the physics-informed loss may work as a safeguard, providing a noise-free source of underlying dynamics.
+Third, collocation points can stabilize the model's long-term predictions, allowing for accurate forecasting far beyond the training time horizon.
+Finally, together with using neural-ODE-based nonlinear latent dynamics, adding physics-informed loss leads to the discovery of more compact latent space representations that also yield more accurate models.
+Simultaneous stability and compactness is especially important if one aims to use models with compressive sensing and control algorithms.
+With respect to the computational complexity, we note that adding $Tk$ collocation points to the training imposes less of a computational burden than adding $k$ data trajectories, because, unlike data trajectories, collocation points do not require computing integrals forward in time.
+The last section served to illustrate how PINODE models that were developed in Section~\ref{sec:method} can help in applied modeling.
+In particular, we showed how one can employ a PINODE model for compressive-sensing purposes.
+We illustrated its performance on two examples: Burger's equation and Kolmogorov flow.
+In these cases, it is clear that PINODE can serve as a powerful regularizer and allow users to go significantly below the limits of the Nyquist sampling theorem.
+
+One clear limitation of the current work is that the choice of an efficient collocation family is a design decision that a practitioner makes.
+We believe that such decisions can be automated by adopting existing approaches from classic works on numerical approximations of PDEs, which we leave for future research.
+Another automation that prompts future research is deriving efficient ways of sampling collocation points, possibly via applying modern adaptive learning techniques \cite{subramanian2022adaptive}.
+Finally, although Section~\ref{sec:burger_noise} provides some rationale for why one may expect robustness of hybrid models under noise, we believe that a more rigorous analysis is possible, particularly one that provides conditions under which such robustness is guaranteed.

### 2023-05-06 conclusion.tex
Incorporate Alanna typo fixing

@@
-In this work I gave two examples of how partial knowledge can be utilized to improve performance of machine learning models.
-First, in Section~\ref{sec:MSR3} we showed how complex priors can be embedded into a linear mixed-effects model via a suitable proximal operator, and how the resulting algorithm can be implemented using a proximal gradient descent.
-We note that a complex combination of priors often yields ill-conditioned problems, and we propose a MSR3 relaxation to address it.
+In this work, I gave two examples of how partial knowledge can be utilized to improve performance of machine learning models.
+First, in Section~\ref{sec:MSR3}, we showed how complex priors can be embedded into a linear mixed-effects model via a suitable proximal operator, and how the resulting algorithm can be implemented using a proximal gradient descent.
+We note that a complex combination of priors often yields ill-conditioned problems, and we propose MSR3 relaxation to address it.
 We also prove that a solution of such relaxation approaches the solution of the original problem as the relaxation tightens.
 Next, we showed how partial knowledge of physics can be embedded into deep-learning-based reduced-order models.
-Namely, we show how to project known physical equations that govern behavior of the system in an observable space into a chosen latent space.
-In that space one can require that the gradient field should match the projected true dynamics.
-We show that this methodology leads to identification of more stable latent manifolds, and using such manifolds for forecasting leads to more accurate predictions and lesser sensitivity to noise.
+Namely, we show how to project known physical equations that govern the behavior of the system in an observable space into a chosen latent space.
+In that space one can require that the gradient field match the projected true dynamics.
+We show that this methodology leads to identification of more stable latent manifolds, and using such manifolds for forecasting leads to more accurate predictions and less sensitivity to noise.
 We also showed how the reduced-order models described above can be used for compressive sensing.
-We demonstrated that they can accurately reconstruct the dynamics from partial observations, significantly surpassing current state-of-the-art compressive-sensing methods.
-We hope that these developments pave the way to higher-quality machine learning models which effectively marry data and the humanity's hard-won expert knowledge.
+We demonstrated that they can accurately reconstruct the dynamics from partial observations, significantly surpassing state-of-the-art compressive-sensing methods.
+We hope that these developments pave the way for the creation of higher-quality machine learning models which effectively marry data and the humanity's hard-won expert knowledge.
 
-\paragraph{Perspective Research Directions} Due to the limited time and resources we left many promising directions for future exploration.
+\paragraph{Perspective Research Directions} We would like to note many promising directions for future exploration.
 One such direction is extending sparse relaxed regularized regression (SR3) to generalized linear and generalized mixed-effects models.
@@
 For example, Algorithm~\ref{alg:pgd_for_value} had two nested optimization methods: for evaluating the value function and for optimizing over it.
-However we showed that one can ``blend'' both levels of optimization and achieve superior performance results.
+However, we showed that one can ``blend'' both levels of optimization and achieve superior performance results.
 The same methodology of ``blending'' could possibly be applied to more than two nested optimization loops to yield algorithms that, in certain scenarios, work faster than non-nested optimization methods.
@@
 First, a more systematic way of choosing collocation points would greatly benefit the method.
-In Section~\ref{sec:collocations_conditions} we define collocation points and provide criteria for a good choice of a family of colocation points.
+In Section~\ref{sec:collocations_conditions}, we define collocation points and provide criteria for a good choice of a family of colocation points.
 However, we ultimately leave the reader without a constructive algorithm for identifying candidate families of collocation points suitable for their problem.
 At the same time, certain differential equations admit families of basis functions that can span their solutions.
-Such basis functions frequently poses easily-computable derivatives, which makes them perfect candidates for collocation points for PINODE.
-Bridging the gap between this new methodology and those classic results would undoubtedly yield a fruitful line of works.
-Finally, one could utilize PINODE models not only in compressive sensing applications but in any application where one needs to quickly search over a large space of possible simulations.
+Such basis functions frequently have easily-computable derivatives, which makes them perfect candidates for collocation points for PINODE.
+Bridging the gap between this new methodology and those classic results would undoubtedly yield a fruitful line of work.
+Finally, one could utilize PINODE models not only in compressive sensing application, but in any application where one needs to quickly search over a large space of possible simulations.
 These examples include medical imaging, model discovery, online control, fast simulations, and many others.

### 2023-05-06 msr3.tex
Incorporate Alanna typo fixing

@@
 made available in an open source library, \texttt{pysr3}\footnote{https://github.com/aksholokhov/pysr3} that is 
-easy to use and to compare different methods. All experiments in the paper can be reproduced 
+easy to use and to compare different methods. All experiments in this work can be reproduced 
 using \texttt{pysr3} and code in the reproducibility guide\footnote{https://github.com/aksholokhov/msr3-paper}.
@@
 We begin in Section \ref{sec:LMEM} by giving a precise description of the LME
-model and set the notation for the remainder of the paper. 
+model and set the notation for the remainder of the section. 
 This is followed by a brief discussion of prior work on LMEs.
@@
 
-This paper focuses the case where $\Gamma$ is diagonal, 
+Our work focuses the case where $\Gamma$ is diagonal, 
   (often referred to as \textit{the diagonal setup}) 
@@
 
-    In this paper, we developed and implemented a first-order variable selection framework for LMEs that handles convex and nonconvex regularizers. We also showed that 
+    In this work, we developed and implemented a first-order variable selection framework for LMEs that handles convex and nonconvex regularizers. We also showed that 
     the MSR3 relaxation \eqref{eq:value_function_optimization} 

### 2023-05-06 phd_thesis.tex
Address multiple TODO items in MSR3 chapter
Incorporate Alanna typo fixing

--- 2023-04-19 phd_thesis.tex
@@
 
-My abstract goes here. I will probably write it at the very end.
+This work summarizes two projects focused on incorporating prior knowledge into machine learning models.
+In the first project, we developed universal feature selection methods for linear mixed-effect models.
+Namely, we extended Sparse Relaxed Regularized Regression (SR3) to Linear Mixed-Effect (LME) likelihood and showed how one can minimize such likelihood with proximal gradient descent.
+We also developed theoretical underpinnings of the proposed extension, including consistency results, variational properties, implementability of optimization methods, and convergence results.
+In particular, we provided convergence analyses for a basic implementation of SR3 for LME and an accelerated hybrid algorithm.
+Numerical results show the utility and speed of these algorithms on realistic simulated datasets.
+Finally, we provide an open-source implementation of both algorithms in an open-source python package \texttt{pysr3}.
+This package offers complete compatibility with \texttt{scikit-learn}, so all \texttt{pysr3} models can be used in pipeline with classic modelling blocks such as data pre-processors, randomized grid search, cross-validation, and quality metrics.
+
+The second line of work develops a framework for training Reduced-Order Models (ROMs) with Physics-Informed Neural Ordinary Differential Equations (PINODE).
+In particular, we illustrate how one can utilize a classic technique of collocation points to transfer knowledge from a known equation to a model that approximates solutions of that equation.
+We show that the addition of our physics-informed loss allows for exceptional data supply strategies that improve the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible.
+The resulting ROMs extrapolate forward in time considerably more effectively,  perform better for unseen initial conditions, and exhibit less sensitivity to noise.
+Finally, we show how such ROMs can be used as strong regularizers in single-pixel imaging (SPI) enabling the reduction of sampling-per-frame rate by an order of magnitude relative to the state-of-the-art algorithms.
 

### 2023-05-06 pinode.tex
Incorporate Alanna typo fixing

@@
 \edit{Write a big-picture intro to machine learning for physics}
-\edit{Replace mentions of paper with work}
 \subsection{Related Work}
@@
 A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}.
-At the time of writing this paper, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation}.
+At the time of writing this document, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation}.
 Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}
@@
 
-        In this paper, we employ autoencoders to perform nonlinear model reduction along with NODE in the latent space to model complex and nonlinear dynamics.
+        In this work, we employ autoencoders to perform nonlinear model reduction along with NODE in the latent space to model complex and nonlinear dynamics.
 We choose Neural ODEs in the latent space dynamics representation because of their ability to model highly non-linear dynamics, which is especially important when applications limit the size of the latent space dimension.
@@
 This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}.
-To simplify the notation, without loss of generality, in the rest of the paper we assume that all trajectories are recorded over the same time-frame with the same uniform step-size.
+To simplify the notation, without loss of generality, in the rest of the section we assume that all trajectories are recorded over the same time-frame with the same uniform step-size.
 To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs)\cite{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}).
@@
 The coefficients $\omega_i$ are hyper-parameters which need to be tuned using a validation dataset.
-However, in all experiments of this paper we set $\omega_i$ to be either $0$ or $1$, and we balance $\LL^{physics}_\theta$ and $\LL^{data}_\theta$ the choice of samples in a batch of training data.
+However, in all experiments of this section we set $\omega_i$ to be either $0$ or $1$, and we balance $\LL^{physics}_\theta$ and $\LL^{data}_\theta$ the choice of samples in a batch of training data.
 Specifically, we set the number of collocation points per batch $N_{batch}$ to be equal to the number of trajectories per batch $k_{batch}$ times the number of time-steps$T$: $N_{batch} = Tk_{batch}$.
@@
 However, the former provides inadequate performance when the data is dominated by noise, whereas a hybrid model in this regime produces a solution that is visually similar to the one that the Physics-Informed model produces.
-A more rigorous analysis of this phenomenon seems possible but lies outside of the scope of this paper.
-
-\section{Discussion and Conclusions}
-
-In this work, we demonstrated how a collocation point-based technique can improve the performance of an emerging class of continuous-time physics-informed neural-network based reduced-order models.
-First, we demonstrated that the incorporation of collocation points in training data can ``cover the gaps'' in training trajectories and inform the model about underrepresented basins of attraction.
-Such an approach alleviates the demand for large volumes of data that is common in network-based models, which is crucial in applications where data is scarce and expensive.
-Second, the physics-informed loss may work as a safeguard, providing a noise-free source of underlying dynamics.
-Third, collocation points can stabilize the model's long-term predictions, allowing for accurate forecasting far beyond the training time horizon.
-Finally, together with using a NODE-based non-linear latent dynamics, adding physics-informed loss leads to the discovery of more compact latent space representations that also yield more accurate models.
-Simultaneous stability and compactness is especially important if one aims to use models together with compressive sensing and control algorithms.
-With respect to the computational complexity, we note that adding $Tk$ collocation points to the training imposes less of a computational burden than adding $k$ data trajectories because collocation points do not require computing integrals forward in time as in the case of data trajectories.
-
-One clear limitation of the current work is that the choice of an efficient collocation family is a design decision that a practitioner makes.
-The authors believe that such decisions can be automated by adopting existing approaches from classic works on numerical approximations of PDEs, which we leave for future research.
-Another automation that prompts future research is deriving efficient ways of sampling collocation points, possibly via applying modern adaptive learning techniques \cite{subramanian2022adaptive}.
-Finally, although Section~\ref{sec:burger_noise} provides some rationale for why one may expect robustness of Hybrid models under noise, the authors believe that a more rigorous analysis is possible; particularly one that provides conditions under which such robustness is guaranteed.
+A more rigorous analysis of this phenomenon seems possible but lies outside of the scope of this work.

### 2023-05-06 pinode.tex
Remove todos

@@
 \section{Introduction}
-\edit{Write a big-picture intro to machine learning for physics}
 \subsection{Related Work}
-\edit{Split original intro into prior works paragraphs: DL for Ph, ROMs, PIML}
-\edit{Add physics-informed ML overview (PINN, DeepONet, NFO)}
         Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs).
@@
 We demonstrate that such an approach not only reduces the need for large training data-sets and produces highly-accurate and long-term stable models, but also leads to the discovery of more compact latent spaces, which is especially important for applications in compressed sensing and control.
-
-        \edit{Add Koopman example to the intro}
-
-        \edit{Add compressive sensing application to the intro}
 

### 2023-05-09 presentation/phd_defense_pres.tex
Isolated conference and thesis presentations to enable different style of presenting; connect Jim's math macroses and thesis's bibliography; address warnings

--- 2023-04-06 presentation/pysr3.tex
@@
+\documentclass[8pt]{beamer}
+\input{useful_packages}
+\input{math_macros}
+
+% \usetheme[progressbar=frametitle]{metropolis}
+\usepackage{appendixnumberbeamer}
+
+% PRESETS OF PACKAGES AND MATH %
+
+\usepackage{xspace}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{lmodern}
+\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
+\newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
+\title{Physics-Informed Neural ODEs (PINODE)}
+\date{\today}
+\author{Aleksei Sholokhov, Steven Brunton, J. Nathan Kutz, Hassan Mansour, Saleh Nabi }
+%\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
+
+\begin{document}
+
+\maketitle
+
+\begin{frame}{TODOs}
+\end{frame}
+
+\section{Introduction}
+\begin{frame}{Plan of the Defense}
+Show topics and published papers. Mention covid
+
+\end{frame}
+
 \section{Linear Mixed-Effect Models}
@@
 \begin{itemize}
-	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \cite{Bondell2010,Lin2013}
+	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \cite{Krishna2008,Lin2013}
 	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \cite{Vaida2005,Jones2011}
@@
 		\includegraphics[width=\textwidth]{Figures/bullying_data_assessment_selection.pdf}
-		\caption{Fixed and random covariate selection for Bullying dataset from~\cite{GBD}.
+		\caption{Fixed and random covariate selection for Bullying dataset\footnote{Institute for Health Metrics and Evaluation (IHME).
+Bullying Victimization Relative Risk Bundle GBD 2020.
+Seattle, United States of America (USA), 2021.}.
 The model selected 9 covariates, 7 of which were historically significant, and did not select 4 covariates, 1 of which was historically significant.}
@@
 \end{frame}
+
+\section{Reduced-Order Models}
+\subsection{Intro to Reduced-Order Models (ROMs)}
+
+\begin{frame}{Reduced-Order Models}
+
+Schematics, terminology, notation
+
+\end{frame}
+
+\begin{frame}{Variations of ROMs}
+	Linear, nonlinear, citations
+\end{frame}
+
+\begin{frame}{Physics-Informed Loss}
+Definition of Physics-informed ML.
+
+Derivation of PI loss through chain rule.
+
+Problem with evaluating RHS of differential equations.
+
+Interpretation with cross-borrowing strength. 
+\end{frame}
+
+\begin{frame}{Physics-Informed Koopman Networks}
+	Linear latent dynamics
+
+	Paper
+
+	Not a rom so not suitable for downstream tasks. 
+\end{frame}
+
+\begin{frame}{Physics-Informed Neural ODEs (PINODE)}
+	Nonlinearity in latent space
+	- require a lot of data
+	- picky to initialization
+	- good interpolation bad extrapolation	
+\end{frame}
+
+\begin{frame}{PINODE Results on Burgers}
+	Add PI loss => Fixes most of the issues
+	(example with Burgers)
+	Points learned
+	- Inform the model about unseen conditions
+	Paper reference
+\end{frame}
+
+\begin{frame}{PINODE Active}
+	> PIKN: "we use it and it works"
+	> PINODE: "this is how you need to use it right"
+	> PINODE-Active: can we use it right automatically?
+
+	Loop scheme
+\end{frame}
+
+\begin{frame}{PINODE Active Results}
+	Best result (Burgers)
+\end{frame}
+
+\begin{frame}{PINODE-CS}
+	Application of the above 
+	Loss and setup
+\end{frame}
+
+\begin{frame}{PINODE-CS Results}
+	Results on Burgers
+	Maybe results on a harder problem
+\end{frame}
+
+\begin{frame}{References}
+	\textbf{References}: 
+	\bibliographystyle{plain}
+	\bibliography{../bibliography.bib}
+\end{frame}
+
+\end{document}

### 2023-05-09 presentation/useful_packages.tex
Isolated conference and thesis presentations to enable different style of presenting; connect Jim's math macroses and thesis's bibliography; address warnings

@@
 \usepackage{ulem} % For strikethrough text \st{}
-\usepackage[colorinlistoftodos]{todonotes} % useful for leaving todonotes
 \usepackage{textcomp} %Text Com­pan­ion fonts
@@
 
-%%%%%%%%%%%%%%%%%
-% TODO COMMANDS %
-%%%%%%%%%%%%%%%%%
-\usepackage[colorinlistoftodos]{todonotes} % useful for leaving todonotes
-\newcommand{\hightodo}[1]{\todo[inline,backgroundcolor=red!90!yellow!80!white]{\textbf{\userId}: #1}}
-\newcommand{\lowtodo}[1]{\todo[inline,backgroundcolor=blue!30!white]{\textbf{\userId}: #1}}
-\newcommand{\ask}[2]{\todo[inline,backgroundcolor=Orchid!40!white]{
-	\textbf{Ask #1} #2}}
-\newcommand{\critical}[1]{\todo[inline, backgroundcolor=BrickRed!50!white]{\textbf{Critical:} #1}}
-\newcommand{\search}[1]{\todo[inline, backgroundcolor=blue!30!white]{\textbf{Search:} #1}}
-\newcommand{\edit}[1]{\todo[inline, backgroundcolor=CornflowerBlue!30!white]{\textbf{Edit:} #1}}
-\newcommand{\verify}[1]{\todo[inline, backgroundcolor=Apricot!50!white]{\textbf{Verify:} #1}}
-\newcommand{\experiment}[1]{\todo[inline, backgroundcolor=Green!50!white]{\textbf{Verify:} #1}}
-\newcommand{\delegate}[2]{\todo[inline, background=Lavender]{\textbf{For #1:} #2}}
-
-% Tips
-\newtcolorbox{tip}{colframe=white, colback=gray!10!white}
-
-% Mettings
-\newcounter{meetingscounter}
-\setcounter{meetingscounter}{1}
-\newtcolorbox{meeting}[4]{colframe=white, colback=white, colbacktitle=gray!10!white, 
-	coltitle=black, fonttitle=\bfseries, breakable,
-	subtitle style={boxrule=0.4pt, colback=white},
-	title={
-		\begin{tabular}{p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}l}
-		\underline{Meeting \arabic{meetingscounter}}  &  \stepcounter{meetingscounter} \faCalendar\ #2 &  \faClockO\ #3 & \faMapMarker\ #4 & \faUsers\ #1
-		\end{tabular}
-		}
-	}
-
 % Thick strike out

### 2023-05-11 presentation/phd_defense_pres.tex
Create slides for ROM introduction; rename slides according to new schematics

@@
 
-\section{Reduced-Order Models}
-\subsection{Intro to Reduced-Order Models (ROMs)}
+\section{PINODE}
 
-\begin{frame}{Reduced-Order Models}
+\subsection{Intro to Physics-Informed Reduced-Order Models (ROMs)}
 
-Schematics, terminology, notation
+\begin{frame}{Data-Driven Modeling of Physical Systems}
+
+1) People used to model physical systems with first-principle knowledge
+2) Data-Driven modelling of dynamical systems became a big thing  
+3) However, it requires a lot of data 
+4) Incorporating prior knowledge is a big recent trend, so history does a spiral
 
@@
 
-\begin{frame}{Variations of ROMs}
-	Linear, nonlinear, citations
+\begin{frame}{Incorporating Knowledge into Models}
+1) There are multiple ways of incorporating knowledge into system 
+4) The overall umbrella term for it is physics-informed machine learning
+2) Some use the equations that model phenomena
+3) Some take aspects of it, e.g. symmetries and preservation laws, and forces
+A network to respect those
+5) Our work falls into the first category of approaches
+\end{frame}
+
+\begin{frame}{Reduced-Order Models (ROMs)}
+\includegraphics[width=\textwidth]{Figures/roms_1}
+\end{frame}
+
+\begin{frame}{Reduced-Order Models (ROMs)}
+\includegraphics[width=\textwidth]{Figures/roms_2}
+\end{frame}
+
+\begin{frame}{Reduced-Order Models (ROMs)}
+\includegraphics[width=\textwidth]{Figures/roms_3}
+\end{frame}
+
+\begin{frame}{Reduced-Order Models (ROMs)}
+\includegraphics[width=\textwidth]{Figures/roms_4}
 \end{frame}
@@
 \begin{frame}{Physics-Informed Loss}
-Definition of Physics-informed ML.
+1) We introduce physics to this system by adding a term which regularises latent gradient field. 
+2) In particular, it forces it to be equal to what a true physics should be under such projection.
+3) We can not evaluate physics everywhere but we can at particular carefully-selected points. We call these points collocation points.
+4) We feed a lot of collocation points and ask a network to do interpolation \end{frame}
 
-Derivation of PI loss through chain rule.
-
-Problem with evaluating RHS of differential equations.
-
-Interpretation with cross-borrowing strength. 
+\begin{frame}{Results: Extrapolation to Unknown Regimes}
+1) We show that network can indeed interpolate between collocations. Moreover, it can fill the whole unknown regimes of behaviour. (Duffing example with explanation)
 \end{frame}
 
-\begin{frame}{Physics-Informed Koopman Networks}
-	Linear latent dynamics
-
-	Paper
-
-	Not a rom so not suitable for downstream tasks. 
+\begin{frame}{Results: Stable Long-Term Predictions}
+Fig 3.3.7
 \end{frame}
 
-\begin{frame}{Physics-Informed Neural ODEs (PINODE)}
-	Nonlinearity in latent space
-	- require a lot of data
-	- picky to initialization
-	- good interpolation bad extrapolation	
+\begin{frame}{Results: Learning From Collocations}
+1) Finally we show that collocations can be even more useful than the data itself.
+2) The difference is especially prominent in low-data regime.
+3) It shows that collocations are powerful source of information and that the network can indeed interpolate between them.
 \end{frame}
 
-\begin{frame}{PINODE Results on Burgers}
-	Add PI loss => Fixes most of the issues
-	(example with Burgers)
-	Points learned
-	- Inform the model about unseen conditions
-	Paper reference
+\begin{frame}{Single-Pixel Imaging}
+1) Now we switch gears to show you an application to single-pixel imaging
+2) It shows how powerful these differentiable reduced-order models can be in applications besides forecasting and motivate the need to improve those.
+3) SPI setup is a single-pixel camera with mirror array… 
 \end{frame}
 
-\begin{frame}{PINODE Active}
-	> PIKN: "we use it and it works"
-	> PINODE: "this is how you need to use it right"
-	> PINODE-Active: can we use it right automatically?
-
-	Loop scheme
+\begin{frame}{Compressive Sensing with Reduced-Order Models}
+	1) We introduce a ROM as a relaxation of a PDE-constrained compressive sensing.
+	2) We search in the latent space of a ROM for the right trajectory that matches both the ROMs predictions and the compressive-sensing observations
 \end{frame}
 
-\begin{frame}{PINODE Active Results}
-	Best result (Burgers)
+\begin{frame}{Results: Burger's Equation}
+	Results on Burgers
+	Maybe results on a harder problem
 \end{frame}
 
-\begin{frame}{PINODE-CS}
-	Application of the above 
-	Loss and setup
+\begin{frame}{Results: Kolmogorov Flow OR Real Example}
+
 \end{frame}
 
-\begin{frame}{PINODE-CS Results}
+\begin{frame}{Conclusion}
 	Results on Burgers

### 2023-05-16 presentation/phd_defense_pres.tex
Add CS plots

@@
 \maketitle
-
-\begin{frame}{TODOs}
-\end{frame}
 
@@
 \begin{frame}{Single-Pixel Imaging}
-1) Now we switch gears to show you an application to single-pixel imaging
-2) It shows how powerful these differentiable reduced-order models can be in applications besides forecasting and motivate the need to improve those.
-3) SPI setup is a single-pixel camera with mirror array… 
+%1) Now we switch gears to show you an application to single-pixel imaging
+%2) It shows how powerful these differentiable reduced-order models can be in applications besides forecasting and motivate the need to improve those.
+%3) SPI setup is a single-pixel camera with mirror array… 
+\begin{figure}
+	\includegraphics[width=\textwidth]{Figures/SPI_setup.pdf}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Single-Pixel Imaging}
+	% This is how SPI setup translates into Math
+	\includegraphics[width=\textwidth]{Figures/spi_math.png}
 \end{frame}
@@
 \begin{frame}{Compressive Sensing with Reduced-Order Models}
-	1) We introduce a ROM as a relaxation of a PDE-constrained compressive sensing.
-	2) We search in the latent space of a ROM for the right trajectory that matches both the ROMs predictions and the compressive-sensing observations
+	%1) We introduce a ROM as a relaxation of a PDE-constrained compressive sensing.
+	%2) We search in the latent space of a ROM for the right trajectory that matches both the ROMs predictions and the compressive-sensing observations
+	\includegraphics[width=\textwidth]{Figures/cs_schematics.png}
 \end{frame}

### 2023-05-17 presentation/phd_defense_pres.tex
Add slides for physics-informed loss

@@
 \begin{frame}{Physics-Informed Loss}
-1) We introduce physics to this system by adding a term which regularises latent gradient field. 
-2) In particular, it forces it to be equal to what a true physics should be under such projection.
-3) We can not evaluate physics everywhere but we can at particular carefully-selected points. We call these points collocation points.
-4) We feed a lot of collocation points and ask a network to do interpolation \end{frame}
+%1) We introduce physics to this system by adding a term which regularizes latent gradient field. 
+%2) In particular, it forces it to be equal to what a true physics should be under such projection.
+%3) We can not evaluate physics everywhere but we can at particular carefully-selected points. We call these points collocation points.
+%4) We feed a lot of collocation points and ask a network to do interpolation 
+\includegraphics[width=\textwidth]{Figures/collocations_1.png}
+\end{frame}
+
+\begin{frame}{Physics-Informed Loss}
+	\includegraphics[width=\textwidth]{Figures/collocations_2.png}
+\end{frame}
+
+\begin{frame}{Physics-Informed Loss}
+	\includegraphics[width=\textwidth]{Figures/collocations_3.png}
+\end{frame}
+
+\begin{frame}{Physics-Informed Loss}
+	\includegraphics[width=\textwidth]{Figures/collocations_4.png}
+\end{frame}
+
+\begin{frame}{Physics-Informed Loss}
+	\includegraphics[width=\textwidth]{Figures/collocations_5.png}
+\end{frame}
 
@@
 \begin{frame}{Results: Burger's Equation}
-	Results on Burgers
-	Maybe results on a harder problem
+	When we capture 32 samples per frame:
+	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_32.pdf}
+\end{frame}
+
+\begin{frame}{Results: Burger's Equation}
+	When we capture 32 samples per frame:
+	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_32.pdf}
+
+	When we capture 2 samples per frame:
+	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_2.pdf}
+\end{frame}
+
+\begin{frame}{Results: Burger's Equation}
+	Aggregated results
+\end{frame}
+
+\begin{frame}{Results: Interpretation}
+
 \end{frame}
@@
 \begin{frame}{Results: Kolmogorov Flow OR Real Example}
-
 \end{frame}

### 2023-05-18 presentation/phd_defense_pres.tex
Add Duffing example

@@
 \newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
-\title{Physics-Informed Neural ODEs (PINODE)}
+\title{PhD Defense}
 \date{\today}
-\author{Aleksei Sholokhov, Steven Brunton, J. Nathan Kutz, Hassan Mansour, Saleh Nabi }
+\author{Aleksei Sholokhov}
 %\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
@@
 
-\begin{frame}{Results: Extrapolation to Unknown Regimes}
-1) We show that network can indeed interpolate between collocations. Moreover, it can fill the whole unknown regimes of behaviour. (Duffing example with explanation)
+\begin{frame}{Results: Extrapolation to Unknown Regions}
+% 1) We show that network can indeed interpolate between collocations. Moreover, it can fill the whole unknown regimes of behavior. (Duffing example with explanation)
+Duffing Oscillator on a low-dimensional (2D) manifold:
+\begin{equation}
+    \label{eq:duffing_definition}
+    \begin{split}
+    \frac{dz_1}{dt} & = z_2 \\ 
+    \frac{dz_2}{dt} & = z_1 - z_1^3
+    \end{split}
+\end{equation}
+
+Projection to a high-dimensional (128) space:
+\begin{equation}
+    \label{eq:duffing_true_decoder}
+    \bd{x} := \mathcal{A}(\bd{z}) = A\bd{z}^3, \quad A \in \mathbb{R}^{128 \times 2}, \quad A_{ij} \sim_{i.i.d.} \mathcal{N}(0, 1)
+\end{equation}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{Figures/duff_results.png}
+\end{figure}
+
 \end{frame}
@@
 \begin{frame}{Results: Stable Long-Term Predictions}
-Fig 3.3.7
+\begin{figure}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{Figures/duff_full_train.pdf}
+	\end{subfigure}%
+	\begin{subfigure}[b]{0.7\textwidth}
+		\centering
+		\includegraphics[width=0.7\textwidth]{Figures/duffing_periods.pdf}
+	\end{subfigure}
+\end{figure}
+\begin{figure}
+	\includegraphics[width=0.7\textwidth]{Figures/duffing_periods_examples.png}
+\end{figure}
 \end{frame}

### 2023-05-23 cs.tex
Move current results on CS to the thesis

@@
 The former ensures that each single-pixel measurement contains as much information as possible by focusing the mirrors on the selected regions of the frame~\cite{zhang2017fast,sun2017russian,xu20181000}.
-Although highly effective, such an approach cannot be applied to detectors that do not offer fine-grained control of their mirrors.
-The latter group focuses on creating more efficient optimization algorithms for the decoding step~\cite{katz2009compressive}.
+Although highly effective, such an approach cannot be applied to monitoring scenes that don't have a ``main zone'' where the information acquisition could be focused.
+For example, a camera monitoring a gas tank for leaks can not have a mask that works well for leaks appearing only in the middle of the scene.
+The second group of approaches focuses on creating more efficient optimization algorithms for the decoding step~\cite{katz2009compressive}.
 It includes non-iterative algorithms such as differential ghost imaging (DGI) \cite{gong2010method}, as well as iterative methods such as total variation (TV)~\cite{suo2016signal} regularization.
@@
 Figure~\ref{fig:cs_abstract} illustrates the architecture of the ROM.
-Since this section assumes that such a ROM comes pre-trained, we consider the weights $\theta$ to be frozen and omit them from the notation.
+We assume that such ROM comes fully trained, so we consider its weights $\theta$ to be frozen and omit them from the notation.
 
@@
 
-In particular, we consider a single pixel camera setup in which, for every time instance $t$, $p$ acquisitions $y(t)$ are obtained by a high sampling rate photo-detector using the projection matrix $A_t$.
+We consider a single pixel camera setup in which, for every time instance $t$, $p$ acquisitions $y(t)$ are obtained by a high sampling rate photo-detector using the projection matrix $A_t$.
 The rows of the matrix $A_t$ correspond to a binary mask pattern that can be encoded using a digital micro-mirror device (DMD, \cite{pittman1995optical, sampsell1993overview}).
@@
 A modern DMD supports snapshot frequencies of around 10~kHz (\cite{wang2022single}).
-When the dynamics of the recorded scene are much slower, one can assume that such a device can take multiple measurements simultaneously\footnote{When the dynamics change rapidly, one would have to use an array of DMD devices to achieve more than one sample per frame}.
+When the scene changes much slower one can assume that such a device can take multiple measurements of a scene simultaneously\footnote{When the dynamics change rapidly, one would have to use an array of DMD devices to achieve more than one sample per frame}.
 The number of samples recorded per  frame is referred to as \textit{samples per frame}, or SPF.
 It is often divided by the total number of mirrors of the DMD device and multiplied by 100\%, in which case it is called a \textit{sample per frame rate} $\beta$.
-For a compressive-sensing-based approach to be practically useful it should require a relatively small $\beta$ (e.g.
+For a compressive-sensing-based image reconstruction approach to be practically useful it should require a relatively small rate $\beta$ (e.g.
 2-10\%) for a successful reconstruction  of the scene.
@@
 
-We compare our approach (PINODE) with three alternatives: Digital Ghost Imaging by~\cite{ferri2010differential} and~\cite{gong2010method}, Total Variation Regularization  (TVR) by~\cite{bian2018experimental}, and an approach by~\cite{bora2017compressed} which uses an autoencoder (AE).
+We compare our approach (PINODE) with four alternatives: Digital Ghost Imaging (DGI) by~\cite{ferri2010differential} and~\cite{gong2010method}, its Autoencoder-enhanced variation (DGI+Conv.
+Decoder) by~\cite{wang2022single}, Total Variation Regularization  (TVR) by~\cite{bian2018experimental}, and an approach by~\cite{bora2017compressed} which uses an autoencoder (AE).
 We measure reconstruction accuracy with three commonly-used metrics: the residual mean-squared error (RMSE), the peak signal-to-noise ratio (PSNR), and the structural similarity index measure (SSIM).
@@
     	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_64.pdf}
-    \caption{\label{fig:cs_burgers_example_large_spf} Example reconstruction with 64 samples per frame (SPF). Unsurprisingly, all handle the job well since SPF is equal to the number of pixels.}
+    \caption{\label{fig:cs_burgers_example_large_spf} Example reconstruction with 32 samples per frame (SPF), an SPF rate of 50\%.
+Rather unsurprisingly, all algorithms complete the reconstruction faithfully due to a large amount of samples.}
 \end{figure}
@@
     	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_4.pdf}
-    \caption{\label{fig:cs_burgers_example_small_spf}Example reconstruction with 4 samples per frame (SPF).
-All algorithms except our proposed method fail to recover the image due to the insufficient SPF rate (~6\%).
-In contrast, our proposed method achieves nearly the same reconstruction quality as in Figure~\ref{fig:cs_burgers_example_large_spf} with SPF rate of 100\%.}
+    \caption{\label{fig:cs_burgers_example_small_spf}Example reconstruction with 2 samples per frame (SPF).
+All algorithms except our proposed method fail to recover the image due to the insufficient SPF rate (~3\%).
+In contrast, our proposed method achieves nearly the same reconstruction quality as in Figure~\ref{fig:cs_burgers_example_large_spf} with SPF rate of 50\%.}
 \end{figure}
@@
 We plot median PSNR, RMSE, and SSIM over different number of samples per frame taken.
-We see that PINODE achieves 70\% higher peak signal-to-noise-ratio (PSNR) and 4 times higher structural similarity index measure (SSIM) when SPF is low relative to its nearest competitor (AE).
+We see that PINODE achieves 70\% higher peak signal-to-noise-ratio (PSNR) and 4 times higher structural similarity index measure (SSIM)  relative other algorithms when SPF is low.
 We also see that the DGI and TVR approaches do not reconstruct images faithfully until SPF approaches the number of pixels in a snapshot; Figure~\ref{fig:cs_burgers_example_large_spf} provides an example of high-SPF reconstruction.
 In contrast, PINODE is able to reconstruct most trajectories with high accuracy using a very small number of samples per frame e.g.
-4, as visualized on Figure~\ref{fig:cs_burgers_example_small_spf}.
-This happens because the reduced-order model of PINODE serves as a strong prior: the model reconstructs the trajectory as a whole instead of reconstructing every snapshot separately, as all other algorithms do in this comparison.
+two samples, as visualized on Figure~\ref{fig:cs_burgers_example_small_spf}.
+It happens because the reduced-order model of PINODE serves as a strong prior: the model reconstructs the trajectory as a whole instead of reconstructing every snapshot separately, as all other algorithms do in this comparison.
 It allows PINODE to borrow strength across snapshots, even when SPF per each snapshot is small.
 
+We also study robustness of PINODE approach to noise in compressive samples $y$ relatively to other algorithms.
+Figure~\ref{fig:cs_burgers_noise_aggregate} presents an aggregated comparison of reconstruction of solutions of Burger's equations under noise for SPF=12.5\%; the setup is equivalent to the one for SPF = 8 on Figure~\ref{fig:cs_burgers_spf} .
+Namely, we plot performance metrics (PSNR, RMSE, SSIM) of reconstructions from different models against Signal-to-Noise Ratio (SNR) in the vector of compressive sensing single-pixel observations $y$.
+The performance of PINODE remains superior until all models start reconstructing the scene very poorly.
+Figure~\ref{fig:cs_burgers_noise_examples} visualizes reconstructions for selected levels of SNR.
+We see that PINODE is able to provide a faithful reconstruction even under a substantial noise (e.g.
+third column).
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{figures/cs_burgers_noise_aggregate.png}
+	\caption{\label{fig:cs_burgers_noise_aggregate} Performance of reconstruction of solutions of Burger's equations under noise for SPF=12.5\%.
+We plot the performance metrics of different models against Signal-to-Noise Ratio (SNR) in the vector of compressive sensing single-pixel observations $y$.
+The performance of PINODE remains superior until all models start reconstructing the scene very poorly.
+See Figure~\ref{fig:cs_burgers_noise_examples} for example visualizations.}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{figures/cs_burgers_noise_examples.pdf}
+	\caption{\label{fig:cs_burgers_noise_examples} Examples of compressive-sensing reconstructions of trajectories of Burger's equation under noise from Figure~\ref{fig:cs_burgers_noise_aggregate}.}
+\end{figure}
+
 \subsubsection{Kolmogorov Flow}
+\label{sec:kolmogorov_flow}
 We study the performance of our framework on Kolmogorov-like flow first-described by~\cite{dovzhenko1981generation} and later analyzed by \cite{tithof2017bifurcations,suri2017forecasting}.
@@
 As in Section~\ref{sec:method}, the ROM consists of an encoder $\phi(u)$, a decoder $\psi(z)$, and the latent dynamics $h(z)$.
-It was trained by minimizing the data-driven loss~\eqref{eq:loss_data_driven} until the prediction RMSE on a holdout set stopped improving.
+It was trained by minimizing the data-driven loss~\eqref{eq:loss_data_driven} until the prediction RMSE on a holdout set stopped improving, which took about 60 epochs.
 
@@
 The first column represents the true flow.
-The second column represents the prediction of the ROM if the initial state of the system were to be known.
-It is unknown, so the second row serves as a reference for the maximal theoretical performance of a ROM.
+The second column represents the prediction of the ROM if the initial state of the system were to be known to the model.
+It is unknown, so the second row serves as a reference performance of an underlying ROM.
 The third column displays the results of CS reconstruction of our algorithm with 512 samples per frame.
@@
 Moreover, we see that the CS algorithm is not limited in its performance by the performance of the underlying ROM: the third row indicates that the model is unable to completely recover shifting dynamics, whereas the CS algorithm aids the model's predictions with CS observations and fully reconstructs the signal.
+
+\subsubsection{Methane Leaks Data}
+In this section we use our technique to reconstruct videos of gas leaks using observations from a single-pixel camera.
+We use a subset of GasVid dataset by~\cite{wang2020machine} which consists of recordings of a methane gas leaks.
+10 videos 10 to 20 seconds each.
+We split them into 1-second-long intervals, where each interval contains 10 steps: $T = 1$ sec., $dt = 0.1$ sec.
+Next, we remove backgrounds from each video using Gaussian Mixture-based Background/Foreground Segmentation (MOG2) algorithm by~\cite{zivkovic2006efficient,zivkovic2004improved} implemented in OpenCV library~(\cite{opencv_library}).
+Finally, we split all intervals into four non-overlapping batches: train (390), dev (50), test (49), and compressive sensing (2).
+We use the first three to train, fine-tune, and select a final ROM respectively, and then we use the fourth batch for compressive sensing reconstruction.
+We use the same architecture of ROM as in Section~\ref{sec:kolmogorov_flow}.
+At the compressive sensing step we sample 512 single-pixel observations per each $240 \times 320$ video frame; it corresponds to SPF ratio of $0.66\%$ (2/3 of one percent).
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{figures/cs_gas_examples.pdf}
+	\caption{\label{fig:cs_gas_examples}}
+\end{figure}
+
+The example reconstructions are presented on Figure~\ref{fig:cs_gas_examples}.
+We see that the algorithm was able to faithfully reconstruct the trajectories.
+Although it was unable to capture the details of turbulent flows, it was able to predict the altitude, the direction, and the spread of the plumes correctly, which are the most important aspects for practical applications in real-world gas monitoring applications.
+
+We note the threefold difficulty of the task.
+First, the background subtraction struggles to separate grey smoke from a dynamic grey background, which leads to fantom clouds of smoke in processed data.
+Second, the amount of training data (3900 frames) is extremely small for training a large convolutional ROM.
+For example, \cite{wang2020machine} used the same dataset for training a simpler classification network (leak/non-leak) using ~11000 frames.
+Finally, an SPF of 0.66\% is vastly insufficient for frame-by-frame reconstruction, as most state-of-the-art methods require ~5-10\% SPF, depending on the nature of the picture (	\cite{wang2022single}).
+Thus, the inductive bias of the ROM becomes a crucial component for a successful video reconstruction.
 

### 2023-05-29 phd_thesis.tex
Remove todos
Final (sent out) version of the abstract

--- 2023-04-19 phd_thesis.tex
@@
 
-My abstract goes here. I will probably write it at the very end.
+This work summarizes two projects focused on incorporating prior knowledge into machine learning models.
+In the first project, a universal feature selection method for linear mixed-effect models is developed.
+Namely, Sparse Relaxed Regularized Regression (SR3) is extended to the case of Linear Mixed-Effect (LME) likelihoods, and we prove that one can minimize such likelihoods with proximal gradient descent.
+Theoretical underpinnings of the proposed extension are also presented, including consistency results, variational properties, implementability of optimization methods, and convergence results.
+In particular, convergence analyses are provided for a basic implementation of SR3 for LME and an accelerated hybrid algorithm.
+Numerical results show the utility and speed of these algorithms on real and simulated datasets.
+Finally, both algorithms are implemented in an open-source python package \texttt{pysr3}.
+Conveniently, this package offers complete compatibility with \texttt{scikit-learn}, so all \texttt{pysr3} models can be used in a pipeline with classic modeling blocks such as data pre-processors, randomized grid search, cross-validation, and quality metrics.
+
+The second line of work develops a framework for training Reduced-Order Models (ROMs) with Physics-Informed Neural Ordinary Differential Equations (PINODE).
+In particular, a classic technique of collocation points is adapted to transfer knowledge from a known equation to a model that approximates solutions of that equation.
+The addition of a physics-informed loss allows for exceptional data supply strategies that improve the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible.
+The resulting ROMs extrapolate forward in time more accurately, perform better for unseen initial conditions, and exhibit less sensitivity to noise.
+Finally, I show how such ROMs can be used as strong regularizers in single-pixel imaging (SPI), enabling the reduction of samples-per-frame rate by an order of magnitude relative to other state-of-the-art algorithms.
 
 \tableofcontents
-
-\listoftodos
 

### 2023-05-29 extras/table_glmm_current.tex
Update Pysr3 part

@@
 \toprule
-Algorithm &        MSR3-Fast ($\ell_1$)&         \texttt{glmmLasso}\footnote{\href{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}} \cite{groll2014variable}  &            \texttt{lmmLasso}\footnote{\href{https://rdrr.io/cran/lmmlasso/}{https://rdrr.io/cran/lmmlasso/}}\cite{schelldorfer2011estimation} & PGD ($\ell_1$) \\
+Algorithm &        MSR3-Fast ($\ell_1$)&         \texttt{glmmLasso}\footnote{\href{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}} \footcite{groll2014variable}  &            \texttt{lmmLasso}\footnote{\href{https://rdrr.io/cran/lmmlasso/}{https://rdrr.io/cran/lmmlasso/}}\footcite{schelldorfer2011estimation} & PGD ($\ell_1$) \\
 \midrule

### 2023-05-29 presentation/phd_defense_pres.tex
Update Pysr3 part

@@
 \usepackage{appendixnumberbeamer}
+\addbibresource{../bibliography.bib}
 
@@
 
-\section{Introduction}
+%\section{Introduction}
 \begin{frame}{Plan of the Defense}
 Show topics and published papers. Mention covid
-
-\end{frame}
-
-\section{Linear Mixed-Effect Models}
-\subsection{ Mixed-Effects Models}
-\begin{frame}{Feature Selection for Mixed-Effect Models}
-
-Mixed-effect models
-\begin{itemize}
-	\item Used for analyzing \textbf{combined data} across a range of \textbf{groups}.
-	\item Use covariates to separate the \textbf{population variability} from the \textbf{group variability}.
-	\item \textbf{Borrow strength} across groups to estimate key statistics. % when data within units are sparse or highly variable.
-\end{itemize}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\textwidth]{Figures/ihme_predictions.png}\footnote{Picture is taken from \href{covid19.healthdata.org}{covid19.healthdata.org} }
-\end{figure}
-
-\end{frame}
-
-\begin{frame}{Feature Selection for Mixed-Effect Models}
-Practitioners:
-\begin{itemize}
-	\item Often seek \textbf{sparse models} that select \textbf{most informative} covariates. 
-	\item Want to be \textbf{flexible but efficient} in using various sparsity-promoting terms.
-	\item Want a library to be \textbf{universal and compatible} with e.g. \texttt{sklearn}.
-\end{itemize}
-\vspace{1em}
-
-Sparse Relaxed Regularized Regression ($\mathcal{SR}3$) \cite{Zheng2018RelaxAndSplit} showed great results for t linear models:
-
-\begin{figure}
-	\includegraphics[width=0.7\textwidth]{Figures/intuition_prev_paper.png}
-\end{figure}
-
-\vspace{1em}
-\textit{\textbf{Goal}: create a feature selection library that uses a relaxation approach for feature-selection in mixed-effect models.}
-
 \end{frame}
@@
 		\item 	$y_i \in \R^{n_i}$ -- group $i$ observations  
+		\item   $u_i \in \R^q$ -- random effects
+		\item   $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects, often $\Gamma = \diag{(\gamma)}$
+		\item   $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for observation noise
 	\end{itemize}
@@
 
-   	Equivalently:
-   		\[
-   		\begin{split}
-   			y_i & = X_i\beta + \textcolor{red}{\omega_i} \\
-   			{\color{red} \omega_i} & {\color{red}\sim \NN(0, Z_i\Gamma Z_i^T + \Lambda_i)}
-   		\end{split}
-   		\]
-
-   	Simplifying assumption: 
-   	\[
-   		\Gamma = \diag(\gamma)
-   	\]
+	Unknowns: $\beta$, $u_i$, $\gamma$, sometimes $\Lambda_i$.
 
@@
 
-\begin{frame}{Notation}
-	\eq{
-   		y_i & = X_i\beta + Z_iu_i + \varepsilon_i \quad i = 1\dots m \\
-   		\varepsilon_i & \sim \NN(0, \Lambda_i) \\
-   		u_i & \sim \NN(0, \Gamma)
-   	}
+\begin{frame}{Mixed-Effect Models}
+Mixed-effect models
+\begin{itemize}
+	\item Used for analyzing \textbf{combined data} across a range of \textbf{groups}.
+	\item Use covariates to separate the \textbf{population variability} from the \textbf{group variability}.
+	\item \textbf{Borrow strength} across groups to estimate key statistics. % when data within units are sparse or highly variable.
+\end{itemize}
 
-   	\begin{itemize}
-   		\item $p$ -- number of fixed features, $q$ -- number of random effects.
-   		\item $\beta \in \R^p$ -- fixed effects, or mean effects
-   		\item $u_i \in \R^q$ -- random effects
-   		\item $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects, often $\Gamma = \diag{(\gamma)}$
-   		\item $\varepsilon_i \in \R^{n_i}$ -- observation noise
-   		\item $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for noise
-   	\end{itemize}
-   	Unknowns: $\beta$, $u_i$, $\gamma$, sometimes $\Lambda_i$.
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{Figures/ihme_predictions.png}\footnote{Picture is taken from \href{covid19.healthdata.org}{covid19.healthdata.org} }
+\end{figure}
+
 \end{frame}
 
-\begin{frame}{Likelihood for Mixed Models}
-%Maximum likelihood estimates for $\beta$ and $\gamma$ solve the problem:
-%\eq{
-%	\mathcal{LME} \quad \min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma)
-%}
+%\begin{frame}{Notation}
+%	\eq{
+%   		y_i & = X_i\beta + Z_iu_i + \varepsilon_i \quad i = 1\dots m \\
+%   		\varepsilon_i & \sim \NN(0, \Lambda_i) \\
+%   		u_i & \sim \NN(0, \Gamma)
+%   	}   	
+%   	
+%   	\begin{itemize}
+%   		\item $p$ -- number of fixed features, $q$ -- number of random effects.
+%   		\item $\beta \in \R^p$ -- fixed effects, or mean effects
+%   		\item $u_i \in \R^q$ -- random effects
+%   		\item $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects, often $\Gamma = \diag{(\gamma)}$
+%   		\item $\varepsilon_i \in \R^{n_i}$ -- observation noise
+%   		\item $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for noise
+%   	\end{itemize}
+%   	Unknowns: $\beta$, $u_i$, $\gamma$, sometimes $\Lambda_i$.
+%\end{frame}
 
+%\section{Sparse Relaxed Regularized Regression (SR3) for Linear Mixed-Effects Models}
+%\subsection{ Mixed-Effects Models}
+
+\begin{frame}{Feature Selection for Mixed-Effect Models}
+Practitioners:
+\begin{itemize}
+	\item<1-> Often seek \textbf{sparse models} which only use \textbf{most informative} covariates. 
+	\item<2-> Want the algorithm to be \textbf{efficient} but also \textbf{flexible} in using various regularizers.
+	\item<3-> Want a library to be \textbf{universal and compatible} with e.g. \texttt{scikit-learn}.
+\end{itemize}
+\vspace{1em}
+\uncover<4->{
 Optimization problem:
@@
 	}
-
+}
+\uncover<5->{
 \begin{itemize}
@@
 \end{itemize}
+}
 
@@
 \begin{itemize}
-	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \cite{Krishna2008,Lin2013}
-	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \cite{Vaida2005,Jones2011}
-	\item $R(x)$ -- SCAD penalty (\cite{Fan2001,Fan2012})
+	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \footcite{Krishna2008,Lin2013}
+	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \footcite{Jones2011}
+	\item $R(x)$ -- SCAD penalty \footcite{Fan2012}
 \end{itemize}
@@
 
+\begin{frame}{Proximal Gradient Descent for Feature Selection in MLE}
+\begin{algorithm}[H]
+\TitleOfAlgo{PGD for standard LMEs}
+\SetAlgoLined
+$\beta^+ \leftarrow\beta_0, \quad \gamma^+\leftarrow\gamma_0, \quad \alpha \leftarrow 1/L$ \tcp*[f]{Initialization}\\
+$x^+ = [\beta^+, \gamma^+]$;\\
+\While{\text{making progress}}{
+	$x^+ \leftarrow \prox_{\alpha^{-1}R + \delta_{\mathcal{C}}}(x^+ - \alpha \nabla_x\LL(x^+))$	\tcp*[f]{PGD iterations}\\	
+}
+\Return{$x^+ = [\beta^+, \gamma^+]$}
+\end{algorithm}
+\only<2>{
+\vspace{2em}
+Basic Assumptions for the PGD Algorithm (Theorem 10.15 from \footcite*{AB17})
+\begin{enumerate}
+	\item $R$ is a closed proper convex function
+	\item $\LL$ is closed and proper, $\dom{\LL}$ convex, $\dom R \subset int(\dom\LL)$, and $\LL$ is $L$-smooth over $int(\dom\LL)$.
+	\item The problem has an optimal solution with an optimal value $\LL^*$
+\end{enumerate}
+}
+\only<3>{
+	\begin{columns}[T,onlytextwidth]
+    \column{0.5\textwidth}
+    \vspace{2em}
+    Synthetic Benchmark:
+    \begin{itemize}
+    	\item 100 randomly-generated problems.
+    	\item $p = q = 20$.
+		\item $\beta = \gamma = \frac{1}{2}[1,2,3,\dots,10, 0\dots,0]$
+		\item 9 groups from $3$ to $15$ observations
+		\item $X_i \sim \NN(0, I)^p$, $Z_i = X_i$, $\varepsilon_i \sim \NN(0, 0.3^2I)$
+		\item Golden search for $\lambda \in [0, 10^5]$
+		\item Final model is chosen to maximize BIC
+    \end{itemize}
+
+    \column{0.5\textwidth}
+    	\vspace{1em}
+    	\centering  
+		\begin{figure}
+			\includegraphics[width=\textwidth]{Figures/benchmark_pgd.jpg}
+		\end{figure}
+  \end{columns}
+}
+\end{frame}
+
+\begin{frame}{Sparse Relaxed Regularized Regression ($\mathcal{SR}3$) \footcite{Zheng2018RelaxAndSplit}}
+\begin{equation}
+	\min_{x} f(x) + R(x)\quad \goto \quad \min_{x, w} f(x) + \frac{\eta}{2} \|x-w\|_2^2 + R(w)
+\end{equation}
+
+\begin{figure}
+	\includegraphics[width=0.7\textwidth]{Figures/intuition_prev_paper.png}
+\end{figure}
+
+\textbf{Objectives:}
+\begin{itemize}
+	\item Introduce $\mathcal{SR}3$ relaxation to linear mixed-effects models - $\mathcal{MSR}3$ (see \footcite{sholokhov2022relaxation}).
+	\item Develop theoretical foundations for it (see \footcite{aravkin2022jimtheory}).
+	\item Implement it as a \texttt{scikit-learn}-compatible Python package -- \texttt{pysr3} (see \footcite{sholokhov2023pysr3}).
+\end{itemize}
+\vspace{1em}
+
+\end{frame}
+
 \begin{frame}{SR3-Relaxation for Mixed-Effect Models ($\ouralgo$)}
@@
 
-\begin{frame}{Value Function Reformulation}
+\begin{frame}{Value Function of $\mathcal{MSR}3$}
 $\ouralgo$-relaxation replaces the original likelihood $\LL$ with a \textit{value function} $u_{\eta,\mu}$:
@@
 \eq{
-	\min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} v_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
+	\min_{\tbeta, \tgamma \in \CC} v_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
 }
-When $\eta$ is larger than the weak-convexity constant
-\begin{itemize}
-	\item $v_{\eta,\mu}$ is well-defined and continuously differentiable.
-	\item As $\mu \rightarrow 0$ and $\eta \rightarrow \infty$, cluster points of solutions to $\ouralgo$ are first-order stationary points for $\mathcal{FS-LME}$ \\ 
-	\item $v_{\eta, \mu}$ don't need to be evaluated precisely.
-\end{itemize}
+
+\textbf{NB:} $v_{\eta,\mu}(\tbeta, \tgamma)$ is smooth on $\CC$ and can be evaluated using Interior Point (IP) method
+
+%When $\eta$ is larger than the weak-convexity constant
+%\begin{itemize}
+%	\item $v_{\eta,\mu}$ is well-defined and continuously differentiable.
+%	\item As $\mu \rightarrow 0$ and $\eta \rightarrow \infty$, cluster points of solutions to $\ouralgo$ are first-order stationary points for $\mathcal{FS-LME}$ \\ 
+%	\item $v_{\eta, \mu}$ don't need to be evaluated precisely.
+%\end{itemize}
 
@@
 
-\begin{frame}{Value Function Reformulation}
+\begin{frame}{Value Function of $\mathcal{MSR}3$}
+\[
+	\min_{\beta, \gamma \in \CC} \LL(\beta, \gamma) + R(\beta, \gamma) \quad \text{vs} \quad \min_{\tbeta, \tgamma \in \CC} v_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
+\]
+
 \begin{figure}
 	\includegraphics[width=0.7\textwidth]{Figures/intuition_current.pdf}
-	\caption{\label{fig:intuition_sr3} Comparison of the level-sets for the original likelihood (left) and $\ouralgo$-likelihood (right), for fixed (top) and random (bottom) effects.}
+	\label{fig:intuition_sr3}
 \end{figure}
@@
 
-\begin{frame}{Designing an Algorithm}
-	$G_{\nu,\eta}$ encodes both gradient of a Lagrangian (lines 1-2) and the complementarity condition (line 3):
-	\eq{
-		G_{\nu,\eta}((\beta,\gamma,v),(\tbeta,\tgamma)) := \begin{bmatrix}
-			\nabla_\beta \LL(\beta, \gamma) + \eta(\beta-\tbeta) \\
-			\nabla_\gamma \LL(\beta, \gamma) + \eta(\gamma-\tgamma) - v\\
-			v \bigodot \gamma - \mu\textbf{1}
-		\end{bmatrix}
+\begin{frame}{$\mathcal{MSR}3$: Algorithm}
+
+\begin{algorithm}[H]
+\TitleOfAlgo{PGD for $\mathcal{MSR}3$}
+\SetAlgoLined
+$\tbeta^+ \leftarrow \tbeta_0, \quad \tgamma^+ \leftarrow \tgamma_0, \quad \alpha \leftarrow 1/\eta, \quad \eta > \bar{\eta}$ \tcp*[f]{Initialization}\\
+$\tw^+ := [\tbeta^+, \tgamma^+], \quad x^+ := [\beta,\gamma]$ \\	
+\While{\text{making progress in $\tw$}}{
+	\text{$x^+ \leftarrow$ IP solution on $\LL_{\eta,\mu}(x^+, \tw^+)$ s.t. $x^+ \in \CC$} \tcp*[f]{IP Iterations}\\	
+	$\nabla_\tw v_{\eta,0}(\tw^+) \leftarrow  \nabla_\tw\LL_{\eta,0}(x^+, \tw^+)$ \tcp*[f]{Evaluate Gradient} \\ 
+	$\tw^+ \leftarrow \prox_{\alpha^{-1}R + \delta_{\mathcal{C}}}(\tw^+ - \alpha \nabla_\tw v_{\eta,0}(\tw^+))$	\tcp*[f]{PGD on Value Function}\\
 }
-	We apply Newton method to $G$ while geometrically decreasing $\mu$. \\
-\textbf{Lemma:} For every $(\mu,\eta) \in \R_+\times\R_{++}$,
-\eq{
-	&(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
-	& \iff \\
-	& \exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
+\Return{$\tw^+ = [\tbeta^+, \tgamma^+]$}
+\end{algorithm}
+\only<2>{
+\vspace{1em}
+Theoretical Results \footcite{aravkin2022jimtheory}:
+\begin{enumerate}
+	\item The problem has an optimal solution with an optimal value $\Phi^*$ (\textbf{Theorem 5})
+	\item $v_{\eta,\mu}$ is well-defined (\textbf{Theorem 5}) and continuously differentiable (\textbf{Theorem 10})
+	\item $\nabla v_{\eta,\mu}$ is locally $\widetilde{L}$-continuous when $R$ is 1-coercive (\textbf{Theorem 14})
+	\item As $\mu \rightarrow 0$ (\textbf{Theorem 6}) or $\eta \rightarrow \infty$ (\textbf{Theorem 7}), cluster points of solutions to $\ouralgo$ are FOSPs for $\mathcal{FS-LME}$ 
+\end{enumerate}
 }
-If $\mu > 0$, then $\hat{v} = -\nabla\phi_\mu(\hat{\gamma})$, and if $\mu = 0$, then $\hat{v}$ is the unique KKT multiplier associated with the constraint $0 \leq \gamma$.
 \end{frame}
 
-\begin{frame}{$\ouralgo$-fast Algorithm}
-\input{../extras/08Pseudocode}
+\begin{frame}{$\mathcal{MSR}3$: Results}
+    	\centering  
+		\begin{figure}
+			\includegraphics[width=0.7\textwidth]{Figures/benchmark_pgd_msr3.jpg}
+		\end{figure}
+\end{frame}
+
+\begin{frame}{$\mathcal{MSR}3$: Algorithm}
+
+\begin{algorithm}[H]
+\TitleOfAlgo{PGD for $\mathcal{MSR}3$}
+\SetAlgoLined
+$\tbeta^+ \leftarrow \tbeta_0, \quad \tgamma^+ \leftarrow \tgamma_0, \quad \alpha \leftarrow 1/\eta, \quad \eta > \bar{\eta}$ \tcp*[f]{Initialization}\\
+$\tw^+ := [\tbeta^+, \tgamma^+], \quad x^+ := [\beta,\gamma]$ \\	
+\While{\text{making progress in $\tw$}}{
+	\text{$x^+ \leftarrow$ \textcolor{red}{IP solution on $\LL_{\eta,\mu}(x^+, \tw^+)$ s.t. $x^+ \in \CC$}} \tcp*[f]{IP Iterations}\\	
+	$\nabla_\tw v_{\eta,0}(\tw^+) \leftarrow  \nabla_\tw\LL_{\eta,0}(x^+, \tw^+)$ \tcp*[f]{Evaluate Gradient} \\ 
+	$\tw^+ \leftarrow \prox_{\alpha^{-1}R + \delta_{\mathcal{C}}}(\tw^+ - \alpha \nabla_\tw v_{\eta,0}(\tw^+))$	\tcp*[f]{PGD on Value Function}\\
+}
+\Return{$\tw^+ = [\tbeta^+, \tgamma^+]$}
+\end{algorithm}
+\vspace{2em}
+\only<2>{
+\textbf{Key Observation:} $\nabla_\tw v(\tw)$ does not need to be evaluated exactly. We only need to come close enough to the central path.
+}
+
+\end{frame}
+
+\begin{frame}{$\mathcal{MSR}3$-fast: Algorithm}
+\begin{algorithm}[H]
+\TitleOfAlgo{$\mathcal{MSR}3$-fast}
+\SetAlgoLined
+$\tbeta^+ \leftarrow \tbeta_0, \quad \tgamma^+ \leftarrow \tgamma_0, \quad \alpha \leftarrow 1/\eta, \quad \eta > \bar{\eta}$ \tcp*[f]{Initialization}\\
+$\tw^+ := [\tbeta^+, \tgamma^+], \quad x^+ := [\beta,\gamma]$ \\	
+\While{\text{making progress}}{
+	\While{\textcolor{red}{not close enough to the central path}}{
+		\text{$x^+ \leftarrow$ IP iteration on $\LL_{\eta,\mu}(x^+, \tw^+)$ s.t. $x^+ \in \CC$} \tcp*[f]{IP Iterations}\\	
+	}
+	\text{\textcolor{red}{Decrease $\mu$}}\\
+	$\nabla_\tw v_{\eta,\mu}(\tw^+) \leftarrow  \nabla_\tw\LL_{\eta,\mu}(x^+, \tw^+)$ \tcp*[f]{Evaluate Gradient} \\ 
+	$\tw^+ \leftarrow \prox_{\alpha^{-1}R + \delta_{\mathcal{C}}}(\tw^+ - \alpha \nabla_\tw v_{\eta,\mu}(\tw^+))$	\tcp*[f]{PGD on Value Function}\\
+}
+\Return{$\tw^+ = [\tbeta^+, \tgamma^+]$}
+\end{algorithm}
 \end{frame}
@@
 
-\begin{frame}{Application to Synthetic Problems}
+\begin{frame}{$\ouralgo$-fast: Results}
 \begin{itemize}
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{Figures/performance_without_msr3.pdf}
+	\includegraphics[width=0.7\textwidth]{Figures/benchmark.jpg}
 \end{figure}
@@
 \begin{frame}{Software}
-	The code is available on GitHub: \href{github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}
+	\includegraphics[width=\textwidth]{Figures/pysr3_screenshot.png}
+	The code is available on GitHub: \href{github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}\footfullcite{sholokhov2023pysr3}
 	\begin{itemize}
-		\item All estimators are fully compatible to \texttt{sklearn} library.
+		\item All estimators are fully compatible to \texttt{scikit-learn} library.
 		\item Implements SR3 for linear, generalized-linear, and linear mixed-effect models.
@@
 
+\appendix
+\begin{frame}{Designing an Algorithm}
+	$G_{\nu,\eta}$ encodes both gradient of a Lagrangian (lines 1-2) and the complementarity condition (line 3):
+	\eq{
+		G_{\nu,\eta}((\beta,\gamma,v),(\tbeta,\tgamma)) := \begin{bmatrix}
+			\nabla_\beta \LL(\beta, \gamma) + \eta(\beta-\tbeta) \\
+			\nabla_\gamma \LL(\beta, \gamma) + \eta(\gamma-\tgamma) - v\\
+			v \bigodot \gamma - \mu\textbf{1}
+		\end{bmatrix}
+	}
+	We apply Newton method to $G$ while geometrically decreasing $\mu$. \\
+\textbf{Lemma:} For every $(\mu,\eta) \in \R_+\times\R_{++}$,
+\eq{
+	&(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
+	& \iff \\
+	& \exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
+}
+If $\mu > 0$, then $\hat{v} = -\nabla\phi_\mu(\hat{\gamma})$, and if $\mu = 0$, then $\hat{v}$ is the unique KKT multiplier associated with the constraint $0 \leq \gamma$.
+\end{frame}
+
+\begin{frame}{$\ouralgo$-fast Algorithm}
+\input{../extras/08Pseudocode}
+\end{frame}
+
 \begin{frame}{References}
-	\textbf{References}: 
-	\bibliographystyle{plain}
-	\bibliography{../bibliography.bib}
+\printbibliography
 \end{frame}

### 2023-05-30 presentation/phd_defense_pres.tex
Populate PINODE part

@@
 
+\AtBeginSection[]{
+  \begin{frame}
+  \vfill
+  \centering
+  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
+    \usebeamerfont{title}\insertsectionhead\par%
+  \end{beamercolorbox}
+  \vfill
+  \end{frame}
+}
+
 \begin{document}
@@
 \end{frame}
+
+\section{$\mathcal{MSR}3$ -- Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}
 
@@
 \begin{itemize}
-	\item Introduce $\mathcal{SR}3$ relaxation to linear mixed-effects models - $\mathcal{MSR}3$ (see \footcite{sholokhov2022relaxation}).
+	\item Extend $\mathcal{SR}3$ relaxation to linear mixed-effects models - $\mathcal{MSR}3$ (see \footcite{sholokhov2022relaxation}).
 	\item Develop theoretical foundations for it (see \footcite{aravkin2022jimtheory}).
@@
 
-\section{Experiments}
-\subsection{Application to Synthetic Problems}
-
 \begin{frame}{$\ouralgo$-fast: Results}
@@
 
-\section{PINODE}
+\section{Physics-Informed Neural ODE (PINODE): Embedding Physics into Models using Collocation Points}
 
@@
 
-1) People used to model physical systems with first-principle knowledge
-2) Data-Driven modelling of dynamical systems became a big thing  
-3) However, it requires a lot of data 
-4) Incorporating prior knowledge is a big recent trend, so history does a spiral
+%1) People used to model physical systems with first-principle knowledge
+%2) Data-Driven modelling of dynamical systems became a big thing  
+%3) However, it requires a lot of data 
+%4) Incorporating prior knowledge is a big recent trend, so history does a spiral
+
+\uncover<2->{
+\begin{columns}[T,onlytextwidth]
+\column{0.6\textwidth}
+\textcolor{orange}{First-Principle Models:}
+\begin{itemize}
+	\item Require extensive knowledge of the phenomenon
+	\item Require a lot of compute for simulating large-scale phenomena
+\end{itemize}
+\column{0.4\textwidth}
+\centering
+\includegraphics[width=\textwidth]{Figures/NN.jpg}
+\end{columns}
+}
+\vspace{2em}
+\uncover<3->{
+\begin{columns}[T,onlytextwidth]
+\column{0.6\textwidth}
+\textcolor{green}{Data-Driven Models:}
+\begin{itemize}
+	\item Require a lot of data
+	\item Often struggle to extrapolate
+\end{itemize}
+\column{0.4\textwidth}
+\centering
+\includegraphics[width=0.8\textwidth]{Figures/nn_schematics.pdf}
+\end{columns}
+}
+\vspace{2em}
+\uncover<4->{
+\begin{columns}[T,onlytextwidth]
+\column{0.6\textwidth}
+\textcolor{blue}{Hybrid Models: }
+\begin{itemize}
+	\item Incorporate elements of both approaches
+	\item Supplement data with knowledge or priors
+\end{itemize}
+\column{0.4\textwidth}
+\centering
+\includegraphics[width=0.7\textwidth]{Figures/NS_pic.png}
+\end{columns}
+}
+\end{frame}
+
+\begin{frame}{Incorporating Knowledge of Physics into Neural Networks}
+\begin{enumerate}
+	\item<2-> \textbf{Symmetry Based:} incorporate symmetries and conservation laws as hard constraints into a network: E3NNs\footcite{geiger2022e3nn}, LieConv\footcite{Finzi2020}, RiCNN \footcite{chidester2018rotation}
+	\item<3-> \textbf{Model Discovery Based:} use a library of terms to discover simple equations: SINDy\footcite{champion2019data}, Genetic Programming\footcite{schmidt2009distilling}
+	\item<4-> \textbf{Equations Based:} incorporate first-principle models to aid training of networks: PINNs\footcite{raissi2017physics}, UDEs\footcite{rackauckas2020udes}
+\end{enumerate}
+
+\begin{columns}[T,onlytextwidth]
+\column{0.3\textwidth}
+	\uncover<2->{\includegraphics[width=0.7\textwidth]{Figures/ricnn.png}}
+\column{0.7\textwidth}
+	\uncover<3->{\includegraphics[width=\textwidth]{Figures/sindy.png}}
+\end{columns}
 
@@
 
-\begin{frame}{Incorporating Knowledge into Models}
-1) There are multiple ways of incorporating knowledge into system 
-4) The overall umbrella term for it is physics-informed machine learning
-2) Some use the equations that model phenomena
-3) Some take aspects of it, e.g. symmetries and preservation laws, and forces
-A network to respect those
-5) Our work falls into the first category of approaches
-\end{frame}
-
 \begin{frame}{Reduced-Order Models (ROMs)}
-\includegraphics[width=\textwidth]{Figures/roms_1}
-\end{frame}
-
-\begin{frame}{Reduced-Order Models (ROMs)}
-\includegraphics[width=\textwidth]{Figures/roms_2}
-\end{frame}
-
-\begin{frame}{Reduced-Order Models (ROMs)}
-\includegraphics[width=\textwidth]{Figures/roms_3}
-\end{frame}
-
-\begin{frame}{Reduced-Order Models (ROMs)}
-\includegraphics[width=\textwidth]{Figures/roms_4}
+\only<1>{\includegraphics[width=\textwidth]{Figures/roms_1}}
+\only<2>{\includegraphics[width=\textwidth]{Figures/roms_2}}
+\only<3>{\includegraphics[width=\textwidth]{Figures/roms_3}}
+\only<4>{\includegraphics[width=\textwidth]{Figures/roms_4}}
 \end{frame}
@@
 %4) We feed a lot of collocation points and ask a network to do interpolation 
-\includegraphics[width=\textwidth]{Figures/collocations_1.png}
-\end{frame}
-
-\begin{frame}{Physics-Informed Loss}
-	\includegraphics[width=\textwidth]{Figures/collocations_2.png}
-\end{frame}
-
-\begin{frame}{Physics-Informed Loss}
-	\includegraphics[width=\textwidth]{Figures/collocations_3.png}
-\end{frame}
-
-\begin{frame}{Physics-Informed Loss}
-	\includegraphics[width=\textwidth]{Figures/collocations_4.png}
-\end{frame}
-
-\begin{frame}{Physics-Informed Loss}
-	\includegraphics[width=\textwidth]{Figures/collocations_5.png}
+Using chain rule:
+\only<1>{\includegraphics[width=\textwidth]{Figures/collocations_1.png}}
+\only<2>{\includegraphics[width=\textwidth]{Figures/collocations_2.png}}
+\only<3>{\includegraphics[width=\textwidth]{Figures/collocations_3.png}}
+\only<4>{\includegraphics[width=\textwidth]{Figures/collocations_4.png}}
+\only<5>{\includegraphics[width=\textwidth]{Figures/collocations_5.png}}
 \end{frame}
@@
 \end{equation}
-
+\uncover<2->{
 \begin{figure}
@@
 \end{figure}
-
+}
 \end{frame}
@@
 \end{figure}
+\uncover<2->{
 \begin{figure}
@@
 \end{figure}
+}
+\end{frame}
+
+\begin{frame}{Results: Burgers' Equation}
+Burgers' Equation:
+\begin{equation}
+	\begin{split}
+		& u_t  + uu_x = \nu u_{xx} \\
+	    & u(-\pi, t) = u(\pi, t),\quad \forall t \in [0, T]
+	\end{split}
+\end{equation}
+\uncover<2->{
+\includegraphics[width=\textwidth]{Figures/burgers_examples_of_ics.pdf}
+}
+\uncover<3->{
+\includegraphics[width=\textwidth]{Figures/example_burgers.pdf}
+}
+
 \end{frame}
@@
 \begin{frame}{Results: Learning From Collocations}
-1) Finally we show that collocations can be even more useful than the data itself.
-2) The difference is especially prominent in low-data regime.
-3) It shows that collocations are powerful source of information and that the network can indeed interpolate between them.
+\includegraphics[width=\textwidth]{Figures/burgers_examples_of_ics.pdf}
+\includegraphics[width=\textwidth]{Figures/data_vs_collocations.pdf}
 \end{frame}
+
+\begin{frame}{Discussion and Limitations}
+We showed that:
+\begin{itemize}
+	\item Physics-informed loss improves accuracy and forecasting stability of ROMs
+	\item Collocations can supplement data to improve model's performance for unseen initial conditions.
+\end{itemize}
+\vspace{2em}
+Limitations:
+\begin{itemize}
+	\item Optimal choice of collocations is problem-specific
+	\item Need a lot of collocations
+	\begin{itemize}
+		\item \textit{Could} be possible to overcome with smarter sampling techniques
+	\end{itemize}
+\end{itemize}
+\end{frame}
+
+\section{Single pixel imaging of spatio-temporal flows using differentiable latent dynamics}
 
@@
 \appendix
+\section{Appendix: $\mathcal{MSR}3$}
 \begin{frame}{Designing an Algorithm}
@@
 
-\begin{frame}{References}
-\printbibliography
+\begin{frame}[noframenumbering,plain,allowframebreaks]{References}
+    \printbibliography[heading=none]
 \end{frame}

### 2023-05-30 presentation/phd_defense_pres.tex
Add pictures souces

@@
 \column{0.6\textwidth}
-\textcolor{orange}{First-Principle Models:}
+\textcolor{orange}{First-Principle Models}
 \begin{itemize}
@@
 \column{0.6\textwidth}
-\textcolor{green}{Data-Driven Models:}
+\textcolor{green}{Data-Driven Models}
 \begin{itemize}
@@
 \column{0.6\textwidth}
-\textcolor{blue}{Hybrid Models: }
+\textcolor{blue}{Hybrid Models}
 \begin{itemize}
@@
 \centering
-\includegraphics[width=0.7\textwidth]{Figures/NS_pic.png}
+\includegraphics[width=0.5\textwidth]{Figures/NS_pic.png}
 \end{columns}
+
+Pictures sources: \footnote{\href{https://en.wikipedia.org/wiki/Navier–Stokes_equations}{Wikipedia: Navier-Stokes}},\footnote{\href{https://www.tibco.com/reference-center/what-is-a-neural-network}{tibco.com}},\footnote{\href{https://www.comsol.com/model/water-purification-reactor-14049}{COMSOL Simulations}}
 }
@@
 	\begin{itemize}
-		\item \textit{Could} be possible to overcome with smarter sampling techniques
+		\item \textit{Seems} possible to overcome with smarter sampling techniques
 	\end{itemize}

### 2023-05-31 presentation/phd_defense_pres.tex
Add Intuition and Kolmogorov slides for CS

@@
 \begin{frame}{Results: Burger's Equation}
+	\uncover<1->{
 	When we capture 32 samples per frame:
 	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_32.pdf}
+	}
+	\uncover<2->{
+	When we capture 2 samples per frame:
+	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_2.pdf}
+	}
+
 \end{frame}
@@
 \begin{frame}{Results: Burger's Equation}
-	When we capture 32 samples per frame:
-	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_32.pdf}
-
-	When we capture 2 samples per frame:
-	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_2.pdf}
+	\includegraphics[width=\textwidth]{Figures/cs_burgers_psnr.png}
 \end{frame}
 
-\begin{frame}{Results: Burger's Equation}
-	Aggregated results
+\begin{frame}[label=again]{Results: Kolmogorov Flow}
+	\animategraphics[loop,width=\textwidth]{5}{Figures/kolm_example_anim_}{0}{19}
+\end{frame}
+
+\begin{frame}{Results: Kolmogorov Flow}
+	\includegraphics[width=\textwidth]{Figures/cs_kolm_aggregate.png}
 \end{frame}
@@
 \begin{frame}{Results: Interpretation}
+	\only<1>{\includegraphics[width=\textwidth]{Figures/cs_int_1.png}}
+	\only<2>{\includegraphics[width=\textwidth]{Figures/cs_int_2.png}}
+	\only<3>{\includegraphics[width=\textwidth]{Figures/cs_int_3.png}}
+\end{frame}
+
+\againframe{again}
+
+\begin{frame}{Results: Interpretation}
+	\only<1>{\includegraphics[width=\textwidth]{Figures/cs_int_3.png}}
+	\only<2>{\includegraphics[width=\textwidth]{Figures/cs_int_4.png}}
+	\only<3>{\includegraphics[width=\textwidth]{Figures/cs_int_5.png}}
+\end{frame}
+
+\againframe{again}
+
+\begin{frame}{Results: Interpretation}
+	\only<1>{\includegraphics[width=\textwidth]{Figures/cs_int_5.png}}
+	\only<2>{\includegraphics[width=\textwidth]{Figures/cs_int_6.png}}
+	\only<3>{\includegraphics[width=\textwidth]{Figures/cs_int_7.png}}
+\end{frame}
+
+\againframe{again}
+
+\begin{frame}{Results: Gas Monitoring}
 
@@
 
-\begin{frame}{Results: Kolmogorov Flow OR Real Example}
-\end{frame}
+\begin{frame}{Conclusion and Acknowledgments}
 
-\begin{frame}{Conclusion}
-	Results on Burgers
-	Maybe results on a harder problem
 \end{frame}

### 2023-06-02 presentation/phd_defense_pres.tex
SIAM OPT presentation complete

@@
 \newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
-\title{PhD Defense}
+%\title{PhD Defense}
+%\date{\today}
+%\author{Aleksei Sholokhov}
+%\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
+\title{$\mathcal{MSR}3$ -- Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}
 \date{\today}
-\author{Aleksei Sholokhov}
-%\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
+\author{Aleksei Sholokhov \\ \vspace{1em}  Together with James V. Burke, Peng Zheng, Damian Santomauro, \\ and Aleksandr Aravkin}
+\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
 
-\AtBeginSection[]{
-  \begin{frame}
-  \vfill
-  \centering
-  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
-    \usebeamerfont{title}\insertsectionhead\par%
-  \end{beamercolorbox}
-  \vfill
-  \end{frame}
-}
+%\AtBeginSection[]{
+%  \begin{frame}
+%  \vfill
+%  \centering
+%  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
+%    \usebeamerfont{title}\insertsectionhead\par%
+%  \end{beamercolorbox}
+%  \vfill
+%  \end{frame}
+%}
 
@@
 %\section{Introduction}
-\begin{frame}{Plan of the Defense}
-Show topics and published papers. Mention covid
+%\begin{frame}{Plan of the Defense}
+%Show topics and published papers. Mention covid
+%\end{frame}
+
+\section{$\mathcal{MSR}3$ -- Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}
+
+\begin{frame}{Research Objectives and Plan}
+
+\textbf{Objectives:}
+\begin{itemize}
+	\item Extend $\mathcal{SR}3$ relaxation to linear mixed-effects models - $\mathcal{MSR}3$ (see \footcite{sholokhov2022relaxation}).
+	\item Develop theoretical foundations for it (see \footcite{aravkin2022jimtheory}).
+	\item Implement it as a \texttt{scikit-learn}-compatible Python package -- \texttt{pysr3} (see \footcite{sholokhov2023pysr3}).
+\end{itemize}
+
+\vspace{3em}
+
+\textbf{Plan of the Talk:}
+\begin{enumerate}
+	\item Proximal Gradient Descent (PGD) for Linear Mixed-Effects Models (LMEs)
+	\item $\mathcal{MSR}3$ --  PGD on $\mathcal{SR}3$-relaxation for LMEs
+	\item $\mathcal{MSR}3$-fast -- practical algorithm for feature selection
+\end{enumerate}
 \end{frame}
 
-\section{$\mathcal{MSR}3$ -- Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}
+\begin{frame}{Mixed-Effect Models}
+Mixed-effect models
+\begin{itemize}
+	\item Used for analyzing \textbf{combined data} across a range of \textbf{groups}.
+	\item \textbf{Borrow strength} across groups to estimate key statistics. % when data within units are sparse or highly variable.
+	\item Use covariates to separate the \textbf{population variability} from the \textbf{group variability}.
+
+\end{itemize}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{Figures/ihme_predictions.png}\footnote{Picture is taken from \href{covid19.healthdata.org}{covid19.healthdata.org} }
+\end{figure}
+
+\end{frame}
 
@@
 
-\begin{frame}{Mixed-Effect Models}
-Mixed-effect models
-\begin{itemize}
-	\item Used for analyzing \textbf{combined data} across a range of \textbf{groups}.
-	\item Use covariates to separate the \textbf{population variability} from the \textbf{group variability}.
-	\item \textbf{Borrow strength} across groups to estimate key statistics. % when data within units are sparse or highly variable.
-\end{itemize}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\textwidth]{Figures/ihme_predictions.png}\footnote{Picture is taken from \href{covid19.healthdata.org}{covid19.healthdata.org} }
-\end{figure}
-
-\end{frame}
-
 %\begin{frame}{Notation}
@@
 
-\begin{frame}{Feature Selection for Mixed-Effect Models}
-Practitioners:
-\begin{itemize}
-	\item<1-> Often seek \textbf{sparse models} which only use \textbf{most informative} covariates. 
-	\item<2-> Want the algorithm to be \textbf{efficient} but also \textbf{flexible} in using various regularizers.
-	\item<3-> Want a library to be \textbf{universal and compatible} with e.g. \texttt{scikit-learn}.
-\end{itemize}
-\vspace{1em}
-\uncover<4->{
+\begin{frame}{Negative Log-Likelihood for Mixed-Effect Models}
 Optimization problem:
@@
 	}
-}
-\uncover<5->{
 \begin{itemize}
@@
 \end{itemize}
-}
 
@@
 \begin{frame}{Regularization}
-\begin{itemize}
-	\item $R(\beta, \gamma)$ is closed, proper, with easily computed \textit{prox operator}
-\end{itemize}
+%Practitioners:
+%\begin{itemize}
+%	\item<1-> Often seek \textbf{sparse models} which only use \textbf{most informative} covariates. 
+%	\item<2-> Want the algorithm to be \textbf{efficient} but also \textbf{flexible} in using various regularizers.
+%	\item<3-> Want a library to be \textbf{universal and compatible} with e.g. \texttt{scikit-learn}.
+%\end{itemize}
+%\vspace{1em}
+$R(\beta, \gamma)$ is closed, proper, with easily computed \textit{prox operator}
 
@@
 
-Examples: 
-\begin{itemize}
-	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \footcite{Krishna2008,Lin2013}
-	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \footcite{Jones2011}
-	\item $R(x)$ -- SCAD penalty \footcite{Fan2012}
-\end{itemize}
+%Examples: 
+%\begin{itemize}
+%	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \footcite{Krishna2008,Lin2013}
+%	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \footcite{Jones2011}
+%	\item $R(x)$ -- SCAD penalty \footcite{Fan2012}
+%\end{itemize}
 
@@
 \begin{frame}{Proximal Gradient Descent for Feature Selection in MLE}
+\only<1-2>{
+Let
+\eq{
+	x := [\beta, \gamma], \quad \CC := \R^p \times R^q_+
+}
+Optimization problem $\mathcal{FS-LME}$:
+
+\eq{
+	\min_{x \in \CC} \LL(x) + R(x)
+}
+}
+\uncover<2->{
 \begin{algorithm}[H]
@@
 \end{algorithm}
-\only<2>{
-\vspace{2em}
-Basic Assumptions for the PGD Algorithm (Theorem 10.15 from \footcite*{AB17})
+}
+\only<3>{
+\vspace{1em}
+Basic Assumptions for the PGD Algorithm (Theorem 10.15 from \footcite*{AB17}):
 \begin{enumerate}
 	\item $R$ is a closed proper convex function
-	\item $\LL$ is closed and proper, $\dom{\LL}$ convex, $\dom R \subset int(\dom\LL)$, and $\LL$ is $L$-smooth over $int(\dom\LL)$.
-	\item The problem has an optimal solution with an optimal value $\LL^*$
+	\item $\LL$ is closed and proper, $\dom{\LL}$ convex, $\dom R \subset int(\dom\LL)$, and $\LL$ is $L$-smooth over $int(\dom\LL)$. [$\times$].
+	\item The problem has an optimal solution\footcite{aravkin2022jimtheory} with an optimal value $\LL^*$
 \end{enumerate}
+Algorithm converges with backtracking~\footcite{Burke-Engle18}. 
 }
-\only<3>{
+\only<4>{
 	\begin{columns}[T,onlytextwidth]
     \column{0.5\textwidth}
-    \vspace{2em}
+    \vspace{1em}
     Synthetic Benchmark:
@@
 \begin{frame}{Sparse Relaxed Regularized Regression ($\mathcal{SR}3$) \footcite{Zheng2018RelaxAndSplit}}
-\begin{equation}
-	\min_{x} f(x) + R(x)\quad \goto \quad \min_{x, w} f(x) + \frac{\eta}{2} \|x-w\|_2^2 + R(w)
-\end{equation}
+
+\begin{equation*}
+		\begin{array}{c}
+			\text{Original} \\
+			 \\
+			\min\limits_{x} f(x) + R(x)
+		\end{array}
+	\quad \goto \quad 
+	\begin{array}{c}
+		\text{Decoupled} \\
+		\\
+		\min\limits_{x, w} f(x) + \frac{\eta}{2} \|x-w\|_2^2 + R(w) 
+	\end{array}
+	\quad \goto \quad 
+	\begin{array}{c}
+		\text{Value Function} \\
+		\\
+		\min\limits_{w} v_\eta(w) + R(w)
+	\end{array}
+\end{equation*}
 
@@
 \end{figure}
-
-\textbf{Objectives:}
-\begin{itemize}
-	\item Extend $\mathcal{SR}3$ relaxation to linear mixed-effects models - $\mathcal{MSR}3$ (see \footcite{sholokhov2022relaxation}).
-	\item Develop theoretical foundations for it (see \footcite{aravkin2022jimtheory}).
-	\item Implement it as a \texttt{scikit-learn}-compatible Python package -- \texttt{pysr3} (see \footcite{sholokhov2023pysr3}).
-\end{itemize}
-\vspace{1em}
 
@@
 \eq{
-	\min_{\beta \in \R^p,\, \gamma \in \R^{q}_+} \LL(\beta, \gamma) + R(\beta, \gamma)
+	\min_{x \in \CC} \LL(x) + R(x)
 }
-Relaxed problem $\ouralgo$:
+\uncover<2->{
+Decoupled problem:
 \eq{
 	\label{eq:msr3_formulation_explicit}
-	\min_{\beta, \tbeta \in \R^p,\, \gamma, \tgamma \in \R^{q}_+} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma) + R(\tbeta, \tgamma)
+	\min_{x, w \in \CC} \LL(x) + \frac{\eta}{2}\|x-w\|^2_2 + R(w)
 }
-where the \textit{relaxation} $\kappa_\eta$ decouples the likelihood and the regularizer 
+where $w = [\tbeta, \tgamma]$ and $\eta > \bar\eta$. 
+}
+
+\uncover<3->{
+By partially minimizing w.r.t. $x$ we get a value function $v_\eta(w)$:
+
 \eq{
-	\kappa_\eta(\beta - \tbeta, \gamma - \tgamma) := \frac{\eta}{2}\|\beta - \tbeta\|^2_2 + \frac{\eta}{2}\|\gamma - \tgamma\|_2^2, \quad \eta > \bar\eta
+	\label{eq:value_function_definition}
+	v_{\eta}(w) & := \min_{x \in \CC} \LL(x) + \frac{\eta}{2}\|x-w\|^2_2
 }
-and the \textit{perspective mapping} $\phi_\mu$ replaces $\gamma \geq 0$ with a log-barrier 
+
+so $\mathcal{SR}3$-formulation~(\ref{eq:msr3_formulation_explicit}) becomes
+
 \eq{
-	\phi_\mu(\gamma) := \begin{cases}
+	\min_{w \in \CC} v_{\eta}(w) + R(w)
+}
+
+}
+\uncover<4->{
+\textbf{NB:} $v_{\eta}(w)$ is smooth on $\CC$ and can be evaluated using Interior Point (IP) method
+}
+\end{frame}
+
+\begin{frame}{Value Function of $\mathcal{MSR}3$}
+$\ouralgo$-relaxation replaces the original likelihood $\LL$ with a \textit{value function} $v_{\eta,\mu}$:
+
+\begin{equation}
+	v_{\eta,\mu}(w) := \min_{x \in \CC} \LL(x) + \frac{\eta}{2}\|x-w\|^2_2 + \phi_\mu(x)
+\end{equation}
+
+where \textit{perspective mapping} $\phi_\mu$ replaces $\gamma \geq 0$ with a log-barrier 
+\eq{
+	\phi_\mu(x) = \phi_\mu(\gamma) := \begin{cases}
 		-\mu\sum_{i=1}^q \ln(\gamma_i/\mu), & \mu > 0 \\
@@
 \end{frame}
-
-\begin{frame}{Value Function of $\mathcal{MSR}3$}
-$\ouralgo$-relaxation replaces the original likelihood $\LL$ with a \textit{value function} $u_{\eta,\mu}$:
-\eq{
-	\label{eq:value_function_definition}
-	v_{\eta,\mu}(\tbeta,\tgamma) & := \min_{(\beta, \gamma)} \LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\  & := \min_{(\beta, \gamma)} \LL(\beta, \gamma) + \phi_\mu(\gamma) + \kappa_\eta(\beta - \tbeta, \gamma - \tgamma)
-}
-so $\ouralgo$-formulation~(\ref{eq:msr3_formulation_explicit}) becomes
-\eq{
-	\min_{\tbeta, \tgamma \in \CC} v_{\eta,\mu}(\tbeta, \tgamma) + R(\tbeta, \tgamma)
-}
-
-\textbf{NB:} $v_{\eta,\mu}(\tbeta, \tgamma)$ is smooth on $\CC$ and can be evaluated using Interior Point (IP) method
 
@@
 % \textbf{Key observation}: in practice, we don't need accurate solutions for~(\ref{eq:value_function_definition}): a few Newton iterations keep the solution close to the central path.
-\end{frame}
 
@@
 \While{\text{making progress in $\tw$}}{
-	\text{$x^+ \leftarrow$ IP solution on $\LL_{\eta,\mu}(x^+, \tw^+)$ s.t. $x^+ \in \CC$} \tcp*[f]{IP Iterations}\\	
-	$\nabla_\tw v_{\eta,0}(\tw^+) \leftarrow  \nabla_\tw\LL_{\eta,0}(x^+, \tw^+)$ \tcp*[f]{Evaluate Gradient} \\ 
+	\text{$x^+ \leftarrow$ IP on $\LL_{\eta,\mu}(x^+, \tw^+)$ s.t. $x^+ \in \CC$ and $\mu \goto 0$} \tcp*[f]{IP Iterations}\\	
+	$\nabla_\tw v_{\eta,0}(\tw^+) \leftarrow  \nabla_\tw\LL_{\eta,0}(x^+, \tw^+)$ \tcp*[f]{Evaluate Gradient of VF} \\ 
 	$\tw^+ \leftarrow \prox_{\alpha^{-1}R + \delta_{\mathcal{C}}}(\tw^+ - \alpha \nabla_\tw v_{\eta,0}(\tw^+))$	\tcp*[f]{PGD on Value Function}\\
@@
 \vspace{1em}
-Theoretical Results \footcite{aravkin2022jimtheory}:
+Theoretical Results \footfullcite{aravkin2022jimtheory}:
 \begin{enumerate}
@@
 \While{\text{making progress in $\tw$}}{
-	\text{$x^+ \leftarrow$ \textcolor{red}{IP solution on $\LL_{\eta,\mu}(x^+, \tw^+)$ s.t. $x^+ \in \CC$}} \tcp*[f]{IP Iterations}\\	
+	\text{$x^+ \leftarrow$ \textcolor{red}{IP on $\LL_{\eta,\mu}(x^+, \tw^+)$ s.t. $x^+ \in \CC$ and $\mu \goto 0$}} \tcp*[f]{IP Iterations}\\	
 	$\nabla_\tw v_{\eta,0}(\tw^+) \leftarrow  \nabla_\tw\LL_{\eta,0}(x^+, \tw^+)$ \tcp*[f]{Evaluate Gradient} \\ 
@@
 
-\begin{frame}{Software}
+\begin{frame}{Thank you!}
 	\includegraphics[width=\textwidth]{Figures/pysr3_screenshot.png}
@@
 \only<5>{\includegraphics[width=\textwidth]{Figures/collocations_5.png}}
+\end{frame}
+
+\begin{frame}{Physics-Informed, Data-Driven, and Hybrid Models}
+
 \end{frame}

### 2023-06-05 presentation/phd_defense_pres.tex
Last-minute edits to the pres

@@
 
-\begin{frame}{Research Objectives and Plan}
+\begin{frame}{Research Objectives and Plan of the Talk}
 
@@
 }
-where $w = [\tbeta, \tgamma]$ and $\eta > \bar\eta$. 
+where $w = [\tbeta, \tgamma]$. 
 }
@@
 \begin{enumerate}
-	\item The problem has an optimal solution with an optimal value $\Phi^*$ (\textbf{Theorem 5})
+	\item When $\eta > \bar\eta$ the problem has an optimal solution $\Phi^*$ (\textbf{Theorem 5})
 	\item $v_{\eta,\mu}$ is well-defined (\textbf{Theorem 5}) and continuously differentiable (\textbf{Theorem 10})
-	\item $\nabla v_{\eta,\mu}$ is locally $\widetilde{L}$-continuous when $R$ is 1-coercive (\textbf{Theorem 14})
+	\item When $R$ is 1-coercive $\nabla v_{\eta,\mu}$ is locally $\widetilde{L}$-continuous (\textbf{Theorem 14})
+	\item Algorithm converges to a stationary point of $v_{\eta,\mu}$ (\textbf{Theorem 15})
 	\item As $\mu \rightarrow 0$ (\textbf{Theorem 6}) or $\eta \rightarrow \infty$ (\textbf{Theorem 7}), cluster points of solutions to $\ouralgo$ are FOSPs for $\mathcal{FS-LME}$ 

### 2023-06-06 presentation/phd_defense_pres.tex
PhD defense version 1

@@
 %\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
-\title{$\mathcal{MSR}3$ -- Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}
+\title{Optimization Methods for Parameter Identifications \\ in Settings with Only Partial Knowledge}
+\subtitle{Ph.D. Defense}
 \date{\today}
-\author{Aleksei Sholokhov \\ \vspace{1em}  Together with James V. Burke, Peng Zheng, Damian Santomauro, \\ and Aleksandr Aravkin}
-\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
+\author{Aleksei Sholokhov}
+\titlegraphic{\vspace{3em}\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}}
 
-%\AtBeginSection[]{
-%  \begin{frame}
-%  \vfill
-%  \centering
-%  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
-%    \usebeamerfont{title}\insertsectionhead\par%
-%  \end{beamercolorbox}
-%  \vfill
-%  \end{frame}
-%}
+\AtBeginSection[]{
+  \begin{frame}
+  \vfill
+  \centering
+  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
+    \usebeamerfont{title}\insertsectionhead\par%
+  \end{beamercolorbox}
+  \vfill
+  \end{frame}
+}
 
@@
 
-%\section{Introduction}
-%\begin{frame}{Plan of the Defense}
-%Show topics and published papers. Mention covid
-%\end{frame}
+\begin{frame}{Plan of the Defense}
+\uncover<2->{
+\includegraphics[height=35pt]{figures/sasha.png}
+\includegraphics[height=35pt]{figures/jim.jpg}
+\includegraphics[height=35pt]{figures/peng.jpg}
+}
+\uncover<3->{
+\includegraphics[height=35pt]{figures/hassan.jpg}
+\includegraphics[height=35pt]{figures/saleh.jpeg}
+\includegraphics[height=35pt]{figures/yuying.jpeg}
+}
+\uncover<4->{
+\includegraphics[height=35pt]{figures/nathan.jpg}
+\includegraphics[height=35pt]{figures/steve.jpeg}
+\includegraphics[height=35pt]{figures/josh.jpg}
+}
+
+\vspace{1em}
+\uncover<2->{
+\textbf{Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}\\
+{\small
+Together with Aleksandr Aravkin, James Burke, Damian Santomauro, and Peng Zheng
+\begin{itemize}
+	\item \only<1-4>{``A Relaxation Approach to Feature Selection for Linear Mixed Effects Models''}\only<5->{\underline{\textcolor{teal}{``A Relaxation Approach to Feature Selection for Linear Mixed Effects Models''}}} \\ \textit{submitted to Journal of Computational and Graphical Statistics (JCGS)}
+	\item ``Analysis of Relaxation Methods for Feature Selection in Mixed Effects Models'' \\ \textit{submitted to Journal of Computational Optimization and Applications (COAP)}
+	\item ``\texttt{pysr3}: A Python Package for Spare Relaxed Regularized Regression'' \\ \textit{Journal of Open-Source Software (JOSS), 2023, ICCOPT 2022, SIAM OPT 2023}
+\end{itemize}
+}
+}
+
+\vspace{1em}
+\uncover<3->{
+\textbf{PINODE: Physics-Informed Neural ODEs}\\
+{\small
+Together with Hassan Mansour, Saleh Nabi, and Yuying Liu
+\begin{itemize}
+	\item ``Physics-Informed Koopman Network'' \\
+	\textit{SIAM ADS 2023, arXiv:2211.09419}
+	\item \only<1-4>{``Physics-Informed Neural ODE: Embedding Physics into Models using Collocation Points''}\only<5->{\underline{\textcolor{teal}{``Physics-Informed Neural ODE: Embedding Physics into Models using Collocation Points''}}} \\
+	\textit{Submitted to Nature Special Issue on Physics-Informed Machine Learning}
+\end{itemize}
+}
+}
+\uncover<4->{
+\vspace{1em}
+\textbf{Single-Pixel Imaging with Reduced-Order Models}\\
+{\small
+Together with J. Nathan Kutz, Steven Brunton, Joshua Rapp, Hassan Mansour, and Saleh Nabi
+\begin{itemize}
+	\item \only<1-4>{``Single pixel imaging of spatio-temporal flows using differentiable latent dynamics''}\only<5->{\underline{\textcolor{teal}{``Single pixel imaging of spatio-temporal flows using differentiable latent dynamics''}}} \textit{in preparation}
+\end{itemize}
+}
+}
+
+\end{frame}
 
@@
 
+\begin{frame}{Designing an Algorithm}
+	$G_{\nu,\eta}$ encodes both gradient of a Lagrangian (lines 1-2) and the complementarity condition (line 3):
+	\eq{
+		G_{\nu,\eta}((\beta,\gamma,v),(\tbeta,\tgamma)) := \begin{bmatrix}
+			\nabla_\beta \LL(\beta, \gamma) + \eta(\beta-\tbeta) \\
+			\nabla_\gamma \LL(\beta, \gamma) + \eta(\gamma-\tgamma) - v\\
+			v \bigodot \gamma - \mu\textbf{1}
+		\end{bmatrix}
+	}
+	We apply Newton method to $G$ while geometrically decreasing $\mu$. \\
+\textbf{Lemma:} For every $(\mu,\eta) \in \R_+\times\R_{++}$,
+\eq{
+	&(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
+	& \iff \\
+	& \exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
+}
+If $\mu > 0$, then $\hat{v} = -\nabla\phi_\mu(\hat{\gamma})$, and if $\mu = 0$, then $\hat{v}$ is the unique KKT multiplier associated with the constraint $0 \leq \gamma$.
+\end{frame}
+
+\begin{frame}{$\ouralgo$-fast Algorithm}
+\input{../extras/08Pseudocode}
+\end{frame}
+
 \begin{frame}{$\ouralgo$-fast: Results}
@@
 
-\begin{frame}{Thank you!}
+\begin{frame}{Software}
 	\includegraphics[width=\textwidth]{Figures/pysr3_screenshot.png}
@@
 
-\begin{frame}{Data-Driven Modeling of Physical Systems}
+\begin{frame}{Modeling of Physical Systems}
 
@@
 \column{0.6\textwidth}
-\textcolor{green}{Data-Driven Models}
+\textcolor{teal}{Data-Driven Models}
 \begin{itemize}
@@
 	\item<3-> \textbf{Model Discovery Based:} use a library of terms to discover simple equations: SINDy\footcite{champion2019data}, Genetic Programming\footcite{schmidt2009distilling}
-	\item<4-> \textbf{Equations Based:} incorporate first-principle models to aid training of networks: PINNs\footcite{raissi2017physics}, UDEs\footcite{rackauckas2020udes}
+	\item<4-> \textbf{Equations Based:} incorporate first-principle models to aid training of networks: PINNs\footcite{raissi2017physics}, UDEs\footcite{rackauckas2020udes}, \textit{PINODE}
 \end{enumerate}
@@
 
+\begin{frame}{Reduced-Order Models: Data-Driven Loss}
+\includegraphics[width=\textwidth]{Figures/just_rom.png}
+\vspace{0.5em}
+
+Notation:
+\begin{itemize}
+	\item $k$ trajectories $x_i(t_j)$, $p$ timesteps each ($i = 1, \dots, k$, $j = 1, \dots, p$)
+	\item $\sigma$ -- observation noise variance
+	\item $\omega_1$, $\omega_2$ -- weights (hyper-parameters) 
+\end{itemize}
+
+\begin{align}
+    \LL^{data}_\theta & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\underbrace{\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2}_{\text{reconstruction loss}}\right.
++ \\
+     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \underbrace{\left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2}_{\text{prediction loss}} \right]
+\end{align}
+
+\end{frame}
+
 \begin{frame}{Physics-Informed Loss}
@@
 \only<5>{\includegraphics[width=\textwidth]{Figures/collocations_5.png}}
+\only<6>{\includegraphics[width=\textwidth]{Figures/collocations_6.png}}
+\only<7>{\includegraphics[width=\textwidth]{Figures/collocations_7.png}}
 \end{frame}
@@
 \begin{frame}{Physics-Informed, Data-Driven, and Hybrid Models}
-
+	Values of $\omega_1$, $\omega_2$, $\omega_3$, and $\omega_4$ control the type of the ROM:
+	\begin{itemize}
+		\item If $\omega_1 = \omega_2 = 0$ then $\LL_\theta^\text{data} = 0$. Thus, the model is purely \textcolor{orange}{Physics-Informed}
+		\item If $\omega_3 = \omega_4 = 0$ then $\LL_\theta^\text{physics} = 0$. Thus, the model is purely \textcolor{teal}{Data-Driven}
+		\item Otherwise the model is \textcolor{blue}{Hybrid}
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{Results: Burgers' Equation}
-Burgers' Equation:
 \begin{equation}
@@
 		& u_t  + uu_x = \nu u_{xx} \\
-	    & u(-\pi, t) = u(\pi, t),\quad \forall t \in [0, T]
+	    & u(-\pi, t) = u(\pi, t),\quad \forall t \in [0, T], \quad \nu = 0.01
 	\end{split}
@@
 }
+
+\only<2>{
+\small
+ICs and Collocations:
+\begin{columns}[T,onlytextwidth]
+\column{0.3\textwidth}
+Harmonic:
+\begin{equation*}
+	\begin{split}
+		& u(x) = \sum_{k=1}^{30}a_k \sin(2\pi x/k) \\ & + b_k \cos(2\pi x/k) \\
+		& a_k = \alpha_ks_k \\
+		& \alpha_k \sim \mathcal{U}(0, 1) \\
+		& s_k \sim \mathcal{B}e(0.5)
+	\end{split}
+\end{equation*}
+
+\column{0.3\textwidth}
+Bell-curve:
+\begin{equation*}
+	\begin{split}
+		& u(x) = \frac{1}{\sqrt{2\pi \sigma^2}}\exp^{\frac{-(x-x_0)^2}{2\sigma^2}} \\
+		& \sigma \sim \mathcal{U}(0, 3) \\
+		& x_0 \sim \mathcal{U}(-\pi, \pi) 
+	\end{split}
+\end{equation*}
+
+\column{0.3\textwidth}
+Bumps:
+	\begin{equation*}
+		\begin{split}
+			& u(x) = \frac{a}{1+\exp^{-k(x-x_0)}} \\ & - \frac{a}{1 + \exp^{-k(x-x_1)}} \\
+			& k \sim \mathcal{U}(5, 20) \\
+			& x_0, x_1 \sim \mathcal{U}(-\pi, \pi) 
+		\end{split}
+	\end{equation*}
+\end{columns}
+}
+
 \uncover<3->{
@@
 	\includegraphics[width=\textwidth]{Figures/spi_math.png}
+	\uncover<2->{
+		\centering
+		\textit{Single-Pixel Image Recovery}: find $\textbf{x}$ using $\textbf{y}$ and $\textbf{A}$.
+
+		\vspace{1em}
+
+		\textbf{SPF rate}, \% = Number of samples per frame / number of pixels per frame * 100
+	}
 \end{frame}
@@
 	\uncover<1->{
-	When we capture 32 samples per frame:
+	Recovery with SPF=25\%:
 	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_32.pdf}
@@
 	\uncover<2->{
-	When we capture 2 samples per frame:
+	Recovery with SPF=6.25\%:
+	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_8.pdf}
+	}
+	\uncover<3->{
+	Recovery with SPF=1.5\%:
 	\includegraphics[width=\textwidth]{Figures/cs_burgers_comparison_2.pdf}
@@
 	\includegraphics[width=\textwidth]{Figures/cs_burgers_psnr.png}
+	\vspace{2em}
+	Metrics:
+	\begin{itemize}
+		\item \textbf{PSNR}: Peak Signal-to-Noise Ratio 
+		\item \textbf{RMSE}: Residual Mean Square Error
+		\item \textbf{SSIM}: Structural Similarity Index Measure
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}[label=again]{Results: Kolmogorov Flow}
+	A 2D velocity field $\bd{u}(x, y, t)$:
+
+	\begin{align}
+	\label{eq:kolmogorov_flow_definition}
+	& \partial_t \bd{u} + \bd{u} \cdot \nabla \bd{u} = -\nabla p + \nu\nabla^2\bd{u} + f \\
+	& \nabla \cdot \bd{u} = 0
+	\end{align}
+	where 
+	\begin{itemize}
+		\item $p$ is a 2D pressure field
+		\item $f = \alpha\sin(ky)x$ - driving force with amplitude $A$ and wavenumber $k$ ($k=4$)
+		\item $\nu = 1/\text{Re}$ is the non-dimensional viscosity. Re=40
+	\end{itemize}
+	\vspace{1em}
+	Reconstruction with SPF rate = $0.18$\% (8 SPI samples per a frame of $66\times 66$ pixels):  
 	\animategraphics[loop,width=\textwidth]{5}{Figures/kolm_example_anim_}{0}{19}
@@
 \begin{frame}{Results: Gas Monitoring}
-
+	\vspace{2em}
+	Gas monitoring problem:
+	\begin{itemize}
+		\item 494 240$\times$320 video recordings, 10 steps each
+		\item 512 samples per frame, SPF rate of 0.67\%
+		\item Background noise
+	\end{itemize}
+	\centering
+	\animategraphics[loop,width=0.6\textwidth]{5}{Figures/gas_frames_}{1}{9}
 \end{frame}
 
-\begin{frame}{Conclusion and Acknowledgments}
+\begin{frame}{Thank You!}
+\includegraphics[height=35pt]{figures/sasha.png}
+\includegraphics[height=35pt]{figures/jim.jpg}
+\includegraphics[height=35pt]{figures/peng.jpg}
+\includegraphics[height=35pt]{figures/hassan.jpg}
+\includegraphics[height=35pt]{figures/saleh.jpeg}
+\includegraphics[height=35pt]{figures/yuying.jpeg}
+\includegraphics[height=35pt]{figures/nathan.jpg}
+\includegraphics[height=35pt]{figures/steve.jpeg}
+\includegraphics[height=35pt]{figures/josh.jpg}
 
-\end{frame}
+\vspace{1em}
+\textbf{Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}\\
+{\small
+Together with Aleksandr Aravkin, James Burke, Damian Santomauro, and Peng Zheng
+\begin{itemize}
+	\item ``A Relaxation Approach to Feature Selection for Linear Mixed Effects Models''\\ \textit{submitted to Journal of Computational and Graphical Statistics (JCGS)}
+	\item ``Analysis of Relaxation Methods for Feature Selection in Mixed Effects Models'' \\ \textit{submitted to Journal of Computational Optimization and Applications (COAP)}
+	\item ``\texttt{pysr3}: A Python Package for Spare Relaxed Regularized Regression'' \\ \textit{Journal of Open-Source Software (JOSS), 2023, ICCOPT 2022, SIAM OPT 2023}
+\end{itemize}
+}
 
-\appendix
-\section{Appendix: $\mathcal{MSR}3$}
-\begin{frame}{Designing an Algorithm}
-	$G_{\nu,\eta}$ encodes both gradient of a Lagrangian (lines 1-2) and the complementarity condition (line 3):
-	\eq{
-		G_{\nu,\eta}((\beta,\gamma,v),(\tbeta,\tgamma)) := \begin{bmatrix}
-			\nabla_\beta \LL(\beta, \gamma) + \eta(\beta-\tbeta) \\
-			\nabla_\gamma \LL(\beta, \gamma) + \eta(\gamma-\tgamma) - v\\
-			v \bigodot \gamma - \mu\textbf{1}
-		\end{bmatrix}
+\vspace{1em}
+\textbf{PINODE: Physics-Informed Neural ODEs}\\
+{\small
+Together with Hassan Mansour, Saleh Nabi, and Yuying Liu
+\begin{itemize}
+	\item ``Physics-Informed Koopman Network'' \\
+	\textit{SIAM ADS 2023, arXiv:2211.09419}
+	\item ``Physics-Informed Neural ODE: Embedding Physics into Models using Collocation Points''\\
+	\textit{Submitted to Nature Special Issue on Physics-Informed Machine Learning}
+\end{itemize}
 }
-	We apply Newton method to $G$ while geometrically decreasing $\mu$. \\
-\textbf{Lemma:} For every $(\mu,\eta) \in \R_+\times\R_{++}$,
-\eq{
-	&(\hat\beta,\hat\gamma) = \argmin_{(\beta,\gamma)}\LL_{\eta,\mu}((\beta, \gamma),(\tbeta, \tgamma)) \\
-	& \iff \\
-	& \exists \hat{v} \in \R_{+}^q \text{ s.t. } G_{\nu,\eta}((\beta,\gamma,\hat{v}),(\tbeta,\tgamma)) = 0
+
+\vspace{1em}
+\textbf{Single-Pixel Imaging with Reduced-Order Models}\\
+{\small
+Together with J. Nathan Kutz, Steven Brunton, Joshua Rapp, Hassan Mansour, and Saleh Nabi
+\begin{itemize}
+	\item ``Single pixel imaging of spatio-temporal flows using differentiable latent dynamics'' \textit{in preparation}
+\end{itemize}
 }
-If $\mu > 0$, then $\hat{v} = -\nabla\phi_\mu(\hat{\gamma})$, and if $\mu = 0$, then $\hat{v}$ is the unique KKT multiplier associated with the constraint $0 \leq \gamma$.
-\end{frame}
-
-\begin{frame}{$\ouralgo$-fast Algorithm}
-\input{../extras/08Pseudocode}
 \end{frame}
@@
 
+\appendix
+\section{Appendix: $\mathcal{MSR}3$}
+
 \end{document}

### 2023-06-06 presentation/phd_defense_pres.tex
Final Version of the pres (after Alanna's edits)

@@
 % \usetheme[progressbar=frametitle]{metropolis}
+\addtobeamertemplate{navigation symbols}{}{%
+    \usebeamerfont{footline}%
+    \usebeamercolor[fg]{footline}%
+    \hspace{1em}%
+    \insertframenumber/\inserttotalframenumber
+}
+
 \usepackage{appendixnumberbeamer}
@@
 %\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
-\title{Optimization Methods for Parameter Identifications \\ in Settings with Only Partial Knowledge}
+\title{Optimization Methods for Parameter Identification \\ in Settings with Only Partial Knowledge}
 \subtitle{Ph.D. Defense}
@@
 		\item   $u_i \in \R^q$ -- random effects
-		\item   $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects, often $\Gamma = \diag{(\gamma)}$
+		\item   $\Gamma \in \R^{q \times q}$ -- covariance matrix of random effects, often $\Gamma = \diag{\gamma}$
 		\item   $\Lambda_i \in R^{n_i \times n_i}$ -- covariance matrix for observation noise
@@
 \eq{
-	\prox_{\alpha R + \delta_{\CC}}(\tbeta, \tgamma) & := \argmin_{(\beta, \gamma) \in \CC} R(\beta, \gamma) + \frac{1}{2\alpha}\|(\beta, \gamma) - (\tbeta, \tgamma)\|_2^2, \\ & \text{ where } \CC := \R^p \times R^q_+ \\
+	\prox_{\alpha R + \delta_{\CC}}(\tbeta, \tgamma) & := \argmin_{(\beta, \gamma) \in \CC} R(\beta, \gamma) + \frac{1}{2\alpha}\|(\beta, \gamma) - (\tbeta, \tgamma)\|_2^2, \\ & \text{ where } \CC := \R^p \times \mathbb{R}^q_+ \\
 }
@@
 \eq{
-	x := [\beta, \gamma], \quad \CC := \R^p \times R^q_+
+	x := [\beta, \gamma], \quad \CC := \R^p \times \R^q_+
 }
@@
 \vspace{1em}
-Basic Assumptions for the PGD Algorithm (Theorem 10.15 from \footcite*{AB17}):
+Basic Assumptions for the PGD Algorithm\footcite*[Theorem 10.15]{AB17}:
 \begin{enumerate}
@@
 	\item<3-> \textbf{Model Discovery Based:} use a library of terms to discover simple equations: SINDy\footcite{champion2019data}, Genetic Programming\footcite{schmidt2009distilling}
-	\item<4-> \textbf{Equations Based:} incorporate first-principle models to aid training of networks: PINNs\footcite{raissi2017physics}, UDEs\footcite{rackauckas2020udes}, \textit{PINODE}
+	\item<4-> \textbf{Equation Based:} incorporate first-principle models to aid training of networks: PINNs\footcite{raissi2017physics}, UDEs\footcite{rackauckas2020udes}, \textit{PINODE}
 \end{enumerate}
@@
 \only<7>{\includegraphics[width=\textwidth]{Figures/collocations_7.png}}
+\uncover<6->{\textbf{Collocations:} a set of pairs $(\tilde{x}_i, \dot{\tilde{x}}_i)$, where $\tilde{x}_i$ are s.t. $\dot{\tilde{x}}_i$ are cheap to evaluate.}
 \end{frame}
@@
 	\includegraphics[width=\textwidth]{Figures/cs_burgers_psnr.png}
-	\vspace{2em}
+
+	\vspace{1em}
+
+	\hspace{2em} Higher is better \hspace{2.6em} Lower is better \hspace{2.6em} Higher is better
+
+	\vspace{1em}
 	Metrics:
@@
 	\includegraphics[width=\textwidth]{Figures/cs_kolm_aggregate.png}
+
+	 \hspace{5em} Lower is better \hspace{9em} Higher is better
 \end{frame}

### 2023-06-08 appendix/acknowledgements.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

@@
-I am extremely grateful to my advisor, Sasha Aravkin, for providing support, guidance, and enthusiasm during my work on this project in the Department of Applied Mathematics and in Institute for Health Metrics and Evaluation.\\
-I would also like to thank Jim Burke for his detailed review of my work which would hopefully lead to fruitful collaboration. \\ 
-Last, but not least: I am grateful to Damian Santomauro, who introduced me to his work on the consequences of bullying, provided with a dataset, and with a valuable feedback on the performance of the proposed method.
-\\
+I would like to thank my research advisors -- Dr.
+Aleksandr Aravkin, and Dr.
+Nathan Kutz, for their excellent leadership, continuous mentorship, and support in my journey as a Ph.D.
+student.
+I would also like to thank my collaborators -- Dr.
+James V.
+Burke, and Dr.
+Steven Brunton for their outstanding contributions to this work.
+I am grateful to Dr.
+Hassan Mansour and Dr.
+Saleh Nabi -- my colleagues and collaborators from Mitsubishi Electrics Research Labs who guided me to success throughout my internship in Spring 2023 and beyond.
+I wholeheartedly thank Biraj Pandey, Ryan Vogt, and Natalie Wellen -- my fellow Ph.D.
+students and dear friends from the Department of Applied Mathematics at UW: you turned what could have been a daunting and isolating journey into a wonderful experience full of happy memories.
+And even more so I would like to thank my family: my mom Svetlana Sholokhova, my grandmother Raisa Zakirova, and my wife Alanna Sholokhova.
+You have been surrounding me with unbelievable amounts of support and love, from close and afar, and I would have never succeeded in this journey without you.

### 2023-06-08 conclusion.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

@@
 In this work, I gave two examples of how partial knowledge can be utilized to improve performance of machine learning models.
-First, in Section~\ref{sec:MSR3}, we showed how complex priors can be embedded into a linear mixed-effects model via a suitable proximal operator, and how the resulting algorithm can be implemented using a proximal gradient descent.
-We note that a complex combination of priors often yields ill-conditioned problems, and we propose MSR3 relaxation to address it.
-We also prove that a solution of such relaxation approaches the solution of the original problem as the relaxation tightens.
-Next, we showed how partial knowledge of physics can be embedded into deep-learning-based reduced-order models.
-Namely, we show how to project known physical equations that govern the behavior of the system in an observable space into a chosen latent space.
+First, in Section~\ref{sec:MSR3}, it was shown how complex priors can be embedded into a linear mixed-effects model via a suitable proximal operator, and how the resulting algorithm can be implemented using a proximal gradient descent.
+Many real-world problems are ill-conditioned and the MSR3 relaxation was developed to address it.
+Next, it was shown how partial knowledge of physics can be embedded into deep-learning-based reduced-order models.
+Namely, Section~\ref{sec:collocations_conditions} shows how to project known physical equations that govern the behavior of the system in an observable space into a chosen latent space.
 In that space one can require that the gradient field match the projected true dynamics.
 We show that this methodology leads to identification of more stable latent manifolds, and using such manifolds for forecasting leads to more accurate predictions and less sensitivity to noise.
-We also showed how the reduced-order models described above can be used for compressive sensing.
+We also showed how the reduced-order models described above can be used for single-pixel imaging.
 We demonstrated that they can accurately reconstruct the dynamics from partial observations, significantly surpassing state-of-the-art compressive-sensing methods.
@@
 
-\paragraph{Perspective Research Directions} We would like to note many promising directions for future exploration.
+I would like to note many promising directions for future exploration.
 One such direction is extending sparse relaxed regularized regression (SR3) to generalized linear and generalized mixed-effects models.
@@
 It implies that one would have three nested levels of optimization when they apply an SR3 relaxation on top of such likelihood: evaluating the likelihood, evaluating a value function, and optimizing over the value function.
-Re-using the recipe that we developed in Section~\ref{sec:MSR3} could help to surmount such obstacle.
+Re-using the recipe that was developed in Section~\ref{sec:MSR3} could help to surmount such obstacle.
 For example, Algorithm~\ref{alg:pgd_for_value} had two nested optimization methods: for evaluating the value function and for optimizing over it.
-However, we showed that one can ``blend'' both levels of optimization and achieve superior performance results.
+However, it was shown that one can ``blend'' both levels of optimization and achieve superior performance results.
 The same methodology of ``blending'' could possibly be applied to more than two nested optimization loops to yield algorithms that, in certain scenarios, work faster than non-nested optimization methods.
-
 Our development of physics-informed reduced-order models also invites future work in several promising directions.
 First, a more systematic way of choosing collocation points would greatly benefit the method.
-In Section~\ref{sec:collocations_conditions}, we define collocation points and provide criteria for a good choice of a family of colocation points.
-However, we ultimately leave the reader without a constructive algorithm for identifying candidate families of collocation points suitable for their problem.
+In Section~\ref{sec:collocations_conditions}, collocations are defined and criteria for a good choice of a family of colocation points are provided.
+However, the reader is left without a constructive algorithm for identifying candidate families of collocation points suitable for their problem.
 At the same time, certain differential equations admit families of basis functions that can span their solutions.
@@
 Bridging the gap between this new methodology and those classic results would undoubtedly yield a fruitful line of work.
-Finally, one could utilize PINODE models not only in compressive sensing application, but in any application where one needs to quickly search over a large space of possible simulations.
+Finally, one could utilize PINODE models not only in SPI applications, but in any application where one needs to quickly search over a large space of possible simulations.
 These examples include medical imaging, model discovery, online control, fast simulations, and many others.
-
-\paragraph{Acknowledgements} I would like to thank ... (TBD)

### 2023-06-08 cs.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

--- 2023-04-06 cs.tex
@@
-\section{Application to Compressive Sensing}
+\section{Application to Single-Pixel Imaging}
+\label{ch:cs}
 \subsection{Introduction}
-\paragraph{Prior Work}
-\subsection{Method}
-\paragraph{Compressive Sensing} In many applications we can not observe $x(t)$ in real time.
-Instead, we observe $p$ detectors, and each detector provides a linear combination of a small number of coordinates of $x(t)$ at a time:
+% SPI definition and main stages
+Single-pixel imaging (SPI) is a relatively recent imaging technique that replaces the classic two dimensional sensor array with a single photodetector~\citep{bian2018experimental}.
+An SPI camera modulates the incoming light from a scene using a light modulator, such as a diffuser~\citep{guo2016multilayer} or a digital micro-mirror device (DMD)~\citep{pittman1995optical, sampsell1993overview}, and focuses the resulting correlated light patterns onto a photodiode.
+The configuration of the modulator and the recorded intensity of light are then processed to reconstruct an image of the target scene.
+
+% advantages of SPI
+SPI owes its popularity to multiple advantages it achieves over traditional imaging approaches that employ 2D sensor arrays.
+First, since all reflected light is focused onto a single detector, SPI has better sensitivity under dim lighting conditions and enjoys a broader spectral range\citep{Edgar2019}.
+Second, SPI moves the cost of acquisition and processing from the encoding/compression stage to the decoding/reconstruction stage.
+For example, a DMD array orients its mirrors to reflect light either towards the detector (+1) or away from it (0).
+Importantly, these orientations do not depend on the scene that is being acquired to ensure proper reconstruction: \citep{baraniuk2008simple}~showed that DMD patterns can be drawn as independent identically distributed (i.i.d.) samples from a uniform Bernoulli distribution.
+In contrast, sample-then-compress approaches\citep{skodras2001jpeg} require an expensive projection of multiple single-pixel snapshots onto a chosen basis, such as Fourier or top-K singular vectors, during their encoding step.
+Doing the heavy-lifting job during decoding is often preferred in applications where sensors have to be small and low-power, whereas decoding happens offline on high-performance computers.
+The simplicity of encoding keeps the computational requirements of SPI sensors (and thus their price) low, inviting their application in multispectral imaging \citep{bian2016multispectral,li2017efficient}, optical encryption~\citep{chen2013ghost}, remote sensing~\citep{zhao2012ghost}, object tracking~\citep{li2014ghost}, and many other fields~\citep{sun20133d,zhang2010correlated,cheng2009ghost}.
+
+% introducing CS
+Compressive Sensing (CS)~\citep{donoho2006compressed, candes2006compressive, baraniuk2014introduction, baraniuk2008simple} emerged in the mid-2000s as an alternative signal acquisition regime to classical Shannon-Nyquist sampling.
+In the classical regime, a band-limited signal is sampled at $N$ specific points in time (and/or space) so that the signal can be reconstructed from the acquired samples using a simple Sinc interpolation.
+Compressive sensing, on the other hand, proposes to acquire the signal using a small number of inner products between the signal and general non-adaptive sampling vectors.
+If the signal to be acquired has a $K$-sparse, or more generally, parsimonious, representation some transform domain, then the signal can be guaranteed to be recovered using as few as $\mathcal{O}(K\log(N/K))$ non-adaptive measurements, which is much smaller than the $N$ samples that are required by sample-then-compress approaches such as JPEG and JPEG-2000~\citep{skodras2001jpeg}.
+The development of compressive sensing theory and efficient reconstruction algorithms has further promoted the deployment of this technique in SPI~\citep{duarte2008single}.
+
+% Compressive Sensing (CS) is an approach to SPI which assumes that a signal is spare in a certain basis.
+Under this assumption a small collection of non-adaptive linear measurements can contain enough information for reconstructing the signal~\citep{duarte2008single, candes2006compressive}.
+Namely, when an original signal made of $N$ pixels has a $K$-sparse representation in a certain basis, the basis coefficients can be recovered using only $\mathcal{O}(K\log(N/K))$ which is much smaller than $N$ required by sample-then-compress approaches such as JPEG and JPEG-2000~\citep{skodras2001jpeg}.
+
+% Recent progress on deep SPI
+Since every SPI measurement contains highly-compressed information about the observed frame, one needs to sample a large number of such measurements to allow for faithful reconstruction.
+Thus, many recent works propose approaches to reduce the number of required samples per frame (SPF).
+These approaches can be roughly grouped in two buckets: signal-dependent control of the mirrors, and more sophisticated reconstruction techniques.
+The former ensures that each single-pixel measurement contains as much information as possible by focusing the mirrors on the selected regions of the frame~\citep{zhang2017fast,sun2017russian,xu20181000}.
+Although highly effective, such approach can not be applied to detectors that do not offer fine-grained control of their mirrors.
+The latter group focuses on creating more efficient optimization algorithms for the decoding step~\citep{katz2009compressive}.
+It includes non-iterative algorithms such as differential ghost imaging (DGI) \citep{gong2010method}, as well iterative methods such as total variation (TV)~\citep{suo2016signal} regularized reconstruction.
+Notably, recent approaches adopted deep learning (DL) for SPI reconstruction and showed that using deep neural networks can dramatically reduce the sampling ratio and offer near-real-time performance \citep{lyu2017deep,higham2018deep,wang2019learning,wang2022single}.
+However, most of these methods use multiple samples for single-frame reconstruction and do not yield satisfactory reconstruction of multiple sequential frames when the number of samples per frame is small even if the total number of samples collected for the full video sequence is large.
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=\textwidth]{figures/cs_ga_current.png}
+    \caption{\label{fig:cs_abstract}We propose a compressive-sensing algorithm that reconstructs observation by identifying their trajectories in a latent space of a Reduced-Order Model (ROM).
+The model provides strong inductive bias and reduces data intake requirements for compressive-sensing hardware.}
+\end{figure*}
+
+% Our place among CS literature
+This work aims to enable high quality reconstruction of videos from a small number of samples per frame.
+Our approach falls into the second group of decoding-centered approaches which aid SPI reconstruction with deep learning.
+In particular, we show how a Reduced-Order Model (ROM) can be used to regularize SPI reconstruction of a video of a physical phenomenon, even if all measurements are taken at different moments in time, i.e.
+when the number of samples per frame is one.
+We achieve that by pre-training a ROM to compress simulations into a latent space and calculate the system's evolution within the latent space using neural ordinary differential equations (Neural ODEs).
+The reconstruction procedure minimizes the measurement mismatch between the samples recorded by the SPI detector and the synthesized measurements obtained by sampling the reconstructed trajectory using a measurement operator $A$.
+Effectively, we solve an ODE-regularized SPI reconstruction in an efficient way using the adjoint-sensitivity method to obtain the necessary derivatives~\citep{chen2018neuralode}.
+Such ODE-based regularization provides a strong signal prior and allows for the reduction in the number of samples per frame required for successful reconstruction below the requirements of current state-of-the-art methods.
+We introduce the methodology in Section~\ref{sec:cs_methods} and study its performance on three examples: Burger's equation, Kolmogorov flow, and methane leaks videos from GasVid dataset \citep{wang2020machine}.
+
+\subsection{Methods}
+\label{sec:cs_methods}
+\paragraph{Single-Pixel Imaging}
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.5\textwidth]{figures/SPI_setup.pdf}
+\caption{A schematics for Single-Pixel Imaging setup with Digital Micro-mirror Device (DMD) \label{fig:SPI}}
+\end{figure}
+
+Let $x(t) \in \XX \subseteq \mathbb{R}^n$ be a representation of a high-dimensional spatio-temporal system, e.g.
+a video recording of a scene that we observe.
+In many applications we can not record $x(t)$ in real time due, for example, to monetary costs that a necessary equipment would impose.
+Instead, we collect records from $p$ detectors, where each detector provides a linear combination of coordinates of $x(t)$ at a time:
 \begin{equation}
     \label{eq:cs_definition}
-    y(t) = A_t x(t), \quad A_t \in \mathbb{R}^{p\times n}
+    y(t_i) = A_i x(t_i), \quad A_i \in \mathbb{R}^{p\times n}, \quad t_i \in [0, T],\quad i \in 0 \dots K
 \end{equation}
 
-In particular, we consider a single pixel camera setup where for every time instance $t$, $p$ acquisitions $y(t)$ are obtained by a high sampling rate photo-detector using the projection matrix $A_t$.
-The rows of the matrix $A_t$ correspond to a binary mask pattern that can be encoded using a digital micro-mirror device (DMD) where the incoming light from $x(t)$ is reflected from the DMD array and focused onto the photo-detector.
+In particular, we consider a single pixel camera setup where for every time instance $t_i$, a vector of $p$ acquisitions $y_i \in \mathbb{R}^p$ are obtained by a high sampling rate photo-detector using the projection matrix $A_i$.
+The rows of the matrix $A_i$ correspond to a binary mask pattern that can be encoded using a digital micro-mirror device (DMD, ~\citep{pittman1995optical, sampsell1993overview}).
+In this setup, the mirrors of the DMD array represent pixels of the future reconstructed image and slopes of the mirrors represent the weights.
+The light from a scene lands on a DMD which slopes its every mirror to reflect light either towards the detector (+1) or away from it (0) according to a uniform Bernoulli variable.
+Next, the (+1) pixels go through a double-convex lens that focuses it on a signal photon detector.
+Finally, the detector integrates the signal and provides the final observation as its output voltage, which is later digitized by an A/D converter~\citep{duarte2008single}.
+The slopes of each mirror are also recorded in the detector's memory.
 Figure~\ref{fig:SPI} illustrates an example of the single pixel imaging setup where a gas plume is imaged using a DMD array and a medium infra-red (MIR) photo-detector.
-\begin{figure}
-\centering
-\includegraphics[width = 0.5\textwidth]{figures/SPI_setup.pdf}\label{fig:SPI}
-\end{figure}
 
-However, it is often possible to have access to a complete state $x(t)$ at the moment of model training.
-Thus, one can develop a ROM $(\psi_\theta, \phi_\theta, h_\theta)$ using the full state $x(t)$, and then utilize this ROM for real-time compressive sensing applications.
+Multiple techniques attempt reconstruct $x(t)$ directly by inverting the relation~\ref{eq:cs_definition} under mild regularity priors; examples include Differential Ghost Imaging (DGI, \citep{ferri2010differential}), Total Variation Regularization (TVR, \citep{bian2018experimental}), and Fourier Domain Regularization (FDRI, \citep{pastuszczak2021differential}).
+Alternatively, one can recover $x(t)$ by recovering its projection $z(t) \in \ZZ \subseteq \mathbb{R}^m$ on a low-dimensional manifold ($m \ll n$).
+Namely, if we know  an invertible mapping $\psi(z): \ZZ \to \XX$ then we can trade a problem of recovery of a high-dimensional vector $x$ using $p$ linear observations to recovery of a low-dimensional vector $z$ using $p$ non-linear observations:
 
-\paragraph{Training Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses.
+\begin{equation}
+    \label{eq:cs_with_ae}
+    y(t) = A_t\psi(z(t))
+\end{equation}
+
+ We call the space $\XX$ an observable space, $\ZZ$ a latent space, and the mapping $\psi(z)$ a true decoder.
+Often $\psi(z)$ is not known, in which case an approximation $\psi_\theta(z)$ is trained using a dataset of collected images $\{x_1, \dots, x_N\}$, where $\theta$ denotes the parameters of the model, e.
+g.
+the weights of an autoencoder network.
+However, if $y(t)$ is insufficient for a unique reconstruction of $x(t)$ then it will remain insufficient for reconstructing $z(t)$.
+Thus, introducing additional structural sparsity or priors is required to make the autoencoder-based reconstruction work, such as joint training of the sampling matrices and the decoder by \citep{wang2022single}.
+In this work we take a different route and supplement the encoder $\psi(z)$ with knowledge of the dynamics of $z(t)$ on the manifold over time.
+
+\paragraph{Reduced-Order Model with Non-Linear Latent Dynamics}
+Let $x(t)$ be modelled as an autonomous dynamical system on a finite space $\XX \subseteq \mathbb{R}^n$
+
+\begin{equation}
+    \label{eq:cs_generic_stationary_ode}
+        \ddt\bd{x}(t) = \bd{f}(\bd{x}(t))
+\end{equation}
+
+In real-world applications it is often expensive to use the relationship~(\ref{eq:generic_stationary_ode}) for predicting the behaviour of the system even when $f(x)$ is known because $x(t)$ can be very high-dimensional.
+For example, if $x(t)$ is a flow of 2D liquid then $\bd{f}(x)$ may be a discretization of Navier-Stokes equations, in which case running a simulation~\ref{eq:cs_generic_stationary_ode} may take months on a cluster.
+However, a variety of works provided both theoretical (\citep{holmes2012turbulence}) and practical (\citep{noack2011reduced,chen2021discovering}) evidence that many physical systems evolve on a lower-dimensional manifold $\ZZ$.
+In that space, the dynamics evolve according to a (generally unknown) function  $\bd{h}(\bd{z})$:
+
+\begin{equation}
+        \ddt\bd{z}(t) = \bd{h}(\bd{z}(t))
+\end{equation}
+
+ Thus, one can predict the dynamics of the system $\bd{x}$ at a future time $T$ by projecting the initial condition $\bd{x}(0)$ into the latent space, performing an integration there, and mapping the resulting trajectory back to the observable space:
+\begin{equation}
+\label{eq:cs_integration_in_latent_space}
+\begin{split}
+    \bd{z}(0) & = \psi^{-1}(\bd{x}(0)) \\
+    \bd{z}(T) & = \bd{z}(0) + \int_{0}^T\bd{h}(\bd{z}(t))dt \\
+    \bd{x}(T) & = \psi(\bd{z}(T))
+\end{split}
+\end{equation}
+
+When $m \ll n$ we refer to the triplet $(\psi, \psi^{-1}, \bd{h})$ as a Reduced-Order Model (ROM) of $\bd{f}$.
+It is often the case that for a given system $\bd{f}$ there exists no ROM $(\psi, \psi^{-1}, \bd{h})$ such that the relation~(\ref{eq:cs_integration_in_latent_space}) holds exactly.
+In this case, we seek an \textit{approximation} ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ that minimizes the difference between the data $x(t)$ and the prediction $\hat{x}(t)$ over a chosen class of models $(\psi_\theta, \phi_\theta, h_\theta)$ parameterized by $\theta$.
+
+\paragraph{Architecture} In this work we model $\psi$, $\psi^{-1}$, and $\bd{h}$ with neural networks $\psi_\theta$, $\phi_\theta$, and $h_\theta$, respectively.
+Specifically, the pair ($\psi$, $\psi^{-1}$) is modelled with an auto-encoder $(\psi_\theta, \phi_\theta)$, and $\bd{h}$ is modelled with a fully-connected network $h_\theta$, as illustrated in the first frame of Figure~\ref{fig:cs_abstract}.
+All considered models share this architecture; however, the exact architectures of the functions $\psi_\theta$, $\phi_\theta$, and $h_\theta$ are problem-dependent and discussed more closely in the related sections.
+
+\paragraph{Training Loss} Similar to prior works by \citep{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{training loss} $\LL$ as a sum of reconstruction and prediction losses.
 The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data.
-Formally, for a given set of trajectories $\bd{x}_i$, $i \in [1 \dots k]$, where each trajectory $\bd{x}_i \in \mathbb{R}^{n \times p}$ is a set of $p$ snapshots that correspond to the recorded states of the system for $p$ time-steps, $t_j$, $j \in [1, \dots, p]$, the loss function $\LL^{data}_\theta$ is defined as:
+Formally, for a given set of trajectories $\bd{x}_i$, $i \in [1 \dots N]$, where each trajectory $\bd{x}_i \in \mathbb{R}^{n \times K}$ is a set of $p$ snapshots of the system for each time-step, for $K$ time-steps, $t_i$, $i \in [0, \dots, K]$, the loss function $\LL^{data}_\theta$ is defined as:
 
 \begin{align}
-    \label{eq:loss_training}
-    \LL^{training}(\theta) & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
-     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
+    \label{eq:cs_loss_data_driven}
+    \LL(\theta) & = \sum_{i = 1}^N \left[\sum_{j=0}^K\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
+     & + \left.\sum_{j=1}^K \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_0)) + \int_{t_0}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
 \end{align}
@@
 
-To obtain a ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$, we minimize the loss above:
+To obtain a ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$, we minimize the loss above.
+We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:cs_loss_data_driven}).
+This part is handled by \texttt{torchdiffeq}~\citep{chen2018neural} library, which supports non-uniform time-frames within a batch}.
+To simplify the notation without loss of generality, in the rest of the work we assume that all trajectories were recorded over the same time-frame with an equal and uniform step-size.
+
+\paragraph{Recovery Loss} 
+We use the ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ above to regularize the latent-space dynamics of the single-pixel imaging reconstructions.
+Namely, we obtain the reconstruction $x^* = \psi(z^*)$ by minimizing the following loss with respect to $z = [z_0, \dots, z_K] \in \mathbb{R}^{m \times K}$.
 
 \begin{equation}
-    \label{eq:training_minimization}
-    \theta^* = \arg\min_{\theta} \LL^{training}(\theta)
-\end{equation}
-
-We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}).
-This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}.
-To simplify the notation without loss of generality, in the rest of the paper we assume that all trajectories were recorded over the same time-frame with an equal and uniform step-size.
-
-\paragraph{Reconstruction Loss} 
-We use the ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ above to forecast the dynamics based on partial observations in real time.
-Namely, instead of reconstructing $x(t)$ based on compressive-sensing observations $y(t)$ directly, we first reconstruct the latent dynamics $z(t)$ and then project it to the observable space using the decoder $\psi_{\theta^*}(z)$.
-
-\begin{align}
-    \label{eq:reconstruction_problem_differential}
-    \min_{\{z_t\}_{t=1, \dots, T}} & \frac{1}{2}\sum_{t=1}^T \left\|y_t - A\psi_{\theta^*}(z_t)\right\|_2^2 \\
-    \text{s.t. } & \dot{z} = h_{\theta^*}(z)
-\end{align}
-
-We integrate the constraint and write it in its Lagrangian form:
-\begin{align}
-\label{eq:reconstruction_problem_integral}
-    \min_{\{z_t\}_{t=1, \dots, T}} \LL_{\theta^*}^{recon}(z)
-\end{align}
-
-where
-
-\begin{equation}
-    \label{eq:reconstruction_loss}
-     \LL_{\theta^*}^{recon}(z) = \frac{1}{2}\sum_{t=1}^T \left\|y_t - A\psi_{\theta^*}(z_t)\right\|_2^2 + \frac{\lambda}{2} \sum_{t=1}^T \left\|z_{t-1} + \int_{t-1}^{t}h_{\theta^*}(z)dz - z_t\right\|_2^2
+    \label{eq:cs_reconstruction_loss}
+     \LL_{\theta^*}^{recon}(z) = \sum_{j=0}^K  \left\|y_j - A\psi_{\theta^*}(z_j)\right\|_2^2 + \lambda \left\|z_0
+     + \int_{t_0}^{t_j}h_{\theta^*}(z)dz - z_j\right\|_2^2
 \end{equation}
@@
 where the parameter $\lambda$ controls the degree on which the compressing sensing algorithm relies on the latent dynamics $h_{\theta^*}$ during the signal reconstruction phase.
-We minimize the loss~\ref{eq:reconstruction_loss} using a gradient-based technique, with the gradients obtained using automatic differentiation frameworks.
+We minimize the loss~\ref{eq:cs_reconstruction_loss} using a gradient-based technique, with the gradients obtained using automatic differentiation frameworks.
 
 \subsection{Experiments}
-\subsection{Discussion and Conclusion}
+
+\subsubsection{Burger's Equation}
+We study the performance of our framework on Burger's equation with $[-\pi, \pi]$-periodic boundary conditions:
+
+\begin{equation}
+\begin{split}
+\label{eq:cs_burgers_equation}
+    & u_t  + uu_x = \nu u_{xx} \\
+    & u(-\pi, t) = u(\pi, t),\quad \forall t \in [0, T]
+\end{split}
+\end{equation}
+where $u_t$, $u_x$, and $u_{xx}$ represent partial derivatives in time, the first, and second spatial derivatives, respectively.
+
+\paragraph{ROM Training} To obtain train dataset for the ROM we replicate the experimental setup from Section~\ref{sec:burger_eqn}.
+Namely, we generate 1024 trajectories on a discretized spacial domain $[-\pi,\,\pi]$ with 128 grid-points.
+To generate a diverse set of initial conditions we sum the first 10 harmonic terms with random coefficients:
+\begin{equation}
+    \label{eq:burger_initial_condition_sine}
+    \begin{split}
+        u(x, 0) = & \frac{1}{10}\sum_{k = 1}^{10} a_k\cos(kx) + b_k\sin((k+1)x)\\ & a_k, b_k \sim \mathcal{N}(0, 1)
+    \end{split}
+\end{equation}
+ We solve Equation~\ref{eq:cs_burgers_equation} for $t \in [0, 2]$ with $\Delta t = 0.1$ using a spectral solver by~\citep{trefethen2000spectral}.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figures/cs_burgers_psnr.png}
+    \caption{\label{fig:cs_burgers_spf} Median PSNR, RMSE, and SSIM with 25-27th percentile intervals over different number of samples per frame taken.}
+\end{figure}
+
+\begin{figure}[t]
+    \centering
+    	\includegraphics[width=\textwidth]{figures/cs_burgers_comparison_all.pdf}
+    \caption{\label{fig:cs_burgers_example_large_spf} Example reconstruction of a trajectory of Burgers' equation for three SPF rates: 25\%, 6\%, and 2\%.
+All algorithms complete the reconstruction faithfully when SPF is large.
+However, all algorithms except our proposed method, fail to recover the image due to the insufficient SPF rate ($2/128 \lesssim$ 2\%).
+In contrast, our proposed method achieves nearly the same reconstruction quality as in with SPF rate of 25\%.}
+\end{figure}
+
+\paragraph{Recovery}
+For the SPI Recovery phase, we generate 128 trajectories with ``bump'' initial conditions -- a smooth approximation of a bump with two opposing steeply-curved sigmoids:
+
+\begin{equation}
+    \label{eq:burger_initial_condition_bumps}
+    u(x, 0) = \frac{1}{1 + exp(-k(x-a))} - \frac{1}{1 + exp(-k(x-b))}
+\end{equation}
+where $a < b$ are sampled uniformly in $[-\pi, \pi]$ and $k = 20$.
+We choose this shape to ensure that the training and sensing trajectories are sufficiently different.
+We choose sensing matrices $A_t$ to be binary (0-1) matrices with each row having 64 non-zero components out of 128 sampled uniformly.
+
+We compare our approach (PINODE) with four alternatives: an Autoencoder-enhanced variation of Digital Ghost Imaging, \citep{ferri2010differential,gong2010method}(DGI+Conv.
+Decoder, \citep{wang2022single}) Total Variation Regularization  (TVR, \citep{bian2018experimental}), Fourier Domain Regularization (FDRI, \citep{czajkowski2018real}), and an approach which uses an autoencoder (AE, \citep{bora2017compressed}).
+We measure reconstruction accuracy with three commonly-used metrics: the residual mean-squared error (RMSE), the peak signal-to-noise ratio (PSNR), and the structural similarity index measure (SSIM).
+We evaluate both metrics for Burger's trajectories as 2D-images in spatial and in temporal domains.
+
+For two 2D images $a$ and $b$, PSNR is defined as the log-ratio of the maximal value from the true image:
+
+\[
+	\text{PSNR}(a, b) = 20\log\left(\frac{\max_{i,j}(|a[i,j]|)}{\|a - b\|_2}\right)
+\]
+
+SSIM is defined as a product of relative luminance $l(a,b)$, contrast $c(a,b)$, and structure $s(x,y)$:
+
+\[
+	\text{SSIM}(a, b) = l(a,b) \cdot c(a,b) \cdot s(a,b) = \left(\frac{2\mu_a\mu_b + c_1}{\mu_a^2 + \mu_b^2 + c_1}\right) \cdot \left(\frac{2\sigma_a\sigma_b + c_2}{\sigma_a^2 + \sigma_b^2 + c_2} \right) \cdot \left(\frac{\sigma_{ab} + c_2/2}{\sigma_a \sigma_b + c_2/2}\right)
+\]
+where, for a picture $x$, $\mu_x$ is the pixel sample mean, $\sigma_x$ is the standard deviation,  and $c_i$ are the constants which stabilize the division when the denominator approaches 0.
+Typically, $c_i$ are set to be proportional to the square of the dynamic range of the pixel values, e.g.
+$c_i = (k_i*2^{\text{bits per pixel}} - 1)^2$, with $k_1 = 0.01$ and $k_2 = 0.03$ being default choices.
+Finally, RMSE is defined as the mean residual squares for every state, averaged over the temporal axis:
+
+\[
+	\text{RMSE}(a,b) = \|a - b\|_F.
+\]
+
+\paragraph{Results} The results are presented in Figure~\ref{fig:cs_burgers_spf}.
+We plot median PSNR, RMSE, and SSIM over different number of samples per frame taken.
+We see that PINODE achieves 70\% higher peak signal-to-noise-ratio (PSNR) and 4 times higher structural similarity index measure (SSIM)  relative other algorithms when SPF is low.
+We also see that the DGI and TVR approaches do not reconstruct images faithfully until SPF approaches the number of pixels in a snapshot; Figure~\ref{fig:cs_burgers_example_large_spf} provides an example of high-SPF reconstruction.
+In contrast, PINODE is able to reconstruct most trajectories with high accuracy using a very small number of samples per frame e.g.
+two samples.
+It happens because the reduced-order model of PINODE serves as a strong prior: the model reconstructs the trajectory as a whole instead of reconstructing every snapshot separately, as all other algorithms do in this comparison.
+It allows PINODE to borrow strength across snapshots, even when SPF per each snapshot is small.
+
+\subsubsection{Kolmogorov Flow}
+
+\label{sec:kolmogorov_flow}
+We study the performance of our framework on Kolmogorov-like flow first-described by~\citep{dovzhenko1981generation} and later analyzed by \citep{tithof2017bifurcations,suri2017forecasting}.
+The following equation describes the behavior of a 2D velocity field $\bd{u}(x, y, t)$:
+
+\begin{align}
+\label{eq:kolmogorov_flow_definition}
+	& \partial_t \bd{u} + \bd{u} \cdot \nabla \bd{u} = -\nabla p + \nu\nabla^2\bd{u} + f \\
+	& \nabla \cdot \bd{u} = 0
+\end{align}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figures/cs_kolm_aggregate_all.png}
+    \caption{Caption}
+    \label{fig:cs_kolm_agg}
+\end{figure}
+
+where $p$ is a 2D pressure field, $f = \alpha\sin(ky)x$ represents the driving force with amplitude $A$ and wavenumber $k$, and $\nu = 1/\text{Re}$ is the non-dimensional viscosity equal to inverse of the Reynolds number.
+In all our experiments, $k = 4$ and $\text{Re} = 40$; we chose this setup to match the one from~\citep{wan2018data}, who identified that this combination of hyper-parameters leads to the occurrence of extreme instability events, making the prediction of such flow truly challenging.
+
+To train a ROM which predicts $u(x, y, t)$ we generate 8192 data trajectories as solutions of~\eqref{eq:kolmogorov_flow_definition} using a spectral solver\footnote{\href{https://github.com/zhong1wan/data-assisted/blob/master/Kolmogorov/kol2d_odd.py}{https://github.com/zhong1wan/data-assisted/blob/master/Kolmogorov/kol2d\_odd.py}} from~\citep{wan2018data}.
+Each solution was simulated from t=0 to t=100 to get past the transient regime and then recorded from t=100 to t=110, with the time-step $dt=0.5$.
+As in Section~\ref{sec:method}, the ROM consists of an encoder $\phi(u)$, a decoder $\psi(z)$, and the latent dynamics $h(z)$.
+It was trained by minimizing the data-driven loss~\eqref{eq:loss_data_driven} until the prediction RMSE on a holdout set stopped improving, which took about 60 epochs.
+
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=\textwidth]{figures/cs_kolm_example.pdf}
+	\caption{\label{fig:kolmogorov_example} An example of a reconstruction of Kolmogorov with three different methods using 8 samples per frame of $66\times 66$ pixels, an SPF rate of 0.18\%.}
+\end{figure}
+
+A sample result is displayed on Figure~\ref{fig:kolmogorov_example}.
+The columns represent different timesteps.
+The first row represents the true flow, the next three rows represent our method, FDRI, and Autoencoder methods respectively.
+We acquire 8 samples per frame where each frame consists of $66\times 66=4356$ pixels, which yields a sampling-per-frame rate of 0.18\%.
+We observe that the reconstruction faithfully recovered the signal whereas two other algorithms fail to recover the image.
+The aggregated results on Figure~\ref{fig:cs_kolm_agg} show that other methods require more than 2\% SPF rate for a faithful recovery of the signal, whereas our method can provide a decent reconstruction with SPF as low as 0.18\%, a difference of one order of magnitude in required data intake.
+
+\subsubsection{Methane Leaks Data}
+In this section we use our technique to reconstruct videos of gas leaks using observations from a single-pixel camera.
+We use a subset of GasVid dataset by~\citep{wang2020machine} which consists of recordings of a methane gas leaks.
+10 videos 10 to 20 seconds each.
+We split them into 1-second-long intervals, where each interval contains 10 steps: $T = 1$ sec., $dt = 0.1$ sec.
+Next, we remove backgrounds from each video using Gaussian Mixture-based Background/Foreground Segmentation (MOG2) algorithm by~\citep{zivkovic2006efficient,zivkovic2004improved} implemented in OpenCV library~(\citep{opencv_library}).
+Finally, we split all intervals into four non-overlapping batches: train (390), dev (50), test (49), and SPI (2).
+We use the first three to train, fine-tune, and select a final ROM respectively, and then we use the fourth batch for SPI reconstruction.
+We use the same architecture of ROM as in Section~\ref{sec:kolmogorov_flow}.
+At the recovery step we sample 512 single-pixel observations per each $240 \times 320$ video frame; it corresponds to SPF ratio of $0.66\%$ (2/3 of one percent).
+
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=\textwidth]{figures/cs_gas_examples.pdf}
+	\caption{\label{fig:cs_gas_examples} Results of a Single-Pixel Imaging reconstruction for Methane Leaks Data using Equation~\ref{eq:cs_reconstruction_loss}.
+We trained a ROM (Equation~\ref{eq:cs_loss_data_driven}) using 489 training trajectories, each having 10 video frames 0.1 seconds apart.
+This dataset is a subset of GasVid dataset by \citep{wang2020machine}}.
+\end{figure}
+
+The example reconstructions are presented on Figure~\ref{fig:cs_gas_examples}.
+We see that the algorithm was able to faithfully reconstruct the trajectories.
+Although it was unable to capture the details of turbulent flows, it was able to predict the altitude, the direction, and the spread of the plumes correctly, which are the most important aspects for practical applications in real-world gas monitoring applications.
+
+We note the threefold difficulty of the task.
+First, the background subtraction struggles to separate grey smoke from a dynamic grey background, which leads to fantom clouds of smoke in processed data.
+Second, the amount of training data (3900 frames) is extremely small for training a large convolutional ROM.
+For example, \citep{wang2020machine} used the same dataset for training a simpler classification network (leak/non-leak) using ~11000 frames.
+Finally, an SPF of 0.66\% is vastly insufficient for frame-by-frame reconstruction, as most state-of-the-art methods require ~5-10\% SPF, depending on the nature of the picture (	\citep{wang2022single}).
+Thus, the inductive bias of the ROM becomes a crucial component for a successful video reconstruction.
+
+\subsection{Discussion} 
+In this work, we demonstrated how a collocation-point-based technique can improve the performance of an emerging class of continuous-time physics-informed neural-network-based reduced-order models.
+First, we demonstrated that the incorporation of collocation points in training data can ``cover the gaps'' in training trajectories and inform the model about underrepresented basins of attraction.
+Such an approach alleviates the demand for large volumes of data that is common in network-based models.
+Reducing the amount of data required to faithfully reconstruct a signal is extremely valuable in applications where data is scarce and expensive.
+Second, the physics-informed loss may work as a safeguard, providing a noise-free source of underlying dynamics.
+Third, collocation points can stabilize the model's long-term predictions, allowing for accurate forecasting far beyond the training time horizon.
+Finally, together with using neural-ODE-based nonlinear latent dynamics, adding physics-informed loss leads to the discovery of more compact latent space representations that also yield more accurate models.
+Simultaneous stability and compactness is especially important if one aims to use models with SPI and control algorithms.
+With respect to the computational complexity, we note that adding $Tk$ collocation points to the training imposes less of a computational burden than adding $k$ data trajectories, because, unlike data trajectories, collocation points do not require computing integrals forward in time.
+The last section served to illustrate how PINODE models that were developed in Section~\ref{sec:method} can help in applied modeling.
+In particular, we showed how one can employ a PINODE model for compressive-sensing purposes.
+We illustrated its performance on two examples: Burger's equation and Kolmogorov flow.
+In these cases, it is clear that PINODE can serve as a powerful regularizer and allow users to go significantly below the limits of the Nyquist sampling theorem.
+
+One clear limitation of the current work is that the choice of an efficient collocation family is a design decision that a practitioner makes.
+We believe that such decisions can be automated by adopting existing approaches from classic works on numerical approximations of PDEs, which we leave for future research.
+Another automation that prompts future research is deriving efficient ways of sampling collocation points, possibly via applying modern adaptive learning techniques (\citep{subramanian2022adaptive}).
+Finally, although Section~\ref{sec:burger_noise} provides some rationale for why one may expect robustness of hybrid models under noise, we believe that a more rigorous analysis is possible, particularly one that provides conditions under which such robustness is guaranteed.

### 2023-06-08 introduction.tex
Remove todos
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

@@
-\edit{Add citations all over the intro}
-
-Machine learning has emerged as an important tool in applied mathematics problems as data-driven approaches have begun to outperform  first-principle modeling in the tasks of classification, recognition, autonomous control, protein folding, and even video games.
-However, due to the absence of prior knowledge,  purely data-driven approaches typically require a large amount of data to approximate the behavior of first-principle models even in the simplest scenarios.
-Thus, the most promising modeling approaches of today are those that integrate both data-driven and first-principle components.
+Machine learning has emerged as an important tool in applied mathematics problems as data-driven approaches have begun to outperform  first-principle modeling in the tasks of classification (\cite{lecun1999object}), recognition (\cite{deng2009imagenet,shafiq2022deep}),  autonomous control (\cite{janai2020computer}), protein folding (\cite{jumper2021highly}), and even video games (\cite{chen2016evolution}).
+However, due to the absence of prior knowledge,  purely data-driven approaches typically require a large amount of data to approximate the behavior of first-principle models (\cite{kutz2013data,pan2018long}) .
+Thus, the most promising modeling approaches of today are those that integrate both data-driven and first-principle components (\cite{rackauckas2020udes,raissi2019physics}).
 These latter components frequently stem from domain expertise and pre-existing beliefs.
@@
 Feature selection problems typically arise when one believes that a small subset of features within a dataset captures the majority of the variance of the quantity of interest; this subset of features is commonly known as sparse support.
-The canonical feature selection methods typically provide control over the  sparsity of the solution and output a candidate support for it.
+The canonical feature selection methods, e.g.
+\cite{ibrahim2011fixed,Fan2012}, typically provide control over the  sparsity of the solution and output a candidate support for it.
 In the absence of additional information about the support apart from its probable sparsity, these methods furnish a satisfactory toolkit for practitioners.
@@
 The second example serves to demonstrate how one can incorporate a degree of partial knowledge of physics into a deep neural network.
-Methods for data-driven modeling of physical phenomena have been increasingly popular in large part due to the advancements in GPU computations and automatic differentiation tools.
-These methods have proven particularly useful in reduced-order modeling tasks.
+Methods for data-driven modeling of physical phenomena have been increasingly popular in large part due to the advancements in GPU computations and automatic differentiation tools, such as pytorch (\cite{pytorch}), jax (\cite{jax2018github}), and Neural ODE frameworks (\cite{chen2018neuralode}).
+These methods have proven particularly useful in reduced-order modeling tasks (\cite{rackauckas2020udes}).
 Using data, a neural network can find a representation of a system on a reduced-order manifold via a non-linear transformation from an observable space to that manifold.
-Classic techniques, such as Galerkin Proper Orthogonal Decomposition (POD) or Dynamic Mode Decomposition (DMD), are fundamentally incapable of finding such spaces because these methods are confined to linear, albeit optimal, transformations.
+Classic techniques, such as Galerkin Proper Orthogonal Decomposition (POD) (\cite{fukunaga2013introduction}) or Dynamic Mode Decomposition (DMD) (\cite{schmid2022dynamic}), are fundamentally incapable of finding such spaces because these methods are confined to linear, albeit optimal, transformations (\cite{benner2015survey}).
 However, the exceptional approximation capabilities of deep networks come at the cost of increased instability.
 In particular, the networks tend to identify manifolds that overfit the data and, consequently, fail at extrapolation tasks.
-Such deficiencies compromise the efficacy of the entire system in which the network is utilized, including applications in control, model identification, and compressive sensing.
+Such deficiencies compromise the efficacy of the entire system in which the network is utilized, including applications in control, model identification, and single-pixel imaging.
 
-Recent works have shown that one can improve a neural-network based model by incorporating partial knowledge of physics.
+Recent works, e.g.
+\cite{raissi2019physics,liu2022physics}, have shown that one can improve a neural-network based model by incorporating partial knowledge of physics.
 For example, a model can find a better reduced-order manifold if it knows that it predicts the behavior of a shock wave that obeys Burger's equation.
@@
 The resulting ROMs are able to extrapolate forward in time considerably more effectively,  perform better for unseen initial conditions, and exhibit less sensitivity to noise.
-Finally, in Chapter~\ref{ch:cs} I demonstrate the utility of PINODE models in compressive sensing applications, where they effectively "stitch" together separate timeframes and go below the Nyquist sampling limit.
+Finally, in Chapter~\ref{ch:cs} I demonstrate the utility of PINODE models in applications to single-pixel imaging (SPI), where they effectively "stitch" together separate timeframes and go below the Nyquist sampling limit.
 
@@
 Saleh Nabi.
-The PINODE framework was developed in \cite{sholokhov2023pinode} and is based on our earlier joint work with Dr.
+The PINODE framework was developed in \cite{sholokhov2023pinode} (under peer review at the moment of writing this document) and is based on our earlier joint work with Dr.
 Yuying Liu on Physics-Informed Koopman Networks~(\cite{liu2022physics}).
@@
 Nathan Kutz, and Dr.
-Steven Brunton at the University of Washington.
+Steven Brunton at the University of Washington, and is an ongoing work to date.
 Many parts of the works listed above were taken verbatim to this document per the explicit permission of my co-authors.

### 2023-06-08 msr3-theory.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

@@
 \end{proof}
-For the remainder of the paper, we assume that 
+For the remainder of the work, we assume that 
 \eq{\label{eq:blam1}\eta> \nu m=:\bar\eta} 
@@
 Hence, the consequences of \cite[Theorem 10.15]{AB17} hold.
-By translating the notion of \cite[Theorem 10.15]{AB17} to that of this paper,
+By translating the notion of \cite[Theorem 10.15]{AB17} to that of this work,
 we obtain the result.
@@
 then leave it fixed over all iterations. Our method for choosing $\eta$ is described
-in \cite[Section 4, Figure 5]{Practice}. Briefly, we maximize the 
+in \cite[Section 4, Figure 5]{sholokhov2022relaxation}. Briefly, we maximize the 
 Baysian Information Criterion (BIC) over a grid of values for $\eta$. 

### 2023-06-08 msr3.tex
Remove todos
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

@@
-% \edit{Rewrite the MSR3 abstract into a summary of whole work with citations}
 \paragraph{Abstract for Chapter~\ref{ch:msr3}} Linear Mixed-Effects (LME) models are a fundamental tool for modeling correlated data, 
@@
 The computational framework only requires the proximal operator for each regularizer to be available, and the implementation is available in an open source \texttt{python} package \texttt{pysr3}, consistent
-with the \texttt{sklearn} standard. 
+with the \texttt{scikit-learn} standard. 
 The numerical results on simulated data sets indicate that the proposed strategy improves on the state of the art for both accuracy and compute time. 
@@
 accelerates and improves the performance of regularization strategies.
-This extension and its mathematical foundations in \cite{Theory1} constitute
+This extension and its mathematical foundations in \cite{aravkin2022jimtheory} constitute
 the major innovations of this work.
@@
 mathematical 
-foundations are presented in \cite{Theory1}.
+foundations are presented in \cite{aravkin2022jimtheory}.
 The SR3 framework introduces auxiliary variables $x$ and $w$ to 
@@
 However, it is convex with respect to $\beta$, and weakly convex with respect to $\gamma$, with a weak convexity constant $\overline \eta$
-computed in~\cite[Section 5.1]{Theory1} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation
+computed in~\cite[Section 5.1]{aravkin2022jimtheory} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation
 
@@
 and, more generally, \eqref{eq:ml_lme_optimization_setup} follows from the work of \cite{zheng2021trimmed}.
-Standalone proofs for the existence of minimizers are developed in~\cite[Theorem 1]{Theory1},
-and extended to the presence of regularizers in~\cite[Theorem 2]{Theory1}. %, restated below for convenience.
+Standalone proofs for the existence of minimizers are developed in~\cite[Theorem 1]{aravkin2022jimtheory},
+and extended to the presence of regularizers in~\cite[Theorem 2]{aravkin2022jimtheory}. %, restated below for convenience.
 
@@
 %
-%Feature selection methods  include sparsity-inducing penalties or constraints. See~\cite[Theorem 1]{Theory1}
+%Feature selection methods  include sparsity-inducing penalties or constraints. See~\cite[Theorem 1]{aravkin2022jimtheory}
 %for an existence result under weak assumptions on the regularizer. 
@@
 %\]  
-By \cite[Theorem 2]{Theory1}, solutions to \eqref{eq:lme_loss_original_in_x}
+By \cite[Theorem 2]{aravkin2022jimtheory}, solutions to \eqref{eq:lme_loss_original_in_x}
 always exist when $R$ has compact lower level sets.
@@
 with $\eta\ge\bar\eta$ where $\bar \eta$ is the weak convexity constant
-computed in~\cite[Section 5.1]{Theory1}.
+computed in~\cite[Section 5.1]{aravkin2022jimtheory}.
 As $\eta \uparrow \infty$, the extended objective~\eqref{eq:lme_loss_sr3} converges in an epigraphical sense to the original objective~\eqref{eq:lme_loss_original_in_x}. 
@@
 in order to optimize~\eqref{eq:value_function_optimization}. 
-Indeed, in \cite[Section 5]{Theory1} it is shown that 
+Indeed, in \cite[Section 5]{aravkin2022jimtheory} it is shown that 
 there exists a computable $\bar\eta>0$, which we have called the weak convexity constant, such that
@@
 }
-The results in \cite{Theory1} show that all components of the iteration~(\ref{eq:pgd_for_value_function}) are well-defined. 
+The results in \cite{aravkin2022jimtheory} show that all components of the iteration~(\ref{eq:pgd_for_value_function}) are well-defined. 
 The equivalence of Algorithm \ref{alg:pgd_for_value} and 
@@
 
-In \cite[Theorem 6]{Theory1}, it is shown that
+In \cite[Theorem 6]{aravkin2022jimtheory}, it is shown that
 for any sequence $\eta_k\uparrow\infty$ the associated optimal solutions $(x^k,w^k)$ 
@@
 Therefore, one must take care in selecting $\eta$. For this, we recall from
-\cite[Lemma 3]{Theory1} that
+\cite[Lemma 3]{aravkin2022jimtheory} that
 \[
@@
 
+\newpage
 \section{Theoretical Analysis}
@@
 
+\newpage
 \section{Software Implementation}
@@
 This library implements functionality for fitting linear mixed models and selecting covariates.
-The user interface was designed to be fully compliant with the standards\footnote{\href{https://scikit-learn.org/stable/developers/develop.html}{https://scikit-learn.org/stable/developers/develop.html}} of \texttt{sklearn} library to minimize learning time.
+The user interface was designed to be fully compliant with the standards\footnote{\href{https://scikit-learn.org/stable/developers/develop.html}{https://scikit-learn.org/stable/developers/develop.html}} of \texttt{scikit-learn} library to minimize learning time.
 

### 2023-06-08 phd_thesis.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

@@
 \begin{document}
-
+\thispagestyle{empty}
 \title{\doctitle}
@@
 \userName
-\\[20mm]
+\\[15mm]
 {A dissertation \\
  submitted in partial fulfillment of the\\ requirements for the degree of}
+\\[10mm]
+{Doctor of Philosophy}
+\\[10mm]
+\institution
 \\[5mm]
-{Doctor of Philosophy}
-\\[20mm]
+2023
+\\[15mm]
 {\center
-\textbf{Reading Committee}\\
-Aleksandr Aravkin (co-chair),\\
-J. Nathan Kutz (co-chair), \\
-James Burke, \\
-Jonathan Wakefield (GSR)
+Reading Committee:\\
+Aleksandr Aravkin (chair),\\
+Jose Kutz (chair), \\
+James Burke
 }
-\\[20mm]
+\\[15mm]
 Program Authorized to Offer Degree:
@@
 \\[5mm]
-\institution
 }
-\\[10mm]
+}
+
+\newpage
+\thispagestyle{empty}
+{
+\center
+
 ©Copyright 2023\\
@@
 \newpage
+\thispagestyle{empty}
 {
@@
 Conveniently, this package offers complete compatibility with \texttt{scikit-learn}, so all \texttt{pysr3} models can be used in a pipeline with classic modeling blocks such as data pre-processors, randomized grid search, cross-validation, and quality metrics.
-
 The second line of work develops a framework for training Reduced-Order Models (ROMs) with Physics-Informed Neural Ordinary Differential Equations (PINODE).
@@
 
+\newpage
 \tableofcontents
 
-\chapter{Introduction}
+\chapter*{Introduction}
 \include{introduction}
@@
 
+\addtocontents{toc}{\protect\newpage}
 \chapter{PINODE: Physics-Informed Neural Ordinary Differential Equations}
@@
 
-\chapter{Conclusion}
+\chapter*{Conclusion}
 \include{conclusion}
@@
 
+\addtocontents{toc}{\protect\newpage}
+
 \appendix
-\chapter{Appendix}
-\section{Acknowledgement}
+\chapter{MSR3}
+
+\section*{Acknowledgement}
 \input{appendix/acknowledgements}
@@
 
+\chapter{PINODE}
 \section{Detailed Design of Experiments for PINODE}

### 2023-06-08 pinode.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

@@
         Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs).
-        High-fidelity simulations of such dynamical systems can become intractable, especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}.
-A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}.
-At the time of writing this document, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation}.
-Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}
+        High-fidelity simulations of such dynamical systems can become intractable, especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \citep{rowley2017model,lucia2004reduced,benner2015survey}.
+A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \citep{farahmand2016learning,nabi2022robust}.
+At the time of writing this document, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \citep{otterness2017evaluation}.
+Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \citep{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}
 
-        In order to enable control of high dimensional dynamical systems, a ROM training method needs to  identify a low-dimensional manifold along with dynamics on the manifold that together yield high-accuracy predictions and long-term stability \cite{ahmed2021closures,noack2011reduced}.
+        In order to enable control of high dimensional dynamical systems, a ROM training method needs to  identify a low-dimensional manifold along with dynamics on the manifold that together yield high-accuracy predictions and long-term stability \citep{ahmed2021closures,noack2011reduced}.
 Most traditional ROMs are projection-based, e.g.
-dynamic mode decomposition (DMD) \cite{kutz2016dynamic,tu2013dynamic} and proper orthogonal decomposition (POD) \cite{holmes2012turbulence}, which transform the trajectories of a high-dimensional dynamical system into a suitable, and in some sense optimal, low-dimensional subspace.
-This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of the basis functions (spatial modes) \cite{benner2015survey}.
+dynamic mode decomposition (DMD) \citep{kutz2016dynamic,tu2013dynamic} and proper orthogonal decomposition (POD) \citep{holmes2012turbulence}, which transform the trajectories of a high-dimensional dynamical system into a suitable, and in some sense optimal, low-dimensional subspace.
+This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of the basis functions (spatial modes) \citep{benner2015survey}.
 One challenge for POD methods is their intrusive nature, i.e.
 requiring access to the solver codes.
-To overcome this, operator inference approaches \cite{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and  exploit lifting to fit the latent space dynamics data into polynomial, typically quadratic, models.
+To overcome this, operator inference approaches \citep{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and  exploit lifting to fit the latent space dynamics data into polynomial, typically quadratic, models.
 These models, however, are (i) limited in representation power (up to quadratic, e.g.
@@
 
-        In a thrust to overcome these challenges, significant effort has been invested into developing autoencoder-based reduced-order models, as a popular nonlinear ROM technique, which can yield both accurate and stable ROMs \cite{lee2020model,gin2021deep,champion2019data,kim2019deep}.
+        In a thrust to overcome these challenges, significant effort has been invested into developing autoencoder-based reduced-order models, as a popular nonlinear ROM technique, which can yield both accurate and stable ROMs \citep{lee2020model,gin2021deep,champion2019data,kim2019deep}.
 In practice, however, autoencoder-based ROMs require datasets that densely cover a hypothetical infinite dimensional phase portrait of the dynamical system.
@@
 
-        Another severe challenge of utilizing ROMs comes from their poor out-of-distribution performance \cite{fries2022lasdi,cranmer2020discovering,gin2021deep}, especially when it is fundamentally impossible for a practitioner to obtain data that covers the entire distribution of possible data inputs.
+        Another severe challenge of utilizing ROMs comes from their poor out-of-distribution performance \citep{fries2022lasdi,cranmer2020discovering,gin2021deep}, especially when it is fundamentally impossible for a practitioner to obtain data that covers the entire distribution of possible data inputs.
 For example, in HVAC applications, one may collect data from a room with two windows but not from a room for every possible number of windows.
-In atmospheric LiDAR applications, we may conduct experiments on a certain terrain but we can never conduct experiments on all sorts of terrains \cite{nabi2020improving}.
+In atmospheric LiDAR applications, we may conduct experiments on a certain terrain but we can never conduct experiments on all sorts of terrains \citep{nabi2020improving}.
 In such situations embedding the knowledge of physics into a model becomes necessary to improve the extrapolation performance, and for which several approaches have recently been proposed.
-        For instance, the seminal works \cite{bongard2007automated,schmidt2009distilling} have tried to determine the underlying structure of a nonlinear dynamical system from data using symbolic regression.
-Recently, Cranmer et al.
-\cite{cranmer2020discovering} employed symbolic regression in conjunction with graph neural network (GNN), while encouraging sparse latent representation, to extract explicit physical relations.
+        For instance, the seminal works \citep{bongard2007automated,schmidt2009distilling} have tried to determine the underlying structure of a nonlinear dynamical system from data using symbolic regression.
+Recently \citep{cranmer2020discovering} employed symbolic regression in conjunction with graph neural network (GNN), while encouraging sparse latent representation, to extract explicit physical relations.
 They showed that the symbolic expressions extracted from the GNN generalized to out-of-distribution data better than the GNN itself.
@@
         Another example of incorporating physics in ROMs is the use of parametric models at the latent space, e.g.
-by using the sparse identification of nonlinear dynamics (SINDy) \cite{brunton2016discovering,champion2019data}.
-For instance, \cite{fries2022lasdi,he2022glasdi} used a chain-rule based loss that ties latent-space derivatives to the observable-space derivatives for simultaneous training of the auto-encoder and the latent dynamics.
-However, such loss is highly sensitive to noise in the data, especially when evaluating time-derivatives with finite differences is required\cite{delahunt2022sindynoise}.
+by using the sparse identification of nonlinear dynamics (SINDy) \citep{brunton2016discovering,champion2019data}.
+For instance, \citep{fries2022lasdi,he2022glasdi} used a chain-rule based loss that ties latent-space derivatives to the observable-space derivatives for simultaneous training of the auto-encoder and the latent dynamics.
+However, such loss is highly sensitive to noise in the data, especially when evaluating time-derivatives with finite differences is required\citep{delahunt2022sindynoise}.
         Collocation-based enforcement of the physics, i.e.
@@
 Recently, Liu et al.
-\cite{liu2022physics} used an auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM.
-However, linearity may not be expressive enough for complex dynamics with multiple basins of attraction \cite{page2019koopman}.
-        Finally, recent works on NeuralODE (NODE) \cite{chen2018neuralode,rackauckas2020udes} show a way to fit an arbitrary non-linear model (e.g.
+\citep{liu2022physics} used an auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM.
+However, linearity may not be expressive enough for complex dynamics with multiple basins of attraction \citep{page2019koopman}.
+        Finally, recent works on NeuralODE (NODE) \citep{chen2018neuralode,rackauckas2020udes} show a way to fit an arbitrary non-linear model (e.g.
 a network) as a latent space dynamics model, significantly extending the set of models for the latent dynamics that one can train efficiently.
@@
 In real-world applications, it is often expensive to solve equation~(\ref{eq:generic_stationary_ode}) directly because $x(t)$ can be very high-dimensional.
-However, a variety of works provided both theoretical \cite{holmes2012turbulence} and empirical \cite{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$.
+However, a variety of works provided both theoretical \citep{holmes2012turbulence} and empirical \citep{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$.
 In that space, the dynamics evolve according to a (generally unknown) function  $\bd{h}(\bd{z})$:
@@
 by utilizing a remote cluster), executing control over the resulting model, which is an end-goal for an HVAC system, may still be intractable.
-Indeed, executing control requires \textit{multiple} evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\cite{duriez2017machine}.
+Indeed, executing control requires \textit{multiple} evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\citep{duriez2017machine}.
 
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{figures/abstract_data_driven.png}
+    \includegraphics[width=\textwidth]{figures/abstract_data_driven.png}
     \caption{Illustration of the autoencoer structure with neural ODE in the latent space.
@@
 
-\paragraph{Data-Driven Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses.
+\paragraph{Data-Driven Loss} Similar to prior works \citep{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses.
 The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data, as illustrated on Figure~\ref{fig:data_driven_loss}.
@@
 We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and may use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}).
-This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}.
+This part is handled by \texttt{torchdiffeq}~\citep{chen2018neural} library, which supports non-uniform time-frames within a batch}.
 To simplify the notation, without loss of generality, in the rest of the section we assume that all trajectories are recorded over the same time-frame with the same uniform step-size.
-To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs)\cite{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}).
+To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs by~\citep{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}).
 Neural ODEs have demonstrated a better ability to model highly non-linear dynamics compared to linear models when the dimensionality of the dynamics variable is limited.
-This is especially useful in applications where the size of the latent space dimension needs to be small~\cite{lee2020model,gin2021deep,champion2019data,kim2019deep}.
+This is especially useful in applications where the size of the latent space dimension needs to be small~\citep{lee2020model,gin2021deep,champion2019data,kim2019deep}.
 
-\paragraph{Physics-Informed Loss} In their recent work, Liu et al.~\cite{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics.
+\paragraph{Physics-Informed Loss} In their recent work, Liu et al.~\citep{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics.
 To extend this approach to the non-linear regime, we note that for a true mapping $\phi$ the following holds:
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{figures/abstract_physics_informed.png}
+    \includegraphics[width=\textwidth]{figures/abstract_physics_informed.png}
     \caption{The physics-informed loss function compares gradient fields in the current latent space with what a correctly-learned field should be in this latent space on set of collocation points.}
@@
 
-The above definition of collocation points is not to be confused with a classic notion of collocation points for finding numerical solutions for differential equations~\cite{fornberg1998practical, trefethen2022numerical}.
+The above definition of collocation points is not to be confused with a classic notion of collocation points for finding numerical solutions for differential equations~\citep{fornberg1998practical, trefethen2022numerical}.
 The classic notion refers to a set of points in time $[t_0, t_0 + c_1h, t_0 + c_2h, \dots, t_0 + h]$, $0 < c1 < c2 < \dots < 1$ which are chosen to obtain an optimal local interpolant of a solution of a differential equation for a time-period between $t_0$ and $t_0 + h$.
@@
 In contrast, we define collocation points as pairs $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$ which are  examples of mapping $x \to f(x)$.
-Our definition is built around solving an \textit{inverse} problem of approximating $\dot{x} = f(x)$ with $f_\theta(x)$ and follows a recent work~\cite{liu2022physics} which develops upon a definition from \cite{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.
+Our definition is built around solving an \textit{inverse} problem of approximating $\dot{x} = f(x)$ with $f_\theta(x)$ and follows a recent work~\citep{liu2022physics} which develops upon a definition from \citep{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.
 
@@
 
-We use a \texttt{pytorch}~\cite{NEURIPS2019_9015} implementation of the Adam algorithm~\cite{kingma2014adam} for optimization.
-To evaluate $\nabla_\theta \LL^{physics}_\theta$ and $\nabla_\theta \LL^{data}_\theta$ we use \texttt{torchdiffeq}~\cite{chen2018neural} -- a \texttt{pytorch}-compatible implementation of the Neural ODE framework.
+We use a \texttt{pytorch}~\citep{NEURIPS2019_9015} implementation of the Adam algorithm~\citep{kingma2014adam} for optimization.
+To evaluate $\nabla_\theta \LL^{physics}_\theta$ and $\nabla_\theta \LL^{data}_\theta$ we use \texttt{torchdiffeq}~\citep{chen2018neural} -- a \texttt{pytorch}-compatible implementation of the Neural ODE framework.
 
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{figures/duff_first_exp_abstract.png}
+    \includegraphics[width=\textwidth]{figures/duff_first_exp_abstract.png}
     \caption{We use a toy example -- a Lifted Duffing Oscillator -- to show that it is possible to ``fill the gaps'' in data with collocation points.
@@
 First, to illustrate the ideas behind the framework we study its performance on a high-dimensional ODE -- a lifted Duffing oscillator.
-We show how a non-linear latent dynamics $\bd{h}(\bd{z})$ overcomes the limitations of DMD and Koopman networks from \cite{liu2022physics} by handling multiple basins of attraction within one model.
+We show how a non-linear latent dynamics $\bd{h}(\bd{z})$ overcomes the limitations of DMD and Koopman networks from \citep{liu2022physics} by handling multiple basins of attraction within one model.
 We also show that using physics-informed loss is sufficient for reconstructing the behaviour for basins of attraction that are not represented by the data.
@@
 
-\begin{wrapfigure}{r}{0.25\textwidth}
-  \begin{center}
-    \includegraphics[width=0.23\textwidth]{figures/duffing_comparison.pdf}
-  \end{center}
-  \caption{Non-linearity in the latent dynamics and the autoencoder employed in teh PINODE Hybrid model are important for accurate long-term extrapolation.
-The DMD model and PIKN Hybrid model were unable to extrapolate the dynamics from collocation points.}%\vspace{-0.5in}
-  \label{fig:duffing_comparison}
-\end{wrapfigure}
 A phase portrait for 300 randomly sampled trajectories from this system is visualized on Figure~\ref{fig:duffing_pinode}, left frame.
 Depending on the total energy, each trajectory always stays in one of three regions: the left lobe, the right lobe, or the outer area, visualized in red, green, and blue, respectively.
-To create a synthetic high-dimensional system that retains this property, we lift the Duffing trajectories into a higher-dimensional space by applying an invertible transformation $\mathcal{A}(\bd{z})$:
+To create a synthetic high-dimensional system that retains this property, we lift the Duffing trajectories into a higher-dimensional space by applying an invertible transformation~$\mathcal{A}(\bd{z})$:
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{figures/duffing_comparison.pdf}
+  \caption{\label{fig:duffing_comparison}Non-linearity in the latent dynamics and the autoencoder employed in teh PINODE Hybrid model are important for accurate long-term extrapolation.
+The DMD model and PIKN Hybrid model were unable to extrapolate the dynamics from collocation points.}%\vspace{-0.5in}
+\end{figure}
 
@@
 \end{equation}
+
 Thus, we created a high-dimensional dynamical system with multiple basins of attraction for which the dynamics $\bd{f}$ are known.
@@
 First, a purely data-driven model is unable to extrapolate outside its training region using only the data from that region.
-This observation is consistent with the conclusions from related works~\cite{gin2021deep} that neural networks interpolate well but struggle with extrapolation tasks.
+This observation is consistent with the conclusions from related works~\citep{gin2021deep} that neural networks interpolate well but struggle with extrapolation tasks.
 Second, we see that collocation points provided enough extra information for the model to predict nearly perfectly in regions from which no trajectories were provided.
@@
 The ability of Neural ODE to model nonlinear dynamics in the latent space is demonstrated in Figure~\ref{fig:duffing_comparison}.
-The figure shows a comparison between the Hybrid PINODE model, the Hybrid PIKN model~\cite{liu2022physics}, and DMD, all of which have been trained using the same dataset.
+The figure shows a comparison between the Hybrid PINODE model, the Hybrid PIKN model~\citep{liu2022physics}, and DMD, all of which have been trained using the same dataset.
 PIKN differs from PINODE in that it uses linear latent dynamics $\frac{dz}{dt} = Lz$, where $L$ is a finite-dimensional approximation of the Koopman operator, instead of a general non-linear dynamics operator $\frac{dz}{dt} = h_\theta(z)$.
@@
 where $u_t$, $u_x$, and $u_{xx}$ represent partial derivatives in time, the first, and second spatial derivatives, respectively.
-Burgers' equation is a PDE occurring in applications in acoustics, gas and fluid dynamics, and traffic flows \cite{burgers1948mathematical}.
+Burgers' equation is a PDE occurring in applications in acoustics, gas and fluid dynamics, and traffic flows \citep{burgers1948mathematical}.
 When $\nu$ is significantly smaller than one, the system exhibits strong non-linear behaviour and is called ``advection-dominated'', otherwise when $\nu$ is large the system is called ``diffusion-dominated''.
-In the case of the former, linear projection methods such as POD become inaccurate as the true solution space has a slow decaying Kolmogorov n-width, manifesting itself in slow decaying singular values \cite{peherstorfer2022breaking}.
+In the case of the former, linear projection methods such as POD become inaccurate as the true solution space has a slow decaying Kolmogorov n-width, manifesting itself in slow decaying singular values \citep{peherstorfer2022breaking}.
 Therefore, in this section we focus on the advection-dominated Burgers' equation for which we set $\nu = 0.01$.
 
-To generate trajectories, we discretize the spatial domain $[-\pi,\,\pi]$ into 128 grid-points, and solve Equation~\ref{eq:burgers_equation} for $t \in [0, 2]$ with $\Delta t = 0.1$ using a spectral solver~\cite{trefethen2000spectral}.
+To generate trajectories, we discretize the spatial domain $[-\pi,\,\pi]$ into 128 grid-points, and solve Equation~\ref{eq:burgers_equation} for $t \in [0, 2]$ with $\Delta t = 0.1$ using a spectral solver~\citep{trefethen2000spectral}.
 To generate a diverse set of initial conditions we sum the first 10 harmonic terms with random coefficients:
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{figures/compressibility.pdf}
+    \includegraphics[width=\textwidth]{figures/compressibility.pdf}
     \caption{PINODE Hybrid model utilized the latent space dimension 5 times more efficiently in terms of MSE than PIKN Hybrid model when modelling low-viscosity (highly-nonlinear) Burgers' equation (left frame).
@@
 First, we notice that DMD does not perform well on the test data, despite achieving a training loss ($\sim 10^{-3}$).
-This observation is consistent with earlier works~(\cite{kalur2021robust,kutz2016dynamic}); and illustrates well that a combination of a linear encoder and a linear latent dynamics operator may not be sufficient for modelling highly-nonlinear phenomena.
+This observation is consistent with earlier works~(\citep{kalur2021robust,kutz2016dynamic}); and illustrates well that a combination of a linear encoder and a linear latent dynamics operator may not be sufficient for modelling highly-nonlinear phenomena.
 Second, we notice that PINODE achieves better performance for a given latent space dimension compared to PIKN.
@@
 This is a consequence of the latent-dynamics matrix ($h(z) = Lz$) of PIKN having eigenvalues with positive real parts, which implies long-term instability (Figure~\ref{fig:burgers_compressibility}, right pane).
-Although there has been progress in the literature~\cite{kojimalearning}, further research is needed to understand (i) how to enforce stability constraints for PIKN, and (ii) why one does not need the same enforcement for PINODE to exhibit stable behaviour.
+Although there has been progress in the literature~\citep{kojimalearning}, further research is needed to understand (i) how to enforce stability constraints for PIKN, and (ii) why one does not need the same enforcement for PINODE to exhibit stable behaviour.
 
-	\begin{figure}[ht!]
+	\begin{figure}[t]
         \centering
@@
 In this section, we show that, when chosen appropriately, collocation points can be effectively used for training a model in the low-data regime, and their contribution to a model's accuracy may even surpass the contribution of the data.
-\begin{figure}[ht]
+\begin{figure}[t]
     \centering
@@
     \label{fig:burgers_examples_of_ics}
-\end{figure}
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.9\textwidth]{figures/data_vs_collocations.pdf}
-    \caption{Comparison of the achievable MSE relative to the full data regime (1024 trajectories).
-When the data is scarce, collocations-based physics-informed loss improves the forecasting accuracy of ROMs by an average of 5 times lower MSE compared to the data-only regime, as shown in this experiment with Burgers' equation.
-When other types of initial conditions (``harmonic'', ``bell-curve'') are used, the physics-only model (top-right corner of the right frame) outperformed the most data-rich model in our experiment (bottom-left corner).
-}
-    \label{fig:burger_data_vs_collocations}
 \end{figure}
@@
 The PINODE models all use a latent space dimension $m=16$.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/data_vs_collocations.pdf}
+    \caption{Comparison of the achievable MSE relative to the full data regime (1024 trajectories).
+When the data is scarce, collocations-based physics-informed loss improves the forecasting accuracy of ROMs by an average of 5 times lower MSE compared to the data-only regime, as shown in this experiment with Burgers' equation.
+When other types of initial conditions (``harmonic'', ``bell-curve'') are used, the physics-only model (top-right corner of the right frame) outperformed the most data-rich model in our experiment (bottom-left corner).
+}
+    \label{fig:burger_data_vs_collocations}
+\end{figure}
 
@@
 Thus, one can think about optimizing a hybrid model~(\ref{eq:loss_combined}) as about training a Physics-Informed model~(\ref{eq:loss_physics_informed}) using a noisy gradient descent with a fixed-variance noise.
-From the optimization literature \cite{friedlander2012hybrid,patel2021global,shapiro2021lectures} we know that, under certain conditions, such SGD converges to a neighbourhood of a local minimum of its loss (in this case $\LL_{\theta}^{physics}$) with high probability.
+From the optimization literature \citep{friedlander2012hybrid,patel2021global,shapiro2021lectures} we know that, under certain conditions, such SGD converges to a neighbourhood of a local minimum of its loss (in this case $\LL_{\theta}^{physics}$) with high probability.
 So instead of diverging, a hybrid model turns into a Physics-Informed model; where the latter works as a performance safeguard in the high-noise regime.

### 2023-06-08 presentation/iccopt22.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

--- 2023-04-06 presentation/pysr3.tex
@@
+\documentclass[8pt]{beamer}
+\input{useful_packages}
+\input{math_macros}
+
+% \usetheme[progressbar=frametitle]{metropolis}
+\usepackage{appendixnumberbeamer}
+\addbibresource{../bibliography.bib}
+
+% PRESETS OF PACKAGES AND MATH %
+
+\usepackage{xspace}
+\usepackage{caption}
+
+\usepackage{subcaption}
+\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
+\newcommand{\ouralgo}{\ensuremath{\mathcal{MSR}3}}
+\title{A Relaxation Approach to Feature Selection \newline for Linear Mixed-Effects Models}
+\date{\today}
+\author{Aleksei Sholokhov, James V. Burke, Peng Zheng, Damian Santomauro, and Aleksandr Aravkin}
+\titlegraphic{\includegraphics[width=0.5\textwidth]{Figures/UW_amath.pdf}\hspace{3em}\includegraphics[height=1cm]{Figures/IHME_Logo.jpg}}
+
+\begin{document}
+
+\maketitle
+
+% pysr3 part
 \section{Linear Mixed-Effect Models}
@@
 
-Sparse Relaxed Regularized Regression ($\mathcal{SR}3$) \cite{Zheng2018RelaxAndSplit} showed great results for t linear models:
+Sparse Relaxed Regularized Regression ($\mathcal{SR}3$) \footcite{Zheng2018RelaxAndSplit} showed great results for t linear models:
 
@@
 \begin{itemize}
-	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \cite{Bondell2010,Lin2013}
-	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \cite{Vaida2005,Jones2011}
-	\item $R(x)$ -- SCAD penalty (\cite{Fan2001,Fan2012})
+	\item $R(x) = \lambda\sum_{j=1}^p w_j\|x_j\|_1$ -- LASSO and Adaptive LASSO penalties \footcite{Krishna2008,Lin2013}
+	\item $R(x) = \lambda \|x\|_0$ -- $\ell_0$ penalty \footcite{Vaida2005,Jones2011}
+	\item $R(x)$ -- SCAD penalty (\footcite{Fan2001,Fan2012})
 \end{itemize}
@@
 %	\includegraphics[width=\textwidth]{Figures/intuition_prev_paper}
-%	\caption{\label{fig:intuition_prev} Picture from \cite{Zheng2018RelaxAndSplit}: for a linear problem, value function relaxation ``squashes'' level-sets simplifying the optimization landscape.}
+%	\caption{\label{fig:intuition_prev} Picture from \footcite{Zheng2018RelaxAndSplit}: for a linear problem, value function relaxation ``squashes'' level-sets simplifying the optimization landscape.}
 %\end{figure}
@@
 		\includegraphics[width=\textwidth]{Figures/bullying_data_assessment_selection.pdf}
-		\caption{Fixed and random covariate selection for Bullying dataset from~\cite{GBD}.
+		\caption{Fixed and random covariate selection for Bullying dataset from~\footnote{Institute for Health Metrics and Evaluation (IHME).
+Bullying Victimization Relative Risk Bundle GBD 2020.
+Seattle, United States of America (USA), 2021.}.
 The model selected 9 covariates, 7 of which were historically significant, and did not select 4 covariates, 1 of which was historically significant.}
@@
 \end{frame}
+
+\begin{frame}[noframenumbering,plain,allowframebreaks]{References}
+    \printbibliography[heading=none]
+\end{frame}
+
+\end{document}

### 2023-06-08 presentation/phd_defense_pres.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

@@
 \usepackage{subcaption}
-\usepackage{lmodern}
+
 \newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
@@
 
-\section{$\mathcal{MSR}3$ -- Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}
+\section{\texorpdfstring{$\mathcal{MSR}3$}{MSR3} -- Sparse Relaxed Regularized Regression for Linear Mixed-Effect Models}
 
@@
 
-\begin{frame}[label=again]{Results: Kolmogorov Flow}
+\begin{frame}{Results: Kolmogorov Flow}
 	A 2D velocity field $\bd{u}(x, y, t)$:
@@
 	\begin{align}
-	\label{eq:kolmogorov_flow_definition}
 	& \partial_t \bd{u} + \bd{u} \cdot \nabla \bd{u} = -\nabla p + \nu\nabla^2\bd{u} + f \\
@@
 
-\againframe{again}
+%\againframe{again}
 
@@
 \appendix
-\section{Appendix: $\mathcal{MSR}3$}
+\section{Appendix: \texorpdfstring{$\mathcal{MSR}3$}{MSR3}}
 
